\documentclass{article}
\usepackage[a6paper, width=166pt, height=280pt]{geometry}
\usepackage{shandean}
\title{The Life and Opinions of Tristram Shandy, Gentleman}
\author{Laurence Sterne}
\date{Vol 8}
\begin{document}
\pagestyle{empty}
\enlargethispage{72pt}
\[\vbox{\openup 12pt\halign{#\cr
\noalign{\vskip -20pt}
\hss\large{\addfontfeature{LetterSpace=20.0}\textsc{the}}\hss\cr
\hss\Large{\addfontfeature{LetterSpace=120.0}LIFE}\hss\cr
\hss\large{\addfontfeature{LetterSpace=20.0}\textsc{and}}\hss\cr
\hss\Large{\addfontfeature{LetterSpace=60.0}OPINIONS}\hss\cr
\hss\large{\addfontfeature{LetterSpace=20.0}\textsc{of}}\hss\cr
\hss\Large{\addfontfeature{LetterSpace=32.0}TRISTRAM SHANDY},\hss\cr
\hss\large{\addfontfeature{LetterSpace=26.0}\textsc{Gentleman}}\hss\cr
\small\textit{Non enim excursus hic ejus, sed opus ipsum est}.\hss\cr
\noalign{\vskip -12pt}
\hss\small\textsc{Plin}.\@ Lib.\@ quintus Epistola sexta.\cr
\hss\large{\addfontfeature{LetterSpace=30.0}VOL}. VIII.\hss\cr
\noalign{\vskip 6pt}
\hss\large{\addfontfeature{LetterSpace=30.0}LONDON}:\hss\cr
\noalign{\vskip -10pt}
\hss Printed for \textsc{T.\@ Becket} and \textsc{P.\@ A.\@ Dehondt},\hss\cr
\noalign{\vskip -14pt}
\hss in the Strand. M\,DCC\,LXV.\hss\cr
}}\]

\newpage\null
\cleardoublepage
\setcounter{page}{1}
\pagestyle{fancy}
\thispagestyle{empty}
\hbox{}\vskip -36pt

\[\vbox{\openup 12pt\halign{\hss # \hss\cr
{\addfontfeature{LetterSpace=12.0}\textsc{the}}\cr
{\addfontfeature{LetterSpace=12.0}LIFE and OPINIONS}\cr
{\addfontfeature{LetterSpace=12.0}\textsc{of}}\cr
{\addfontfeature{LetterSpace=16.0}TRISTRAM SHANDY}, Gent.\cr}}\]

\vskip 12pt
\hrule
\sloppy
\setlength{\baselineskip}{14pt}

\section{\lsss{CHAP}.\enspace I.}

\lettrine{\Tsk B}{ut} softly\tsk for in these sportive plains,
and under this genial sun, where at this instant all flesh is
running out piping, fiddling, and dancing to the vintage, and
every step that’s taken, the judgment is surprised~by the
imagination, I defy, notwithstanding\pb 
all that has been said upon \textit{straight
lines}~\footnote{Vid.\@ Vol.\@ VI. p.\@ 152} in sundry pages of
my book\tsk I defy the best cabbage planter that ever existed,
whether he plants backwards or forwards, it makes little
difference in the account (except that he will have more to
answer for in the one case than in the other)\tsk
I defy him to go on coolly, critically, and canonically,
planting his cabbages one by one, in straight lines, and stoical
distances, especially if slits in petticoats are unsew’d up\tsk
without ever and anon straddling out, or sidling into some
bastardly digression\tsh In \textit{Freeze-land},
\textit{Fog-land}, and some other lands I wot of\tsk it may be done\tsh

But in this clear climate of fantasy\break and perspiration, where
every idea,\pb
sensible and insensible, gets vent\tsk in this land,
my dear Eugenius\tsk in this\break fertile land of chivalry and
romance, where I now sit, unskrewing my ink-horn to write my uncle
Toby’s amours, and with all the meanders of
\textsc{Julia}’s track in quest of her
\textsc{Diego}, in full view of my study window\tsk if thou
comest not and takest me by the hand\tsh

What a work it is likely to turn out!

Let us begin it.




\newpage
\section{\lsss{CHAP}.\enspace II.}

\lettrine{I}{t} is with \textsc{love} as
with \textsc{cuck}\textsc{oldom}\tsh

\bigskip
\tsh But now I am talking of beginning a book, and have long had a
thing upon my mind to be imparted to the reader, which, if not
imparted now, can never be imparted to him as long as I live
(whereas the \textsc{comparison} may be imparted to him any
hour in the day)\tsh I’ll just mention it, and begin
in good earnest.

The thing is this.

That of all the several ways of beginning a book which are now
in practice\pb throughout the known world, I am confident my own way
of doing it is the\break best\tsh I’m sure it is the most
religious \hbox{}\tsh for I begin with writing the first
sentence\tsh and trusting to Almighty God for the
second.

’Twould cure an author for ever of\break
the fuss and folly of opening his street-door,
and calling in his neighbours and friends, and kinsfolk, with the devil and all his
imps, with their hammers and engines, \etc only to observe how one sentence of mine
follows another, and\break how the plan follows the whole.

I wish you saw me half starting out of my chair, with what confidence, as I grasp
the elbow of it, I look up\tsh\pb catching the idea, even sometimes before it half way
reaches me\tsh

I believe in my conscience I intercept many a thought which
heaven intended for another man.


Pope and his Portrait \fnast\  are fools to\break
me\tsh no martyr is ever so full of faith\break or
fire\tsh I wish I could say of good works too\tsh but I have
no\\
\stick{\kern 40pt Zeal of Anger\tsh or\hfill}
\stick{\kern 40pt Anger or Zeal\tsh\hfill}\break
And till gods and men agree together to call it by the same
name\tsh the errantest \textsc{Tartuffe}, in science\tsk in
politics\break
\stick{\tsk or in religion, shall never kindle a}
spark within
me, or have a worse word, or a more unkind greeting,\break
\stick{\hfill}\break
\stick{\footnotesize \hfill \fnast\ Vid. Pope’s Portrait.\hfill}

\newpage
\noindent
than what he will read in the next\break chapter.

\section{\lsss{CHAP}.\enspace III.}

\noindent\quad
\tsh Bon jour!\tsh good
morrow?\break\tsh so you have got your cloak on
betimes!\tsh but ’tis a cold morning, and you judge
the matter rightly\tsh ’tis better to be well
mounted, than go \hbox{o’\thinspace foot}\break
\tsh and obstructions in the
glands are dangerous\tsh And how goes it with thy
concubine\tsk thy wife\tsk and thy little ones o’\thinspace both
sides? and when did you hear from the old gentleman and
lady\tsk your sister, aunt, uncle, and cousins\tsk I~hope 
they have got better of their colds, coughs, claps,
tooth-aches, fevers, stranguries, sciaticas, swellings, and sore
eyes.\pb
\tsh What a devil of an apothecary! to take so much blood\tsk
give such a vile purge\tsk puke\tsk poultice\tsk plaister\tsk
night-draught\tsk glister\tsk blister?\tsh And why so many
grains of calomel? santa Maria! and such a dose of opium!
periclitating, pardi! the whole family of ye, from head to tail\tsh 
By my great aunt Dinah’s old black velvet mask! I think
there is no occasion for it.

Now this being a little bald about the chin, by frequently
putting off and on,\break \textit{before} she was got with child by the
coachman\tsk not one of our family would wear it after. To cover
the \textsc{mask} afresh, was more than the mask was
worth\tsh and to wear a mask which was bald,\pb or which
could be half seen through, was as bad as having no mask at
all\tsh

This is the reason, may it please your reverences, that in all
our numerous family, for these four generations, we count no more
than one archbishop, a Welch judge, some three or four
aldermen, and a single mountebank\tsh

In the sixteenth century, we boast of no less than a dozen
alchymists.

\newpage
\section{\lsss{CHAP}.\enspace IV.}

\lettrine{\lower-10pt\hbox{\large \lqq}I}{\,t} is with Love as with
Cuckol-\break dom”\tsh the suffering party is at\break least the
\textit{third}, but generally the last in the house who knows any
thing about the matter: this comes, as all the world knows, from
having half a dozen words for one thing; and so long, as what in
this vessel of the human frame, is \textit{Love}\tsk may be
\textit{Hatred}, in that\tsh \textit{Sentiment}\break
half a yard higher\tsh and \textit{Nonsense}\tsh\break
\tsh\!\tsk no, Madam,\tsk not there\tsk I\break
mean at the part I am now
pointing to with my forefinger\tsh how can we help\break
ourselves?

Of all mortal, and immortal men too, if you please, who ever soliloquized upon\pb this
mystic subject, my uncle Toby was the worst fitted, to have push’d his researches,
thro’ such a contention of feelings; and he had infallibly let them all run on, as
we do worse matters, so see what they would turn out\tsh had not Bridget’s
pre-notification of them to\break
Susannah, and Susannah’s repeated manifesto’s thereupon
to all the world, made it necessary for my uncle Toby to look into the affair.

\newpage
\section{\lsss{CHAP}.\enspace V.}

\lettrine{W}{\,\ls{hy}} weavers, gardeners, and
gla\-diators\tsk or a man with a pined leg (proceeding from some
ailment in\break
the \textit{foot})\tsk should ever have had some tender
nymph breaking her heart in secret for them, are points well and
duely settled and accounted for, by ancient and modern
physiologists.

A water-drinker, provided he is a profess’d one, and does
it without fraud or covin, is precisely in the same predicament:
not that, at first sight, there is any consequence, or shew of
logic in it,\break
\lqq That a rill of cold water dribbling\break
\lqq through my inward parts, should light\break
\lqq up a torch in my Jenny’s\tsk

\newpage
\tsh The proposition does not strike one; on the
contrary, it seems to run opposite to the natural workings of causes and
effects\tsh

But it shews the weakness and imbecility of human reason.

\tsh \lqq And in perfect good health\break
\lqq with it?\,”

\tsk The most perfect,\tsk Madam, that\break
friendship herself could wish me\tsh

\tsk \lqq And drink nothing!\tsk nothing\break\lqq but water?\,”

\tsk Impetuous fluid! the moment thou\break presses against the
flood-gates of the brain\tsk see how they give
way!\tsh

\newpage
In swims \textsc{Curiosity}, beckoning to her damsels to
follow\tsk they dive into the center of the
current\tsh

\textsc{Fancy} sits musing upon the bank, and with her
eyes following the stream, turns straws and bulrushes into masts
and bowsprits\tsh And \textsc{Desire}, with vest
held up to the knee in one hand, snatches at them, as they swim by
her, with the other\tsh

O ye water-drinkers! is it then by this delusive fountain, that
ye have so often governed and turn’d this world about like a mill-wheel\tsk grinding the faces of the
impotent\tsk bepowdering their ribs\tsk be-peppering their
noses, and changing sometimes even the very frame and face of
nature\tsh

\newpage
If I was you, quoth Yorick, I would drink more water,
Eugenius\tsk And, if I was you, Yorick, replied
Eugenius, so would I.

Which shews they had both read Longinus\tsh

For my own part, I am resolved never to read any book but my
own, as long as I live.

\newpage
\section{\lsss{CHAP}.\enspace VI.}

\lettrine{I}{ wish} my uncle Toby had been
a water-drinker; for then the thing had been accounted for, That
the first moment Widow Wadman saw him, she felt something
stirring within her in his
favour\tsk Something!\tsk something.

\noindent
\stick{\hfill \tsk Something perhaps more than friend-}
ship\tsk less than
love\tsk some\-thing\tsk no matter what\tsk no matter
where\tsk I would not give a single hair off my mule’s tail, and
be obliged to pluck it off myself (indeed the villain has not many
to spare, and is not a little vicious into the bargain), to be let
by your worships into the secret\tsh

\newpage
But the truth is, my uncle Toby was not a water-drinker;
he drank it neither\break pure nor mix’d, or any how, or any where,
except fortuitously upon some advanced posts, where better liquor
was not to be had\tsh or during the time he was under
cure; when the surgeon telling him it would extend the fibres, and
bring them sooner into contact\tsh my uncle Toby
drank it for quietness sake.

Now as all the world knows, that no\break effect in nature can be
produced without a cause and as it is as well known, that\break my uncle
Toby was neither a weaver\tsk\break a gardener, or a
gladiator\tsk unless as~a captain, you will needs have him
one\tsk\break
\stick{but then he was only a captain of foot\tsk}\\
and besides,
the whole is an equivocation\break\tsh There is nothing left for
us to sup-\pb pose, but that my uncle Toby’s
leg\tsh but that will avail us little in the present
hypothesis, unless it had proceeded from some ailment \textit{in the
foot}\tsk whereas his leg was not emaciated from any disorder
in his foot\tsk for my uncle Toby’s leg was not
emaciated at all. It was a little stiff and awkward, from a total
disuse of it, for the three years he lay confined at my
father’s house in town; but it was plump and muscular, and in
all other respects as good and promising a leg as the other.

I declare, I do not recollect any one\break
\stick{opinion or passage of my life, where my} 
\stick{understanding was more at a loss to make}
ends meet, and torture the chapter I had been writing, to the
service of the chapter following it, than in the present case:
one would think I took a pleasure in run-\pb ning into
difficulties of this kind, merely to make fresh experiments of
getting out of ’em\tsh Inconsiderate soul that thou art! What!
are not the unavoidable distresses with which, as an author and
a man, thou art hemm’d in on every side of thee\tsh are they,
Tristram, not sufficient, but thou must entangle thyself still
more?


Is it not enough that thou art in debt, and that thou hast ten
cart-loads of thy fifth and sixth volumes still\tsk still
unsold, and art almost at thy wit’s ends, how to get them off
thy hands.


To this hour art thou not tormented with the vile asthma
thou gattest in skating against the wind in Flanders?\break and is
it but two months ago, that in a\pb fit of laughter, on seeing a
cardinal make\break water like a quirister (with both hands) thou brakest
a vessel in thy lungs, where\-by, in two hours, thou lost as many
quarts of blood; and hadst thou lost as much more, did not the
faculty tell thee\break
\stick{\tsh it would have amounted to a}
gallon?\tsh

\section{\lsss{CHAP}.\enspace VII.}

\quad\tsk But for heaven’s sake, let us not talk of quarts or
gallons\tsh let us take\break the story straight before us; it
is so nice and intricate a one, it will scarce bear the
transposition of a single tittle; and, somehow or other, you have
got me thrust almost into the middle of it\tsh

\tsh I beg we may take more care.

\section{\lsss{CHAP}.\enspace VIII.}

\lettrine{M}{y} uncle Toby and the corporal
had posted down with so much\break heat and precipitation, to take
possession of the spot of ground we have so often spoke of, in
order to open their campaign as early as the rest of the allies;
that they had forgot one of the most necessary articles of the
whole affair; it was neither a pioneer’s spade, a pick-ax, or
a\break shovel\tsk

\tsh It was a bed to lie on: so that as\break Shandy-Hall was
at that time unfurn\-ish\-ed; and the little inn where poor Le
Fever died, not yet built; my uncle Toby was constrained
to accept of a bed at Mrs.\@ Wadman’s, for a night or
two,\pb till corporal Trim (who to the character of an
excellent valet, groom, cook, sempster, surgeon, and engineer,
superadded that of an excellent upholsterer too), with
the help of a carpenter and a couple of taylors, constructed one in
my uncle Toby’s house.

A daughter of Eve, for such was widow Wadman, and
’tis all the character I intend to give of her\tsk 

\tsk\lqq\textit{That she was a perfect woman};”\\[6pt]
\stick{had better be fifty leagues off\tsk or in her}
warm bed\tsk or playing with a case-knife\break\tsk or any thing
you please\tsk than make a man the object of her attention,
when\break
the house and all the furniture is her own.

\newpage
There is nothing in it out of doors and in broad day-light,
where a woman has a power, physically speaking, of viewing a man in
more lights than one\tsk but here, for her soul, she can see him
in no light without mixing something of her own goods and chattels
along with him\tsh till by reiterated acts of such
combination, he gets foisted into her inventory\tsh

\tsh And then good night.

But this is not matter of \textsc{System}; for I have
delivered that above\tsh nor is it matter of
\textsc{Breviary}\tsh for I make no man’s
creed but my own\tsh nor matter of
\textsc{Fact}\tsh at least that I know of; but
’tis matter copulative and introductory to what follows.

\section{\lsss{CHAP}.\enspace IX.}

\lettrine{I}{ do} not speak it with regard to the
coarseness or cleanness of them\tsh or\break the strength of their
gussets\tsh but pray do not night-shifts differ from
day-shifts as much in this particular, as in any thing else in the
world; that they so far exceed the others in length, that when you
are laid down in them, they fall almost as much below the feet, as
the day-shifts fall short of them?

Widow Wadman’s night-shifts (as was the mode I
suppose in King William’s and Queen
Anne’s reigns) were cut however after this fashion;
and if the fashion is changed, (for in Italy they are come to
nothing)\tsh so much the worse\pb for the public; they were two Flemish ells
and a half in length, so that allowing a moderate woman two ells,
she had half an ell to spare, to do what she would with.

Now from one little indulgence gain’d after another, in the many
bleak and decemberly nights of a seven years widow-hood, things
had insensibly come to this pass, and for the two last years had
got establish’d into one of the ordinances of the
bed-chamber\tsk That as soon as Mrs.\@ 
\stick{Wadman was put to bed, and had got her}
\stick{legs stretched down to the bottom of it, of}
which she always gave Bridget notice\tsk Bridget, with all
suitable decorum, having first open’d the bed cloaths at the
feet, took hold of the half ell of cloath we are speaking of,
and having gently, and\pb with both her hands, drawn it
downwards to its furthest extension, and then contracted it
again side long by four or five even plaits, she took a large
corking pin out of her sleeve, and with the point directed
towards her, pin’d the plaits all fast together a little above
the hem; which done, she tuck’d all in tight at the feet, and
wish’d her mistress\break a good night.

This was constant, and without any other variation than this;
that on shivering and tempestuous nights, when Bridget
untuck’d the feet of the bed, \etc to do
this\tsk she consulted no thermometer but that of her own
passions; and so performed it standing\tsk kneeling\break\tsk or
squatting, according to the different degrees of faith, hope, and
charity, she\pb was in, and bore towards her mistress that night. In
every other respect, the \textit{etiquette} was sacred, and might
have vied with the most mechanical one of the most inflexible
bed-chamber in\break Christendom.

The first night, as soon as the corporal had conducted my uncle
Toby up stairs, which was about ten\tsh Mrs.\break
Wadman threw herself into her arm chair, and crossing her
left knee with her right, which formed a resting-place for her
elbow, she reclin’d her cheek upon the palm of her hand, and
leaning forwards, ruminated till midnight upon both sides of
the question.

The second night she went to her bureau, and having ordered
Bridget to\pb bring her up a couple of fresh candles and leave
them upon the table, she took out her marriage-settlement, and read
it over with great devotion: and the third night (which was the
last of my uncle Toby’s stay) when Bridget had
pull’d down the night-shift, and was assaying to stick in the
corking pin\tsh

\tsh With a kick of both heels at once, but at the same
time the most natural kick that could be kick’d in her
\stick{situation\tsk for supposing\enspace\astfill}
\stick{\astfill\enspace to be the sun in its meridian, it}
was a north-east kick\tsh she kick’d the pin out of her
fingers\tsh the \textit{etiquette} which hung upon it, down\tsh
down it fell to the ground, and was shiver’d into a thousand
atoms.

\newpage
From all which it was plain that widow Wadman was in love
with my uncle Toby.

\section{\lsss{CHAP}.\enspace X.}

\lettrine{M}{y} uncle Toby’s head at
that time was full of other matters,\break
so that it was not till the
demolition of Dunkirk, when all the other civilities of
Europe were settled, that he found leisure to return
this.

This made an armistice (that is, speaking with regard to my uncle Toby\tsk but with
respect to Mrs.\@ Wadman, a vacancy)\tsk of almost eleven years. But in all cases of
this nature, as it is the second blow, happen at what distance of\pb time it will,
which makes the fray\tsh\break 
I chuse for that reason to call these\break 
the amours of my uncle Toby with\break
Mrs.\@ Wadman, rather than the amours\break
of Mrs.\@ Wadman with my uncle Toby.

This is not a distinction without a difference.

It is not like the affair of \textit{an old hat cock’d}\tsk and
\textit{a cock’d
old hat}, about which your reverences have so often been at odds with one another
\tsh but there is a difference here in the nature of\break
things\tsh

And let me tell you, gentry, a wide one too.

\newpage
\section{\lsss{CHAP}.\enspace XI.}

\lettrine{N}{\,\ls{ow}} as widow Wadman did love
my uncle Toby\tsh and my uncle Toby did not
love widow Wadman, there was nothing for widow Wadman
to do, but to go on and love my uncle Toby\tsh or
let it alone.

Widow Wadman would do neither the one or the other\tsk

\tsk Gracious heaven!\tsk but I forget I am a
little of her temper myself; for whenever it so falls out, which it
sometimes does about the equinoxes, that an earthly goddess is so
much this, and that, and t’other, that I cannot eat my\pb
breakfast for her\tsh and that she careth not three
halfpence whether I eat my breakfast or no\tsh

\tsh Curse on her! and so I send her to Tartary,
and from Tartary to \textit{Terra del Fuogo}, and so on to the
devil: in short, there is not an infernal nitch where I do not take
her divinityship and stick\break it.

But as the heart is tender, and the passions in these tides ebb
and flow ten times in a minute, I instantly bring her back again;
and as I do all things in extremes, I place her in the very
centre\break
of the milky-way\tsh

Brightest of stars! thou wilt shed thy influence upon some
one\tsh

\newpage
\tsh The duce take her and her influence
too\tsh for at that word I lose all
patience\tsh much good may it do him! \tsh By all
that is hirsute and gashly!\break
I cry, taking off my furr’d cap, and twisting it round my
finger\tsh I would not give sixpence for a dozen such!

\tsh But ’tis an excellent cap too (putting it upon my head, and
pressing it close to my ears)\tsk and warm\tsk and soft;
especially if you stroke it the right way\tsk\break
\stick{but alas! that will never be my luck\tsh}
(so here my philosophy is shipwreck’d again)

\tsh No; I shall never have a finger in the pye (so
here I break my meta-\break phor)\tsh

\newpage
Crust and Crumb\\
\null\indent Inside and out\\
\null\indent Top and bottom\tsh I detest it, I hate it, I repudiate
it\tsh I’m sick at the sight of it \tsh 

’Tis all pepper,\\
\null\hbox to 42pt{\hss}garlick,\\
\null\hbox to 42pt{\hss}staragen,\\
\null\hbox to 42pt{\hss}salt, and\\
\null\hbox to 42pt{\hss}devil’s dung\tsh by
the great arch-cook of cooks, who does nothing, I think, from
morning to night, but sit down by the fire-side and invent
inflammatory dishes for us, I would not touch it for the
world\tsh

\tsh O Tristram! Tristram! cried Jenny.

\newpage
O Jenny! Jenny! replied I, and so went on
with the twelfth chapter.

\section{\lsss{CHAP}.\enspace XII.}

\qquad\tsh \lqq Not touch it for the world”\break
did I say\tsh

Lord, how I have heated my imagination with this metaphor!

\section{\lsss{CHAP}.\enspace XIII.}

\lettrine{W}{\,\ls{hich}} shews, let your reverences and
worships say what you\break will of it (for as for
\textit{thinking}\tsh all who \textit{do} think\tsk think pretty
much alike both upon it and other matters)\tsh\pb Love is
certainly, at least alphabetically speaking, one of the most

\medskip
\vbox{\halign{\quad #\hss\cr
A gitating\cr
B ewitching\cr
C onfounded\cr
D evilish affairs of life\tsh the most\cr
E xtravagant\cr
F utilitous\cr
G alligaskinish\cr
H andy-dandyish\cr
I racundulous (there is no K to it) and\cr
L yrical of all human passions: at\cr
\kern 42pt  the same time, the most\cr
M isgiving\cr
N innyhammering\cr
O bstipating\cr
P ragmatical\cr
S tridulous\cr
}}

\vskip -\baselineskip
\noindent
\quad R idiculous\tsh though by the bye the\pb 
\stick{R should have gone first\tsk But in short}
\stick{’tis of such a nature, as my father once}
\stick{told my uncle Toby upon the close of a}
\stick{long dissertation upon the subject\tsh}
\lqq You can scarce,” said he, \lqq combine\break
\lqq two ideas together upon it, brother\break
\lqq Toby, without an hypallage\,”\tsh\break
What’s that? cried my uncle Toby.

The cart before the horse, replied my father\tsh

\tsh And what is he to do there? cried my uncle Toby.

Nothing, quoth my father, but to get in\tsh or let it
alone.

Now widow Wadman, as I told you before, would do neither
the one or the other.

\newpage
She stood however ready harnessed and caparisoned at all points,
to watch accidents.

\section{\lsss{CHAP}.\enspace XIV.}

\lettrine{T}{ he fates}, who certainly all
fore-knew of these amours of widow Wadman and my uncle
Toby, had, from the first creation of matter and motion (and
with more courtesy than they usually do things of this kind),
established such a chain of causes and effects hanging so fast to
one another, that it was scarce possible for my uncle Toby
to have dwelt in any other house in the world, or to have occupied
any other garden in Christendom, but the very house and
garden which join’d and\pb laid parallel to Mrs.\@ Wadman’s; this, with the advantage of a thickset
arbour in Mrs.\@ Wadman’s garden, but planted in the
hedge-row of my uncle Toby’s, put all the occasions
into her hands which Love-militancy wanted; she could observe my uncle Toby’s motions, and
was mistress likewise of his councils of war; and as his
unsuspecting heart had given leave to the corporal, through the
mediation of Bridget, to make her a wicker-gate of
communication to enlarge her walks, it enabled her to carry on her
approaches to the very door of the sentry-box; and sometimes out of
gratitude, to make the attack, and endeavour to blow my uncle
Toby up in the very sentry-box itself.

\section{\lsss{CHAP}.\enspace XV.}

\lettrine{I}{t} is a great pity\tsh but
’tis certain from every day’s observation of man, that
he may be set on fire like a candle, at either end\tsk provided
there is a sufficient wick standing out; if there is
not\break\tsk there’s an end of the affair; and if there
is\tsk by lighting it at the bottom, as the flame in that case
has the misfortune generally to put out itself\tsk there’s
an end of the affair again.

For my part, could I always have the ordering of it which way I
would be burnt myself\tsk for I cannot bear the thoughts of being
burnt like a beast\tsk\break I would oblige a housewife constantly to\pb
light me at the top; for then I should burn down decently to the
socket; that is, from my head to my heart, from my heart to my
liver, from my liver to my bowels, and so on by the meseraick veins
and arteries, through all the turns and lateral insertions of the
intestines and\break
their tunicles to the blind gut\tsh

\noindent
\tsh I beseech you, doctor Slop, quoth my uncle
Toby, interrupting him as he mentioned the \textit{blind gut},
in a discourse with my father the night my mother was brought to
bed of me\tsh I beseech you, quoth my uncle Toby,
to tell me which is the blind gut; for, old as I am, I vow I do not
know to this day where it lies.

\newpage
The \textit{blind gut}, answered doctor Slop, lies betwixt the
\textit{Illion} and \textit{Colon}\tsh

\tsh In a man? said my father.

\tsh ’Tis precisely the same, cried
doctor Slop, in a woman\tsh

That’s more than I know; quoth my father.

\newpage
\section{\lsss{CHAP}.\enspace XVI.}

\quad
\tsh And so to make sure of both systems, Mrs.
Wadman predetermined to light my uncle Toby neither
at this end or that; but, like a prodigal’s candle, to light
him, if possible, at both ends at\break once.

Now, through all the lumber rooms of military furniture,
including both of horse and foot, from the great arsenal\break 
of Venice to the Tower of London\break (exclusive), if
Mrs.\@ Wadman had been rummaging for seven years together, and
with Bridget to help her, she could not have found any one
\textit{blind} or \textit{man\-telet} so fit for her purpose, as that
which\pb the expediency of my uncle Toby’s affairs had
fix’d up ready to her hands.

I believe I have not told you\tsh but I don’t know\tsh possibly I have\tsh be it
as it will, ’tis one of the number of those many things, which a man had better do
over again, than dispute about it\tsk That whatever town or fortress the corporal was
at work upon, during the course of their campaign, my uncle Toby always took care,
on the inside of his sentry-box, which was towards his left hand, to have a plan of
the place, fasten’d up with two or three pins at the top, but loose at the bottom,
for the conveniency of holding it up to the eye, \etc .\@ .\@
.\@ as\break occasions required;
so that when an attack was resolved upon, Mrs.\@ Wadman had nothing more to do, when
she had\pb got advanced to the door of the sentry-box, but to extend her right hand;
and edging in her left foot at the same movement, to take hold of the map or plan,
or upright, or whatever it was, and with out-stretched neck meeting it half way, \tsk 
to advance it towards her; on which my uncle Toby’s passions
were sure to catch fire\tsh for he would instantly take hold of
the other corner of the map in his left hand, and with the end
of his\break pipe, in the other, begin an expla-\break nation.

When the attack was advanced to this point;\tsh the
world will naturally enter into the reasons of Mrs.
Wadman’s next stroke of generalship\tsh which
was, to take my uncle Toby’s tobacco-pipe out of his
hand as soon as she possibly could;\pb which, under one pretence or
other, but generally that of pointing more distinctly at some
redoubt or breast-work in the map, she would effect before my uncle
Toby (poor soul!) had well march’d above half a dozen
toises with it.

\tsk  It obliged my uncle Toby to make use of his
forefinger.

The difference it made in the attack was this; That in going upon it, as in the
first case, with the end of her fore-finger against the end of my uncle Toby’s
tobacco-pipe, she might have travelled with it, along the lines, from Dan to
Beersheba, had my uncle Toby’s lines reach’d so far, without any effect: For as
there was no arterial or vital heat in the end of the tobacco-pipe, it could excite
no\pb sentiment\tsh it could neither give fire by pulsation \tsh or receive it by
sympathy\break\null\tsh ’twas nothing but smoak.

Whereas, in following my uncle Toby’s forefinger
with hers, close thro’ all the little turns and indentings of
his works\tsh pressing sometimes against the side of
it\tsh then treading upon it's nail \tsh then
tripping it up\tsh then touching it here \tsh then
there, and so on\tsh it set something at least in
mo-\break tion.

This, tho’ slight skirmishing, and at a distance from the
main body, yet drew on the rest; for here, the map usually falling
with the back of it, close to the side of the sentry-box, my uncle
Toby, in the simplicity of his soul, would lay his\pb hand flat
upon it, in order to go on with his explanation; and Mrs.\@
Wadman, by a manœuvre as quick as thought, would as
certainly place her’s close besides it; this at once opened a
communication, large enough for any sentiment to pass or repass,
which a person skill’d in the elementary and practical part of love-making, has
occasion for\tsh

By bringing up her forefinger parallel\break (as before) to my uncle Toby’s\tsk it
\stick{unavoidably brought the thumb into}
\stick{action\tsk and the forefinger and thumb}
being once engaged, as naturally brought in the whole hand. Thine, dear uncle
Toby!\@ was never now in it’s right place\break\tsh Mrs.\@ Wadman had it ever to
take up, or, with the gentlest pushings, protrusions, and equivocal compressions,\pb
that a hand to be removed is capable of receiving\tsh to get it press’d a hair
breadth of one side out of her way.

Whilst this was doing, how could she forget to make him
sensible, that it was her leg (and no one’s else) at the
bottom of the sentry-box, which slightly press’d against the
calf of his\tsh So that my uncle Toby being thus
attack’d and sore push’d on both his
wings\tsh was it a wonder, if now and then, it put his
centre into disorder?\tsh

\tsh The duce take it! said my uncle Toby.

\newpage
\section{\lsss{CHAP}.\enspace XVII.}

\lettrine{T}{hese} attacks of Mrs.\@ Wadman,
you will readily conceive to be of different kinds; varying from
each other, like the attacks which history is full of, and from the
same reasons. A general looker~on would scarce allow them to be
attacks at all\tsh or if he did, would confound them all
together\tsh but I write not to them: it will be time
enough to be a little more exact in my descriptions of them, as I
come up to them, which will not be for some chapters; having
nothing more to add in this, but that in a bundle of original
papers and drawings which my father took care to roll up by
themselves, there is a plan of\pb 
Bouchain in perfect preservation (and
\stick{shall be kept so, whilst I have power to}
\stick{preserve any thing) upon the lower corner}
\stick{of which, on the right hand side, there}
\stick{is still remaining the marks of a snuffy}
\stick{finger and thumb, which there is all the} 
\stick{reason in the world to imagine, were}
\stick{Mrs.\@ Wadman’s; for the opposite side}
\stick{of the margin, which I suppose to have}
\stick{been my uncle Toby’s, is absolutely clean:}
This seems an
authenticated record of one of these attacks; for there are
vestigia of the two punctures partly grown up, but still visible on
the opposite corner of the map, which are unquestionably the very
holes, through which it has been pricked up in the
sentry-box\tsh

By all that is priestly! I value this precious relick, with it’s
\textit{stigmata} and\pb
\textit{pricks}, more than all the relicks of the
Romish church\tsh always excepting, when I am
writing upon these matters, the pricks which entered the flesh of
St.\@ \textit{Radagunda} in the desert, which in your road from
\textsc{Fesse} to \textsc{Cluny}, the nuns of that name
will shew you for love.

\medskip

\section{\lsss{CHAP}.\enspace XVIII.}

\lettrine{I}{ think}, an’ please your
honour, quoth Trim, the fortifications are quite
destroyed\tsh and the bason is upon a level with the
mole\tsh I think so too; replied my uncle Toby with
a sigh half suppress’d\tsh but step into the
parlour, Trim, for the stipulation\tsh it lies upon
the table.

\newpage
\noindent
\stick{\indent\tight{It has lain there these six weeks,
replied}}
the corporal, till
this very morning that the old woman kindled the fire with
it\tsk

\tsk Then, said my uncle Toby, there is no
further occasion for our services. The more, an’ please your
honour, the pity, said the corporal; in uttering which he cast his
spade into the wheel-barrow, which was beside him, with an air the
most expressive of disconsolation that can be imagined, and was
heavily turning about to look for his pick-ax, his pio\-neer’s
shovel, his picquets and other little military stores, in order to
carry them off the field\tsh when a heigh ho!\@ from the
sentry-box, which being made of thin slit deal, reverberated the
sound more sorrowfully to his ear, forbad him.

\newpage
\tsk No; said the corporal to himself,\break I’ll do it
before his honour rises to-mor\-row morning; so taking his spade out
of the wheel-barrow again, with a little earth in it, as if to
level something at the foot of the glacis\tsh but with a
real intent to approach nearer to his master, in order to divert
him\tsh he loosen’d a sod or two\tsh pared
their edges with his spade, and having given them a gentle blow or
two with the back of it, he sat himself down close by my uncle
Toby’s feet, and began as follows.

\newpage
\section{\lsss{CHAP}.\enspace XIX.}

\lettrine{I}{t} was a thousand
pities\tsh though I believe, an’ please your honour,
I am going to say but a foolish kind of a thing for a
soldier\tsh

A soldier, cried my uncle Toby, interrupting the
corporal, is no more exempt from saying a foolish thing, Trim, than a man
of letters\tsh But not so often; an’ please your
honour, replied the corporal\tsh My uncle Toby gave
a nod.

It was a thousand pities then, said the corporal, casting his
eye upon Dunkirk, and the mole, as \textit{Servius Sulpicius},
in returning out of Asia (when he sailed from\pb
Ægina towards Megara), did upon Co\-rinth
and Pyreus\tsh

\tsk \lqq It was a thousand pities, an’ please
your honour, to destroy these works\tsk\break and a thousand
pities to have let them stood.”\tsh

\tsh Thou art right, Trim, in both cases; said
my uncle Toby\tsh This, continued the corporal, is
the reason, that from the beginning of their demolition to the
end\tsh I have never once whistled, or sung, or
laugh’d, or cry’d, or talk’d of pass'd done deeds,
or told your honour one story good or bad\tsh

\tsh Thou hast many excellencies, Trim, said my
uncle Toby, and I hold it not the least of them, as thou
happenest\pb to be a story-teller, that of the number thou hast told me, either to amuse me in my
painful hours, or divert me in my grave ones\tsk thou hast seldom
told me a bad one\tsh

\tsh Because, an’ please your honour, except one
of a \textit{King of Bohemia and his seven castles},\tsk they are
all true; for they are about myself\tsh

I do not like the subject the worse, Trim, said my uncle
Toby, on that score: But prithee what is this story? thou
hast excited my curiosity.

I’ll tell it your honour, quoth the corporal,
directly\tsh Provided, said my uncle Toby, looking
earnestly towards Dun\-kirk and the mole
again\tsh provided it\pb is not a merry one; to such,
Trim, a man should ever bring one half of the entertainment
along with him; and the disposition I am in at present would wrong
both thee, Trim, and thy story\tsk It is\break
not a merry one by any means, replied the corporal\tsk Nor would
I have it alto-\break gether a grave one, added my uncle Toby\tsh It
is neither the one nor the other, replied the corporal, but will
suit your honour exactly\tsh Then I’ll thank thee for it with
all my heart, cried my uncle Toby, so prithee begin it, Trim.

The corporal made his reverence;\break
and though it is not so easy a
matter as the world imagines, to pull off a lank montero cap
with grace\tsk or a whit less difficult, in my
conceptions, when a man is sitting squat upon the ground, to make\pb 
a bow so teeming with respect as the 
\stick{corporal was wont; yet by suffering the}
\stick{palm of his right hand, which was to-}
\stick{wards his master, to slip backwards upon}
\stick{\tight{the grass, a little beyond his body, in order}}
\stick{to allow it the greater sweep\tsk and by}
\stick{\tight{an unforced compression, at the same time,}}
\stick{of his cap with the thumb and the two}
\stick{forefingers of his left, by which the dia-}
\stick{meter of the cap became reduced, so that}
\stick{it might be said, rather to be insensibly}
\stick{squeez’d\tsk than pull’d off with a flatus}
\tsh the corporal acquitted himself of both in a
better manner than the posture of his affairs promised; and having
hemmed twice, to find in what key his story would best go, and best
suit his master’s humour\tsk he exchanged a single
look of kindness with him, and set off thus.

\newpage
\centerline{The Story of the king of Bohemia}
\centerline{and his seven castles.}

\lettrine{T}{\,\ls{here}} was a certain king of\break
Bo - - he\tsh

As the corporal was entering the confines of Bohemia, my
uncle Toby obliged him to halt for a single moment; he had
set out bare-headed, having since he pull’d off his
Montero-cap in the latter end of the last chapter, left it
lying beside him on the ground.

\tsh The eye of Goodness espieth all things\tsh so that before the corporal had
well got through the first five words\pb of his story, had my uncle Toby twice
touch’d his Montero-cap with the end of his cane, interrogatively\tsh as much as to
say, Why don’t you put it on, Trim?\break Trim took it up with the most respectful
slowness, and casting a glance of humiliation as he did it, upon the embroidery of
the fore-part, which being dismally tarnish’d and fray’d moreover in some of the
principal leaves and boldest parts of the pattern, he lay’d it down again between
his two feet, in order to moralize upon the subject.

\tsh ’Tis every word of it but too true, cried my
uncle Toby, that thou art about to observe\tsh\

\newpage\indent
\lqq\textit{Nothing in this world, Trim, is}\break
\lqq\textit{made to last for ever.}”

\tsh But when tokens, dear Tom, of thy love and
remembrance wear out, said Trim, what shall we say?

There is no occasion, Trim, quoth my uncle Toby,
to say any thing else; and was a man to puzzle his brains till
Doom’s~day, I believe, Trim, it would be
impossible.

The corporal, perceiving my uncle Toby was in the right,
and that it would be in vain for the wit of man to think of
extracting a purer moral from his cap, without further attempting
it, he put it on; and passing his hand across his fore-\pb head to rub
out a pensive wrinkle, which the text and the doctrine between them
had engender’d, he return’d, with the same look and
tone of voice, to his story of the king of Bohemia and his
seven castles.

\newpage
\centerline{The story of the king of Bohemia and}
\centerline{his seven castles, continued.}

\lettrine{T}{\,\ls{here}} was a certain king of\break
Bohemia, but in whose reign, except his
own, I am not able to inform your honour\tsh

I do not desire it of thee, Trim, by any means, cried my
uncle Toby.

\tsh It was a little before the time, an’ please
your honour, when giants were beginning to leave off
breeding;\tsk\break but in what year of our Lord that\break
was\tsh

\newpage
\tsh I would not give a halfpenny to know, said my uncle
Toby.

\tsh Only, an’ please your honour, it makes a
story look the better in the face\tsh

\tsh ’Tis thy own, Trim, so orna-\break ment it
after thy own fashion; and take any date, continued my uncle Toby,\break
looking pleasantly upon him\tsk take any date in the
whole world thou choosest, and put it to\tsk thou art heartily
welcome\tsh

The corporal bowed; for of every century, and of every year of
that century, from the first creation of the world down to
Noah’s flood; and from Noah’s\pb flood to
the birth of Abraham; through all the pilgrimages of the
patriarchs, to the departure of the Israelites out of
Egypt\tsh and throughout all the Dynasties,
Olympiads, Urbecondita’s, and other memorable epochas of the
different nations of the world, down to the coming of Christ, and
from thence to the very moment in which the corporal was telling
his story\tsh had my uncle Toby subjected this vast
empire of time and all its abysses at his feet; but as \textsc{modesty} scarce touches with a
finger what \textsc{liberality} offers her with both hands
open\tsk the corporal contented himself with the very \textit{worst
year} of the whole bunch; which, to prevent your honours of the
Majority and Minority from tearing the very flesh off your bones\pb in
contestation, ‘Whether that year is not always the last
cast-year of the last cast-almanack’\tsh I tell you
plainly it was; but from a different reason than you\break wot of\tsh

\tsk It was the year next him\tsk\break which being
the year of our Lord\break 
seventeen hundred and twelve, when\break
the duke of Ormond was playing the devil in
Flanders\tsh the corporal took it, and set out with
it afresh on his expedition to Bohemia.

\newpage
\centerline{The story of the king of Bohemia and}
\centerline{his seven castles, continued.}

\lettrine{I}{\,n} the year of our Lord one thousand seven hundred and twelve, there
was, an’ please your honour\tsh

\tsh To tell thee truly, Trim, quoth my uncle
Toby, any other date would have pleased me much better, not
only on account of the sad stain upon our history that year, in
marching off our troops, and refusing to cover the siege of
Quesnoi, though Fagel was carrying on the works with
such incredible vigour\tsk but likewise on the score,
Trim, of thy own story; because if there are\tsk and
which, from what thou hast dropt, I\pb partly suspect to be the
fact\tsk if there are giants in it\tsh

There is but one, an’ please your honour\tsh

\tsh ’Tis as bad as twenty, replied my uncle
Toby\tsh thou should’st have carried him back
some seven or eight hundred years out of harm’s way, both of
critics and other people: and therefore I would advise thee, if
ever thou\break tellest it again\tsh

\tsh If I live, an’ please your honour, but once
to get through it, I will never tell it again, quoth Trim,
either to man, woman, or child\tsh Poo\tsk poo! said my uncle Toby\tsk but with accents of
such sweet encouragement did he\pb utter it, that the corporal went on
with his story with more alacrity than ever.

\bigskip
\centerline{The story of the king of Bohemia and}
\centerline{his seven castles, continued.}

\lettrine{T}{\,\ls{here}} was, an’ please your
honour, said the corporal, raising his voice and rubbing the palms
of his two hands cheerily together as he begun, a certain king of
Bohemia\tsh

\tsh Leave out the date entirely, Trim, quoth my
uncle Toby, leaning forwards, and laying his hand gently
upon the corporal’s shoulder to temper the
interruption\tsk leave it out entirely, Trim; a story
passes very well without these niceties, unless one is pretty sure\pb
of ’em\tsh Sure of ’em! said the corporal,
shaking his head\tsh

Right; answered my uncle Toby, it is not easy,
Trim, for one, bred up as thou and I have been to arms, who
seldom looks further forward than to the end of his
musket, or backwards beyond his knapsack, to know much about this
matter\tsh God bless your honour! said the corporal, won
by the \textit{manner} of my uncle Toby’s reasoning, as much as
by the reasoning itself, he has something else to do; if not on
action, or a march, or upon duty in his garrison\tsk he has his
firelock, an’ please your honour, to furbish\tsk his
accoutrements to take care of\tsk his regimentals to
mend\tsk himself to shave and keep clean, so as to appear\pb always
like what he is upon the parade; what business, added the corporal
triumphantly, has a soldier, an’ please your honour, to know
any thing at all of \textit{geography ?}

\tsh Thou would’st have said \textit{chro\-no\-logy},
Trim, said my uncle Toby; for as for geography,
’tis of absolute use to him; he must be acquainted intimately
with every country and its boundaries where his profession carries
him; he should know every town and city, and village and hamlet,
with the canals, the roads, and hollow ways which lead up to them; there
is not a river or a rivulet he passes, Trim, but he should
be able at first sight to tell thee what is its name\tsk in what
mountains it takes its\pb rise\tsk what is its course\tsk how far
it is navigable\tsk where fordable\tsk where not; he should
know the fertility of every valley, as well as the hind who ploughs
it; and be able to describe, or, if it is required, to give thee an
exact map of all the plains and defiles, the forts, the
acclivities, the woods and morasses, thro’ and by which his
army is to march; he should know their produce, their plants, their
minerals, their waters, their animals, their seasons, their
climates, their heats and cold, their inhabitants, their customs,
their language, their policy, and even their religion.

Is it else to be conceived, corporal, continued my uncle
Toby, rising up in his sentry-box, as he began to warm in
this\pb part of his discourse\tsk how Marlborough could have
marched his army from the banks of the Maes to
Belburg; from Belburg to Kerpenord\tsk (here the
corporal could sit no longer) from Kerpenord, Trim, to
Kalsaken; from Kalsaken to Newdorf; from
Newdorf to Landenbourg; from Landenbourg to
Mildenheim; from Mildenheim to Elchingen; 
\stick{from Elchingen to Gingen; from Gingen}
\stick{to Balmerchoffen; from Balmerchoffen to}
\stick{Skellenburg, where he broke in upon the}
enemy’s works; forced his passage over the \textit{Danube}; cross’d the
\textit{Lech}\tsk push’d on his troops into the heart of the empire, marching at the
head of them through Fribourg, Hokenwert, and Schonevelt, to the plains of Blenheim
and Hochstet?\break\null\tsh Great as he was, corporal, he could\pb 
not have advanced a step, or made one single day’s march without the aids of\break
\textit{Geography}\tsh As for \textit{Chronology}, I own, Trim, continued my uncle
Toby, sitting down again coolly in his sentry-box, that of all others, it seems a
science which the soldier might best spare, was it not for the lights which that
science must one day give him, in determining the invention of powder; the furious
execution of which, renversing every thing like thunder before it, has become a new
æra to us of military improvements, changing so totally the nature of attacks and
defences both by sea and land, and awakening so much art and skill in doing it, that
the world cannot be too exact in ascertaining the precise time of its discovery, or
too inquisitive in knowing what great man\pb was the discoverer, and what occasions
gave birth to it.

I am far from controverting, conti\-nued my uncle Toby,
what historians agree in, that in the year of our Lord 1380, under
the reign of Wencelaus, 
\stick{son of Charles the fourth\tsh a certain}
\stick{priest, whose name was Schwartz, shew’d}
\stick{the use of powder to the Venetians, in}
\stick{their wars against the Genoese; but ’tis}
\stick{certain he was not the first; because if}
\stick{we are to believe Don Pedro, the bishop}
\stick{of Leon\tsk How came priests and bishops,}
\stick{an’ please your honour, to trouble their}
\stick{heads so much about gun-powder? God}
\stick{knows, said my uncle Toby\tsh his pro-}
\stick{vidence brings good out of every thing}
\stick{\tsk and he avers, in his chronicle of King}
Alphonsus, who reduced Toledo, That\pb in the year
1343, which was full thirty seven years before that time, the secret of powder was
well known, and employed with success, both by Moors and Christians, not only in
their sea-combats, at that period, but in many of their most memorable sieges in
Spain and Barbary\tsk\break And all the world knows, that Friar Bacon had wrote expressly
about it, and had generously given the world a receipt to make it by, above a
hundred and fifty years before even Schwartz was born\tsk And that the Chinese,
added my uncle Toby, embarrass us, and all accounts of it, still more, by boasting
of the invention some hundreds of years even before him\tsh

\tsk They are a pack of liars, I believe, cried
Trim\tsh

\tsh They are somehow or other deceived, said my uncle
Toby, in this matter, as is plain to me from the present
miserable state of military architecture amongst them; which
consists of nothing more than a fossè with a brick wall without
flanks\tsk and for what they give us as a bastion at each angle
of it, ’tis so barbarously constructed, that it looks for all
the world\tsh \tsh Like one of my seven
castles, an’ please your honour, quoth Trim.

My uncle Toby, tho’ in the utmost distress for a
comparison, most courteously refused Trim’s
offer\tsk till Trim telling him, he had half a dozen more
in Bohemia, which he knew not how to get off his
hands\tsh my uncle Toby was so\pb touch’d with
the pleasantry of heart of the corporal\tsh that he
discontinued his dissertation upon gunpowder\tsh and
begged the corporal forthwith to go on with his story of the King
of Bohemia and his seven castles.

\bigskip
\centerline{The story of the king of Bohemia and}
\centerline{his seven castles, continued.}

\lettrine{T}{\,\ls{his}} \textit{unfortunate} King of
Bohemia, said Trim,\tsh Was he unfortunate,
then? cried my uncle Toby, for he had been so wrapt up in
his dissertation upon gun-powder, and other military affairs, that
tho’ he had desired the corporal to go on, yet the many
interruptions he had given, dwelt not so strong upon his fancy as
to account for the\pb epithet\tsh Was he \textit{unfortunate},
then, Trim? said my uncle Toby,
pathetically\break\tsh The corporal, wishing first the \textit{word} and
all its synonimas at the devil, forthwith began to run back in his
mind, the principal events in the King of Bohemia’s
story; from every one of which, it appearing that he was the most
fortunate man that ever existed in the world\tsh\break
it put the corporal to a stand: for not caring to retract his epithet\tsh and less
to explain it \tsh and least of all, to twist his tale (like men of lore) to serve a
system\tsh he looked up in my uncle Toby’s face for assistance\tsh but seeing it was
the very thing, my uncle Toby sat in expectation of himself\tsh after a hum and a
haw, he went on\tsh

\newpage
The King of Bohemia, an’ please your honour,
replied the corporal, was \textit{unfortunate}, as
thus\tsh That taking great pleasure and delight in
navigation and all sort of sea-affairs\tsh and there
\textit{happening} throughout the whole kingdom of Bohemia, to
be no sea-port town whatever\tsh

How the duce should there\tsk Trim? cried my
uncle Toby; for Bohemia being totally inland, it
could have happen’d no otherwise\tsh It might, said
Trim, if it had pleased God\tsh

My uncle Toby never spoke of the\break being and natural
attributes of God, but with diffidence and
hesitation\tsh

\newpage
\tsk I believe not, replied my uncle Toby, after
some pause\tsk\ for being inland, as I said, and having
Silesia and Moravia to the east; Lusatia and
Upper Saxony to the north; Franconia to the
west; and Bavaria to the south; Bohemia could not
have been propell’d to the sea without ceasing to be
Bohemia\tsh nor could the sea, on the other hand,
have come up to Bohemia, without overflowing a great part of
Germany, and destroying millions of unfortunate inhabitants
who could make no defence against\break it\tsh Scandalous! cried
Trim\tsk Which would bespeak, added my uncle Toby,
mildly, such a want of compassion in him who is the father of
it\tsh that, I think, Trim \tsh the thing
could have happen’d no way.

\newpage
The corporal made the bow of unfeign’d conviction; and
went on.

Now the King of Bohemia with his queen and courtiers
\textit{happening} one fine summer’s evening to walk
out\tsh Aye! there the word \textit{happening} is right,
Trim, cried my uncle Toby; for the King of
Bohemia and his queen might have walk’d out, or let it
alone:\tsh ’twas a matter of contingency, which
might happen, or not, just as chance ordered it.

King William was of an opinion, an’ please your
honour, quoth Trim, that every thing was predestined for us
in this world; insomuch, that he would often say to his soldiers,
that \lqq every ball had its billet.” He was a great man,
said my uncle Toby\tsh And I believe, con-\pb tinued
Trim, to this day, that the shot which disabled me at the
battle of Landen, was pointed at my knee for no\break other
purpose, but to take me out of his service, and place me in your
honour’s, where I should be taken so much better care of in
my old age\tsh It shall never, Trim, be construed
otherwise, said my uncle Toby.

The heart, both of the master and the man, were alike subject to
sudden over-flowings;\tsh a short silence ensued.

Besides, said the corporal, resuming the discourse\tsk but in
a gayer accent\tsh if it had not been for that single
shot, I had never, ’an please your honour, been in
love\tsh

\newpage
So, thou wast once in love, Trim! said my uncle Toby, smiling\tsh

Souse!\@ replied the corporal\tsk over head and ears! an’ please your honour. Prithee
when?\@ where?\tsk and how came it to pass?\tsk I never heard one word of it before; quoth
my uncle Toby:\tsh I dare say, answered Trim, that every drummer and
serjeant’s son in the regiment knew of it\tsh Its high time I should \tsh said my
uncle Toby.

Your honour remembers with concern, said the corporal, the total rout and confusion
of our camp and army at the affair of Landen; every one was left to shift for
himself; and if it had not been for the regiments of Wyndham, Lumley, and Galway,
which covered the re-\pb treat over the bridge Neerspeeken, the king himself could
scarce have gained it\tsh he was press’d hard, as your honour knows, on every side
of him\tsh

Gallant mortal! cried my uncle Toby, caught up with
enthusiasm\tsk this moment, now that all is lost, I see him
galloping across me, corporal, to the left, to bring up the remains
of the English horse along with him to support the right,
and tear the laurel from Luxembourg’s
brows, if yet ’tis possible\tsh I see him with the
knot of his scarfe just shot off, infusing fresh spirits into poor
Galway’s regiment\tsk riding along the
line\tsk then wheeling about, and charging Conti at the
head of it\tsh Brave! brave by heaven! cried my uncle
Toby\tsk he deserves\pb a crown \tsh As richly, as a
thief a halter; shouted Trim.

My uncle Toby knew the corporal’s
loyalty;\tsk otherwise the comparison was not at all to his
mind\tsh it did not altogether strike the corporal’s
fancy when he had made it\tsh but it could not be
recall’d\tsh so he had nothing to do, but
proceed.

As the number of wounded was prodigious, and no one had time to
think of any thing but his own safety\tsk Though Talmash,
said my uncle Toby, brought off the foot with great
prudence\tsh\break But I was left upon the field, said the
corporal. Thou wast so; poor fellow! replied my uncle
Toby\tsk So that it was noon the next day, continued the
cor-\pb poral, before I was exchanged, and put into a cart
with thirteen or fourteen more, in order to be convey’d to
our\break hospital.

There is no part of the body, an’ please your honour,
where a wound occasions more intolerable anguish than upon the
knee\tsh

Except the groin; said my uncle Toby. An’ please
your honour, replied the corporal, the knee, in my opinion, must
certainly be the most acute, there being so many tendons and
what-d’ye-call-’ems all about it.

It is for that reason, quoth my uncle Toby, that the
groin is infinitely more sensible\tsh there being not only
as ma-\pb ny tendons and what-d’ye-call-’ems (for I know
their names as little as thou do’st)\tsh about
it\tsh but moreover\break \astW3\tsh

Mrs.\@ Wadman, who had been all the time in her
arbour\tsk instantly stopp’d her\break
breath\tsk unpinn’d her mob at the chin, and stood upon one
leg\tsh

The dispute was maintained with ami\-cable and equal force betwixt
my uncle Toby and Trim for some time; till
Trim at length recollecting that he had often cried at his
master’s sufferings, but never shed a tear at his
own\tsk was for giving up the point, which my uncle Toby
would not allow\tsh ’Tis a proof of nothing,
Trim, said he, but the generosity of thy
temper\tsh

So that whether the pain of a wound in the groin (cæteris
paribus) is greater than the pain of a wound in the
knee\tsh or

Whether the pain of a wound in the knee is not greater than the
pain of a wound in the groin\tsh are points which to this
day remain unsettled.

\section{\lsss{CHAP}.\enspace XX.}

\lettrine{T}{\,\ls{he}} anguish of my knee, continued the
corporal, was excessive in itself; and the uneasiness of the
cart, with the roughness of the roads, which were terribly cut
up\tsk making bad still worse\tsk every step was death to me: so
that with the loss of blood, and the\pb want of care-taking of
me, and a fever I~felt coming on besides\tsh (Poor soul!\break
said my uncle Toby) all together, an’ please your honour, was
more than I could sustain.

I was telling my sufferings to a young woman at a
peasant’s house, where our cart, which was the last of the
line, had halted; they had help’d me in, and the young woman
had taken a cordial out of her pocket and dropp’d it upon
some sugar, and seeing it had cheer’d me, she had given it me
a second and a third time\tsh So I was telling her,
an’ please your honour, the anguish I was in, and was saying
it was so intolerable to me, that I had much rather lie down upon
the bed, turning my face towards one which was in the corner of the
room\tsk and die,\pb than go on\tsh when, upon her
attempting to lead me to it, I fainted away in her arms. She was a
good soul! as your honour, said the corporal, wiping his eyes, will
hear.

I thought \textit{love} had been a joyous thing, quoth my uncle
Toby.

’Tis the most serious thing, an’ please your honour
(sometimes) that is in the world.

By the persuasion of the young woman, continued the corporal,
the cart with the wounded men set off without me: she had assured
them I should expire immediately if I was put into the cart. So
when I came to myself\tsh I found myself in a still quiet
cottage, with\pb no one but the young woman, and the peasant and his
wife. I was laid across the bed in the corner of the room, with my
wounded leg upon a chair, and the young woman beside me, holding
the corner of her handkerchief dipp’d in vinegar to my nose
with one hand, and rubbing my temples with the other.

I took her at first for the daughter of the peasant (for it was
no inn)\tsk so had offer’d her a little purse with eighteen
florins, which my poor brother Tom (here Trim
wip’d his eyes) had sent me as a token, by a recruit, just before he set out for
Lisbon\tsh

\tsh I never told your honour that piteous story
yet\tsh here Trim wiped his eyes a third time.

The young woman call’d the old man and his wife into the
room, to shew them the money, in order to gain me credit for a bed
and what little necessaries I should want, till I should be in a
condition to be got to the hospital\tsh Come then! said
she, tying up the little purse\tsk I’ll be your
banker\tsk but as that office alone will not keep me
employ’d, I’ll be your nurse too.

I thought by her manner of speaking\break this, as well as by her
dress, which~I then began to consider more
attentively\tsh that the young woman could not be the
daughter of the peasant.

She was in black down to her toes, with her hair conceal’d
under a cambrick border, laid close to her forehead: she\pb was one of
those kind of nuns, an’ please your honour, of which, your
honour knows, there are a good many in Flanders,
which they let go loose\tsh By thy description,
Trim, said my uncle Toby, I dare say she was a young
Beguine, of which there are none to be found any where but
in the Spanish Netherlands\tsk\break
\stick{except at Amsterdam\tsh they differ}
\stick{from nuns in this, that they can quit}
their cloister if they choose to marry; they visit
and take care of the sick by profession\tsh I had rather,
for my own part, they did it out of good-nature.

\tsh She often told me, quoth Trim, she did it
for the love of Christ\tsk I did not like it.\tsh I
believe, Trim, we are both wrong, said my uncle
Toby\tsk we’ll\pb ask Mr.\@ Yorick about it
to-night at my brother Shandy’s\tsh so put me
in mind; added my uncle Toby.

The young Beguine, continued the corporal, had scarce
given herself time to tell me \lqq she would be my nurse,”\break
when she hastily turned about to begin the office of one, and
prepare something for me\tsh and in a short
time\tsk though I\break thought it a long one\tsk she came back with
flannels, \etc \etc and having fomented my knee soundly for a
couple of hours, \etc and made me a thin basin of gruel for my
supper\tsk she wish’d me rest, and promised to be with me
early in the morning.\tsh She wish’d me, an’
please your honour, what was not to be had.\pb 
\stick{My fever ran very high that night\tsk}
\stick{her figure made sad disturbance within}
\stick{me\tsk I was every moment cutting the} 
\stick{world in two\tsk to give her half of it\tsk}
and every moment was I crying, That I~had nothing but a
knapsack and eighteen florins to share with her\tsh The
whole night long was the fair Beguine, like an angel, close
by my bed-side, holding back my curtain and offering me
cordials\tsk and I was only awakened from my dream by her coming
there at the hour promised, and giving them in reality. In truth,
she was scarce ever from me; and so accustomed was I to receive
life from her hands, that my heart sickened, and I lost colour when
she left the room: and yet, continued the corporal \pb
(making one of the strangest reflections upon it in
the world)\tsh

\tsh \lqq\textit{It was not love}”\tsh for du-\break
\stick{ring the three weeks she was almost con-}
\stick{stantly with me, fomenting my knee}
\stick{with her hand, night and day\tsk I can}
\stick{honestly say, an’ please your honour\tsk}
\stick{that\enspace \starfill}
\hbox to 0.7\hsize{\starfill\enspace once.}

That was very odd, Trim, quoth my uncle Toby\tsh

I think so too\tsk said Mrs.\@ Wadman.

It never did, said the corporal.

\newpage
\section{\lsss{CHAP}.\enspace XXI.}

\quad\tsh But ’tis no marvel, continued\break 
\stick{the corporal\tsk seeing my uncle Toby}
\stick{musing upon it\tsk for Love, an’ please} 
\stick{your honour, is exactly like war, in this;}
\stick{that a soldier, though he has escaped}
\stick{three weeks complete o’\,Saturday-night,}
\tsk may nevertheless be shot
through his heart on Sunday morning\tsh \textit{It
happened so here}, an’ please your honour, with this
difference only\tsk that it was on Sunday in the afternoon, when I fell in love all
at once with a sisserara\tsh it burst upon me, an’
please your honour, like a bomb\tsh scarce giving me time
to say,\break\lqq God bless me.”

\newpage
I thought, Trim, said my uncle\break
Toby, a man never fell in love so very\break
suddenly.

Yes, an’ please your honour, if\break
he is in the way of it\tsh replied\break
Trim.

I prithee, quoth my uncle Toby,\break
inform me how this matter happened.

\tsh With all pleasure, said the corporal, making a
bow.

\newpage
\section{\lsss{CHAP}.\enspace XXII.}

\lettrine{I}{ had} escaped, continued the
corporal, all that time from falling in love, and had gone on to
the end of the chapter, had it not been predestined
otherwise\tsh\break there is no resisting our fate.

It was on a Sunday, in the afternoon, as I told your
honour.

The old man and his wife had walked
out\tsh

Every thing was still and hush as midnight about the
house\tsh

There was not so much as a duck or a duckling about the
yard\tsh

\newpage
\tsh When the fair Beguine came in to see
me.

My wound was then in a fair way of doing well\tsh the
inflammation had been gone off for some time, but it was succeeded
with an itching both above and below my knee, so insufferable, that
I had not shut my eyes the whole night for it.

Let me see it, said she, kneeling down upon the ground parallel
to my knee, and laying her hand upon the part below
it\tsh It only wants rubbing a little, said the
Beguine; so covering it with the bed cloaths, she began with
the forefinger of her right-hand to rub under my knee, guiding her
fore-finger backwards and\pb forwards by the edge of the flannel which
kept on the dressing.

In five or six minutes I felt slightly the end of
her second finger\tsk and presently it was laid flat with the
other, and she continued rubbing in that way round and round for a
good while; it then came into my head, that I should fall in
love\tsk\break I blush’d when I saw how white a hand she
had\tsk I shall never, an’ please your honour, behold
another hand so white whilst I live\tsh

\tsh Not in that place: said my uncle
Toby\tsh

Though it was the most serious des\-pair in nature to the
corporal\tsk he could not forbear smiling.

The young Beguine, continued the corporal, perceiving it
was of great service to me\tsk from rubbing for some time, with
two fingers\tsk proceeded to rub at length, with three\tsk till
by little and little she brought down the fourth, and then
rubb’d with her whole hand: I will never say another word,
an’ please your honour, upon hands again\tsk but it was
softer than satin\tsh

\tsh Prithee, Trim, commend it as much as thou
wilt, said my uncle Toby; I shall hear thy story with the
more delight\tsh The corporal thank’d his master
most unfeignedly; but having nothing to say upon the
Beguine’s hand, but the same over again\tsh he
proceeded to the effects of it.

\newpage
The fair Beguine, said the corporal, continued rubbing with her whole
hand under my knee\tsk till I fear’d her zeal would weary
her\tsh \lqq I would do a\break
\lqq thousand times more,” said she, \lqq for\break
\lqq the love of Christ”\tsh In saying which,\break
she pass’d her hand across the flannel, to the part above my
knee, which I had equally complain’d of, and rubb’d it also.


I perceiv’d, then, I was beginning to be in
love\tsh

As she continued rub-rub-rubbing\tsk I felt it spread from
under her hand, an’ please your honour, to every part of my
frame\tsh

The more she rubb’d, and the longer strokes she
took\tsh the more the fire\pb kindled in my veins \tsh till at length, by
two or three strokes longer than the\break rest\tsh my passion
rose to the highest pitch\tsh I seiz’d her
hand\tsh

\tsh And then thou clapped’st it to\break thy lips,
Trim, said my uncle Toby\tsh\break and madest a
speech.

Whether the corporal’s amour terminated precisely in the
way my uncle Toby described it, is not material; it is
enough that it contain'd in it the essence of all the love-romances
which ever have been wrote since the beginning of the world.

\newpage
\section{\lsss{CHAP}.\enspace XXIII.}

\lettrine{A}{s} soon as the corporal had finished\break
the story of his amour\tsh or rather my uncle Toby for
him\tsk Mrs.\break Wadman silently sallied forth from her
arbour, replaced the pin in her mob, pass’d the wicker gate,
and advanced slowly towards my uncle Toby’s
sentry-box: the disposition which Trim had made in my uncle
Toby’s mind, was too favourable a crisis to be let
slipp’d\tsh

\tsh The attack was determin’d upon: it was facilitated still more by my uncle
Toby’s having ordered the corporal to wheel off the pioneer’s shovel, the spade, the
pick-axe, the picquets, and other\pb military stores which lay scatter’d upon the
ground where Dunkirk stood\tsk The 
\stick{corporal had march’d\tsk the field was clear.\hss}

Now consider, sir, what nonsense it is, either in fighting, or
writing, or any thing else (whether in rhyme to it, or not) which a
man has occasion to do\tsk\break to act by plan: for if ever Plan,
independent of all circumstances, deserved registering in letters
of gold (I mean in the archives of Gotham)\tsk it was
certainly the \textsc{Plan} of Mrs.\@ Wadman’s
attack of my uncle Toby in his sentry-box, \textsc{by}
\textsc{Plan}\tsh Now the plan hanging up in it at
this juncture, being the Plan of Dunkirk\tsk and the tale
of Dunkirk a tale of relaxation, it opposed every impression
she could make: and besides, could she have gone upon it\tsk the
manœuvre of\pb fingers and hands in the attack of the
sentry-box, was so outdone by that of the fair Beguine’s, in Trim’s
story\tsk that just then, that particular attack, however
successful before\tsk became the most heartless attack that could
be made\tsh

O! let woman alone for this. Mrs.\@ Wadman had scarce
open’d the wicker-gate, when her genius sported with the
change of circumstances.

\tsh She formed a new attack in a moment.

\newpage
\section{\lsss{CHAP}.\enspace XXIV.}

\quad\tsh I am half distracted, captain Shandy, said
Mrs.\@ Wadman, holding up her cambrick handkerchief to her
left eye, as she approach’d the door of my uncle
Toby’s sentry-box\tsh a mote \tsh\break
\stick{or sand\tsh or something \tsh I know not}
\stick{what, has got into this eye of mine\tsh}
do look into it\tsk it is not in the white\tsk 

In saying which, Mrs.\@ Wadman edged herself close in
beside my uncle Toby, and squeezing herself down upon the
corner of his bench, she gave him an opportunity of doing it without rising up 
\tsh Do look into it\tsk said she.

\newpage
Honest soul! thou didst look into it with as much innocency of
heart, as ever child look’d into a raree-shew-box; and
’twere as much a sin to have hurt thee.

\tsh If a man will be peeping of his own accord into
things of that nature\tsh I’ve nothing to say to
it\tsh


My uncle Toby never did: and I will answer for him, that he would have sat quietly
upon a sopha from June to January (which, you know, takes in both the hot and cold
months), with an eye as fine as the Thracian \starnote{\quad Rodope Thracia tam
inevitabili fascino instructa, tam exacte oculus intuens attraxit, ut si in illam
quis incidisset, fieri non posset, quin caperetur.\tsh I know not who.} Rodope’s
besides\pb him, without being able to tell, whether it was a black or blue one.

The difficulty was to get my uncle Toby, to look at one,
at all.

’Tis surmounted. And

I see him yonder with his pipe pendulous in his hand, and the
ashes falling out of it\tsk looking\tsk and looking\tsk then
rubbing his eyes\tsh and looking again, with twice the
good-nature that ever Galileo look’d for a spot in the
sun.

\tsh In vain! for by all the powers\break
which animate the organ\tsk Widow\break
Wadman’s left eye shines this
moment as lucid as her right\tsk there is neither mote, or
sand, or dust, or chaff, or speck, or particle of opake matter
floating in it\pb\tsk there is nothing, my dear paternal uncle! but
one lambent delicious fire, furtively shooting out from every part
of it, in all directions, into thine\tsh

\tsh If thou lookest, uncle Toby, in\break
search of this mote one moment longer\break
\tsh thou art undone.


\section{\lsss{CHAP}.\enspace XXV.}

\lettrine{A}{n eye} is for all the world exactly like a cannon, in this respect;
That it is not so much the eye or the cannon, in themselves, as it is the carriage
of the eye\tsh and the carriage of the cannon, by which both the one and the other
are enabled to do so much execution. I don’t think the comparison a bad one:
However, as ’tis made and placed at the\pb head of the chapter, as much for use as
ornament, all I desire in return, is, that whenever I speak of Mrs.\@ Wadman’s eyes
(except once in the next period), that you keep it in your fancy.

I protest, Madam, said my uncle\break Toby, I can see nothing
whatever in\break your eye.

It is not in the white; said Mrs Wadman: my uncle
Toby look’d with might and main into the
pupil\tsh

Now of all the eyes which ever were created\tsh from
your own, Madam, up to those of Venus herself, which
certainly were as venereal a pair of eyes as ever stood in a
head\tsh there never was an eye of them all, so fitted to
rob my uncle\pb Toby of his repose, as the very eye, at which
he was looking\tsh it was not, Madam a rolling
eye\tsh a romping or\break a wanton one\tsk nor was it an eye
spark\-ling\tsk petulant or imperious\tsk of high claims and
terrifying exactions, which would have curdled at once that milk of human
nature, of which my uncle Toby was made up\tsh but
’twas an eye full of\break gentle salutations\tsh and soft
responses\break\tsh speaking \tsh not like the trumpet\break
stop of some ill-made organ, in which many an eye I talk to, holds
coarse converse\tsh but whispering soft \tsh like
the last low accent of an expiring
saint\tsh\break 
\lqq How can you live comfortless, cap-\break
\lqq tain Shandy, and alone, without a\break
\lqq bosom to lean your head on\tsh or\break
\lqq trust your cares to?\,”

\newpage
It was an eye\tsh

But I shall be in love with it myself,\break
if I say another word about it.

\tsh It did my uncle Toby’s business.

\section{\lsss{CHAP}.\enspace XXVI.}

\lettrine{T}{\,\ls{here}} is nothing shews the character
of my father and my uncle Toby, in a more entertaining
light, than their different manner of deportment, under the same
accident\tsh for I call not love a misfortune, from a
persuasion, that a man’s heart is ever the better for
it\break
\tsh Great God! what must my uncle Toby’s
have been, when ’twas all benignity without it.

\newpage
My father, as appears from many of his papers, was very subject
to this passion, before he married\tsh but from a little
subacid kind of drollish impatience in his nature, whenever it
befell him, he would never submit to it like a christian; but would
pish, and huff, and bounce, and kick, and play the Devil, and write
the bitterest Philippicks against the eye that ever man
wrote\tsh there is one in verse upon some body’s eye
or other, that for two or three nights together, had put him by his
rest; which in his first transport of resentment against it, he
begins thus:

\bigskip
\vbox{\fontsize{8}{10}\selectfont\halign{\indent\lqq #\cr
A Devil ’tis\tsk and mischief such doth work\cr
As never yet did Pagan, Jew, or Turk.”\enspace\fnast\hfill\cr}}

\indent
\bgroup\fontsize{9}{11}\selectfont
\fnast\enspace This will be printed with my
father’s\break life of Socrates, \etc \etc\par
\egroup

\newpage
In short, during the whole paroxism, my father was all abuse and foul language,
approaching rather towards malediction\tsh only he did not do it with as much method
as Ernulphus\tsh he was\break too impetuous; nor with Ernulphus’s policy\tsh for tho’ my
father, with the most intolerant spirit, would curse both this and that, and every
thing under heaven, which was either aiding or abetting to his love\tsh yet never
concluded his chapter of curses upon it, without cursing himself in at the bargain,
as one of the most egregious fools and cox-\break combs, he would say, that ever was let
loose in the world.

My uncle Toby, on the contrary, took it like a
lamb\tsh sat still and let the\pb poison work in his veins
without resistance\tsh in the sharpest exacerbations of
his wound (like that on his groin) he never dropt one fretful or
discontented word\tsh he blamed neither heaven nor
earth\tsh or thought or spoke an inju\-rious thing of any
body, or any part of it; he sat solitary and pensive with his
pipe\tsh looking at his lame leg \tsh\break 
\stick{then whiffing out a sentimental heigh ho!}
which mixing with the smoak, incommoded no one
mortal.

He took it like a lamb\tsh I say.


In truth he had mistook it at first; for having taken a ride with my father, that
very morning, to save if possible a beautiful wood, which the dean and chapter\pb were
hewing down to give to the poor\,\starnote{\enspace Mr Shandy must mean the poor \textit{in
spirit}; inasmuch as they divided the money amongst themselves.}; which said wood
being in full view of my uncle Toby’s house, and of singular service to him in his
description of the battle of Wynnendale\tsk by trotting on too hastily to save it\tsh
upon an uneasy saddle\tsh worse horse, \etc \etc .\@ .\@  it had\break so happened, that the
serous part of the blood had got betwixt the two skins, in the nethermost part of my
uncle Toby\break\tsk the first shootings of which (as my uncle Toby had no
experience of love) he had taken for a part of the passion\tsk\break till the blister
breaking in the one case\tsk\break and the other remaining\tsk my uncle Toby was presently
convinced, that his\pb wound was not a skin-deep wound\tsh\break but that it had
gone to his heart.


\section{\lsss{CHAP}.\enspace XXVII.}

\lettrine{T}{\,\ls{he}} world is ashamed of being
vir\-tuous\tsh My uncle Toby knew little of the
world; and therefore when he felt he was in love with widow
Wadman, he had no conception that the thing was any more to
be made a mystery of, than if Mrs.\@ Wadman had given him a
cut with a gap’d knife across his finger: Had it been
otherwise\tsh yet as\break he ever look’d upon Trim
as a humble friend; and saw fresh reasons every day of his life, to
treat him as such\tsh it would have made no variation in
the manner in which he informed him of the\break affair.

\newpage
\lqq I am in love, corporal!” quoth my uncle
Toby.

\section{\lsss{CHAP}.\enspace XXVIII.}

\lettrine{I}{\,n} love!\tsh said the
corporal\tsk your honour was very well the day before yesterday,
when I was telling your honour of the story of the King of
Bohemia\break\null\tsk Bohemia! said my uncle Toby - - - -\break
musing a long time - - -\enspace What became of that story,
Trim?

\tsh We lost it, an’ please your honour, somehow betwixt
us\tsk but your honour was as free from love then, as I
am\tsh\break ’twas, just whilst thou went’st off with
the wheel-barrow\tsh with Mrs.\@ Wadman, quoth my
uncle Toby\tsh She has left a\pb
\stick{ball here\tsk added
my uncle Toby\tsk} pointing to his breast\tsh

\tsh She can no more, an’ please your\break
honour, stand a siege, than she can fly\tsk\break
cried the corporal\tsh

\tsh But as we are neighbours, Trim,\break
\tsk the
best way I think is to let her know it civilly first\tsk quoth my
uncle Toby.

Now if I might presume, said the corporal, to differ from your
honour\tsh

\tsh Why else, do I talk to thee Trim: said my uncle
Toby, mildly\tsh

\tsh Then I would begin, an’ please your honour, with
making a good thundering attack upon her, in return\tsk and
telling her civilly afterwards\tsk for if she knows\pb any thing of
your honour’s being in love, before
hand\tsh L\tsk d help her!\tsk she\break knows no more at
present of it, Trim, said my uncle Toby\tsk than
the child\break
unborn\tsh

Precious souls!\tsh

Mrs.\@ Wadman had told it, with all its circumstances, to
Mrs.\@ Bridget twenty-four hours before; and was at that very\break
moment sitting in council with her, touching some slight misgivings
with regard to the issue of the affair, which the Devil, who never
lies dead in a ditch, had put into her head\tsk before he would
allow half time, to get quietly through her, \textit{te Deum}\tsh

I am terribly afraid, said widow Wadman, in case I should
marry him, Bridget\pb 
\tsk that the poor captain will not enjoy his
health, with the monstrous wound upon his groin\tsh

It may not, Madam, be so very large, 
\stick{replied Bridget, as you think\tsh and I}
\stick{believe, besides, added she\tsk that ’tis dried}
up\tsh

\tsh I could like to know\tsk merely for his sake,
said Mrs.\@ Wadman\tsh

\tsk We’ll know and long and the broad of it, in ten
days\tsk answered Mrs.\@ Bridget, for whilst the captain is
paying his addresses to you\tsk I’m confident Mr.
Trim will be for making love to me\tsk and I’ll let
him as much as he will\tsk added\break
Bridget\tsk to get it
all out of him\tsh

The measures were taken at once\tsh\break and my uncle
Toby and the corporal\break went on with theirs.

\newpage
Now, quoth the corporal, setting his left hand a kimbo, and
giving such a flourish with his right, as just promised
success\tsk and no more\tsh if your honour will give me
leave to lay down the plan of this attack\tsh

\tsh Thou wilt please me by it, Trim, said my
uncle Toby, exceedingly\tsk and as I foresee thou must act
in it as my \textit{aid de camp}, here’s a crown, corporal, to
begin with, to steep thy commission.

Then, an’ please your honour, said the\break
\stick{corporal (making a bow first for his com-}\break
\stick{mission)\tsk we will begin with getting your}\break
honour’s laced cloaths out of the great campaign-trunk, to be
well-air’d, and have the blue and gold taken up at the
sleeves\tsk and I’ll put your white ramallie-\pb 
wig fresh into
pipes\tsk and send for a taylor, to have your honour’s thin
scarlet breeches turn’d\tsh

\tsh I had better take the red plush ones,\break
quoth my uncle Toby\tsk They will be\break
too clumsy\tsk said the corporal.

\section{\lsss{CHAP}.\enspace XXIX.}

\quad\tsh Thou wilt get a brush and a little chalk to my
sword\tsh ’Twill be only in your honour’s way,
replied Trim.

\section{\lsss{CHAP}.\enspace XXX.}

\quad\tsk But your honour’s two razors shall be new
set\tsk and I will get my Montero cap furbish’d up,
and put on poor lieutenant Le Fever’s regimental coat,
which your honour gave me to\pb 
wear for his sake\tsk and as soon as your honour is clean
shaved\tsk and has got your clean shirt on, with your blue and
gold, or your fine scarlet\tsh sometimes one and sometimes
t’other\tsk and every thing is ready for the attack\tsk we’ll
march up boldly, as if ’twas to the face of a bastion; and
whilst your honour engages Mrs.\@  Wadman in the parlour, to the
right\tsh I’ll attack Mrs.\@ Bridget in the kitchen, to the
left; and having seiz’d that pass, I’ll answer for it, said the
corporal, snapping his fingers over his head\break\tsk that the
day is our own.

I wish I may but manage it right; said my uncle
Toby\tsk but I declare, corporal, I had rather march up to
the very edge of a trench\tsh

\newpage
\tsk A woman is quite a different thing\break\tsk said the
corporal.

\tsk I suppose so, quoth my uncle Toby.

\section{\lsss{CHAP}.\enspace XXXI.}

\lettrine{I}{f} any thing in this world, which my\break
father said, could have provoked my uncle Toby, during the
time he was in love, it was the perverse use my father was always
making of an expression of Hilarion the hermit; who, in
speaking of his abstinence, his watchings, flagellations, and other
instrumental parts of his religion\tsk would say\tsk tho’
with more facetiousness than became an
hermit\tsk \lqq That they were the means he used, to make
his \textit{ass} (meaning his body) leave off kicking.”

It pleased my father well; it was not only a laconick way of
expressing\tsh\break
but of libelling, at the same time, the desires and appetites of
the lower part of us; so that for many years of my father’s
life, ’twas his constant mode of expression\tsk he never used
the word \textit{passions} once\tsk but \textit{ass} always
instead of them\tsh So that he might be said truly, to have been
upon the bones, or the back of his own ass, or else of some
other man’s, during all that time.

I must here observe to you the difference betwixt\\
\null\qquad My father’s ass\\
\null\qquad and my hobby-horse\tsk in order to keep
characters as separate as may be, in our fancies as we go
along.

\newpage
For my hobby-horse, if you recollect a little, is no way a
vicious beast; he has scarce one hair or lineament of the ass about
him\tsh ’Tis the sporting little filly-folly which
carries you out for the present hour\tsk a maggot, a butterfly, a
picture, a fiddle-stick\tsk an uncle Toby’s
siege\tsk or an \textit{any thing}, which a man makes a shift to
get a stride on, to canter it away from the cares and solicitudes
of life\tsk ’Tis as useful a beast as is in the whole
creation\tsk nor do I really see how the world could do without
it\tsh

\noindent
\tsh But for my father’s ass\tsh oh!
mount him\tsk mount him\tsk mount him\tsk \break
(that’s three times, is it not?)\tsk mount him not:— ’tis a beast
concupiscent\tsk and foul befall the man, who does not hinder him
from kicking.

\section{\lsss{CHAP}.\enspace XXXII.}

\lettrine{W}{\,\ls{ell}}! dear brother Toby, said\break
my father, upon his first seeing him after he fell in
love\tsk and how goes it with your \textsc{Asse} ?

Now my uncle Toby thinking more 
\stick{of the \textit{part} where he had had the blister,}
than of Hilarion’s metaphor\tsk and our preconceptions having (you know) as great a
power over the sounds of words as the shapes of things, he had imagined, that my
father, who was not very ceremonious in his choice of words, had enquired after the
part by its proper name: so notwithstanding my mother, doctor Slop, and Mr.\@
Yorick, were sitting in the parlour, he thought it rather civil to\pb conform to the
term my father had made use of than not. When a man is hemm’d in by two indecorums,
and must commit one of ’em\tsk I always observe\tsk\break let him choose which he will, the
world will blame him\tsk so I should not be astonished if it blames my uncle Toby.

My A—e, quoth my uncle Toby, is much
better\tsk brother Shandy\tsk My father had formed great
expectations from his Asse in this onset; and would have brought
him on again; but doctor Slop setting up an intemperate
laugh\tsk and my mother crying out L\tsk bless us!\tsk it
drove my father’s Asse off the field\tsk and the laugh then
becoming general\tsk there was no bringing him back to the
charge, for some time\tsh

\newpage
And so the discourse went on without him.

Every body, said my mother, says\break
you are in love, brother Toby\tsk and we\break
hope it is true.

I am as much in love, sister, I believe, replied my uncle
Toby, as any man usually is\tsk Humph! said my
father\tsk\break and when did you know it? quoth my
mother\tsh

\tsh When the blister broke; replied my uncle
Toby.

My uncle Toby’s reply put my father into good
temper\tsk so he charg’d o’ foot.

\section{\lsss{CHAP}.\enspace XXXIII.}

\lettrine{A}{s} the ancients agree, brother Toby, said my father, that there are two
different and distinct kinds of \textit{love}, according to the different parts
which are affected by it\tsk the Brain or Liver\tsh I think when a man is in love, it
behoves him a little to consider which of the two he is fallen into.

What signifies it, brother Shandy, replied my uncle
Toby, which of the two it is, provided it will but make a
man marry, and love his wife, and get a few children?

\tsh A few children! cried my father, rising out of his
chair, and looking full\pb in my mother’s face, as he forced his
way betwixt her’s and doctor Slop’s\tsk a few children! cried my father, repeating my uncle
Toby’s words as he walk’d to and
fro’\tsh

\tsh Not, my dear brother Toby, cried my father, recovering
himself all at once, and coming close up to the back of my uncle
Toby’s chair\tsk not that I should be\break sorry had’st thou a
score\tsk on the contrary, I should rejoice\tsk and be as kind,
Toby, to every one of them as a father\tsk 

My uncle Toby stole his hand unperceived behind his
chair, to give my father’s a squeeze\tsh

\tsh Nay, moreover, continued he,\break
\stick{keeping hold of my uncle Toby’s hand}
\stick{\tsk so much do’st thou possess, my dear}\pb
Toby, of the milk of human nature, and so little of
its asperities\tsk ’tis piteous the world is not peopled by creatures which resemble
thee; and was I an Asia\-tick monarch, added my father, heating himself with his new
project\tsk I would oblige thee, provided it would not impair thy strength\tsk or dry up
thy radical moisture too fast\tsk or weaken thy memory or fancy, brother Toby, which
these gymnicks inordinately taken are apt to do\tsk else, dear Toby, I would procure
thee the most beautiful woman in my empire, and I would oblige thee, \textit{nolens,
volens}, to beget for me one subject every \textit{month}\tsh

As my father pronounced the last word of the sentence\tsk my
mother took a pinch of snuff.

\newpage
Now I would not, quoth my uncle Toby, get a child,
\textit{nolens, volens}, that is, whether I would or no, to please
the greatest prince upon earth\tsh

\tsh And ’twould be cruel in me, brother
Toby, to compell thee; said my father\tsk but ’tis a
case put to shew thee, that it is not thy begetting a
child\break
\tsk in case thou should’st be able\tsk but the
system of Love and Marriage thou goest upon, which I would set thee
right in\tsh

There is at least, said Yorick, a great deal of reason
and plain sense in captain Shandy’s opinion of love;
and ’tis amongst the ill spent hours of my life, which I have
to answer for, that I have read so many flourishing poets and
rhetoricians in my time, from whom I never could
extract so much\tsh

\newpage
I wish, Yorick, said my father, you\break had read
Plato; for there you would have learnt that there are two
\textsc{Loves}\tsk\break 
\stick{I know there were two \textsc{Religions}, re-}
\stick{plied Yorick, amongst the ancients\tsh}
one\tsk for the vulgar, and another for the
learned;\tsk but I think \textsc{one Love}
might have served both of them very well\tsk 

It could not; replied my father\tsk and for the same reasons:
for of these Loves, according to Ficinus’s comment
upon Velasius, the one is \textit{rational}\tsh\\
\null\quad\tsh the other is \textit{natural}\tsh \\
the first ancient\tsh without mother \tsh\break where
Venus had nothing to do: the second, begotten of
Jupiter and Dione\tsk

\tsh Pray, brother, quoth my uncle Toby, what
has a man who believes in God to do with this? My father could\pb
\bgroup\parskip=3pt plus 3pt%
not
stop to answer, for fear of breaking the thread of his
discourse\tsh

This latter, continued he, partakes wholly of the nature of
Venus.

The first, which is the golden chain let down from heaven,
excites to love heroic, which comprehends in it, and excites to the
desire of philosophy and truth\tsh the second, excites to
\textit{desire}, simply\tsh

\tsh I think the procreation of children as beneficial
to the world, said Yorick, as the finding out the
longitude\tsh

\tsh To be sure, said my mother, \textit{love} keeps
peace in the world\tsh

\tsh In the \textit{house}\tsk my dear, I
own\tsk \break
\null\tsh It replenishes the earth; said my
mother\tsh

But it keeps heaven empty\tsk my dear; replied my father.
\egroup
\newpage
\tsk ’Tis Virginity, cried Slop,
triumphantly, which fills paradise.

Well push’d nun! quoth my father.

\section{\lsss{CHAP}.\enspace XXXIV.}

\lettrine{M}{y} father had such a skirmishing, cutting kind of a slashing way 
\stick{with him in his disputations, thrusting} 
\stick{and ripping, and giving every one a}
\stick{stroke to remember him by in his turn\tsk}
\stick{that if there were twenty people in com-}
\stick{pany\tsk in less than half an hour he was}
\stick{\tight{sure to have every one of ’em against him.}\hss}

\noindent
\stick{\indent\tight{What did not a little contribute to leave}}
\stick{him thus without an ally, was, that if}
\stick{there was any one post more untenable}
\stick{than the rest, he would be sure to throw}
\stick{himself into it; and to do him justice,}
\stick{when he was once there, he would defend}\pb it so gallantly,
that ’twould have been a concern, either to a brave man or
a\break
good-natured one, to have seen him\break
driven out.

Yorick, for this reason, though he would often attack
him\tsk yet could never bear to do it with all his force.

Doctor Slop’s \textsc{Virginity}, in the
close of the last chapter, had got him for once on the right side
of the rampart; and he was beginning to blow up all the convents in
Christendom about Slop’s ears, when corporal
Trim came into the parlour to inform my uncle Toby,
that his thin scarlet breeches, in which the attack was to be made
upon Mrs.\@ Wadman, would not do; for that the taylor, in
ripping them up, in order to turn them, had found they had been
turn’d\pb before\tsh Then turn them again, bro-\break ther,
said my father, rapidly, for there will be many a turning of
’em yet before all’s done in the
affair\tsh They are as rotten as dirt, said the
corporal\tsh\break Then by all means, said my father, bespeak a
new pair, brother\tsh for though I know, continued my
father, turning himself to the company, that widow Wadman
has been deeply in love with my brother Toby for many years,
and has used every art and circumvention of woman to outwit him
into the same passion, yet now that she has caught
him\tsh her fever will be pass’d it’s\break
height\tsh

\tsh She has gain’d her point.

In this case, continued my father, which Plato, I am
persuaded, never\pb thought of\tsh Love, you see, is not so
much a \textsc{Sentiment} as a \textsc{Situation}, into
which a man enters, as my brother Toby would do, into a
\textit{corps}\tsh no matter whether he loves the service or
no\tsh being once in it\tsk he acts as if he did; and
takes every step to shew himself a man of prowesse.

The hypothesis, like the rest of my\break father’s, was
plausible enough, and my uncle Toby had but a single word to
object to it\tsk in which Trim stood ready to second
him\tsh but my father had not drawn his
conclusion\tsh

For this reason, continued my father (stating the case over
again) notwithstanding all the world knows, that Mrs.\pb
Wadman \textit{affects} my brother Toby\tsk and my brother
Toby contrariwise \textit{affects} Mrs.\@ Wadman, and no
obstacle in nature to forbid the music striking up this very night,
yet will I answer for it, that this self-same tune will not be
play’d this twelvemonth.

We have taken our measures badly, quoth my uncle Toby,
looking up interrogatively in Trim’s face.

I would lay my Montero-cap, said
Trim\tsh Now Trim’s
Montero-cap, as\break
\stick{I once told you, was his constant wager;}
and having furbish’d it up that very night, in
order to go upon the attack\tsk it made the odds look more
considerable\break\null\tsh I would lay, an’ please your
honour, my Montero-cap to a shilling\tsk \pb was it proper,
continued Trim (making a bow), to offer a wager before your
honours\tsh

\tsk There is nothing improper in it,\break said my
father\tsk ’tis a mode of expression; for in saying thou
would’st lay thy Montero-cap to a shilling\tsk all
thou meanest is this\tsk that thou believest\tsh

\tsh Now, What do’st thou believe?

That Widow Wadman, an’ please your worship, cannot
hold it out ten days\tsh

\noindent
\stick{\hfill And whence, cried Slop, jeeringly, hast}
\stick{\tight{thou all this knowledge of woman, friend?}\hss}

By falling in love with a popish clergy-woman; said
Trim.

’Twas a Beguine, said my uncle Toby.

\newpage
Doctor Slop was too much in wrath to listen to the
distinction; and my father taking that very crisis to fall in
helter-skelter upon the whole order of Nuns and Beguines, a
set of silly, fusty, bag-\break gages\tsh Slop could not stand
it\tsh\break and my uncle Toby having some measures to take
about his breeches\tsk and Yorick about his fourth general
division\break\tsk in order for their several attacks next
day\tsk the company broke up: and my father being left alone, and
having half an hour upon his hands betwixt that and bed-time; he
called for pen, ink, and paper, and wrote my uncle Toby the
following letter of instructions.

\newpage
\setlength{\baselineskip}{13.36pt} % 21*14 = 294/22 \simeq 13.3636 
My dear brother Toby,

\vskip -6pt

\lettrine{W}{\,\ls{hat}} I am going to say to\break thee is
upon the nature of wo-\break men, and of love-making to them; and perhaps
it is as well for thee\tsk tho’ not so well for
me\tsk that thou hast occasion for a letter of instructions upon
that head, and that I am able to write it to\break thee.

Had it been the good pleasure of him who disposes of our
lots\tsk and thou no sufferer by the knowledge, I had been well content that thou should’st have
dipp’d the pen this moment into the ink, instead of myself;
but that not being the case\tsh\tsk Mrs Shandy
being now close beside me, preparing for bed\tsh\pb 
I have thrown together without order, and just as they have come into my mind, such
hints and documents as I deem may be of use to thee; intending, in this, to give
thee a token of my love; not doubting, my dear Toby, of the manner in which it will
be accepted.

In the first place, with regard to all which concerns religion
in the affair\tsh\break though I perceive from a glow in my
cheek, that I blush as I begin to speak to thee upon the subject,
as well knowing, notwithstanding thy unaffected secrecy, how few of
its offices thou neglectest\tsk yet I would remind thee of one
(during the continuance of thy courtship) in a particular manner,
which I would not have omitted; and that is, never to go forth upon
the enterprize, whether it be in the morning or the afternoon,
with-\pb
out first recommending thyself to the protection of Almighty
God, that he may defend thee from the evil one.

Shave the whole top of thy crown clean once at least every four
or five days, but oftner if convenient; lest in taking off thy wig
before her, thro’ absence of mind, she should be able to
discover how much has been cut away by Time\tsh how much
by Trim.

\tsh ’Twere better to keep ideas of baldness out of her
fancy.

Always carry it in thy mind, and act upon it as a sure maxim,
Toby\tsh

\lqq\textit{That women are timid:}” And ’tis well they
are\tsh else there would be no dealing with them.

\newpage
Let not thy breeches be too tight, or hang too loose about thy
thighs, like the trunk-hose of our ancestors.

\tsh A just medium prevents all conclusions.

Whatever thou hast to say, be it more or less, forget not to
utter it in a low soft tone of voice. Silence, and whatever
approaches it, weaves dreams of midnight secrecy into the brain:
For this cause, if thou canst help it, never throw down the
tongs and poker.

Avoid all kinds of pleasantry and facetiousness in thy discourse
with her, and do whatever lies in thy power at the same time, to
keep her from all books and writings which tend thereto: there are
some devotional tracts, which if thou\pb canst entice her to read
over\tsk it will be well: but suffer her not to look into
Rabelais, or Scarron, or Don Quixote\tsh

\tsh They are all books which excite laughter; and thou
knowest, dear Toby, \tight{that there is no passion so serious, as
lust.}

Stick a pin in the bosom of thy shirt, before thou enterest her
parlour.

And if thou art permitted to sit upon the same sopha with her,
and she gives thee occasion to lay thy hand upon hers\tsk beware
of taking it\tsh thou can’st not lay thy hand on hers, but she
will feel the temper of thine. Leave that and as many other
things as thou canst, quite undetermined; by so doing, thou wilt
have her curiosity on thy side; and if she is not conquer’d by
that, and thy \textsc{Asse}\pb continues still kicking, which
there is great reason to suppose\tsh Thou must begin, with first
losing a few ounces of blood below the ears, according to the
practice of the ancient Scythians, who cured the most
intemperate fits of the appetite by that means.

Avicenna, after this, is for having the\break part anointed
with the syrrup of helle\-bore, using proper evacuations and
purges\tsh and I believe rightly. But thou must eat little
or no goat’s flesh,\break nor red deer\tsh nor even
foal’s flesh by\break any means; and carefully
abstain\tsh that is, as much as thou canst, from peacocks,
cranes, coots, didappers, and water-hens\tsh

As for thy drink\tsk I need not tell thee, it must be the
infusion of \textsc{Vervain} and the herb
\textsc{Hanea}, of which Ælian\pb 
\stick{relates such effects\tsk but if thy stomach}
\stick{palls with it\tsk discontinue it from time to}
time, taking cucumbers, melons, purs\-lane, water-lillies, woodbine, and lettice,
in the stead of them.

There is nothing further for thee, which occurs to me at
present\tsh

\tsh Unless the breaking out of a fresh
war\tsh So wishing every thing, dear Toby, for
best,\\[3pt]
\indent I rest thy affectionate brother,\\[3pt]
\rightline{\textsc{Walter Shandy}.}

\section{\lsss{CHAP}.\enspace XXXV.}

\lettrine{W}{\,\ls{hilst}} my father was writing his
letter of instructions, my uncle Toby and the corporal were
busy in preparing every thing for the attack. As the turning of the
thin scarlet\pb breeches was laid aside (at least for the present),
there was nothing which should put it off beyond the next morning;
so accordingly it was resolv’d upon, for eleven
o’clock.

Come, my dear, said my father to my mother\tsk ’twill be
but like a brother and sister, if you and I take a walk down to my
brother Toby’s\tsh to countenance him in this
attack of his.

\vskip -6pt

My uncle Toby and the corporal had been accoutred both some time, when my father and
mother enter’d, and the clock striking eleven, were that moment in motion to sally
forth\tsk but the account of this is worth more than to be wove into the fag end of
the eighth volume of such a work as this.\tsh My father had no time but to put the
letter of in-\pb structions into my uncle Toby’s coat-pocket\tsh and join with my mother
in wishing his attack prosperous.


I could like, said my mother, to look through the key-hole out
of \textit{curiosity}\tsh\break Call it by it’s right name, my
dear,\break quoth my father\tsk 

\textit{And look through the key-hole} as long as you will.

\vfill
\centerline{\textsc{End} of the \ls{\textsc{Eighth Volume}}.}
\vfill

\end{document}
