\documentclass{article}
\usepackage[a6paper, width=166pt, height=294pt]{geometry}
\usepackage{graphicx}
\usepackage{shandean}
\title{The Life and Opinions of Tristram Shandy, Gentleman}
\author{Laurence Sterne}
\date{Vol 1}
\begin{document}
\pagestyle{fancy}
\thispagestyle{empty}
\[\vbox{\openup 6pt\halign{\hss # \hss\cr
\addfontfeature{LetterSpace=12.0}\textsc{the}\cr
\addfontfeature{LetterSpace=12.0}LIFE and OPINIONS\cr
\addfontfeature{LetterSpace=12.0}\textsc{of}\cr
{\addfontfeature{LetterSpace=16.0}TRISTRAM SHANDY}, Gent.\cr}}\]

\vskip 6pt
\setlength{\baselineskip}{14pt}  % 21*14 = 294 

\section{Chap. I.}

\lettrine{I}{ have} begun a new book, on purpose that I might
have room enough to explain the nature of the perplexities in
which my uncle \textit{Toby} was involved, from the many
discourses and interrogations about the siege of \textit{Namur},
where he received his wound.

I must remind the reader, in case he has read the history of
King \textit{William}’s\pb 
\stick{wars,\tsk but if he has not,\tsk I then inform}
\stick{him, that one of the most memorable at-}
\stick{tacks in that siege, was that which was}
\stick{made by the \textit{English} and \textit{Dutch} upon the}
point of the advanced counterscarp,
be\-tween the gate of \textit{St.\@ Nicolas}, which inclo\-sed the
great sluice or water-stop, where the \textit{English} were
terribly exposed to the shot of the counter-guard and
demi-\break
\stick{bastion of \textit{St.\@ Roch:} The issue of which}
hot dispute, in three words, was this;\break That the \textit{Dutch}
lodged themselves upon the counter-guard,\tsk and that the
\textit{English} made themselves masters of the covered way
before \textit{St.\@ Nicolas}'s gate, notwithstanding the gallantry
of the \textit{French} offi\-cers, who exposed themselves upon the\break
glacis sword in hand.

As this was the principal attack of which my uncle \textit{Toby}
was an eye-witness\pb at \textit{Namur},\tsh the army of the
besiegers being cut off, by the confluence of the \textit{Maes}
and \textit{Sambre}, from seeing much of each other’s
operations,\tsk my uncle \textit{Toby}\break was generally more
eloquent and particular in his account of it; and the many
perplexities he was in, arose out of the\break
almost insurmountable difficulties he\break
found in telling his story intelligibly,\break
and giving such clear ideas of the differences and distinctions
between the scarp and counterscarp,\tsh the glacis and
covered way,\tsh the half-moon and rave-\break
lin,\tsh as to make his company fully\break
comprehend where and what he was\break about.

Writers themselves are too apt to confound these terms;\tsh so that
you will the less wonder, if in his endeavours to explain them, and
in opposition to many\pb misconceptions, that my uncle \textit{Toby} did
oft times puzzle his visiters, and sometimes himself too.

To speak the truth, unless the compa\-ny my father led up stairs
were tolerably clear-headed, or my uncle \textit{Toby} was in one of
his explanatory moods, ’twas a difficult thing, do what he could, to keep the
discourse free from obscurity.

What rendered the account of this\break 
affair the more intricate to my uncle\break
\textit{Toby}, was this,\tsk that in the attack of the
counterscarp, before the gate of \textit{St.\@ Nicolas}, extending
itself from the bank of the \textit{Maes}, quite up to the great
water-stop;\tsk the ground was cut and cross-cut with such a
multitude of dykes, drains, rivulets, and sluices, on all
sides,\tsk and he would get so sadly bewilder’d, and set\pb fast
amongst them, that frequently he could neither get backwards or
forwards to save his life; and was oft times obliged to give up the
attack upon that very account only.

These perplexing rebuffs gave my\break
uncle \textit{Toby Shandy} more perturbations\break
than you would imagine; and as my\break
father’s kindness to him was continually\break
dragging up fresh friends and fresh in-\break 
quirers,\tsk he had but a very uneasy task\break of it.

No doubt my uncle \textit{Toby} had great command of
himself,\tsk and could guard appearances, I believe, as well as
most men;\tsk yet any one may imagine, that when he could not
retreat out of the ravelin without getting into the half-moon,
or get out of the covered way without\pb falling down the
counterscarp, nor cross the dyke without danger of slipping into
the ditch, but that he must have fretted and fumed inwardly:\tsk
He did so;\tsk and the little and hourly vexations, which may
seem trifling and of no account to the man who has not read
\textit{Hippocrates}, yet, whoever has read
\textit{Hippocrates}, or Dr.\@ \textit{James Mackenzie},
and has considered well the effects which the passions and
affections of the mind have upon the digestion,\tsk (Why not of a
wound as well as of a dinner?)\tsk may easily conceive what
sharp paroxysms and exacerbations of his wound my uncle
\textit{Toby} must have undergone upon that score only.

\tsk My uncle \textit{Toby} could not philoso\-phize upon
it;\tsk ’twas enough he felt it was so,\tsk and having
sustained the pain and sorrows of it for three months\pb together, he was resolved some
way or other to extricate himself.

He was one morning lying upon his back in his bed, the anguish and nature\break
of the wound upon his groin suffering him to lie in no other position, when a
thought came into his head, that if he could purchase such a thing, and have it
pasted down upon a board, as a large map of the fortification of the town and
citadel of \textit{Namur}, with its environs, it might be a means of giving him
ease.\tsk\break I take notice of his desire to have the environs along with the
town and cita\-del, for this reason,\tsk because my uncle \textit{Toby}’s
wound was got in one of the traverses, about thirty toises from the returning
angle of the trench, opposite to the salient angle of the demi-bastion of
\textit{St.\@ Roch};\tsh so that he was pretty con-\pb fident he could stick a
pin upon the identical spot of ground where he was standing on when the stone
struck him.


All this succeeded to his wishes, and not only freed him from a
world of sad explanations, but, in the end, it proved the happy means, as you will read, of procuring my
uncle \textit{Toby} his \textsc{Hobby-}\break\textsc{Horse}.


\section{Chap. II.}

\lettrine{T}{ here} is nothing so foolish, when you
are at the expence of making an entertainment of this kind, as to
order things so badly, as to let your criticks and gentry of
refined taste run it down: Nor is there any thing so likely\pb
\stick{to make them do it, as that of leaving}
\stick{them out of the party, or, what is full as} 
\stick{offensive, of bestowing your attention}
\stick{upon the rest of your guests in so parti-}
\stick{cular a way, as if there was no such}
thing as a critick (by occupation) at\break table.

\tsh I guard against both; for, in the first place, I have left
half a dozen places purposely open for them;\tsk and in the next
place, I pay them all court,\tsk\break Gentlemen, I kiss your
hands,\tsh I pro-\break test no company could give me half the pleasure,\tsk
by my soul I am glad to see\break you,\tsh I beg only you will make no
strangers of yourselves, but sit down without any ceremony, and
fall on hear-\break tily.

\newpage
I said I had left six places, and I was upon the point of
carrying my complaisance so far, as to have left a seventh open for
them,\tsk and in this very spot I stand on; but being told by a
critick (tho’ not by occupation,\tsk but by nature) that I
had acquitted myself well enough, I shall fill it up directly,
hoping, in the mean time, that I shall be able to make a great deal
of more room next year.

\tsh How, in the name of wonder! could your
uncle \textit{Toby}, who, it seems, was a military man, and whom you
have represented as no fool,\tsk be at the same time such a
confused, pudding-headed, muddle-headed, fellow, as\tsk Go
look.

So, Sir Critick, I could have replied; but I scorn it.\tsh ’Tis
language un\-urbane,\tsk and only befitting the man\pb who
cannot give clear and satisfactory accounts of things, or dive
deep enough into the first causes of human ignorance and
confusion. It is moreover the reply valiant\tsk and therefore I
reject it; for tho’ it might have suited my uncle
\textit{Toby}’s character as a soldier excellently well,\tsk and
had he not accustomed himself, in such attacks, to whistle the
\textit{Lilla\-bullero}, as he wanted no courage, ’tis the very
answer he would have given; yet it would by no means have done
for me. You see as plain as can be, that I write as a man of
erudition;\tsk that even my similies, my allusions, my
illustrations, my metaphors, are erudite,\tsk and that I must
sustain my character properly, and contrast it properly too,\tsk
else what would become of me? Why, Sir, I should be undone;\tsk
at this very moment that I am going here to fill up one\pb place
against a critick,\tsk I should have made an opening for a
couple.

\tsh Therefore I answer thus:

Pray, Sir, in all the reading which you have ever read, did you
ever read such a book as \textit{Locke}’s Essay upon the Human
Understanding?\tsh Don’t answer me rashly,\tsk because many, I
know, quote the book, who have not read it,\tsk and many have
read it who understand it not:\tsk If either of these is your
case, as I write to instruct, I will tell you in three words
what the book is.\tsk It is a history.\tsk A\break
\stick{history! of who? what? where? when?}
Don’t hurry yourself.\tsk It is a
history-book, Sir, (which may possibly recommend it to the
world) of what passes in a man’s own mind; and if you will say
so much of the book, and no more, believe\pb me, you will cut no
contemptible figure in a metaphysic circle.

But this by the way.

Now if you will venture to go along with me, and look down into
the bottom of this matter, it will be found that the cause of
obscurity and confusion, in the mind of a man, is threefold.

Dull organs, dear Sir, in the first place. 
\stick{Secondly, slight and transient impressions}
\stick{made by objects when the said organs are}
\stick{not dull. And thirdly, a memory like}
\stick{unto a sieve, not able to retain what it has}
\stick{received.\tsk Call down \textit{Dolly} your cham-}
\stick{ber-maid, and I will give you my cap}
\stick{and bell along with it, if I make not this}
\stick{matter so plain that \textit{Dolly} herself shall}
\stick{understand it as well as \textit{Malbranch.}\tsh}\pb
When \textit{Dolly} has indited her epistle to \textit{Robin},
and has thrust her arm into the bottom of her pocket hanging by
her right side;\tsk take that opportunity to recollect that the
organs and faculties of perception can, by nothing in this
world, be so aptly typified and explained as by that one thing
which \textit{Dolly}’s hand is in search of.\tsk Your organs are
not so dull that I should inform you\tsk ’tis an inch, Sir, of
red seal-wax.

When this is melted and dropped upon the letter, if \textit{Dolly}
fumbles too long for her thimble, till the wax is over harden’d, it
will not receive the mark of her thimble from the usual impulse
which was wont to imprint it. Very well. If \textit{Dolly}’s
wax, for want of better, is bees-wax, or of a temper too
soft,\tsk tho’ it may receive,\tsk it will not hold the
impression,\pb how hard soever \textit{Dolly} thrusts against it; and
last of all, supposing the wax good, and eke the thimble, but
applied thereto in careless haste, as her Mistress rings the bell;\tsh in any one of
these three cases the print left by the thimble will be as unlike
the prototype as a brass-jack.

Now you must understand that not one of these was the true cause
of the confusion in my uncle \textit{Toby}’s discourse; and it
is for that very reason I enlarge upon them so long, after the
manner of great physiologists\tsk to shew the world, what it did
\textit{not} arise from.

What it did arise from, I have hinted above, and a fertile
source of obscurity it is,\tsk and ever will be,\tsk and that
is the unsteady uses of words, which have\pb perplexed the clearest
and most exalted understandings.

It is ten to one (at \textit{Arthur}’s) whether you have
ever read the literary histories of past ages;\tsk if you have,
what terrible battles, ’yclept logomachies, have they
occasioned and perpetuated with so much gall and
ink-shed,\tsk that a good natured man cannot read the accounts of
them without tears in his eyes.

\noindent
\stick{\hfill Gentle critick! when thou hast weigh’d}
\stick{all this, and consider’d within thyself how}
\stick{much of thy own knowledge, discourse,} 
\stick{and conversation has been pestered and}
\stick{disordered, at one time or other, by this,}
\stick{and this only:\tsh What a pudder and}
\stick{racket in \textsc{Councils} about οὐσία and ὑπό-}
\stick{στασις; and in the \textsc{Schools} of the learned}
about power and about spirit;\tsk about\pb
\stick{essences, and about quintessences;\tsh}
\stick{about substances, and about space.\tsh}
\stick{What confusion in greater \ls{\textsc{Theatres}}}
from words of little meaning, and as
indeterminate a sense;\tsk when thou consi\-ders this, thou wilt not
wonder at my uncle \textit{Toby}’s perplexities,\tsk thou
wilt drop a tear of pity upon his scarp and his
counterscarp;\tsk his glacis and his covered-way;\tsk his
ravelin and his half-moon:\break
’Twas not by ideas,\tsh by heaven!\break
his life was put in jeopardy by words.

\bigskip

\section{Chap. III.}

\lettrine{W}{\,\ls{hen}} my uncle \textit{Toby} got his map
of \textit{Namur} to his mind, he began immediately to apply himself,
and with the utmost diligence, to the study of it; for
nothing being of more importance\pb to him than his recovery, and his
recovery depending, as you have read, upon the passions and
affections of his mind, it behoved him to take the nicest care to
make himself so far master of his subject, as to be able to talk
upon it without emotion.

In a fortnight’s close and painful application, which, by
the bye, did my uncle \textit{Toby}’s wound, upon his groin, no
\stick{good,\tsk he was enabled, by the help of}
some marginal documents at the feet of the elephant, together with \textit{Gobesius}’s
mi\-litary architecture and pyroballogy,\break
translated from the
\textit{Flemish}, to form his\break discourse with passable perspicuity;
and\break before he was two full months gone,\tsk\break
he was right eloquent upon it, and could make not only the
attack of the advan-\pb 
ced counterscarp with great order;\tsh\break 
but having, by that time, gone much\break deeper into
the art, than what his first motive made necessary,\tsk my uncle \textit{Toby} was able
to cross the \textit{Maes} and \textit{Sambre}; make diversions as far as
\textit{Vauban}’s line, the abbey of \textit{Salsines}, \etc and give his visiters
as distinct a history of each of their attacks, as of that of the gate of
\textit{St.\@ Nicolas}, where he had the honour to\break receive his wound.

\noindent
\stick{\hfill But the desire of knowledge, like the}
\stick{thirst of riches, increases ever with the ac-}
\stick{quisition of it. The more my uncle \textit{Toby}}
\stick{pored over his map, the more he took a}
liking to it;\tsk by the same process and\break
electrical assimilation, as I told you, thro’\break
which I ween the souls of connoisseurs\break
themselves, by long friction and incum-\break
bition, have the happiness, at length, to\pb get all
be-virtu’d\tsk be-pictur’d,\tsk be-butterflied, and
be-fiddled.

The more my uncle \textit{Toby} drank of this\break
\stick{sweet fountain of science, the greater was}
the heat and impatience of his thirst,\break
so that, before the first year of his con-\break
finement had well gone round, there was\break
\stick{scarce a fortified town in \textit{Italy} or \textit{Flanders},}
of which, by one means or other, he had not
procured a plan, reading over as he got them, and carefully
collating therewith the histories of their sieges, their
demolitions, their improvements, and new works, all which he
would read with that intense application and delight, that he
would forget himself, his wound, his confinement, his dinner.

In the second year my uncle \textit{Toby} purchased \textit{Ramelli}
and \textit{Cataneo}, translated\pb 
\stick{from the \textit{Italian};\tsk likewise \textit{Stevinus},}
\textit{Moralis}, the Chevalier \textit{de Ville}, \textit{Lorini},\break 
\textit{Cochorn}, \textit{Sheeter}, the Count \textit{de Pagan}, 
the Marshal \textit{Vauban}, Mons.\@ \textit{Blondel}, with\break
almost as many more books of military architecture, as Don
\textit{Quixote} was found to have of chivalry, when the
curate and\break
barber invaded his library.

Towards the beginning of the third year, which was in
\textit{August}, ninety-nine, my uncle \textit{Toby} found it necessary
to understand a little of projectiles:\tsk And ha-\break\stick{ving judged it
best to draw his know-}\break ledge from the fountain-head, he began with
\textit{N.\@ Tartaglia}, who it seems was the first man who detected
the imposition of a canon-ball’s doing all that mischief
under the notion of a right line.\tsk This\break
\textit{N.\@ Tartaglia} proved to my uncle \textit{Toby} to be an
impossible thing.

\newpage
\tsh Endless is the Search of Truth!

\noindent
\stick{\hfill No sooner was my uncle \textit{Toby} satisfied}
which road the
cannon-ball did not go, but he was insensibly led on, and resolv\-ed
in his mind to enquire and find out\break which road the ball did go: For
which purpose he was obliged to set off afresh with old
\textit{Maltus}, and studied him devout\-ly.\tsk He proceeded next to
\textit{Galileo} and \textit{Torricellius}, wherein, by certain
geometrical rules, infallibly laid down, he found the precise path
to be a \ls{\textsc{Parabola}},\tsk or else an
\ls{\textsc{Hyperbola}},\tsk and that the parameter, or \textit{latus
rectum}, of the conic section of the said path, was to the
quantity and amplitude in a direct \textit{ratio}, as the whole line
to the sine of double the angle of incidence, form’d by the breech
upon an horizontal plane;\tsk and that the
semi-\pb parameter,\tsh stop! my dear uncle
\textit{Toby}\tsh stop!\tsk go not one foot further\break 
into this thorny and bewilder’d track,\tsk\break
intricate are the steps!  intricate are the\break
mases of this labyrinth! intricate are the\break
\stick{troubles which the pursuit of this bewit-}
\stick{ching phantom \textsc{Knowledge} will bring}
\stick{upon thee.\tsk O my uncle;\tsk fly\tsk fly,\tsk}
\stick{fly from it as from a serpent.\tsh Is it fit,}
\stick{good-natur’d man!\@ thou should’st sit up,}
\stick{with the wound upon thy groin, whole}
\stick{nights baking thy blood with hectic}
\stick{watchings?\tsh Alas! ’twill exasperate thy}
\stick{symptoms,\tsk check thy perspirations\tsk}
\stick{evaporate thy spirits\tsk waste thy animal}
\stick{strength, dry up thy radical moisture,}
\stick{\tsk bring thee into a costive habit of body,}
\stick{impair thy health,\tsh and hasten all the}
\stick{infirmities of thy old age.\tsk O my uncle!}
my uncle \textit{Toby!}

\newpage
\section{Chap. IV.}

\lettrine{I}{ would} not give a groat for that
man’s knowledge in pen-craft, who does not\break 
understand this,\tsk That the best plain
\stick{narrative in the world, tack’d very close}
\stick{to the last spirited apostrophe to my un-}
\stick{cle \textit{Toby},\tsh would have felt both cold and}
vapid upon the reader’s palate;\tsk there\-fore I forthwith put an end to the
chapter,\tsk though I was in the middle of my story.

\tsh Writers of my stamp have one principle in common
with painters.\tsk\break Where an exact copying makes our pictures less
striking, we choose the less evil; deeming it even more pardonable
to trespass against truth, than beauty.\tsk\break This is to be understood
\textit{cum grano salis};\pb but be it as it will,\tsk as the parallel
is made more for the sake of letting the apostrophe cool, than any
thing else,\tsk\break ’tis not very material whether upon any
other score the reader approves of it or not.

In the latter end of the third year, my uncle \textit{Toby}
perceiving that the parameter and semi-parameter of the conic
section angered his wound, he left off the study of projectiles in
a kind of a huff, and betook himself to the practical part of
fortification only; the pleasure of which, like a spring held back,
returned upon him with redoubled force.

It was in this year that my uncle began to break in upon the daily regularity of a clean
shirt,\tsh to dismiss his barber
unshaven,\tsh and to allow his surgeon\pb
scarce time
sufficient to dress his wound, concerning himself so little about
it, as not to ask him once in seven times dressing, how it went on:
When, lo!\tsk all of a sudden, for the change was quick as
lightening, he began to sigh heavily for his
recovery,\tsk complain’d to my father, grew impatient with
the surgeon:\tsk and one morning, as he heard his foot
com-\break ing up stairs, he shut up his books, and thrust aside his
instruments, in order to expostulate with him upon the protraction
of the cure, which, he told him,\break might surely have been
accomplished at least by that time:\tsk He dwelt long upon the
miseries he had undergone, and the sorrows of his four years
melancholy\break
imprisonment;\tsh adding, that had it not been
for the kind looks, and fraternal chearings of the best of
brothers,\tsk he had long since sunk under his
misfor-\pb tunes.\tsh My father was by: My uncle
\textit{Toby}’s eloquence brought tears into his eyes;\tsh ’twas
unexpected.\tsh My uncle \textit{Toby}, by nature was not
eloquent;\tsk\break it had the greater effect:\tsk The surgeon
was confounded;\tsk not that there wanted grounds for
such, or greater, marks of impatience,\tsk but ’twas
unexpected too; in the four years he had attended him, he had never
seen any thing like it in my uncle \textit{Toby}’s carriage;\tsk he
had never once dropp’d one fretful or discontented
word;\tsk he had been all patience,\tsk all
submission.

\tsk We lose the right of complaining sometimes by forbearing
it;\tsk but we often treble the force:\tsk The surgeon was
astonished;\tsk but much more so, when he heard my uncle \textit{Toby} go on,
and peremptorily insist upon his healing up\pb the wound
directly,\tsh or sending for\break Monsieur \textit{Ronjat}, the
King’s Serjeant-\break Surgeon, to do it for him.

The desire of life and health is implanted in man’s
nature;\tsh the love of liberty and enlargement is a
sister-passion to it: These my uncle \textit{Toby} had in common with
his species;\tsk and either of them had been sufficient to account for his earnest
desire to get well and out of doors;\tsk but I have told
you before, that nothing wrought with our family after the common
way;\tsh and from the time and manner in which this eager
desire shew’d itself in the present case, the penetrating reader
will suspect there was some other cause or crotchet for it in my
uncle \textit{Toby}’s head:\tsk There was so, and
’tis the subject of the next chapter to set forth what that
cause and crotchet was.\pb I own, when that’s done, ’twill
be time to return back to the parlour fire-side, where we left my
uncle \textit{Toby} in the middle of his sentence.

\bigskip

\section{Chap. V.}

\lettrine{W}{hen} a man gives himself up to\break 
the government of a ruling pas-\break
\stick{sion,\tsk or, in other words, when his}
\stick{\textsc{Hobby-Horse} grows headstrong,\tsh }
\stick{farewell cool reason and fair discretion!}

\noindent
\stick{\hfill My uncle \textit{Toby}’s wound was near well,}
\stick{and as soon as the surgeon recovered his} 
\stick{surprize, and could get leave to say as}
\stick{much\tsh he told him, ’twas just beginning}
\stick{to incarnate; and that if no fresh exfolia-}
\stick{tion happened, which there was no sign}
\stick{of,\tsk it would be dried up in five or six}\pb
weeks. The sound of as many olym-\break
piads, twelve hours before, would have
convey’d an idea of shorter duration to my uncle
\textit{Toby}’s mind.\tsh The succession of his ideas
was now rapid,\tsk he broil’d with impatience to put his design
in execution;\tsk and so, without consulting farther with
any soul living,\tsk which, by the bye, I think is right, when
you are predetermined to take no one soul’s
advice,\tsk he privately ordered \textit{Trim}, his man, to
pack up a bundle of lint and dressings, and hire a chariot and four
to be at the door exactly by twelve o’clock that day, when he
knew my father would be upon ’Change.\tsk So leaving
a bank-note upon the table for the surgeon’s care of him, and
a letter of tender thanks\break 
for his brother’s\tsk he pack’d up his\break
maps, his books of fortification, his in-\break
struments, \etc\tsk and, by the help of a\pb 
crutch on one side, and \textit{Trim} on the\break
other,\tsh my uncle \textit{Toby} embark’d for\break
\textit{Shandy-Hall}.

The reason, or rather the rise of this sudden demigration, was as
follows:

The table in my uncle \textit{Toby}’s room, and at which,
the night before this\break 
\stick{change happened, he was sitting with his}
maps, \etc about him,\tsk being somewhat\break
of the smallest, for that infinity of great\break
and small instruments of knowledge\break
which usually lay crouded upon it;\tsk he\break
had the accident, in reaching over for\break
his tobacco-box, to throw down his com-\break
passes, and in stooping to take the com-\break
passes up, with his sleeve he threw down\break
his case of instruments and snuffers;\tsk\break
and as the dice took a run against him,\break
in his endeavouring to catch the snuffers\pb in
falling,\tsh he thrust Monsieur \textit{Blondel} off the
table, and Count \textit{de Pagan} o’\,top of him.

’Twas to no purpose for a man, lame as my uncle
\textit{Toby} was, to think of redressing these evils by
himself,\tsk he rung his bell for his man
\textit{Trim};\tsk \textit{Trim!} quoth my uncle \textit{Toby},
pri’thee see what
confusion I have here been making\tsk I must have some better
contrivance, \textit{Trim.}\tsk\break Can’st not thou take my rule,
and measure the length and breadth of this table, and then go and
bespeak me one as big again?\tsh Yes, an’ please
your \hbox{Honour,} replied \textit{Trim}, making a bow;\tsk but I\break 
\stick{hope your Honour will be soon well}
\stick{enough to get down to your country seat,}
\stick{where,\tsk as your Honour takes so much}
pleasure in
fortification, we could ma\-nage this matter to a T.

\newpage
I must here inform you, that this servant of my uncle
\textit{Toby}’s, who went by the name of \textit{Trim}, had been
a Corporal in my uncle’s own company,\tsh his real name was
\textit{James Butler},\tsk but having got the nick-name of
\textit{Trim}, in the regiment, my uncle \textit{Toby}, unless when he
happened to be very angry with him, would never call him by any
other name.

The poor fellow had been disabled for the service, by a
wound on his left knee by a musket-bullet, at the battle of
\textit{Landen}, which was two years before the affair of
\textit{Namur};\tsk and as the fellow was well beloved in
the regiment, and a\break
handy fellow into the bargain, my uncle \textit{Toby} took
him for his servant; and of excellent use was he, attending
my uncle \textit{Toby} in the camp and in his quarters as
valet, groom, barber, cook, sempster,\pb and nurse; and
indeed, from first to last, waited upon him and served him
with great fidelity and affection.

My uncle \textit{Toby} loved the man in return, and what attached
him more to him still, was the similitude of their\break
knowledge:\tsh For Corporal \textit{Trim}, (for\break so, for the
future, I shall call him) by\break
four years occasional attention to his\break
Master’s discourse upon fortified towns,\break
and the advantage of prying and peeping\break
\stick{continually into his Master’s plans, \etc}
\stick{exclusive and besides what he gained}
\textsc{Hobby-Horsically}, as a body-servant,
\textit{Non Hobby-Horsical per se};\tsh had become no mean
proficient in the science;\break and was thought, by the cook and
cham\-ber-maid, to know as much of the nature of
strong-holds as my uncle \textit{Toby}\break himself.

\newpage
I have but one more stroke to give to finish Corporal
\textit{Trim}’s character,\tsh and it is the only dark
line in it.\tsk The fellow lov’d to advise,\tsk or rather to
hear himself talk; his carriage, however, was so perfectly
respectful, ’twas easy to keep him silent when you had him
so; but set his tongue a-going,\tsk you had no hold of
him;\tsk he was voluble;\tsk the eternal interlardings of
\textit{your Honour}, with the respectfulness of Corporal
\textit{Trim}’s manner, interceding so strong in behalf of his
elocution,\tsk that tho’ you might have been
incommoded,\tsk you could not well be angry. My uncle
\textit{Toby} was seldom either the one or the other with
him,\tsk\break or, at least, this fault, in \textit{Trim}, broke no
squares with ’em. My uncle \textit{Toby}, as I said, loved the
man;\tsh and besides, as he ever looked upon a faithful
servant,\tsk\break but as a humble friend,\tsk he could not\pb bear to
stop his mouth.\tsh Such was Corporal \textit{Trim.}

If I durst presume, continued \textit{Trim}, 
to give your Honour my advice, and\break
speak my opinion in this matter.\tsk Thou\break
art welcome, \textit{Trim}, quoth my uncle\break
\stick{\textit{Toby},\tsk speak,\tsh speak what thou think-}
\stick{est upon the subject, man, without fear.}
\stick{Why then, replied \textit{Trim}, (not hanging}
his ears and scratching his head like a
country lout, but) stroking his hair back from his forehead, and
standing erect as before his division,\tsh I think, quoth
\textit{Trim}, advancing his left, which was his lame leg, a little
forwards,\tsk and pointing with his right hand open towards
a\break
\stick{map of \textit{Dunkirk}, which was pinn’d against}
the
hangings,\tsh I think, quoth Corporal \textit{Trim}, with
humble submission to your Honour’s better
judgment,\tsh that these\pb
\stick{ravelins, bastions, curtins, and horn-}
\stick{works, make but a poor, contemptible,}
fiddle faddle piece of work of it here\break
upon paper, compared to what your Honour and I
could make of it, were we in the country by ourselves, and had but a
rood, or a rood and a half of ground to do what we pleased with: As
summer is coming on, continued \textit{Trim}, your Honour might sit out
of doors, and give me\break
the nography\tsh (call it ichnography,
quoth my uncle)\tsk of the town or cita\-del, your Honour
was pleased to sit down before,\tsk and I will be shot by your
Honour upon the glacis of it, if I did not fortify it to your
Honour’s mind.\tsh I dare say thou would’st,
\textit{Trim}, quoth my uncle.\break
\tsk For if your Honour, continued the 
Corporal, could but mark me the polygon,
with its exact lines and angles.\tsk\break
That I could do very well, quoth my\pb 
uncle.\tsk I would begin with the fossé,\break
and if your Honour
could tell me the proper depth and breadth\tsk I can to a
hair’s breadth, \textit{Trim}, replied my uncle.\tsk\break
I would
throw out the earth upon this hand towards the town for the
scarp,\tsk\break and on that hand towards the campaign for the
counterscarp.\tsk Very right, \textit{Trim}; quoth my uncle
\textit{Toby}.\tsk And when I had sloped them to your
mind,\tsh an’ please 
\stick{your Honour, I would face the glacis, as}
the finest fortifications are done in \textit{Flan\-ders},
with sods,\tsk and as your Honour
\stick{knows they should be,\tsk and I would}
\stick{make the walls and parapets with sods}
\stick{too;\tsk The best engineers call them ga-}
\stick{zons, \textit{Trim}, said my uncle \textit{Toby};\tsk Whe-}
\stick{ther they are gazons or sods, is not much}
matter, replied \textit{Trim}, your Honour knows
they are ten times beyond a facing either
of brick or stone;\tsh I know they are,\pb
\stick{\textit{Trim} in some respects,\tsh quoth my uncle}
\stick{\textit{Toby}, nodding his head;\tsk for a cannon-}
ball enters into the gazon right onwards, 
without bringing any rubbish down with
it, which might fill the fossé, (as was the
case at \textit{St.\@ Nicolas}’s gate) and facilitate\break
the passage over it.

Your Honour understands these matters, replied Corporal
\textit{Trim}, better than any officer in his Majesty’s
service;\tsk\break
but would your Honour please to let the bespeaking of the
table alone, and let us 
\stick{but go into the country, I would work}
under your Honour’s directions like a horse, and make
fortifications for you
\stick{something like a tansy, with all their}
batteries, saps, ditches, and palisadoes, that it
should be worth all the world’s\break
riding twenty miles to go and see it.

\newpage
My uncle \textit{Toby} blushed as red as scarlet as \textit{Trim}
went on;\tsk but it was not a blush of guilt,\tsk of
modesty,\tsk or of anger,\tsk it was a blush of joy;\tsk he
was fired with Corporal \textit{Trim}’s project and
description.\tsk \textit{Trim!} said my uncle \textit{Toby}, thou
hast said enough.\tsk We might begin the campaign, continued
\textit{Trim}, on\break
\stick{the very day that his Majesty and the}
Allies take the field, and demolish them town by town as fast
as\tsh \textit{Trim}, quoth\break
my uncle \textit{Toby}, say no more.\tsk Your
Honour, continued \textit{Trim}, might sit in your 
\stick{arm-chair ( pointing to it ) this fine}
\stick{weather, giving me your orders, and I}
would\tsh Say no more, \textit{Trim}, quoth my uncle
\textit{Toby}\tsh Besides, your Honour would get not only
pleasure and good pastime,\tsk but good air, and good exercise,
and good health,\tsk and your Honour’s
wound would be well in a month.\pb
Thou hast said enough, \textit{Trim},\tsk quoth my
uncle \textit{Toby} (putting his hand into his breeches-pocket)\tsh I like thy project
mightily;\tsk And if your Honour pleases, I’ll this moment
go and buy a pioneer’s spade to take down with us, and
I’ll bespeak a shovel and a pick-ax, and a couple
of\tsh Say no more, \textit{Trim}, quoth my uncle
\textit{Toby}, leaping up upon one leg, quite overcome with
rapture,\tsk and\break
\stick{thrusting a guinea into \textit{Trim}’s hand,\tsk}
\textit{Trim}, said my uncle \textit{Toby}, say no
more;\tsk\break
but go down, \textit{Trim}, this moment, my lad, and
bring up my supper this in-\break stant.

\textit{Trim} ran down and brought up his Master’s
supper,\tsk to no purpose:\tsh\break
\stick{\textit{Trim}’s plan of operation ran so in my}
\stick{uncle \textit{Toby}’s head, he could not taste}
\stick{it.\tsk \textit{Trim}, quoth my uncle \textit{Toby}, get me}\pb
to-bed.\tsk ’twas all one.\tsk Corporal \textit{Trim}’s
description had fired his imagination,\tsk\break
my uncle \textit{Toby} could not shut his eyes.\tsk\break
The more he considered it, the more 
bewitching the scene appeared to him;\tsk\break
so that, two full hours before day-light, he had
come to a final determination and had concerted the whole plan of his\break
and Corporal \textit{Trim}’s decampment.

My uncle \textit{Toby} had a little neat coun-\break
try-house of his own, in the village where my father’s estate lay at \textit{Shandy},
which had been left him by an old uncle,\break with a small estate of
about one hundred pounds a year. Behind this house, and contiguous
to it, was a kitchen-garden of about half an acre;\tsk and at the
bottom of the garden, and cut off from it\break
by a tall yew hedge, was a bowling-\break 
green, containing just about as much\pb
ground as Corporal
\textit{Trim} wished for;\tsk\break
so that as \textit{Trim} uttered the
words, “ A rood and a half of ground to do what\break
they would with:”\tsh this identical bowling-green instantly presented
itself, and became curiously painted all at once, upon the retina
of my uncle \textit{Toby}’s fancy;\break\tsh which was the physical
cause of making him change colour, or at least of heightening his
blush, to that immo-\break
derate degree I spoke of.

Never did lover post down to a belov’d mistress with more heat
and expectation, than my uncle \textit{Toby} did, to enjoy this
self-same thing in private;\tsk I say in private;\tsk for it was sheltered from the house, as
I told you, by a tall yew hedge, and was covered on the other three
sides, from mortal sight, by rough holly and thickset flowering
shrubs;\tsk so that the\pb
idea of not being seen, did not a little contribute to the
idea of pleasure pre-conceived in my uncle \textit{Toby}’s
mind.\tsk\break 
\stick{Vain thought! however thick it was}
planted about,\tsh or private soever it\break might seem,\tsk to think,
dear uncle \textit{Toby}, of enjoying a thing which took up
a whole rood and a half of ground,\tsk and\break
not have it known!

How my uncle \textit{Toby} and Corporal \textit{Trim} managed this
matter,\tsk with the history of their campaigns, which were no
way barren of events,\tsh may make no uninteresting
under-plot in the epitasis and working up of this drama.\tsk At
present the scene must drop,\tsk and change for the parlour
fire-side.

\section{Chap. VI}

\quad\tsh \textsc{What} can they be doing?\break
brother, said my father.\tsk I think, replied my uncle \textit{Toby},\tsk taking, as I told you,
his pipe from his mouth, and striking the ashes out of it as he
began his sentence;\tsh I think, replied he,\tsk it
would not be amiss, brother, if we rung the\break bell.

Pray, what’s all that racket over our heads,
\textit{Obadiah?}\tsk quoth my father;\tsk\break
my brother and I can scarce hear ourselves speak.

Sir, answer’d \textit{Obadiah}, making a bow
towards his left-shoulder,\tsk my Mistress
is taken very badly;\tsk and where’s \textit{Susannah}
running down the garden there,\pb
as if they were going to ravish her?\tsh\break
Sir, she is running
the shortest cut into the town, replied \textit{Obadiah}, to
fetch the old midwife.\tsh Then saddle a horse, quoth my
father, and do you go directly for Dr.\@
\textit{Slop}, the man-midwife, with all our services,\tsk
and let him know your mistress is fallen into labour\tsk and
that I desire he will return with you with all speed.

It is very strange, says my father, addressing himself to my
uncle \textit{Toby}, as \textit{Obadiah} shut the door,\tsh as there
is so expert an operator as Dr.\@ \textit{Slop} so
near,\tsk\break
that my
wife should persist to the very last in this obstinate humour of
hers, in trusting the life of my child, who has had one misfortune
already, to the ignorance of an old woman;\tsh and not
only the life of my child, brother,\tsh but her own life,
and with it the lives of all the children I might, peradventure,
have begot out of her hereafter.

Mayhap, brother, replied my uncle\break
\textit{Toby}, my sister does it to save the expence:\tsk A
pudding’s end,\tsk replied my father,\tsh the doctor must be
paid the same for inaction as action,\tsk if not better,\tsk
to keep him in temper.

\noindent
\stick{\indent\tsh Then it can be out of no-}
\stick{thing in the whole world, quoth my}
\stick{uncle \textit{Toby}, in the simplicity of his heart,}
\stick{\tsk but \textsc{Modesty}:\tsk My sister, I dare}
\stick{say, added he, does not care to let a}
man come so near her \astiv. I will
not say whether my uncle \textit{Toby} had completed the
sentence or not;\tsh ’tis for his advantage to suppose he
had,\tsh\break as, I think, he could have added no\pb
\ls{\textsc{One Word}} which would have improved it.

If, on the contrary, my uncle \textit{Toby} had not fully arrived
at the period’s end\tsk\break
then the world stands indebted to
the sudden snapping of my father’s tobacco-pipe for one of
the neatest examples of that ornamental figure in oratory, which
Rhetoricians stile the \textit{Aposiopesis.}\tsh Just
heaven! how does the \textit{Poco piu} and the \textit{Poco meno} of
the \textit{Italian} artists;\tsk the insensible \textsc{more or
less}, determine the precise line of beauty in the sentence,
as well as in the statue! How do the slight touches of the chisel,
the pencil, the pen, the fiddle-stick, \textit{et
cætera},\tsk give the true swell, which gives the true
pleasure!\tsk O my countrymen!\tsk be nice; be cautious of your
language;\tsh and never, O! never let it be forgotten upon
what\pb small
particles your eloquence and your fame depend.


\tsh “\kern 3pt My sister, mayhap,” quoth my uncle \textit{Toby},
“\kern 3pt does not choose to let a
\stick{“\kern 3pt man come so near her \astiv” Make}
\stick{this dash,\tsk ’tis an Aposiopesis:\tsk Take}
\stick{the dash away, and write \textit{Backside},\tsh}
’tis Bawdy.\tsk Scratch Backside out, and put
\textit{Cover’d-way} in,\tsk ’tis a Metaphor;\tsk\break and, I dare say, as
fortification ran so much in my uncle \textit{Toby}’s head, that if he had
been left to have added one word to\break the sentence,\tsh that word was
it.


But whether that was the case or not the case;\tsk or whether
the snapping of my father’s tobacco-pipe, so critically,
happened thro’ accident or anger,\tsk will be seen in due time.

\newpage
\section{Chap. VII.}

\lettrine{T}{\,\ls{ho}}’ my father was a good natural
philosopher,\tsk yet he was something of a moral philosopher
too; for which reason, when his tobacco-pipe\break snapp’d short
in the middle,\tsk he had nothing to do,\tsk as such,\tsk
but to have taken hold of the two pieces, and thrown them
gently upon the back of the fire.\tsh\break He did no such
thing;\tsh he threw them with all the violence in the
world;\tsk and, to give the action still more
emphasis,\tsk\break he started upon both his legs to\break
do it.

This looked something like heat;\tsh\break
and the manner of his reply to what 
my uncle \textit{Toby} was saying, prov’d it\break
was so.

\newpage
\tsh “\kern 3pt Not choose,” quoth my father,
repeating my uncle \textit{Toby}’s words) “\, to let\break
“\, a man come so near her.”\tsh By hea-\break
ven, brother \textit{Toby!} you would try the patience of a
\textit{Job};\tsk and I think I have the plagues of one
already, without it.\break
\tsh Why?\tsh Where?\tsh Wherein?\tsh\break
Wherefore?\tsh Upon what account? re-\break plied my uncle \textit{Toby}: in the utmost
astonishment.\tsk To think, said my father, of a man living to
your age, brother, and knowing so little about
women!\tsh\break
I know nothing at all about them,\tsk re-\break
plied
my uncle \textit{Toby}; and I think, continued he, that the shock I
received the year after the demolition of \textit{Dunkirk}, in my
affair with widow \textit{Wadman};\tsk which shock you know I
should not have received, but from my total ignorance of the
sex,\tsk has given me just cause to say, That I neither
know, nor do pretend to\pb
know any thing about ’em, or their concerns either.\tsk Methinks, brother,
replied my father, you might, at least, know so much as the right
end of a woman from the wrong.

It is said in \textit{Aristotle}’s \textit{Master-Piece},
\stick{“\kern 2pt That when a man doth think of any}
\stick{“\kern 2pt thing which is past,\tsk he looketh down}
\stick{“\kern 2pt upon the ground;\tsh but that when he}
\stick{“\kern 2pt thinketh of something that is to come,}
\stick{“\kern 2pt he looketh up towards the heavens.”}

My uncle \textit{Toby}, I suppose, thought of neither, for he
look’d horizontally.\tsk\break
Right end! quoth my uncle \textit{Toby}, muttering the two words low to himself, and fixing his
two eyes insensibly as he muttered them, upon a small crevice,
formed by a bad joint in the chimney-piece\tsk\break
Right end of a woman!\tsh I declare,\pb
quoth my uncle, I know no more
which it is than the man in the moon;\tsh and if I was to
think, continued my uncle \textit{Toby} (keeping his eyes still fixed
upon the bad joint) this month together, I am sure I should not be able to find it out.

Then, brother \textit{Toby}, replied my father, I will tell you.

Every thing in this world, continued my father (filling a fresh
pipe)\tsk every thing in this earthly world, my dear
bro-\break ther \textit{Toby},
has two handles.\tsk Not always, quoth my uncle
\textit{Toby}.\tsh At least, replied my father, every one
has two hands,\tsk\break
which comes to the same
thing.\tsh Now, if a man was to sit down cooly, and con-\break
sider within himself the make, the shape, the construction,
com-at-ability, and convenience of all the parts which
constitute\pb
the whole of that animal, call’d Woman, and compare them
analogically.\tsk I never understood rightly the meaning of that
word,\tsk quoth my uncle \textit{Toby}.\tsh\break
\textsc{Analogy}, replied my father, is the certain
relation and agreement which different\tsh Here a devil of
a rap at the door snapped my father’s definition (like his
tobacco-pipe) in two,\tsk and, at the same time, crushed the head
of as notable and curious a dissertation as ever was
engendered in the womb of speculation;\tsk it was some months
before my father could get an opportunity to be safely
deliver’d of
it:\tsk And, at this hour, it is a thing full as problematical as
the subject of the dissertation itself,\tsk (considering the
confusion and distresses of our domestic misadventures, which are
now coming thick one upon the back of another) whether\pb I shall be
able to find a place for it in the third volume or not.

\section{Chap. VIII.}

\lettrine{I}{\,t} is about an hour and a
half’s tolerable good reading since my uncle \textit{Toby} rung
the bell, when \textit{Obadiah} was order’d to saddle a horse, and go
for Dr.\@ \textit{Slop}, the man-midwife;\tsk so that no one can say,
with reason, that I have not allowed \textit{Obadiah} time enough,
poetically speaking, and considering the emergen\-cy too, both to go
and come;\tsh tho’, morally and truly speaking, the
man perhaps, has scarce had time to get on\break
his boots.

If the hypercritic will go upon this; and is resolved after all
to take a pen-\pb dulum, and measure the true distance betwixt the
ringing of the bell, and the rap at the door;\tsk and, after
finding it to be no more than two minutes, thirteen seconds, and
three fifths,\tsk should take upon him to insult over me for such
a breach in the unity, or rather probability, of time;\tsk I would
remind him, that the idea of duration, and of its simple modes, is
got merely from the train and succession of our
ideas,\tsk and is the true scholastic
pendulum,\tsh and by which, as a scholar, I will be tried
in this matter,\tsk\break abjuring and detesting the jurisdiction of
all other pendulums whatever.

I would therefore desire him to consider that it is but poor
eight miles from \textit{Shandy-Hall} to Dr.\@ \textit{Slop}, the
man-mid\-wife’s house;\tsk and that whilst \textit{Obadiah} has
been going those said miles and back,\pb
I have brought my uncle
\textit{Toby} from \textit{Namur}, quite across all \textit{Flanders}, into
\textit{England}:\tsk That I have had him ill upon my hands near
four years;\tsk and have since travelled him and Corporal
\textit{Trim} in a chariot and four, a journey of near two hundred
miles down into \textit{Yorkshire};\tsk all which put together,
must have prepared the reader’s imagination for the entrance
of Dr.\@ \textit{Slop} upon the stage,\tsk as much, at least (I hope)
as a dance, a song, or a concerto between the acts.

If my hypercritic is intractable,\tsk al\-ledging, that two minutes
and thirteen seconds are no more than two minutes and thirteen
seconds,\tsk when I have said all I can about them;\tsk and that this
plea, tho’ it might save me dramatically, will damn me
biographically, rendering my book, from this very moment, a
pro-\pb
fess’d \textsc{Romance}, which, before, was a book
apocryphal:\tsk If I am thus pres-\break
sed.\tsk I then put an
end to the whole objection and controversy about it all
at\break
\stick{once,\tsk by acquainting him, that \textit{Obadiah}}
had not got above threescore yards from the stable-yard before he met
with Dr.\@ \textit{Slop};\tsk and indeed he gave a dirty
proof that he had met with him,\tsk and was within an ace of giving a
tragical one\break
too.  

Imagine to yourself;\tsk but this had better begin a new
chapter.


\section{Chap. IX.}

\lettrine{I}{\ls{magine}} to yourself a little, squat, uncourtly figure of a Doctor
\textit{Slop}, of\break
about four feet and a half perpendicular\pb
height, with a breadth of back, and a sesqui\-pedality of belly, which might have done honour to a serjeant in
the horse-guards.

Such were the out-lines of Dr.\@ \textit{Slop}’s figure, which\tsk if you have read
\textit{Ho\-garth}’s analysis of beauty, and if you have not, I wish you would;\tsh
you must know, may as certainly be caricatur’d, and convey’d to the mind by three
strokes as three hundred.

Imagine such a one,\tsk for such, I say, were the out-lines of
Dr.\@ \textit{Slop}’s figure, coming slowly along, foot by foot,
waddling thro’ the dirt upon the vertebræ of a little
diminutive pony,\tsk of a pretty colour;\tsk but of strength,\tsk
alack!\tsh\break
scarce able to have made an amble of it,\break
under such a fardel, had the roads been\pb
in an ambling condition.\tsh They were not.\tsh Imagine
to yourself, \textit{Obadiah}
\stick{mounted upon a strong monster of a}
\stick{coach-horse, prick’d into a full gallop,}
and making all practicable speed the ad-\break verse way.

Pray, Sir, let me interest you a moment in this description.

Had Dr.\@ \textit{Slop} beheld \textit{Obadiah} a mile\break
off, posting in a narrow lane directly to-\break
wards him, at that monstrous rate,\tsh\break
\stick{splashing and plunging like a devil thro’}
\stick{thick and thin, as he approach’d, would}
\stick{not such a phænomenon, with such a vor-}
\stick{tex of mud and water moving along with}
\stick{it, round its axis,\tsk have been a subject}
\stick{of juster apprehension to Dr.\@ \textit{Slop} in his}
\stick{situation, than the \textit{worst} of \textit{Whiston}’s co-}
\stick{mets?\tsk To say nothing of the \textsc{Nucleus};}\pb
\stick{that is, of \textit{Obadiah} and the coach-horse.}
\tsh In my idea, the vortex alone of\break
’em was enough to have involved and carried, if not the doctor, at least the
doctor’s pony, quite away with it. What then do you think
must the terror and hydrophobia of Dr.\@ \textit{Slop} have been, when
you read (which you are just going to do) that he was advancing
thus warily along towards \textit{Shandy-Hall}, and had
approach’d to
within sixty yards of it, and within five yards of a sudden turn,
made by an acute angle of the garden-wall,\break
\tsk and in the dirtiest part of a dirty lane,\break
\tsk when \textit{Obadiah} and his coach-horse 
turn’d the corner, rapid, furious,\tsk pop,\break
\tsk full upon him!\tsk Nothing, I think, in
nature, can be supposed more terrible,
than such a rencounter,\tsk so imprompt! 
so ill prepared to stand the shock
of it as Dr.\@ \textit{Slop} was!

\newpage
What could Dr.\@ \textit{Slop} do?\tsh he cross’d
\stick{himself ✝︎\tsk Pugh!\tsk but the doctor,}
Sir, was a Papist.\tsk No matter; he had better have kept
hold of the pummel.\break He had so;\tsk nay, as it
happened, he had better have done nothing at all;\tsk for in
crossing himself he let go his whip,\tsk\break and in
attempting to save his whip betwixt his knee and his
saddle’s skirt, as it slipp’d, he lost his stirrup,\tsk in
losing which, he lost his seat;\tsk and in the multitude of
all these losses (which, by the bye, shews what little
advantage there is in crossing) the unfortunate doctor lost
his presence of mind. So that without waiting for
\textit{Obadiah}’s onset, he left his pony to its destiny,
tumbling off it diagonally, something in the stile and
manner of a pack of wool, and without any other consequence
from the fall, save that of being left (as it would have
been)\pb
with the broadest part of him sunk about twelve inches deep in the mire.

\textit{Obadiah} pull’d off his cap twice to Dr.\@
\textit{Slop};\tsh once as he was falling,\tsk and then again
when he saw him seated.\tsk Ill-timed
complaisance!\tsh had not the fellow better have stopp’d his
horse, and got off and help’d him?\tsk Sir, he did all that
his situation would allow;\tsk but the\break
\textsc{Momentum} of the coach-horse was so great, that \textit{Obadiah} could not do it
all at once;\tsh he rode in a circle three times round Dr.\@ \textit{Slop},
before he could fully accomplish it any how;\tsk and at the last, when
he did stop his beast, ’twas done with such an explosion of
mud, that \textit{Obadiah} had better have been a league
off.\break
In short, never was a Dr.\@ \textit{Slop} so beluted, and so
transubstantiated, since that affair\break
came into fashion.

\newpage
\section{Chap. X.}

\lettrine{W}{\,\ls{hen}} Dr.\@ \textit{Slop} entered the back-\break
parlour, where my father and my\break
\stick{uncle \textit{Toby} were discoursing upon the na-}\break
\stick{ture of women,\tsh it was hard to deter-}
\stick{mine whether Dr.\@ \textit{Slop}’s figure, or Dr.}
\stick{\textit{Slop}’s presence, occasioned more surprize}
to them; for as the accident happened so near the house, as not to make it worth while for
\textit{Obadiah} to remount him,\tsk\break \textit{Obadiah} had led
him in as he was, \textit{un\-wiped, unappointed, unanealed}, with all
his stains and blotches on him.\tsk He stood like
\textit{Hamlet}’s ghost, motionless and speechless, for a full
minute and a half, at the parlour door (\textit{Obadiah} still holding
his hand) with all the majesty of mud. His hinder parts, upon which he had
received his fall, totally besmear’d,\tsh and\pb
in every other part of him, blotched over in such a manner with
\textit{Obadiah}’s explosion, that you would have sworn
(without mental reservation) that every grain of it had taken
effect.


Here was a fair opportunity for my uncle
\textit{Toby} to have triumph’d over my
father in his turn;\tsk for no mortal,
who had beheld Dr.\@ \textit{Slop} in
that pickle, could have dissented from
so much, at least, of my uncle
\textit{Toby}’s opinion, “\kern 2pt That may-\break
\stick{“\kern 2pt hap his sister might not care to let}
\stick{“\kern 2pt such a Dr.\@ \textit{Slop} come so near her}
\stick{“\kern 2pt \astiv\,” \qquad But it was the \textit{Argu-}}
\textit{mentum ad hominem}; and if my uncle
\textit{Toby} was not very expert at it,
you may think, he might not care to use
it.\tsk\break
No; the reason was,\tsk ’twas not his nature to insult.

\newpage
Dr.\@ \textit{Slop}’s presence at that time, was no less
problematical than the mode of it; tho’ it is certain, one
moment’s reflection in my father might have solved it; for he
had apprized Dr.\@ \textit{Slop} but the week before, that my mother was at her full reckoning;
and as the doctor had heard nothing since, ’twas natural and
very political too in him, to have taken a ride to
\textit{Shandy-Hall}, as he did, merely\break to see how matters went
on.

But my father’s mind took unfortunately a wrong turn in
the investigation; running, like the hypercritic’s,
altogether upon the ringing of the bell and the rap upon the
door,\tsk measuring their distance,\tsk and keeping his mind so
intent upon the operation, as to have power to think of nothing
else,\tsh common-place infirmity of the greatest
mathematicians!\pb working with might and main at the demonstration,
and so wasting all their strength upon it, that they have none left
in them to draw the corollary, to do\break good with.

The ringing of the bell, and the rap upon the door, struck
likewise strong upon the sensorium of my uncle
\textit{Toby},\tsk\break but it excited a very different train of
thoughts;\tsk the two irreconcileable pulsations instantly
brought \textit{Stevinus}, the\break
\stick{great engineer, along with them, into}
my uncle \textit{Toby}’s mind:\tsk What business
\textit{Stevinus} had in this affair,\tsk is the greatest
problem of all;\tsh it shall be solved,\tsk\break
but not in the next chapter.

\newpage

\section{Chap. XI.}

\lettrine{W}{\,\ls{riting}}, when properly managed, (as
you may be sure I think mine is) is but a different name for
conversation: As no one, who knows what he is about in good
company, would venture to talk all;\tsk so no author, who
understands the just boundaries of decorum and good breeding, would
presume to think all: The truest respect which you can pay to the
reader’s understanding, is to halve this matter amicably, and
leave him something to imagine, in his turn,\break
as well as yourself.

For my own part, I am eternally paying him compliments of this
kind, and do all that lies in my power to keep his\break
imagination as busy as my own.

\newpage
’Tis his turn now;\tsk I have given an ample description of Dr.\@ \textit{Slop}’s sad
overthrow, and of his sad appearance in the back-parlour;\tsk his
imagination must now go on with it for a while.

Let the reader imagine then, that Dr.\@ \textit{Slop} has told his
tale;\tsk and in what words, and with what aggravations his fancy
chooses:\tsk Let him suppose, that \textit{Obadiah} has told his
tale also, and with such rueful looks of affected concern, as he
thinks best will contrast the two\break
figures as they stand by each other:\tsk\break
Let him imagine, that my father has\break
stepp’d up stairs to see my mother.\tsk\break 
And, to conclude this work of imagination,\tsk let him
imagine the doctor wash’d,\break
\tsk rubb’d down,\tsk condoled with,\tsk\break
felicitated,\tsk got into a pair of \textit{Obadiah}’s\break
pumps, stepping forwards towards the\pb
door, upon the very point of entering\break
upon action.

Truce!\tsk truce, good Dr.\@ \textit{Slop!}\tsk stay thy
obstetrick hand;\tsk return it safe into thy bosom to keep
it warm;\tsk little dost thou know what obstacles,\tsk
little dost thou think what hidden causes, retard its
operation!\tsk Hast thou, Dr.\@ \textit{Slop},\tsk hast thou
been entrusted with the secret ar-\break
\stick{ticles of the solemn treaty which has} 
brought thee into this place?\tsk Art thou aware
that, at this instant, a daughter of \textit{Lucina} is put
obstetrically over thy head?\break
Alas!\tsk ’tis too true.\tsk
Besides, great son of \textit{Pilumnus!} what canst thou
do?\tsh\break
Thou hast come forth unarm’d;\tsk thou hast left thy
\textit{tire téte},\tsk thy new-invented
\textit{forceps},\tsk thy \textit{crotchet},\tsk thy
\textit{squirt}, and all thy instruments of salvation and
deliverance, behind thee.\tsk By heaven!\pb
at this moment they are hanging up in a green bays bag,
betwixt thy two pistols, at the bed’s head!\tsk Ring;\tsk
call;\tsk send \textit{Obadiah} back upon the coach-horse to
bring them with all speed.

\tsh Make great haste, \textit{Obadiah}, quoth my father,
and I’ll give thee a crown;\tsk\break
and, quoth my uncle \textit{Toby}, I’ll give him another.

\bigskip

\section{Chap. XII.}

\lettrine{Y}{\,\ls{our}} sudden and unexpected ar-\break
rival, quoth my uncle \textit{Toby}, ad-\break
dressing himself to Dr.\@ \textit{Slop},
(all three of them sitting down to the fire together, as my uncle
\textit{Toby} began to speak)\tsk in-\break
stantly brought the great \textit{Stevinus} into\break
my head, who, you must know, is a fa-\pb
\stick{vourite author with me.\tsh Then,}
added my father, making use of the ar\-gument \textit{Ad
Crumenam},\tsh I will lay\break
twenty guineas to a single crown piece (which will serve to
give away to \textit{Obadiah} when he gets back) that this
same \textit{Stevinus} was some engineer or other\tsk\break
or has wrote something or other, either directly or indirectly,
upon the science of\break
fortification.

He has so,\tsk replied my uncle \textit{Toby.}\tsk\break
I knew it, said my father;\tsk tho’, for the soul of me, I cannot see what kind
of connection there can be betwixt Dr.\@ \textit{Slop}’s sudden
coming, and a discourse upon fortification;\tsk yet I
fear’d it.\tsk\break
Talk of what we will, brother,\tsk or let the occasion be
never so foreign or unfit for the subject,\tsk you are sure
to bring it in: I would not, brother \textit{Toby}, con-\pb
tinued my father,\tsh I declare I would not have my head
so full of curtins and horn-works.\tsh That, I dare say, you
would not, quoth Dr.\@ \textit{Slop}, interrupting him, and
laughing most immoderately at\break
his pun.

\textit{Dennis} the critic could not detest and abhor a pun, or
the insinuation of a pun, more cordially than my father;\tsh he
would grow testy upon it at any time;\tsk\break
but to be broke in upon by one, in a serious discourse, was
as bad, he would say, as a fillip upon the nose;\tsh he saw\break
no difference.

Sir, quoth my uncle \textit{Toby}, addressing himself to Dr.\@ \textit{Slop},\tsk 
the curtins my brother \textit{Shandy} mentions
here, have no-\break
\stick{thing to do with bed-steads;\tsk tho’, I} know,
\textit{Du Cange} says, “\, That bed-cur-\pb
\stick{“\kern 2pt tains, in all probability, have taken}
\stick{“\kern 2pt their name from them;”\tsk nor have}
the horn-works he speaks of, any thing in the world to do with the
horn-works of cuckoldom:\tsk But the \textit{curtin}, Sir, is the
word we use in fortification, for that part of the wall or rampart
which lies between the two bastions and joins them.\tsh Be\-siegers
seldom offer to carry on their attacks directly against the curtin,
for this reason, because they are so well
\textit{flanked}.\break
(’Tis the case of other curtins, quoth Dr.\@ \textit{Slop}, laughing). 
However, continued my uncle \textit{Toby}, to
make them sure, we generally choose to place ravelins before them,
taking care only to extend them beyond the fossé or
ditch:\tsh The common men, who know very little of
fortification, confound the ravelin and the half-moon
together,\tsk tho’ they are very different
things;\tsk not in their figure or\pb
construction, for we make them exactly alike, in all
points;\tsk for they always consist of two faces,
making a salient angle, with the gorges, not straight, but in form
of a crescent;\tsk Where then lies the difference? (quoth
my father, a little testily.)\tsk In their situations, answered
my uncle \textit{Toby:}\tsk For when a ravelin, brother, stands
before the curtin, it is a ravelin; and when a ravelin
stands before a bastion, then the ravelin is not a
ravelin;\tsk it is a half-moon;\tsk a half-moon likewise is a
half-moon, and no more, so long as it stands before its
bastion;\tsh but was it to change place, and get before
the curtin,\tsk ’twould be no longer a half-moon; a
half-moon, in that case, is not a half-moon;\tsk ’tis no
more than a ravelin.\tsh I think, quoth my father, that
the noble science of defence has its weak sides\tsh as
well as others.

\newpage
\tsh As for the horn-work (high! ho! sigh’d my father)
which, continued my uncle \textit{Toby}, my brother was speaking of,
they are a very considerable part of an 
\stick{outwork;\tsh they are called by the \textit{French}}
engineers, \textit{Ouvrage à
corne}, and we generally make them to cover such places as we
suspect to be weaker than the rest;\tsk\break ’tis formed by two
epaulments or demi-bastions\tsk they are very pretty, and
if you will take a walk, I’ll engage to shew you one well
worth your trouble.\tsh I own, continued my uncle \textit{Toby},
when we crown them,\tsk they are much stronger, but
then they are very expensive, and take up a great deal of
ground; so that,\break
in my opinion, they are most of use to\break
cover or defend the head of a camp;\break
otherwise the double tenaille\tsh By\break
the mother who bore us!\tsh brother\pb
\textit{Toby}, quoth my father, not able to hold out any
longer,\tsk you would provoke a saint;\tsk here have you
got us, I know not how, not only souse into the middle of
the old subject again:\tsk But so full is your head of these
confounded works, that tho’ my wife is this moment in the
pains of labour,\tsk and you hear her cry out,\tsk\break
yet nothing will serve you but to carry\break
off the man-midwife.\tsh \textit{Accoucheur},\tsk\break
if you please, quoth Dr.\@ \textit{Slop.}\tsk With all 
my heart, replied my father, I don’t care
what they call you,\tsk but I wish the whole
science of fortification, with all its inventors, at the
devil;\tsk it has been the death of thousands,\tsk and it
will be mine in the end.\tsk I would not, I would not,
brother \textit{Toby}, have my brains so full of saps,
mines, blinds, gabions, palisadoes, ravelins, half-moons,
and such trum-\pb 
pery, to be proprietor of \textit{Namur}, and of\break
all the towns in \textit{Flanders} with it.

My uncle \textit{Toby} was a man patient of injuries;\tsk not
from want of courage,\tsk I have told you in the chapter of
this\break
second book, “\kern 2pt That he was a man of\break
“\kern 2pt courage:”\tsk And will add here, that where just
occasions presented, or called it forth,\tsk I know no man
under whose arm I would have sooner taken shelter;\break
nor did this arise from any insensibility or obtuseness of
his intellectual parts;\tsk\break for he felt this insult of my
father’s as feelingly as a man could do;\tsk but he was of a
peaceful, placid nature,\tsk no\break
jarring element in it,\tsk all was mix’d up\break
so kindly within him; my uncle \textit{Toby}\break
had scarce a heart to retaliate upon\break a fly.

\newpage
\tsk Go\tsk says he, one day at dinner, to an over-grown one
which had buzz’d about his nose, and tormented him cruelly all
dinner-time,\tsk and which, after infinite attempts, he had caught
at last, as it flew by him;\tsk I’ll not hurt thee,
says my uncle \textit{Toby}, rising from his chair, and
going a-cross
the room, with the fly in his hand,\tsk I’ll not hurt a
hair of thy head:\tsk Go, says he, lifting up the sash, and
opening his hand as he spoke, to let it escape;\tsk go poor
devil, get thee gone, why should I hurt thee?\tsh This
world surely is wide enough to hold both thee\break and me.

I was but ten years old when this happened;\tsk but whether it was,
that the action itself was more in unison to my nerves at that age
of pity, which instantly set my whole frame into one
vibration\pb of
most pleasurable sensation;\tsk or how far the manner and
expression of it might go towards it;\tsk or in what degree, or
by what secret magic,\tsk a tone of voice and harmony of
movement, attuned by mercy, might find a passage to my heart,
\stick{I know not;\tsk this I know, that the lesson}
of universal good-will
then taught and imprinted by my uncle \textit{Toby}, has never since
been worn out of my mind: And 
\stick{tho’ I would not depreciate what the study}
of the \textit{Literæ humaniores}, at the university,
have done for me in that
respect, or discredit the other helps of an expensive education
bestowed upon me, both at home and abroad since;\tsk yet I often
think that I owe one half of my philanthropy to that one accidental
impression.

☞\quad This is to serve for parents and governors instead of a whole
volume\break upon the subject.

I could not give the reader this stroke in my uncle
\textit{Toby}’s picture, by the instrument with which I drew
the other parts of it,\tsh that taking in no more than
\stick{the mere
\ls{\textsc{Hobby-Horsical}} likeness;}\break\tsh this
is a part of his moral character. My father, in this patient
endurance of wrongs, which I mention, was very different, as the
reader must long ago have noted; he had a much more acute and quick
sensibility of nature, attended with a little soreness of temper;
tho’ this ne\-ver transported him to any thing which looked
like malignancy;\tsh yet in the little rubs and vexations of
life, ’twas apt to shew itself in a drollish and witty kind of
peevishness:\tsh He was, however, frank and generous in
his nature;\break
\stick{\tsh at all times open to conviction;}
and in the little ebullitions of this sub\-acid humour towards others, but
parti-\pb
cularly towards my uncle \textit{Toby}, whom he truly
loved:\tsh he would feel more pain, ten times told (except
in the affair of my aunt \textit{Dinah}, or where an
hy-\break 
pothesis was concerned) than what he ever gave.

The characters of the two brothers, in this view of them,
reflected light upon each other, and appear’d with great advantage
in this affair which arose about \textit{Stevinus.}

I need not tell the reader, if he keeps a
\textsc{Hobby-Horse},\tsk that a man’s
\textsc{Hobby-Horse} is as tender a part as he
has about him; and that these unprovoked strokes, at my uncle
\textit{Toby}’s could not be unfelt by
him.\tsh No;\tsk as I said above, my uncle
\textit{Toby} did feel them, and very\break sensibly too.

\newpage
Pray, Sir, what said he?\tsk How did he behave?\tsk O,
Sir!\tsk it was great: For as soon as my father had done insulting his
\textsc{Hobby-Horse},\tsh he turned
his head without the least emotion, from Dr.\@ \textit{Slop}, to whom
he was addressing his discourse, and look’d up into my
father’s face, with a countenance spread over with so much
good nature;\tsh so placid;\tsh so
fraternal;\tsk so inexpressibly tender towards
him:\tsk it penetrated my father to his heart: He rose up hastily
from his chair, and seizing hold of both my uncle
\textit{Toby}’s hands as he spoke:\tsh\break
Brother \textit{Toby}, said he:\tsk I beg thy pardon;\tsk forgive, I pray thee,
this rash humour which my mother gave me.\tsh\break
My dear, dear brother, answer’d my uncle \textit{Toby}, rising up by my
father’s help, say no more about it;\tsk you are heartily
welcome, had it been ten times as much,\pb
brother. But ’tis
ungenerous, replied my father, to hurt any man;\tsh a
brother worse;\tsk but to hurt a brother of such gentle
manners,\tsk so unprovoking,\tsk and so
unresenting;\tsk ’tis base:\tsk By heaven,
’tis cowardly.\tsh You are heartily welcome, brother, quoth
my uncle \textit{Toby},\tsk\break
had it been fifty times as much.\tsh\break
Besides, what have I to do, my
dear \textit{Toby}, cried my father, either with your amusements or
your pleasures, unless it was in my power (which it is not) to
increase\break
their measure?

\tsk Brother \textit{Shandy}, answer’d my uncle
\textit{Toby}, looking wistfully in his face,\tsk you are
much mistaken in this point:\tsk for you do increase my pleasure
very much, in begetting children for the \textit{Shandy} family at
your time of life.\tsh But, by that, Sir, quoth Dr.\@ \textit{Slop},
Mr.\@ \textit{Shandy} in-\pb
creases his own.\tsh Not a jot, quoth\break
my father.

\bigskip

\section{Chap. XIII.}

\lettrine{M}{y} brother does it, quoth my uncle
\textit{Toby}, out of \textit{principle.}\tsh In a family-way,
I suppose, quoth Dr.\@ \textit{Slop.}\tsh\break
Pshaw!\tsk said my father,\tsk ’tis not worth\break
talking of.

\bigskip

\section{Chap. XIV.}

\lettrine{A}{t} the end of the last chapter, my\break
father and my uncle \textit{Toby} were left both standing, like
\textit{Brutus} and \textit{Cassius}, at the close of the scene, making
up their accounts.

\newpage
\noindent
\stick{\hfill As my father spoke the three last words,}\break
\tsk he sat down;\tsk my uncle \textit{Toby}
exactly followed his example, only, that before he took his chair, he rung the bell,
to order Corporal \textit{Trim}, who was in waiting, to step home for
\textit{Stevinus}:\tsk my uncle \textit{Toby}’s house being no farther off than the
opposite side of the way.

Some men would have dropp’d the subject of
\textit{Stevinus};\tsk but my uncle \textit{Toby} had no resentment
in his heart, and he went on with the subject, to shew my father
that he had none.

Your sudden appearance, Dr.\@ \textit{Slop},\break
quoth my uncle,
resuming the discourse, instantly brought \textit{Stevinus} into my
head. (My father, you may be sure, did not offer to lay any more
wagers upon \textit{Stevinus}’s head.)\tsh Because, continued my\pb
uncle \textit{Toby}, the celebrated sailing chariot,
which belonged to Prince \textit{Maurice}, and was of such wonderful
contrivance and velocity, as to carry half a dozen people thirty
\textit{German} miles, in I don’t know how few
minutes,\tsh was invented by \textit{Stevinus}, that great
mathematician and engineer.

You might have spared your servant the trouble, quoth Dr.\@ \textit{Slop} (as the fellow is lame) of going for
\textit{Stevinus}’s account of it, because, in my return from
\textit{Leyden} thro’ the \textit{Hague}, I walked as far as
\textit{Schev\-ling}, which is two long miles, on
purpose\break 
to take a view of it.

\tsk That’s nothing, replied my uncle \textit{Toby}, to what the
learned \textit{Peireskius} did, who walked a matter of five hundred
miles, reckoning from \textit{Paris} to
\textit{Schevling},\pb and from
\textit{Schevling} to \textit{Paris} back again, in order to see
it,\tsk and nothing else.

Some men cannot bear to be out-gone.

The more fool \textit{Peireskius}, replied Dr.\@ \textit{Slop.} But
mark, ’twas out of no contempt of \textit{Peireskius} at
all;\tsh but that \textit{Peireskius}’s indefatigable labour in
trudging so far on foot, out of love for the sciences, reduced the
exploit of Dr.\@ \textit{Slop}, in that affair, to nothing;\tsk the
more fool \textit{Peireskius}, said he again:\tsk Why so?\tsk\break
replied my father, taking his brother’s part, not
only to make reparation as fast as he could for the insult he had
given him, which sat still upon my father’s\break
mind;\tsk but partly, that my father began really to
interest himself in the discourse.
\stick{\tsk Why so?\tsk said he. Why is \textit{Peireskius},}
or any man else, to be abused for an ap-\pb
petite for that, or any other morsel of
sound knowledge? For notwithstand-\break
ing I know nothing of the chariot in\break
question, continued he, the inventor of it must have had a very mechanical head;\break
and tho’ I cannot guess upon what prin\-ciples of philosophy he has atchiev’d\break
it;\tsh yet certainly his machine has\break
been constructed upon solid ones, be\break
they what they will, or it could not\break
have answer’d at the rate my brother\break
mentions.

It answered, replied my uncle \textit{Toby}, as well, if not better; for, as \textit{Peireskius}
elegantly expresses it, speaking of the velocity of its motion,
\textit{Tam citus erat, quam erat ventus}; which, unless I have
forgot my Latin, is, \textit{that it was as swift\break as the wind
itself}.

\newpage
But pray, Dr.\@ \textit{Slop}, quoth my father,\break
interrupting my uncle (tho’ not with-\break
out begging pardon for it at the same
time) upon what principles was this self-\break
same chariot set a-going?\tsk Upon very pretty principles to be sure, replied
Dr.\break
\stick{\textit{Slop};\tsh and I have often wondered,}\break
continued he, evading
the question, why none of our gentry, who live upon large plains
like this of ours,\tsk (especially they whose wives are not past
child-bearing) attempt nothing of this kind; for it would not only
be infinitely expeditious upon sudden calls, to which the sex is
subject,\tsk if the wind only served,\tsk but would be
excellent good husbandry to make use of the winds, which cost
no-\break
thing, and which eat nothing, rather\break
than horses, which (the devil take ’em)\break 
both cost and eat a great deal.

\newpage
For that very reason, replied my father,
“\kern 2pt Because they cost nothing, and\break
“\kern 2pt because they eat nothing,”\tsk the scheme\break
is bad;\tsk it is the
consumption of our products, as well as the manufactures of them,
which gives bread to the hungry, circulates trade,\tsk brings in
money, and supports the value of our lands;\tsk and tho’, I
own, if I was a prince, I would
ge\-nerously recompense the scientific head which brought
forth such contrivances;\break\tsk yet I
would as peremptorily suppress the use of them.

My father here had got into his element,\tsh and was going on
as pros\-per\-ously with his dissertation upon trade, as my uncle
\textit{Toby} had before, upon his\break of fortification;\tsh but to the
loss\break
of much sound knowledge, the destinies in the morning had
decreed that no\pb
dissertation of any kind should be spun
\stick{by my father that day,\tsh for as}
\stick{he opened his mouth to begin the next}\break sentence,

\medskip

\section{Chap. XV.}

\lettrine{I}{\,n} popp’d Corporal \textit{Trim} with
\textit{Stevinus}:\tsk But ’twas too late,\tsk all the
discourse had been exhausted without him, and was running into a
new channel.

\tsk You may take the book home again, \textit{Trim}, said my
uncle \textit{Toby}, nodding to\break him.

But pri’thee, Corporal, quoth my father, drolling,\tsk look
first into it, and see if thou canst spy aught of a
sailing\break
chariot in it.

\newpage
Corporal \textit{Trim}, by being in the service, had learned to
obey,\tsk and not to remonstrate,\tsk so taking the book to a
side-table, and running over the leaves; an’ please your
Honour, said \textit{Trim}, I can see no such thing;\tsk however,
continued the Corporal, drolling a little in his turn, I’ll
make sure work of it, an’ please your Honour;\tsk so taking
hold of the two\break covers of the book, one in each hand, and letting
the leaves fall down as he bent the covers back, he gave the
book\break a good sound shake.

There is something fallen out, however, said \textit{Trim},
an’ please your Honour; but it is not a chariot, or any
thing like one:\tsk Pri’thee, Corporal, said my father, smiling,
what is it then?\tsk I think, answered \textit{Trim}, stooping to
take it up,\tsk\break  
’tis more like a sermon,\tsk for it begins,\pb
with a text of scripture, and
the chapter and verse;\tsk and then goes on, not as a chariot,
but like a sermon directly.

The company smiled.

I cannot conceive how it is possible, quoth my uncle
\textit{Toby}, for such a thing as a sermon to have got into my
\textit{Stevinus.}

I think ’tis a sermon, replied \textit{Trim}:\tsk\break
but if it please your Honours, as it is a fair hand, I will read you a
page;\tsk for\break \textit{Trim}, you must know, loved to
hear\break himself
read almost as well as talk.

I have ever a strong propensity, said my father, to look into
things which\break cross my way, by such strange fatalities\break
as these;\tsk and as we have nothing better to do, at least
till \textit{Obadiah} gets back,\pb
I should be obliged to you, brother, if Dr.\@ \textit{Slop} has no
objection to it, to order the Corporal to give us a page or two of
it,\tsk if he is as able to do it, as he seems willing. An’
please your honour, quoth \textit{Trim}, I officiated two whole
campaigns, in \textit{Flanders}, as clerk to the chaplain of the
regiment.\tsh He can read it, quoth my uncle \textit{Toby},
as well as I can.\tsh \textit{Trim}, I assure you, was the
best scholar in my company, and should have had the next halberd,
but for the poor fellow’s mis\-fortune. Corporal \textit{Trim}
laid his hand upon his heart, and made an humble bow to his
master;\tsk
then laying down his hat upon the floor, and taking up the sermon
in his left hand, in order to have his right at liberty,\tsk he
advanced, nothing doubting, into the middle of the room, where he
could best see, and be best seen\break by his audience.

\newpage
\section{Chap. XVI}

\quad\tsh If you have any
objection,\tsk said my father, addressing himself to Dr.\@
\textit{Slop}.\break
Not in the least, replied Dr.\@ \textit{Slop};\tsk for
it does not appear on which side of the question it is
wrote;\tsk it may be a composition of a divine of our church, as
well as yours,\tsk so that we run equal
risks.\tsh ’Tis wrote upon neither side, quoth
\textit{Trim}, for ’tis only upon \textit{Conscience},\break
an’ please your Honours.

\textit{Trim}’s reason put his audience into good
humour,\tsk all but Dr.\@ \textit{Slop}, who,
turning his head about towards \textit{Trim},\break
look’d a little angry.

Begin, \textit{Trim},\tsh and read distinctly, quoth my
father;\tsk I will, an’ please your\pb
Honour, replied the
Corporal, making a bow, and bespeaking attention with a slight
movement of his right hand.

\section{Chap. XVII}

\quad\tsh But before the Corporal
begins, I must first give you a description of his
attitude;\tsh otherwise he will naturally stand
represented, by your imagination, in an uneasy
posture,\tsk stiff,\tsk perpendi\-cular,\tsk dividing the
weight of his body equally upon both legs;\tsh his eye fix’d,
as if on duty;\tsk his look determined,\tsk\break clenching the
sermon in his left hand, like his firelock:\tsk In a word,
you would be apt to paint \textit{Trim}, as if he was standing in his
platoon ready for action:\tsh\break
His attitude was as unlike all this as you\break
can conceive.

\newpage
He stood before them with his body swayed, and bent forwards
just so far, as to make an angle of 85 degrees and a half upon the
plain of the horizon;\tsh\break which sound orators, to whom I address
this, know very well to be the true persuasive angle of
incidence;\tsk in any other angle you may talk and
preach;\tsk ’tis certain,\tsk and it is done every
day;\tsk\break but with what effect,\tsk I leave the world to
judge!

The necessity of this precise angle of 85 degrees and a half to
a mathematical exactness,\tsh does it not shew us, by the
way,\tsk how the arts and sciences mutually befriend each other?

How the duce Corporal \textit{Trim}, who\break
knew not so much as an acute angle\break
from an obtuse one, came to hit it so ex-\pb 
actly;\tsh or whether it was chance or nature, or good
sense or imitation, \etc\break shall be commented upon in that part of
the cyclopædia of arts and sciences, where the instrumental
parts of the eloquence of the senate, the pulpit, and the bar, the
coffee-house, the bed-chamber, and fire-side, fall under
consideration.

He stood,\tsh for I repeat it, to take the picture of
him in at one view, with his body sway’d, and somewhat bent
forwards,\tsk his right leg from under him, sustaining
seven\,-\,eighths of his whole\break
weight,\tsk the foot of his
left leg, the defect of which was no disadvantage to his attitude,
advanced a little,\tsk not laterally, nor forwards, but in a line
betwixt them;\tsk his knee bent, but that not
violently,\tsk but so as to fall within the limits of the line of
beauty;\tsk and I add,\pb
of the line of science too;\tsk for
consider, it had one eighth part of his body to bear up;\tsk so
that in this case the position of the leg is
determined,\tsk because the foot could be no farther advanced, or
the knee more bent, than what would allow him mechanically,
to receive an eighth part of his whole weight under\break
it,\tsk and to carry it too.

☞ This I recommend to painters?\tsk\break
need I add,\tsk to orators?\tsk I think not;\break
for unless they practise it,\tsh they must\break
fall upon their noses.

So much for Corporal \textit{Trim}’s body and
legs.\tsk He held the sermon loosely,\tsk not carelessly, in
his left hand, raised something above his stomach, and detached a
little from his breast;\tsh his right-arm
\stick{falling negligently by his side, as nature}\pb
and the laws of gravity ordered
it,\tsh but with the palm of it open and turned towards
his audience, ready to aid the sentiment in case it stood in
need.

Corporal \textit{Trim}’s eyes and the muscles of his face
were in full harmony with the other parts of him;\tsk he looked
frank,\tsk\break
unconstrained,\tsk something assured,\tsh\break
but not bordering upon assurance.

Let not the critic ask how Corporal \textit{Trim} could come by
all this. I’ve told him it should be explained;\tsk but so he
stood before my father, my uncle \textit{Toby}, and Dr.\@ \textit{Slop},\tsk so swayed his body, so contrasted his limbs, and
with such an oratorical sweep throughout the whole
figure,\tsh a statuary might have modell’d from
it;\tsh nay, I doubt whether the oldest Fellow of a
College,\tsk or the \textit{He}-\pb\textit{brew} Professor himself, could have
much mended it.

\textit{Trim} made a bow, and read as follows:

\bigskip
\centerline{The\quad \lsss{SERMON}.}
\bigskip
\centerline{\ls{\textsc{Hebrews}}\, xiii. 18.}
\tsh \textit{For we} trust \textit{we have a good
Conscience}.\tsh

\lettrine{\rlap{\normalsize “}\lower-12pt\hbox{\normalsize “ }T}{\,rust!}\tsh Trust we have a good conscience!”

[Certainly, \textit{Trim}, quoth my father, interrupting him, you
give that sentence a very improper accent; for you curl up your
nose, man, and read it with such a sneering tone, as if the Parson
was going to abuse the Apostle.

\newpage
He is, an’ please your Honour, replied \textit{Trim.} Pugh!
said my father, smiling.

Sir, quoth Dr.\@ \textit{Slop, Trim} is certainly in the right; for
the writer (who I perceive is a Protestant) by the snappish
\stick{manner in which he takes up the Apostle,}
\stick{is certainly going to abuse him,\tsk if this}
treatment of him has not done it already. But
from whence, replied my father,\break
have you concluded so soon, Dr.\@ \textit{Slop},\break
that the writer is of our Church?\tsk for\break aught I
can see yet,\tsk he may be of any\break Church:\tsh Because,
answered Dr.\@ \textit{Slop}, if he was of ours.\tsk he durst no more
take such a licence,\tsk than a bear by his beard:\tsk If, in
our communion, Sir, a man was to insult an Apostle,\tsh a
saint,\tsh or even the paring of a saint’s
nail,\tsk he would have his eyes scratched out.\tsk What, by
the saint, quoth my\pb
uncle \textit{Toby.} No, replied Dr.\@ \textit{Slop},\tsk
he would have an old house over his head. Pray is the Inquisition
an ancient building, answered my uncle \textit{Toby}, or is it a
modern one?\tsk I know nothing of architecture, replied
Dr.\@ \textit{Slop.}\tsh
An’ please your Honours, quoth \textit{Trim}, the Inquisition is the
vilest\tsh Pri’thee spare thy description, \textit{Trim}, I
hate the very name of it, said my father.\tsk No matter for that,
answered Dr.\@ \textit{Slop},\tsk it has its uses; for tho’
I’m no great advocate for it, yet, in such a case as this, he
would soon be taught better manners; and I can tell him, if he went
on at that rate, would be flung into the Inquisition for his pains.
God help him then, quoth my uncle \textit{Toby.} Amen, added
\textit{Trim}; for heaven above knows, I have a poor brother who has
been fourteen years a captive in it.\tsk\break
I never heard one word of it before, said\pb
my uncle \textit{Toby}, hastily:\tsk How came he
there, \textit{Trim?}\tsh O, Sir!\@ the story will make your
heart bleed,\tsk as it has made mine a thousand times;\tsk but
it is too long to be told now;\tsk your Honour shall hear it from
first to last some day when I am working beside you in our
fortifications;\tsk but the short of the story is
this;\tsk That my brother \textit{Tom} went over a servant to
\textit{Lisbon},\tsk and then married a \textit{Jew}’s widow, who kept
a small shop, and sold sausages, which, somehow or other, was the cause of
his being taken in the middle of the night out of his bed, where he
was lying with his wife and two small children, and carried
directly to the Inquisition, where, God help him, continued
\textit{Trim}, fetching a sigh from the bottom of his
heart,\tsk the poor honest lad lies confined at this
hour;\tsh\break he was
as honest a soul, added \textit{Trim},\pb (pulling out his handkerchief)
as ever blood warm’d.\tsh

\tsh The tears trickled down \textit{Trim}’s cheeks faster
than he could well wipe them away.\tsk A dead silence in the room
ensued for some minutes.\tsh Certain\break proof of pity!

Come \textit{Trim}, quoth my father, after he saw the poor
fellow’s grief had got a little vent,\tsk read
on,\tsk and put this melancholy story out of thy head:\tsk I
grieve that I interrupted thee;\tsk but pri’thee begin the sermon
again;\tsk for if the first sentence in it is matter of abuse, as
thou sayest, I have a great desire to know what\break kind of provocation
the Apostle has given.

Corporal \textit{Trim} wiped his face, and returned his
handkerchief into his pocket,\pb and, making a bow as he did
it,\tsk he began again.]

\bigskip
\centerline{The\quad \lsss{SERMON}.}
\bigskip
\centerline{\ls{\textsc{Hebrews}}\, xiii. 18.}
\tsh \textit{For we} trust \textit{we have a good
Conscience}.\tsh

\lettrine{\rlap{\normalsize “}\lower-12pt\hbox{\normalsize “
}T}{\,rust!} trust we have a good con-\break
science! Surely if there is any\break
\stick{“\kern 2pt thing in this life which a man may}
\stick{“\kern 2pt depend upon, and to the knowledge}
\stick{“\kern 2pt of which he is capable of arriving}
\stick{“\kern 2pt upon the most indisputable evidence,}
\stick{“\kern 2pt it must be this very thing,\tsk whether}
\stick{“\kern 2pt he has a good conscience or no.”\hfill}

[I am positive I am right, quoth Dr.\@ \textit{Slop.}]

\newpage
\begin{story}{If a man thinks at all, he cannot}
well be a stranger to the true state of 
this account:\tsk he must be privy to
his own thoughts and desires;\tsh he
must remember his past pursuits, and
know certainly the true springs and
motives, which, in general, have governed the actions of his life.”
\end{story}
\vspace\parskip

[I defy him, without an assistant, quoth Dr.\@ \textit{Slop.}]

\begin{story}{In other matters we may be deceived}
by false appearances; and, as the
    wise man complains,
    \textit{hardly do we guess\break
aright at the things that are upon the earth, and with labour do we
find the things that are before us}. But here the mind has all
the evidence and facts with\-in
    herself;\tsk is conscious of
the web she has wove;\tsk knows its
    texture and\pb
fineness,
and the exact share which every passion has had in working upon the
several designs which virtue or vice
    has plann’d before
her.”
\end{story}
\vspace\parskip

[The language is good, and I declare \textit{Trim} reads very
well, quoth my father.]

\begin{story}{Now,\tsk as conscience
    is nothing else} but the
    knowledge which the mind has
    within herself of this; and the
    judgment, either of approbation
    or censure, which it unavoidably
    makes upon the successive
    actions of our lives; ’tis plain
    you will say, from the very
    terms of the proposition,\tsk
    whenever this inward testimony
    goes against a man, and he
    stands self-accused,\tsk that he
    must necessarily be a guilty
    man.\tsk And, on the contrary,
    when the report is fa\-vour\-able on
    his side, and his heart con-\pb demns
    him not:\tsh that it is not a
    matter of \textit{trust}, as the Apostle
    intimates,\tsk but a matter of
    \textit{certainty} and fact,
    that the conscience is good, and
    that the man must be good also.”
\end{story}
\vspace\parskip

[Then the Apostle is altogether in the wrong, I suppose, quoth
Dr.\@ \textit{Slop}, and the Protestant divine is in the right. Sir,
have patience, replied my father, for I think it will presently
appear that St.\@ \textit{Paul} and the Protestant divine are both of
an opinion.\tsk As nearly so, quoth Dr.\@ \textit{Slop}, as east is
to west;\tsk but this, continued he, lifting both hands, comes
from the\break liberty of the press.

It is no more at the worst, replied my uncle \textit{Toby}, than
the liberty of the pulpit; for it does not appear that the sermon is
printed, or ever likely to be.

\newpage
Go on, \textit{Trim}, quoth my father.]

\begin{story}{At first sight this may seem to be a}
    true state of the case: and I make no doubt but the
    knowledge of right and wrong is so truly impressed
    upon the mind of man,\tsk that did no such thing
    ever happen, as that the conscience of a man, by
    long habits of sin, might\break (as the scripture
    assures it may) insensibly become hard;\tsk and,
    like some tender parts of his body, by much stress
    and continual hard usage, lose,\break by degrees
    that nice sense and per-\break ception with which
    God and nature\break endow’d it:\tsk Did this never
    happen;\break \tsk or was it certain that self-love
    could never hang the least bias upon the\break
    judgment;\tsk or that the little interests\break
    below, could rise up and perplex the\break faculties
    of our upper regions, and\pb encompass them about
    with clouds and thick darkness:\tsh Could no
    such\break
    thing as favour and affection enter this sacred
    \textsc{Court}:\tsh  Did \textsc{Wit} disdain to
    take a bribe in it;\tsk or was asham’d to shew its
    face as an advocate for an unwarrantable
    enjoyment:\tsk Or, lastly, were we assured that
    \textsc{Interest} stood always unconcern’d whilst
    the cause\break was hearing\tsk and that passion never got
    into the judgment-seat, and pronounc’d sentence in
    the stead of reason, which is supposed always to
    preside and determine upon the case:\tsk Was this
    truly so, as the objection must suppose;\tsk no
    doubt then the religious and moral state of a man
    would be exactly what he himself esteem’d
    it:\tsk\break 
    and the guilt or innocence of every\break
    man’s life could be known, in general,\break
    by no better measure, than the de-\pb grees of his own
    approbation and censure.
\end{story}
\vspace\parskip

\begin{story}{I own, in one case, whenever a man’s}
conscience does accuse him (as it seldom errs on that side) that he
is guilty; and, unless in melancholy and hypocondriac cases,
we may safely pronounce upon it, that there is always sufficient
grounds for the accusation.
\end{story}
\vspace\parskip

\begin{story}{But the converse of the proposition}
will not hold
true;\tsh namely, that whenever there is guilt, the conscience
must accuse; and if it does not, that a man is therefore
innocent.\tsh This is not fact:\tsk So that the
common consolation which some good christian or other is hourly
administering to himself,\tsk that he thanks God his mind does
not misgive him; and that, con-\pb sequently, he has a good conscience,
because he hath a quiet one,\tsk is fallacious;\tsk and as
current as the inference is, and as infallible as the rule appears
at first sight, yet when you look nearer to it, and try the truth
of this rule upon plain facts,\tsk you see it liable to so
much error from a false application;\tsk the principle
upon which it goes so often perverted;\tsk the whole force
of it lost, and sometimes so vilely cast away, that it is painful
to produce the common examples from human life, which confirm the
account.
\end{story}
\vspace\parskip

\begin{story}{A man shall be vicious and utterly} debauched in his
principles;\tsk excep-\break tionable in his conduct to the world; shall live
shameless, in the open commission of a sin which no reason or pretence can
justify;\tsh a sin by which\pb contrary to all the workings of humanity,
he shall ruin for ever the deluded partner of his guilt;\tsk rob her of her
best dowry; and not only cover her own head with dishonour;\tsk but involve a
whole virtuous family in shame and sorrow for her sake. Surely, you will think
conscience must lead such a man a troublesome life; he can have no rest night
and day from its reproaches.
\end{story}
\vspace\parskip

\begin{story}{Alas! \textsc{Conscience} had something}
    else to do, all this time, than break in upon him;
    as \textit{Elijah} reproached the God
    \textit{Baal},\tsh this domestic God \textit{was
    either talking, or pursuing, or was in a journey, or
    peradventure he slept and could not be awoke.}
\end{story}
\vspace\parskip

\begin{story}{Perhaps \textsc{He} was gone out in com-}
    pany
with \textsc{Honour} to fight a duel;\pb to pay off some debt at
play;\tsh or dirty annuity, the bargain of his lust;
Perhaps \textsc{Conscience} all this time was engaged at
home, talking aloud against petty larceny, and executing vengeance
upon some such puny crimes as his fortune and rank of life secured
him\break
    against all temptation of committing;\break
    \hbox to 160pt{so that he lives as merrily;” [If he was}
\end{story}
\noindent of our church, tho’, quoth Dr.\@ \textit{Slop}, he\break
could not]\tsk 
\begin{story}{sleeps as soundly in his} 
bed;\tsk and at last meets death as
unconcernedly;\tsk perhaps much more so\break
    than a much better man.”
\end{story}
\vspace\parskip
All this is impossible with us, quoth Dr.\@
\textit{Slop}, turning to my father,\tsk the case could
not happen in our church.\tsh It happens in ours,
however, replied my father, but too often.\tsh I own,
quoth Dr.\@ \textit{Slop}, (struck a little with my
father’s\pb frank acknowledgment)\tsk that a man in the
\textit{Romish} church may live as badly;\tsk\break but
then he cannot easily die so.\tsh ’Tis little matter,
replied my father, with an air of indifference,\tsk how
a rascal dies.\tsk\break I mean, answer’d Dr.\@ \textit{Slop},
he would be denied the benefits of the last
sacraments.\break
\tsk Pray how many have you in all, said my uncle
\textit{Toby},\tsk for I always forget?\tsk\break
Seven, answered Dr.\@ \textit{Slop}.\tsk
Humph!\tsk\break said my
uncle \textit{Toby};\tsk tho’ not accented as a note of
acquiescence,\tsk but as an interjection of that
particular species of surprize, when a man in looking
into a drawer, finds more of a thing than he
ex-\break pected.\tsk Humph! replied my uncle \textit{Toby}.
Dr.\@ \textit{Slop}, who had an ear, understood my uncle
\textit{Toby} as well as if he had wrote a whole volume
against the seven sacraments.\tsh Humph! replied Dr.\@
\textit{Slop}, (stating my uncle \textit{Toby}’s
argument over\pb again to him)\tsh Why, Sir, are there not
seven cardinal virtues?\tsh Seven mortal sins?\tsh Seven
golden candlesticks?\tsh Seven heavens?\tsh ’Tis\break
more than I know, replied my uncle\break
\textit{Toby}.\tsh Are
there not seven wonders of the world?\tsh Seven days of
the creation?\tsh Seven planets?\tsh Seven plagues?\tsh
That there are, quoth my father with a most affected
gravity. But pri’thee, continued he, go on with the rest
of thy characters, \textit{Trim}.]

“Another is sordid, unmerciful,” (here
\textit{Trim} waved his right hand)
\begin{story}{a strait-}
    hearted, selfish wretch, incapable either of private
    friendship or public spirit. Take notice how he
    passes by the wi\-dow and orphan in their distress,
    and sees all the miseries incident to human\break
    \hbox to 160pt{life without a sigh or a prayer.\,” [And}
\end{story}
\newpage\noindent
please your honours, cried
\textit{Trim}, I think this a viler man than the other.]

\begin{story}{Shall not conscience rise up and sting}
    him on such occasions?\tsh No; thank God there is no
    occasion, \textit{I pay every man his own;\tsk I
    have no fornication to answer to my conscience;\tsk
    no faithless vows or promises to make up;\tsk I have
    debauched no man’s wife or child; thank God, I am
    not as other men, adulterers, unjust, or even as
    this libertine, who stands before me.}
\end{story}
\vspace\parskip

\begin{story}{A third is crafty and designing in}
    his nature. View
his whole life;\tsk ’tis nothing but a cunning
contexture “of dark arts and unequitable
subterfuges, basely to defeat the true intent of all
laws,\tsh plain dealing and the safe enjoyment of our
several properties.\tsh\pb You will see such a one working
out a frame of little designs upon the ignorance and perplexities
of the poor and needy man;\tsk shall raise a fortune upon the
inexperience of a youth, or the unsuspecting temper of his friend,
who would have trusted him with his\break life.
\end{story}
\vspace\parskip

\begin{story}{When old age comes on, and re-}
    pentance calls him to look back upon this black
    account, and state it over\break
    again with his conscience,\tsh \textsc{Con\-science} looks into the
    \textsc{Statutes} at\break
    \textsc{Large};\tsk finds no express law
    broken by what he has done;\tsk perceives no penalty
    or forfeiture of goods and chat\-tels incurred;\tsk
    sees no scourge waving 
    over his head, or prison opening his\break
    gates upon him:\tsk What is there to\break
    affright his conscience?\tsk Conscience\pb has got
    safely entrenched behind the Letter of the Law; sits
    there invulnerable, fortified with \gothic{Cases}
    and \gothic{Reports}\break so strongly on all sides;\tsk
that it\break is not preaching can dispossess it of its hold.\,”
\end{story} \vspace\parskip

[Here Corporal \textit{Trim} and my uncle \textit{Toby} exchanged
looks with each other.\tsk\break
Aye, \tsk aye, \textit{Trim!} quoth my uncle \textit{Toby}, shaking
his head,\tsh these are but sorry fortifications,
\textit{Trim}.\tsh O! very poor work, answered \textit{Trim}, to
what your Honour and I make of it.\tsh The char-\break acter of this
last man, said Dr.\@ \textit{Slop}, interrupting \textit{Trim},
is more detestable than\break
all the rest;\tsh and seems to have\break
been taken from some pettifogging Law-\break
yer amongst you:\tsk Amongst us, a\break
man’s conscience could not possibly continue so long
\textit{blinded};\tsk three times in\pb
a year, at least, he must go to confession. Will that restore it
to sight, quoth my uncle
\textit{Toby?}\tsk Go on, \textit{Trim}, quoth my father, or
\textit{Obadiah} will have got back before thou has got to the
end of thy sermon.\tsh ’tis a very short one, replied
\textit{Trim.}\tsk I wish it was longer, quoth my uncle
\textit{Toby}, for I like it hugely.\tsk \textit{Trim} went on.]

\begin{story}{A fourth man shall want even this}
refuge;\tsk shall break through all their ceremony of slow
chicane;\tsk scorns the doubtful workings of secret plots and
cautious trains to bring about his purpose:\tsh See the
bare-faced villain, how he cheats, lies, perjures, robs,
murders!\tsh Horrid!\tsh But indeed much better was not to be
expected, in the present case,\tsk the poor man was in the
dark!\tsh his priest had got\pb the keeping of his
conscience;\tsh and all he would let him know of it, was,
That he must believe in the Pope;\tsk\break
    go to Mass;\tsk cross
himself;\tsk tell his beads;\tsh be a good Catholic, and that
this, in all conscience, was enough to carry him to heaven.
What;\tsk if\break he perjures!\tsk Why;\tsk he had a men-\break tal
reservation in it.\tsk But if he is so wicked and abandoned a
wretch as you represent him;\tsk if he robs,\tsk if he
    stabs,\tsk 
will not conscience, on every such act, receive a wound
itself? Aye,\break
    \tsk but the man has carried it to con-\break
    fession;\tsk the wound digests there,\break
    and will do well enough, and in a\break
    short time be quite healed up by absolution. O Popery!
what hast thou to answer for?\tsk when not content with
the too many natural and fatal ways, thro’ which the heart of
man is\pb 
    every day thus treacherous to itself\break
    above all things;\tsk thou hast wilfully set open the wide gate of deceit
before the face of this unwary traveller, too apt, God knows, to go
astray of himself; and confidently speak peace to himself, when
there is no peace.
\end{story}
\vspace\parskip
\begin{story}{Of this the common instances which}
    I have drawn out of life, are too notorious to require much
    evidence. If any man doubts the reality of them, or thinks
    it impossible for a man to be such a bubble to himself,\tsk
    I must refer him a moment to his own reflections, and will
    then venture to trust my appeal with his own heart.  
\end{story}
\vspace\parskip
\begin{story}{Let him consider in how different} a degree of
detestation, numbers of\break
    wicked actions stand \textit{there}, tho’ equally\pb
    bad and vicious in their own natures;\break
    \tsk he will soon find, that such of them as strong inclination and custom
have prompted him to commit, are generally dress’d out and painted
with all the false beauties which a soft and a flattering hand can
give them;\tsk and\break that the others, to which he feels no
propensity, appear, at once, naked and deformed, surrounded with
all the\break
    true circumstances of folly and dis-\break honour.
\end{story}
\vspace\parskip
\begin{story}{When \textit{David} surprized \textit{Saul} sleep-}
    ing in the cave, and cut off the skirt of his robe\tsk we
    read his heart smote him for what he had done:\tsh But in\break
    the matter of \textit{Uriah}, where a faithful and gallant
    servant, whom he ought to have loved and honoured, fell to
    make way for his lust,\tsk where conscience\pb had so much
    greater reason to take the alarm, his heart smote him
    not.\break
    A whole year had almost passed from first commission of that
    crime, to the time \textit{Nathan} was sent to reprove him;
    and we read not once of the least sorrow or compunction of
    heart which he testified, during all that time, for what he
    had done.
\end{story}
\vspace\parskip
\begin{story}{Thus conscience, this once able mon-}
    itor,\tsh placed on high as a judge within us, and
intended by our maker as a just and equitable one too,\tsk by an
unhappy train of causes and impediments, takes often such imperfect
cognizance of what passes,\tsh does its office so
negligently,\tsh sometimes so corruptly,\tsk that it is
not to be trusted alone; and therefore we find there is\break
    a necessity, an absolute necessity of\pb
    joining another principle with
it, to aid, if not govern, its determinations.
\end{story}
\vspace\parskip
\begin{story}{So that if you would form a just}
    judgment of what is of infinite importance to you not to be
    misled in,\tsh\break namely, in what degree of real merit
    you stand either as an honest man, an useful citizen, a
    faithful subject to your king, “or a good servant to your
    God,\tsk\break call in religion and morality.\tsk Look, 
    \tsk What is written in the law of God?\break
    \tsh How readest thou?\tsk
    Consult calm reason and the unchangeable ob\-ligations of
    justice and truth;\tsh what say they?
\end{story}
\vspace\parskip
\begin{story}{Let \lss{\textsc{Conscience}} determine the}
    matter
upon these reports;\tsh and then if thy heart condemns
thee not, which is the case the Apostle supposes,\tsh the\pb
    \hbox to 160pt{rule will be infallible;” [Here Dr.\@
    \textit{Slop}}
\end{story}

\noindent fell asleep] 
\begin{story}{\textit{thou wilt have confidence to-}}
    \textit{wards God};\tsh that is, have just grounds to believe the
judgment thou hast past upon thyself, is the judgment of God; and
nothing else but an anticipation of that righteous sentence which
will be pronounced upon thee hereafter by that Being, to whom thou
art finally to give an account of thy actions.
\end{story}
\vspace\parskip
\begin{story}{\textit{Blessed is the man}, indeed, then, as} the
author of the book of \textit{Ecclesiasticus} expresses it, \textit{who is
not prick’d with the multitude of his sins: Blessed is the
man whose heart hath not condemn’d him;
whether he be rich, or whether he be poor, if he have a good
heart} (a heart thus guided and informed) \textit{he shall at all
times rejoice in a chearful countenance; his mind shall tell him
    more than\pb seven watch-men that sit above upon a} 
    \hbox to 160pt{\textit{tower on high}.”\tsh [A tower has}
\end{story}

\noindent
no strength, quoth my uncle \textit{Toby}, un-\break
less ’tis flank’d.]
\begin{story}{in the darkest doubts} it shall
conduct him safer than a thousand casuists, and give the state
    he\break
lives in, a better security for his behaviour than all the causes
and restrictions put together, which law-makers are forced to
multiply:\tsh \textit{Forced}, I say, as things stand; human laws
not being a matter of original choice, but of pure necessity,
brought in to fence against the mischievous effects of those
consciences which are no law unto themselves; well intending, by
the many provisions made,\tsh that in all such corrupt and
misguided cases, where prin\-ciples and the checks of
    conscience\break
    will not make us upright,\tsh to\pb
    supply their force, and, by the terrors of gaols and
halters, oblige us to it.”
\end{story}
\vspace\parskip

[I see plainly, said my father, that\break
this sermon has been composed to be\break 
preached at the Temple,\tsh or at some
Assize.\tsh I like the reasoning,\tsh and
am sorry that Dr.\@ \textit{Slop} has fallen asleep
before the time of his conviction;\tsh\break
for it is now clear, that the Parson, as
I thought at first, never insulted St.\@ \textit{Paul}
in the least;\tsk nor has there been, bro\-ther, the least difference
between them.\break
\tsh A great matter, if they had differed,
replied my uncle \textit{Toby},\tsk the best friends in the world
may differ sometimes.\tsh True,\tsk brother \textit{Toby}
quoth my father, shaking hands with him,\tsk\break
we’ll fill our pipes, brother, and then \textit{Trim} shall go on.

\newpage
Well,\tsk what dost thou think of it? said my father,
speaking to Corporal \textit{Trim}, as he reach’d his
tobacco-box.

\noindent
\stick{\hfill I think, answer’d the Corporal, that the}
seven watch-men upon the tower, who, I suppose, are all centinels there,\tsk
are more,
an’ please your Honour, than were necessary;\tsk and, to go
on at that rate, would harrass a regiment all to pieces, which
a commanding officer, who loves his men, will never do, if he can
help it;\break
\stick{because two centinels, added the Corporal,}
are as good as
twenty.\tsk I have been a commanding officer myself in the
\textit{Corps de Garde} a hundred times, continued \textit{Trim},
rising an inch higher in his figure, as he spoke,\tsk and all the
time I had the honour to serve his Majesty King \textit{William}, in
relieving the most considerable posts, I never left more than two
in my life.\tsk\pb
Very right, \textit{Trim}, quoth my uncle \textit{Toby},\break
\tsk but you do not consider, \textit{Trim}, that the
towers, in \textit{Solomon}’s days, were not such things as our
bastions, flank’d and defended by other works;\tsk this,
\textit{Trim}, was an invention since \textit{Solomon}’s death;
nor had they horn-works, or ravelins before the curtin, in his
time;\tsk or such a fossé as we make with a cuvette
in the middle of it, and with cover’d ways and counterscarps
pallisadoed along it, to guard against a \textit{Coup de
main:}\tsk So that the seven men upon the tower were a party,
I dare say, from the \textit{Corps de Garde}, set there, not only to
look out, but to defend it.\tsk They could be no more, an’
please your Honour, than a Corporal’s Guard.\tsk My father
smiled inwardly,\tsk\break but not outwardly\tsk the subject being rather
too serious, considering what had\pb happened, to make a jest
of.\tsk So putting his pipe into his mouth, which he had just
lighted,\tsk he contented himself with ordering \textit{Trim} to
read on. He read on as follows:]

\begin{story}{To have the fear of God before our}
    eyes, and, in our mutual dealings with each other, to govern
    our actions by the eternal measures of right and
    wrong:\tsk\break
    The first of these will comprehend the duties of
    religion;\tsh the second, those of morality, which are so
    inseparably connected together, that you cannot divide these
    two \textit{tables}, even in imagination, (tho’ the attempt
    is often made in practice) without breaking and mutually
    destroying them both.  
\end{story}
\vspace\parskip
\begin{story}{I said the attempt is often made,} and so it
is;\tsh there being nothing\pb more common than to see a man who has
no sense at all of religion,\tsk and indeed has so much honesty as to
pretend to none, who would take it as the bitterest affront, should
you but hint at a suspicion of his moral character,\tsh\break or
imagine he was not conscientiously just and scrupulous to the
uttermost mite.
\end{story}
\vspace\parskip
\begin{story}{When there is some appearance that} it is
so,\tsk tho’ one is unwilling even to suspect the
appearance of so amiable a virtue as moral honesty, yet were we to
look into the grounds of it, in the present case, I am persuaded we
should find little reason to envy such a one the honour of his
motive.
\end{story}
\vspace\parskip
\begin{story}{Let him declaim as pompously as}
    he chooses upon the subject, it will\pb
    be found to rest upon no better foundation than
either his interest, his pride, his ease, or some such little and
changeable passion as will give us but small dependence upon his
actions in matters of great distress.
\end{story}
\vspace\parskip

“I will illustrate this by an example.

\noindent
\stick{\indent “\kern 2pt I know the banker I deal with, or}
\stick{“\kern 2pt the physician I usually call in,” [There}
\stick{is no need, cried Dr.\@ \textit{Slop}, (waking) to call}
in any physician in this case] 
\begin{story}{to be}
    neither of them men of much
religion: I hear them make a jest of it every day, and treat all
its sanctions with so much scorn, as to put the matter past doubt.
Well;\tsk notwithstanding this, I put my fortune into the hands
of the one;\break\tsk and what is dearer still to me,\pb I trust my life
to the honest skill of the other.
\end{story}
\vspace\parskip
\begin{story}{Now let me examine what is my}
    reason for this great confidence.\tsh\break Why, in the
    first place, I believe there is no probability that either
    of them will employ the power I put into their hands to my
    disadvantage;\tsk I consider that honesty serves the
    purposes of this life:\tsk I know their success in the world
    depends upon the fairness of their characters.\tsh In a
    word,\tsh I’m persuaded that they cannot hurt me without hurting
    themselves more.
\end{story}
\vspace\parskip
\begin{story}{But put it otherwise, namely, that}
    interest lay, for once, on the other side; that a case
    should happen, wherein the one, without stain to his
    reputation,\break
    could secrete my fortune, and leave\pb
    me naked in
    the world;\tsk or that the other could send me out of it,
    and enjoy an estate by my death, without dishonour to
    himself or his art:\tsk In this case, what hold have I of
    either of them?\break\tsk Religion, the strongest of all motives,
    is out of the question;\tsk Interest, the next most powerful
    motive in the world, is strongly against me:\tsh\break What
    have I left to cast into the opposite scale to balance this
    temptation?\tsh\break
    Alas! I have nothing,\tsk nothing but what
    is lighter than a bubble\tsk I must lay at the mercy of
    \ls{\textsc{Honour}}, or some such capricious principle\tsk
    Strait security for two of the most valuable blessings!\tsk
    my property and myself.
\end{story}
\vspace\parskip
\begin{story}{As, therefore, we can have no de-}
    pendence upon morality without religion;\tsk so, on the
    other hand, there is\pb nothing better to be expected from
    religion without morality;\tsk neverthe\-less, ’tis no prodigy to
    see a man whose real moral character stands very low, who
    yet entertains the highest notion of himself in the light of
    a religious\break man.
\end{story}
\vspace\parskip
\begin{story}{He shall not only be covetous, re-}
    vengeful, implacable,\tsk but even\break
    wanting in points of common ho-\break
    nesty; yet inasmuch as he talks aloud\break
    against the infidelity of the age,\tsh\break
    is zealous for some points of reli-\break
    gion,\tsh goes twice a day to church,\break
    \tsh attends the sacraments,\tsh and\break
    amuses himself with a few instrumental\break
    parts of religion,\tsk shall cheat his\break
    conscience into a judgment, that, for\break
    this, he is a religious man, and has\break
    discharged truly his duty to God:\pb
    And you will find that such a man,\break
    through force of this delusion, generally looks
    down with spiritual pride\break
    upon every other man who has less
    affectation of piety,\tsk though, perhaps, ten times more
    real honesty than\break
    himself.
\end{story}
\vspace\parskip
\begin{story}{\textit{This likewise is a sore evil under the}}
    \textit{sun}; and I believe, there
is no one mistaken principle, which, for its time, has wrought more serious
mischiefs.\tsh\break
    For a general proof of this,\tsk examine 
    \hbox to 160pt{the history of the \textit{Romish} church;”\tsk} 
\end{story}

\noindent
\stick{[Well, what can you make of that, cried}
\stick{Dr.\@ \textit{Slop?}]\tsk “\kern 2pt see what scenes of cru-}
\stick{“\kern 2pt elty, murder, rapine, blood-shed,”}
\stick{[They may thank their own obstinacy,}
\stick{cried Dr.\@ \textit{Slop}] “\kern 2pt have all been sanctified}
\stick{“\kern 2pt by a religion not strictly governed by}
\stick{“\kern 2pt morality.}\newpage

“In how many kingdoms of the world,”\break
[Here \textit{Trim} kept waving his right-hand from the sermon
to the extent of his arm, returning it backwards and forwards to
the conclusion of the paragraph.]

\begin{story}{In how many kingdoms of the world}
    has the crusading sword of this mis\-guided saint-errant,
    spared neither age or merit, or sex, or condition?\tsk and,
    as he fought under the banners of a religion which set him
    loose from justice and humanity, he shew’d none; mercilessly
    trampled upon both,\tsh\break heard neither the cries of the
    unfortunate, nor pitied their distresses.” 
\end{story}
\vspace\parskip
[I have been in many a battle, an’ please your Honour, quoth \textit{Trim},
sighing, but never in so melancholy a one as this.\tsk I would not have drawn
a tricker\pb
in it against these poor souls,\tsh to have been made a general
officer.\tsh\break Why? what do you understand of the affair?
said Dr.\@ \textit{Slop}, looking towards \textit{Trim}, with
something more of contempt than the Corporal’s honest heart
deserved.\tsh\break What do you know, friend, about this battle
you talk of?\tsh I know, replied \textit{Trim}, that I never
refused quarter in my life to any man who cried out for
it;\tsh\break but to a woman or a child, continued
\textit{Trim}, before I would level my musket at them, I would
lose my life a thousand times.\tsk Here’s a crown for thee,
\textit{Trim}, to drink with \textit{Obadiah} to-night, quoth my
uncle \textit{Toby}, and I’ll give \textit{Obadiah} another
too.\tsk God bless your Honour, replied \textit{Trim},\tsh I had
rather these poor women and children had it.\tsh\break Thou art
an honest fellow, quoth my\break
uncle \textit{Toby.}\tsh My father
nodded his\pb head,\tsh as much as to say\tsh and so\break he is.\tsh

But pri’thee, \textit{Trim}, said my father, make an end,\tsk for
I see thou hast but a leaf or two left.]

\vspace\parskip
\centerline{Corporal \textit{Trim} read on.}

\begin{story}{If the testimony of past centuries}
    in this matter is not sufficient,\tsk consider at this
    instant, how the votaries of that religion are every day
    thinking to do service and honour to God, by actions which
    are a dishonour and scandal to themselves.  
\end{story}
\vspace\parskip

\noindent
\stick{\indent “\kern 2pt To be convinced of this, go with}
\stick{“\kern 2pt me for a moment into the prisons of}
\stick{“\kern 2pt the Inquisition.” [God help my poor}
\stick{brother \textit{Tom}.]\tsk “\kern 2pt Behold \textit{Religion}, with}
\newpage

\stick{“\kern 3pt \textit{Mercy} and \textit{Justice} chained down under}
\stick{“\kern 3pt her feet,\tsh there sitting ghastly upon}
\stick{“\kern 3pt a black tribunal, propp’d up with racks}
\stick{“\kern 3pt and instruments of torment. Hark!}
\stick{“\kern 3pt \tsk hark!  what a piteous groan!” [Here}
\stick{\textit{Trim}’s face turned as pale as ashes.]}
\stick{“\kern 3pt See the melancholy wretch who ut-}
\stick{“\kern 3pt ter’d it”\tsk [Here the tears began to}
\stick{trickle down]\tsh “\kern 3pt just brought forth to un-}
\stick{“\kern 3pt dergo the anguish of a mock trial, and}
\stick{“\kern 3pt endure the utmost pains that a studied}
\stick{“\kern 3pt system of cruelty has been able to in-}
\stick{“\kern 3pt vent.”\tsk [D\tsk n them all, quoth \textit{Trim},}
\stick{his colour returning into his face as red}
\stick{as blood.]\tsk “\kern 3pt Behold this helpless victim}
\stick{“\kern 3pt delivered up to his tormentors,\tsk his}
\stick{“\kern 3pt body so wasted with sorrow and con-}
\stick{“\kern 3pt finement.\,”\tsk [Oh!\@ ’tis my brother,}
\stick{cried poor \textit{Trim} in a most passionate}
\stick{exclamation, dropping the sermon upon}
\stick{the ground, and clapping his hands to-}
\newpage
\noindent
gether\tsk I fear ’tis poor
\textit{Tom.} My father’s and my uncle \textit{Toby}’s heart
yearn’d with sympathy for the poor fellow’s distress,\tsk even
\textit{Slop} himself acknowledged pity for him.\tsh Why,
\textit{Trim}, said my father, this is not a history,\tsh
’tis a sermon thou art reading;\tsk pri’thee begin the sentence
again.]\tsh “\kern 3pt Behold this helpless\break
\stick{“\kern 3pt victim deliver’d up to his tormentors,}
\stick{“\kern 3pt \tsk his body so wasted with sorrow and}
\stick{“\kern 3pt confinement, you will see every nerve}
\stick{“\kern 3pt and muscle as it suffers.\hfill}

\noindent
\stick{\indent “\kern 3pt Observe the last movement of that}
\stick{“\kern 3pt horrid engine!” [I would rather face}
\stick{a cannon, quoth \textit{Trim}, stamping.]\tsh}
\begin{story}{See what convulsions it has thrown}
    him into!\tsh Consider the nature of
    the posture in which he how lies
    stretched,\tsk what exquisite tortures he
    \hbox to 160pt{endures by it!”\tsh [I hope ’tis not in}
\end{story}
\newpage
\noindent
\stick{\textit{Portugal.}]\tsk “\kern 2pt ’Tis all nature can bear!} 
\stick{“\kern 3pt Good God! see how it keeps his weary}
\stick{“\kern 3pt soul hanging upon his trembling lips!”}
[I would not read another line of it,\break
quoth \textit{Trim} for all this \textit{world};\tsk I fear,
an’ please your Honours, all this is in
\textit{Portugal}, where my poor brother \textit{Tom} is.\break
I tell thee, \textit{Trim}, again, quoth my father, 
’tis not an historical account,\tsk ’tis a de\-scription.\tsk
’Tis only a description, honest 
man, quoth \textit{Slop}, there’s not a word of
truth in it.\tsk That’s another story, replied
my father.\tsk However, as \textit{Trim} reads it
with so much concern,\tsk ’tis cruelty to
force him to go on with it.\tsk Give me
hold of the sermon, \textit{Trim},\tsk I’ll finish it 
for thee, and thou may’st go. I must stay
and hear it too, replied \textit{Trim}, if your Hon\-our
will allow me;\tsk tho’ I would not
read it myself for a Colonel’s pay.\tsh\pb
Poor \textit{Trim!} quoth my uncle \textit{Toby}. My
father went on.]

\begin{story}{\tsk Consider the nature of the posture}
    in which he now lies stretched,\tsk what exquisite torture he endures by
    it!\tsk\break ’Tis all nature can bear! Good God! See how it keeps his weary
    soul hanging upon his trembling lips,\tsk willing to take its leave,\tsh but not
    suffered\break to depart!\tsk Behold the unhappy\break 
    \hbox to 160pt{wretch led back to his cell!” [Then}
\end{story}
\noindent
thank God, however, quoth \textit{Trim}, they\break
have not killed him]\tsk
\begin{story}{See him dragg’d}
    out of it again to meet the flames, and the insults in his last
agonies, which this principle,\tsh this principle, that there can be religion
without mercy,
    \hbox to 160pt{has prepared for him.” [Then, thank}
\end{story}

\noindent God,\tsk he is dead, quoth \textit{Trim},\tsk he is\pb
out of his pain,\tsk and they have
done their worst at him.\tsk O Sirs!\tsk Hold your peace,
\textit{Trim}, said my father, going on with the sermon, lest
\textit{Trim} should incense Dr.\@ \textit{Slop},\tsk we shall never
have done at this rate.]

\begin{story}{The surest way to try the merit of} any disputed notion is, to trace down
the consequences such a notion has produced, and compare them with
the spirit of Christianity;\tsh ’tis the short\break
    and decisive rule which our Saviour\break
    hath left us, for these and such-like\break cases,
    and it is worth a thousand ar\-guments,\tsk \textit{By their fruits ye shall
    know\break
    them.} \end{story}
\vspace\parskip
\begin{story}{I will add no farther to the length}
    of this sermon, than by two or three\pb
    short and independent rules deducible from it.
\end{story}
\vspace\parskip
\begin{story}{\textit{First}, Whenever a man talks loudly} against religion, always suspect that
it is not his reason, but his passions, which have got the better of his
    \ls{\textsc{Creed}}. A bad life and a good belief are disagreeable and troublesome
neighbours, and where they separate, depend upon it, ’tis for no other cause but
quietness sake.
\end{story}
\vspace\parskip
\begin{story}{\textit{Secondly}, When a man, thus represen-}
    ted, tells you in any particular instance,\break
    \tsk That such a thing goes \textit{against} his conscience,\tsk always believe he means exactly the same
    thing, as when he tells you such a thing goes \textit{against} his stomach;\tsk
    a present want of appetite being generally the true cause of both.
\end{story}
\newpage
\begin{story}{In a word,\tsk trust that man in no-}thing, who has not a
    \ls{\textsc{Conscience}} in every
thing.
\end{story}
\vspace\parskip
\begin{story}{And, in your own case, remember} 
    this plain distinction, a mistake in\break
    which has ruined thousands,\tsk that\break
    your conscience is not a law:\tsk No,\break 
    God and reason made the
law, and have placed conscience within you to determine;\tsh not, like an
\textit{Asiatic} Cadi, according to the ebbs and flows of his own passions,\tsk but
like a \textit{British} judge in this land of liberty and good sense, who makes no
new law, but faithfully declares that law which he knows already written.”
\end{story}

\vskip 32pt
\centerline{\lsss{\textit{F I N I S}}.}

\newpage
Thou hast read the sermon extremely well, \textit{Trim}, quoth
my father.\tsk If he had spared his comments, replied Dr.\@
\textit{Slop}, he would have read it much better.  I should have
read it ten times better, Sir, answered \textit{Trim}, but that
my heart was so full.\tsk That was the very reason,
\textit{Trim}, replied my father, which has made thee 
\stick{read the sermon as well as thou hast done;}
\stick{and if the clergy of our church, continued}
\stick{my father, addressing himself to Dr.\@ \textit{Slop},}
would take part in what they deliver, as deeply as
this poor fellow has done,\tsk as their compositions are fine;
(I deny it, quoth Dr.\@ \textit{Slop}) I maintain it,
that the eloquence of our pulpits, with such subjects to enflame
it,\tsk would be a model for the whole world:\tsk But alas!
continued my father, and I own it, Sir, with sorrow, that, like
\textit{French} politicians in this respect, what they gain in
the cabinet\pb they lose in the field.\tsh ’Twere a pity, quoth my
uncle, that this should be lost. I like the sermon well, replied
my father,\tsh ’tis dramatic,\tsk and there is something in
that way of writing, when skilfully managed, which catches the
attention.\tsh We preach much in that way with us, said Dr.\@
\textit{Slop.}\tsk I know that very well, said my father,\tsk
but in a tone and manner which disgusted Dr.\@ \textit{Slop},
full as much as his assent, simply, could have pleased him.\tsh
But in this, added Dr.\@ \textit{Slop}, a little piqued,\tsk our
sermons have greatly the advantage, that we never introduce any
character into them below a patriarch or a patriarch’s wife, or
a martyr or a saint.\tsk There are some very bad characters in
this, however, said my father, and I do not think the sermon a
jot the worse for ’em.\tsh But pray, quoth my uncle
\textit{Toby},\tsk who’s can this\pb be?\tsk How could it get into
my \textit{Stevinus}? A man must be as great a conjurer as
\textit{Stevinus}, said my father, to resolve the second
question:\tsk The first, I think, is not so difficult;\tsh for
unless my judg\-ment greatly deceives me,\tsh I know the author,
for ’tis wrote, certainly, by the parson of the parish.

The similitude of the stile and man-\break
ner of it, with those my father con-\break
stantly had heard preach’d in his pa-\break
rish-church, was the ground of his con-\break
jecture,\tsh proving it as strongly, as an argument
\textit{à priori} could prove such a thing to a philosophic
mind, That it was \textit{Yorick}’s and no one’s
else:\tsh\break 
It was proved to be so, \textit{à posteriori},\break
the day after, when \textit{Yorick} sent a servant to my uncle
\textit{Toby}’s house to enquire\break
after it.

It seems that \textit{Yorick}, who was inquisitive after all kinds
of knowledge, had borrowed \textit{Stevinus} of my uncle \textit{Toby},
and had carelesly popped his sermon, as soon as he had made it,
into the middle of \textit{Stevinus}; and by an act of forgetfulness,
to which he was ever subject, he had sent \textit{Stevinus} home, and
his sermon to keep him company.

Ill-fated sermon! Thou wast lost, after this recovery of thee, a
second time, dropped thro’ an unsuspected fissure in thy
master’s pocket, down into a treacherous and a tatter’d
lining,\tsk trod deep into the dirt by the left hind foot of
his Rosinante, inhumanly stepping upon thee as thou
falledst;\tsk buried ten days in the mire,\tsh raised up
out of it by a beggar, sold for a halfpenny to a
parish-clerk,\tsk\break transferred to his
parson,\tsh lost for ever\pb to thy own, the remainder of his
days,\tsk\break nor restored to his restless \ls{\textsc{Manes}} till
this very moment, that I tell the world the story.

Can the reader believe, that this sermon of
\textit{Yorick}’s was preach’d at an assize, in the cathedral
of \textit{York}, before a thousand witnesses, ready to give oath of
it, by a certain prebendary of that church, and actually printed by
him when he had done,\tsh and within so short a space as
two years and three months after \textit{Yorick}’s
death.\tsk \textit{Yorick} indeed, was never better served in his
life;\tsh but it was a little hard to male-treat him after,
and plunder him after he was laid in his grave.

However, as the gentleman who did it was in perfect charity with
\textit{Yorick},\tsk\break
and, in conscious justice, printed but a\pb
few copies to give away;\tsk and that, I am told, he could
moreover have made as good a one himself, had he thought
fit,\break \tsk I declare I would not have published this
anecdote to the world;\tsk nor do I publish it with an intent to
hurt his character and advancement in the church;\tsk\break I
leave that to others;\tsh but I find myself impelled by two
reasons, which I cannot withstand.

The first is, That, in doing justice, I may give rest to
\textit{Yorick}’s ghost;\tsh which, as the
country people,\tsk and some others, believe,\tsh \textit{still
walks}.

The second reason is, That, by laying open this story to the
world, I gain an opportunity of informing it,\tsk That in case
the character of parson \textit{Yorick}, and this sample of his
sermons is liked,\tsh that\pb
there are now in the possession of
the \textit{Shandy} family, as many as will make a handsome volume,
at the world’s service,\tsh and much good may they\break
do it.

\bigskip

\section{Chap. XVIII.}

\lettrine{O}{\ls{badiah}} gain’d the two crowns\break without
dispute;\ for he came in jingling, with all the instruments in
the green bays bag we spoke of, slung across his body, just as
Corporal \textit{Trim} went out of the room.

It is now proper, I think, quoth Dr.\@ \textit{Slop}, (clearing up
his looks) as we are in a condition to be of some service to
Mrs.\@ \textit{Shandy}, to send up stairs to know\break how she goes on.

\newpage
I have ordered, answered my father, the old midwife to come down
to us upon the least difficulty;\tsh for you\break
must know, Dr.\@ \textit{Slop}, continued my father, with a perplexed kind of a
smile upon his countenance, that by express treaty, solemnly
ratified between me and my wife, you are no more than an
auxiliary in this affair,\tsk and not so much as that,\tsk
unless the lean old mother of a midwife above stairs cannot do
without you.\tsh Women have their particular fancies, and in
points of this nature, continued my father, where they bear the
whole burden, and suffer so much acute pain for the advantage of
our families, and the good of the species,\tsk they claim a
right of deciding, \textit{en Souveraines}, in whose hands, and
in
what fashion, they chuse to undergo it.

\newpage
They are in the right of it,\tsh quoth\break
my uncle \textit{Toby.} But, Sir, replied Dr.\@ \textit{Slop}, not taking notice of
my uncle \textit{Toby}’s opinion, but turning to my
father,\tsk they had better govern in other
points;\tsh and a father of a family, who wished its
perpetuity, in my opinion, had better exchange this prerogative
with them, and give up some other rights in lieu of
it.\tsh\break
I know not, quoth my father, answering a little too testily, to
be quite dispassionate in what he said,\tsk I know not, quoth
he, what we have left to give up, in lieu of who shall bring our
children into the world,\tsk unless that,\tsk of who shall beget
them.\tsh One would almost give up any thing, replied Dr.\@
\textit{Slop.}\tsk I beg your pardon,\tsh answered my uncle
\textit{Toby.}\break\tsk Sir, replied Dr.\@ \textit{Slop}, it
would astonish you to know what Improvements we have made of
late years in all branches\pb
of obstetrical knowledge, but partic\-u\-larly in that
one single point of the safe\break and expeditious extraction of the
\textit{fœtus},\break
\tsh which has received such lights, that, 
for my part (holding up his hand) I 
declare I wonder how the world has\tsh\break
I wish, quoth my uncle \textit{Toby}, you had
seen what prodigious armies we had in \textit{Flanders}.

\bigskip

\section{Chap. XIX.}

\lettrine{I}{\, have} dropp’d the curtain over this\break
scene for a minute,\tsk to remind you of one
thing,\tsk and to inform you of an\-other.

What I have to inform you, comes, I own, a little out of its due
course;\tsh for it should have been told a hundred and\pb
fifty pages ago, but that I foresaw then ’twould come in pat hereafter, and be of
more advantage here than elsewhere.\tsh\break
Writers had need look before them, to keep up the spirit and
connection of what they have in hand.

When these two things are done,\tsk the curtain shall be drawn
up again, and my uncle \textit{Toby}, my father, and Dr.\@ \textit{Slop},
shall go on with their discourse, without any more
interruption.

First, then, the matter which I have to remind you of, is
this;\tsh that from the specimens of singularity in my
father’s notions in the point of Christian-names, and that
other previous point thereto,\tsk\break
you was led, I think, into an opinion,
(and I am sure I said as much) that my father was a
gentleman altogether as odd\pb
and whimsical in fifty other opinions.
In truth, there was not a stage in the life of man, from the very
first act of his begetting,\tsh down to the lean and
slippered pantaloon in his second childishness, but he had some
favourite notion to himself, springing out of it, as sceptical, and
as far out of the high-way of thinking, as these two which have been
explained.

\tsk Mr.\@ \textit{Shandy}, my father, Sir, would see nothing in
the light in which others placed it;\tsk he placed things in his
own light;\tsk he would weigh nothing in common scales;\tsk
no,\tsk he was too refined a researcher to lay open to so gross an
imposition.\tsk To come at the exact weight of things in the
scientific steel-yard, the fulcrum, he would say, should be almost
invisible, to avoid all friction from popular tenets;\tsk without
this the minutiæ\pb of philosophy, which would always turn the
balance, will have no weight at all.\tsk\break
Knowledge, like matter, he would affirm,
\stick{was divisible \textit{in infinitum};\tsk that the grains}
and scruples were as much a part of it, as the gravitation of the whole world.\tsk In
a word, he would say, error was error,\tsk\break
no matter where it fell,\tsh whether in a fraction,\tsh or a pound,\tsk ’twas alike
fatal to truth, and she was kept down at the bottom of her well,
as inevitably by a mistake in the dust of a butterfly’s
wing,\break\tsk as in the disk of the sun, the moon, and all the stars
of heaven put together.

He would often lament that it was for
\stick{want of considering this properly, and of}
applying it skilfully to civil matters, as well
as to speculative truths, that so many things in this world were
out of joint;\tsk\break
that the political arch was giving way;\tsk\pb
and that the very foundations of our excellent constitution in
church and state, were so sapp’d as estimators had reported.

You cry out, he would say, we are a ruined, undone people.\tsk
Why? he would 
\stick{ask, making use of the sorites or syllogism} 
\stick{of \textit{Zeno} and \textit{Chrysippus}, without knowing}
\stick{it belonged to them.\tsk Why? why are we}
a ruined people?\tsk Because we are
corrup\-ted.\tsk Whence is it, dear Sir, that we are
corrupted?\tsk Because we are needy;\tsk our
poverty, and not our wills, consent.\tsk\break
And wherefore, he would add, are we needy?\tsk From the neglect, he would answer,
of our pence and our halfpence:\tsk\break
Our bank notes, Sir, our guineas,\tsk nay our shillings,
take care of themselves.

’Tis the same, he would say, throughout the whole circle
of the sciences;\tsk\pb
the great, the established points of them, are not to be broke
in upon.\tsk The laws of nature will defend themselves;\tsk but
error\tsk (he would add, looking earnestly at my mother)\tsk
error, Sir, creeps in thro’ the minute holes, and small
crevices, which human nature leaves unguarded.

This turn of thinking in my father, is what I had to remind you
of:\tsk The point you are to be informed of, and which I have
reserved for this place, is as follows.

Amongst the many and excellent reasons, with which my father had
urged\break my mother to accept of Dr.\@ \textit{Slop}’s as\-sistance
preferably to that of the old wo\-man,\tsk there was one of a very
singular nature; which, when he had done arguing the matter with
her as a Christian,\break
and came to argue it over again with her\pb
as a philosopher, he had put his whole strength to, depending
indeed upon it as his sheet-anchor.\tsh It failed him, tho’ from
no defect in the argument itself; but that, do what he could, he
was not able for his soul to make her comprehend the drift of
it.\tsh Cursed luck!\tsh said he to himself, one afternoon, as
he walk’d out of the room, after he had been stating it for an
hour and a half to her, to no manner of purpose;\tsk cursed
luck! said he, biting his lip as he shut the door,\break
\tsk for a man
to be master of one of the finest chains of reasoning in
nature,\tsh\break
and have a wife at the same time with such a
head-piece, that he cannot hang up a single inference within
side of it, to save his soul from destruction.

This argument, tho’ it was entirely lost upon my
mother,\tsk had more weight with\pb him, than all his other
arguments joined together:\tsk I will therefore endeavour to do
it justice,\tsk and set it forth with all the perspicuity I am
master of.

My father set out upon the strength of these two following
axioms:

\noindent
\stick{\hfill\textit{First}, That an ounce of a man’s own wit,}
was worth a tun of other peoples; and,

\textit{Secondly}, (Which, by the bye, was the ground-work of the
first axiom,\tsk tho’ it comes last) That every
man’s wit must come from every man’s own
soul,\tsh and no other body’s.

Now, as it was plain to my father, that all souls were by nature
equal,\tsk and that the great difference between the most
acute and the most obtuse understanding\pb
\tsk was from no
original sharpness or bluntness of one thinking substance above or
below another,\tsk but arose merely from the lucky or
unlucky organization of the body, in that part where the soul
principally took up her residence,\tsh he had made it the
subject of his enquiry to find out the identical place.

Now, from the best accounts he had been able to get of this
matter, he was satisfied it could not be where \textit{Des
Cartes} had fixed it, upon the top of the \textit{pineal}
gland of the brain; which, as he philo\-sophised, formed a cushion
for her about the size of a marrow pea; tho’ to speak the
truth, as so many nerves did terminate all in that one
place,\tsk ’twas no bad conjecture;\tsk and my
father had certainly fallen with that great philosopher plumb into
the center of the mistake, had it not\pb been for my uncle
\textit{Toby}, who rescued him out of it, by a story he told him of a
\textit{Walloon} officer at the battle of \textit{Landen}, who had one
part of his brain shot away by a musket-ball,\tsk and another
part of it taken out after by a \textit{French} surgeon; and, after
all, recovered, and did his duty very well without it.

If death, said my father, reasoning with himself, is nothing but
the separation of the soul from the body;\tsk and if it is true
that people can walk about and do their business without
brains,\tsk then certes the soul does not inhabit there.
\quad Q.\,E.\,D.

As for that certain, very thin, subtle, 
\stick{and very fragrant juice which \textit{Coglionissimo}}
\stick{\textit{Borri}, the great \textit{Milaneze} physician af-}
\stick{firms, in a letter to \textit{Bartholine}, to have dis-}
\stick{covered in the cellulæ of the occipital parts}\pb 
of the cerebellum, and which he likewise
\stick{affirms to be the principal seat of the rea-}
\stick{sonable soul, (for, you must know, in these}
\stick{latter and more enlightened ages, there}
\stick{are two souls in every man living,\tsk the}
one, according to the great \textit{Metheglingius},
being called the \textit{Animus}, the other, the
\textit{Anima});\tsk as for the opinion, I say of
\textit{Borri},\tsk my father could never subscribe to it by any
means; the very idea of so 
\stick{noble, so refined, so immaterial, and so ex-}
\stick{alted a being as the \textit{Anima}, or even the \textit{Ani-}}
\stick{\textit{mus}, taking up her residence, and sitting}
\stick{dabbling, like a tad-pole all day long, both}
\stick{summer and winter, in a puddle,\tsk or in a}
\stick{liquid of any kind, how thick or thin soever,}
\stick{he would say, shock’d his imagination; he}
would scarce give the doctrine a hearing.

\noindent
\stick{\hfill What, therefore, seem’d the least liable}
to objections of any, was, that the chief\pb sensorium, or
head-quarters of the soul, and to which place all intelligences
were referred, and from whence all her mandates were issued,\tsk
was in, or near, the\break
cerebellum,\tsk or rather some-where about
the \textit{medulla oblongata}, wherein it was ge\-nerally agreed
by \textit{Dutch} anatomists, that
\stick{all the minute nerves from all the organs}
\stick{of the seven senses concentered, like streets} and
winding alleys, into a square.

So far there was nothing singular in my father’s
opinion,\tsk he had the best of philosophers, of all ages and
climates, to go along with him.\tsk But here he took a\break
road of his own, setting up another \textit{Shan\-dean} hypothesis upon
these corner-stones they had laid for him;\tsk and which
said hypothesis equally stood its ground; whe\-ther the subtilty and
fineness of the soul depended upon the temperature and clear-\pb ness of
the said liquor, or of the finer net-work and texture in the
cerebellum itself; which opinion he favoured.

He maintained, that next to the due care to be taken in the act
of propagation of each individual, which required all the thought
in the world, as it laid the foundation of this incomprehensible
contexture, in which wit, memory, fancy, eloquence, and what is
usually meant by the name of good natural parts, do
consist;\tsk that next to this and his Christian-name, which were
the two original and
\stick{most efficacious causes of all;\tsk that the} 
third cause, or rather what logicians
call the \textit{Causa sina quâ non}, and without which all that was
done was of no manner of 
\stick{significance,\tsk was the preservation of this} delicate and fine-spun web, from the havock
which was generally made in it by\pb
the violent compression and crush
which the head was made to undergo, by the nonsensical method of
bringing us into the world by that foremost.

\tsh This requires explanation.

My father, who dipp’d into all kinds of books, upon looking into \textit{Lithopædus
Senonesis de Partu difficili}\ \footnote{\fontsize{8}{11}\selectfont\quad 
The author is here twice mistaken;\tsh for \textit{Lithopædus} should be wrote thus,
\textit{Lithopædii Seno\-nensis Icon.} The second mistake is, that this
\textit{Li\-thopædus} is not an author, but a drawing of a petrified child. The
account of this, published by \textit{Athosius}, 1580, may be seen at the end of
\textit{Cordæus}’s works in \textit{Spachius.} Mr.  \textit{Tristram Shandy} has
been led into this error, either from seeing \textit{Lithopædus}’s name of late in a
catalogue of learned writers in \hbox{Dr. \tsh}, or by mistaking \textit{Lithopædus}
for \textit{Trinecavellius},\tsk from the too great similitude of the names.},
published by\pb
\textit{Adrianus Smelvgot}, had found out, That\break
the lax and pliable state of a child’s\break
head in parturition, the bones of the\break
cranium having no sutures at that time,\break
was such,\tsk that by force of the woman’s\break
efforts, which, in strong labour-pains,\break
was equal, upon an average, to the
weight of 470 pounds averdupoise acting perpendicularly upon it;\tsk it so happened,
that in 49 instances out of 50, the said head was compressed and moulded into the
shape of an oblong conical piece of dough, such as a pastry-cook generally rolls up
in order to make a pye of.\tsh\break
Good God!  cried my father, what havock and destruction
must this make in the infinitely fine and tender texture of the cerebellum!\tsk Or if
there is such a juice as \textit{Borri} pretends\tsk is it not enough to make the
clearest liquid in the world both feculent and mothery?

\newpage
But how great was his apprehension, when he farther understood,
that this force acting upon the very vertex of the head, not only
injured the brain itself, or cerebrum,\tsk but that it
necessarily squeez’d and propell’d the cerebrum towards the
cerebellum, which was the immediate seat of the
understanding!\tsh\break
Angels and ministers of grace defend
us! cried my father,\tsk can any soul withstand this
shock?\tsk No wonder the intellectual web is so rent and tattered
as we see it; and that so many of our best heads are no better than
a puzzled skein of silk,\tsk all
perplexity,\tsk all confusion within side.

But when my father read on, and was let into the secret, that
when a child was turn’d topsy-turvy, which was easy for an operator
to do, and was extracted by the feet;\tsk that instead of the
cerebrum being propell’d towards the cerebellum, the cerebellum, on
the contrary, was propell’d simply towards the cerebrum, where it could
do no manner of hurt:\tsh By heavens! cried he, the world
is in conspiracy to drive out what little wit God has given
us,\tsk and the professors of the obstetric art are listed
into the same conspiracy.\tsk What is it to me which end of my
son comes foremost into the world, provided all goes right after,
and his cerebellum escapes uncrushed?

It is the nature of an hypothesis, when once a man has conceived
it, that it assimilates every thing to itself, as proper
nourishment; and, from the first moment of your begetting it, it
generally grows the stronger by every thing you see, hear, read, or
understand. This is of great\break use.

\newpage
When my father was gone with this about a month, there was
scarce a phæ\-nomenon of stupidity or of genius, which he could
not readily solve by it;\tsk it ac-\break counted for the eldest son
being the greatest blockhead in the family.\tsk Poor
devil, \stick{he would say,\tsk he made way for the capa-}
\stick{city of his younger brothers.\tsk It unriddled}
the observations of drivellers and monstrous heads,\tsk shewing \textit{à
priori}, it could not be otherwise,\tsk unless \astiv\ I
don’t know what. It wonderfully explain’d and accounted for
the acumen of the \textit{Asiatic} genius, and that sprightlier turn,
and a more penetrating intuition of minds, in warmer climates; not
from the loose and common-place solution of a clearer sky, and a
more perpetual sunshine, \etc\tsk\break which, for aught he knew,
might as well rarefy and dilute the faculties of the soul into
nothing, by one extreme,\tsk as they\pb
are condensed in colder climates by the other;\tsh but he traced the affair up to
its spring-head;\tsk shew’d that, in warmer climates, nature had laid a lighter tax
upon the fairest parts of the creation;\tsk\break
their pleasures more;\tsk the necessity of their pains less, insomuch that the
pressure and resistance upon the vertex was so slight, that the whole organization
of the cerebellum was preserved;\tsh nay, he did not believe, in natural births,
that so much as a single thread of the net-work was broke or displaced,\tsh so that
the soul might just act as she liked.

\enlargethispage{\baselineskip}
When my father had got so far,\tsh\break
what a blaze of light did the accounts of the \textit{Caesarian} section, and of the
towering geniuses who had come safe into the world by it, cast upon
this hypothesis? Here you see, he would say, there was no injury
done to the sensorium;\tsk no pres-\pb sure of the head against the
pelvis;\tsh no propulsion of the cerebrum towards the
cerebellum, either by the \textit{os pubis} on this side, or \textit{os
coxygis} on that;\tsk and pray, what were the happy
consequences?\break Why, Sir, your \textit{Julius Caesar}, who gave the
operation a name;\tsk and your \textit{Hermes Trismegistus}, who
was born so before ever the operation had a name;\tsk your
\stick{\textit{Scipio Africanus}; your \textit{Manlius Torquatus};}
\stick{our \textit{Edward} the Sixth,\tsk who, had he lived,}
would have done
the same honour to the hypothesis:\tsk These, and many
more who figur’d high in the annals of fame,\tsk all came
\textit{side-way}, Sir, into the world.

The incision of the \textit{abdomen} and \textit{ute\-rus},
ran for six weeks together in my father’s
head;\tsk he had read, and was satisfied, that wounds in
the \textit{epigastrium}, and those in the \textit{matrix}, were not
mortal;\tsk\break
so that the belly of the mother might be\pb
opened extremely well to give a passage to the child.\tsk He mentioned the thing one
afternoon to my mother,\tsk merely as a matter of fact; but seeing her turn as pale
as ashes at the very mention of it, as much as the operation flattered his
hopes,\tsk he thought it as well to say no more of it,\tsk contenting himself with
admiring,\tsk what he thought was to no purpose to propose.

This was my father Mr. \textit{Shandy}’s hypothesis; concerning
which I have only to add, that my brother \textit{Bobby} did as
great honour to it (whatever he did to the family) as any one of
the great heroes we spoke of:\tsk For happening not only to be
christen’d, as I told you, but to be born too, when my father
was at \textit{Epsom},\tsk\break
being moreover my mother’s \textit{first} child,\break
\tsk coming into the world with his head
\textit{foremost},\tsk and turning out afterwards a\pb lad of
wonderful slow parts,\tsh my father spelt all these together
into his opinion: and as he had failed at one end,\tsk he was
determined to try the other.

This was not to be expected from one of the sisterhood, who are
not easily to be put out of their way,\tsh and was
therefore one of my father’s great reasons in favour of a man
of science, whom he could better deal with.

Of all men in the world, Dr.\@ \textit{Slop} was the fittest for my
father’s purpose;\tsh for though his new-invented
forceps was the armour he had proved, and what he maintained, to be
the safest instrument of deliverance,\tsk yet, it seems, he had
scattered a word or two in his book, in favour of the very thing
which ran in my father’s fancy;\tsh tho’ not
with a view to the soul’s good in extracting by the feet,\pb as
was my father’s system,\tsk but for reasons merely
obstetrical.

This will account for the coalition betwixt my father and Dr.\@
\textit{Slop}, in the ensuing discourse, which went a little
hard against my uncle \textit{Toby.}\tsk In what manner a plain
man, with nothing but common sense, could bear up against two
such allies in science,\tsk is hard to conceive.\tsk\break
You may conjecture upon it, if you please,\break
\tsk and whilst your imagination is in motion, you may encourage
it to go on, and discover by what causes and effects in nature
it could come to pass, that my uncle \textit{Toby} got his
modesty by the wound he received upon his groin.\tsk You may
raise a system to account for the loss of my nose by
marriage articles,\tsk and shew the world how it could happen,
that I should have the misfortune to be called
\ls{\textsc{Tristram}}, in opposition to my father’s\pb
hypothesis, and the wish of the whole family, God-fathers and
God-mothers not excepted.\tsk These, with fifty other points left
yet unravelled, you may endeavour to solve if you have time;\tsh
but I tell you beforehand it will be in vain,\break\tsk for not the sage
\textit{Alquise}, the magician in Don \textit{Belianis} of
\textit{Greece}, nor the no less famous \textit{Urganda}, the
sorceress his wife, (were they alive) could pretend to come
within a league of the truth.

The reader will be content to wait for a full explanation of
these matters till the next year,\tsk when a series of
things will be laid open which he little expects.

\bigskip
\bigskip
\centerline{\textit{\lsss{END}\enspace of\enspace the}\enspace
\textsc{Second\enspace Volume}.}

\end{document}
