\documentclass[twoside]{article}
\usepackage[paperheight=395pt, paperwidth=240pt, width=166pt, height=308pt, top=42pt, headsep=2pt]{geometry}
\usepackage[noinfo, frame, a6, center, lualatex]{crop}
\usepackage{graphicx}
\usepackage{shandean}
\title{The Life and Opinions of Tristram Shandy, Gentleman}
\author{Laurence Sterne}
\date{1760}
\begin{document}
\pagestyle{empty}
\null
\vfill
\eject
\null
\vfill
\eject
\noindent
\stick{\fontsize{09}{14}\selectfont\hfill\lsv{70}{THE}\hfill}\\[16pt]
\stick{\fontsize{16}{24}\selectfont\hfill\lsv{180}{LIFE}\hfill}\\[12pt]
\stick{\fontsize{09}{14}\selectfont\hfill\lsv{70}{AND}\hfill}\\[18pt]
\stick{\fontsize{16}{24}\selectfont\hfill\lsv{70}{OPINIONS}\hfill}\\[12pt]
\stick{\fontsize{09}{14}\selectfont\hfill\lsv{50}{OF}\hfill}\\[17pt]
\stick{\fontsize{14}{20}\selectfont\hfill\lsv{20}{TRISTRAM}\enspace\lsv{20}{SHANDY},\hfill}\\[6pt]
\stick{\fontsize{10}{14}\selectfont\hfill\ss{Gentleman}.\hfill}

\vskip 16pt

\stick{\quad\g{Ταρασσει τοὺς Ἀνθρώπους οὐ τὰ Πράγματα,}}
\stick{\g{αλλα τὰ περι τῶν Πραγμάτων, Δογματα}.\hfill}

\vskip 16pt
\centerline{\lsv{44}{VOL}.\quad II.}
\vskip 10pt
\centerline{The\kern 5pt \ss{Second Edition}.}
\vskip 18pt
\bgroup\fontsize{9}{11}\selectfont
\centerline{\itshape\lsv{60}{LONDON}:}
\vskip 2pt
\centerline{\smaller Printed for R.\@ and J.\@ \lss{\s{Dodsley}} in \i{Pall-Mall}.}
\centerline{M.DCC.LX.}
\par
\egroup
\eject
\null
\vfill
\eject
\setcounter{page}{1}
\pagestyle{folio}
\thispagestyle{empty}
\baselineskip 12.5pt
\parskip 12.5pt

\noindent
\stick{\tstrut 32pt\small\hfill \lsv{42}{THE}\hfill}
\stick{\tstrut 22pt\large\hfill \lsv{40}{LIFE}\enspace and\enspace \lsv{40}{OPINIONS}\hfill}
\stick{\tstrut 28pt\small\hfill \lsv{48}{OF}\hfill}
\stick{\tstrut 24pt\large       \lsv{16}{TRISTRAM}\enspace\lsv{16}{SHANDY}, Gent.}

\chap{I}{26pt}{8pt}

\initial[lines=3,lraise=0.12,loversize=-0.06,nindent=0.6pt]{I}{Have} begun a new book, on purpose\break
that I might have room enough to\break
\tight{explain the nature of the perplexities}\break
\stick{in which my uncle \toby was involved,}
\stick{from the many discourses and interroga-}
\stick{tions about the siege of \i{Namur}, where}
he received his wound.

I must remind the reader, in case he\break
\stick{has read the history of King \i{William}’s}
\catch{wars,}
\stick{wars,\tsk but if he has not,\tsk I then inform}
\stick{him, that one of the most memorable at-}
\stick{tacks in that siege, was that which was}
\stick{made by the \i{English} and \i{Dutch} upon the}
\stick{point of the advanced counterscarp, be-}
\stick{tween the gate of \i{St.\@ Nicolas}, which inclo-}
\stick{sed the great sluice or water-stop, where}
\stick{the \i{English} were terribly exposed to the}
\stick{shot of the counter-guard and demi-}
\stick{bastion of \i{St.\@ Roch:} The issue of which}
\stick{hot dispute, in three words, was this;}
\stick{That the \i{Dutch} lodged themselves upon}
\stick{the counter-guard,\tsk and that the \i{English}}
\stick{made themselves masters of the covered}
\stick{way before \i{St.\@ Nicolas}’\/s gate, notwith-}
\stick{standing the gallantry of the \i{French} offi-}
\stick{cers, who exposed themselves upon the}
glacis sword in hand.

As this was the principal attack of\break
\stick{which my uncle \toby was an eye-witness}
\catch{at}
\stick{at \i{Namur},\tsh the army of the besiegers}
\stick{being cut off, by the confluence of the}
\stick{\i{Maes} and \i{Sambre}, from seeing much of}
\stick{each other’s operations,\tsk my uncle \toby}
\stick{was generally more eloquent and parti-}
\stick{cular in his account of it; and the many}
\stick{perplexities he was in, arose out of the}
\stick{almost insurmountable difficulties he}
\stick{found in telling his story intelligibly,}
\stick{and giving such clear ideas of the differ-}
\stick{ences and distinctions between the scarp}
\stick{and counterscarp,\tsh the glacis and co-}
\stick{vered way,\tsh the half-moon and rave-}
\stick{lin,\tsh as to make his company fully}
\stick{comprehend where and what he was}
about.

Writers themselves are too apt to con-\break
\stick{found these terms;\tsh so that you will}
\stick{the less wonder, if in his endeavours to}
\stick{explain them, and in opposition to ma-}
\catch{ny}
\stick{ny misconceptions, that my uncle \toby}
\stick{did oft times puzzle his visiters, and}
sometimes himself too.

To speak the truth, unless the compa-\break
\stick{ny my father led up stairs were tolerably}
\stick{clear-headed, or my uncle \toby was in}
\stick{one of his best explanatory moods, ’twas}
\stick{a difficult thing, do what he could, to}
keep the discourse free from obscurity.

What rendered the account of this\break
\stick{affair the more intricate to my uncle}
\stick{\toby, was this,\tsk that in the attack of the}
\stick{counterscarp before the gate of \i{St.\@ Ni-}}
\stick{\i{colas}, extending itself from the bank of}
\stick{the \i{Maes}, quite up to the great water-}
\stick{stop;\tsk the ground was cut and cross-cut}
\stick{with such a multitude of dykes, drains,}
\stick{rivulets, and sluices, on all sides,\tsk and}
\stick{he would get so sadly bewilder’d and set}
\catch{fast}
\stick{fast amongst them, that frequently he}
\stick{could neither get backwards or for-}
\stick{wards to save his life; and was oft times}
\stick{obliged to give up the attack upon that}
very account only.

These perplexing rebuffs gave my\break
\stick{uncle \i{Toby Shandy} more perturbations}
\stick{than you would imagine; and as my}
\stick{father’s kindness to him was continually}
\stick{dragging up fresh friends and fresh in-}
\stick{quirers,\tsk he had but a very uneasy task}
of it.

No doubt my uncle \toby had great
\stick{command of himself,\tsk and could guard}
\stick{appearances, I believe, as well as most}
\stick{men;\tsk yet any one may imagine, that}
\stick{when he could not retreat out of the ra-}
\stick{velin without getting into the half-moon,}
\stick{or get out of the covered way without}
\catch{falling}
\stick{falling down the counterscarp, nor cross}
\stick{the dyke without danger of slipping into}
\stick{the ditch, but that he must have fretted}
\stick{and fumed inwardly:\tsk He did so;\tsk and}
\stick{these little and hourly vexations, which}
\stick{may seem trifling and of no account to}
\stick{the man who has not read \i{Hippocrates},}
\stick{yet, whoever has read \i{Hippocrates}, or}
\stick{Dr.\@ \i{James Mackenzie}, and has considered}
\stick{well the effects which the passions and}
\stick{affections of the mind have upon the di-}
\stick{gestion,\tsk (Why not of a wound as well}
\stick{as of a dinner?)\tsk may easily conceive}
\stick{what sharp paroxisms and exacerbations}
\stick{of his wound my uncle \toby must have}
undergone upon that score only.

\tsk My uncle \toby could not philoso-\break
\stick{phize upon it;\tsk ’twas enough he felt it}
\stick{was so,\tsk and having sustained the pain}
\stick{and sorrows of it for three months toge-}
\catch{ther,}
\stick{ther, he was resolved some way or other}
to extricate himself.

He was one morning lying upon his\break
\stick{back in his bed, the anguish and nature}
\stick{of the wound upon his groin suffering}
\stick{him to lye in no other position, when a}
\stick{thought came into his head, that if he}
\stick{could purchase such a thing, and have it}
\stick{pasted down upon a board, as a large}
\stick{map of the fortifications of the town and}
\stick{citadel of \i{Namur}, with its environs, it}
\stick{might be a means of giving him ease.\tsk}
\stick{I take notice of his desire to have the}
\stick{environs along with the town and cita-}
\stick{del, for this reason,\tsk because my uncle}
\stick{\toby's wound was got in one of the tra-}
\stick{verses, about thirty toises from the re-}
\stick{turning angle of the trench, opposite to}
\stick{the salient angle of the demi-bastion of}
\stick{\i{St.\@ Roch};\tsh so that he was pretty con-}
\catch{fident}
\stick{fident he could stick a pin upon the iden-}
\stick{tical spot of ground where he was stand-}
ing in when the stone struck him.

All this succeeded to his wishes, and\break
\stick{not only freed him from a world of sad}
\stick{explanations, but, in the end, it prov’d}
\stick{the happy means, as you will read, of}
\stick{procuring my uncle \toby his \s{Hobby-}}
\s{Horse}.

\chap{II}{36pt}{12pt}

\dropcap{T}{here} is nothing so foolish, when\break
you are at the expence of making\break
\stick{an entertainment of this kind, as to or-}
\stick{der things so badly, as to let your cri-}
\stick{ticks and gentry of refined taste run it}
\stick{down: Nor is there any thing so likely}
\catch{to}
\stick{to make them do it, as that of leaving}
\stick{them out of the party, or, what is full as}
\stick{offensive, of bestowing your attention}
\stick{upon the rest of your guests in so parti-}
\stick{cular a way, as if there was no such}
\stick{thing as a critick (\,by occupation\,) at}
table.

\tsh I guard against both; for, in\break
\stick{the first place, I have left half a dozen}
\stick{places purposely open for them;\tsk and,}
\stick{in the next place, I pay them all court,\tsk}
\stick{Gentlemen, I kiss your hands,\tsk I pro-}
\stick{test no company could give me half the}
\stick{pleasure,\tsk by my soul I am glad to see}
\stick{you,\tsk I beg only you will make no}
\stick{strangers of yourselves, but sit down}
\stick{without any ceremony, and fall on hear-}
tily.
\natch[12pt]{I}

I said I had left six places, and I was\break
\stick{upon the point of carrying my complai-}
\stick{sance so far, as to have left a seventh}
\stick{open for them,\tsk and in this very spot I}
\stick{stand on;\tsk but being told by a critick,}
\stick{(tho’ not by occupation,\tsk but by nature)}
\stick{that I had acquitted myself well enough,}
\stick{I shall fill it up directly, hoping, in the}
\stick{mean time, that I shall be able to make}
a great deal of more room next year.

\tsh How, in the name of wonder!\break
\stick{could your uncle \toby, who, it seems,}
\stick{was a military man, and whom you have}
\stick{represented as no fool,\tsk be at the same}
\stick{time such a confused, pudding-headed,}
muddle-headed fellow, as\tsk Go look.

So, Sir Critick, I could have replied;\break
\stick{but I scorn it.\tsh ’Tis language un-}
\stick{urbane,\tsk and only befitting the man}
\catch{who}
\stick{who cannot give clear and satisfactory}
\stick{accounts of things, or dive deep enough}
\stick{into the first causes of human ignorance}
\stick{and confusion. It is moreover the re-}
\stick{ply valiant,\tsk\kern1.414pt and therefore I reject it;}
\stick{for tho’ it might have suited my uncle}
\stick{\toby's character as a soldier excellently}
\stick{well,\tsk and had he not accustomed him-}
\stick{self, in such attacks, to whistle the \i{Lilla}-}
\stick{\i{bullero},\tsk as he wanted no courage, ’tis}
\stick{the very answer he would have given;}
\stick{yet it would by no means have done for}
\stick{me. You see as plain as can be, that I}
\stick{write as a man of erudition;\tsk that even}
\stick{my similies, my allusions, my illustra-}
\stick{tions, my metaphors, are erudite,\tsk and}
\stick{that I must sustain my character properly,}
\stick{and contrast it properly too,\tsk else what}
\stick{would become of me?  Why, Sir, I}
\stick{should be undone;\tsk at this very mo-}
\stick{ment that I am going here to fill up one}
\catch{place}
\stick{place against a critick,\tsk I should have}
made an opening for a couple.

\tsh Therefore I answer thus:

Pray, Sir, in all the reading which you\break
\stick{have ever read, did you ever read such a}
\stick{book as \i{Locke}’s Essay upon the Human}
\stick{Understanding?\tsh Don’t answer me}
\stick{rashly,\tsk because many, I know, quote the}
\stick{book, who have not read it,\tsk and many}
\stick{have read it who understand it not:\tsk If}
\stick{either of these is your case, as I write to}
\stick{instruct, I will tell you in three words}
\stick{what the book is.\tsk It is a history.\tsk A}
\stck{history! of who? what? where? when?}
\stick{Don’t hurry yourself.\tsh It is a history-}
\stick{book, Sir, (which may possibly recom-}
\stick{mend it to the world) of what passes in a}
\stick{man’s own mind; and if you will say so}
\stick{much of the book, and no more, believe}
\catch{me,}
\stick{me, you will cut no contemptible figure}
in a metaphysic circle.

But this by the way.

Now if you will venture to go along\break
\stick{with me, and look down into the bottom}
\stick{of this matter, it will be found that the}
\stick{cause of obscurity and confusion, in the}
mind of a man, is threefold.

Dull organs, dear Sir, in the first place.\break
\stck{Secondly, slight and transient impressions}
\stick{made by objects when the said organs are}
\stick{not dull. And, thirdly, a memory like}
\stick{unto a sieve, not able to retain what it has}
\stick{received.\tsk Call down \i{Dolly} your cham-}
\stick{ber-maid, and I will give you my cap}
\stick{and bell along with it, if I make not this}
\stick{matter so plain that \i{Dolly} herself shall}
\stick{understand it as well as \i{Malbranch.}\tsh}
\catch{When}
\stick{When \i{Dolly} has indited her epistle to}
\stick{\i{Robin}, and has thrust her arm into the}
\stick{bottom of her pocket hanging by her}
\stick{right-side;\tsk take that opportunity to}
\stick{recollect that the organs and faculties of}
\stick{perception, can, by nothing in this world,}
\stick{be so aptly typified and explained as by}
\stick{that one thing which \i{Dolly}’s hand is in}
\stick{search of.\tsk Your organs are not so dull}
\stick{that I should inform you,\tsk ’tis an inch,}
Sir, of red seal-wax.

\tight{When this is melted and dropped upon}\break
\stick{the letter, if \i{Dolly} fumbles too long for}
\stick{her thimble, till the wax is over harden’d,}
\stick{it will not receive the mark of her thim-}
\stick{ble from the usual impulse which was}
\stick{wont to imprint it. Very well: If \i{Dolly}’s}
\stick{wax, for want of better, is bees-wax, or}
\stick{of a temper too soft,\tsk tho’ it may re-}
\stick{ceive,\tsk it will not hold the impression,}
\catch{how}
\stick{how hard soever \i{Dolly} thrusts against it;}
\stick{and last of all, supposing the wax good,}
\stick{and eke the thimble, but applied there-}
\stick{to in careless haste, as her Mistress rings}
\stick{the bell;\tsk in any one of these three}
\stick{cases, the print, left by the thimble, will}
\stick{be as unlike the prototype as a brass-}
jack.

Now you must understand that not\break
\stick{one of these was the true cause of the}
\stick{confusion in my uncle \toby's discourse;}
\stick{and it is for that very reason I enlarge}
\stick{upon them so long, after the manner of}
\stick{great physiologists\tsk to shew the world,}
what it did \i{not} arise from.

What it did arise from, I have hinted\break
\stick{above, and a fertile source of obscurity}
\stick{it is,\tsk and ever will be,\tsk and that is the}
\stick{unsteady uses of words which have per-}
\catch{plexed}
\stick{plexed the clearest and most exalted un-}
derstandings.

It is ten to one (at \i{Arthur}’s) whether\break
\stick{you have ever read the literary histories}
\stick{of past ages;\tsk if you have,\tsk what ter-}
\stick{rible battles, ’yclept logomachies, have}
\stick{they occasioned and perpetuated with so}
\stick{much gall and ink-shed,\tsk that a good}
\stick{natured man cannot read the accounts}
of them without tears in his eyes.

\tight{Gentle critick! when thou hast weigh’d}\break
\stck{all this, and consider’d within thyself how}
\stick{much of thy own knowledge, discourse,}
\stick{and conversation has been pestered and}
\stick{disordered, at one time or other, by this,}
\stick{and this only:\tsh What a pudder and}
\stick{racket in \s{Councils} about \g{οὐσια} and \g{ὑπο}-}
\stck{\g{στασις}; and in the \s{Schools} of the learned}
\stick{about power and about spirit;\tsk about}
\catch{essences,}
\stick{essences, and about quintessences;\tsh}
\stick{about substances, and about space.\tsh}
\stick{What confusion in greater \s{Theatres}}
\stick{from words of little meaning, and as in-}
\stick{determinate a sense;\tsk when thou consi-}
\stick{ders this, thou wilt not wonder at my}
\stick{uncle \toby's perplexities,\tsk thou wilt drop}
\stick{a tear of pity upon his scarp and his coun-}
\stick{terscarp;\tsk his glacis and his covered-}
\stick{way;\tsk his ravelin and his half-moon:}
\stick{’Twas not by ideas,\tsh by heaven!}
his life was put in jeopardy by words.

\chap{III}{24pt}{8pt}

\dropcap{W}{hen} my uncle \toby got his map\break
of \i{Namur} to his mind, he began\break
\stick{immediately to apply himself, and with}
\stick{the utmost diligence, to the study of it;}
\stick{for nothing being of more importance}
\catch{to}
\stick{to him than his recovery, and his reco-}
\stick{very depending, as you have read, up-}
\stick{on the passions and affections of his}
\stick{mind, it behoved him to take the nicest}
\stick{care to make himself so far master of his}
\stick{subject, as to be able to talk upon it}
without emotion.

In a fortnight’s close and painful ap-
\stick{plication, which, by the bye, did my}
\stick{uncle \toby's wound, upon his groin, no}
\stick{good,\tsk he was enabled, by the help of}
\stick{some marginal documents at the feet of}
\stck{the elephant, together with \i{Gobesius}’s mi-}
\stick{litary architecture and pyroballogy, tran-}
\stick{slated from the \i{Flemish}, to form his dis-}
\stick{course with passable perspicuity; and}
\stick{before he was two full months gone,\tsk}
\stick{he was right eloquent upon it, and could}
\stick{make not only the attack of the advan-}
\catch{ced}
\stick{ced counterscarp with great order;\tsh}
\stick{but having, by that time, gone much}
\stick{deeper into the art, than what his first}
\stick{motive made necessary,\tsk my uncle \i{Toby}}
\stick{was able to cross the \i{Maes} and \i{Sambre};}
\stick{make diversions as far as \i{Vauban}’s line,}
\stick{the abbey of \i{Salsines}, \etc and give his}
\stick{visiters as distinct a history of each of}
\stick{their attacks, as of that of the gate of}
\stick{\i{St.\@ Nicolas}, where he had the honour to}
receive his wound.

But the desire of knowledge, like the
\stck{thirst of riches, increases ever with the ac-}
\stick{quisition of it. The more my uncle \i{Toby}}
\stick{pored over his map, the more he took a}
\stick{liking to it;\tsk by the same process and}
\stick{electrical assimilation, as I told you, thro’}
\stick{which I ween the souls of connoisseurs}
\stick{themselves, by long friction and incum-}
\stick{bition, have the happiness, at length, to}
\catch{get}
\stick{get all be-virtu’d,\tsk be-pictur’d,\tsk be-}
butterflied, and be-fiddled.

The more my uncle \toby drank of this\break
\stck{sweet fountain of science, the greater was}
\stick{the heat and impatience of his thirst,}
\stick{so that, before the first year of his con-}
\stick{finement had well gone round, there was}
\stick{scarce a fortified town in \i{Italy} or \i{Flanders},}
\stick{of which, by one means or other, he had}
\stick{not procured a plan, reading over as he}
\stick{got them, and carefully collating there-}
\stick{with the histories of their sieges, their de-}
\stick{molitions, their improvements, and new}
\stick{works, all which he would read with that}
\stick{intense application and delight, that he}
\stick{would forget himself, his wound, his con-}
finement, his dinner.

In the second year my uncle \toby pur-\break
\stick{chased \i{Ramelli} and \i{Cataneo}, translated}
\catch{from}
\stick{from the \i{Italian};\tsk likewise \i{Stevinus},}
\stick{\i{Marolis}, the Chevalier \i{de Ville}, \i{Lorini},}
\stick{\i{Cochorn}, \i{Sheeter}, the Count \i{de Pagan},}
\stick{the Marshal \i{Vauban}, Mons.\@ \i{Blondel}, with}
\stick{almost as many more books of military}
\stick{architecture, as Don \i{Quixote} was found}
\stick{to have of chivalry, when the curate and}
barber invaded his library.

Towards the beginning of the third\break
\stick{year, which was in \i{August}, ninety-nine,}
\stick{my uncle \toby found it necessary to un-}
\stick{derstand a little of projectiles:\tsk And ha-}
\stick{ving judged it best to draw his know-}
\stick{ledge from the fountain-head, he began}
\stick{with \i{N.\@ Tartaglia}, who it seems was the}
\stick{first man who detected the imposition of}
\stick{a canon-ball’s doing all that mischief}
\stick{under the notion of a right line.\tsk This}
\stick{\i{N.\@ Tartaglia} proved to my uncle \toby to}
be an impossible thing.

\natch[2pt]{\tsk Endless}

\tsh Endless is the Search of\break
Truth!

No sooner was my uncle \toby satisfied\break
\stick{which road the cannon-ball did not go,}
\stck{but he was insensibly led on, and resolved}
\stick{in his mind to enquire and find out}
\stick{which road the ball did go: For which}
\stick{purpose he was obliged to set off afresh}
\stick{with old \i{Maltus}, and studied him devout-}
\stick{ly.\tsk He proceeded next to \i{Galileo} and}
\stick{\i{Torricellius}, wherein, by certain geome-}
\stick{trical rules, infallibly laid down, he found}
\stick{the precise path to be a \s{Parabola},\tsk or}
\stick{else an \s{Hyperbola},\tsk and that the pa-}
\stick{rameter, or \i{latus rectum}, of the conic sec-}
\stick{tion of the said path, was to the quantity}
\stick{and amplitude in a direct \i{ratio}, as the}
\stick{whole line to the sine of double the angle}
\stick{of incidence, form’d by the breech upon}
\stick{an horizontal plane;\tsk and that the semi-}
\catch{parameter,}
\stick{parameter,\tsh stop! my dear uncle}
\stick{\i{Toby}\tsh stop!\tsh go not one foot further}
\stick{into this thorny and bewilder’d track,\tsk}
\stick{intricate are the steps!  intricate are the}
\stick{mases of this labyrinth! intricate are the}
\stck{troubles which the pursuit of this bewitch-}
\stick{ing phantom, \s{Knowledge}, will bring}
\stick{upon thee.\tsk O my uncle! fly\tsk fly\tsk}
\stick{fly from it as from a serpent.\tsk Is it fit,}
\stick{good-natur’d man!\@ thou should’st sit up,}
\stick{with the wound upon thy groin, whole}
\stick{nights baking thy blood with hectic}
\stck{watchings?\tsh Alas! ’twill exasperate thy}
\stick{symptoms,\tsk check thy perspirations,\tsk}
\stick{evaporate thy spirits,\tsk waste thy animal}
\stick{strength,\tsk dry up thy radical moisture,}
\stick{\tsk bring thee into a costive habit of body,}
\stick{impair thy health,\tsk and hasten all the}
\stick{infirmities of thy old age.\tsk O my uncle!}
my uncle \i{Toby!}

\natch[14pt]{\lsss{CHAP}.}

\chap{IV}{24pt}{8pt}

\initial[nindent=0pt,loversize=0.1,lraise=.01]{I}{Would} not give a groat for that man’s\break
knowledge in pen-craft, who does not\break
\stick{understand this,\tsk That the best plain}
\stick{narrative in the world, tack’d very close}
\stick{to the last spirited apostrophe to my un-}
\stick{cle \toby,\tsh would have felt both cold and}
\stick{vapid upon the reader’s palate;\tsk there-}
\stick{fore I forthwith put an end to the chap-}
\stick{ter,\tsk though I was in the middle of my}
story.

\tsh Writers of my stamp have\break
\stick{one principle in common with painters.\tsk}
\stick{Where an exact copying makes our}
\stick{pictures less striking, we choose the less}
\stick{evil; deeming it even more pardonable}
\stick{to trespass against truth, than beauty.\tsk}
\stick{This is to be understood \i{cum grano salis};}
\catch{but}
\stick{but be it as it will,\tsk as the parallel is}
\stick{made more for the sake of letting the}
\stick{apostrophe cool, than any thing else,\tsk}
\stick{’tis not very material whether upon any}
\stick{other score the reader approves of it or}
not.

In the latter end of the third year, my\break
\stick{uncle \toby perceiving that the parameter}
\stick{and semi-parameter of the conic section,}
\stick{angered his wound, he left off the study}
\stick{of projectiles in a kind of a huff, and}
\stick{betook himself to the practical part of}
\stick{fortification only; the pleasure of which,}
\stick{like a spring held back, returned upon}
him with redoubled force.

It was in this year that my uncle be-\break
\stick{gan to break in upon the daily regularity}
\stick{of a clean shirt,\tsh to dismiss his barber}
\stick{unshaven,\tsh and to allow his surgeon}
\catch{scarce}
\stick{scarce time sufficient to dress his wound,}
\stick{concerning himself so little about it, as}
\stick{not to ask him once in seven times dres-}
\stick{sing how it went on: When, lo!\tsk all of}
\stick{a sudden, for the change was quick as}
\stick{lightening, he began to sigh heavily for}
\stick{his recovery,\tsk complain’d to my father,}
\stick{grew impatient with the surgeon;\tsk and}
\stick{one morning as he heard his foot com-}
\stick{ing up stairs, he shut up his books, and}
\stick{thrust aside his instruments, in order to}
\stick{expostulate with him upon the protrac-}
\stick{tion of his cure, which, he told him,}
\stick{might surely have been accomplished at}
\stick{least by that time:\tsk He dwelt long upon}
\stick{the miseries he had undergone, and the}
\stick{sorrows of his four years melancholy}
\stick{imprisonment;\tsk adding, that had it not}
\stick{been for the kind looks, and fraternal}
\stick{chearings of the best of brothers,\tsk he}
\stick{had long since sunk under his misfor-}
\catch{tunes.\tsk}
\stick{tunes.\tsk My father was by: My uncle}
\stick{\toby's eloquence brought tears into his}
\stick{eyes;\tsk ’twas unexpected.\tsk My uncle}
\stick{\toby, by nature was not eloquent;\tsh}
\stick{it had the greater effect.\tsk The surgeon}
\stick{was confounded;\tsk not that there wanted}
\stick{grounds for such, or greater, marks of}
\stick{impatience,\tsk but ’twas unexpected too;}
\stick{in the four years he had attended him,}
\stick{he had never seen any thing like it in}
\stick{my uncle \toby's carriage;\tsk he had never}
\stick{once dropp’d one fretful or discontented}
\stick{word;\tsk he had been all patience,\tsk all}
submission.

\tsk We lose the right of complaining\break
\stick{sometimes by forbearing it;\tsk but we}
\stick{oftner treble the force:\tsk The surgeon}
\stick{was astonished;\tsk but much more so,}
\stick{when he heard my uncle \toby go on, and}
\stick{peremptorily insist upon his healing up}
\catch{the}
\stick{the wound directly,\tsh or sending for}
\stick{Monsieur \i{Ronjat}, the King’s Serjeant-}
Surgeon, to do it for him.

The desire of life and health is im-\break
\stick{planted in man’s nature;\tsk the love of}
\stck{liberty and enlargement is a sister-passion}
\stick{to it: These my uncle \toby had in com-}
\stick{mon with his species;\tsk and either of}
\stick{them had been sufficient to account for}
\stick{his earnest desire to get well and out of}
\stick{doors;\tsk but I have told you before that}
\stick{nothing wrought with our family after}
\stick{the common way;\tsh and from the time}
\stick{and manner in which this eager desire}
\stick{shew’d itself in the present case, the pe-}
\stick{netrating reader will suspect there was}
\stick{some other cause or crotchet for it in my}
\stick{uncle \toby's head:\tsk There was so, and}
\stick{’tis the subject of the next chapter to set}
\stick{forth what that cause and crotchet was.}
\catch{I}
\stick{I own, when that’s done, ’twill be time}
\stick{to return back to the parlour fire-side,}
\stick{where we left my uncle \toby in the mid-}
dle of his sentence.

\chap{V}{24pt}{10pt}

\dropcap{W}{hen} a man gives himself up to\break
the government of a ruling pas-\break
\stick{sion,\tsfill or, in other words, when his}
\stick{\s{Hobby-Horse} grows head-strong,\tsk}
farewell cool reason and fair discretion!

My uncle \toby's wound was near well,\break
\stick{and as soon as the surgeon recovered his}
\stick{surprize, and could get leave to say as}
\stick{much\tsk he told him, ’twas just beginning}
\stick{to incarnate; and that if no fresh exfolia-}
\stick{tion happen’d, which there was no signs}
\stick{of,\tsk it would be dried up in five or six}
\catch{weeks.}
\stick{weeks. The sound of as many olym-}
\stick{piads twelve hours before, would have}
\stick{convey’d an idea of shorter duration to}
\stick{my uncle \toby's mind.\tsh The succession}
\stick{of his ideas was now rapid,\tsk he broil’d}
\stick{with impatience to put his design in}
\stick{execution;\tsk and so, without consulting}
\stick{further with any soul living,\tsk which,}
\stick{by the bye, I think is right, when you}
\stick{are predetermined to take no one soul’s}
\stick{advice,\tsk he privately ordered \trim, his}
\stick{man, to pack up a bundle of lint and}
\stick{dressings, and hire a chariot and four to}
\stick{be at the door exactly by twelve o’clock}
\stick{that day, when he knew my father would}
\stick{be upon ’Change.\tsk So leaving a bank-}
\stick{note upon the table for the surgeon’s}
\stick{care of him, and a letter of tender thanks}
\stick{for his brother’s,\tsk he pack’d up his}
\stick{maps, his books of fortification, his in-}
\stick{struments, \etc\tsk and, by the help of a}
\catch{crutch,\sic}
\stick{crutch on one side, and \trim on the}
\stick{other,\tsh my uncle \toby embark’d for}
\i{Shandy-Hall}.

The reason, or rather the rise, of this\break
sudden demigration, was as follows:

The table in my uncle \toby's room,\break
\stick{and at which, the night before this}
\stick{change happened, he was sitting with his}
\stick{maps, \etc about him,\tsk being somewhat}
\stick{of the smallest, for that infinity of great}
\stick{and small instruments of knowledge}
\stick{which usually lay crouded upon it;\tsk he}
\stick{had the accident, in reaching over for}
\stick{his tobacco-box, to throw down his com-}
\stick{passes, and in stooping to take the com-}
\stick{passes up, with his sleeve he threw down}
\stick{his case of instruments and snuffers;\tsk}
\stick{and as the dice took a run against him,}
\stick{in his endeavouring to catch the snuffers}
\catch{in}
\stick{in falling,\tsh he thrust Monsieur \i{Blondel}}
\stick{off the table and Count \i{de Pagan} o’\,top}
of him.

’Twas to no purpose for a man, lame\break
\stick{as my uncle \toby was, to think of redres-}
\stick{sing all these evils by himself,\tsk he rung}
\stick{his bell for his man \trim;\tsk \i{Trim!} quoth}
\stick{my uncle \toby, pri’thee see what confu-}
\stick{sion I have here been making.\tsk I must}
\stick{have some better contrivance, \trim.\tsk}
\stick{Can’st not thou take my rule and mea-}
\stick{sure the length and breadth of this table,}
\stick{and then go and bespeak me one as big}
\stick{again?\tsk Yes, an’ please your Honour,}
\stick{replied \trim, making a bow;\tsk but I}
\stick{hope your Honour will be soon well}
\stick{enough to get down to your country seat,}
\stick{where,\tsk as your Honour takes so much}
\stick{pleasure in fortification, we could ma-}
nage this matter to a T.
\natch{I}

I must here inform you, that this ser-\break
\stick{vant of my uncle \toby's, who went by}
\stick{the name of \trim, had been a Corporal}
\stick{in my uncle’s own company,\tsh his real}
\stick{name was \i{James Butler},\tsk but having}
\stick{got the nick-name of \trim in the regi-}
\stck{ment, my uncle \toby, unless when he hap-}
\stick{pened to be very angry with him, would}
never call him by any other name.

The poor fellow had been disabled\break
\stick{for the service, by a wound on his left}
\stick{knee by a musket-bullet, at the battle of}
\stick{\i{Landen}, which was two years before the}
\stick{affair of \i{Namur};\tsk and as the fellow was}
\stick{well beloved in the regiment, and a}
\stick{handy fellow into the bargain, my uncle}
\stick{\toby took him for his servant, and of ex-}
\stick{cellent use was he, attending my uncle}
\stick{\toby in the camp and in his quarters as}
\stick{valet, groom, barber, cook, sempster,}
\catch{and}
\stick{and nurse; and indeed, from first to last,}
\stick{waited upon him and served him with}
great fidelity and affection.

My uncle \toby loved the man in re-\break
\stick{turn, and what attached him more to}
\stick{him still, was the similitude of their}
\stick{knowledge:\tsk For Corporal \trim, (for}
\stick{so, for the future, I shall call him) by}
\stick{four years occasional attention to his}
\stick{Master’s discourse upon fortified towns,}
\stick{and the advantage of prying and peeping}
\stick{continually into his Master’s plans, \etc}
\stick{exclusive and besides what he gained}
\stick{\s{Hobby-Horsically}, as a body-servant,}
\stick{\i{Non Hobby-Horsical per se};\tsk had be-}
\stick{come no mean proficient in the science;}
\stick{and was thought, by the cook and cham-}
\stick{ber-maid, to know as much of the na-}
\stick{ture of strong-holds as my uncle \toby}
himself.

\natch{I}

I have but one more stroke to give to
\stick{finish Corporal \trim's character,\tsk and it}
\stick{is the only dark line in it.\tsk The fellow}
\stick{lov’d to advise,\tsk or rather to hear him-}
\stick{self talk; his carriage, however, was so}
\stick{perfectly respectful, ’twas easy to keep}
\stick{him silent when you had him so; but set}
\stick{his tongue a-going,\tsk you had no hold}
\stick{of him;\tsk he was voluble;\tsk the eternal}
\stick{interlardings of \i{your Honour}, with the}
\stck{respectfulness of Corporal \trim's manner,}
\stick{interceding so strong in behalf of his}
\stick{elocution,\tsk that tho’ you might have}
\stick{been incommoded,\tsk you could not well}
\stick{be angry. My uncle \toby was seldom}
\stick{either the one or the other with him,\tsk}
\stick{or, at least, this fault, in \trim, broke no}
\stick{squares with ’em. My uncle \toby, as}
\stick{I said, loved the man;\tsk and besides, as}
\stick{he ever looked upon a faithful servant,\tsk}
\stick{but as a humble friend,\tsk he could not}
\catch{bear}
\stick{bear to stop his mouth.\tsh Such was Cor-}
poral \trim.

If I durst presume, continued \trim,\break
\stick{to give your Honour my advice, and}
\stick{speak my opinion in this matter.\tsk Thou}
\stick{art welcome, \trim, quoth my uncle}
\stck{\toby,\tsk speak,\tsh speak what thou thinkest}
\stick{upon the subject, man, without fear.}
\stick{Why then, replied \trim, (not hanging}
\stick{his ears and scratching his head like a}
\stick{country lout, but) stroking his hair back}
\stick{from his forehead, and standing erect as}
\stick{before his division.\tsh I think, quoth}
\stick{\trim, advancing his left, which was his}
\stick{lame leg, a little forwards,\tsk and point-}
\stick{ing with his right hand open towards a}
\stck{map of \i{Dunkirk}, which was pinn’d against}
\stick{the hangings,\tsk I think, quoth Corporal}
\stick{\trim, with humble submission to your}
\stick{Honour’s better judgment,\tsk that these}
\catch{ravelins,}
\stick{ravelins, bastions, curtins, and horn-}
\stick{works, make but a poor, contemptible,}
\stick{fiddle faddle piece of work of it here}
\stick{upon paper, compared to what your Ho-}
\stick{nour and I could make of it, were we in}
\stick{the country by ourselves, and had but a}
\stick{rood, or a rood and a half of ground to}
\stick{do what we pleased with: As summer is}
\stick{coming on, continued \trim, your Ho-}
\stick{nour might sit out of doors, and give me}
\stick{the nography\tsh (call it ichnography,}
\stick{quoth my uncle)\tsk of the town or cita-}
\stick{del, your Honour was pleased to sit down}
\stick{before,\tsk and I will be shot by your Ho-}
\stick{nour upon the glacis of it, if I did not}
\stick{fortify it to your Honour’s mind.\tsk I dare}
\stick{say thou would’st, \trim, quoth my uncle.}
\stick{\tsk For if your Honour, continued the}
\stick{Corporal, could but mark me the poly-}
\stick{gon, with its exact lines and angles.\tsh}
\stick{That I could do very well, quoth my}
\catch{uncle.}
\stick{uncle.\tsk I would begin with the fossé,}
\stick{and if your Honour could tell me the}
\stick{proper depth and breadth,\tsk I can to a}
\stick{hair’s breadth, \trim, replied my uncle,\tsk}
\stick{I would throw out the earth upon this}
\stick{hand towards the town for the scarp,\tsk}
\stick{and on that hand towards the campaign}
\stick{for the counterscarp.\tsk Very right, \trim;}
\stick{quoth my uncle \toby.\tsk And when I had}
\stick{sloped them to your mind,\tsk an’ please}
\stick{your Honour, I would face the glacis, as}
\stick{the finest fortifications are done in \i{Flan-}}
\stick{\i{ders}, with sods,\tsk and as your Honour}
\stick{knows they should be,\tsk and I would}
\stick{make the walls and parapets with sods}
\stick{too;\tsk The best engineers call them ga-}
\stick{zons, \trim, said my uncle \toby;\tsk Whe-}
\stick{ther they are gazons or sods, is not much}
\stick{matter, replied \trim, your Honour knows}
\stick{they are ten times beyond a facing either}
\stick{of brick or stone;\tsh I know they are,}
\catch{\trim,}
\stck{\trim, in some respects,\tsh quoth my uncle}
\stick{\toby, nodding his head;\tsk for a cannon-}
\stick{ball enters into the gazon right onwards,}
\stick{without bringing any rubbish down with}
\stick{it, which might fill the fossé, (as was the}
\stick{case at \i{St.\@ Nicolas}’s \i{Gate)}\sic and facilitate}
the passage over it.

Your Honour understands these mat-\break
\stick{ters, replied Corporal \trim, better than}
\stick{any officer in his Majesty’s service;\tsk}
\stick{but would your Honour please to let the}
\stick{bespeaking of the table alone, and let us}
\stick{but go into the country, I would work}
\stick{under your Honour’s directions like a}
\stick{horse, and make fortifications for you}
\stick{something like a tansy, with all their}
\stick{batteries, saps, ditches, and palisadoes,}
\stick{that it should be worth all the world’s}
riding twenty miles to go and see it.

\natch[12pt]{My}

My uncle \toby blushed as red as scar-
\stick{let as \trim went on;\tsk but it was not a}
\stick{blush of guilt,\tsk of modesty,\tsk or of an-}
\stick{ger;\tsk it was a blush of joy;\tsk he was}
\stick{fired with Corporal \trim's project and}
\stick{description.\tsk \i{Trim!} said my uncle \toby,}
\stick{thou hast said enough.\tsk We might be-}
\stick{gin the campaign, continued \trim, on}
\stick{the very day that his Majesty and the}
\stick{Allies take the field, and demolish them}
\stick{town by town as fast as\tsh \trim, quoth}
\stick{my uncle \toby, say no more.\tsk Your Ho-}
\stick{nour, continued \trim, might sit in your}
\stick{arm-chair (\,pointing to it\,) this fine}
\stick{weather, giving me your orders, and I}
\stick{would\tsh Say no more, \trim, quoth my}
\stick{uncle \toby.\tsh Besides, your Honour}
\stick{would get not only pleasure and good}
\stick{pastime,\tsk but good air, and good exer-}
\stick{cise, and good health,\tsk and your Ho-}
\stick{nour’s wound would be well in a month.}
\catch{Thou}
\stick{Thou hast said enough, \trim,\tsk quoth}
\stick{my uncle \toby (putting his hand into}
\stick{his breeches-pocket)\tsh I like thy project}
\stick{mightily;\tsk And if your Honour pleases,}
\stick{I’ll, this moment, go and buy a pioneer’s}
\stick{spade to take down with us, and I’ll be-}
\stck{speak a shovel and a pick-ax, and a couple}
\stick{of\tsh Say no more, \trim, quoth my}
\stick{uncle \toby, leaping up upon one leg,}
\stick{quite overcome with rapture,\tsk and}
\stick{thrusting a guinea into \trim's hand,\tsk}
\stick{\trim, said my uncle \toby, say no more;\tsk}
\stick{but go down, \trim, this moment, my}
\stick{lad, and bring up my supper this in-}
stant.

\trim ran down and brought up his\break
\stick{Master’s supper,\tsk to no purpose:\tsh}
\stick{\trim's plan of operation ran so in my}
\stick{uncle \toby's head, he could not taste}
\stick{it.\tsk \trim, quoth my uncle \toby, get me}
\catch{to}
\stick{to-bed;\tsk ’twas all one.\tsk Corporal \trim's}
\stick{description had fired his imagination,\tsk}
\stick{my uncle \toby could not shut his eyes.\tsk}
\stick{The more he consider’d it, the more be-}
\stick{witching the scene appeared to him;\tsk}
\stick{so that, two full hours before day-light,}
\stick{he had come to a final determination,}
\stick{and had concerted the whole plan of his}
and Corporal \trim's decampment.

My uncle \toby had a little neat coun-\break
\stick{try-house of his own, in the village where}
\stick{my father’s estate lay at \i{Shandy}, which}
\stick{had been left him by an old uncle,}
\stick{with a small estate of about one hundred}
\stick{pounds a year. Behind this house, and}
\stick{contiguous to it, was a kitchen-garden}
\stick{of about half an acre;\tsk and at the bot-}
\stick{tom of the garden, and cut off from it}
\stick{by a tall yew hedge, was a bowling-}
\stick{green, containing just about as much}
\catch{ground}
\stick{ground as Corporal \trim wished for;\tsk}
\stick{so that as \trim uttered the words, \lqq A}
\stick{rood and a half of ground to do what}
\stick{they would with:”\tsh this identical}
\stick{bowling-green instantly presented itself,}
\stck{and became curiously painted, all at once,}
\stick{upon the retina of my uncle \toby's fancy;}
\stick{\tsh which was the physical cause of}
\stick{making him change colour, or at least,}
\stick{of heightening his blush to that immo-}
derate degree I spoke of.

Never did lover post down to a belov’d\break
\stick{mistress with more heat and expectation,}
\stick{than my uncle \toby did, to enjoy this}
\stick{self-same thing in private;\tsk I say in pri-}
\stick{vate;\tshh for it was sheltered from the}
\stick{house, as I told you, by a tall yew hedge,}
\stick{and was covered on the other three sides,}
\stick{from mortal sight, by rough holly and}
\stick{thickset flowering shrubs;\tsk so that the}
\catch{idea}
\stick{idea of not being seen, did not a little}
\stick{contribute to the idea of pleasure pre-}
\stick{conceived in my uncle \toby's mind.\tsk}
\stick{Vain thought! however thick it was}
\stick{planted about,\tsh or private soever it}
\stick{might seem,\tsk to think, dear uncle \toby,}
\stick{of enjoying a thing which took up a}
\stick{whole rood and a half of ground,\tsk and}
not have it known!

\tight{How my uncle \toby and Corporal \trim}\break
\stick{managed this matter,\tsk with the history}
\stick{of their campaigns, which were no way}
\stick{barren of events,\tsh may make no unin-}
\stick{teresting under-plot in the epitasis and}
\stick{working up of this drama.\tsk At present}
\stick{the scene must drop,\tsk and change for}
the parlour fire-side.

\natch[36pt]{\lsss{CHAP}.}

\chap{VI}{24pt}{6pt}

\tsh What can they be doing,\break
\stick{brother? said my father.\tsk I think, re-}
\stick{plied my uncle \toby,\tsk taking, as I told}
\stick{you, his pipe from his mouth, and strik-}
\stick{ing the ashes out of it as he began his}
\stick{sentence;\tsk I think, replied he,\tsk it would}
\stick{not be amiss, brother, if we rung the}
bell.

Pray, what’s all that racket over our\break
\stick{heads, \i{Obadiah?}\tsk quoth my father;\tsk}
\stick{my brother and I can scarce hear our-}
selves speak.

Sir, answer’d \i{Obadiah}, making a bow\break
\stick{towards his left-shoulder,\tsk my Mistress}
\stick{is taken very badly;\tsk and where’s \i{Su}-}
\stick{\i{sannah} running down the garden there,}
\catch{as}
\stick{as if they were going to ravish her?\tsh}
\stick{Sir, she is running the shortest cut into}
\stick{the town, replied \i{Obadiah}, to fetch the}
\stick{old midwife.\tsh Then saddle a horse,}
\stick{quoth my father, and do you go direct-}
\stick{ly for \drslop, the man-midwife, with}
\stick{all our services,\tsk and let him know your}
\stick{Mistress is fallen into labour,\tsk and that}
\stick{I desire he will return with you with all}
speed.

It is very strange, says my father, ad-\break
\stick{dressing himself to my uncle \toby, as}
\stick{\i{Obadiah} shut the door,\tsk as there is so}
\stick{expert an operator as \drslop so near\tsk}
\stick{that my wife should persist to the very}
\stick{last in this obstinate humour of hers, in}
\stick{trusting the life of my child, who has had}
\stick{one misfortune already, to the ignorance}
\stick{of an old woman;\tsh and not only the}
\stick{life of my child, brother,\tsk but her own}
\catch{life,}
\stick{life, and with it the lives of all the chil-}
\stick{dren I might, peradventure, have begot}
out of her hereafter.

Mayhap, brother, replied my uncle\break
\stick{\toby, my sister does it to save the ex-}
\stick{pence:\tsk A pudding’s end,\tsk replied my}
\stick{father,\tsh the doctor must be paid the}
\stick{same for inaction as action,\tsk if not bet-}
ter,\tsk to keep him in temper.

\noindent
\stick{\indent\tsh Then it can be out of no-}
\stick{thing in the whole world, quoth my}
\stick{uncle \toby, in the simplicity of his heart,}
\stick{\tsk but \s{Modesty}:\tsk My sister, I dare}
\stick{say, added he, does not care to let a}
\stick{man come so near her \wastiv. I will}
\stick{not say whether my uncle \toby had com-}
\stick{pleted the sentence or not;\tsh ’tis for}
\stick{his advantage to suppose he had,\tsh}
\stick{as, I think, he could have added no}
\catch{\s{One}}
\stick{\s{One Word} which would have im-}
proved it.

If, on the contrary, my uncle \toby had
\stick{not fully arrived at the period’s end,\tsk}
\stick{then the world stands indebted to the}
\stick{sudden snapping of my father’s tobacco-}
\stick{pipe, for one of the neatest examples of}
\stick{that ornamental figure in oratory, which}
\stick{Rhetoricians stile the \i{Aposiopesis.}\tsk Just}
\stick{heaven! how does the \i{Poco piu} and the}
\stick{\i{Poco meno} of the \i{Italian} artists;\tsk the in-}
\stick{sensible \s{more} or \s{less}, determine the}
\stick{precise line of beauty in the sentence, as}
\stick{well as in the statue! How do the slight}
\stick{touches of the chisel, the pencil, the pen,}
\stick{the fiddle-stick, \i{et cætera},\tsk give the true}
\stick{swell, which gives the true pleasure!\tsk O}
\stick{my countrymen!\tsk be nice;\tsk be cau-}
\stick{tious of your language;\tsh and never,}
\stick{O! never let it be forgotten upon what}
\catch{small}
\stick{small particles your eloquence and your}
fame depend.

\tsh \lqq My sister, mayhap,” quoth my
\stick{uncle \toby, \lqq does not choose to let a}
\stick{\lqq man come so near her \wastiv” Make}
\stick{this dash,\tsk ’tis an Aposiopesis:\tsk Take}
\stick{the dash away, and write \i{Backside},\tsh}
\stick{’tis Bawdy.\tsk Scratch Backside out, and}
\stick{put \i{Cover’d-way} in,\tsk ’tis a Metaphor;\tsk}
\stick{and, I dare say, as fortification ran so}
\stick{much in my uncle \toby's head, that if he}
\stick{had been left to have added one word to}
the sentence,\tsk that word was it.

But whether that was the case or not
\stick{the case;\tsk or whether the snapping of}
\stick{my father’s tobacco-pipe, so critically,}
\stick{happened thro’ accident or anger,\tsk will}
be seen in due time.

\natch[12pt]{\lsss{CHAP}.}

\chap{VII}{26pt}{8pt}

\dropcap{T}{ho}’ my father was a good natural\break
philosopher,\tsk yet he was some-\break
\stick{thing of a moral philosopher too; for}
\stick{which reason, when his tobacco-pipe}
\stick{snapp’d short in the middle,\tsk he had}
\stick{nothing to do,\tsk as such,\tsk but to have}
\stick{taken hold of the two pieces, and thrown}
\stick{them gently upon the back of the fire.\tsh}
\stick{He did no such thing;\tsh he threw them}
\stick{with all the violence in the world;\tsk and,}
\stick{to give the action still more emphasis,\tsk}
\stick{he started up upon both his legs to}
do it.

This look’d something like heat;\tsh\break
\stick{and the manner of his reply to what}
\stick{my uncle \toby was saying, prov’d it}
\stick{was so.\hfill}
\catch{\lqq Not\sic}
\stick{\indent\tsk \lqq Not choose,” quoth my father,}
\stick{repeating my uncle \toby's words, \lqq to let}
\stick{\lqq a man come so near her.”\tsh By hea-}
\stick{ven, brother \i{Toby!} you would try the}
\stick{patience of a \i{Job};\tsk and I think I have}
\stick{the plagues of one already, without it.}
\stick{\tsh Why?\tsk Where?\tsk Wherein?\tsh}
\stick{Wherefore?\tsh Upon what account? re-}
\stick{plied my uncle \toby, in the utmost asto-}
\stick{nishment.\tsh To think, said my father,}
\stick{of a man living to your age, brother,}
\stick{and knowing so little about women!\tsh}
\stick{I know nothing at all about them,\tsk re-}
\stick{plied my uncle \toby; and I think, con-}
\stick{tinued he, that the shock I received the}
\stick{year after the demolition of \i{Dunkirk}, in}
\stick{my affair with widow \i{Wadman};\tsk which}
\stick{shock you know I should not have re-}
\stick{ceived, but from my total ignorance of}
\stick{the sex,\tsk has given me just cause to say,}
\stick{That I neither know, nor do pretend to}
\catch{know,}
\stick{know, any thing about ’em, or their}
\stick{concerns either.\tsk Methinks, brother,}
\stick{replied my father, you might, at least,}
\stick{know so much as the right end of a}
woman from the wrong.

It is said in \i{Aristotle}’s \i{Master-Piece},\break
\stick{\lqq That when a man doth think of any}
\stick{\lqq thing which is past,\tsk he looketh down}
\stick{\lqq upon the ground;\tsh but that when he}
\stick{\lqq thinketh of something that is to come,}
\stick{\lqq he looketh up towards the heavens.”\hfill}

My uncle \toby, I suppose, thought of\break
\stick{neither,\tsk for he look’d horizontally.\tsk}
\stick{Right end,\tsk quoth my uncle \toby, mut-}
\stick{tering the two words low to himself, and}
\stick{fixing his two eyes insensibly as he mut-}
\stick{tered them, upon a small crevice, form’d}
\stick{by a bad joint in the chimney-piece.\tsk}
\stick{Right end of a woman!\tsh I declare,}
\catch{quoth}
\stick{quoth my uncle, I know no more which}
\stick{it is, than the man in the moon;\tsk and if}
\stick{I was to think, continued my uncle \toby,}
\stick{(keeping his eyes still fix’d upon the bad}
\stick{joint) this month together, I am sure I}
should not be able to find it out.

Then brother \toby, replied my father,\break
I will tell you.

Every thing in this world, continued\break
\stick{my father (filling a fresh pipe)\tsk every}
\stick{thing in this earthly world, my dear bro-}
\stick{ther \toby, has two handles.\tsk Not always,}
\stick{quoth my uncle \toby.\tsh At least, replied}
\stick{my father, every one has two hands,\tsk}
\stick{which comes to the same thing.\tsh Now,}
\stick{if a man was to sit down cooly, and con-}
\stick{sider within himself the make, the shape,}
\stick{the construction, com-at-ability, and con-}
\stick{venience of all the parts which constitute}
\catch{the}
\stick{the whole of that animal, call’d Woman,}
\stick{and compare them analogically.\tsk I never}
\stick{understood rightly the meaning of that}
\stick{word,\tsh quoth my uncle \toby.\tsfill}
\stick{\s{Analogy}, replied my father, is the cer-}
\stick{tain relation and agreement, which dif-}
\stick{ferent\tsh Here a devil of a rap at the door}
\stick{snapp’d my father’s definition (like his}
\stick{tobacco-pipe) in two,\tsk and, at the same}
\stick{time, crushed the head of as notable and}
\stick{curious a dissertation as ever was engen-}
\stick{dered in the womb of speculation;\tsk it}
\stick{was some months before my father could}
\stick{get an opportunity to be safely deliver’d}
\stick{of it:\tsk And, at this hour, it is a thing}
\stick{full as problematical as the subject of the}
\stick{dissertation itself,\tsk (considering the con-}
\stck{fusion and distresses of our domestic mis-}
\stick{adventures, which are now coming thick}
\stick{one upon the back of another) whether}
\catch{I}
\stick{I shall be able to find a place for it in}
the third volume or not.

\chap{VIII}{26pt}{8pt}

\dropcap{I}{t} is about an hour and a half’s tole-\break
rable good reading since my uncle\break
\stick{\toby rung the bell, when \i{Obadiah} was}
\stick{order’d to saddle a horse, and go for}
\stick{\drslop the man-midwife;\tsk so that no}
\stick{one can say, with reason, that I have not}
\stick{allowed \i{Obadiah} time enough, poetically}
\stck{speaking, and considering the emergency}
\stick{too, both to go and come;\tsh tho’,}
\stick{morally and truly speaking, the man}
\stick{perhaps, has scarce had time to get on}
his boots.

If the hypercritic will go upon this;\break
\stick{and is resolved after all to take a pen-}
\catch{dulum,}
\stick{dulum, and measure the true distance}
\stick{betwixt the ringing of the bell, and the}
\stick{rap at the door;\tsk and, after finding it}
\stick{to be no more than two minutes, thirteen}
\stick{seconds, and three fifths,\tsk should take}
\stick{upon him to insult over me for such a}
\stick{breach in the unity, or rather probability,}
\stick{of time;\tsk I would remind him, that the}
\stick{idea of duration and of its simple modes,}
\stick{is got merely from the train and succes-}
\stick{sion of our ideas,\tsk and is the true scho-}
\stick{lastic pendulum,\tsh and by which, as a}
\stick{scholar, I will be tried in this matter,\tsk}
\stick{abjuring and detesting the jurisdiction of}
all other pendulums whatever.

I would, therefore, desire him to con-\break
\stick{sider that it is but poor eight miles from}
\stick{\i{Shandy-Hall} to \drslop, the man mid-}
\stick{wife’s house;\tsk and that whilst \i{Obadiah}}
\stick{has been going those said miles and back,}
\catch{I}
\stick{I have brought my uncle \toby from \i{Na}-}
\stick{\i{mur},  quite across all \i{Flanders}, into \i{Eng}-}
\stick{\i{land}:\tsk That I have had him ill upon my}
\stick{hands near four years;\tsk and have since}
\stick{travelled him and Corporal \trim, in a}
\stick{chariot and four, a journey of near two}
\stick{hundred miles down into \i{Yorkshire};\tsk all}
\stick{which put together, must have prepared}
\stick{the reader’s imagination for the entrance}
\stick{of \drslop upon the stage,\tsk as much, at}
\stick{least (I hope) as a dance, a song, or a}
concerto between the acts.

If my hypercritic is intractable,\tsk al-\break
\stick{ledging, that two minutes and thirteen}
\stick{seconds are no more than two minutes}
\stick{and thirteen seconds,\tsk when I have said}
\stick{all I can about them;\tsk and that this}
\stick{plea, tho’ it might save me dramatically,}
\stick{will damn me biographically, rendering}
\stick{my book, from this very moment, a pro-}
\catch{fess’d}
\stick{fess’d \s{Romance}, which, before, was a}
\stick{book apocryphal:\tsk If I am thus pres-}
\stick{sed.\tsk I then put an end to the whole}
\stick{objection and controversy about it all at}
\stick{once,\tsk by acquainting him, that \i{Obadiah}}
\stick{had not got above threescore yards from}
\stick{the stable-yard before he met with Dr.}
\stick{\i{Slop};\tsk and indeed he gave a dirty proof}
\stick{that he had met with him,\tsk and was}
\stick{within an ace of giving a tragical one}
too.

Imagine to yourself;\tsk but this had\break
better begin a new chapter.

\chap{IX}{26pt}{8pt}

\initial[findent=1pt, nindent=6pt]{I}{Magine} to yourself a little, squat, un-\break
courtly figure of a Doctor \i{Slop}, of\break
\stick{about four feet and a half perpendicular}
\catch{height,}
\stick{height, with a breadth of back, and a}
\stick{sesquipedality of belly, which might have}
\stick{done honour to a serjeant in the horse-}
guards.

Such were the out-lines of \drslop's\break
\stick{figure, which,\tsk if you have read \i{Ho}-}
\stick{\i{garth}’s analysis of beauty, and if you have}
\stick{not, I wish you would;\tsk you must know,}
\stick{may as certainly be caracatur’d, and con-}
\stick{vey’d to the mind by three strokes as}
three hundred.

Imagine such a one,\tsk for such, I say,
\stick{were the out-lines of \drslop's figure,}
\stick{coming slowly along, foot by foot, wad-}
\stick{dling thro’ the dirt upon the vertebræ}
\stick{of a little diminutive pony,\tsk of a pretty}
\stick{colour;\tsk but of strength,\tsk alack!\tsh}
\stick{scarce able to have made an amble of it,}
\stick{under such a fardel, had the roads been}
\catch{in}
\stick{in an ambling condition.\tsk They were}
\stick{not.\tsh Imagine to yourself, \i{Obadiah}}
\stick{mounted upon a strong monster of a}
\stick{coach-horse, prick’d into a full gallop,}
\stick{and making all practicable speed the ad-}
verse way.

Pray, Sir, let me interest you a mo-\break
ment in this description.

Had \drslop beheld \i{Obadiah} a mile\break
\stick{off, posting in a narrow lane directly to-}
\stick{wards him, at that monstrous rate,\tsh}
\stick{splashing and plunging like a devil thro’}
\stick{thick and thin, as he approach’d, would}
\stck{not such a phænomenon, with such a vor-}
\stick{tex of mud and water moving along with}
\stick{it, round its axis,\tsk have been a subject}
\stick{of juster apprehension to \drslop in his}
\stick{situation, than the \i{worst} of \i{Whiston}’s co-}
\stick{mets?\tsk To say nothing of the \s{Nucleus};}
\catch{that}
\stick{that is, of \i{Obadiah} and the coach-horse.}
\stick{\tsh In my idea, the vortex alone of}
\stick{’em was enough to have involved and}
\stick{carried, if not the doctor, at least the}
\stick{doctor’s pony quite away with it. What}
\stick{then do you think must the terror and}
\stick{hydrophobia of \drslop have been, when}
\stick{you read (which you are just going to}
\stick{do) that he was advancing thus warily}
\stick{along towards \i{Shandy-Hall}, and had ap-}
\stick{proach’d to within sixty yards of it, and}
\stick{within five yards of a sudden turn, made}
\stick{by an acute angle of the garden-wall,}
\stick{\tsk and in the dirtiest part of a dirty lane,}
\stick{\tsk when \i{Obadiah} and his coach-horse}
\stick{turn’d the corner, rapid, furious,\tsk pop,}
\stick{\tsk full upon him!\tsk Nothing, I think,}
\stick{in nature, can be supposed more terrible,}
\stick{than such a Rencounter,\tsk so imprompt!}
\stick{so ill prepared to stand the shock of it as}
\stick{\drslop was!\hfill}
\catch{What}
\stick{\indent What could \drslop do?\tsh He cross’d}
\stick{himself ✝︎ \tsk Pugh!\tsk but the doctor,}
\stick{Sir, was a Papist.\tsk No matter; he had}
\stick{better have kept hold of the pummel.}
\stick{He had so;\tsk nay, as it happened, he had}
\stick{better have done nothing at all;\tsk for in}
\stick{crossing himself he let go his whip,\tsk}
\stick{and in attempting to save his whip be-}
\stick{twixt his knee and his saddle’s skirt, as}
\stick{it slipp’d, he lost his stirrup,\tsk in losing}
\stick{which, he lost his seat;\tsk and in the}
\stick{multitude of all these losses (which, by}
\stick{the bye, shews what little advantage}
\stick{there is in crossing) the unfortunate doc-}
\stick{tor lost his presence of mind. So that}
\stick{without waiting for \i{Obadiah}’s onset, he}
\stick{left his pony to its destiny, tumbling off}
\stick{it diagonally, something in the stile and}
\stick{manner of a pack of wool, and without}
\stick{any other consequence from the fall, save}
\stick{that of being left (as it would have been)}
\catch{with}
\stick{with the broadest part of him sunk about}
twelve inches deep in the mire.

\i{Obadiah} pull’d off his cap twice to Dr.\break
\stick{\i{Slop};\tsh once as he was falling,\tsk and}
\stick{then again when he saw him seated.\tsk Ill-}
\stick{timed complaisance!\tsh had not the fel-}
\stick{low better have stopp’d his horse, and got}
\stick{off and help’d him?\tsk Sir, he did all}
\stick{that his situation would allow;\tsk but the}
\stick{\s{Momentum} of the coach-horse was so}
\stick{great, that \i{Obadiah} could not do it all at}
\stick{once;\tsh he rode in a circle three times}
\stick{round \drslop, before he could fully ac-}
\stick{complish it any how;\tsk and at the last,}
\stick{when he did stop his beast, ’twas done}
\stick{with such an explosion of mud, that \i{Oba}-}
\stick{\i{diah} had better have been a league off.}
\stick{In short, never was a \drslop so beluted,}
\stick{and so transubstantiated, since that affair}
came into fashion.

\natch{\lsss{CHAP}.}

\chap{X}{28pt}{10pt}

\dropcap{W}{hen} \drslop entered the back-\break
\tight{parlour, where my father and my}\break
\stick{uncle \toby were discoursing upon the na-}
\stick{ture of women,\tsh it was hard to deter-}
\stick{mine whether \drslop's figure, or Dr.}
\stck{\i{Slop}’s presence, occasioned more surprize}
\stick{to them; for as the accident happened so}
\stick{near the house, as not to make it worth}
\stick{while for \i{Obadiah} to remount him,\tsk}
\stick{\i{Obadiah} had led him in as he was, \i{un-}}
\stick{\i{wiped}, \i{unappointed}, \i{unanealed}, with all his}
\stick{stains and blotches on him.\tsh He}
\stick{stood like \i{Hamlet}’s ghost, motionless and}
\stick{speechless, for a full minute and a half,}
\stick{at the parlour door \i{(Obadiah}\sic still hold-}
\stick{ing his hand) with all the majesty of mud.}
\stick{His hinder parts, upon which he had re-}
\stick{ceived his fall, totally besmear’d,\tsk and}
\catch{in}
\stick{in every other part of him, blotched over}
\stick{in such a manner with \i{Obadiah}’s explo-}
\stick{sion, that you would have sworn (with-}
\stick{out mental reservation) that every grain}
of it had taken effect.

Here was a fair opportunity for my\break
\stick{uncle \toby to have triumph’d over my}
\stick{father in his turn;\tsk for no mortal, who}
\stick{had beheld \drslop in that pickle, could}
\stick{have dissented from so much, at least, of}
\stick{my uncle \toby's opinion, \lqq That may-}
\stick{\lqq hap his sister might not care to let}
\stick{\lqq such a \drslop come so near her}
\stick{\lqq \wastiv\;” \quad But it was the \i{Argu-}}
\stick{\i{mentum ad hominem}; and if my uncle}
\stick{\toby was not very expert at it, you may}
\stick{think, he might not care to use it.\tsk}
\stick{No; the reason was,\tsk ’twas not his na-}
ture to insult.

\natch[21pt]{Dr.}

\drslop's presence at that time, was\break
\stick{no less problematical than the mode of}
\stick{it; tho’, it is certain, one moment’s re-}
\stick{flection in my father might have solved}
\stick{it; for he had apprized \drslop but the}
\stick{week before, that my mother was at her}
\stick{full reckoning; and as the doctor had}
\stick{heard nothing since, ’twas natural and}
\stick{very political too in him, to have taken}
\stick{a ride to \i{Shandy-Hall}, as he did, merely}
to see how matters went on.

But my father’s mind took unfortu-\break
\stick{nately a wrong turn in the investigation;}
\stick{running, like the hypercritic’s, altogether}
\stick{upon the ringing of the bell and the}
\stick{rap upon the door,\tsk measuring their di-}
\stick{stance,\tsk and keeping his mind so intent}
\stick{upon the operation, as to have power to}
\stick{think of nothing else,\tsk common-place}
\stick{infirmity of the greatest mathematicians!}
\catch{work-}
\stick{working with might and main at the de-}
\stick{monstration, and so wasting all their}
\stick{strength upon it, that they have none}
\stick{left in them to draw the corollary, to do}
good with.

The ringing of the bell, and the rap\break
\stick{upon the door, struck likewise strong}
\stick{upon the sensorium of my uncle \toby,\tsk}
\stick{but it excited a very different train of}
\stick{thoughts;\tsk the two irreconcileable pul-}
\stick{sations instantly brought \i{Stevinus}, the}
\stick{great engineer, along with them, into}
\stick{my uncle \toby's mind:\tsk What business}
\stick{\i{Stevinus} had in this affair,\tsk is the greatest}
\stick{problem of all;\tsk it shall be solved,\tsk}
but not in the next chapter.

\natch[42pt]{\lsss{CHAP}.}

\chap{XI}{28pt}{8pt}

\initial[findent=1pt,loversize=0.07]{W}{Riting}, when properly managed,\break
(as you may be sure I think mine\break
\stick{is) is but a different name for conversa-}
\stick{tion: As no one, who knows what he is}
\stick{about in good company, would venture}
\stick{to talk all;\tsk so no author, who under-}
\stick{stands the just boundaries of decorum}
\stick{and good breeding, would presume to}
\stick{think all: The truest respect which you}
\stick{can pay to the reader’s understanding, is}
\stick{to halve this matter amicably, and leave}
\stick{him something to imagine, in his turn,}
as well as yourself.

For my own part, I am eternally pay-\break
\stick{ing him compliments of this kind, and}
\stick{do all that lies in my power to keep his}
\stick{imagination as busy as my own.\hfill}
\catch{’Tis}
\stick{\indent ’Tis his turn now;\tsk I have given an}
\stick{ample description of \drslop's sad over-}
\stick{throw, and of his sad appearance in the}
\stick{back parlour;\tsh his imagination must}
now go on with it for a while.

Let the reader imagine then, that Dr.\break
\stick{\i{Slop} has told his tale;\tsk and in what}
\stick{words, and with what aggravations his}
\stick{fancy chooses:\tsk\!\tsk Let him suppose, that}
\stick{\i{Obadiah} has told his tale also, and with}
\stick{such rueful looks of affected concern,}
\stick{as he thinks will best contrast the two}
\stick{figures as they stand by each other:\tsk}
\stick{Let him imagine, that my father has}
\stick{stepp’d up stairs to see my mother:\tsk}
\stick{And, to conclude this work of imagina-}
\stick{tion,\tsk let him imagine the doctor wash’d,}
\stick{\tsh rubb’d down,\tsh condoled with,\tsk}
\stick{felicitated,\tsk got into a pair of \i{Obadiah}’s}
\stick{pumps, stepping forwards towards the}
\catch{door,}
\stick{door, upon the very point of entring}
upon action.

Truce!\tsk truce, good \drslop\/\i{!}\tsk stay\break
\stick{thy obstetric hand;\tsk return it safe into}
\stick{thy bosom to keep it warm;\tsk little dost}
\stick{thou know what obstacles;\tsk little dost}
\stick{thou think what hidden causes retard its}
\stick{operation!\tsk Hast thou, \drslop,\tsk hast}
\stick{thou been intrusted with the secret ar-}
\stick{ticles of this solemn treaty which has}
\stick{brought thee into this place?\tsk Art thou}
\stick{aware that, at this instant, a daughter of}
\stick{\i{Lucina} is put obstetrically over thy head?}
\stick{Alas! ’tis too true.\tsk Besides, great son}
\stick{of \i{Pilumnus!} what canst thou do?\tsh}
\stick{Thou hast come forth unarm’d;\tsk thou}
\stick{hast left thy \i{tire tête},\tsk thy new-invented}
\stick{\i{forceps},\tsk thy \i{crotchet},\tsk thy \i{squirt}, and}
\stick{all thy instruments of salvation and deli-}
\stick{verance behind thee.\tsh By heaven!}
\catch{at}
\stick{at this moment they are hanging up in a}
\stick{green bays bag, betwixt thy two pistols,}
\stick{at thy bed’s head!\tsk Ring;\tsk call;\tsk send}
\stick{\i{Obadiah} back upon the coach-horse to}
bring them with all speed.

\tsk Make great haste, \i{Obadiah}, quoth
\stick{my father, and I’ll give thee a crown;\tsh}
\stick{and, quoth my uncle \toby, I’ll give him}
another.

\chap{XII}{24pt}{8pt}

\dropcap{Y}{our} sudden and unexpected ar-\break
rival, quoth my uncle \toby, ad-\break
\stick{dressing himself to \drslop (all three of}
\stick{them sitting down to the fire together,}
\stick{as my uncle \toby began to speak)\tsk in-}
\stick{stantly brought the great \i{Stevinus} into}
\stick{my head, who, you must know, is a fa-}
\catch{vourite}
\stick{vourite author with me.\tsh Then,}
\stick{added my father, making use of the ar-}
\stick{gument \i{Ad Crumenam},\tshh I will lay}
\stick{twenty guineas to a single crown piece}
\stick{(which will serve to give away to \i{Oba}-}
\stick{\i{diah} when he gets back) that this same}
\stick{\i{Stevinus} was some engineer or other,\tsk}
\stick{or has wrote something or other, either}
\stick{directly or indirectly, upon the science of}
fortification.

He has so,\tsk replied my uncle \toby.\tsk\break
\stick{I knew it, said my father;\tsk tho’, for}
\stick{the soul of me, I cannot see what kind}
\stick{of connection there can be betwixt Dr.}
\stick{\i{Slop}'s sudden coming, and a discourse}
\stick{upon fortification;\tsk yet I fear’d it.\tsk}
\stick{Talk of what we will, brother,\tsk or let}
\stick{the occasion be never so foreign or unfit}
\stick{for the subject,\tsk you are sure to bring}
\stick{it in: I would not, brother \toby, con-}
\catch{tinued}
\stick{tinued my father,\tsh I declare I would}
\stick{not have my head so full of curtins and}
\stick{horn-works.\tsk That, I dare say, you}
\stick{would not, quoth \drslop, interrupting}
\stick{him, and laughing most immoderately at}
his pun.

\i{Dennis} the critic could not detest and\break
\stick{abhor a pun, or the insinuation of a pun,}
\stick{more cordially than my father;\tshh he}
\stick{would grow testy upon it at any time;\tskk}
\stick{but to be broke in upon by one, in a}
\stick{serious discourse, was as bad, he would}
\stick{say, as a fillip upon the nose;\tskk he saw}
no difference.

Sir, quoth my uncle \toby, addressing\break
\stick{himself to \drslop,\tsk  the curtins my}
\stick{brother \i{Shandy} mentions here, have no-}
\stick{thing to do with bed-steads;\tsk tho’, I}
\stick{know, \i{Du Cange} says, \lqq That bed-cur-}
\catch{\lqq tains,}
\stick{\lqq tains, in all probability, have taken}
\stick{\lqq their name from them;”\tsk nor have}
\stck{the horn-works, he speaks of, any thing in}
\stick{the world to do with the horn-works of}
\stick{cuckoldom:\tsk But the \i{curtin}, Sir, is the}
\stick{word we use in fortification, for that part}
\stick{of the wall or rampart which lies between}
\stick{the two bastions and joins them.\tsh Be-}
\stick{siegers seldom offer to carry on their at-}
\stick{tacks directly against the curtin, for this}
\stick{reason, because they are so well \i{flanked}.}
\stick{(’Tis the case of other curtins, quoth Dr.}
\stick{\i{Slop}, laughing). However, continued my}
\stick{uncle \toby, to make them sure, we ge-}
\stick{nerally choose to place ravelins before}
\stick{them, taking care only to extend them}
\stick{beyond the fossé or ditch:\tsh The com-}
\stick{mon men, who know very little of for-}
\stick{tification, confound the ravelin and the}
\stick{half-moon together,\tsk tho’ they are very}
\stick{different things;\tsk not in their figure or}
\catch{con-}
\stick{construction, for we make them exactly}
\stick{aliks\sic in all points;\tsk for they always con-}
\stick{sist of two faces, making a salient angle,}
\stick{with the gorges, not straight, but in}
\stick{form of a crescent;\tsk Where then lies the}
\stick{difference? (quoth my father, a little}
\stck{testily.)\tsk In their situations, answered my}
\stick{uncle \i{Toby:}\tsk For when a ravelin, bro-}
\stick{ther, stands before the curtin, it is a ra-}
\stick{velin; and when a ravelin stands before}
\stick{a bastion, then the ravelin is not a rave-}
\stick{lin;\tsk it is a half-moon;\tsk a half-moon}
\stick{likewise is a half-moon, and no more, so}
\stick{long as it stands before its bastion;\tsk but}
\stick{was it to change place, and get before the}
\stick{curtin,\tsk ’twould be no longer a half-}
\stick{moon; a half-moon, in that case, is not}
\stick{a half-moon;\tsk ’tis no more than a rave-}
\stick{lin.\tsk I think, quoth my father, that the}
\stick{noble science of defence has its weak}
\stick{sides,\tsk as well as others.\hfill}
\catch{\tsk As}
\stick{\indent \tsk As for the horn-works (high! ho!}
\stick{sigh’d my father) which, continued my}
\stick{uncle \toby, my brother was speaking of,}
\stick{they are a very considerable part of an}
\stick{outwork;\tsh they are called by the \i{French}}
\stick{engineers, \i{Ouvrage à corne}, and we gene-}
\stick{rally make them to cover such places as}
\stick{we suspect to be weaker than the rest;\tsk}
\stick{’tis form’d by two epaulments or demi-}
\stick{bastions\tsk they are very pretty, and if}
\stick{you will take a walk, I’ll engage to shew}
\stick{you one well worth your trouble.\tsh I}
\stick{own, continued my uncle \toby, when we}
\stick{crown them,\tsk they are much stronger,}
\stick{but then they are very expensive, and}
\stick{take up a great deal of ground; so that,}
\stick{in my opinion, they are most of use to}
\stick{cover or defend the head of a camp;}
\stick{otherwise the double tenaille\tsh By}
\stick{the mother who bore us!\tsh brother}
\catch{\toby,}
\stick{\toby, quoth my father, not able to hold}
\stick{out any longer,\tsk you would provoke a}
\stick{saint;\tsk here have you got us, I know not}
\stck{how, not only souse into the middle of the}
\stick{old subject again:\tsk But so full is your}
\stick{head of these confounded works, that}
\stick{tho’ my wife is this moment in the pains}
\stick{of labour,\tsk and you hear her cry out,\tsk}
\stick{yet nothing will serve you but to carry}
\stick{off the man-midwife.\tsh \i{Accoucheur},\tsk}
\stick{if you please, quoth \drslop.\tsk With all}
\stick{my heart, replied my father, I don’t care}
\stick{what they call you,\tsh but I wish the}
\stick{whole science of fortification, with all its}
\stick{inventors, at the devil;\tsk it has been the}
\stick{death of thousands,\tsk and it will be}
\stick{mine, in the end.\tsk I would not, I would}
\stick{not, brother \toby, have my brains so full}
\stck{of saps, mines, blinds, gabions, palisadoes,}
\stick{ravelins, half-moons, and such trum-}
\catch{pery,}
\stick{pery, to be proprietor of \i{Namur}, and of}
all the towns in \i{Flanders} with it.

My uncle \toby was a man patient of
\stick{injuries;\tsk not from want of courage,\tsk I}
\stick{have told you in the fifth\sic chapter of this}
\stick{second book, \lqq That he was a man of}
\stick{\lqq courage:”\tsk And will add here, that}
\stick{where just occasions presented, or called}
\stick{it forth,\tsk I know no man under whose}
\stick{arm I would have sooner taken shelter;}
\stick{nor did this arise from any insensibility}
\stick{or obtuseness of his intellectual parts;\tsk}
\stick{for he felt this insult of my father’s as}
\stick{feelingly as a man could do;\tsk but he}
\stick{was of a peaceful, placid nature,\tsk no}
\stick{jarring element in it,\tsk all was mix’d up}
\stick{so kindly within him; my uncle \i{Toby}}
\stick{had scarce a heart to retaliate upon}
\stick{a fly.\hfill}
\catch{\tsk Go}
\stick{\indent \tsk Go\tsk says he, one day at dinner, to}
\stick{an over-grown one which had buzz’d}
\stick{about his nose, and tormented him cruel-}
\stick{ly all dinner-time,\tsk and which, after in-}
\stick{finite attempts, he had caught at last, as}
\stick{it flew by him;\tsk I’ll not hurt thee, says}
\stick{my uncle \toby, rising from his chair, and}
\stick{going a-cross the room, with the fly in}
\stick{his hand,\tsk I’ll not hurt a hair of thy}
\stick{head:\tsk Go, says he, lifting up the sash,}
\stick{and opening his hand as he spoke, to let}
\stick{it escape;\tsk go poor devil, get thee gone,}
\stick{why should I hurt thee?\tsh This world}
\stick{surely is wide enough to hold both thee}
and me.

I was but ten years old when this hap-
\stick{pened;\tsk but whether it was, that the}
\stick{action itself was more in unison to my}
\stick{nerves at that age of pity, which instant-}
\stick{ly set my whole frame into one vibration}
\catch{of}
\stick{of most pleasurable sensation;\tsk or how}
\stick{far the manner and expression of it might}
\stick{go towards it;\tsk or in what degree, or}
\stick{by what secret magic,\tsk a tone of voice}
\stick{and harmony of movement, attuned by}
\stick{mercy, might find a passage to my heart,}
\stick{I know not;\tsk this I know, that the lesson}
\stick{of universal good-will then taught and}
\stick{imprinted by my uncle \toby, has never}
\stick{since been worn out of my mind: And}
\stck{tho’ I would not depreciate what the study}
\stck{of the \i{Literæ humaniores}, at the university,}
\stick{have done for me in that respect, or dis-}
\stck{credit the other helps of an expensive edu-}
\stick{cation bestowed upon me, both at home}
\stick{and abroad since;\tsk yet I often think that}
\stick{I owe one half of my philanthropy to}
that one accidental impression.

\vskip -2pt

\fist This is to serve for parents and\break
\stick{governors instead of a whole volume}
\stick{upon the subject.\hfill}
\catch{I}
\stick{\indent I could not give the reader this stroke}
\stick{in my uncle \toby's picture, by the instru-}
\stick{ment with which I drew the other parts}
\stick{of it,\tsk that taking in no more than}
\stick{the mere \s{Hobby-Horsical} likeness;}
\stick{\tsk this is a part of his moral character.}
\stick{My father, in this patient endurance of}
\stick{wrongs, which I mention, was very dif-}
\stick{ferent, as the reader must long ago have}
\stick{noted; he had a much more acute and}
\stick{quick sensibility of nature, attended with}
\stick{a little soreness of temper; tho’ this}
\stck{never transported him to any thing which}
\stick{looked like malignancy;\tsk\!\tsk yet, in the}
\stick{little rubs and vexations of life, ’twas}
\stick{apt to shew itself in a drollish and witty}
\stick{kind of peevishness:\tsh He was, how-}
\stick{ever, frank and generous in his nature;}
\stick{\tsk at all times open to conviction;}
\stick{and in the little ebullitions of this sub-}
\stick{acid humour towards others, but parti-}
\catch{cularly}
\stick{cularly towards my uncle \toby, whom}
\stick{he truly loved:\tsh he would feel more}
\stick{pain, ten times told (except in the af-}
\stick{fair of my aunt \i{Dinah}, or where an hy-}
\stick{pothesis was concerned) than what he}
ever gave.

The characters of the two brothers, in\break
\stick{this view of them, reflected light upon}
\stick{each other, and appear’d with great ad-}
\stick{vantage in this affair which arose about}
\i{Stevinus}.

I need not tell the reader, if he keeps a
\stick{\s{Hobby-Horse},\tsk that a man’s \s{Hobby-}}
\stick{\s{Horse} is as tender a part as he has}
\stick{about him; and that these unprovoked}
\stick{strokes, at my uncle \toby's could not be}
\stick{unfelt by him.\tsk No;\tsk as I said above,}
\stick{my uncle \toby did feel them, and very}
\stick{sensibly too.\hfill}
\catch{Pray,}
\stick{\indent Pray, Sir, what said he?\tsk How did}
\stick{he behave?\tsk Oh, Sir!\tsk it was great:}
\stick{For as soon as my father had done in-}
\stick{sulting his \s{Hobby-Horse},\tsh he turned}
\stick{his head, without the least emotion,}
\stick{from \drslop, to whom he was addres-}
\stick{sing his discourse, and look’d up into my}
\stick{father’s face, with a countenance spread}
\stick{over with so much good nature;\tsh so}
\stick{placid;\tsk so fraternal;\tsk so inexpressibly}
\stick{tender towards him;\tsk it penetrated my}
\stick{father to his heart: He rose up hastily}
\stick{from his chair, and seizing hold of both}
\stick{my uncle \toby's hands as he spoke:\tsh}
\stick{Brother \toby, said he,\tsk I beg thy par-}
\stick{don;\tsk forgive, I pray thee, this rash}
\stick{humour which my mother gave me.\tsk}
\stick{My dear, dear brother, answer’d my}
\stick{uncle \toby, rising up by my father’s help,}
\stick{say no more about it;\tsk you are heartily}
\stick{welcome, had it been ten times as much,}
\catch{brother.}
\stick{brother. But ’tis ungenerous, replied}
\stick{my father, to hurt any man;\tsk a brother}
\stick{worse;\tsk but to hurt a brother of such}
\stick{gentle manners,\tsk so unprovoking,\tsk and}
\stick{so unresenting;\tsk ’tis base:\tsk By heaven,}
\stick{’tis cowardly.\tsh You are heartily wel-}
\stick{come, brother, quoth my uncle \toby,\tsk}
\stick{had it been fifty times as much.\tsh}
\stick{Besides, what have I to do, my dear \toby,}
\stick{cried my father, either with your amuse-}
\stick{ments or your pleasures, unless it was in}
\stick{my power (which it is not) to increase}
their measure?

\tsk Brother \i{Shandy}, answer’d my uncle
\stick{\toby, looking wistfully in his face,\tsk you}
\stick{are much mistaken in this point;\tsk for}
\stick{you do increase my pleasure very much,}
\stick{in begetting children for the \i{Shandy} fa-}
\stick{mily at your time of life.\tsk But, by}
\stick{that, Sir, quoth \drslop, Mr.\@ \i{Shandy} in-}
\catch{creases}
\stick{creases his own.\tsh Not a jot, quoth}
my father.

\chap{XIII}{28pt}{8pt}

\dropcap[findent=2pt]{M}{y} brother, does it, quoth my uncle\break
\toby, out of \i{principle.}\tsh In a fa-\break
\stick{mily-way, I suppose, quoth \drslop.\tsk}
\stick{Pshaw!\tsk said my father,\tsk ’tis not worth}
talking of.

\chap{XIV}{28pt}{8pt}

\dropcap[findent=1pt]{A}{t} the end of the last chapter, my\break
father and my uncle \toby were\break
\stick{left both standing, like \i{Brutus} and \i{Cassius}}
\stick{at the close of the scene making up their}
\stick{accounts.\hfill}
\catch{As}
\stck{\indent As my father spoke the three last words,}
\stick{\tsk he sat down;\tsk my uncle \toby exactly}
\stick{followed his example, only, that before}
\stick{he took his chair, he rung the bell, to}
\stick{order Corporal \trim, who was in waiting,}
\stick{to step home for \i{Stevinus};\tsk my uncle}
\stck{\toby's house being no further off than the}
opposite side of the way.

Some men would have dropp’d the\break
\stick{subject of \i{Stevinus};\tsk but my uncle \toby}
\stick{had no resentment in his heart, and he}
\stick{went on with the subject, to shew my}
father that he had none.

Your sudden appearance, \drslop,\break
\stick{quoth my uncle, resuming the discourse,}
\stick{instantly brought \i{Stevinus} into my head.}
\stick{(My father, you may be sure, did not}
\stick{offer to lay any more wagers upon \i{Ste}-}
\stick{\i{vinus}’s head)\tsh Because, continued my}
\catch{uncle}
\stick{uncle \toby, the celebrated sailing chariot,}
\stick{which belonged to Prince \i{Maurice}, and}
\stick{was of such wonderful contrivance and}
\stick{velocity, as to carry half a dozen peo-}
\stick{ple thirty \i{German} miles, in I don’t know}
\stick{how few minutes,\tsh was invented by}
\stick{\i{Stevinus}, that great mathematician and}
engineer.

You might have spared your servant\break
\stick{the trouble, quoth \drslop (as the fellow}
\stick{is lame) of going for \i{Stevinus}’s account}
\stick{of it, because, in my return from \i{Leyden}}
\stick{thro’ the \i{Hague}, I walked as far as \i{Schev}-}
\stick{\i{ling}, which is two long miles, on purpose}
to take a view of it.

\tsk That’s nothing, replied my uncle\break
\stick{\toby, to what the learned \i{Peireskius} did,}
\stick{who walked a matter of five hundred}
\stick{miles, reckoning from \i{Paris} to \i{Schevling},}
\catch{and}
\stick{and from \i{Schevling} to \i{Paris} back again,}
in order to see it,\tsk and nothing else.

Some men cannot bear to be out-gone.

The more fool \i{Peireskius}, replied Dr.\break
\stick{\i{Slop.} But mark, ’twas out of no con-}
\stick{tempt of \i{Peireskius} at all;\tsh but that}
\stick{\i{Peireskius}’s indefatigable labour in trudg-}
\stick{ing so far on foot out of love for the}
\stick{sciences, reduced the exploit of \drslop,}
\stick{in that affair, to nothing;\tsk the more fool}
\stick{\i{Peireskius}, said he again:\tsk Why so?\tsk}
\stick{replied my father, taking his brother’s}
\stick{part, not only to make reparation as fast}
\stick{as he could for the insult he had given}
\stick{him, which sat still upon my father’s}
\stick{mind;\tsk but partly, that my father began}
\stick{really to interest himself in the discourse.}
\stick{\tsk Why so?\tsk said he. Why is \i{Peireskius},}
\stick{or any man else, to be abused for an ap-}
\catch{petite}
\stick{petite for that, or any other morsel of}
\stick{sound knowledge? For notwithstand-}
\stick{ing I know nothing of the chariot in}
\stick{question, continued he, the inventor of}
\stick{it must have had a very mechanical head;}
\stick{and tho’ I cannot guess upon what prin-}
\stick{ciples of philosophy he has atchiev’d}
\stick{it;\tsh yet certainly his machine has}
\stick{been constructed upon solid ones, be}
\stick{they what they will, or it could not}
\stick{have answer’d at the rate my brother}
mentions.

It answered, replied my uncle \toby,\break
\stick{as well, if not better; for, as \i{Peireskius}}
\stick{elegantly expresses it, speaking of the}
\stick{velocity of its motion, \i{Tam citus erat},}
\stick{\i{quam erat ventus}; which, unless I have}
\stick{forgot my Latin, is, \i{that it was as swift}}
\stick{\i{as the wind itself}.\hfill}
\catch{But}
\stick{\indent But pray, \drslop, quoth my father,}
\stick{interrupting my uncle (tho’ not with-}
\stick{out begging pardon for it, at the same}
\stick{time) upon what principles was this self-}
\stick{same chariot set a-going?\tsk Upon very}
\stick{pretty principles to be sure, replied Dr.}
\stick{\i{Slop};\tsk and I have often wondered,}
\stick{continued he, evading the question, why}
\stick{none of our gentry, who live upon large}
\stick{plains like this of ours,\tsk (especially they}
\stick{whose wives are not past child-bearing)}
\stick{attempt nothing of this kind; for it}
\stick{would not only be infinitely expeditious}
\stick{upon sudden calls, to which the sex is}
\stick{subject,\tsk if the wind only served,\tsk but}
\stick{would be excellent good husbandry to}
\stick{make use of the winds, which cost no-}
\stick{thing, and which eat nothing, rather}
\stick{than horses, which (the devil take ’em)}
\stick{both cost and eat a great deal.\hfill}
\catch{For}
\stick{\indent For that very reason, replied my fa-}
\stick{ther, \lqq Because they cost nothing, and}
\stick{\lqq because they eat nothing,”\tsk the scheme}
\stick{is bad;\tsk it is the consumption of our}
\stick{products, as well as the manufactures of}
\stick{them, which gives bread to the hungry,}
\stick{circulates trade,\tsk brings in money, and}
\stick{supports the value of our lands;\tsk and}
\stick{tho’, I own, if I was a prince, I would}
\stck{generously recompense the scientific head}
\stick{which brought forth such contrivances;}
\stick{\tsk yet I would as peremptorily suppress}
the use of them.

My father here had got into his ele-\break
\stick{ment,\tsk and was going on as pro-}
\stck{sperously with his dissertation upon trade,}
\stick{as my uncle \toby had before, upon his}
\stick{of fortification;\tsh but, to the loss}
\stick{of much sound knowledge, the destinies}
\stick{in the morning had decreed that no}
\catch{dis-}
\stick{dissertation of any kind should be spun}
\stick{by my father that day,\tsh for as}
\stick{he opened his mouth to begin the next}
sentence,

\chap{XV}{28pt}{10pt}

\dropcap{I}{n} popp’d Corporal \trim with \i{Stevi}-\break
\i{nus:}\tsk But ’twas too late,\tsk all the\break
\stk{discourse had been exhausted without him,}
and was running into a new channel.

\tsk You may take the book home again,\break
\stick{\trim, said my uncle \toby, nodding to}
him.

But pri’thee, Corporal, quoth my fa-\break
\stick{ther, drolling,\tsk look first into it, and}
\stick{see if thou canst spy aught of a sailing}
chariot in it.
\natch{Corporal}

Corporal \trim, by being in the service,\break
\stick{had learned to obey,\tsk and not to re-}
\stick{monstrate,\tsk so taking the book to a}
\stick{side-table, and running over the leaves;}
\stick{an’ please your Honour, said \trim, I can}
\stick{see no such thing;\tsk however, continued}
\stick{the Corporal, drolling a little in his turn,}
\stick{I’ll make sure work of it, an’ please your}
\stick{Honour;\tsk so taking hold of the two}
\stick{covers of the book, one in each hand,}
\stick{and letting the leaves fall down, as he}
\stick{bent the covers back, he gave the book}
a good sound shake.

There is something fallen out, how-\break
\stick{ever, said \trim, an’ please your Honour;}
\stick{but it is not a chariot, or any thing like}
\stick{one:\tsk Pri’thee Corporal, said my father,}
\stick{smiling, what is it then?\tsk I think, an-}
\stick{swered \trim, stooping to take it up,\tsk}
\stick{’tis more like a sermon,\tsk for it begins,}
\catch{with}
\stick{with a text of scripture, and the chapter}
\stick{and verse;\tsk and then goes on, not as a}
chariot,\tsk but like a sermon directly.

The company smiled.

I cannot conceive how it is possible,\break
\stick{quoth my uncle \toby, for such a thing as}
a sermon to have got into my \i{Stevinus.}

I think ’tis a sermon, replied \trim;\tsk\break
\stick{but if it please your Honours, as it is a}
\stick{fair hand, I will read you a page;\tsk for}
\stick{\trim, you must know, loved to hear}
himself read almost as well as talk.

I have ever a strong propensity, said
\stick{my father, to look into things which}
\stick{cross my way, by such strange fatalities}
\stick{as these;\tsk and as we have nothing bet-}
\stick{ter to do, at least till \i{Obadiah} gets back,}
\catch{I}
\stick{I should be obliged to you, brother, if}
\stick{\drslop has no objection to it, to order}
\stick{the Corporal to give us a page or two of}
\stick{it,\tsk if he is as able to do it, as he seems}
\stick{willing. An’ please your Honour, quoth}
\stick{\trim, I officiated two whole campaigns}
\stick{in \i{Flanders}, as clerk to the chaplain of}
\stick{the regiment.\tsk He can read it, quoth}
\stick{my uncle \toby, as well as I can.\tsk \trim,}
\stick{I assure you, was the best scholar in my}
\stick{company, and should have had the next}
\stick{halberd, but for the poor fellow’s mis-}
\stick{fortune. Corporal \trim laid his hand}
\stick{upon his heart, and made an humble bow}
\stick{to his master;\tsk then laying down his hat}
\stick{upon the floor, and taking up the sermon}
\stick{in his left-hand, in order to have his}
\stick{right at liberty,\tsk he advanced, nothing}
\stick{doubting, into the middle of the room,}
\stick{where he could best see, and be best seen}
by, his audience.

\natch{\lsss{CHAP}.}

\chap{XVI}{28pt}{10pt}

\tsh If you have any objection,\tsk said\break
\stick{my father, addressing himself to \drslop.}
\stick{Not in the least, replied \drslop;\tsk for}
\stick{it does not appear on which side of the}
\stick{question it is wrote;\tsk it may be a}
\stick{composition of a divine of our church,}
\stick{as well as yours,\tsk so that we run equal}
\stick{risks.\tsh ’Tis wrote upon neither side,}
\stick{quoth \trim, for ’tis only upon \i{Conscience},}
an’ please your Honours.

\trim's reason put his audience into\break
\stick{good humour,\tsk all but \drslop, who,}
\stick{turning his head about towards \trim,}
look’d a little angry.

Begin, \trim,\tshh and read distinctly,
\stick{quoth my father;\tsk I will, an’ please your}
\catch{Honour,}
\stick{Honour, replied the Corporal, making a}
\stick{bow, and bespeaking attention with a}
slight movement of his right-hand.

\chap{XVII}{28pt}{10pt}

\tsh But before the Corporal begins,\break
\stick{I must first give you a description of his}
\stick{attitude;\tsh otherwise he will naturally}
\stick{stand represented, by your imagination,}
\stick{in an uneasy posture,\tsk stiff,\tsk perpendi-}
\stick{cular,\tsk dividing the weight of his body}
\stick{equally upon both legs;\tsh his eye fix’d,}
\stick{as if on duty;\tsk his look determined,\tsk}
\stick{clinching the sermon in his left-hand,}
\stick{like his firelock:\tsk In a word, you would}
\stick{be apt to paint \trim, as if he was stand-}
\stick{ing in his platoon ready for action:\tsh}
\stick{His attitude was as unlike all this as you}
can conceive.
\natch{He}

He stood before them with his body\break
\stick{swayed, and bent forwards just so far, as}
\stick{to make an angle of 85 degrees and a}
\stick{half upon the plain of the horizon;\tsh}
\stick{which sound orators, to whom I address}
\stick{this, know very well, to be the true per-}
\stick{suasive angle of incidence;\tsk in any other}
\stick{angle you may talk and preach;\tsk ’tis}
\stick{certain,\tsk and it is done every day;\tsk}
\stick{but with what effect,\tsk I leave the world}
to judge!

The necessity of this precise angle of\break
\stick{85 degrees and a half to a mathematical}
\stick{exactness,\tsk does it not shew us, by the}
\stick{way,\tsk how the arts and sciences mutu-}
ally befriend each other?

How the duce Corporal \trim, who\break
\stick{knew not so much as an acute angle}
\stick{from an obtuse one, came to hit it so ex-}
\catch{actly;}
\stick{actly;\tsh or whether it was chance or}
\stick{nature, or good sense or imitation, \etc}
\stick{shall be commented upon in that part}
\stick{of the cyclopædia of arts and sciences,}
\stick{where the instrumental parts of the elo-}
\stick{quence of the senate, the pulpit, the bar,}
\stick{the coffee-house, the bed-chamber, and}
fire-side, fall under consideration.

He stood,\tsh for I repeat it, to take\break
\stick{the picture of him in at one view, with}
\stick{his body sway’d, and somewhat bent}
\stick{forwards,\tsk his right-leg firm under him,}
\stick{sustaining seven\,-\,eighths of his whole}
\stick{weight,\tsk the foot of his left-leg, the de-}
\stick{fect of which was no disadvantage to his}
\stick{attitude, advanced a little,\tsk not lateral-}
\stick{ly, nor forwards, but in a line betwixt}
\stick{them;\tsk his knee bent, but that not vio-}
\stick{lently,\tsk but so as to fall within the li-}
\stick{mits of the line of beauty;\tsk and I add,}
\catch{of}
\stick{of the line of science too;\tsk for consider,}
\stick{it had one eighth part of his body to}
\stick{bear up;\tsk so that in this case the posi-}
\stick{tion of the leg is determined,\tsk because}
\stick{the foot could be no further advanced,}
\stick{or the knee more bent, than what would}
\stick{allow him mechanically, to receive an}
\stick{eighth part of his whole weight under}
it,\tsk and to carry it too.

\fist This I recommend to painters?\tsk\break
\stick{need I add,\tsk to orators?\tsk I think not;}
\stick{for, unless they practise it,\tsk they must}
fall upon their noses.

So much for Corporal \trim's body and\break
\stick{legs.\tsk He held the sermon loosely,\tsk not}
\stick{carelessly, in his left-hand, raised some-}
\stick{thing above his stomach, and detach’d a}
\stick{little from his breast;\tsh his right-arm}
\stick{falling negligently by his side, as nature}
\catch{and}
\stick{and the laws of gravity ordered it,\tsk but}
\stick{with the palm of it open and turned to-}
\stick{wards his audience, ready to aid the sen-}
timent, in case it stood in need.

\tight{Corporal \trim's eyes and the muscles of}\break
\stick{his face were in full harmony with the}
\stick{other parts of him;\tsk he look’d frank,\tsk}
\stick{unconstrained,\tsk something assured,\tsh}
but not bordering upon assurance.

Let not the critic ask how Corporal\break
\stick{\trim could come by all this; I’ve told}
\stick{him it shall be explained;\tsk but so he}
\stick{stood before my father, my uncle \toby,}
\stick{and \drslop,\tsk so swayed his body, so}
\stick{contrasted his limbs, and with such an}
\stick{oratorical sweep throughout the whole}
\stick{figure,\tsh a statuary might have modell’d}
\stick{from it;\tsh nay, I doubt whether the}
\stick{oldest Fellow of a College,\tsk or the \i{He}-}
\catch{\i{brew}}
\stick{\i{brew} Professor himself, could have much}
mended it.

\trim made a bow, and read as follows:

\bigskip
\centerline{The\quad \lsv{48}{SERMON}.}
\bigskip
\centerline{\s{Hebrews}\, xiii. 18.}
\tsh \i{For we} trust \i{we have a good}\break
\i{Conscience}.\tsh

\initial[findent=1pt,ante=\rlap{“}\smash{\lower12pt\hbox{\lqq}}]{T}{Rust!}\tsk Trust we have a good\break
conscience!”

[Certainly, \trim, quoth my father, in-\break
\stick{terrupting him, you give that sentence a}
\stick{very improper accent; for you curl up}
\stick{your nose, man, and read it with such a}
\stick{sneering tone, as if the Parson was going}
\stick{to abuse the Apostle.\hfill}
\catch{He}
\stick{\indent He is, an’ please your Honour, replied}
\trim.\quad Pugh! said my father, smiling.

Sir, quoth Dr.\@ \i{Slop, Trim} is certainly\break
\stick{in the right; for the writer (who I per-}
\stick{ceive is a Protestant) by the snappish}
\stick{manner in which he takes up the Apostle,}
\stick{is certainly going to abuse him,\tsk if this}
\stick{treatment of him has not done it already.}
\stick{But from whence, replied my father,}
\stick{have you concluded so soon, \drslop,}
\stick{that the writer is of our Church?\tsk for}
\stick{aught I can see yet,\tsk he may be of any}
\stick{Church:\tsh Because, answered \drslop,}
\stick{if he was of ours.\sicc\tsk he durst no more}
\stick{take such a licence,\tsk than a bear by his}
\stick{beard:\tsk If, in our communion, Sir,}
\stick{a man was to insult an Apostle,\tsh a}
\stick{saint,\tsh or even the paring of a saint’s}
\stick{nail,\tsk he would have his eyes scratched}
\stick{out.\tsk What, by the saint, quoth my}
\catch{uncle}
\stick{uncle \toby. No, replied \drslop,\tsk he}
\stick{would have an old house over his head.}
\stick{Pray is the Inquisition an antient build-}
\stick{ing, answered my uncle \toby, or is it a}
\stick{modern one?\tsk I know nothing of archi-}
\stick{tecture, replied \drslop.\tsh An’ please}
\stick{your Honours, quoth \trim, the Inqui-}
\stick{sition is the vilest\tsh Pri’thee spare thy}
\stick{description, \trim, I hate the very name}
\stick{of it, said my father.\tsk No matter for}
\stick{that, answered \drslop,\tsk it has its uses;}
\stick{for tho’ I’m no great advocate for it, yet}
\stick{in such a case as this, he would soon be}
\stick{taught better manners; and I can tell}
\stick{him, if he went on at that rate, would}
\stick{be flung into the Inquisition for his pains.}
\stick{God help him then, quoth my uncle}
\stick{\toby. Amen, added \trim; for, heaven}
\stick{above knows, I have a poor brother who}
\stick{has been fourteen years a captive in it.\tsh}
\stick{I never heard one word of it before, said}
\catch{my}
\stick{my uncle \toby, hastily:\tsk How came he}
\stick{there, \i{Trim?}\tsh O, Sir!\@ the story will}
\stick{make your heart bleed,\tsk as it has made}
\stick{mine a thousand times;\tsk but it is too}
\stick{long to be told now;\tsh your Honour}
\stick{shall hear it from first to last some day}
\stick{when I am working beside you in our}
\stick{fortifications;\tsk but the short of the}
\stick{story is this:\tsk That my brother \i{Tom}}
\stick{went over a servant to \i{Lisbon},\tsk and then}
\stick{married a \i{Jew}’s widow, who kept a small}
\stck{shop, and sold sausages, which, some how}
\stick{or other, was the cause of his being taken}
\stick{in the middle of the night out of his}
\stick{bed, where he was lying with his wife}
\stick{and two small children, and carried di-}
\stick{rectly to the Inquisition, where, God help}
\stick{him, continued \trim, fetching a sigh from}
\stick{the bottom of his heart,\tsk the poor ho-}
\stick{nest lad lies confined at this hour;\tsh}
\stick{he was as honest a soul, added \trim,}
\catch{(pulling}
\stick{(pulling out his handkerchief) as ever}
blood warm’d.\tsh

\tsh The tears trickled down \trim's\break
\stick{cheeks faster than he could well wipe}
\stick{them away.\tsk A dead silence in the room}
\stick{ensued for some minutes.\tsh Certain}
proof of pity!

Come \trim, quoth my father, after\break
\stick{he saw the poor fellow’s grief had got a}
\stick{little vent,\tsk read on,\tsk and put this me-}
\stick{lancholy story out of thy head:\tsk I grieve}
\stick{that I interrupted thee;\tsk but pri’thee}
\stick{begin the sermon again;\tsk for if the first}
\stick{sentence in it is matter of abuse, as thou}
\stick{sayest, I have a great desire to know what}
\tight{kind of provocation the Apostle has given.}

Corporal \trim wiped his face, and re-\break
\stick{turned his handkerchief into his pocket,}
\catch{and,}
\stick{and, making a bow as he did it,\tsk he be-}
gan again.]

\bigskip
\centerline{The\quad \lsv{48}{SERMON}.}
\bigskip
\centerline{\s{Hebrews}\, xiii. 18.}
\tsh \i{For we} trust \i{we have a good}\break
\i{Conscience}.\tsh

\initial[findent=1pt,ante=\rlap{“}\smash{\lower12pt\hbox{\lqq}}]{T}{Rust!} trust we have a good con-\break
science! Surely if there is any\break
\stick{\lqq thing in this life which a man may}
\stick{\lqq depend upon, and to the knowledge}
\stick{\lqq of which he is capable of arriving}
\stick{\lqq upon the most indisputable evidence,}
\stick{\lqq it must be this very thing,\tsk whether}
\stick{\lqq he has a good conscience or no.”\hfill}

[I am positive I am right, quoth Dr.\break
\i{Slop.}]

\natch{\lqq If}

\lqq If a man thinks at all, he cannot\break
\stick{\lqq well be a stranger to the true state of}
\stick{\lqq this account:\tsk he must be privy to}
\stick{\lqq his own thoughts and desires;\tsh he}
\stick{\lqq must remember his past pursuits, and}
\stick{\lqq know certainly the true springs and}
\stick{\lqq motives, which, in general, have go-}
\lqq verned the actions of his life.”

[\,I defy him, without an assistant,\break
quoth \drslop.\,]

\lqq In other matters we may be deceived\break
\stick{\lqq by false appearances; and, as the wise}
\stick{\lqq man complains, \i{hardly do we guess}}
\stick{\lqq\i{aright at the things that are upon the}}
\stick{\lqq\i{earth, and with labour do we find the}}
\stick{\lqq\i{things that are before us}. But here the}
\stick{\lqq mind has all the evidence and facts}
\stick{\lqq within herself;\tsk is conscious of the web}
\stick{\lqq she has wove;\tsk knows its texture and}
\catch{\lqq fine-}
\stick{\lqq fineness, and the exact share which}
\stick{\lqq every passion has had in working upon}
\stick{\lqq the several designs which virtue or vice}
\lqq has plann’d before her.”

[The language is good, and I declare\break
\trim reads very well, quoth my father.]

\lqq Now,\tsk as conscience is nothing else\break
\stick{\lqq but the knowledge which the mind has}
\stick{\lqq within herself of this; and the judg-}
\stick{\lqq ment, either of approbation or censure,}
\stick{\lqq which it unavoidably makes upon the}
\stick{\lqq successive actions of our lives; ’tis plain}
\stick{\lqq you will say, from the very terms of}
\stick{\lqq the proposition,\tsk whenever this inward}
\stick{\lqq testimony goes against a man, and he}
\stick{\lqq stands self-accused,\tsk that he must ne-}
\stick{\lqq cessarily be a guilty man.\tsk And, on}
\stick{\lqq the contrary, when the report is fa-}
\stick{\lqq vourable on his side, and his heart con-}
\catch{\lqq demns}
\stick{\lqq demns him not;\tsh that it is not a}
\stick{\lqq matter of \i{trust}, as the Apostle intimates,}
\stick{\lqq \tsk but a matter of \i{certainty} and fact,}
\stick{\lqq that the conscience is good, and that}
\lqq the man must be good also.”

[Then the Apostle is altogether in the\break
\stick{wrong, I suppose, quoth \drslop, and}
\stick{the Protestant divine is in the right. Sir,}
\stick{have patience, replied my father, for I}
\stick{think it will presently appear that \i{St.\@ Paul}}
\stick{and the Protestant divine are both of an}
\stick{opinion.\tsk As nearly so, quoth \drslop,}
\stick{as east is to west;\tsk but this, continued}
\stick{he, lifting both hands, comes from the}
liberty of the press.

It is no more, at the worst, replied my\break
\stick{uncle \toby, than the liberty of the pulpit;}
\stick{for it does not appear that the sermon is}
\stick{printed, or ever likely to be.\hfill}
\catch{Go}
\stick{\indent Go on, \trim, quoth my father.]\hfill}\\[12pt]
\stick{\indent \lqq At first sight this may seem to be a}
\stick{\lqq true state of the case: and I make no}
\stick{\lqq doubt but the knowledge of right and}
\stick{\lqq wrong is so truly impressed upon the}
\stick{\lqq mind of man,\tsk that did no such thing}
\stick{\lqq ever happen, as that the conscience of}
\stick{\lqq a man, by long habits of sin, might}
\stick{\lqq (as the scripture assures it may) insen-}
\stick{\lqq sibly become hard;\tsk and, like some}
\stick{\lqq tender parts of his body, by much}
\stick{\lqq stress and continual hard usage, lose,}
\stick{\lqq by degrees, that nice sense and per-}
\stick{\lqq ception with which God and nature}
\stick{\lqq endow’d it:\tsk Did this never happen;}
\stick{\lqq \tsk or was it certain that self-love could}
\stick{\lqq never hang the least bias upon the}
\stick{\lqq judgment;\tsk or that the little interests}
\stick{\lqq below, could rise up and perplex the}
\stick{\lqq faculties of our upper regions, and}
\catch{\lqq en-}
\stick{\lqq encompass them about with clouds and}
\stick{\lqq thick darkness:\tsh Could no such}
\stick{\lqq thing as favour and affection enter this}
\stick{\lqq sacred \s{Court}:\tsk  Did \s{Wit} disdain to}
\stick{\lqq take a bribe in it;\tsk or was asham’d to}
\stick{\lqq shew its face as an advocate for an}
\stick{\lqq unwarrantable enjoyment:\tsk Or, lastly,}
\stick{\lqq were we assured, that \s{Interest} stood}
\stick{\lqq always unconcern’d whilst the cause}
\stick{\lqq was hearing,\tsk and that passion never}
\stick{\lqq got into the judgment-seat, and pro-}
\stick{\lqq nounc’d sentence in the stead of reason,}
\stick{\lqq which is supposed always to preside}
\stick{\lqq and determine upon the case:\tshh Was}
\stick{\lqq this truly so, as the objection must}
\stick{\lqq suppose;\tsk no doubt then, the religious}
\stick{\lqq and moral state of a man would be}
\stick{\lqq exactly what he himself esteem’d it;\tsk}
\stick{\lqq and the guilt or innocence of every}
\stick{\lqq man’s life could be known, in general,}
\stick{\lqq by no better measure, than the de-}
\catch{\lqq grees}
\stick{\lqq grees of his own approbation and cen-}
\lqq sure.

\lqq I own, in one case, whenever a man’s\break
\stick{\lqq conscience does accuse him (as it sel-}
\stick{\lqq dom errs on that side) that he is guil-}
\stick{\lqq ty; and, unless in melancholy and hy-}
\stick{\lqq pocondriac cases, we may safely pro-}
\stick{\lqq nounce upon it, that there is always}
\lqq sufficient grounds for the accusation.

\lqq But the converse of the proposition\break
\stick{\lqq will not hold true;\tshh namely, that}
\stick{\lqq whenever there is guilt, the conscience}
\stick{\lqq must accuse; and if it does not, that}
\stick{\lqq a man is therefore innocent.\tsk This is}
\stick{\lqq not fact:\tsk So that the common con-}
\stick{\lqq solation which some good christian or}
\stick{\lqq other is hourly administering to him-}
\stick{\lqq self,\tsk that he thanks God his mind}
\stick{\lqq does not misgive him; and that, con-}
\catch{\lqq sequently,}
\stick{\lqq sequently, he has a good conscience,}
\stick{\lqq because he hath a quiet one,\tsk is falla-}
\stick{\lqq cious;\tsk and as current as the inference}
\stick{\lqq is, and as infallible as the rule appears}
\stick{\lqq at first sight, yet, when you look nearer}
\stick{\lqq to it, and try the truth of this rule}
\stick{\lqq upon plain facts,\tsk you see it liable to}
\stick{\lqq so much error from a false application;}
\stick{\lqq \tsk the principle upon which it goes so}
\stick{\lqq often perverted;\tsk the whole force of}
\stick{\lqq it lost, and sometimes so vilely cast}
\stick{\lqq away, that it is painful to produce the}
\stick{\lqq common examples from human life}
\lqq which confirm the account.

\lqq A man shall be vicious and utterly\break
\stick{\lqq debauched in his principles;\tsk excep-}
\stick{\lqq tionable in his conduct to the world;}
\stick{\lqq shall live shameless, in the open com-}
\stick{\lqq mission of a sin which no reason or pre-}
\stick{\lqq tence can justify;\tsk a sin, by which}
\catch{\lqq con-}
\stick{\lqq contrary to all the workings of huma-}
\stick{\lqq nity, he shall ruin for ever the deluded}
\stick{\lqq partner of his guilt;\tsk rob her of her}
\stick{\lqq best dowry; and not only cover her own}
\stick{\lqq head with dishonour,\tsk but involve a}
\stick{\lqq whole virtuous family in shame and}
\stick{\lqq sorrow for her sake.\tsk Surely, you will}
\stick{\lqq think conscience must lead such a man}
\stick{\lqq a troublesome life;\tsk he can have no}
\lqq rest night or day from its reproaches.

\lqq Alas! \s{Conscience} had something\break
\stick{\lqq else to do, all this time, than break in}
\stick{\lqq upon him; as \i{Elijah} reproached the}
\stick{\lqq God \i{Baal},\tsh this domestic God \i{was}}
\stick{\lqq \i{either talking, or pursuing, or was in a}}
\stick{\lqq \i{journey, or peradventure he slept and}}
\lqq \i{could not be awoke.}

\lqq Perhaps \s{He} was gone out in com-\break
\stick{\lqq pany with \s{Honour} to fight a duel;}
\catch{\lqq to}
\stick{\lqq to pay off some debt at play;\tshh or}
\stick{\lqq dirty annuity, the bargain of his lust:}
\stick{\lqq Perhaps \s{Conscience} all this time was}
\stick{\lqq engaged at home, talking loud against}
\stick{\lqq petty larceny, and executing vengeance}
\stick{\lqq upon some such puny crimes as his}
\stick{\lqq fortune and rank of life secured him}
\stick{\lqq against all temptation of committing;}
\stick{\lqq so that he lives as merrily,” [If he was}
\stick{of our church, tho’, quoth \drslop, he}
\stick{could not]\tskk \lqq sleeps as soundly in his}
\stick{\lqq bed;\tsk and at last meets death as un-}
\stick{\lqq concernedly;\tsk perhaps much more so}
\lqq than a much better man.”

All this is impossible with us, quoth\break
\stick{\drslop, turning to my father,\tsk the case}
\stick{could not happen in our church.\tshh It}
\stick{happens in ours, however, replied my fa-}
\stick{ther, but too often.\tskk I own, quoth Dr.}
\stick{\i{Slop} (struck a little with my father’s}
\catch{frank}
\stick{frank acknowledgment)\tsk that a man in}
\stick{the \i{Romish} church may live as badly;\tsk}
\stick{but then he cannot easily die so.\tsk ’Tis}
\stick{little matter, replied my father, with an}
\stick{air of indifference,\tsk how a rascal dies.\tsk}
\stick{I mean, answer’d \drslop, he would be}
\stick{denied the benefits of the last sacraments.}
\stick{\tsk Pray how many have you in all, said}
\stick{my uncle \toby,\tsk for I always forget?\tsk}
\stick{Seven, answered \drslop.\tsk Humph!\tsk}
\stick{said my uncle \toby;\tsk tho’ not accented}
\stick{as a note of acquiescence,\tsk but as an in-}
\stck{terjection of that particular species of sur-}
\stick{prize, when a man in looking into a}
\stick{drawer, finds more of a thing than he ex-}
\stick{pected.\tsk Humph! replied my uncle \toby.}
\stick{\drslop, who had an ear, understood my}
\stick{uncle \toby as well as if he had wrote a}
\stick{whole volume against the seven sacra-}
\stick{ments.\tsh Humph! replied \drslop,}
\stick{(stating my uncle \toby's argument over}
\catch{again}
\stick{again to him)\tsh Why, Sir, are there}
\stick{not seven cardinal virtues?\tsh Seven}
\stick{mortal sins?\tsh Seven golden candle-}
\stick{sticks?\tsh Seven heavens?\tsh ’Tis}
\stick{more than I know, replied my uncle}
\stick{\toby.\tsh Are there not seven won-}
\stick{ders of the world?\tsh Seven days of the}
\stick{creation?\tsh Seven planets?\tsk\tsk Seven}
\stick{plagues?\tsh That there are, quoth my}
\stick{father, with a most affected gravity. But}
\stick{pri’thee, continued he, go on with the}
rest of thy characters, \trim.]

\lqq Another is sordid, unmerciful,” (here\break
\stick{\trim waved his right-hand) \lqq a strait-}
\stick{\lqq hearted, selfish wretch, incapable either}
\stick{\lqq of private friendship or public spirit.}
\stick{\lqq Take notice how he passes by the wi-}
\stick{\lqq dow and orphan in their distress, and}
\stick{\lqq sees all the miseries incident to human}
\stick{\lqq life without a sigh or a prayer.\,” [And}
\catch{please}
\stick{please your Honours, cried \trim, I think}
this a viler man than the other.]

\lqq \tight{Shall not conscience rise up and sting}\break
\stick{\lqq him on such occasions?\tsh No; thank}
\stick{\lqq God there is no occasion; \i{I pay every}}
\stick{\lqq \i{man his own};\i{\tsk I have no fornication to}}
\stick{\lqq \i{answer to my conscience};\i{\tsk no faithless}}
\stick{\lqq \i{vows or promises to make up};\i{\tsk I have}}
\stick{\lqq \i{debauched no man’s wife or child}; \i{thank}}
\stick{\lqq \i{God, I am not as other men, adulterers,}}
\stick{\lqq \i{unjust, or even as this libertine, who}}
\lqq \i{stands before me.}

\lqq A third is crafty and designing in\break
\stick{\lqq his nature. View his whole life;\tsk ’tis}
\stick{\lqq nothing but a cunning contexture of}
\stick{\lqq dark arts and unequitable subterfuges,}
\stick{\lqq basely to defeat the true intent of all}
\stick{\lqq laws,\tsk plain dealing and the safe en-}
\stick{\lqq joyment of our several properties.\tsh}
\catch{\lqq You}
\stick{\lqq You will see such a one working out}
\stick{\lqq a frame of little designs upon the ig-}
\stick{\lqq norance and perplexities of the poor}
\stick{\lqq and needy man;\tsk shall raise a fortune}
\stick{\lqq upon the inexperience of a youth, or}
\stick{\lqq the unsuspecting temper of his friend,}
\stick{\lqq who would have trusted him with his}
\lqq life.

\lqq When old age comes on, and re-\break
\stick{\lqq pentance calls him to look back upon}
\stick{\lqq this black account, and state it over}
\stick{\lqq again with his conscience,\tsh \s{Con}-}
\stick{\lqq \s{science} looks into the \s{Statutes} at}
\stick{\lqq \s{Large};\tsk finds no express law broken}
\stick{\lqq by what he has done;\tsk perceives no}
\stick{\lqq penalty or forfeiture of goods and chat-}
\stick{\lqq tels incurred;\tsk sees no scourge waving}
\stick{\lqq over his head, or prison opening his}
\stick{\lqq gates upon him:\tsk What is there to}
\stick{\lqq affright his conscience?\tsk Conscience}
\catch{\lqq has}
\stick{\lqq has got safely entrenched behind the}
\stick{\lqq Letter of the Law; sits there invul-}
\stick{\lqq nerable, fortified with \gothic{\large Cases} and \gothic{\large Re-}}
\stick{\lqq \gothic{\large ports} so strongly on all sides;\tsk that}
\stick{\lqq it is not preaching can dispossess it of}
\lqq its hold.\,”

[Here Corporal \trim and my uncle\break
\stick{\toby exchanged looks with each other.\tsk}
\stick{Aye, \tsk aye, \i{Trim!} quoth my uncle \toby,}
\stick{shaking his head,\tsk these are but sorry}
\stick{fortifications, \trim.\tsh O! very poor}
\stick{work, answered \trim, to what your Ho-}
\stick{nour and I make of it.\tsh The cha-}
\stick{racter of this last man, said \drslop, in-}
\stick{terrupting \trim, is more detestable than}
\stick{all the rest;\tsh and seems to have}
\stick{been taken from some pettifogging Law-}
\stick{yer amongst you:\tsk Amongst us, a}
\stick{man’s conscience could not possibly con-}
\stick{tinue so long \i{blinded};\tsk three times in}
\catch{a}
\stick{a year, at least, he must go to confession.}
\stick{Will that restore it to sight, quoth my}
\stick{uncle \i{Toby?}\tsk Go on, \trim, quoth my}
\stick{father, or \i{Obadiah} will have got back}
\stick{before thou has got to the end of thy}
\stick{sermon;\tsh ’tis a very short one, replied}
\stick{\trim.\tsk I wish it was longer, quoth my}
\stick{uncle \toby, for I like it hugely.\tsk \trim}
went on.]

\lqq A fourth man shall want even this\break
\stick{\lqq refuge;\tsk shall break through all this}
\stick{\lqq ceremony of slow chicane;\tsk scorns}
\stick{\lqq the doubtful workings of secret plots}
\stick{\lqq and cautious trains to bring about his}
\stick{\lqq purpose:\tsk See the bare-faced villain,}
\stick{\lqq how he cheats, lies, perjures, robs,}
\stick{\lqq murders.\tsk\!\tsk Horrid!\tsh But indeed}
\stick{\lqq much better was not to be expected,}
\stick{\lqq in the present case,\tsk the poor man}
\stick{\lqq was in the dark!\tsk his priest had got}
\catch{\lqq the}
\stick{\lqq the keeping of his conscience;\tsk and}
\stick{\lqq all he would let him know of it, was,}
\stick{\lqq That he must believe in the Pope;\tsk}
\stick{\lqq go to Mass;\tsk cross himself;\tsk tell his}
\stick{\lqq beads;\tsk be a good Catholic, and}
\stick{\lqq that this, in all conscience, was enough}
\stick{\lqq to carry him to heaven.  What;\tsk if}
\stick{\lqq he perjures!\tsk Why;\tsk he had a men-}
\stick{\lqq tal reservation in it.\tsk But if he is so}
\stick{\lqq wicked and abandoned a wretch as}
\stick{\lqq you represent him;\tsk if he robs,\tsk if}
\stick{\lqq he stabs,\tsk will not conscience, on every}
\stick{\lqq such act, receive a wound itself? Aye,}
\stick{\lqq \tsk but the man has carried it to con-}
\stick{\lqq fession;\tsk the wound digests there,}
\stick{\lqq and will do well enough, and in a}
\stick{\lqq short time be quite healed up by ab-}
\stick{\lqq solution. O Popery! what hast thou}
\stick{\lqq to answer for?\tsk when, not content}
\stick{\lqq with the too many natural and fatal}
\stick{\lqq ways, thro’ which the heart of man is}
\catch{\lqq every}
\stick{\lqq every day thus treacherous to itself}
\stick{\lqq above all things;\tsk thou hast wilfully}
\stick{\lqq set open the wide gate of deceit before}
\stick{\lqq the face of this unwary traveller, too}
\stick{\lqq apt, God knows, to go astray of him-}
\stick{\lqq self; and confidently speak peace to}
\lqq himself, when there is no peace.

\lqq Of this the common instances which\break
\stick{\lqq I have drawn out of life, are too no-}
\stick{\lqq torious to require much evidence. If}
\stick{\lqq any man doubts the reality of them,}
\stick{\lqq or thinks it impossible for a man to be}
\stick{\lqq such a bubble to himself,\tsk I must refer}
\stick{\lqq him a moment to his own reflections,}
\stick{\lqq and will then venture to trust my ap-}
\lqq peal with his own heart.

\lqq Let him consider in how different\break
\stick{\lqq a degree of detestation, numbers of}
\stick{\lqq wicked actions stand \i{there}, tho’ equally}
\catch{\lqq bad}
\stick{\lqq bad and vicious in their own natures;}
\stick{\lqq \tsk he will soon find, that such of them,}
\stick{\lqq as strong inclination and custom have}
\stick{\lqq prompted him to commit, are gene-}
\stick{\lqq rally dress’d out and painted with all}
\stick{\lqq the false beauties, which a soft and a}
\stick{\lqq flattering hand can give them;\tsk and}
\stick{\lqq that the others, to which he feels no}
\stick{\lqq propensity, appear, at once, naked and}
\stick{\lqq deformed, surrounded with all the}
\stick{\lqq true circumstances of folly and dis-}
\lqq honour.

\lqq When \i{David} surprized \i{Saul} sleep-\break
\stick{\lqq ing in the cave, and cut off the skirt}
\stick{\lqq of his robe,\tsk we read his heart smote}
\stick{\lqq him for what he had done:\tsk But in}
\stick{\lqq the matter of \i{Uriah}, where a faithful}
\stick{\lqq and gallant servant, whom he ought}
\stick{\lqq to have loved and honoured, fell to}
\stick{\lqq make way for his lust,\tsk where con-}
\catch{\lqq science}
\stick{\lqq science had so much greater reason to}
\stick{\lqq take the alarm, his heart smote him}
\stick{\lqq not.  A whole year had almost passed}
\stick{\lqq from first commission of that crime,}
\stick{\lqq to the time \i{Nathan} was sent to reprove}
\stick{\lqq him; and we read not once of the least}
\stick{\lqq sorrow or compunction of heart which}
\stick{\lqq he testified, during all that time, for}
\lqq what he had done.

\lqq Thus conscience, this once able mo-\break
\stick{\lqq nitor,\tsk\!\tsk placed on high as a judge}
\stick{\lqq within us, and intended by our maker}
\stick{\lqq as a just and equitable one too,\tsk by}
\stick{\lqq an unhappy train of causes and impe-}
\stick{\lqq diments, takes often such imperfect}
\stick{\lqq cognizance of what passes,\tsh does its}
\stick{\lqq office so negligently,\tsh sometimes so}
\stick{\lqq corruptly,\tsk that it is not to be trusted}
\stick{\lqq alone; and therefore we find there is}
\stick{\lqq a necessity, an absolute necessity of}
\catch{\lqq joining}
\stick{\lqq joining another principle with it to aid,}
\lqq if not govern, its determinations.

\lqq So that if you would form a just\break
\stick{\lqq judgment of what is of infinite impor-}
\stick{\lqq tance to you not to be misled in,\tsh}
\stick{\lqq namely, in what degree of real merit}
\stick{\lqq you stand either as an honest man, an}
\stick{\lqq useful citizen, a faithful subject to your}
\stick{\lqq king, or a good servant to your God,\tsk}
\stick{\lqq call in religion and morality.\tsk Look,}
\stick{\lqq \tsk What is written in the law of God?}
\stick{\lqq \tsh How readest thou?\tsh Consult}
\stick{\lqq calm reason and the unchangeable ob-}
\stick{\lqq ligations of justice and truth;\tsk what}
\lqq say they?

\lqq Let \s{Conscience} determine the\break
\stick{\lqq matter upon these reports;\tsk and then}
\stick{\lqq if thy heart condemns thee not, which}
\stick{\lqq is the case the Apostle supposes,\tsh the}
\catch{\lqq rule}
\stick{\lqq rule will be infallible;” [Here \drslop}
\stick{fell asleep] \lqq \i{thou wilt have confidence to-}}
\stick{\lqq \i{wards God};\tsk that is, have just grounds}
\stick{\lqq to believe the judgment thou hast past}
\stick{\lqq upon thyself, is the judgment of God;}
\stick{\lqq and nothing else but an anticipation}
\stick{\lqq of that righteous sentence which will}
\stick{\lqq be pronounced upon thee hereafter by}
\stick{\lqq that Being, to whom thou art finally}
\lqq to give an account of thy actions.

\lqq \i{Blessed is the man}, indeed then, as\break
\stick{\lqq the author of the book of \i{Ecclesiasticus}}
\stick{\lqq expresses it, \i{who is not prick’d with the}}
\stick{\lqq \i{multitude of his sins: Blessed is the man}}
\stick{\lqq \i{whose heart hath not condemn’d him};}
\stick{\lqq \i{whether he be rich, or whether he be}}
\stick{\lqq \i{poor, if he have a good heart} (a heart}
\stick{\lqq thus guided and informed) \i{he shall at}}
\stick{\lqq \i{all times rejoice in a chearful counte-}}
\stick{\lqq \i{nance; his mind shall tell him more than}}
\catch{\lqq \i{seven}}
\stick{\lqq \i{seven watch-men that sit above upon a}}
\stick{\lqq \i{tower on high}.\,”\tsh [A tower has}
\stick{no strength, quoth my uncle \toby, un-}
\stick{less ’tis flank’d.] \lqq In the darkest doubts}
\stick{\lqq it shall conduct him safer than a thou-}
\stick{\lqq sand casuists, and give the state he}
\stick{\lqq lives in a better security for his beha-}
\stick{\lqq viour than all the causes and re-}
\stick{\lqq strictions put together, which law-}
\stick{\lqq makers are forced to multiply:\tshh}
\stick{\lqq \i{Forced}, I say, as things stand; hu-}
\stick{\lqq man laws not being a matter of ori-}
\stick{\lqq ginal choice, but of pure necessity,}
\stick{\lqq brought in to fence against the mis-}
\stick{\lqq chievous effects of those consciences}
\stick{\lqq which are no law unto themselves;}
\stick{\lqq well intending, by the many provi-}
\stick{\lqq sions made,\tshh that in all such cor-}
\stick{\lqq rupt and misguided cases, where prin-}
\stick{\lqq ciples and the checks of conscience}
\stick{\lqq will not make us upright,\tsh to}
\catch{\lqq supply}
\stick{\lqq supply their force, and, by the ter-}
\stick{\lqq rors of gaols and halters, oblige us}
\lqq to it.”

[I see plainly, said my father, that\break
\stick{this sermon has been composed to be}
\stick{preached at the Temple,\tsh or at some}
\stick{Assize.\tsh I like the reasoning,\tsh and}
\stick{am sorry that \drslop has fallen asleep}
\stick{before the time of his conviction;\tsh}
\stick{for it is now clear, that the Parson, as}
\stick{I thought at first, never insulted St.\@ \i{Paul}}
\stick{in the least;\tsk nor has there been, bro-}
\stick{ther, the least difference between them.}
\stick{\tsh A great matter, if they had dif-}
\stick{fered, replied my uncle \toby,\tsk the}
\stick{best friends in the world may differ some-}
\stick{times.\tsh True,\tsk brother \toby quoth}
\stick{my father, shaking hands with him,\tsk}
\stick{we’ll fill our pipes, brother, and then}
\stick{\trim shall go on.\hfill}
\catch{Well,\tsk}
\stick{\indent Well,\tsk what dost thou think of it?}
\stick{said my father, speaking to Corporal}
\trim, as he reach’d his tobacco-box.

\tight{I think, answer’d the Corporal, that the}\break
\stick{seven watch-men upon the tower, who, I}
\stick{suppose, are all centinels there,\tsk are}
\stick{more, an’ please your Honour, than were}
\stick{necessary;\tsk and, to go on at that rate,}
\stick{would harrass a regiment all to pieces,}
\stick{which a commanding officer, who loves}
\stick{his men, will never do, if he can help it;}
\stck{because two centinels, added the Corporal,}
\stick{are as good as twenty.\tsk I have been a}
\stick{commanding officer myself in the \i{Corps}}
\stck{\i{de Garde} a hundred times, continued \trim,}
\stick{rising an inch higher in his figure, as he}
\stick{spoke,\tsk and all the time I had the ho-}
\stick{nour to serve his Majesty King \i{William},}
\stick{in relieving the most considerable posts,}
\stick{I never left more than two in my life.\tsk}
\catch{Very}
\stick{Very right, \trim, quoth my uncle \toby,}
\stick{\tsk but you do not consider, \trim, that}
\stick{the towers, in \i{Solomon}’s days, were not}
\stick{such things as our bastions, flank’d and}
\stick{defended by other works;\tsk this, \trim,}
\stick{was an invention since \i{Solomon}’s death;}
\stick{nor had they horn-works, or ravelins be-}
\stick{fore the curtin, in his time;\tsk or such a}
\stick{fossé as we make with a cuvette in the}
\stick{middle of it, and with cover’d-ways and}
\stick{counterscarps pallisadoed along it, to}
\stick{guard against a \i{Coup de main:}\tsk So that}
\stick{the seven men upon the tower were a}
\stick{party, I dare say, from the \i{Corps de Garde},}
\stick{set there, not only to look out, but to}
\stick{defend it.\tsk They could be no more, an’}
\stick{please your Honour, than a Corporal’s}
\stick{Guard.\tsk My father smiled inwardly,\tsk}
\stick{but not outwardly;\tsk the subject between}
\stick{my uncle \toby and Corporal \trim being}
\stick{rather too serious, considering what had}
\catch{hap-}
\stick{happened, to make a jest of:\tsk So put-}
\stick{ting his pipe into his mouth, which he}
\stick{had just lighted,\tsk he contented himself}
\stick{with ordering \trim to read on. He read}
on as follows:]

\lqq To have the fear of God before our\break
\stick{\lqq eyes, and, in our mutual dealings with}
\stick{\lqq each other, to govern our actions by the}
\stick{\lqq eternal measures of right and wrong:\tsk}
\stick{\lqq The first of these will comprehend the}
\stick{\lqq duties of religion;\tsh the second, those}
\stick{\lqq of morality, which are so inseparably}
\stick{\lqq connected together, that you cannot}
\stick{\lqq divide these two \i{tables}, even in imagi-}
\stick{\lqq nation (tho’ the attempt is often made}
\stick{\lqq in practice) without breaking and mu-}
\lqq tually destroying them both.

\lqq I said the attempt is often made,\break
\stick{\lqq and so it is;\tsh there being nothing}
\catch{\lqq more}
\stick{\lqq more common than to see a man who}
\stick{\lqq has no sense at all of religion,\tsk and in-}
\stick{\lqq deed has so much honesty as to pre-}
\stick{\lqq tend to none, who would take it as the}
\stick{\lqq bitterest affront, should you but hint at}
\stick{\lqq a suspicion of his moral character,\tsh}
\stick{\lqq or imagine he was not conscientiously}
\stick{\lqq just and scrupulous to the uttermost}
\lqq mite.

\lqq When there is some appearance that\break
\stick{\lqq it is so,\tsk tho’ one is unwilling even to}
\stick{\lqq suspect the appearance of so amiable a}
\stick{\lqq virtue as moral honesty, yet were we}
\stick{\lqq to look into the grounds of it, in the}
\stick{\lqq present case, I am persuaded we should}
\stick{\lqq find little reason to envy such a one}
\lqq the honour of his motive.

\lqq Let him declaim as pompously as\break
\stick{\lqq he chooses upon the subject, it will}
\catch{\lqq be}
\stick{\lqq be found to rest upon no better foun-}
\stick{\lqq dation than either his interest, his pride,}
\stick{\lqq his ease, or some such little and change-}
\stick{\lqq able passion as will give us but small}
\stick{\lqq dependance upon his actions in matters}
\lqq of great stress.

\lqq I will illustrate this by an ex-\break
\lqq ample.

\lqq I know the banker I deal with, or\break
\stick{\lqq the physician I usually call in,” [There}
\stick{is no need, cried \drslop, (waking) to call}
\stick{in any physician in this case] \lqq to be}
\stick{\lqq neither of them men of much religion:}
\stick{\lqq I hear them make a jest of it every}
\stick{\lqq day, and treat all its sanctions with so}
\stick{\lqq much scorn, as to put the matter past}
\stick{\lqq doubt.  Well;\tsk notwithstanding this,}
\stick{\lqq I put my fortune into the hands of the}
\stick{\lqq one;\tsk and what is dearer still to me,}
\catch{\lqq I}
\stick{\lqq I trust my life to the honest skill of}
\lqq the other.

\lqq Now let me examine what is my\break
\stick{\lqq reason for this great confidence.\tsh}
\stick{\lqq Why, in the first place, I believe there}
\stick{\lqq is no probability that either of them}
\stick{\lqq will employ the power I put into their}
\stick{\lqq hands to my disadvantage;\tsk I con-}
\stick{\lqq sider that honesty serves the purposes}
\stick{\lqq of this life:\tsk I know their success in}
\stick{\lqq the world depends upon the fairness of}
\stick{\lqq their characters.\tsh In a word,\tsk I’m}
\stick{\lqq persuaded that they cannot hurt me,}
\lqq without hurting themselves more.

\lqq But put it otherwise, namely, that\break
\stick{\lqq interest lay, for once, on the other side;}
\stick{\lqq that a case should happen, wherein the}
\stick{\lqq one, without stain to his reputation,}
\stick{\lqq could secrete my fortune, and leave}
\catch{\lqq me}
\stick{\lqq me naked in the world;\tsk or that the}
\stick{\lqq other could send me out of it, and en-}
\stick{\lqq joy an estate by my death, without}
\stick{\lqq dishonour to himself or his art:\tsk In}
\stick{\lqq this case, what hold have I of either of}
\stick{\lqq them?\tsk Religion, the strongest of all}
\stick{\lqq motives, is out of the question:\tsk In-}
\stick{\lqq terest, the next most powerful motive}
\stick{\lqq in the world, is strongly against me:\tsk}
\stick{\lqq What have I left to cast into the oppo-}
\stick{\lqq site scale to balance this temptation?\tsk}
\stick{\lqq Alas! I have nothing,\tsk nothing but}
\stick{\lqq what is lighter than a bubble\tsk I must}
\stick{\lqq lay at the mercy of \s{Honour}, or some}
\stick{\lqq such capricious principle.\tsk Strait secu-}
\stick{\lqq rity for two of the most valuable bles-}
\lqq sings!\tsk my property and my life.

\lqq As, therefore, we can have no de-\break
\stick{\lqq pendence upon morality without reli-}
\stick{\lqq gion;\tsk so, on the other hand, there is}
\catch{\lqq nothing}
\stick{\lqq nothing better to be expected from}
\stick{\lqq religion without morality;\tsk neverthe-}
\stick{\lqq less, ’tis no prodigy to see a man whose}
\stick{\lqq real moral character stands very low,}
\stick{\lqq who yet entertains the highest notion}
\stick{\lqq of himself, in the light of a religious}
\lqq man.

\lqq He shall not only be covetous, re-\break
\stick{\lqq vengeful, implacable,\tsk but even}
\stick{\lqq wanting in points of common ho-}
\stick{\lqq nesty; yet, inasmuch as he talks aloud}
\stick{\lqq against the infidelity of the age,\tsh}
\stick{\lqq is zealous for some points of reli-}
\stick{\lqq gion,\tsh goes twice a day to church,}
\stick{\lqq \tsh attends the sacraments,\tsh and}
\stick{\lqq amuses himself with a few instrumental}
\stick{\lqq parts of religion,\tsk shall cheat his}
\stick{\lqq conscience into a judgment that, for}
\stick{\lqq this, he is a religious man, and has}
\stick{\lqq discharged truly his duty to God:}
\catch{\lqq And}
\stick{\lqq And you will find that such a man,}
\stick{\lqq through force of this delusion, gene-}
\stick{\lqq rally looks down with spiritual pride}
\stick{\lqq upon every other man who has less}
\stick{\lqq affectation of piety,\tsk though, perhaps,}
\stick{\lqq ten times more moral honesty than}
\lqq himself.

\lqq \i{This likewise is a sore evil under the}\break
\stick{\lqq \i{sun}; and, I believe there is no one mi-}
\stick{\lqq staken principle, which, for its time,}
\stick{\lqq has wrought more serious mischiefs.\tsh}
\stick{\lqq For a general proof of this,\tsk examine}
\stick{\lqq the history of the \i{Romish} church;”\tsk}
\stick{[Well, what can you make of that, cried}
\stick{Dr.\@ \i{Slop?}]\tsk \lqq see what scenes of cru-}
\stick{\lqq elty, murders, rapines, blood-shed,”}
\stick{[They may thank their own obstinacy,}
\stick{cried \drslop] \lqq have all been sanctified}
\stick{\lqq by a religion not strictly governed by}
\stick{\lqq morality.\hfill}
\catch{\lqq In}
\stck{\indent \lqq In how many kingdoms of the world,”}
\stick{[Here \trim kept waving his right-hand}
\stick{from the sermon to the extent of his arm,}
\stick{returning it backwards and forwards to}
the conclusion of the paragraph.]

\lqq In how many kingdoms of the world\break
\stick{\lqq has the crusading sword of this mis-}
\stick{\lqq guided saint-errant spared neither age,}
\stick{\lqq or merit, or sex, or condition?\tsk and,}
\stick{\lqq as he fought under the banners of a}
\stick{\lqq religion which set him loose from ju-}
\stick{\lqq stice and humanity, he shew’d none;}
\stick{\lqq mercilessly trampled upon both,\tsh}
\stick{\lqq heard neither the cries of the unfortu-}
\lqq nate, nor pitied their distresses.”

[I have been in many a battle, an’\break
\stick{please your Honour, quoth \trim, sighing,}
\stick{but never in so melancholy a one as}
\stick{this.\tsk I would not have drawn a tricker}
\catch{in}
\stick{in it, against these poor souls,\tshh to}
\stick{have been made a general officer.\tsh}
\stick{Why? what do you understand of the}
\stick{affair? said \drslop, looking towards}
\stck{\trim with something more contempt than}
\stick{the Corporal’s honest heart deserved.\tsh}
\stick{What do you know, friend, about this}
\stick{battle you talk of?\tsh I know, replied}
\stick{\trim, that I never refused quarter in my}
\stick{life to any man who cried out for it;\tsh}
\stick{but to a woman or a child, continued}
\stick{\trim, before I would level my musket}
\stick{at them, I would lose my life a thou-}
\stick{sand times.\tsk Here’s a crown for thee,}
\stick{\trim, to drink with \i{Obadiah} to-night,}
\stick{quoth my uncle \toby, and I’ll give \i{Oba}-}
\stick{\i{diah} another too.\tsk God bless your Ho-}
\stick{nour, replied \trim,\tsh I had rather these}
\stick{poor women and children had it.\tsh}
\stick{Thou art an honest fellow, quoth my}
\stick{uncle \toby.\tsh My father nodded his}
\catch{head,}
\stick{head,\tsh as much as to say,\tsh and so}
he is.\hbox to 3em{\tsfill}

\tight{But pri’thee \trim, said my father, make}\break
\stick{an end,\tsk for I see thou hast but a leaf}
or two left.]

\noindent
\stick{\hfill Corporal \trim read on.\hfill}

\lqq If the testimony of past centuries\break
\stick{\lqq in this matter is not sufficient,\tsk con-}
\stick{\lqq sider at this instant, how the votaries}
\stick{\lqq of that religion are every day thinking}
\stick{\lqq to do service and honour to God, by}
\stick{\lqq actions which are a dishonour and scan-}
\lqq dal to themselves.

\lqq To be convinced of this, go with\break
\stick{\lqq me for a moment into the prisons of}
\stick{\lqq the inquisition.” [God help my poor}
\stick{brother \i{Tom}.]\tsk \lqq Behold \i{Religion}, with}
\catch{\lqq\i{Mercy}}
\stick{\lqq \i{Mercy} and \i{Justice} chained down under}
\stick{\lqq her feet,\tsh there sitting ghastly upon}
\stick{\lqq a black tribunal, propp’d up with racks}
\stick{\lqq and instruments of torment. Hark!}
\stick{\lqq \tsk hark!  what a piteous groan!” [Here}
\stick{\trim's face turned as pale as ashes.]}
\stick{\lqq See the melancholy wretch who ut-}
\stick{\lqq ter’d it,\,”\tsk [Here the tears began to}
\stick{trickle down] \lqq just brought forth to un-}
\stick{\lqq dergo the anguish of a mock trial, and}
\stick{\lqq endure the utmost pains that a studied}
\stick{\lqq system of cruelty has been able to in-}
\stick{\lqq vent.”\tsk [D\bzz{am}n them all, quoth \trim,}
\stick{his colour returning into his face as red}
\stick{as blood.]\tsk \lqq Behold this helpless victim}
\stick{\lqq delivered up to his tormentors,\tsk his}
\stick{\lqq body so wasted with sorrow and con-}
\stick{\lqq finement.\,”\tsk [Oh!\@ ’tis my brother,}
\stick{cried poor \trim in a most passionate}
\stick{exclamation, dropping the sermon upon}
\stick{the ground, and clapping his hands to-}
\catch{gether}
\stick{gether\tsk I fear ’tis poor \i{Tom.} My fa-}
\stick{ther’s and my uncle \toby's heart yearn’d}
\stick{with sympathy for the poor fellow’s di-}
\stick{stress,\tsk even \i{Slop} himself acknowledged}
\stick{pity for him.\tsh Why, \trim, said my fa-}
\stick{ther, this is not a history,\tsk ’tis a sermon}
\stick{thou art reading;\tsk pri’thee begin the}
\stick{sentence again.]\tsh \lqq Behold this helpless}
\stick{\lqq victim deliver’d up to his tormentors,}
\stick{\lqq \tsk his body so wasted with sorrow and}
\stick{\lqq confinement, you will see every nerve}
\lqq and muscle as it suffers.

\lqq Observe the last movement of that\break
\stick{\lqq horrid engine!” [I would rather face}
\stick{a cannon, quoth \trim, stamping.]\tsh}
\stick{\lqq See what convulsions it has thrown}
\stick{\lqq him into!\tsh Consider the nature of}
\stick{\lqq the posture in which he how lies}
\stick{\lqq stretched,\tsk what exquisite tortures he}
\stick{\lqq endures by it!\,”\tsh [I hope ’tis not in}
\catch{\i{Portugal}.]}
\stick{\i{Portugal}.]\tsk \lqq ’Tis all nature can bear!}
\stick{\lqq Good God! see how it keeps his weary}
\stick{\lqq soul hanging upon his trembling lips!”}
\stick{[I would not read another line of it,}
\stick{quoth \trim for all this world;\tsk I fear,}
\stick{an’ please your Honours, all this is in}
\stick{\i{Portugal}, where my poor brother \i{Tom} is.}
\stick{I tell thee, \trim, again, quoth my father,}
\stick{’tis not an historical account,\tsk ’tis a de-}
\stck{scription.\tsk ’Tis only a description, honest}
\stick{man, quoth \i{Slop}, there’s not a word of}
\stick{truth in it.\tsk That’s another story, replied}
\stick{my father.\tsk However, as \trim reads it}
\stick{with so much concern,\tsk ’tis cruelty to}
\stick{force him to go on with it.\tsk Give me}
\stick{hold of the sermon, \trim,\tsk I’ll finish it}
\stick{for thee, and thou may’st go. I must stay}
\stick{and hear it too, replied \trim, if your Ho-}
\stick{nour will allow me;\tsk tho’ I would not}
\stick{read it myself for a Colonel’s pay.\tsh}
\catch{Poor}
\stick{Poor \i{Trim!} quoth my uncle \toby. My}
father went on.]

\lqq \tsk Consider the nature of the posture\break
\stick{\lqq in which he now lies stretch’d,\tsk what}
\stick{\lqq exquisite torture he endures by it!\tsk}
\stick{\lqq ’Tis all nature can bear! Good God!}
\stick{\lqq See how it keeps his weary soul hang-}
\stick{\lqq ing upon his trembling lips,\tsk willing}
\stick{\lqq to take its leave,\tsh but not suffered}
\stick{\lqq to depart!\tsk\!\tsk Behold the unhappy}
\stick{\lqq wretch led back to his cell!” [Then,}
\stick{thank God, however, quoth \trim, they}
\stick{have not killed him]\tsk\lqq See him dragg’d}
\stick{\lqq out of it again to meet the flames, and}
\stick{\lqq the insults in his last agonies, which}
\stick{\lqq this principle,\tsh this principle, that}
\stick{\lqq there can be religion without mercy,}
\stick{\lqq has prepared for him.” [Then, thank}
\stick{God,\tsk he is dead, quoth \trim,\tsk he is}
\catch{out}
\stick{out of his pain,\tsk and they have done}
\stick{their worst at him.\tsk O Sirs!\tsk Hold}
\stick{your peace, \trim, said my father, going}
\stick{on with the sermon, lest \trim should in-}
\stick{cense \drslop,\tsk we shall never have done}
at this rate.]

\lqq The surest way to try the merit of\break
\stick{\lqq any disputed notion is, to trace down}
\stick{\lqq the consequences such a notion has}
\stick{\lqq produced, and compare them with the}
\stick{\lqq spirit of Christianity;\tsh ’tis the short}
\stick{\lqq and decisive rule which our Saviour}
\stick{\lqq hath left us, for these and such-like}
\stick{\lqq cases, and it is worth a thousand ar-}
\stick{\lqq guments,\tsk \i{By their fruits ye shall know}}
\lqq \i{them.}

\lqq I will add no further to the length\break
\stick{\lqq of this sermon, than, by two or three}
\catch{\lqq short}
\stick{\lqq short and independent rules deducible}
\lqq from it.

\lqq \i{First}, Whenever a man talks loudly\break
\stick{\lqq against religion,\tsk always suspect that it}
\stick{\lqq is not his reason, but his passions which}
\stick{\lqq have got the better of his \s{Creed}. A}
\stick{\lqq bad life and a good belief are disagree-}
\stick{\lqq able and troublesome neighbours, and}
\stick{\lqq \tight{where they separate, depend upon it, ’tis}}
\lqq for no other cause but quietness sake.

\lqq \tight{\i{Secondly}, When a man, thus represen-}\break
\stick{\lqq ted, tells you in any particular instance,}
\stick{\lqq \tsk That such a thing goes \i{against} his}
\stick{\lqq conscience,\tsk always believe he means}
\stick{\lqq exactly the same thing, as when he tells}
\stick{\lqq you such a thing goes \i{against} his sto-}
\stick{\lqq mach;\tsk a present want of appetite}
\stick{\lqq being generally the true cause of both.\hfill}
\catch{\lqq In}
\stick{\indent \lqq In a word,\tsk trust that man in no-}
\stick{\lqq thing, who has not a \s{Conscience} in}
\lqq every thing.

\lqq And, in your own case, remember\break
\stick{\lqq this plain distinction, a mistake in}
\stick{\lqq which has ruined thousands,\tsk that}
\stick{\lqq your conscience is not a law:\tsk No,}
\stick{\lqq God and reason made the law, and}
\stick{\lqq have placed conscience within you to}
\stick{\lqq determine;\tsh not like an \i{Asiatic} Cadi,}
\stick{\lqq according to the ebbs and flows of his}
\stick{\lqq own passions,\tsk but like a \i{British} judge}
\stick{\lqq in this land of liberty and good sense,}
\stick{\lqq who makes no new law, but faithfully}
\stick{\lqq declares that law which he knows al-}
\lqq ready written.”

\noindent
\centerline{\lsv{80}{\large\i{FINIS}}.}

\natch[25pt]{Thou}

Thou hast read the sermon extremely
\stick{well, \trim, quoth my father.\tsk If he had}
\stick{spared his comments, replied \drslop,}
\stick{he would have read it much better.  I}
\stick{should have read it ten times better, Sir,}
\stick{answered \trim, but that my heart was so}
\stick{full.\tsk That was the very reason, \trim,}
\stick{replied my father, which has made thee}
\stck{read the sermon as well as thou hast done;}
\stick{and if the clergy of our church, continued}
\stick{my father, addressing himself to \drslop,}
\stick{would take part in what they deliver, as}
\stick{deeply as this poor fellow has done,\tsk as}
\stick{their compositions are fine; (I deny it,}
\stick{quoth Dr.\@ \i{Slop)}\sic I maintain it, that the}
\stick{eloquence of our pulpits, with such sub-}
\stick{jects to inflame it,\tsk would be a model}
\stick{for the whole world:\tsk But, alas! con-}
\stick{tinued my father, and I own it, Sir, with}
\stick{sorrow, that, like \i{French} politicians in this}
\stick{respect, what they gain in the cabinet}
\catch{they}
\stick{they lose in the field.\tsh ’Twere a pity,}
\stick{quoth my uncle, that this should be lost.}
\stick{I like the sermon well, replied my fa-}
\stick{ther,\tsk ’tis dramatic,\tsk and there is}
\stick{something in that way of writing, when}
\stick{skilfully managed, which catches the at-}
\stick{tention.\tsh We preach much in that}
\stick{way with us, said \drslop.\tsk I know that}
\stick{very well, said my father,\tsk but in a tone}
\stick{and manner which disgusted \drslop,}
\stick{full as much as his assent, simply, could}
\stick{have pleased him.\tsh But in this, added}
\stick{\drslop, a little piqued,\tsh our sermons}
\stick{have greatly the advantage, that we ne-}
\stick{ver introduce any character into them}
\stick{below a patriarch or a patriarch’s wife,}
\stick{or a martyr or a saint.\tsk There are some}
\stick{very bad characters in this, however, said}
\stick{my father, and I do not think the sermon}
\stick{a jot the worse for ’em.\tsh But pray,}
\stick{quoth my uncle \toby,\tsk who’s can this}
\catch{be?}
\stick{be?\tsk How could it get into my \i{Stevi}-}
\stick{\i{nus}? A man must be as great a conjurer}
\stick{as \i{Stevinus}, said my father, to resolve}
\stick{the second question:\tsk The first, I think,}
\stick{is not so difficult;\tsh for unless my judg-}
\stick{ment greatly deceives me,\tsh I know the}
\stick{author, for ’tis wrote, certainly, by the}
parson of the parish.

The similitude of the stile and man-\break
\stick{ner of it, with those my father con-}
\stick{stantly had heard preach’d in his pa-}
\stick{rish-church, was the ground of his con-}
\stick{jecture,\tshh proving it as strongly, as}
\stick{an argument \i{à priori}, could prove such}
\stick{a thing to a philosophic mind, That}
\stick{it was \i{Yorick}’s and no one’s else:\tsh}
\stick{It was proved to be so \i{à posteriori},}
\stick{the day after, when \i{Yorick} sent a ser-}
\stick{vant to my uncle \toby's house to enquire}
\stick{after it.\hfill}
\catch{It}
\stick{\indent It seems that \i{Yorick}, who was inqui-}
\stick{sitive after all kinds of knowledge, had}
\stick{borrowed \i{Stevinus} of my uncle \toby, and}
\stick{had carelesly popp’d his sermon, as soon}
\stick{as he had made it, into the middle of}
\stick{\i{Stevinus}; and, by an act of forgetfulness,}
\stick{to which he was ever subject, he had}
\stick{sent \i{Stevinus} home, and his sermon to}
keep him company.

Ill-fated sermon! Thou wast lost, after
\stick{this recovery of thee, a second time,}
\stick{dropped thro’ an unsuspected fissure in}
\stick{thy master’s pocket, down into a trea-}
\stick{cherous and a tatter’d lining,\tskk trod deep}
\stick{into the dirt by the left hind foot of his}
\stck{Rosinante, inhumanly stepping upon thee}
\stick{as thou falledst;\tskk buried ten days in the}
\stick{mire,\tskk raised up out of it by a beggar,}
\stick{sold for a halfpenny to a parish-clerk,\tskk}
\stick{transferred to his parson,\tskk lost for ever}
\catch{to}
\stick{to thy own, the remainder of his days,\tsk}
\stick{nor restored to his restless \s{Manes} till}
\stick{this very moment, that I tell the world}
the story.

Can the reader believe, that this ser-\break
\stick{mon of \i{Yorick}’s was preach’d at an assize,}
\stick{in the cathedral of \i{York}, before a thou-}
\stick{sand witnesses, ready to give oath of it,}
\stick{by a certain prebendary of that church,}
\stick{and actually printed by him when he had}
\stick{done,\tsh and within so short a space as}
\stick{two years and three months after \i{Yorick}’s}
\stick{death.\tsk \i{Yorick}, indeed, was never better}
\stick{served in his life!\tsh but it was a little}
\stick{hard to male-treat\sic him after, and plun-}
der him after he was laid in his grave.

However, as the gentleman who did\break
\stick{it, was in perfect charity with \i{Yorick},\tsk}
\stick{and, in conscious justice, printed but a}
\catch{few}
\stick{few copies to give away;\tsk and that, I am}
\stick{told, he could moreover have made as}
\stick{good a one himself, had he thought fit,}
\stick{\tsk I declare I would not have published}
\stick{this anecdote to the world;\tsk nor do I}
\stick{publish it with an intent to hurt his cha-}
\stick{racter and advancement in the church;\tsk}
\stick{I leave that to others;\tsk but I find my-}
\stick{self impelled by two reasons, which I}
cannot withstand.

The first is, That, in doing justice, I\break
\stick{may give rest to \i{Yorick}’s ghost;\tsk which,}
\stick{as the country people,\tsk and some others,}
believe,\tsh \i{still walks}.

The second reason is, That, by laying\break
\stick{open this story to the world, I gain an}
\stick{opportunity of informing it,\tsk That in}
\stick{case the character of parson \i{Yorick}, and}
\stick{this sample of his sermons is liked,\tsk that}
\catch{there}
\stick{there are now in the possession of the}
\stick{\i{Shandy} family, as many as will make}
\stick{a handsome volume, at the world’s ser-}
\stick{vice,\tsh and much good may they}
do it.

\chap{XVIII}{28pt}{10pt}

\dropcap{O}{badiah} gain’d the two crowns\break
without dispute; for he came in\break
\stick{jingling, with all the instruments in the}
\stick{green bays bag we spoke of, slung across}
\stick{his body, just as Corporal \trim went out}
of the room.

It is now proper, I think, quoth Dr.\break
\stick{\i{Slop} (clearing up his looks) as we are in}
\stick{a condition to be of some service to}
\stick{Mrs.\@ \i{Shandy}, to send up stairs to know}
\stick{how she goes on.\hfill}
\catch{I}
\stick{\indent I have ordered, answered my father,}
\stick{the old midwife to come down to us}
\stick{upon the least difficulty;\tsk for you}
\stick{must know, \drslop, continued my fa-}
\stick{ther, with a perplexed kind of a smile}
\stick{upon his countenance, that by express}
\stick{treaty, solemnly ratified between me and}
\stick{my wife, you are no more than an auxi-}
\stick{liary in this affair,\tsk and not so much}
\stick{as that,\tsk\!\tsk unless the lean old mother}
\stick{of a midwife above stairs cannot do}
\stick{without you.\tsh Women have their}
\stick{particular fancies, and in points of this}
\stick{nature, continued my father, where they}
\stick{bear the whole burden, and suffer so}
\stick{much acute pain for the advantage of}
\stick{our families, and the good of the spe-}
\stick{cies,\tsk they claim a right of deciding,}
\stick{\i{en Souveraines}, in whose hands, and in}
\stick{what fashion, they chuse to undergo}
\stick{it.\hfill}
\catch{They}
\stick{\indent They are in the right of it,\tsk quoth}
\stick{my uncle \toby. But, Sir, replied Dr.}
\stick{\i{Slop}, not taking notice of my uncle \toby's}
\stick{opinion, but turning to my father,\tsk they}
\stick{had better govern in other points;\tsh and}
\stick{a father of a family, who wished its per-}
\stick{petuity, in my opinion, had better ex-}
\stick{change this prerogative with them, and}
\stick{give up some other rights in lieu of it.\tsh}
\stick{I know not, quoth my father, answering}
\stick{a little too testily, to be quite dispassion-}
\stick{ate in what he said,\tsk I know not, quoth}
\stick{he, what we have left to give up, in lieu}
\stick{of who shall bring our children into the}
\stick{world,\tsk unless that,\tsk of who shall beget}
\stick{them.\tsk One would almost give up}
\stick{any thing, replied \drslop.\tsk I beg}
\stick{your pardon,\tsk answered my uncle \toby.}
\stick{\tsk\!\tsk Sir, replied \drslop, it would asto-}
\stick{nish you to know what Improvements}
\stck{we have made of late years in all branches}
\catch{of}
\stick{of obstetrical knowledge, but particu-}
\stick{larly in that one single point of the safe}
\stick{ and expeditious extraction of the \i{fœtus},}
\stick{\tsh which has received such lights, that,}
\stick{for my part (holding up his hands) I}
\stick{declare I wonder how the world has\tsh}
\stick{I wish, quoth my uncle \toby, you had}
\stick{seen what prodigious armies we had in}
\i{Flanders}.

\chap{XIX}{24pt}{10pt}

\initial{I}{Have} dropp’d the curtain over this\break
scene for a minute,\tsk to remind you\break
\stick{of one thing,\tsk and to inform you of an-}
other.

What I have to inform you, comes, I\break
\stick{own, a little out of its due course;\tsk for}
\stick{it should have been told a hundred and}
\catch{fifty}
\stick{fifty pages ago, but that I foresaw then}
\stick{’twould come in pat hereafter, and be of}
\stick{more advantage here than elsewhere.\tsh}
\stick{Writers had need look before them to}
\stick{keep up the spirit and connection of what}
they have in hand.

When these two things are done,\tsk the\break
\stick{curtain shall be drawn up again, and my}
\stick{uncle \toby, my father, and \drslop shall}
\stick{go on with their discourse, without any}
more interruption.

First, then, the matter which I have\break
\stick{to remind you of, is this;\tsk that from}
\stck{the specimens of singularity in my father’s}
\stick{notions in the point of Christian-names,}
\stick{and that other point previous thereto,\tsk}
\stick{you was led, I think, into an opinion,}
\stick{(and I am sure I said as much) that my}
\stick{father was a gentleman altogether as odd}
\catch{and}
\stick{and whimsical in fifty other opinions. In}
\stick{truth, there was not a stage in the life of}
\stick{man, from the very first act of his be-}
\stick{getting,\tsh down to the lean and slipper’d}
\stick{pantaloon in his second childishness, but}
\stick{he had some favourite notion to himself,}
\stick{springing out of it, as sceptical, and as}
\stick{far out of the high-way of thinking, as}
these two which have been explained.

\tsk Mr.\@ \i{Shandy}, my father, Sir, would\break
\stick{see nothing in the light in which others}
\stick{placed it;\tsk he placed things in his own}
\stick{light;\tsk he would weigh nothing in com-}
\stick{mon scales;\tsk no,\tsk he was too refined a}
\stick{researcher to lay open to so gross an im-}
\stick{position.\tsk To come at the exact weight}
\stick{of things in the scientific steel-yard, the}
\stick{fulcrum, he would say, should be almost}
\stick{invisible, to avoid all friction from po-}
\stick{pular tenets;\tsk without this the minutiæ}
\catch{of}
\stick{of philosophy, which would always turn}
\stick{the balance, will have no weight at all.\tsk}
\stick{Knowledge, like matter, he would affirm,}
\stck{was divisible \i{in infinitum};\tsk that the grains}
\stick{and scruples were as much a part of it, as}
\stick{the gravitation of the whole world.\tsk In}
\stick{a word, he would say, error was error,\tsk}
\stick{no matter where it fell,\tsk whether in a}
\stick{fraction,\tsk or a pound,\tsk ’twas alike fatal}
\stick{to truth, and she was kept down at the}
\stick{bottom of her well as inevitably by a}
\stick{mistake in the dust of a butterfly’s wing,}
\stick{\tsk as in the disk of the sun, the moon, and}
all the stars of heaven put together.

He would often lament that it was for\break
\stick{want of considering this properly, and of}
\stick{applying it skilfully to civil matters, as}
\stck{well as to speculative truths, that so many}
\stick{things in this world were out of joint;\tsk}
\stick{that the political arch was giving way;\tsk}
\catch{and}
\stick{and that the very foundations of our ex-}
\stick{cellent constitution in church and state,}
\tight{were so sapp’d as estimators had reported.}

You cry out, he would say, we are a\break
\stck{ruined, undone people.\tsk Why? he would}
\stck{ask, making use of the sorites or syllogism}
\stick{of \i{Zeno} and \i{Chrysippus}, without knowing}
\stick{it belonged to them.\tsk Why? why are we}
\stck{a ruined people?\tsk Because we are corrup-}
\stick{ted.\tsk Whence is it, dear Sir, that we are}
\stck{corrupted?\tsk Because we are needy;\tsk our}
\stick{poverty, and not our wills, consent.\tsk}
\stick{And wherefore, he would add,\tsk are we}
\stick{needy?\tsk From the neglect, he would an-}
\stick{swer, of our pence and our halfpence:\tsk}
\stick{Our bank-notes, Sir, our guineas,\tsk nay}
our shillings, take care of themselves.

’Tis the same, he would say, through-\break
\stick{out the whole circle of the sciences;\tsk}
\catch{the}
\stick{the great, the established points of them,}
\stick{are not to be broke in upon.\tsk The laws}
\stick{of nature will defend themselves;\tsk but}
\stick{error\tsk (he would add, looking earnestly}
\stick{at my mother)\tsk error, Sir, creeps in thro’}
\stk{the minute holes, and small crevices, which}
human nature leaves unguarded.

This turn of thinking in my father, is\break
\stick{what I had to remind you of:\tsk The point}
\stick{you are to be informed of, and which I}
have reserved for this place, is as follows:

Amongst the many and excellent rea-\break
\stick{sons, with which my father had urged}
\stick{my mother to accept of \drslop's as-}
\stick{sistance preferably to that of the old wo-}
\stick{man,\tsk there was one of a very singular}
\stick{nature; which, when he had done ar-}
\stick{guing the matter with her as a Christian,}
\stick{and came to argue it over again with her}
\catch{as}
\stick{as a philosopher, he had put his whole}
\stick{strength to, depending indeed upon it as}
\stick{his sheet-anchor.\tsh It failed him, tho’}
\stick{from no defect in the argument itself;}
\stick{but that, do what he could, he was not}
\stick{able for his soul to make her comprehend}
\stick{the drift of it.\tsk Cursed luck!\tsk said he}
\stick{to himself, one afternoon, as he walk’d}
\stick{out of the room, after he had been sta-}
\stick{ting it for an hour and a half to her, to}
\stick{no manner of purpose;\tsk cursed luck!}
\stick{said he, biting his lip as he shut the door,}
\stick{\tsk for a man to be master of one of the}
\stick{finest chains of reasoning in nature,\tsh}
\stick{and have a wife at the same time with}
\stick{such a head-piece, that he cannot hang}
\stick{up a single inference within side of it, to}
save his soul from destruction.

This argument, tho’ it was intirely lost\break
\stick{upon my mother,\tsk had more weight with}
\catch{him,}
\stick{him, than all his other arguments joined}
\stick{together:\tsk I will therefore endeavour to}
\stick{do it justice,\tsk and set it forth with all the}
perspicuity I am master of.

My father set out upon the strength of\break
these two following axioms:

\i{First}, \tight{That an ounce of a man’s own wit,}\break
was worth a tun of other peoples; and,

\i{Secondly}, (Which, by the bye, was the\break
\stick{ground-work of the first axiom,\tsk tho’ it}
\stick{comes last) That every man’s wit must}
\stick{come from every man’s own soul,\tsk and}
no other body’s.

Now, as it was plain to my father, that\break
\stick{all souls were by nature equal,\tsk and that}
\stick{the great difference between the most}
\stck{acute and the most obtuse understanding,}
\catch{\tsk was}
\stick{\tsh was from no original sharpness or}
\stck{bluntness of one thinking substance above}
\stick{or below another,\tsk but arose merely from}
\stick{the lucky or unlucky organization of the}
\stick{body, in that part where the soul princi-}
\stick{pally took up her residence,\tsh he had}
\stick{made it the subject of his enquiry to find}
out the identical place.

Now, from the best accounts he had\break
\stick{been able to get of this matter, he was}
\stick{satisfied it could not be where \i{Des Cartes}}
\stick{had fixed it, upon the top of the \i{pineal}}
\stick{gland of the brain; which, as he philo-}
\stick{sophised, formed a cushion for her about}
\stick{the size of a marrow pea; tho’ to speak}
\stck{the truth, as so many nerves did terminate}
\stick{all in that one place,\tsk ’twas no bad con-}
\stick{jecture;\tsk and my father had certainly}
\stick{fallen with that great philosopher plumb}
\stick{into the center of the mistake, had it not}
\catch{been}
\stick{been for my uncle \toby, who rescued}
\stick{him out of it, by a story he told him of}
\stick{a \i{Walloon} officer at the battle of \i{Landen},}
\stick{who had one part of his brain shot away}
\stick{by a musket-ball,\tsk and another part of}
\stick{it taken out after by a \i{French} surgeon;}
\stick{and, after all, recovered, and did his duty}
very well without it.

\tight{If death, said my father, reasoning with}\break
\stick{himself, is nothing but the separation of}
\stick{the soul from the body;\tsk and if it is true}
\stick{that people can walk about and do their}
\stick{business without brains,\tsk then certes the}
soul does not inhabit there. \quad Q.\,E.\,D.

As for that certain, very thin, subtle,\break
\stk{and very fragrant juice which \i{Coglionissimo}}
\stick{\i{Borri}, the great \i{Milaneze} physician, af-}
\stck{firms, in a letter to \i{Bartholine}, to have dis-}
\stck{covered in the cellulæ of the occipital parts}
\catch{of}
\stick{of the cerebellum, and which he likewise}
\stick{affirms to be the principal seat of the rea-}
\stck{sonable soul (for, you must know, in these}
\stick{latter and more enlightened ages, there}
\stick{are two souls in every man living,\tsk the}
\stick{one, according to the great \i{Metheglingius},}
\stick{being called the \i{Animus}, the other, the}
\stick{\i{Anima\sic)};\tsk as for the opinion, I say, of}
\stick{\i{Borri},\tsk my father could never subscribe}
\stick{to it by any means; the very idea of so}
\stck{noble, so refined, so immaterial, and so ex-}
\stick{\tight{alted a being as the \i{Anima}, or even the \i{Ani-}}}
\stick{\i{mus}, taking up her residence, and sitting}
\stck{dabbling, like a tad-pole all day long, both}
\stick{summer and winter, in a puddle,\tsk or in a}
\stick{\tighter{liquid of any kind, how thick or thin soever,}}
\stick{\tight{he would say, shock’d his imagination; he}}
\tight{would scarce give the doctrine a hearing.}

\tight{What, therefore, seem’d the least liable}\break
\stick{to objections of any, was, that the chief}
\catch{sen-}
\stick{sensorium, or head-quarters of the soul,}
\stick{and to which place all intelligences were}
\stick{referred, and from whence all her man-}
\stick{dates were issued,\tsk was in, or near, the}
\stck{cerebellum,\tsk or rather some-where about}
\stick{the \i{medulla oblongata}, wherein it was ge-}
\stick{nerally agreed by \i{Dutch} anatomists, that}
\stick{all the minute nerves from all the organs}
\stk{of the seven senses concentered, like streets}
and winding alleys, into a square.

So far there was nothing singular in\break
\stick{my father’s opinion,\tsk he had the best of}
\stick{philosophers, of all ages and climates, to}
\stick{go along with him.\tsk But here he took a}
\stick{road of his own, setting up another \i{Shan}-}
\stck{\i{dean} hypothesis upon these corner-stones}
\stick{they had laid for him;\tsk and which said}
\stck{hypothesis equally stood its ground; whe-}
\stick{ther the subtilty and fineness of the soul}
\stck{depended upon the temperature and clear-}
\catch{ness}
\stick{ness of the said liquor, or of the finer}
\stick{net-work and texture in the cerebellum}
itself; which opinion he favoured.

He maintained, that next to the due\break
\stick{care to be taken in the act of propaga-}
\stick{tion of each individual, which required}
\stick{all the thought in the world, as it laid}
\stick{the foundation of this incomprehensible}
\stick{contexture in which wit, memory, fancy,}
\stick{eloquence, and what is usually meant by}
\stick{the name of good natural parts, do con-}
\stick{sist;\tsk that next to this and his Christian-}
\stick{name, which were the two original and}
\stick{most efficacious causes of all;\tsk that the}
\stick{third cause, or rather what logicians call}
\stck{the \i{Causa sine quâ non}, and without which}
\stick{all that was done was of no manner of}
\stck{significance,\tsk was the preservation of this}
\stick{delicate and fine-spun web, from the ha-}
\stick{vock which was generally made in it by}
\catch{the}
\stick{the violent compression and crush which}
\stick{the head was made to undergo, by the}
\stick{nonsensical method of bringing us into}
the world by that part foremost.

\tsh This requires explanation.

My father, who dipp’d into all kinds\break
\stick{of books, upon looking into \i{Lithopædus}}
\stick{\i{Senonesis de Partu difficili} \fnast, published by}
\stick{\hfill\i{Adri-}}

\vskip 4pt

\bgroup\fontsize{8}{10.5}\selectfont\noindent
\stick{\quad\fnast\enspace The author is here twice mistaken;\tsh for}
\stick{\i{Lithopædus} should be wrote thus, \i{Lithopædii Seno}-}
\stick{\i{nensis Icon}. The second mistake is, that this \i{Li}-}
\stick{\i{thopædus} is not an author, but a drawing of a}
\stick{petrified child. The account of this, published}
\stick{by \i{Albosius}, 1580, may be seen at the end of}
\stick{\i{Cordæus}’s works in \i{Spachius}. Mr.\,\i{Tristram Shandy}}
\stick{has been led into this error, either from seeing}
\stick{\i{Lithopædus}’s name of late in a catalogue of lear-}
\stick{ned writers in \hbox to 36pt{Dr.\@ \tsfill}, or by mistaking \i{Li}-}
\stick{\i{thopædus} for \i{Trinecavellius},\tsk from the too great}
similitude of the names.\par\egroup

\vfill
\eject
\noindent
\stick{\i{Adrianus Smelvgot}, had found out, That}
\stick{the lax and pliable state of a child’s}
\stick{head in parturition, the bones of the}
\stick{cranium having no sutures at that time,}
\stick{was such,\tsk that by force of the woman’s}
\stick{efforts, which, in strong labour-pains,}
\stick{was equal, upon an average, to a weight}
\stick{of 470 pounds averdupoise acting per-}
\stick{pendicularly upon it;\tsk it so happened,}
\stick{that, in 49 instances out of 50, the said}
\stick{head was compressed and moulded into}
\stick{the shape of an oblong conical piece of}
\stick{dough, such as a pastry-cook generally}
\stick{rolls up in order to make a pye of.\tsh}
\stick{Good God!  cried my father, what ha-}
\stick{vock and destruction must this make in}
\stick{the infinitely fine and tender texture of}
\stick{the cerebellum!\tsk Or if there is such a}
\stick{juice as \i{Borri} pretends,\tsk is it not enough}
\stick{to make the clearest liquid in the world}
\stick{both feculent and mothery? \hfill}
\catch{But}
\stick{\indent But how great was his apprehension,}
\stick{when he further understood, that this}
\stick{force, acting upon the very vertex of}
\stick{the head, not only injured the brain itself}
\stick{or cerebrum,\tsk but that it necessarily}
\stick{squeez’d and propell’d the cerebrum to-}
\stick{wards the cerebellum, which was the im-}
\stick{mediate seat of the understanding.\tsh}
\stick{Angels and Ministers of grace defend us!}
\stick{cried my father,\tsk can any soul withstand}
\stick{this shock?\tsk No wonder the intellectual}
\stick{web is so rent and tatter’d as we see it;}
\stick{and that so many of our best heads are no}
\stick{better than a puzzled skein of silk,\tsk all}
perplexity,\tsk all confusion within side.

But when my father read on, and was\break
\stick{let into the secret, that when a child was}
\stick{turn’d topsy-turvy, which was easy for}
\stick{an operator to do, and was extracted by}
\stick{the feet;\tsk that instead of the cerebrum}
\catch{being}
\stick{being propell’d towards the cerebellum,}
\stick{the cerebellum, on the contrary, was pro-}
\stck{pell’d simply towards the cerebrum where}
\stick{it could do no manner of hurt:\tsk By}
\stick{heavens! cried he, the world is in con-}
\stick{spiracy to drive out what little wit God}
\stick{has given us,\tsk and the professors of the}
\stick{obstetric art are listed into the same con-}
\stick{spiracy.\tsk What is it to me which end of}
\stick{my son comes foremost into the world,}
\stick{provided all goes right after, and his ce-}
rebellum escapes uncrushed?

It is the nature of an hypothesis, when\break
\stick{once a man has conceived it, that it assi-}
\stick{milates every thing to itself as proper}
\stick{nourishment; and, from the first moment}
\stick{of your begetting it, it generally grows}
\stick{the stronger by every thing you see, hear,}
\stick{read, or understand. This is of great}
\stick{use.\hfill}
\catch{When}
\stick{\indent When my father was gone with this}
\stick{about a month, there was scarce a phæ-}
\stick{nomenon of stupidity or of genius, which}
\stick{he could not readily solve by it;\tsk it ac-}
\stck{counted for the eldest son being the great-}
\stick{est blockhead in the family.\tsk Poor devil,}
\stck{he would say,\tsk he made way for the capa-}
\stck{city of his younger brothers.\tsk It unriddled}
\stk{the observation of drivellers and monstrous}
\stick{heads,\tsk shewing, \i{à priori}, it could not be}
\stck{otherwise,-\kern.5pt-unless\enspace\wastiv\ I don’t know}
\stick{what. It wonderfully explain’d and ac-}
\stick{counted for the acumen of the \i{Asiatic} ge-}
\stick{nius, and that sprightlier turn, and a}
\stick{more penetrating intuition of minds, in}
\stick{warmer climates; not from the loose and}
\stick{common-place solution of a clearer sky,}
\stick{and a more perpetual sun-shine, \etc\!\tsk}
\stick{which, for aught he knew, might as well}
\stick{rarify and dilute the faculties of the soul}
\stick{into nothing, by one extreme,\tsk as they}
\catch{are}
\stick{are condensed in colder climates by the}
\stick{other;\tsh but he traced the affair up to its}
\stick{spring-head;\tsk shew’d that, in warmer}
\stick{climates, nature had laid a lighter tax}
\stick{upon the fairest parts of the creation;\tsk}
\stick{their pleasures more;\tsk the necessity of}
\stick{their pains less, insomuch that the pref-\sic}
\stick{sure and resistance upon the vertex was}
\stick{so slight, that the whole organization of}
\stick{the cerebellum was preserved;\tsh nay, he}
\stick{did not believe, in natural births, that so}
\stick{much as a single thread of the net-work}
\stick{was broke or displaced,\tsk so that the soul}
might just act as she liked.

When my father had got so far,\tsk\break
\stick{what a blaze of light did the accounts of}
\stick{the \i{Caesarian} section, and of the towering}
\stick{geniuses who had come safe into the}
\stick{world by it, cast upon this hypothesis?}
\stick{Here you see, he would say, there was no}
\stick{injury done to the sensorium;\tsk no pres-}
\catch{sure}
\stick{sure of the head against the pelvis;\tsk no}
\stick{propulsion of the cerebrum towards the}
\stick{cerebellum, either by the \i{os pubis} on this}
\stick{side, or the \i{os coxcygis} on that;\tsk and,}
\stck{pray, what were the happy consequences?}
\stick{Why, Sir, your \i{Julius Caesar}, who gave}
\stick{the operation a name;\tsk and your \i{Her}-}
\stick{\i{mes Trismegistus}, who was born so before}
\stick{ever the operation had a name;\tsk your}
\stick{\i{Scipio Africanus}; your \i{Manlius Torquatus};}
\stick{our \i{Edward} the Sixth,\tsk who, had he lived,}
\stick{would have done the same honour to the}
\stck{hypothesis:\tsk These, and many more, who}
\stick{figur’d high in the annals of fame,\tsk all}
came \i{side-way}, Sir, into the world.

The incision of the \i{abdomen} and \i{ute}-\break
\stick{\i{rus}, ran for six weeks together in my fa-}
\stick{ther’s head;\tsk he had read, and was satis-}
\stick{fied, that wounds in the \i{epigastrium}, and}
\stick{those in the \i{matrix}, were not mortal;\tsk}
\stick{so that the belly of the mother might be}
\catch{opened}
\stick{opened extremely well to give a passage}
\stick{to the child.\tsk He mentioned the thing}
\stick{one afternoon to my mother,\tsk merely as}
\stick{a matter of fact;\tsk but seeing her turn as}
\stick{pale as ashes at the very mention of it,}
\stick{as much as the operation flattered his}
\stick{hopes,\tsk he thought it as well to say no}
\stick{more of it,\tsk contenting himself with ad-}
\stick{miring\tsk what he thought was to no pur-}
pose to propose.

\tight{This was my father Mr.\@ \i{Shandy}’s hypo-}\break
\stick{thesis; concerning which I have only to}
\stick{add, that my brother \i{Bobby} did as great}
\stick{honour to it (whatever he did to the fa-}
\stick{mily) as any one of the great heroes we}
\stick{spoke of:\tsk For happening not only to be}
\stick{christen’d, as I told you, but to be born}
\stick{too, when my father was at \i{Epsom},\tsk}
\stick{being moreover my mother’s \i{first} child,}
\stick{\tsk coming into the world with his head}
\stick{\i{foremost},\tsk and turning out afterwards a}
\catch{lad}
\stick{lad of wonderful slow parts,\tsh my father}
\stick{spelt all these together into his opinion;}
\stick{and as he had failed at one end,\tsk he was}
determined to try the other.

This was not to be expected from one\break
\stick{of the sisterhood, who are not easily to be}
\stick{put out of their way,\tsh and was therefore}
\stick{one of my father’s great reasons in favour}
\stick{of a man of science, whom he could bet-}
ter deal with.

Of all men in the world, \drslop was\break
\stick{the fittest for my father’s purpose;\tsh for}
\stick{though his new invented forceps was}
\stick{the armour he had proved, and what he}
\stick{maintained, to be the safest instrument of}
\stick{deliverance,\tsk yet, it seems, he had scat-}
\stick{tered a word or two in his book, in fa-}
\stick{vour of the very thing which ran in my}
\stick{father’s fancy;\tsh tho’ not with a view to}
\stick{the soul’s good in extracting by the feet,}
\catch{as}
\stick{as was my father’s system,\tsk but for rea-}
sons merely obstetrical.

This will account for the coalition be-\break
\stick{twixt my father and \drslop, in the en-}
\stick{suing discourse, which went a little hard}
\stick{against my uncle \toby.\tsk In what manner}
\stick{a plain man, with nothing but common}
\stick{sense, could bear up against two such}
\stick{allies in science,\tsk is hard to conceive.\tsk}
\stick{You may conjecture upon it, if you please,}
\stick{\tsk and whilst your imagination is in mo-}
\stick{tion, you may encourage it to go on, and}
\stick{discover by what causes and effects in}
\stick{nature it could come to pass, that my}
\stick{uncle \toby got his modesty by the wound}
\stick{he received upon his groin.\tsk You may}
\stick{raise a system to account for the loss of}
\stick{my nose by marriage articles,\tsk and shew}
\stick{the world how it could happen, that I}
\stick{should have the misfortune to be called}
\stick{\s{Tristram}, in opposition to my father’s}
\catch{hy-}
\stick{hypothesis, and the wish of the whole}
\stick{family, God-fathers and God-mothers}
\stick{not excepted.\tsk These, with fifty other}
\stick{points left yet unraveled, you may en-}
\stick{deavour to solve if you have time;\tsh but}
\stick{I tell you before-hand it will be in vain,}
\stick{\tsk for not the sage \i{Alquise},\sic the magician}
\stick{in Don \i{Belianis} of \i{Greece}, nor the no less}
\stick{famous \i{Urganda}, the sorceress his wife,}
\stick{(were they alive) could pretend to come}
within a league of the truth.

The reader will be content to wait for\break
\stick{a full explanation of these matters till}
\stick{the next year,\tsk when a series of things}
will be laid open which he little expects.

\vskip 25pt

\noindent
\centerline{\i{\lsss{END}\enspace of\enspace the}\enspace \s{Second\enspace Volume}.}
\vfill
\eject
\pagestyle{empty}
\null
\end{document}
