\documentclass[twoside]{article}
\usepackage[paperheight=395pt, paperwidth=240pt, width=166pt, height=308pt, top=42pt, headsep=2pt]{geometry}
\usepackage[noinfo, frame, a6, center, lualatex]{crop}
\usepackage{graphicx}
\usepackage{shandean}
\usepackage{luamplib}
\mplibtextextlabel{enable}
\title{The Life and Opinions of Tristram Shandy, Gentleman}
\author{Laurence Sterne}
\date{1762}
\begin{document}
\pagestyle{empty}
% half title
\null
\vfill
\noindent
\centerline{\lsv{50}{THE}}
\centerline{\tstrut 32pt\large \ls{LIFE}\enspace and\enspace \ls{OPINIONS}}
\centerline{\tstrut 27pt\lsv{80}{OF}}
\centerline{\tstrut 26pt\large \lss{TRISTRAM SHANDY}, Gent.}
\vfill
\eject
\null
\vfill
\eject
 % full title
\noindent
\centerline{\fontsize{10}{14}\selectfont\lsv{100}{THE}}
\centerline{\tstrut 26pt\fontsize{20}{28}\selectfont\lsv{150}{LIFE}}
\centerline{\tstrut 18pt\fontsize{10}{14}\selectfont\lsv{80}{AND}}
\centerline{\tstrut 26pt\fontsize{20}{28}\selectfont\lsv{55}{OPINIONS}}
\centerline{\tstrut 16pt\fontsize{08}{14}\selectfont\lsv{80}{OF}}
\centerline{\tstrut 22pt\fontsize{14}{20}\selectfont\lsv{20}{TRISTRAM}\enspace\lsv{20}{SHANDY},}
\centerline{\tstrut 14pt\fontsize{12}{14}\selectfont\lsv{16}{\textsc{Gentleman}}.}

\vskip 16pt

\bgroup\fontsize{8}{10}\selectfont\noindent
\stick{\enspace\i{Dixero si quid fortè jocosius, hoc mihi juris}\hfill}
\stick{\enspace\i{Cum venia dabis}.\hfill \textsc{Hor.}\enspace}
\stick{\tstrut 13pt\enspace\hbox to 2em{\tsfill}\i{Si quis calumnietur levius esse quam decet}\enspace}
\stick{\enspace\quad\i{theologum, aut mordacius quam deceat Chris-}\enspace}
\stick{\enspace\quad\i{tianum\tsfill non Ego, sed Democritus dixit}.\tsfill\enspace}
\stick{\hfill\s{Erasmus}.}
\par\egroup

\vskip 14pt

\centerline{\lsss{VOL}.\quad VI.}

\vskip 14pt

\centerline{\small \lsv{120}{LONDON}:}
\centerline{\small Printed for T. \s{Becket} and P. A. \s{Dehondt},}
\centerline{\small in the Strand.\quad MDCCLXII.}
\eject
\hbox{}
\vfill
\eject
\baselineskip 14pt  % 22*14 = 308
\parskip 12pt
\setcounter{page}{1}
\pagestyle{fancy}
\thispagestyle{empty}
\noindent
\centerline{\lsv{50}{THE}}
\centerline{\tstrut 26pt\large \lsss{LIFE}\enspace and\enspace \lsss{OPINIONS}}
\centerline{\tstrut 26pt\lsv{80}{OF}}
\centerline{\tstrut 26pt\large \lss{TRISTRAM SHANDY}, Gent.}

\vskip 20pt

\hrule

\chap{I}{18pt}{6pt}

\dropcap[nindent=1pt, ante=\lower6pt\hbox{\tsh}]{W}{e’ll} not stop two mo-\break
\tight{ments, my dear Sir,\tsk only,}\break
\stick{as we have got thro’ these five volumes,}
\stick{(do, Sir, sit down upon a set\tshh they}
\stick{are better than nothing) let us just look}
\stick{back upon the country we have pass’d}
through.\tush

\tsh What a wilderness has it been!\break
\stick{and what a mercy that we have not both}
\catch{of}
\stick{of us been lost, or devoured by wild}
beasts in it.

Did you think the world itself, Sir,\break
\stick{had contained such a number of Jack}
\stick{Asses?\tshh How they view’d and re-}
\stick{view’d us as we passed over the rivulet}
\stick{at the bottom of that little valley!\tsh}
\stick{and when we climbed over that hill, and}
\stick{were just getting out of sight\tsk good}
\stick{God! what a braying did they all set up}
together!

\tshh Prithee, shepherd! who keeps\break
all those Jack Asses? \raise2pt\hbox{\enspace\ast\enspace\ast\enspace\ast}

\tsh Heaven be their comforter\tsh\break
\stick{What! are they never curried?\tsk Are}
\stick{they never taken in in winter?\tsk Bray}
\stick{bray\tsk bray. Bray on,\tsk the world is}
\stick{deeply your debtor;\tsh louder still\tsk}
\catch{that’s}
\stick{that’s nothing:\tsk in good sooth, you}
\stick{are ill-used:\tsk Was I a Jack Asse, I}
\stick{solemnly declare, I would bray in}
\stick{G-sol-re-ut from morning, even unto}
night.

\chap{II}{12pt}{0pt}

\dropcap[nindent=0pt]{W}{hen} my father had danced his\break
\tighter{white bear backwards and forwards}\break
\stick{through half a dozen pages, he closed}
\stick{the book for good an’ all,\tsk and in a}
\stick{kind of triumph redelivered it into \trim's}
\stick{hand, with a nod to lay it upon the ’scru-}
\stick{toire where he found it.\tsk \i{Tristram},}
\stick{said he, shall be made to conjugate every}
\stick{word in the dictionary, backwards and}
\stick{forwards the same way;\tsh every word,}
\stick{\i{Yorick}, by this means, you see, is con-}
\stick{verted into a thesis or an hypothesis;\tsk}
\stick{every thesis and hypothesis have an off-}
\catch{spring}
\stick{spring of propositions;\tsk and each pro-}
\stick{position has its own consequences and}
\stick{conclusions; every one of which leads}
\stick{the mind on again, into fresh tracks of}
\stick{enquiries and doubtings.\tsh The force}
\stick{of this engine, added my father, is in-}
\stick{credible, in opening a child’s head.\tsh}
\stick{’Tis enough, brother \i{Shandy}, cried my}
\stick{uncle \toby, to burst it into a thousand}
splinters.\tsh

I presume, said \i{Yorick}, smiling,\tsk it\break
\stick{must be owing to this,\tsk (for let logi-}
\stick{cians say what they will, it is not to be}
\stick{accounted for sufficiently from the bare}
\stick{use of the ten predicaments)\tsk That}
\stick{the famous \i{Vincent Quirino}, amongst the}
\stick{many other astonishing feats of his child-}
\stick{hood, of which the Cardinal \i{Bembo} has}
\stick{given the world so exact a story,\tsk should}
\stick{be able to paste up in the publick schools}
\catch{at}
\stick{at \i{Rome}, so early as in the eighth year of}
\stick{his age, no less than four thousand, five}
\stick{hundred, and sixty different theses, upon}
\stick{the most abstruse points of the most ab-}
\stick{struse theology;\tsk and to defend and}
\stick{maintain them in such sort, as to cramp}
\stick{and dumbfound his opponents.\tsk}
\stick{What is that, cried my father, to what}
\stick{is told us of \i{Alphonsus Tostatus}, who,}
\stick{almost in his nurse’s arms, learned all}
\stick{the sciences and liberal arts without be-}
\stick{ing taught any one of them?\tsh What}
\stick{shall we say of the great \i{Piereskius?}\tsk}
\stick{That’s the very man, cried my uncle}
\stick{\toby, I once told you of, brother \i{Shandy},}
\stick{who walked a matter of five hundred}
\stick{miles, reckoning from \i{Paris} to \i{Schevling},}
\stick{and from \i{Schevling} back again, merely}
\stick{to see \i{Stevinus}’s flying chariot.\tsh He}
\stick{was a very great man! added my uncle}
\stick{\toby; (meaning \i{Stevinus})\tsk He was so;}
\catch{bro-}
\stick{brother \toby, said my father, (meaning}
\stick{\i{Piereskius})\tsk and had multiplied his}
\stick{ideas so fast, and increased his know-}
\stick{lege\sic to such a prodigious stock, that,}
\stick{if we may give credit to an anecdote}
\stick{concerning him, which we cannot with-}
\stck{hold here, without shaking the authority}
\stck{of all anecdotes whatever\tsk at seven years}
\stick{of age, his father committed entirely to}
\stick{his care the education of his younger bro-}
\stck{ther, a boy of five years old,\tsk with the sole}
\stck{management of all his concerns.\tsk Was the}
\stck{father as wise as the son? quoth my uncle}
\stck{\i{Toby:}\tsk I should think not, said \i{Yorick:}\tsk}
\stck{But what are these, continued my father\tsk}
\stick{(breaking out in a kind of enthusiasm)}
\stick{\tsk what are these, to those prodigies of}
\stick{childhood in \i{Grotius}, \i{Scioppius}, \i{Heinsius},}
\stick{\i{Politian}, \i{Pascal}, \i{Joseph Scaliger}, \i{Fer}-}
\stick{\i{dinand de Cordouè}, and others\tsk some of}
\stick{which left off their \i{substantial forms} at}
\catch{nine}
\stick{nine years old, or sooner, and went on}
\stick{reasoning without them;\tsk others went}
\stick{through their classics at seven;\tsk wrote}
\stick{tragedies at eight;\tskk \i{Ferdinand de Cor-}}
\stick{\i{douè} was so wise at nine,\tsk ’twas thought}
\stick{the Devil was in him;\tsh and at \i{Venice}}
\stick{gave such proofs of his knowledge and}
\stick{goodness, that the monks imagined he}
\stick{was \i{Antichrist}, or nothing.\tshh Others}
\stick{were masters of fourteen languages at}
\stick{ten,\tsk finished the course of their rheto-}
\stick{ric, poetry, logic, and ethics at eleven,}
\stick{\tskk put forth their commentaries upon}
\stick{\i{Servius} and \i{Martianus Capella} at twelve,}
\stick{\tskk and at thirteen received their degrees}
\stick{in philosophy, laws, and divinity:\tsh}
\stick{But you forget the great \i{Lipsius}, quoth}
\stick{\i{Yorick}, who composed a work \fnast\ the day}
\stick{\hfill he}

\noindent\bgroup\fontsize{8}{10}\selectfont
\stick{\enspace\fnast\ Nous aurions quelque interêt, says \i{Baillet}, de}
\stick{montrer qu’il n’ a rien de ridicule s’il étoit vérita-}
\stick{\hfill ble,}
\par\egroup\eject\noindent
\stick{he was born;\tsh They should have}
\stick{wiped it up, said my uncle \toby, and}
said no more about it.

\chap{III}{16pt}{0pt}

\dropcap[findent=2pt, nindent=0pt]{W}{hen} the cataplasm was ready, a\break
\tight{scruple of \i{decorum} had unseasona-}\break
\stck{bly rose up in \susannah's conscience, about}
\stick{holding the candle, whilst \slop tied it on;}
\stick{\slop had not treated \susannah's distem-}
\stick{per with anodines,\tsk and so a quarrel}
had ensued betwixt them.

\noindent\bgroup\fontsize{8}{10}\selectfont
\stick{ble, au moins dans le sens énigmatique que \i{Nicius}}
\stick{\i{Erythræus} a tâché de lui donner. Cet auteur dit}
\stick{que pour comprendre comme \i{Lipse}, a pû com-}
\stick{poser un ouvrage le premier jour de sa vie, il faut}
\stick{s’imaginer, que ce premier jour n’est pas celui de}
\stick{sa naissance charnelle, mais celui au quel il a com-}
\stick{mencé d’user de la raison; il veut que ç’ait été a}
\stick{l’age de \i{neuf} ans; et il nous veut persuader que ce}
\stick{fut en cet âge, que \i{Lipse} fit un poem.\tsh Le tour}
est ingenieux, \&c. \&c.\par\egroup
\patch{\tsh Oh!}

\topstrut\tsh Oh! oh!\tsh said \slop, casting\break
\stick{a glance of undue freedom in \susannah's}
\stick{face, as she declined the office;\tsk then,}
\stick{I think I know you, madam\tsk You}
\stick{know me, Sir! cried \susannah fastidi-}
\stick{ously, and with a toss of her head, le-}
\stick{velled evidently, not at his profession,}
\stick{but at the doctor himself,\tsh you know}
\stick{me! cried \susannah again.\tsk Doctor}
\stick{\slop clapped his finger and his thumb}
\stick{instantly upon his nostrils;\tsk \i{Susan}-}
\stick{\i{nah}’s spleen was ready to burst at it;\tsh}
\stick{’Tis false, said \susannah.\tsk Come, come,}
\stick{Mrs.\@ Modesty, said \slop, not a little}
\stick{elated with the success of his last thrust,}
\stick{\tsk if you won’t hold the candle, and}
\stick{look\tsk you may hold it and shut your}
\stick{eyes:\tsk That’s one of your popish shifts,}
\stick{cried \i{Susannah:}\tsk ’Tis better, said \slop,}
\stick{with a nod, than no shift at all, young}
\catch{woman;}
\parskip 6pt
\stick{woman;\tsh I defy you, Sir, cried \i{Su}-}
\stick{\i{sannah}, pulling her shift sleeve below}
her elbow.

It was almost impossible for two per-\break
\stick{sons to assist each other in a surgical}
case with a more splenetic cordiality.

\slop snatched up the cataplasm,\tsh\break
\stick{\susannah snatched up the candle;\tsh}
\stick{A little this way, said \slop; \susannah}
\stick{looking one way, and rowing another,}
\stick{instantly set fire to \slop's wig, which}
\stick{being somewhat bushy and unctuous}
\stick{withal, was burnt out before it was well}
\stick{kindled.\tsk You impudent whore!}
\stick{cried \slop,\tsk (for what is passion, but a}
\stick{wild beast)\tsk you impudent whore, cried}
\stick{\slop, getting upright, with the cataplasm}
\stick{in his hand;\tsk I never was the de-}
\stick{struction of any body’s nose, said \i{Susan-}}
\stick{\i{nah},\tsk which is more than you can say:}
\catch{\tsh Is}
\stick{\tsh Is it?  cried \slop, throwing the ca-}
\stick{taplasm in her face;\tsh Yes, it is, cried}
\stick{\susannah, returning the compliment}
with what was left in the pan.\tsh

\chap{IV}{14pt}{0pt}

\dropcap[findent=0pt, nindent=2pt]{D}{octor} \slop and \susannah filed\break
cross-bills against each other in the\break
\stick{parlour; which done, as the cataplasm}
\stick{had failed, they retired into the kitchen}
\stick{to prepare a fomentation for me;\tsk and}
\stick{whilst that was doing, my father deter-}
mined the point as you will read.

\chap{V}{14pt}{0pt}

\dropcap{Y}{ou} \tight{see ’tis high time, said my father,}\break
\tighter{addressing himself equally to my un-}\break
\stick{cle \toby and \i{Yorick}, to take this young}
\stick{creature out of these women’s hands,}
\stick{and put him into those of a private go-}
\stick{vernor. \i{Marcus Antoninus} provided four-}
\catch{teen}
\parskip 12pt
\stick{teen governors all at once to superintend}
\stick{his son \i{Commodus}’s education,\tsk and in}
\stick{six weeks he cashiered five of them;\tsk}
\stick{I know very well, continued my father,}
\stick{that \i{Commodus}’s mother was in love with}
\stick{a gladiator at the time of her concep-}
\stick{tion, which accounts for a great many}
\stick{of \i{Commodus}’s cruelties when he became}
\stick{emperor;\tsk but still I am of opinion,}
\stick{that those five whom \i{Antoninus} dismiss-}
\stick{ed, did \i{Commodus}’s temper in that short}
\stick{time, more hurt than the other nine}
were able to rectify all their lives long.

Now as I consider the person who is\break
\stick{to be about my son, as the mirror in}
\stick{which he is to view himself from morn-}
\stick{ing to night, and by which he is to adjust}
\stick{his looks, his carriage, and perhaps the}
\stick{inmost sentiments of his heart;\tsk I would}
\stick{have one, \i{Yorick}, if possible, polished at}
\catch{all}
\stick{all points, fit for my child to look into.}
\stick{\tsh This is very good sense, quoth my}
uncle \toby to himself.

\tsh There is, continued my father,\break
\stick{a certain mien and motion of the body}
\stick{and all its parts, both in acting and}
\stick{speaking, which argues a man \i{well}}
\stick{\i{within}; and I am not at all surprized}
\stick{that \i{Gregory} of \i{Nazianzum}, upon ob-}
\stick{serving the hasty and untoward gestures}
\stick{of \i{Julian}, should foretel he would one}
\stick{day become an apostate;\tsh or that St.}
\stick{\i{Ambrose} should turn his \i{Amanuensis} out}
\stick{of doors, because of an indecent mo-}
\stick{tion of his head, which went backwards}
\stick{and forwards like a flail;\tsk or that}
\stick{\i{Democritus} should conceive \i{Protagoras}}
\stick{to be a scholar, from seeing him bind up}
\stick{a faggot, and thrusting, as he did it,}
\stick{the small twigs inwards.\tsk There are a}
\catch{thou-}
\stick{thousand unnoticed openings, continued}
\stick{my father, which let a penetrating eye}
\stick{at once into a man’s soul; and I main-}
\stick{tain it, added he, that a man of sense}
\stick{does not lay down his hat in coming}
\stick{into a room,\tsk or take it up in going out}
\stick{of it, but something escapes, which dis-}
covers him.

It is for these reasons, continued my\break
\stick{father, that the governor I make choice}
\stick{of shall neither \fnast\ lisp, or squint, or wink,}
\stick{or talk loud, or look fierce, or foolish;}
\stick{\tsh or bite his lips, or grind his teeth,}
\stick{or speak through his nose, or pick it,}
or blow it with his fingers.\tsh

He shall neither walk fast,\tsk or slow,\break
\stick{or fold his arms,\tsk for that is laziness; \tsk}
\stick{or hang them down,\tsk for that is folly;}
\centerline{\footnotesize\fnast\enspace Vid. \i{Pellegrina}.}
\catch{or}
\stick{or hide them in his pocket, for that is}
nonsense.\tsh

He shall neither strike, or pinch, or\break
\stick{tickle,\tsk or bite, or cut his nails, or}
\stick{hawk, or spit, or snift, or drum with his}
\stick{feet or fingers in company;\tsh nor (ac-}
\stick{cording to \i{Erasmus)}\sic shall he speak to}
\stick{any one in making water,\tsk nor shall he}
\stick{point to carrion or excrement.\tsh Now}
\stick{this is all nonsense again, quoth my}
uncle \toby to himself.\tsh

I will have him, continued my father,\break
\stick{cheerful, faceté, jovial; at the same}
\stick{time, prudent, attentive to business, vi-}
\stick{gilant, acute, argute, inventive, quick}
\stick{in resolving doubts and speculative ques-}
\stick{tions;\tsh he shall be wise and judicious,}
\stick{and learned:\tsh And why not humble,}
\stick{and moderate, and gentle tempered, and}
\catch{good?}
\stick{good? said \i{Yorick:}\tsh And why not,}
\stick{cried my uncle \toby, free, and generous,}
\stick{and bountiful, and brave?\tsh He shall,}
\stick{my dear \toby, replied my father, get-}
\stick{ting up and shaking him by his hand.\tsk}
\stick{Then, brother \i{Shandy}, answered my un-}
\stick{cle \toby, raising himself off the chair, and}
\stick{laying down his pipe to take hold of my}
\stick{father’s other hand,\tsk I humbly beg I}
\stick{may recommend poor \i{Le Fever}’s son to}
\stick{you;\tsh a tear of joy of the first water}
\stick{sparkled in my uncle \toby's eye,\tsk and}
\stick{another, the fellow to it, in the corpo-}
\stick{ral’s, as the proposition was made;\tsh}
\stick{you will see why when you read \i{Le Fe}-}
\stick{\i{ver}’s story:\tsh fool that I was! nor}
\stick{can I recollect, (nor perhaps you) with-}
\stick{out turning back to the place, what it}
\stick{was that hindered me from letting the}
\stick{corporal tell it in his own words;\tsk but}
\catch{the}
\stick{the occasion is lost,\tsk I must tell it now}
in my own.

\chap{VI}{24pt}{4pt}

\centerline{\tstrut18pt\hbox to 100pt{\i{The Story of} \s{Le Fever}}.}

\dropcap{I}{t} was some time in the summer of\break
that year in which \i{Dendermond} was\break
\stick{taken by the allies,\tsk which was about}
\stick{seven years before my father came into}
\stick{the country,\tsk and about as many, after}
\stick{the time, that my uncle \toby and \trim}
\stick{had privately decamped from my father’s}
\stick{house in town, in order to lay some of}
\stick{the finest sieges to some of the finest}
\stick{fortified cities in \i{Europe}\tsk when my}
\stick{uncle \toby was one evening getting his}
\stick{supper, with \trim sitting behind him at}
\stick{a small sideboard,\tsk I say, sitting\tsk for}
\catch{in}
\stick{in consideration of the corporal’s lame}
\stick{knee (which sometimes gave him exqui-}
\stick{site pain)\tsk when my uncle \toby dined or}
\stick{supped alone, he would never suffer the}
\stick{corporal to stand; and the poor fellow’s}
\stick{veneration for his master was such, that,}
\stick{with a proper artillery, my uncle \toby}
\stick{could have taken \i{Dendermond} itself,}
\stick{with less trouble than he was able to}
\stick{gain this point over him; for many a}
\stick{time when my uncle \toby supposed the}
\stick{corporal’s leg was at rest, he would look}
\stick{back, and detect him standing behind}
\stick{him with the most dutiful respect: this}
\stick{bred more little squabbles betwixt them,}
\stick{than all other causes for five and twenty}
\stick{years together\tsk But this is neither here}
\stick{nor there\tsk why do I mention it?\tsh}
\stick{Ask my pen,\tsk it governs me,\tsk I govern}
not it.
\patch{He}

He was one evening sitting thus at\break
\stick{his supper, when the landlord of a little}
\stick{inn in the village came into the parlour}
\stick{with an empty phial in his hand, to beg}
\stick{a glass or two of sack; ’Tis for a poor}
\stick{gentleman,\tsk I think, of the army, said}
\stick{the landlord, who has been taken ill at}
\stick{my house four days ago, and has never}
\stick{held up his head since, or had a desire}
\stick{to taste any thing, till just now, that}
\stick{he has a fancy for a glass of sack and a}
\stick{thin toast,\tsk \i{I~think}, says he, taking}
\stick{his hand from his forehead, \i{it would}}
\i{comfort me.}\tush

\tshh If I could neither\sic\ beg, borrow,\break
\stick{or buy such a thing,\tsk added the land-}
\stick{lord,\tsk I~would almost steal it for the}
\stick{poor gentleman, he is so ill.\tsh I hope}
\stick{in God he will still mend, continued he,}
\tsk we are all of us concerned for him.
\patch{Thou}

\tstrut 22pt Thou art a good natured soul, I will\break
\stick{answer for thee, cried my uncle \toby;}
\stick{and thou shalt drink the poor gentle-}
\stick{man’s health in a glass of sack thyself,}
\stick{\tsk and take a couple of bottles with my}
\stick{service, and tell him he is heartily wel-}
\stick{come to them, and to a dozen more if}
they will do him good.

Though I am persuaded, said my un-\break
\stick{cle \toby, as the landlord shut the door,}
\stick{he is a very compassionate fellow\tsk \trim,}
\stick{\tsk yet I cannot help entertaining a high}
\stick{opinion of his guest too; there must be}
\stick{something more than common in him,}
\stick{that in so short a time should win so}
\stick{much upon the affections of his host;}
\stick{\tsk And of his whole family, added}
\stick{the corporal, for they are all concerned}
\stick{for him.\tsk Step after him, said my}
\catch{uncle\enspace\sic}
\stick{uncle \toby,\tsk do \trim,\tsk and ask if he}
knows his name.

\tsh I have quite forgot it, truly,\break
\stick{said the landlord, coming back into the}
\stick{parlour with the corporal,\tsk but I can}
\stick{ask his son again:\tsh Has he a son with}
\stick{him then? said my uncle \toby.\tsk A boy,}
\stick{replied the landlord, of about eleven or}
\stick{twelve years of age;\tsk but the poor crea-}
\stick{ture has tasted almost as little as his fa-}
\stick{ther; he does nothing but mourn and}
\stick{lament for him night and day:\tsh He}
\stick{has not stirred from the bedside these}
two days.

My uncle \toby laid down his knife\break
\stick{and fork, and thrust his plate from be-}
\stick{fore him, as the landlord gave him the}
\stick{account; and \trim, without being or-}
\stick{dered, took away without saying one}
\catch{word,}
\stick{word, and in a few minutes after brought}
him his pipe and tobacco.

\tshh Stay in the room a little, said\break
my uncle \toby.\tsh

\i{Trim!}\tshh said my uncle \toby, after\break
\stick{he lighted his pipe, and smoak’d about}
\stick{a dozen whiffs.\tsk \trim came in front}
\stick{of his master, and made his bow;\tsk my}
\stick{uncle \toby smoak’d on, and said no}
\stick{more.\tsk Corporal! said my uncle \toby}
\stick{\tsk the corporal made his bow.\tsk}
\stick{My uncle \toby proceeded no farther, but}
finished his pipe.

\i{Trim!} said my uncle \toby, I have a\break
\stick{project in my head, as it is a bad night,}
\stick{of wrapping myself up warm in my ro-}
\stick{quelaure, and paying a visit to this poor}
\stick{gentleman.\tsk Your honour’s roque-}
\catch{laure,}
\stick{laure, replied the corporal, has not}
\stick{once been had on, since the night before}
\stick{your honour received your wound, when}
\stick{we mounted guard in the trenches be-}
\stick{fore the gate of St.\@ \i{Nicholas};\tsk and}
\stick{besides it is so cold and rainy a night,}
\stick{that what with the roquelaure, and what}
\stick{with the weather, ’twill be enough to}
\stick{give your honour your death, and bring}
\stick{on your honour’s torment in your groin.}
\stick{I fear so; replied my uncle \toby, but I}
\stick{am not at rest in my mind, \trim, since}
\stick{the account the landlord has given me.}
\stick{\tsh I wish I had not known so much of}
\stick{this affair,\tsk added my uncle \toby,\tsk or}
\stick{that I had known more of it:\tsh How}
\stick{shall we manage it?  Leave it, an’t}
\stick{please your honour, to me, quoth the}
\stick{corporal;\tsh I’ll take my hat and stick}
\stick{and go to the house and reconnoitre, and}
\stick{act accordingly; and I will bring your}
\catch{honour}
\stick{honour a full account in an hour.\tsh}
\stick{Thou shalt go, \trim, said my uncle}
\stick{\toby, and here’s a shilling for thee to}
\stick{drink with his servant.\tsh I shall get it}
\stick{all out of him, said the corporal, shut-}
ting the door.

My uncle \toby filled his second pipe;\break
\stick{and had it not been, that he now and}
\stick{then wandered from the point, with}
\stick{considering whether it was not full as}
\stick{well to have the curtain of the tennaile}
\stick{a straight line, as a crooked one,\tsk he}
\stick{might be said to have thought of no-}
\stick{thing else but poor \i{Le Fever} and his boy}
the whole time he smoaked it.
\patch[28pt]{\lsss{CHAP}.}
\parskip 10pt

\chap{VII}{28pt}{4pt}

\centerline{\tstrut18pt\hbox to 144pt{\i{The Story of} \s{Le Fever} \i{continued}}.}

\vskip -4pt

\dropcap{I}{t} was not till my uncle \toby had\break
knocked the ashes out of his third\break
\stick{pipe, that corporal \trim returned from}
\stick{the inn, and gave him the following}
account.

I despaired at first, said the corporal,\break
\stick{of being able to bring back your honour}
\stick{any kind of intelligence concerning the}
\stick{poor sick lieutenant\tsk Is he in the army}
\stick{then?  said my uncle \i{Toby}\tsh He is:}
\stick{said the corporal\tsh And in what re-}
\stick{giment?  said my uncle \i{Toby}\tsh I’ll tell}
\stick{your honour, replied the corporal, every}
\stick{thing straight forwards, as I learnt it.\tsk}
\stick{Then, \trim, I’ll fill another pipe, said}
\stick{my uncle \toby, and not interrupt thee}
\catch{till}
\stick{till thou hast done; so sit down at thy}
\stick{ease, \trim, in the window seat, and be-}
\stick{gin thy story again.  The corporal made}
\stick{his old bow, which generally spoke as}
\stick{plain as a bow could speak it\tsk \i{Your}}
\stick{\i{honour is good:}\tsk And having done}
\stick{that, he sat down, as he was ordered,\tsk}
\stick{and begun the story to my uncle \toby}
\stick{over again in pretty near the same}
words.

I despaired at first, said the corporal,\break
\stick{of being able to bring back any intelli-}
\stick{gence to your honour, about the lieute-}
\stick{nant and his son; for when I asked}
\stick{where his servant was, from whom I}
\stick{made myself sure of knowing every}
\stick{thing which was proper to be asked,\tsk}
\stick{That’s a right distinction, \trim, said my}
\stick{uncle \i{Toby}\tsk I was answered, an’ please}
\stick{your honour, that he had no servant}
\catch{with}
\stick{with him;\tsh that he had come to the}
\stick{inn with hired horses, which, upon}
\stick{finding himself unable to proceed, (to}
\stick{join, I suppose, the regiment) he had}
\stick{dismissed the morning after he came.\tsk}
\stick{If I get better, my dear, said he, as he}
\stick{gave his purse to his son to pay the man,}
\stick{\tsk we can hire horses from hence.\tsh}
\stick{But alas! the poor gentleman will never}
\stick{get from hence, said the landlady to me,}
\stick{\tsk for I heard the death-watch all night}
\stick{long;\tsh and when he dies, the youth,}
\stick{his son, will certainly die with him;}
for he is broken-hearted already.

I was hearing this account, continued\break
\stick{the corporal, when the youth came into}
\stick{the kitchen, to order the thin toast the}
\stick{landlord spoke of;\tsh but I will do it}
\stick{for my father myself, said the youth.}
\stick{\tsk Pray let me save you the trouble,}
\catch{young}
\stick{young gentleman, said I, taking up a}
\stick{fork for the purpose, and offering him}
\stick{my chair to sit down upon by the fire,}
\stick{whilst I did it.\tsh I believe, Sir, said he,}
\stick{very modestly, I can please him best}
\stick{myself.\tsh I am sure, said I, his honour}
\stick{will not like the toast the worse for being}
\stick{toasted by an old soldier.\tsh The youth}
\stick{took hold of my hand, and instantly}
\stick{burst into tears.\tsk Poor youth! said}
\stick{my uncle \toby,\tsk he has been bred up}
\stick{from an infant in the army, and the}
\stick{name of a soldier, \trim, sounded in his}
\stick{ears like the name of a friend;\tsk I wish}
I had him here.

\tshh I never in the longest march,\break
\stick{said the corporal, had so great a mind to}
\stick{my dinner, as I had to cry with him}
\stick{for company:\tsk What could be the mat-}
\stick{ter with me, an’ please your honour?}
\catch{Nothing}
\stick{Nothing in the world, \trim, said my}
\stick{uncle \toby, blowing his nose,\tsk but that}
thou art a good natured fellow.

\vskip 4pt

When I gave him the toast, continued\break
\stick{the corporal, I thought it was proper to}
\stick{tell him I was Captain \i{Shandy}’s servant,}
\stick{and that your honour (though a stran-}
\stick{ger) was extremely concerned for his}
\stick{father;\tsk and that if there was any thing}
\stick{in your house or cellar\tshh (And thou}
\stick{might’st have added my purse too, said}
\stick{my uncle \toby)\tsh he was heartily wel-}
\stick{come to it:\tsk He made a very low}
\stick{bow, (which was meant to your honour)}
\stick{but no answer,\tsk for his heart was full\tsk}
\stick{so he went up stairs with the toast;\tsk I}
\stick{warrant you, my dear, said I, as I open-}
\stick{ed the kitchen door, your father will be}
\stick{well again.\tsk Mr.\@ \i{Yorick}’s curate was}
\catch{smoak-}
\stick{smoaking a pipe by the kitchen fire,\tsk}
\stick{but said not a word good or bad to com-}
\stick{fort the youth.\tsk I thought it wrong;}
\stick{added the corporal\tsk I think so too,}
said my uncle \i{Toby.}

When the lieutenant had taken his\break
\stick{glass of sack and toast, he felt himself}
\stick{a little revived, and sent down into the}
\stick{kitchen, to let me know, that in about}
\stick{ten minutes he should be glad if I}
\stick{would step up stairs.\tsk I believe, said}
\stick{the landlord, he is going to say his}
\stick{prayers,\tsk for there was a book laid}
\stick{upon the chair by his bedside, and as I}
\stick{shut the door, I saw his son take up a}
cushion.\tsh

I thought, said the curate, that you\break
\stick{gentlemen of the army, Mr.\@ \trim, never}
\stick{said your prayers at all.\tsh I heard the}
\catch{poor}
\stick{poor gentleman say his prayers last}
\stick{night, said the landlady, very devoutly,}
\stick{and with my own ears, or I could not}
\stick{have believed it.\tsh Are you sure of it?}
\stick{replied the curate.\tsk A soldier, an’}
\stick{please your reverence, said I, prays as}
\stick{often (of his own accord) as a parson;}
\stick{\tsk and when he is fighting for his}
\stick{king, and for his own life, and for his}
\stick{honour too, he has the most reason to}
\stick{pray to God, of any one in the whole}
\stick{world\tsk ’Twas well said of thee, \trim,}
\stick{said my uncle \i{Toby.}\tsk But when a sol-}
\stick{dier, said I, an’ please your reverence,}
\stick{has been standing for twelve hours toge-}
\stick{ther in the trenches, up to his knees in}
\stick{cold water,\tsk or engaged, said I, for}
\stick{months together in long and dangerous}
\stick{marches;\tsk harassed, perhaps, in his}
\stick{rear to-day;\tsk harassing others to-mor-}
\stick{row;\tsk detached here;\tsk countermanded}
\catch{there;}
\stick{there;\tsk resting this night out upon his}
\stick{arms;\tsk beat up in his shirt the next;\tsk}
\stick{benumbed in his joints;\tsk perhaps with-}
\stick{out straw in his tent to kneel on;\tsk must}
\stick{say his prayers \i{how} and \i{when} he can.\tsk}
\stick{I believe, said I,\tsk for I was piqued,}
\stick{quoth the corporal, for the reputation}
\stick{of the army,\tsk I believe, an’ please your}
\stick{reverence, said I, that when a soldier}
\stick{gets time to pray,\tsk he prays as heartily}
\stick{as a parson,\tsk though not with all his}
\stick{fuss and hypocrisy.\tsh Thou shouldst}
\stick{not have said that, \trim, said my uncle}
\stick{\toby,\tsk for God only knows who is a hy-}
\stick{pocrite, and who is not:\tsk At the}
\stick{great and general review of us all, cor-}
\stick{poral, at the day of judgment, (and}
\stick{not till then)\tsk it will be seen who has}
\stick{done their duties in this world,\tsk and}
\stick{who has not; and we shall be advanced,}
\stick{\trim, accordingly.\tsh I hope we shall,}
\catch{said}
\stick{said \i{Trim.}\tsh It is in the Scripture, said}
\stick{my uncle \toby; and I will shew it thee}
\stick{to-morrow:\tskk In the mean time we}
\stick{may depend upon it, \trim, for our}
\stick{comfort, said my uncle \toby, that God}
\stick{Almighty is so good and just a gover-}
\stick{nor of the world, that if we have but}
\stick{done our duties in it,\tsk it will never be}
\stick{enquired into, whether we have done}
\stick{them in a red coat or a black one:\tshh}
\stick{I hope not; said the corporal\tshh But}
\stick{go on, \trim, said my uncle \toby, with}
thy story.

When I went up, continued the cor-\break
\stick{poral, into the lieutenant’s room, which}
\stick{I did not do till the expiration of the}
\stick{ten minutes,\tsk he was lying in his bed}
\stick{with his head raised upon his hand,}
\stick{with his elbow upon the pillow, and a}
\stick{clean white cambrick handkerchief be-}
\catch{side}
\stick{side it:\tsh The youth was just stooping}
\stick{down to take up the cushion, upon}
\stick{which I supposed he had been kneeling,}
\stick{\tsk the book was laid upon the bed,\tsk}
\stick{and as he rose, in taking up the cushion}
\stick{with one hand, he reached out his other}
\stick{to take it away at the same time.\tsh}
\stick{Let it remain there, my dear, said the}
lieutenant.

He did not offer to speak to me, till\break
\stick{I had walked up close to his bed-side:\tsk}
\stick{If you are captain \i{Shandy}’s servant, said}
\stick{he, you must present my thanks to your}
\stick{master, with my little boy’s thanks}
\stick{along with them, for his courtesy to}
\stick{me;\tsk if he was of \i{Levens}’s\tsk said the}
\stick{lieutenant.\tsk I told him your honour}
\stick{was\tsk Then, said he, I served three cam-}
\stick{paigns with him in \i{Flanders}, and re-}
\stick{member him,\tsk but ’tis most likely, as}
\catch{I had}
\stick{I had not the honour of any acquaint-}
\stick{ance with him, that he knows nothing}
\stick{of me.\tshh You will tell him, however,}
\stick{that the person his good nature has laid}
\stick{under obligations to him, is one \i{Le}}
\stick{\i{Fever}, a lieutenant in \i{Angus}’s\tsk but}
\stick{he knows me not,\tsk said he, a second}
\stick{time, musing;\tsk possibly he may}
\stick{my story\tsk added he\tsk pray tell the cap-}
\stick{tain, I was the ensign at \i{Breda}, whose}
\stick{wife was most unfortunately killed with}
\stick{a musket shot, as she lay in my arms in}
\stick{my tent.\tsh I remember the story, an’t}
\stick{please your honour, said I, very well.}
\stick{\tsh Do you so? said he, wiping his}
\stick{eyes with his handkerchief,\tsk then well}
\stick{may I.\tsk In saying this, he drew a little}
\stick{ring out of his bosom, which seemed}
\stick{tied with a black ribband about his neck,}
\stick{and kiss’d it twice\tshh Here, \i{Billy}, said}
\stick{he,\tshh the boy flew across the room to}
\catch{the}
\stick{the bed-side,\tsk and falling down upon}
\stick{his knee, took the ring in his hand, and}
\stick{kissed it too,\tsk then kissed his father,}
and sat down upon the bed and wept.

I wish, said my uncle \toby, with a\break
deep sigh,\tsk I wish, \trim, I was asleep.

Your honour, replied the corporal,\break
\stick{is too much concerned;\tsk shall I pour}
\stick{your honour out a glass of sack to your}
\stick{pipe?\tsk Do, \trim, said my uncle}
\i{Toby.}

I remember, said my uncle \toby,\break
\stick{sighing again, the story of the ensign}
\stick{and his wife, with a circumstance his}
\stick{modesty omitted;\tsk and particularly well}
\stick{that he, as well as she, upon some ac-}
\stick{count or other, (I forget what) was uni-}
\stick{versally pitied by the whole regiment;}
\catch{\tsk but}
\baselineskip 13.5pt
\stick{\tsk but finish the story thou art upon:}
\stick{\tsk ’Tis finished already, said the corpo-}
\stick{ral,\tsk for I could stay no longer,\tsk so}
\stick{wished his honour a good night; young}
\stick{\i{Le Fever} rose from off the bed, and saw}
\stick{me to the bottom of the stairs; and as}
\stick{we went down together, told me, they}
\stick{had come from \i{Ireland}, and were on}
\stick{their route to join the regiment in \i{Flan}-}
\stick{\i{ders}.\tsh But alas! said the corporal,\tsk}
\stick{the lieutenant’s last day’s march is over.}
\stick{\tsk Then what is to become of his poor}
boy? cried my uncle \i{Toby.}

\chap{VIII}{18pt}{4pt}

\centerline{\tstrut18pt\hbox to 144pt{\i{The Story of} \s{Le Fever} \i{continued}}.}

\dropcap{I}{t} was to my uncle \toby's eternal ho-\break
nour,\tshh though I tell it only for\break
\stick{the sake of those, who, when coop’d}
\stick{in betwixt a natural and a positive law,}
\catch{know}
\stick{know not for their souls, which way in}
\stick{the world to turn themselves\tsk That}
\stk{notwithstanding my uncle \toby was warm-}
\stick{ly engaged at that time in carrying on}
\stick{the siege of \i{Derdermond}\sic, parallel with}
\stick{the allies, who pressed theirs on so vigo-}
\stick{rously, that they scarce allowed him}
\stick{time to get his dinner\tsk that never-}
\stick{theless he gave up \i{Dendermond}, though}
\stick{he had already made a lodgment upon}
\stick{the counterscarp;\tsh and bent his whole}
\stck{thoughts towards the private distresses at}
\stick{the inn; and, except that he ordered}
\stick{the garden gate to be bolted up, by}
\stick{which he might be said to have turned}
\stick{the siege of \i{Dendermond} into a blockade,}
\stick{\tsk he left \i{Dendermond} to itself,\tsk to be}
\stick{relieved or not by the \i{French} king, as}
\stick{the \i{French} king thought good; and}
\stick{only considered how he himself should}
relieve the poor lieutenant and his son.
\patch{\tsh That}
\parskip 12pt

\topstrut\tsk That kind \s{Being}, who is a\break
\stick{friend to the friendless, shall recompence}
thee for this.

Thou hast left this matter short, said\break
\stick{my uncle \toby to the corporal, as he was}
\stick{putting him to bed,\tsk and I will tell}
\stick{thee in what, \i{Trim.}\tsk In the first}
\stick{place, when thou madest an offer of my}
\stick{services to \i{Le Fever},\tsk as sickness and}
\stick{travelling are both expensive, and thou}
\stick{knowest he was but a poor lieutenant,}
\stick{with a son to subsist as well as himself,}
\stick{out of his pay,\tsk that thou didst not}
\stick{make an offer to him of my purse; be-}
\stick{cause, had he stood in need, thou know-}
\stick{est, \trim, he had been as welcome to it}
\stick{as myself.\tsk Your honour knows, said}
\stick{the corporal, I had no orders;\tsk True,}
\stick{quoth my uncle \toby,\tsk thou didst very}
\catch{right,}
\stick{right, \trim, as a soldier,\tsk but certainly}
very wrong as a man.

In the second place, for which, indeed,\break
\stick{thou hast the same excuse, continued my}
\stick{uncle \toby,\tshh when thou offeredst him}
\stick{whatever was in my house,\tskk thou}
\stick{shouldst have offered him my house too:}
\stick{\tshh A sick brother officer should have}
\stick{the best quarters, \trim, and if we had}
\stick{him with us,\tsk we could tend and look}
\stick{to him:\tsh Thou art an excellent nurse}
\stick{thyself, \trim,\tsk and what with thy care}
\stick{of him, and the old woman’s, and his}
\stick{boy’s, and mine together, we might re-}
\stick{cruit him again at once, and set him}
upon his legs.\tush

\tsh In a fortnight or three weeks,\break
\stick{added my uncle \toby, smiling,\tskk he}
\stick{might march.\tsh He will never march,}
\catch{an’}
\stick{an’ please your honour, in this world,}
\stick{said the corporal:\tsk He will march;}
\stick{said my uncle \toby, rising up from the}
\stick{side of the bed, with one shoe off:\tsh}
\stick{An’ please your honour, said the cor-}
\stick{poral, he will never march, but to his}
\stick{grave:\tsk He shall march, cried my}
\stick{uncle \toby, marching the foot which}
\stick{had a shoe on, though without advanc-}
\stick{ing an inch,\tsk he shall march to his regi-}
\stick{ment.\tsk He cannot stand it, said the}
\stck{corporal;\tshh He shall be supported, said}
\stick{my uncle \toby;\tshh He’ll drop at last,}
\stick{said the corporal, and what will become}
\stick{of his boy?\tshh He shall not drop, said}
\stick{my uncle \toby, firmly.\tshh A-well-o’day,}
\stick{\tskk do what we can for him, said \trim,}
\stick{maintaining his point,\tsk the poor soul}
\stick{will die:\tsk He shall not die, by G\bzz{od},}
cried my uncle \i{Toby.}
\patch{\tsh The}

\tsk\tstrut 26pt The \s{accusing spirit}, which flew\break
\stick{up to heaven’s chancery with the oath,}
\stick{blush’d as he gave it in;\tskk and the}
\stick{\s{recording angel} as he wrote it down,}
\stick{dropp’d a tear upon the word, and blot-}
ted it out for ever.

\chap{IX}{24pt}{8pt}

\dropcap[ante=\lower 5pt\hbox{\tsh}]{M}{y} uncle \toby went to his\break
\tight{bureau,\tsk put his purse into}\break
\stick{his breeches pocket, and having order-}
\stick{ed the corporal to go early in the morn-}
\stick{ing for a physician,\tsk he went to bed,}
and fell asleep.
\patch[42pt]{\lsss{CHAP}.}

\chap{X}{34pt}{8pt}

\centerline{\i{The Story of} \s{Le Fever} \i{concluded}.\tstrut18pt}

\dropcap{T}{he} sun looked bright the morn-\break
ing after, to every eye in the vil-\break
\stick{lage but \i{Le Fever}’s and his afflicted son’s;}
\stick{the hand of death press’d heavy upon}
\stick{his eye-lids,\tsk and hardly could the}
\stick{wheel at the cistern turn round its circle,}
\stick{\tsk when my uncle \toby, who had rose}
\stick{up an hour before his wonted time, en-}
\stick{tered the lieutenant’s room, and without}
\stick{preface or apology, sat himself down}
\stick{upon the chair by the bed-side, and inde-}
\stick{pendantly of all modes and customs,}
\stick{opened the curtain in the manner an old}
\stick{friend and brother officer would have}
\stick{done it, and asked him how he did,\tsk}
\stick{how he had rested in the night,\tsk what}
\catch{was}
\stick{was his complaint,\tsk where was his}
\stick{pain,\tsk and what he could do to help}
\stick{him:\tsh and without giving him time}
\stick{to answer any one of the enquiries, went}
\stick{on and told him of the little plan which}
\stick{he had been concerting with the corpo-}
ral the night before for him.\tsh

\tsh You shall go home directly, \i{Le}\break
\stick{\i{Fever}, said my uncle \toby, to my house,}
\stick{\tsk and we’ll send for a doctor to see}
\stick{what’s the matter,\tsk and we’ll have an}
\stick{apothecary,\tsk and the corporal shall be}
\stick{your nurse;\tsh and I’ll be your servant,}
\i{Le Fever}.

There was a frankness in my uncle\break
\stick{\toby,\tsk not the \i{effect} of familiarity,\tsk but}
\stick{the \i{cause} of it,\tsk which let you at once}
\stick{into his soul, and shewed you the good-}
\stick{ness of his nature; to this, there was}
\catch{some-}
\stick{something in his looks, and voice, and}
\stick{manner, superadded, which eternally}
\stick{beckoned to the unfortunate to come and}
\stick{take shelter under him; so that before}
\stick{my uncle \toby had half finished the kind}
\stick{offers he was making to the father, had}
\stick{the son insensibly pressed up close to his}
\stick{knees, and had taken hold of the breast}
\stick{of his coat, and was pulling it towards}
\stick{him.\tsk The blood and spirits of \i{Le}}
\stick{\i{Fever}, which were waxing cold and}
\stick{slow within him, and were retreating to}
\stick{their last citadel, the heart,\tsk rallied}
\stick{back,\tsk the film forsook his eyes for a}
\stick{moment,\tsk he looked up wishfully\sic in my}
\stick{uncle \toby's face,\tsk then cast a look upon}
\stick{his boy,\tsk and that \i{ligament}, fine as}
it was,\tsk was never broken.\tush

Nature instantly ebb’d again,\tsh the\break
\stick{film returned to its place,\tsh the pulse}
\catch{flut-}
\stick{fluttered\tsh stopp’d\tsh went on\tsh}
\stick{\tsh throb’d\tsh stopp’d again\tsh mo-}
\stick{ved\tsh stopp’d\tsh shall I go on?\tsh}
No.

\chap{XI}{18pt}{0pt}

\initial[nindent=1pt]{I}{Am} so impatient to return to my own\break
story, that what remains of young\break
\stick{\i{Le Fever}’s, that is, from this turn of}
\stick{his fortune, to the time my uncle \i{Toby}}
\stick{recommended him for my preceptor,}
\stick{shall be told in a very few words, in the}
\stick{next chapter.\tsk All that is necessary to}
be added to this chapter is as follows.\tsk

That my uncle \toby, with young \i{Le}\break
\stick{\i{Fever} in his hand, attended the poor}
\stick{lieutenant, as chief mourners, to his}
grave.
\patch{That}

\topstrut That the governor of \i{Dendermond} paid\break
\stick{his obsequies all military honours,\tsk and}
\stick{that \i{Yorick}, not to be behind hand\tsk}
\stick{paid him all ecclesiastic\tsk for he buried}
\stick{him in his chancel:\tsk And it appears}
\stick{likewise, he preached a funeral sermon}
\stick{over him\tsh I say it \i{appears},\tsk for it}
\stick{was \i{Yorick}’s custom, which I suppose a}
\stick{general one with those of his profession,}
\stick{on the first leaf of every sermon which}
\stick{he composed, to chronicle down the}
\stick{time, the place, and the occasion of its}
\stick{being preached: to this, he was ever}
\stick{wont to add some short comment or}
\stick{stricture upon the sermon itself, seldom,}
\stick{indeed, much to its credit:\tsk For in-}
\stick{stance, \i{This sermon upon the jewish dispen}-}
\stick{\i{sation\tsk I don’t like it at all};\tsk \i{Though I}}
\stick{\i{own there is a world of} \s{water-landish}}
\stick{\i{knowlege\sic in it},\tsk \i{but ’tis all tritical}, \i{and}}
\catch{\i{most}}
\stick{\i{most tritically put together.\tsh This is}}
\stick{\i{but a flimsy kind of a composition}; \i{what}}
\i{was in my head when I made it?}

\tsk N.\@ B.\@ \i{The excellency of this text}\break
\stick{\i{is, that it will suit any sermon,\tsk and of}}
\stick{\i{this sermon,\tsk that it will suit any}}
\i{text}.\tsh

\tsh \i{For this sermon I shall be hanged},\break
\stick{\tsk \i{for I have stolen the greatest part of it}.}
\stick{\i{Doctor} Paidagunes \i{found me out.\hfill \fist  Set}}
\i{a thief to catch a thief}.\tsh

On the back of half a dozen I find\break
\stick{written, \i{So}, \i{so}, and no more\tsk and}
\stick{upon a couple \i{Moderato}; by which, as}
\stick{far as one may gather from \i{Altieri’s Ita}-}
\stick{\i{lian} dictionary,\tsk but mostly from the}
\stick{authority of a piece of green whipcord,}
\stick{which seemed to have been the unravel-}
\catch{ling}
\stick{ling of \i{Yorick}’s whip-lash, with which}
\stick{he has left us the two sermons marked}
\stick{\i{Moderato}, and the half dozen of \i{So}, \i{so},}
\stick{tied fast together in one bundle by them-}
\stck{selves,\tsk one may safely suppose he meant}
pretty near the same thing.

\vskip 4pt

There is but one difficulty in the way\break
\stick{of this conjecture, which is this, that}
\stick{the \i{moderato}’s are five times better than}
\stick{the \i{so}, \i{so}’s;\tsk shew ten times more know-}
\stick{lege\sic of the human heart;\tsk have seventy}
\stick{times more wit and spirit in them;\tsk}
\stick{(and, to rise properly in my climax)\tsk}
\stick{discover a thousand times more genius;}
\stick{\tsk and to crown all, are infinitely more}
\stck{entertaining than those tied up with them;}
\stick{\tsk for which reason, whene’er \i{Yorick}’s}
\stck{\i{dramatic} sermons are offered to the world,}
\stick{though I shall admit but one out of the}
\stick{whole number of the \i{so}, \i{so}’s, I shall,}
\catch{never-}
\stick{nevertheless, adventure to print the two}
\i{moderato}’s without any sort of scruple.

What \i{Yorick} could mean by the words\break
\stick{\i{lentamente},\tsk \i{tenutè},\tsk \i{grave},\tsk and some-}
\stick{times \i{adagio},\tsk as applied to theological}
\stick{compositions, and with which he has}
\stick{characterized some of these sermons, I}
\stick{dare not venture to guess.\tsh I am more}
\stick{puzzled still upon finding \i{a l’octava alta!}}
\stick{upon one;\tsh \i{Con strepito} upon the back}
\stick{of another;\tsh \i{Scicilliana} upon a third;}
\stick{\tsk \i{Alla capella} upon a fourth;\tsk}
\stick{\i{Con l’arco} upon this;\tsk \i{Senza l’arco}}
\stick{upon that.\tsh All I know is, that they}
\stick{are musical terms, and have a mean-}
\stick{ing;\tsh and as he was a musical man,}
\stick{I will make no doubt, but that by some}
\stick{quaint application of such metaphors to}
\stck{the compositions in hand, they impressed}
\stick{very distinct ideas of their several cha-}
\catch{racters}
\stick{racters upon his fancy,\tsk whatever they}
may do upon that of others.

Amongst these, there is that particular\break
\stick{sermon which has unaccountably led me}
\stick{into this digression\tsh The funeral ser-}
\stick{mon upon poor \i{Le Fever}, wrote out}
\stick{very fairly, as if from a hasty copy.\tsk I}
\stick{take notice of it the more, because it}
\stick{seems to have been his favourite compo-}
\stick{sition\tsk It is upon mortality; and is}
\stick{tied length-ways and cross-ways with a}
\stick{yarn thrum, and then rolled up and}
\stick{twisted round with a half sheet of dirty}
\stick{blue paper, which seems to have been}
\stick{once the cast cover of a general review,}
\stick{which to this day smells horribly of}
\stick{horse-drugs.\tsh Whether these marks}
\stick{of humiliation were designed,\tsk I some-}
\stick{thing doubt;\tsh because at the end of}
\stick{the sermon, (and not at the beginning}
\catch{of}
\stick{of it)\tsk very different from his way of}
\stick{treating the rest, he had wrote\tsk\qquad}
\stick{\hfill Bravo! \hfill}
\stick{\indent\tsk Though not very offensively,}
\stick{\tsk for it is at two inches, at least,}
\stick{and a half’s distance from, and below}
\stick{the concluding line of the sermon, at}
\stick{the very extremity of the page, and in}
\stick{that right hand corner of it, which, you}
\stick{know, is generally covered with your}
\stick{thumb; and, to do it justice, it is wrote}
\stick{besides with a crow’s quill so faintly in a}
\stick{small \i{Italian} hand, as scarce to sollicit}
\stick{the eye towards the place, whether your}
\stick{thumb is there or not,\tsk so that from the}
\stick{\i{manner of it}, it stands half excused;}
\stick{and being wrote moreover with very pale}
\stick{ink, diluted almost to nothing,\tsk ’tis}
\stick{more like a \i{ritratto} of the shadow of va-}
\stick{nity, than of \s{Vanity} herself\tsk of the}
\stick{two; resembling rather a faint thought}
\catch{of}
\stick{of transient applause, secretly stirring}
\stick{up in the heart of the composer, than a}
\stick{gross mark of it, coarsely obtruded}
upon the world.

With all these extenuations, I am\break
\stick{aware, that in publishing this, I do no}
\stick{service to \i{Yorick}’s character as a modest}
\stick{man;\tsk but all men have their failings!}
\stick{and what lessens this still farther, and}
\stick{almost wipes it away, is this; that the}
\stick{word was struck through sometime af-}
\stick{terwards (as appears from a different}
\stick{tint of the ink) with a line quite across}
\stick{it in this manner, \struck{\ls{BRAVO}}\tsk as if}
\stick{he had retracted, or was ashamed of the}
opinion he had once entertained of it.

These short characters of his sermons\break
\stick{were always written, excepting in this}
\stick{one instance, upon the first leaf of his}
\catch{sermon,}
\stick{sermon, which served as a cover to it;}
\stick{and usually upon the inside of it, which}
\stick{was turned towards the text;\tsk but at}
\stick{the end of his discourse, where, perhaps,}
\stick{he had five or six pages, and sometimes,}
\stick{perhaps, a whole score to turn himself}
\stick{in,\tsk he took a large circuit, and, in-}
\stick{deed, a much more mettlesome one;\tsk}
\stick{as if he had snatched the occasion of un-}
\stick{lacing himself with a few more frolick-}
\stick{some strokes at vice, than the straitness}
\stick{of the pulpit allowed.\tsk These, though}
\stick{hussar-like, they skirmish lightly and}
\stick{out of all order, are still auxiliaries on the}
\stick{side of virtue\tsk\!; tell me then, Mynheer}
\stick{Vander Blonederdondergewdenstronke,}
\stck{why they should not be printed together?}


\patch[28pt]{\lsss{CHAP}.}

\chap{XII}{28pt}{0pt}

\dropcap{W}{hen} my uncle \toby had turned
every thing into money, and\break
\stick{settled all accounts betwixt the agent}
\stick{of the regiment and \i{Le Fever}, and be-}
\stick{twixt \i{Le Fever} and all mankind,\tshh}
\stick{there remained nothing more in my un-}
\stick{cle \toby's hands, than an old regimental}
\stick{coat and a sword; so that my uncle \toby}
\stick{found little or no opposition from the}
\stick{world in taking administration. The}
\stick{coat my uncle \toby gave the corporal;}
\stick{\tshh Wear it, \trim, said my uncle \toby,}
\stick{as long as it will hold together, for the}
\stick{sake of the poor lieutenant\tshh And}
\stick{this,\tshh said my uncle \toby, taking up}
\stick{the sword in his hand, and drawing it}
\stick{out of the scabbard as he spoke\tshh and}
\stick{this, \i{Le Fever}, I’ll save for thee,\tskk ’tis}
\catch{all}
\stick{all the fortune, continued my uncle}
\stick{\toby, hanging it up upon a crook, and}
\stick{pointing to it,\tsk ’tis all the fortune, my}
\stick{dear \i{Le Fever}, which God has left thee;}
\stick{but if he has given thee a heart to fight}
\stick{thy way with it in the world,\tsk and thou}
\stick{doest it like a man of honour,\tskk ’tis}
enough for us.

As soon as my uncle \toby had laid a\break
\stick{foundation, and taught him to inscribe}
\stick{a regular polygon in a circle, he sent}
\stick{him to a public school, where, except-}
\stick{ing \i{Whitsontide} and \i{Christmas}, at which}
\stick{times the corporal was punctually dis-}
\stick{patched for him,\tsk he remained to the}
\stick{spring of the year, seventeen; when the}
\stick{stories of the emperor’s sending his army}
\stick{into \i{Hungary} against the \i{Turks}, kindling}
\stick{a spark of fire in his bosom, he left his}
\stick{\i{Greek} and \i{Latin} without leave, and}
\catch{throw-}
\stick{throwing himself upon his knees before}
\stck{my uncle \toby, begged his father’s sword,}
\stick{and my uncle \toby's leave along with it,}
\stick{to go and try his fortune under \i{Eu}-}
\stick{\i{gene}.\tsk Twice did my uncle \toby forget}
\stick{his wound and cry out, \i{Le Fever!} I will}
\stick{go with thee, and thou shalt fight be-}
\stick{side me\tsh And twice he laid his hand}
\stick{upon his groin, and hung down his head}
in sorrow and disconsolation.\tush

My uncle \toby took down the sword\break
\stick{from the crook, where it had hung un-}
\stick{touched ever since the lieutenant’s death,}
\stick{and delivered it to the corporal to}
\stick{brighten up;\tsk and having detained}
\stick{\i{Le Fever} a single fortnight to equip him,}
\stick{and contract for his passage to \i{Leghorn},}
\stick{\tsk he put the sword into his hand,\tshh}
\stick{If thou art brave, \i{Le Fever}, said my}
\stick{uncle \toby, this will not fail thee,\tshh}
\catch{but}
\stick{but Fortune, said he, (musing a little),}
\stick{\tsh Fortune may\tsh And if she does,}
\stick{\tsk added my uncle \toby, embracing}
\stick{him, come back again to me, \i{Le Fever},}
and we will shape thee another course.

The greatest injury could not have op-\break
\stick{pressed the heart of \i{Le Fever} more than}
\stick{my uncle \toby's paternal kindness;\tsh}
\stick{he parted from my uncle \toby, as the}
\stick{best of sons from the best of fathers\tsh}
\stick{both dropped tears\tsh and as my uncle}
\stick{\toby gave him his last kiss, he slipped}
\stick{sixty guineas, tied up in an old purse of}
\stick{his father’s, in which was his mother’s}
\stick{ring, into his hand,\tsk and bid God bless}
him.

\patch[14pt]{\lsss{CHAP}.}

\chap{XIII}{28pt}{0pt}

\dropcap[findent=0pt,nindent=6pt,loversize=-.11,lraise=.25]{\itshape L}{\itshape e} \i{Fever} got up to the Imperial army\break
just time enough to try what me-\break
\stick{tal his sword was made of, at the defeat}
\stick{of the \i{Turks} before \i{Belgrade}; but a se-}
\stick{ries of unmerited mischances had pur-}
\stick{sued him from that moment, and trod}
\stick{close upon his heels for four years toge-}
\stick{ther after: he had withstood these buf-}
\stick{fetings to the last, till sickness overtook}
\stick{him at \i{Marseilles}, from whence he wrote}
\stick{my uncle \toby word, he had lost his}
\stick{time, his services, his health, and, in}
\stick{short, every thing but his sword;\tsh}
\stick{and was waiting for the first ship to re-}
turn back to him.

As this letter came to hand about six\break
\stick{weeks before \susannah's accident, \i{Le}}
\catch{\i{Fever}}
\stick{\i{Fever} was hourly expected; and was}
\stick{uppermost in my uncle \toby's mind all}
\stick{the time my father was giving him and}
\stick{\i{Yorick} a description of what kind of a}
\stick{person he would chuse for a preceptor}
\stick{to me: but as my uncle \toby thought}
\stick{my father at first somewhat fanciful in}
\stick{the accomplishments he required, he}
\stick{forbore mentioning \i{Le Fever}’s name,}
\stick{\tsh till the character, by \i{Yorick}’s inter-}
\stick{position, ending unexpectedly, in one,}
\stick{who should be gentle tempered, and ge-}
\stick{nerous, and good, it impressed the image}
\stick{of \i{Le Fever}, and his interest upon my}
\stick{uncle \toby so forceably, he rose instant-}
\stick{ly off his chair; and laying down his}
\stick{pipe, in order to take hold of both my}
\stick{father’s hands\tsh I beg, brother \i{Shandy},}
\stick{said my uncle \toby, I may recommend}
\stick{poor \i{Le Fever}’s son to you\tsk I beseech}
\stick{you, do, added \i{Yorick}\tsh He has a good}
\catch{heart,}
\stick{heart, said my uncle \i{Toby}\tsk And a}
\stick{brave one too, an’ please your honour,}
said the corporal.

\tshh The best hearts, \trim, are\break
\stick{ever the bravest, replied my uncle \toby.}
\stick{\tsh And the greatest cowards, an’ please}
\stick{your honour, in our regiment, were the}
\stick{greatest rascals in it.\tshh There was ser-}
jeant \i{Kumbur}, and ensign\tush

\tshh We’ll talk of them, said my fa-\break
ther, another time.

\chap{XIV}{18pt}{0pt}

\dropcap[nindent=3pt]{W}{hat} a jovial and a merry world\break
would this be, may it please your\break
\stick{worships, but for that inextricable laby-}
\stick{rinth of debts, cares, woes, want, grief,}
\catch{dis-}
\parskip 8pt
\stick{discontent, melancholy, large jointures,}
impositions, and lies!

Doctor \slop, like a son of a w\bzz{hore}, as\break
\stick{my father called him for it,\tsk to exalt}
\stick{himself,\tsk debased me to death,\tsk and}
\stick{made ten thousand times more of \i{Susan}-}
\stick{\i{nah}’s accident, than there was any}
\stick{grounds for; so that in a week’s time,}
\stick{or less, it was in every body’s mouth,}
\stick{\i{That poor Master Shandy}\enspace\vastfill{1.48em}}
\stick{\vastfill{1.48em}\enspace entirely.\tsk}
\stick{And \textsc{Fame}, who loves to double every}
\stick{thing,\tsk in three days more, had sworn,}
\stick{positively she saw it,\tsk and all the world,}
\stick{as usual, gave credit to her evidence\tsh}
\stick{\lqq That the nursery window had not}
\stick{only\quad\vastfill{1.45em}}
\stick{\vastfill{1.45em}}
\stick{\hbox to 4.2em{\vastfill{1.4em};}\tsh but that\enspace \vastfill{1.35em}}
\stick{\vastfill{1.45em}}
\stick{\hbox to 6em{\vastfill{1.45em}’s} also.\,”\hfill}

\patch{Could}
\setlength{\baselineskip}{14pt}

\topstrut Could the world have been sued like\break
\stick{a \s{body-corporate},\tskk my father had}
\stick{brought an action upon the case, and}
\stick{trounced it sufficiently; but to fall foul}
\stick{of individuals about it\tsh as every soul}
\stick{who had mentioned the affair, did it}
\stick{with the greatest pity imaginable;\tsh}
\stick{’twas like flying in the very face of his}
\stick{best friends:\tshh And yet to acquiesce}
\stick{under the report, in silence\tsk was to ac-}
\stick{knowledge it openly,\tsk at least in the}
\stick{opinion of one half of the world; and}
\stick{to make a bustle again, in contradicting}
\stick{it,\tsk was to confirm it as strongly in the}
opinion of the other half.\tush

\tsh Was ever poor devil of a coun-\break
\stick{try gentleman so hampered? said my}
father.
\patch{I would}

\topstrut  I would shew him publickly, said my\break
uncle \toby, at the market cross.

\tsk ’Twill have no effect, said my\break
father.

\chap{XV}{12pt}{0pt}

\tsk I’ll put him, however, into\break
breeches said my father,\tsk let the world\break
say what it will.

\chap{XVI}{12pt}{0pt}

\dropcap[loversize=-.2,lraise=.31]{T}{here} are a thousand resolutions,\break
Sir, both in church and state, as\break
\stick{well as in matters, Madam, of a more}
\stick{private concern;\tsk which, though they}
\stick{have carried all the appearance in the}
\stick{world of being taken, and entered upon}
\stick{in a hasty, hare-brained, and unadvised}
\stick{manner, were, notwithstanding this,}
\catch{(and}
\stick{(and could you or I have got into the}
\stick{cabinet, or stood behind the curtain, we}
\stick{should have found it was so) been weigh-}
\stick{ed, poized, and perpended\tshh argued}
\stick{upon\tsh canvassed through\tsh entered}
\stick{into, and examined on all sides with so}
\stck{much coolness, that the \textsc{goddess} of \textsc{cool}-}
\stick{\s{ness} herself (I do not take upon me to}
\stick{prove her existence) could neither have}
wished it, or done it better.

Of the number of these was my fa-\break
\stick{ther’s resolution of putting me into}
\stick{breeches; which, though determined}
\stick{at once,\tsk in a kind of huff, and a defi-}
\stick{ance of all mankind, had, nevertheless,}
\stick{been \i{pro}’\i{d} and \i{conn}’\i{d}, and judicially talk-}
\stick{ed over betwixt him and my mother}
\stick{about a month before, in two several}
\stick{\i{beds of justice}, which my father had held}
\stick{for that purpose. I shall explain the}
\catch{nature}
\stick{nature of these beds of justice in my}
\stick{next chapter; and in the chapter follow-}
\stick{ing that, you shall step with me, Madam,}
\stick{behind the curtain, only to hear in what}
\stick{kind of manner my father and my mo-}
\stick{ther debated between themselves, this}
\stick{affair of the breeches,\tsk from which you}
\stick{may form an idea, how they debated all}
lesser matters.

\chap{XVII}{24pt}{0pt}

\dropcap[loversize=-.23,lraise=.3]{T}{he} ancient \i{Goths} of \i{Germany}, who\break
(the learned \i{Cluverius} is positive)\break
\stick{were first seated in the country between}
\stick{the \i{Vistula} and the \i{Oder}, and who after-}
\stick{wards incorporated the \i{Herculi}, the \i{Bu}-}
\stick{\i{gians}, and some other \i{Vandallick} clans}
\stick{to ’em,\tsk had all of them a wise custom}
\stick{of debating every thing of importance}
\catch{to}
\stick{to their state, twice; that is,\tsk once}
\stick{drunk, and once sober:\tsh Drunk\tsk}
\stick{that their councils might not want vi-}
\stick{gour;\tsh and sober\tsk that they might}
not want discretion.

Now my father being entirely a water-\break
\stick{drinker,\tskk was a long time gravelled}
\stick{almost to death, in turning this as much}
\stick{to his advantage, as he did every other}
\stick{thing, which the ancients did or said;}
\stick{and it was not till the seventh year of his}
\stick{marriage, after a thousand fruitless ex-}
\stick{periments and devices, that he hit upon}
\stick{an expedient which answered the pur-}
\stick{pose;\tshh and that was when any diffi-}
\stick{cult and momentous point was to be set-}
\stick{tled in the family, which required great}
\stick{sobriety, and great spirit too, in its de-}
\stick{termination,\tshh he fixed and set apart}
\stick{the first \i{Sunday} night in the month, and}
\catch{the}
\stick{the \i{Saturday} night which immediately}
\stick{preceded it, to argue it over, in bed}
\stick{with my mother: By which contrivance,}
\stick{if you consider, Sir, with yourself,\quad \vastfill{1.4em}}
\stick{\vastfill{1.414em}}
\stick{\vastfill{1.414em}}
\stick{\vastfill{1.414em}}
\hbox to 80pt{\vastfill{1.414em}}.

These my father, humorously enough,\break
\stick{called his \i{beds of justice};\tsh for from}
\stick{the two different counsels taken in these}
\stick{two different humours, a middle one}
\stick{was generally found out, which touched}
\stick{the point of wisdom as well, as if he}
\stick{had got drunk and sober a hundred}
times.

It must not be made a secret of to the\break
\stick{world, that this answers full as well in}
\stick{literary discussions, as either in military}
\catch{or}
\stick{or conjugal; but it is not every author}
\stick{that can try the experiment as the \i{Goths}}
\stick{and \i{Vandals} did it\tsk or if he can,}
\stick{may it be always for his body’s health;}
\stick{and to do it, as my father did it,\tsk}
\stick{am I sure it would be always for his}
soul’s.\tsh

My way is this:\tush

In all nice and ticklish discussions,\tsk\break
\stick{(of which, heaven knows, there are}
\stick{but too many in my book)\tskk where I}
\stick{find I cannot take a step without the}
\stick{danger of having either their worships}
\stick{or their reverences upon my back\tsh}
\stick{I write one half \i{full},\tsk and t’other \i{fast}-}
\stick{\i{ing};\tsh or write it all full,\tsk and cor-}
\stick{rect it fasting;\tsh or write it fasting,\tsk}
\stick{and correct it full, for they all come to}
\stick{the same thing:\tsh So that with a less}
\catch{variation}
\stick{variation from my father’s plan, than}
\stick{my father’s from the \i{Gothick}\tsk I feel}
\stick{myself upon a par with him in his first}
\stick{bed of justice,\tsk and no way inferior to}
\stick{him in his second.\tsh These different}
\stick{and almost irreconcileable effects, flow}
\stick{uniformly from the wise and wonderful}
\stick{mechanism of nature,\tsk of which,\tsk be}
\stick{her’s the honour.\tsh All that we can do,}
\stick{is to turn and work the machine to the}
\stick{improvement and better manufactory}
of the arts and sciences.\tush

Now, when I write full,\tsk I write as if\break
\stick{I was never to write fasting again as long}
\stick{as I live;\tsk that is, I write free from}
\stick{the cares, as well as the terrors of the}
\stick{world.\tsk I count not the number of}
\stick{my scars,\tsk nor does my fancy go forth}
\stick{into dark entries and bye corners to ante-}
\stick{date my stabs.\tsk In a word, my pen}
\catch{takes}
\stick{takes its course; and I write on as much}
\stick{from the fullness of my heart, as my}
stomach.\tsh

But when, an’ please your honours,\break
\stick{I indite fasting, ’tis a different history.}
\stick{\tsh I pay the world all possible atten-}
\stick{tion and respect,\tsk and have as great a}
\stick{share (whilst it lasts) of that understrap-}
\stick{ping virtue of discretion, as the best of}
\stick{you.\tsh So that betwixt both, I write a}
\stick{careless kind of a civil, nonsensical, good}
\stick{humoured \i{Shandean} book, which will}
do all your hearts good\tush

\tsh And all your heads too,\tsk pro-\break
vided you understand it.

\patch[28pt]{\lsss{CHAP}.}
\parskip 12pt

\chap{XVIII}{28pt}{8pt}

\dropcap{W}{e} should begin, said my father,\break
turning himself half round in\break
\stick{bed, and shifting his pillow a little to-}
\stick{wards my mother’s, as he opened the}
\stick{debate\tsk We should begin to think,}
\stick{Mrs.\@ \i{Shandy}, of putting this boy into}
breeches.\tsh

We should so,\tsk said my mother.\tsh\break
\stick{We defer it, my dear, quoth my father,}
shamefully.\tush

I think we do, Mr.\@ \i{Shandy},\tsk said my\break
mother.

\tsh Not but the child looks extreme-\break
\stick{ly well, said my father, in his vests and}
tunicks.\tush

\patch{\tsh He}

\topstrut
\tshh He does look very well in them,\break
\tsk replied my mother.\tush

\tshh And for that reason it would be\break
\stick{almost a sin, added my father, to take}
him out of ’em.\tsh

\tshh It would so,\tsk said my mother:\break
\stick{\tshh But indeed he is growing a very}
tall lad,\tsk rejoin’d my father.

\tshh He is very tall for his age, in-\break
deed,\tsk said my mother.\tsh

\tshh I can not (making two syllables\break
\stick{of it) imagine, quoth my father, who}
the duce he takes after.\tsh

I cannot conceive, for my life,\tsk said\break
my mother.\tush


\patch{Humph!}

Humph!\tsh\topstrut said my father.

(The dialogue ceased for a moment.)

\tshh I am very short myself,\tsk conti-\break
nued my father, gravely.

You are very short, Mr.\@ \i{Shandy},\tsk\break
said my mother.

Humph! quoth my father to himself,\break
\stick{a second time: in muttering which, he}
\stick{plucked his pillow a little further from}
\stick{my mother’s,\tsk and turning about again,}
\stick{there was an end of the debate for three}
minutes and a half.

\tshh When he gets these breeches\break
\stick{made, cried my father in a higher tone,}
he’ll look like a beast in ’em.


\patch{He}

\topstrut He will be very aukward in them at\break
first, replied my mother.\tush

\tshh And ’twill be lucky, if that’s\break
the worst on’t, added my father.

It will be very lucky, answered my\break
mother.

I suppose, replied my father,\tsk mak-\break
\stick{ing some pause first,\tsk he’ll be exactly}
like other people’s children.\tush

Exactly, said my mother.\tush

\tshh Though I should be sorry for\break
\stick{that, added my father: and so the de-}
bate stopped again.

\tshh They should be of leather, said\break
\stick{my father, turning him about again.\tsk}
\catch{They}
\parskip 10pt
\stick{They will last him, said my mother, the}
longest.

But he can have no linings to ’em,\break
replied my father.\tush

He cannot, said my mother.

’Twere better to have them of fustian,\break
quoth my father.

Nothing can be better, quoth my\break
mother.\tush

\tshh Except dimity,\tsk replied my fa-\break
\stick{ther:\tsh ’Tis best of all,\tsk replied my}
mother.

\tshh One must not give him his death,\break
however,\tsk interrupted my father.

By no means, said my mother:\tshh\break
and so the dialogue stood still again.

\patch{I am}
\parskip 12pt

\topstrut I am resolved, however, quoth my\break
father, breaking silence the fourth time,\break
he shall have no pockets in them.\tsh

\tshh There is no occasion for any,\break
said my mother.\tush

I mean in his coat and waistcoat,\tsk\break
cried my father.

\tshh I mean so too,\tsk replied my\break
mother.

\tshh Though if he gets a gig or top\break
\stick{\tshh Poor souls! it is a crown and a}
\stck{scepter to them,\tsk they should have where}
to secure it.\tush

Order it as you please, Mr.\@ \i{Shandy}, replied my
mother.\tush

\patch{\tsh But}

\tshh But don’t you think it right?\break
\stick{added my father, pressing the point}
home to her.

Perfectly, said my mother, if it pleases\break
you, Mr. \i{Shandy.}\tush

\tshh \tight{There’s for you! cried my father,}\break
\stick{losing temper\tsh Pleases me!\tsh You}
\stick{never will distinguish, Mrs.\@ \i{Shandy}, nor}
\stick{shall I ever teach you to do it, betwixt}
\stick{a point of pleasure and a point of con-}
\stick{venience.\tsk This was on the \i{Sunday}}
\stick{night;\tsk and further this chapter say-}
eth not.

\chap{XIX}{18pt}{8pt}

\dropcap[findent=-2pt,nindent=4pt,lraise=0]{A}{fter} \hbox{\tight{my father had debated the af-}}\break
\tight{fair of the breeches with my mother,}\break
\stick{\tsk he consulted \i{Albertus Rubenius} upon}
\catch{it;}
\stick{it; and \i{Albertus Rubenius} used my father}
\stick{ten times worse in the consultation (if}
\stck{possible) than even my father had used my}
\stick{mother: For as \i{Rubenius} had wrote a}
\stick{quarto \i{express, De re Vestiaria Veterum},\tsk}
\stick{it was \i{Rubenius}’s business to have given}
\stick{my father some lights.\tsk On the contrary,}
\stick{my father might as well have thought of}
\stick{extracting the seven cardinal virtues out}
\stick{of a long beard,\tsk as of extracting a single}
word out of \i{Rubenius} upon the subject.

Upon every other article of ancient\break
\stick{dress, \i{Rubenius} was very communicative}
\stick{to my father;\tsk gave him a full and sa-}
tisfactory account of
\parskip 10pt

\noindent
\short{The Toga, or loose gown.}
\short{The Chlamys.}
\short{The Ephod.}
\short{The Tunica, or Jacket.}
\catch{The}
\short{The Synthesis.}
\short{The Pænula.}
\short{The Lacema, with its Cucullus.}
\short{The Paludamentum.}
\short{The Prætexta.}
\short{The Sagum, or soldier’s jerkin.}
\indent\enspace The Trabea: of which, according\break
to \i{Suetonius}, there were three kinds.\tsk

\tsk But what are all these to the\break
breeches? said my father.

\i{Rubenius} threw him down upon the\break
\stick{counter all kinds of shoes which had}
\stick{been in fashion with the \i{Romans.}\tush}
There was,

\noindent
\short{The open shoe.}
\short{The close shoe.}
\short{The slip shoe.}
\short{The wooden shoe.}
\catch{The}
\baselineskip 13pt
\stick{\hbox to 51.2pt{\hfill}The soc.\hfill}
\stick{\hbox to 51.2pt{\hfill}The buskin.\hfill}
\stick{\hbox to 51.2pt{\hfill And }The military shoe with hob-}
\stick{\hbox to 63pt{\hfill}nails in it, which \i{Juvenal}}
\stick{\hbox to 63pt{\hfill}takes notice of.\hfill}
\stick{There were, The clogs.\hfill}
\stick{\hbox to 51.2pt{\hfill}The patins.\hfill}
\stick{\hbox to 51.2pt{\hfill}The pantoufles.\hfill}
\stick{\hbox to 51.2pt{\hfill}The brogues.\hfill}
\stick{\hbox to 51.2pt{\hfill}The sandals, with latchets to}
\stick{\hbox to 58.0pt{\hfill}them.\hfill}
\stick{\hbox to 51.2pt{\hfill There was, }The felt shoe.\hfill}
\stick{\hbox to 51.2pt{\hfill}The linen shoe.\hfill}
\stick{\hbox to 51.2pt{\hfill}The laced shoe.\hfill}
\stick{\hbox to 51.2pt{\hfill}The braided shoe.\hfill}
\stick{\hbox to 51.2pt{\hfill}The calceus incisus.\hfill}
\stick{\hbox to 51.2pt{\hfill And }The calceus rostratus.\hfill}
\stick{\i{Rubenius} shewed my father how well}
\stick{they all fitted,\tskk in what manner they}
\stick{laced on,\tskk with what points, straps,}
\stick{thongs, latchets, ribands, jaggs, and}
ends.\tush

\patch{\tsh But}
\baselineskip 14pt

\tsh\topstrut But I want to be informed about\break
the breeches, said my father.

\i{Albertus Rubenius} informed my father\break
\stick{that the \i{Romans} manufactured stuffs of}
\stick{various fabricks,\tsh some plain,\tsk some}
\stick{striped,\tsk others diapered throughout the}
\stick{whole contexture of the wool, with silk}
\stick{and gold\tsk That linen did not begin}
\stick{to be in common use, till towards the}
\stick{declension of the empire, when the \i{Egyp}-}
\stick{\i{tians} coming to settle amongst them,}
brought it into vogue.

\tshh That persons of quality and\break
\stick{fortune distinguished themselves by the}
\stick{fineness and whiteness of their cloaths;}
\stick{which colour (next to purple, which}
\stick{was appropriated to the great offices)}
\stick{they most affected and wore on their}
\catch{birth}
\stick{birth-days and public rejoicings.\tsh}
\stick{That it appeared from the best historians}
\stick{of those times, that they frequently sent}
\stick{their cloaths to the fuller, to be cleaned}
\stick{and whitened;\tsh but that the inferior}
\stick{people, to avoid that expence, generally}
\stick{wore brown cloaths, and of a something}
\stick{coarser texture,\tsk till towards the begin-}
\stick{ning of \i{Augustus}’s reign, when the slave}
\stick{dressed like his master, and almost every}
\stick{distinction of habiliment was lost, but}
the \i{Latus Clavus.}

And what was the \i{Latus Clavus?} said\break
my father.

\i{Rubenius} told him, that the point was\break
\stick{still litigating amongst the learned:\tsh}
\stick{That \i{Egnatius}, \i{Sigonius}, \i{Bossius Tici}-}
\stick{\i{nensis}, \i{Bayfius}, \i{Budæus}, \i{Salmasius}, \i{Lip}-}
\stick{\i{sius}, \i{Lazius}, \i{Isaac Causonbon}, and \i{Jo}-}
\catch{\i{seph}}
\stick{\i{seph Scaliger}, all differed from each}
\stick{other,\tsk and he from them: That some}
\stick{took it to be the button,\tsk some the coat}
\stick{itself,\tsk others only the colour of it:\tsk}
\stick{That the great \i{Bayfius}, in his Wardrobe}
\stick{of the ancients, chap.\@ 12.\tsk honestly}
\stick{said, he knew not what it was,\tsk whether}
\stick{a tibula,\tsk a stud,\tsk a button,\tsk a loop,\tsk}
a buckle,\tsk or clasps and keepers.\tsh

\tsh My father lost the horse, but not\break
\stick{the saddle\tsh They are \i{hooks and eyes},}
\stick{said my father\tsh and with hooks and}
eyes he ordered my breeches to be made.

\chap{XX}{12pt}{0pt}

\dropcap{W}{e} are now going to enter upon a\break
new scene of events.\tush

\vskip -2pt

\tshh Leave we then the breeches in\break
\stick{the taylor’s hands, with my father stand-}
\catch{ing}
\stick{ing over him with his cane, reading him}
\stick{as he sat at work a lecture upon the}
\stick{\i{latus clavus}, and pointing to the precise}
\stick{part of the waistband, where he was de-}
termined to have it sewed on.\tsh

Leave we my mother\tsk (truest of all the\break
\stick{\tight{\i{Poco-curante’s} of her sex!)\tsk careless about}}
\stick{it, as about every thing else in the world}
\stick{which concerned her;\tsk that is,\tsk indiffe-}
\stck{rent whether it was done this way or that,}
\tsk provided it was but done at all.\tsh

Leave we \slop likewise to the full pro-\break
fits of all my dishonours.\tush

Leave we poor \i{Le Fever} to recover,\break
\stick{and get home from \i{Marseilles} as he can.}
\stick{\tsh And last of all,\tsk because the hard-}
est of all\tsh
\patch[12pt]{Let}

Let us leave, if possible, \i{myself:}\tsh\break
\stick{But ’tis impossible,\tsk I must go along}
with you to the end of the work.

\chap{XXI}{18pt}{0pt}

\dropcap{I}{f} the reader has not a clear concep-\break
tion of the rood and the half of\break
\stick{ground which lay at the bottom of my}
\stick{uncle \toby's kitchen-garden, and which}
\stick{was the scene of so many of his delicious}
\stick{hours,\tsk the fault is not in me,\tsk but in}
\stick{his imagination;\tsk for I am sure I gave}
\stick{him so minute a description, I was al-}
most ashamed of it.

When \s{Fate} was looking forwards one\break
\stick{afternoon, into the great transactions of}
\stick{future times,\tsk and recollected for what}
\stick{purposes, this little plot, by a decree}
\stick{fast bound down in iron, had been de-}
\catch{stined,}
\stick{stined,\tsk she gave a nod to \s{Nature}\tsk}
\stick{’twas enough\tsk Nature threw half a spade}
\stick{full of her kindliest compost upon it,}
\stick{with just so \i{much} clay in it, as to retain}
\stick{the forms of angles and indentings,\tsk}
\stick{and so \i{little} of it too, as not to cling to}
\stick{the spade, and render works of so much}
glory, nasty in foul weather.

My uncle \toby came down, as the\break
\stick{reader has been informed, with plans}
\stick{along with him, of almost every forti-}
\stick{fied town in \i{Italy} and \i{Flanders}; so let}
\stick{the Duke of \i{Marlborough}, or the allies,}
\stick{have set down before what town they}
\stick{pleased, my uncle \toby was prepared for}
them.

His way, which was the simplest one\break
\stick{in the world, was this; as soon as ever}
\stick{a town was invested\tsk (but sooner when}
\catch{the}
\stick{the design was known) to take the plan}
\stick{of it, (let it be what town it would)}
\stick{and enlarge it upon a scale to the exact}
\stick{size of his bowling-green; upon the sur-}
\stick{face of which, by means of a large role}
\stick{of packthread, and a number of small}
\stick{piquets driven into the ground, at the}
\stick{several angles and redans, he transferred}
\stick{the lines from his paper; then taking}
\stick{the profile of the place, with its works,}
\stick{to determine the depths and slopes of}
\stick{the ditches,\tsk the talus of the glacis,}
\stick{and the precise height of the several}
\stick{banquets, parapets, \&c.\tsk he set the}
\stick{corporal to work\tsh and sweetly went}
\stick{it on:\tsh The nature of the soil,\tsk}
\stick{the nature of the work itself,\tsk and above}
\stick{all, the good nature of my uncle \toby}
\stick{sitting by from morning to night, and}
\stick{chatting kindly with the corporal upon}
\catch{past-}
\stick{past-done deeds,\tsk left \s{labour} little else}
but the ceremony of the name.

When the place was finished in this\break
\stick{manner, and put into a proper posture}
\stick{of defence,\tsk it was invested,\tsk and my}
\stick{uncle \toby and the corporal began to run}
\stick{their first parallel.\tsk I beg I may not}
\stick{be interrupted in my story, by being}
\stick{told, \i{That the first parallel should be}}
\stick{\i{at least three hundred toises distant from}}
\stick{\i{the main body of the place,\tsk and that I}}
\stick{\i{have not left a single inch for it};\tsh for}
\stick{my uncle \toby took the liberty of in-}
\stick{croaching upon his kitchen garden, for}
\stick{the sake of enlarging his works on the}
\stick{bowling green, and for that reason ge-}
\stick{nerally ran his first and second parallels}
\stick{betwixt two rows of his cabbages and}
\stick{his collyflowers; the conveniences and}
\stick{inconveniences of which will be consi-}
\catch{dered}
\baselineskip 13.5pt
\stick{dered at large in the history of my}
\stick{uncle \toby's and the corporal’s cam-}
\stick{paigns, of which, this I’m now writing}
\stick{is but a sketch, and will be finished, if}
\stick{I conjecture right, in three pages (but}
\stick{there is no guessing)\tsh The campaigns}
\stick{themselves will take up as many books;}
\stick{and therefore I apprehend it would be}
\stick{hanging too great a weight of one kind}
\stick{of matter in so flimsy a performance as}
\stick{this, to rhapsodize them, as I once in-}
\stick{tended, into the body of the work\tsh}
\stick{surely they had better be printed apart,}
\stick{\tsh we’ll consider the affair\tsh so take}
\stick{the following sketch of them in the}
mean time.

\chap{XXII}{10pt}{8pt}

\dropcap{W}{hen} the town, with its works,\break
was finished, my uncle \i{Toby}\break
\stick{and the corporal began to run their first}
\catch{parallel}
\setlength{\baselineskip}{15pt}%
\stick{parallel \tsh not at random, or any how}
\stick{\tsh but from the same points and dis-}
\stick{tances the allies had begun to run theirs;}
\stick{and regulating their approaches and at-}
\stick{tacks, by the accounts my uncle \i{Toby}}
\stick{received from the daily papers,\tsk they}
\stick{went on, during the whole siege, step}
by step with the allies.

When the duke of \i{Marlborough} made\break
\stick{a lodgment,\tsh my uncle \toby made a}
\stick{lodgment too.\tsh And when the face of}
\stick{a bastion was battered down, or a de-}
\stick{fence ruined,\tsk the corporal took his}
\stick{mattock and did as much,\tsk and so on;}
\stick{\tsh gaining ground, and making them-}
\stick{selves masters of the works one after}
\stick{another, till the town fell into their}
hands.
\patch{To}

\setlength{\baselineskip}{14pt}%
\topstrut To one who took pleasure in the hap-\break
\stick{py state of others,\tsk there could not}
\stick{have been a greater sight in the world,}
\stick{than, on a post-morning, in which a}
\stick{practicable breach had been made by the}
\stick{duke of \i{Marlborough}, in the main body}
\stick{of the place,\tsk to have stood behind the}
\stick{horn beam hedge, and observed the spi-}
\stick{rit with which my uncle \toby, with}
\stick{\trim behind him, sallied forth;\tsh the}
\stick{one with the \i{Gazette} in his hand,\tsk the}
\stick{other with a spade on his shoulder to ex-}
\stick{ecute the contents.\tsh What an honest}
\stick{triumph in my uncle \toby's looks as he}
\stick{marched up to the ramparts! What in-}
\stick{tense pleasure swimming in his eye as he}
\stick{stood over the corporal, reading the pa-}
\stick{ragraph ten times over to him, as he was}
\stick{at work, lest, peradventure, he should}
\stick{make the breach an inch too wide,\tsk or}
\catch{leave}
\stick{leave it an inch too narrow\tshh But}
\stick{when the \i{chamade} was beat, and the cor-}
\stick{poral helped my uncle up it, and fol-}
\stick{lowed with the colours in his hand, to}
\stick{fix them upon the ramparts\tsk Heaven!}
\stick{Earth!  Sea!\tshh but what avails apos-}
\stick{trophes?\tshh with all your elements,}
\stick{wet or dry, ye never compounded so in-}
toxicating a draught.

In this track of happiness for many\break
\stick{years, without one interruption to it,}
\stick{except now and then when the wind}
\stick{continued to blow due west for a week}
\stick{or ten days together, which detained the}
\stick{\i{Flanders} mail, and kept them so long in}
\stick{torture,\tskk but still ’twas the torture of}
\stick{the happy\tshh In this track, I say, did}
\stick{my uncle \toby and \trim move for many}
\stick{years, every year of which, and some-}
\stick{times every month, from the invention}
\catch{of}
\stick{of either the one or the other of them,}
\stick{adding some new conceit or quirk of}
\stick{improvement to their operations, which}
\stick{always opened fresh springs of delight}
in carrying them on.

The first year’s campaign was carried\break
\stick{on from beginning to end, in the plain}
and simple method I’ve related.

In the second year, in which my un-\break
\stick{cle \toby took \i{Liege} and \i{Ruremond}, he}
\stick{thought he might afford the expence of}
\stick{four handsome draw-bridges, two of}
\stick{which I have given an exact description,}
in the former part of my work.

At the latter end of the same year\break
\stick{he added a couple of gates with port-}
\stick{cullises:\tsh These last were converted}
\stick{afterwards in orgues, as the better thing;}
\catch{and}
\stick{and during the winter of the same year,}
\stick{my uncle \toby, instead of a new suit of}
\stick{cloaths, which he always had at \i{Christ}-}
\stick{\i{mas}, treated himself with a handsome}
\stick{sentry-box, to stand at the corner of the}
\stick{bowling-green, betwixt which point and}
\stick{the foot of the glacis, there was left a}
\stick{little kind of an esplanade for him and}
\stick{the corporal to confer and hold councils}
of war upon.

\tshh The sentry-box was in case of\break
rain.

All these were painted white three\break
\stick{times over the ensuing spring, which}
\stick{enabled my uncle \toby to take the field}
with great splendour.

My father would often say to \i{Yorick},\break
\stick{that if any mortal in the whole universe}
\catch{had}
\stick{had done such a thing, except his bro-}
\stick{ther \toby, it would have been looked}
\stick{upon by the world as one of the most}
\stck{refined satyrs upon the parade and pranc-}
\stick{ing manner, in which \i{Lewis} XIV.\@ from}
\stick{the beginning of the war, but particu-}
\stick{larly that very year, had taken the field}
\stick{\tsh But ’tis not my brother \toby's na-}
\stick{ture, kind soul!\@ my father would add,}
to insult any one.

\tsh But let us go on.

\chap{XXIII}{10pt}{0pt}

\initial{I}{Must} observe, that although in the\break
first year’s campaign, the word \i{town}\break
\stick{is often mentioned,\tsk yet there was no}
\stick{town at that time within the polygon;}
\stick{that addition was not made till the sum-}
\stick{mer following the spring in which the}
\stick{bridges and sentry-box were painted,}
\catch{which}
\stick{which was the third year of my uncle}
\stick{\toby's campaigns,\tsk when upon his tak-}
\stick{ing \i{Amberg}, \i{Bonn}, and \i{Rhinberg}, and}
\stick{\i{Huy} and \i{Limbourg}, one after another,}
\stick{a thought came into the corporal’s head,}
\stick{that to talk of taking so many towns,}
\stick{\i{without one} \textsc{town} \i{to show for it},\tsk was a}
\stick{very nonsensical way of going to work,}
\stick{and so proposed to my uncle \toby, that}
\stick{they should have a little model of a}
\stick{town built for them,\tsk to be run up to-}
\stick{gether of slit deals, and then painted,}
\stick{and clapped within the interior polygon}
to serve for all.

My uncle \toby felt the good of the\break
\stick{project instantly, and instantly agreed to}
\stick{it, but with the addition of two sin-}
\stick{gular improvements, of which he was}
\stick{almost as proud, as if he had been the}
original inventor of the project itself.
\patch{The}

\topstrut The one was to have the town built\break
\stick{exactly in the style of those, of which it}
\stick{was most likely to be the representative:}
\stick{\tsh with grated windows, and the}
\stick{gable ends of the houses, facing the}
\stick{streets, \&c. \&c.\tsk as those in \i{Ghent} and}
\stick{\i{Bruges}, and the rest of the towns in}
\i{Brabant} and \i{Flanders}.

The other was, not to have the houses\break
\stick{run up together, as the corporal propos-}
\stick{ed, but to have every house indepen-}
\stick{dant, to hook on, or off, so as form}
\stick{into the plan of whatever town they}
\stick{pleased. This was put directly into}
\stick{hand, and many and many a look of}
\stick{mutual congratulation was exchanged}
\stick{between my uncle \toby and the corpo-}
ral, as the carpenter did the work.
\patch{\tsh It}

\tsh\topstrut It answered prodigiously the next\break
\stick{summer\tsh the town was a perfect \i{Pro}-}
\stick{\i{teus}\tsh It was \i{Landen}, and \i{Trerebach},}
\stick{and \i{Santvliet}, and \i{Drusen}, and \i{Hagenau},}
\stick{\tsk and then it was \i{Ostend} and \i{Menin}, and}
\i{Aeth} and \i{Dendermond}.\tsh

\tsh Surely never did any \s{town} act\break
\stick{so many parts, since \i{Sodom} and \i{Gomor}-}
\i{rah}, as my uncle \toby's town did.

In the fourth year, my uncle \toby\break
\stick{thinking a town looked foolishly with-}
\stick{out a church, added a very fine one}
\stick{with a steeple.\tsh \trim was for having}
\stick{bells in it;\tsh my uncle \toby said, the}
mettle had better be cast into cannon.

\tight{This led the way the next campaign for}\break
\stick{half a dozen brass field pieces,\tsk to be}
\catch{planted}
\stck{planted three and three on each side of my}
\stick{uncle \toby's sentry-box; and in a short}
\stck{time, these led the way for a train of some-}
\stick{what larger,\tsk and so on\tsk (as must al-}
\stck{ways be the case in hobby-horsical affairs)}
\stick{from pieces of half an inch bore, till it}
came at last to my father’s jack boots.

The next year, which was that in\break
\stick{which \i{Lisle} was besieged, and at the}
\stick{close of which both \i{Ghent} and \i{Bruges}}
\stick{fell into our hands,\tsk my uncle \toby was}
\stick{sadly put to it for \i{proper} ammunition;}
\stick{\tsh I say proper ammunition\tsh be-}
\stick{cause his great artillery would not bear}
\stick{powder; and ’twas well for the \i{Shandy}}
\stick{family they would not\tsh For so full}
\stick{were the papers, from the beginning to}
\stick{the end of the siege, of the incessant}
\stick{firings kept up by the besiegers,\tsh and}
\stick{so heated was my uncle \toby's imagina-}
\catch{tion}
\stick{tion with the accounts of them, that he}
had infallibly shot away all his estate.

\s{Something} therefore was wanting,\break
\stick{as a \i{succedaneum}, especially in one or}
\stick{two of the more violent paroxysms of}
\stick{the siege, to keep up something like a}
\stick{continual firing in the imagination,\tsh}
\stick{and this \i{something}, the corporal, whose}
\stick{principal strength lay in invention, sup-}
\stick{plied by an entire new system of batter-}
\stick{ing of his own,\tsk without which, this}
\stick{had been objected to by military critics,}
\stick{to the end of the world, as one of the}
\stick{great \i{desiderata} of my uncle \toby's ap-}
paratus.

This will not be explained the worse,\break
\stick{for setting off, as I generally do, at a}
little distance from the subject.

\patch{\lsss{CHAP}.}
\baselineskip 13.5pt
\parskip 8pt
\chap{XXIV}{28pt}{8pt}

\dropcap{W}{ith} two or three other trinkets,\break
small in themselves, but of great\break
\stick{regard, which poor \i{Tom}, the corporal’s}
\stick{unfortunate brother, had sent him over,}
\stick{with the account of his marriage with}
the \i{Jew}’s widow\tsh there was

\tight{A \i{Montero}-cap and two \i{Turkish} tobacco}\break
pipes.

The \i{Montero}-cap I shall describe by\break
\stick{and bye.\tsk The \i{Turkish} tobacco pipes}
\stick{had nothing particular in them, they}
\stick{were fitted up and ornamented as usual,}
\stick{with flexible tubes of \i{Morocco} leather}
\stick{and gold wire, and mounted at their}
\stick{ends, the one of them with ivory,\tsk the}
\stick{other with black ebony, tipp’d with}
\stick{silver.\hfill}
\catch{My}
\baselineskip 14pt
\parskip 10pt
\stick{\indent\tstrut 20pt My father, who saw all things in}
\stick{lights different from the rest of the}
\stick{world, would say to the corporal, that}
\stick{he ought to look upon these two presents}
\stick{more as tokens of his brother’s nicety,}
\stick{than his affection.\tsh \i{Tom} did not care,}
\stick{\trim, he would say, to put on the cap, or}
\stick{to smoak in the tobacco-pipe of a \i{Jew}.}
\stick{\tsh God bless your honour, the corpo-}
\stick{ral would say, (giving a strong reason}
to the contrary)\tsk how can that be.\tsh

The Montero-cap was scarlet, of a\break
\stick{superfine \i{Spanish} cloth, died in grain,}
\stick{and mounted all round with furr, except}
\stick{about four inches in the front, which}
\stick{was faced with a light blue, slightly em-}
\stick{broidered,\tsk and seemed to have been}
\stick{the property of a \i{Portuguese} quarter-}
\catch{master,}
\setlength{\baselineskip}{14.7pt}% 294/20 = 14.7  294/21 = 14
\setlength{\parskip}{8pt plus 2pt minus 2pt}%
\stick{master, not of foot, but of horse, as}
the word denotes.

The corporal was not a little proud\break
\stick{of it, as well for its own sake, as the}
\stick{sake of the giver, so seldom or never}
\stick{put it on but upon \s{Gala}-days; and}
\stick{yet never was a Montero-cap put to so}
\stick{many uses; for in all controverted points,}
\stick{whether military or culinary, provided}
\stick{the corporal was sure he was in the right,}
\stick{\tsk it was either his \i{oath},\tsk his \i{wager},\tsk}
or his \i{gift}.

\tsh ’Twas his gift in the present case.

I’ll be bound, said the corporal, speak-\break
\stick{ing to himself, to \i{give} away my Montero-}
\stick{cap to the first beggar who comes to the}
\stick{door, if I do not manage this matter to}
his honour’s satisfaction.
\patch{The}
\baselineskip 13.5pt
\parskip 12pt

The completion was no further off,\break
\stick{than the very next morning; which was}
\stick{that of the storm of the counterscarp be-}
\stick{twixt the \i{Lower Deule}, to the right, and}
\stick{the gate St.\@ \i{Andrew},\tsk and on the left,}
between St.\@ \i{Magdalen}’s and the river.

As this was the most memorable at-\break
\stick{tack in the whole war,\tsk the most gallant}
\stick{and obstinate on both sides,\tsk and I must}
\stick{add the most bloody too, for it cost the}
\stick{allies themselves that morning above}
\stick{eleven hundred men,\tsk my uncle \i{Toby}}
\stick{prepared himself for it with a more than}
ordinary solemnity.

The eve which preceded, as my un-\break
\stick{cle \toby went to bed, he ordered his}
\stick{ramallie wig, which had laid inside out}
\stick{for many years in the corner of an old}
\stick{campaigning trunk, which stood by his}
\catch{bed-}
\stick{bedside, to be taken out and laid upon}
\stick{the lid of it, ready for the morning;\tsk}
\stick{and the very first thing he did in his}
\stick{shirt, when he had stepped out of bed,}
\stick{my uncle \toby, after he had turned the}
\stick{rough side outwards,\tsk put it on:\tsh}
\stick{This done, he proceeded next to his}
\stick{breeches, and having buttoned the waist-}
\stick{band, he forthwith buckled on his sword}
\stick{belt, and had got his sword half way in,}
\stick{\tsk when he considered he should want}
\stick{shaving, and that it would be very in-}
\stick{convenient doing it with his sword on,\tsk}
\stick{so took it off:\tsh In assaying to put on}
\stick{his regimental coat and waistcoat, my}
\stick{uncle \toby found the same objection in}
\stick{his wig,\tsk so that went off too:\tsk So that}
\stick{what with one thing, and what with}
\stick{another, as always falls out when a man}
\stick{is in the most haste,\tsk ’twas ten o’clock,}
\stick{which was half an hour later than his}
\catch{usual}
\baselineskip 12.3333pt
\parskip 10pt
\stick{usual time, before my uncle \toby sallied}
out.

\chap{XXV}{12pt}{6pt}

\dropcap{M}{y} uncle \toby had scarce turned\break
the corner of his yew hedge,\break
\stick{which separated his kitchen garden from}
\stick{his bowling green, when he perceived}
\stick{the corporal had begun the attack with-}
out him.\tsh

Let me stop and give you a picture of\break
\stick{the corporal’s apparatus; and of the}
\stick{corporal himself in the height of this at-}
\stick{tack just as it struck my uncle \toby, as}
\stick{he turned towards the sentry box, where}
\stick{the corporal was at work,\tshh for in}
\stick{nature there is not such another,\tshh}
\stick{nor can any combination of all that is}
\stick{grotesque and whimsical in her works}
produce its equal.

The corporal\tush

\patch{\tsh Tread}

\topstrut\tshh Tread lightly on his ashes, ye\break
\stick{men of genius,\tsk for he was your}
kinsman:

Weed his grave clean, ye men of\break
\stick{goodness,\tsk for he was your brother.\tsk}
\stick{Oh corporal! had I thee, but now,\tsk}
\stick{now, that I am able to give thee a din-}
\stick{ner and protection,\tsk how would I che-}
\stick{rish thee! thou should’st wear thy Mon-}
\stick{tero-cap every hour of the day, and}
\stick{every day of the week,\tsk and when it}
\stick{was worn out, I would purchase thee a}
\stick{couple like it:\tsh But alas! alas! alas!}
\stick{now that I can do this, in spight of their}
\stick{reverences\tsk the occasion is lost\tsk for}
\stick{thou art gone;\tsk thy genius fled up to}
\stick{the stars from whence it came;\tsk and}
\stick{that warm heart of thine, with all its}
\stick{generous and open vessels, compressed}
into a \i{clod of the valley!}
\patch[10pt]{\tsh But}

\noindent\enspace\topstrut\tshh But what\tsh what is this, to that\break
\stick{future and dreaded page, where I look}
\stick{towards the velvet pall, decorated with}
\stick{the military ensigns of thy master\tsk the}
\stick{first\tsk the foremost of created beings;}
\stick{\tsh where, I shall see thee, faithful ser-}
\stck{vant! laying his sword and scabbard with}
\stick{a trembling hand across his coffin, and}
\stick{then returning pale as ashes to the door,}
\stick{to take his mourning horse by the bridle,}
\stick{to follow his hearse, as he directed thee;}
\stick{\tsh where\tsk all my father’s systems shall}
\stick{be baffled by his sorrows; and, in spight}
\stick{of his philosophy, I shall behold him,}
\stick{as he inspects the lackered plate, twice}
\stick{taking his spectacles from off his nose,}
\stick{to wipe away the dew which nature has}
\stick{shed upon them\tshh When I see him}
\stck{cast in the rosemary with an air of discon-}
\stick{solation, which cries through my ears,}
\stick{\tshh O \i{Toby!} in what corner of the}
world shall I seek thy fellow?

\patch{\tsh Gra-}
\topstrut\tshh Gracious powers! which erst\break
\stick{have opened the lips of the dumb in his}
\stick{distress, and made the tongue of the}
\stick{stammerer speak plain\tsk when I shall}
\stick{arrive at this dreaded page, deal not}
with me, then, with a stinted hand.

\chap{XXVI}{18pt}{8pt}

\dropcap{T}{he} corporal, who the night be-\break
fore had resolved in his mind, to\break
\stick{supply the grand \i{desideratum}, of keeping}
\stick{up something like an incessant firing}
\stick{upon the enemy during the heat of the}
\stick{attack,\tsk had no further idea in his fancy}
\stick{at that time, than a contrivance of smoak-}
\stick{ing tobacco against the town, out of}
\stick{one of my uncle \toby's six field pieces,}
\stick{which were planted on each side of his}
\stick{sentry-box; the means of effecting which}
\stick{occurring to his fancy at the same time,}
\stick{though he had pledged his cap, he}
\catch{thought}
\stick{thought it in no danger from the miscar-}
riage of his projects.

Upon turning it this way, and that,\break
\stick{a little in his mind, he soon began to}
\stick{find out, that by means of his two \i{Tur}-}
\stick{\i{kish} tobacco-pipes, with the supplement}
\stick{of three smaller tubes of wash-leather at}
\stick{each of their lower ends, to be tagg’d}
\stick{by the same number of tin pipes fitted}
\stick{to the touch holes, and sealed with clay}
\stick{next the cannon, and then tied herme-}
\stick{tically with waxed silk at their several}
\stick{insertions into the \i{Morocco} tube,\tsk he}
\stick{should be able to fire the six field pieces}
\stick{all together, and with the same ease as}
to fire one.\tush

\tsh Let no man say from what taggs\break
\stick{and jaggs hints may not be cut out for}
\stick{the advancement of human knowledge.}
\stick{Let no man, who has read my father’s}
\stick{first and second \i{beds of justice}, ever rise}
\stick{up and say again, from collision of what}
\catch{kinds}
\stick{kinds of bodies, light may, or may not}
\stick{be struck out, to carry the arts and sci-}
\stick{ences up to perfection.\tshh Heaven!}
\stick{thou knowest how I love them;\tshh thou}
\stick{knowest the secrets of my heart, and}
\stick{that I would this moment give my shirt}
\stick{\tshh Thou art a fool, \i{Shandy}, says \i{Eu}-}
\stick{\i{genius},\tsk for thou hast but a dozen in the}
world,\tsk and ’twill break thy set.\tsh

No matter for that, \i{Eugenius}; I would\break
\stick{give the shirt off my back to be burnt}
\stick{into tinder, were it only to satisfy one}
\stick{feverish enquirer, how many sparks at}
\stick{one good stroke, a good flint and steel}
\stick{could strike into the tail of it.\tshh}
\stick{Think ye not that in striking these \i{in},\tsk}
\stick{he might, peradventure, strike something}
\i{out?} as sure as a gun.\tsh

\tsh But this project, by the bye.

The corporal sat up the best part of\break
\stick{the night in bringing \i{his} to perfection;}
\catch{and}
\parskip 8pt
\stick{and having made a sufficient proof of}
\stick{his cannon, with charging them to the}
\stick{top with tobacco,\tsk he went with con-}
tentment to bed.

\chap{XXVII}{10pt}{0pt}

\dropcap{T}{he} corporal had slipped out about\break
ten minutes before my uncle \toby,\break
\stick{in order to fix his apparatus, and just}
\stick{give the enemy a shot or two before my}
uncle \toby came.

He had drawn the six field-pieces for\break
\stick{this end, all close up together in front}
\stick{of my uncle \toby's sentry-box, leaving}
\stick{only an interval of about a yard and a}
\stick{half betwixt the three, on the right and}
\stick{left, for the convenience of charging,}
\stick{\&c.\tsk and the sake possibly of two bat-}
\stick{teries, which he might think double the}
honour of one.

In the rear, and facing this opening,\break
with his back to the door of the sentry-
\catch{box,}
\stick{box, for fear of being flanked, had the}
\stick{corporal wisely taken his post:\tsh He}
\stick{held the ivory pipe, appertaining to the}
\stick{battery on the right, betwixt the finger}
\stick{and thumb of his right hand,\tsk and the}
\stick{ebony pipe tipp’d with silver, which ap-}
\stick{pertained to the battery on the left, be-}
\stick{twixt the finger and thumb of the other}
\stick{\tsh and with his right knee fixed firm}
\stick{upon the ground, as if in the front rank}
\stick{of his platoon, was the corporal, with}
\stick{his montero-cap upon his head, furiously}
\stick{playing off his two cross batteries at the}
\stick{same time against the counterguard,}
\stick{which faced the counterscarp, where the}
\stick{attack was to be made that morning.}
\stick{His first intention, as I said, was no}
\stick{more than giving the enemy a single puff}
\stick{or two;\tsk but the pleasure of the \i{puffs},}
\stick{as well as the \i{puffing}, had insensibly got}
\stick{hold of the corporal, and drawn him on}
\stick{from puff to puff, into the very height}
\catch{of}
\stick{of the attack, by the time my uncle \toby}
joined him.

’Twas well for my father, that my\break
\stick{uncle \toby had not his will to make that}
day.

\chap{XXVIII}{18pt}{10pt}

\dropcap[findent=1pt]{M}{y} uncle \toby took the ivory pipe\break
out of the corporal’s hand,\tskk\break
\stick{looked at it for half a minute, and re-}
turned it.

In less than two minutes my uncle\break
\stick{\toby took the pipe from the corporal}
\stick{again, and raised it half way to his mouth}
\stick{\tsh then hastily gave it back a second}
time.

The corporal redoubled the attack,\break
\stick{\tsh my uncle \toby smiled,\tsh then}
\stick{looked grave,\tsh then smiled for a mo-}
\stick{ment,\tsh then looked serious for a long}
\catch{time;}
\stick{time;\tsk Give me hold of the ivory}
\stick{pipe, \trim, said my uncle \i{Toby}\tsk my}
\stick{uncle \toby put it to his lips,\tsk drew}
\stick{it back directly,\tsk gave a peep over}
\stick{the horn-beam hedge;\tsk never did}
\stick{my uncle \toby's mouth water so much}
\stick{for a pipe in his life.\tsk My uncle \toby}
\stick{retired into the sentry-box with the pipe}
in his hand.\tsh

\tshh Dear uncle \i{Toby!} don’t go into\break
\stick{the sentry-box with the pipe,\tsk there’s}
\stick{no trusting a man’s self with such a}
thing in such a corner.

\chap{XXIX}{14pt}{8pt}

\initial{I}{Beg} the reader will assist me here, to\break
wheel off my uncle \toby's ordnance\break
\stick{behind the scenes,\tsk to remove his}
\stick{sentry-box, and clear the theatre, \i{if pos}-}
\stick{\i{sible}, of horn-works and half moons,}
\stick{and get the rest of his military appa-}
\stick{ratus out of the way;\tsk that done,}
\catch{my}
\parskip 12pt
\stick{my dear friend \i{Garrick}, we’ll snuff the}
\stick{candles bright,\tsk sweep the stage with a}
\stick{new broom,\tsk draw up the curtain, and}
\stick{exhibit my uncle \toby dressed in a new}
\stick{character, throughout which the world}
\stick{can have no idea how he will act: and}
\stick{yet, if pity be akin to love,\tsk and bra-}
\stick{very no alien to it, you have seen enough}
\stick{of my uncle \toby in these, to trace these}
\stick{family likenesses, betwixt the two passi-}
\stick{ons (in case there is one) to your heart’s}
content.

Vain science! thou assists us in no
\stick{case of this kind\tsk and thou puzzlest us}
in every one.

There was, Madam, in my uncle\break
\stick{\toby, a singleness of heart which misled}
\stick{him so far out of the little serpentine}
\stick{tracks in which things of this nature}
\stick{usually go on; you can\tsk you can have}
\stick{no conception of it: with this, there}
\catch{was}
\stick{was a plainness and simplicity of think-}
\stick{ing, with such an unmistrusting igno-}
\stick{rance of the plies and foldings of the}
\stick{heart of woman;\tsh and so naked and}
\stick{defenceless did he stand before you,}
\stick{(when a siege was out of his head) that}
\stick{you might have stood behind any one of}
\stick{your serpentine walks, and shot my un-}
\stick{cle \toby ten times in a day, through his}
\stick{liver, if nine times in a day, Madam,}
had not served your purpose.

With all this, Madam,\tsk and what\break
\stick{confounded every thing as much on the}
\stick{other hand, my uncle \toby had that un-}
\stick{paralleled modesty of nature I once told}
\stick{you of, and which, by the bye, stood}
\stick{eternal sentry upon his feelings, that}
\stick{you might as soon\tsh But where am I}
\stck{going? these reflections croud in upon me}
\stick{ten pages at least too soon, and take up}
\stick{that time, which I ought to bestow upon}
facts.

\patch{\lsss{CHAP}.}

\chap{XXX}{26pt}{8pt}

\dropcap[findent=1pt]{O}{f} the few legitimate sons of \i{Adam},\break
whose breasts never felt what the\break
\stick{sting of love was,\tsk (maintaining first,}
\stick{all mysogynists to be bastards)\tsk the}
\stick{greatest heroes of ancient and modern}
\stck{story have carried off amongst them, nine}
\stick{parts in ten of the honour; and I wish}
\stick{for their sakes I had the key of my}
\stick{study out of my draw-well, only for}
\stick{five minutes, to tell you their names\tsk}
\stick{recollect them I cannot\tsk so be content}
\stick{to accept of these, for the present, in}
their stead.\tush

There was the great king \i{Aldrovandus},\break
\stick{and \i{Bosphorus}, and \i{Cappadocius}, and \i{Dar}-}
\stick{\i{danus}, and \i{Pontus}, and \i{Asius},\tsk to}
\stick{say nothing of the iron-hearted \i{Charles}}
\stick{the XIIth, whom the Countess of K\raise.3ex\hbox{\small\nastv}}
\stick{herself could make nothing of.\tsk}
\catch{There}
\stck{There was \i{Babylonicus}, and \i{Mediterraneus},}
\stick{and \i{Polixenes}, and \i{Persicus}, and \i{Prusicus},}
\stick{not one of whom (except \i{Cappadocius}}
\stick{and \i{Pontus}, who were both a little sus-}
\stick{pected) ever once bowed down his breast}
\stick{to the goddess\tsk The truth is, they}
\stick{had all of them something else to do\tsk}
\stick{and so had my uncle \i{Toby}\tsk till Fate\tsk}
\stick{till Fate I say, envying his name the}
\stick{glory of being handed down to posterity}
\stick{with \i{Aldrovandus}’s and the rest,\tsk she}
basely patched up the peace of \i{Utrecht}.

\tsh Believe me, Sirs, ’twas the worst\break
deed she did that year.

\chap{XXXI}{14pt}{6pt}

\dropcap{A}{mongst} the many ill conse-\break
quences of the treaty of \i{Utrecht},\break
\stick{it was within a point of giving my uncle}
\stick{\toby a surfeit of sieges; and though he}
\stick{recovered his appetite afterwards, yet}
\catch{\i{Calais}}
\stick{\i{Calais} itself left not a deeper scar in}
\stick{\i{Mary}’s heart, than \i{Utrecht} upon my}
\stick{uncle \toby's. To the end of his life he}
\stick{never could hear \i{Utrecht} mentioned upon}
\stick{any account whatever,\tsk or so much as}
\stick{read an article of news extracted out of}
\stick{the \i{Utrecht Gazette}, without fetching a}
\stick{sigh, as if his heart would break in}
twain.

My father, who was a great \s{motive}-\break
\stick{\s{monger}, and consequently a very dan-}
\stick{gerous person for a man to sit by, either}
\stick{laughing or crying,\tsk for he generally}
\stick{knew your motive for doing both, much}
\stick{better than you knew it yourself\tsk would}
\stck{always console my uncle \toby upon these}
\stick{occasions, in a way, which shewed}
\stick{plainly, he imagined my uncle \i{Toby}}
\stick{grieved for nothing in the whole affair,}
\stick{so much as the loss of his \i{hobby-horse.}}
\stick{\tsk Never mind, brother \toby, he}
\stick{would say,\tsk by God’s blessing we shall}
\catch{have}
\parskip 10pt
\stick{have another war break out again some}
\stick{of these days; and when it does,\tsk the}
\stick{belligerent powers, if they would hang}
\stick{themselves, cannot keep us out of play.}
\stick{\tsh I defy ’em, my dear \toby, he would}
\stick{add, to take countries without taking}
towns,\tsh or towns without sieges.

My uncle \toby never took this back-\break
\stick{stroke of my father’s at his hobby horse}
\stick{kindly.\tsh He thought the stroke unge-}
\stick{nerous; and the more so, because in}
\stick{striking the horse, he hit the rider too,}
\stick{and in the most dishonourable part a}
\stick{blow could fall; so that upon these oc-}
\stick{casions, he always laid down his pipe}
\stick{upon the table with more fire to defend}
himself than common.

I told the reader, this time two years,\break
\stick{that my uncle \toby was not eloquent;}
\stick{and in the very same page gave an in-}
\stick{stance to the contrary:\tshh I repeat the}
\stick{observation, and a fact which contradicts}
\catch{it}
\baselineskip 13pt
\parskip 13pt
\stick{it again.\tsk He was not eloquent,\tsk it was}
\stick{not easy to my uncle \toby to make long}
\stick{harangues,\tsk and he hated florid ones;}
\stck{but there were occasions where the stream}
\stick{overflowed the man, and ran so counter}
\stick{to its usual course, that in some parts}
\stick{my uncle \toby, for a time, was at least}
\stick{equal to \i{Tertullus}\tshh but in others, in}
my own opinion, infinitely above him.

My father was so highly pleased with\break
\stick{one of these apologetical orations of my}
\stick{uncle \toby's, which he had delivered one}
\stick{evening before him and \i{Yorick}, that he}
wrote it down before he went to bed.

I have had the good fortune to meet\break
\stick{with it amongst my father’s papers, with}
\stick{here and there an insertion of his own,}
\stick{betwixt two crooks, thus \hbox{\Large [\quad ],} and is \tstrut 13pt}
endorsed,\tstrut 13pt
\patch{\i{My}}

\noindent
\stick{\tstrut 16pt\i{My brother} \textsc{Toby}’\i{s justification of his}}
\stick{\indent\i{own principles and conduct in wishing to}}
\stick{\indent\i{continue the war.\hfill}}
\stick{\tstrut 18pt I may safely say, I have read over this}
\stick{apologetical oration of my uncle \toby's}
\stick{a hundred times, and think it so fine a}
\stick{model of defence,\tsk and shews so sweet a}
\stick{temperament of gallantry and good prin-}
\stick{ciples in him, that I give it the world,}
\stick{word for word, (interlineations and all)}
\stick{as I find it.\hfill}
\chap{XXXII}{14pt}{0pt}
\centerline{\tstrut 16pt \i{My  uncle} \s{Toby}’\i{s apologetical  oration}.}

\vskip -10pt

\initial[nindent=1pt]{I}{Am} not insensible, brother \i{Shandy},\break
that when a man, whose profession\break
\stick{is arms, wishes, as I have done, for}
\stick{war,\tsk it has an ill aspect to the world;}
\stick{\tsh and that, how just and right soever}
\stick{his motives and intentions may be,\tsk he}
\stck{stands in an uneasy posture in vindicating}
himself from private views in doing it.

\patch{For}
\baselineskip 12.5pt
\parskip 8pt
\topstrut For this cause, if a soldier is a prudent\break
\stick{man, which he may be, without being}
\stick{a jot the less brave, he will be sure not}
\stick{to utter his wish in the hearing of an}
\stick{enemy; for say what he will, an enemy}
\stick{will not believe him.\tsk He will be}
\stick{cautious of doing it even to a friend,\tsk}
\stick{lest he may suffer in his esteem:\tsk But}
\stick{if his heart is overcharged, and a secret}
\stick{sigh for arms must have its vent, he}
\stick{will reserve it for the ear of a brother,}
\stick{who knows his character to the bottom,}
\stick{and what his true notions, dispositions,}
\stick{and principles of honour are: What, I}
\stick{\i{hope}, I have been in all these, brother}
\stick{\i{Shandy}, would be unbecoming in me to}
\stick{say:\tsk much worse, I know, have I}
\stick{been than I ought,\thinspace\tskk\thinspace and something}
\stick{worse, perhaps, than I think: But such}
\stick{as I am, you, my dear brother \i{Shandy},}
\stick{who have sucked the same breasts with}
\stick{me,\thinspace\tskk\thinspace and with whom I have been}
\catch{brought}
\stick{brought up from my cradle,\tsk and from}
\stick{whose knowledge, from the first hours of}
\stick{our boyish pastimes, down to this, I}
\stick{have concealed no one action of my life,}
\stick{and scarce a thought in it\tsh Such as I}
\stick{am, brother, you must by this time}
\stick{know me, with all my vices, and with}
\stick{all my weaknesses too, whether of my}
\stick{age, my temper, my passions, or my}
understanding.

Tell me then, my dear brother \i{Shandy},\break
\stick{upon which of them it is, that when I}
\stick{condemned the peace of \i{Utrecht}, and}
\stick{grieved the war was not carried on with}
\stick{vigour a little longer, you should think}
\stick{your brother did it upon unworthy}
\stick{views; or that in wishing for war, he}
\stick{should be bad enough to wish more of}
\stick{his fellow creatures slain,\tsk more slaves}
\stick{made, and more families driven from}
\stick{their peaceful habitations, merely for his}
\stick{own pleasure:\,\tsk Tell me, brother}
\catch{\i{Shandy},}
\stick{\i{Shandy}, upon what one deed of mine do}
\stick{you ground it? [\i{The devil a deed do I}}
\stick{\i{know of, dear} Toby, \i{but one for a hun}-}
\stick{\i{dred pounds, which I lent thee to carry on}}
\i{these cursed sieges.}]

\medskip

If, when I was a school-boy, I could\break
\stick{not hear a drum beat, but my heart beat}
\stick{with it\tsk was it my fault?\tshh Did I}
\stick{plant the propensity there?\tshh did I}
sound the alarm within, or Nature?

\medskip

When \i{Guy}, Earl of \i{Warwick}, and\break
\stick{\i{Parismus} and \i{Parismenus}, and \i{Valentine}}
\stick{and \i{Orson}, and the \i{Seven Champions of}}
\stick{\i{England}, were handed around the school,}
\stick{\tsk were they not all purchased with my}
\stick{own pocket money? Was that selfish,}
\stick{brother \i{Shandy?} When we read over the}
\stick{siege of \i{Troy}, which lasted ten years}
\stick{and eight months,\tsh though with such}
\stick{a train of artillery as we had at \i{Namur},}
\stick{the town might have been carried in a}
\catch{week}
\stick{week\tsk was I not as much concerned for}
\stick{the destruction of the \i{Greeks} and \i{Trojans}}
\stick{as any boy of the whole school? Had I}
\stick{not three strokes of a ferula given me,}
\stick{two on my right hand and one on my}
\stick{left, for calling \i{Helena} a bitch for it?}
\stick{Did any one of you shed more tears for}
\stick{\i{Hector?} And when king \i{Priam} came to}
\stick{the camp to beg his body, and returned}
\stick{weeping back to \i{Troy} without it,\tsk you}
\stick{know, brother, I could not eat my}
dinner.\tush

\medskip

\tshh Did that bespeak me cruel? Or\break
\stick{because, brother \i{Shandy}, my blood flew}
\stick{out into the camp, and my heart panted}
\stick{for war,\tsk was it a proof it could not}
ache for the distresses of war too?

\medskip

O brother! ’tis one thing for a sol-\break
\stick{dier to gather laurels,\tsk and ’tis another}
\stick{to scatter cypress.\tsk [\i{Who told thee},}
\catch{\i{my}}
\stick{\i{my dear} Toby, \i{that cypress was used by}}
\i{the ancients on mournful occasions?} ]

\tsh ’Tis one thing, brother \i{Shandy},\break
\stick{for a soldier to hazard his own life\tsk to}
\stick{leap first down into the trench, where he}
\stick{is sure to be cut in pieces:\tsh ’Tis one}
\stick{thing, from public spirit and a thirst of}
\stick{glory, to enter the breach the first man,}
\stick{\tsk to stand in the foremost rank, and}
\stick{march bravely on with drums and trum-}
\stick{pets, and colours flying about his ears:}
\stick{\tsk ’Tis one thing, I say, brother}
\stick{\i{Shandy}, to do this\tsk and ’tis another}
\stick{thing to reflect on the miseries of war;}
\stick{\tsk to view the desolations of whole coun-}
\stick{tries, and consider the intolerable fa-}
\stick{tigues and hardships which the soldier}
\stick{himself, the instrument who works them,}
\stick{is forced (for six-pence a day, if he can}
get it) to undergo.

Need I be told, dear \i{Yorick}, as I\break
\stick{was by you, in \i{Le Fever}’s funeral ser-}
\catch{mon,}
\stick{mon, \i{That so soft and gentle a creature},}
\stick{\i{born to love}, \i{to mercy}, \i{and kindness}, \i{as}}
\stick{\i{man is}, \i{was not shaped for this?}\tsh But}
\stick{why did you not add, \i{Yorick},\tsk if not by}
\stick{\s{nature}\tsk that he is so by \s{necessity}?}
\stick{\tshh For what is war? what is it, \i{Yorick},}
\stick{when fought as ours has been, upon}
\stick{principles of \i{liberty}, and upon principles}
\stick{of \i{honour}\tsk what is it, but the getting}
\stick{together of quiet and harmless people,}
\stick{with their swords in their hands, to keep}
\stick{the ambitious and the turbulent within}
\stick{bounds? And heaven is my witness,}
\stick{brother \i{Shandy}, that the pleasure I have}
\stick{taken in these things,\tsk and that infinite}
\stick{delight, in particular, which has attend-}
\stick{ed my sieges in my bowling green, has}
\stick{arose within me, and I hope in the cor-}
\stick{poral too, from the consciousness we}
\stick{both had, that in carrying them on, we}
\stick{were answering the great ends of our}
creation.
\patch[10pt]{\lsss{CHAP}.}

\chap{XXXIII}{26pt}{10pt}

\initial{I}{Told} the Christian reader\tsk I say\break
\i{Christian}\tsh hoping he is one\tsh\break
\stick{and if he is not, I am sorry for it\tshh}
\stick{and only beg he will consider the matter}
\stick{with himself, and not lay the blame}
entirely upon this book,\tush

\medskip

I told him, Sir\tsh for in good truth,\break
\stick{when a man is telling a story in the}
\stick{strange way I do mine, he is obliged}
\stick{continually to be going backwards and}
\stick{forwards to keep all tight together in the}
\stick{reader’s fancy\tsh which, for my own}
\stick{part, if I did not take heed to do more}
\stick{than at first, there is so much unfixed}
\stick{and equivocal matter starting up, with}
\stick{so many breaks and gaps in it,\tsk and so}
\stick{little service do the stars afford, which,}
\stick{nevertheless, I hang up in some of the}
\stick{darkest passages, knowing that the}
\catch{world}
\stick{world is apt to lose its way, with all the}
\stick{lights the sun itself at noon day can}
\stick{give it\tsh and now, you see, I am lost}
myself!\tush

\tshh But ’tis my father’s fault; and\break
\stck{whenever my brains come to be dissected,}
\stick{you will perceive, without spectacles,}
\stick{that he has left a large uneven thread, as}
\stick{you sometimes see in an unsaleable piece}
\stick{of cambrick, running along the whole}
\stick{length of the web, and so untowardly,}
\stick{you cannot so much as cut out a \fnast\;\fnast,}
\stick{(here I hang up a couple of lights again)}
\stick{\tshh or a fillet, or a thumb-stall, but it}
is seen or felt.\tush

\i{Quanto id diligentias\sic in liberis procrean}-\break
\stick{\i{dis cavendum}, sayeth \i{Cardan}. All which}
\stick{being considered, and that you see ’tis}
\stick{morally impracticable for me to wind}
this round to where I set out\tsh

I begin the chapter over again.
\patch{\lsss{CHAP}.}

\chap{XXXIV}{26pt}{10pt}

\initial[nindent=0pt]{I}{Told} \tight{the Christian reader in the be-}\break
\tight{ginning of the chapter which preceded}\break
\stick{my uncle \toby's apologetical oration,\tsk}
\stick{though in a different trope from what}
\stick{I should make use of now, That the peace}
\stick{of \i{Utrecht} was within an ace of creating}
\stick{the same shyness betwixt my uncle \toby}
\stick{and his hobby-horse, as it did betwixt}
\stick{the queen and the rest of the confede-}
rating powers.

\medskip

There is an indignant way in which a\break
\stick{man sometimes dismounts his horse,}
\stick{which as good as says to him, \lqq I’ll go}
\stick{\lqq afoot, Sir, all the days of my life,}
\stick{\lqq before I would ride a single mile upon}
\stick{\lqq your back again.” Now my uncle}
\stick{\toby could not be said to dismount his}
\stick{horse in this manner; for in strictness of}
\stick{language, he could not be said to dis-}
\catch{mount}
\stick{mount his horse at all\tsh his horse ra-}
\stick{ther flung him\tsk and somewhat \i{vi}-}
\stick{\i{ciously}, which made my uncle \toby take}
\stick{it ten times more unkindly. Let this}
\stick{matter be settled by state jockies as they}
\stick{like.\tsk It created, I say, a sort of shy-}
\stick{ness betwixt my uncle \toby and his}
\stick{hobby-horse.\tsk He had no occasion}
\stick{for him from the month of \i{March} to}
\stick{\i{November}, which was the summer after}
\stick{the articles were signed, except it was}
\stick{now and then to take a short ride out,}
\stick{just to see that the fortifications and har-}
\stick{bour of \i{Dunkirk} were demolished, ac-}
cording to stipulation.

\smallskip

The \i{French} were so backwards all that\break
\stick{summer in setting about that affair, and}
\stick{Monsieur \i{Tugghe}, the deputy from the}
\stck{magistrates of \i{Dunkirk}, presented so many}
\stck{affecting petitions to the queen,\tsk beseech-}
\stck{ing her majesty to cause only her thunder-}
\stck{bolts to fall upon the martial works, which}
\catch{might}
\stick{might have incurred her displeasure,\tsk}
\stick{but to spare\tsk to spare the mole, for the}
\stick{mole’s sake; which, in its naked situa-}
\stick{tion, could be no more than an object}
\stick{of pity\tsk and the queen (who was}
\stick{but a woman) being of a pitiful dispo-}
\stick{sition,\tsk and her ministers also, they not}
\stick{wishing in their hearts to have the town}
\stck{dismantled, for these private reasons, \hbox to 14pt{\fnast\hss\fnast}}
\stick{\wastfill}
\hbox to 132pt{\vastfill{1.5em}\tush}

\medskip

\noindent
\stick{\indent\wastfill}
\stick{\wastfill}
\stick{\wastfill; so that the whole went hea-}
\stick{vily on with my uncle \toby; insomuch,}
\stick{that it was not within three full months,}
\stick{after he and the corporal had constructed}
\stick{the town, and put it in a condition to}
\stick{be destroyed, that the several com-}
\stick{mandants, commissaries, deputies, ne-}
\stick{gociators, and intendants, would permit}
\catch{him}
\stick{him to set about it.\tsh Fatal interval of}
inactivity!

\vskip -2pt

The corporal was for beginning the\break
\stick{demolition, by making a breach in the}
\stick{ramparts, or main fortifications of the}
\stick{town\tsh No,\tsk that will never do, cor-}
\stick{poral, said my uncle \toby, for in going}
\stick{that way to work with the town, the}
\stick{\i{English} garrison will not be safe in it an}
\stick{hour; because if the French are treache-}
\stick{rous\tsh They are as treacherous as de-}
\stick{vils, an’ please your honour, said the}
\stick{corporal\tsh It gives me concern always}
\stick{when I hear it, \trim, said my uncle}
\stick{\toby,\tsk for they don’t want personal bra-}
\stick{very; and if a breach is made in the}
\stick{ramparts, they may enter it, and make}
\stick{themselves masters of the place when}
\stick{they please:\tsk Let them enter it, said}
\stick{the corporal, lifting up his pioneer’s}
\stick{spade in both his hands, as if he was}
\stick{going to lay about him with it,\tsk let}
\stick{them enter, an’ please your honour, if}
\catch{they}
\stick{they dare.\tsh In cases like this, corpo-}
\stick{ral, said my uncle \toby, slipping his}
\stick{right hand down to the middle of his}
\stick{cane, and holding it afterwards trun-}
\stck{cheon-wise, with his forefinger extended,}
\stick{\tsh ’tis no part of the consideration of}
\stick{a commandant, what the enemy dare,\tsk}
\stick{or what they dare not do; he must act}
\stick{with prudence. We will begin with the}
\stick{outworks both towards the sea and the}
\stick{land, and particularly with fort \i{Louis},}
\stick{the most distant of them all, and demo-}
\stick{lish it first,\tsk and the rest, one by one,}
\stick{both on our right and left, as we retreat}
\stick{towards the town;\tsh then we’ll demo-}
\stick{lish the mole,\tsk next fill up the harbour,}
\stick{\tsk then retire into the citadel, and blow}
\stick{it up into the air; and having done}
\stick{that, corporal, we’ll embark for \i{Eng}-}
\stick{\i{land}.\tsh We are there, quoth the corpo-}
\stick{ral, recollecting himself\tsh Very true,}
\stick{said my uncle \i{Toby}\tsk looking at the}
church.

\patch{\lsss{CHAP}.}

\chap{XXXV}{28pt}{8pt}

\initial[findent=2pt]{A}{Delusive}, delicious consultation or\break
two of this kind, betwixt my uncle\break
\stick{\toby and \trim, upon the demolition of}
\stick{\i{Dunkirk},\tsk for a moment rallied back the}
\stick{ideas of those pleasures, which were slip-}
\stick{ping from under him:\tsh still\tsk still all}
\stick{went on heavily\tsk the magic left the}
\stick{mind the weaker\tsk \s{Stillness}, with \s{Si}-}
\stick{\s{lence} at her back, entered the solitary}
\stick{parlour, and drew their gauzy mantle}
\stick{over my uncle \toby's head;\tsk\!\tsk and}
\stick{\s{Listlessness}, with her lax fibre and un-}
\stick{directed eye, sat quietly down beside him}
\stick{in his arm chair.\tsh No longer \i{Amberg},}
\stick{and \i{Rhinberg}, and \i{Limbourg}, and \i{Huy},}
\stick{and \i{Bonn}, in one year,\tsk and the prospect}
\stick{of \i{Landen}, and \i{Trerebach}, and \i{Drusen}, and}
\i{Dendermond}, the next,\tsk hurried on the
\stick{blood:\tsk No longer did saps, and mines,}
\stick{and blinds, and gabions, and palisadoes,}
\catch{keep}
\stick{keep out this fair enemy of man’s re-}
\stick{pose:\tsk No more could my uncle}
\toby, after passing the \i{French} lines, as
\stick{he eat his egg at supper, from thence}
\stick{break into the heart of \i{France},\tsk cross}
\stick{over the \i{Oyes}, and with all \i{Picardie} open}
\stick{behind him, march up to the gates of}
\stick{\i{Paris}, and fall asleep with nothing but}
\stick{ideas of glory:\tsh No more was he to}
\stick{dream, he had fixed the royal standard}
\stick{upon the tower of the \i{Bastile}, and awake}
with it streaming in his head.

\medskip

\tsh Softer visions,\tsk gentler vibra-\break
\stick{tions stole sweetly in upon his slumbers;}
\stick{\tsk the trumpet of war fell out of his}
\stick{hands,\tsk he took up the lute, sweet in-}
\stick{strument! of all others the most delicate!}
\stick{the most difficult!\tsk how wilt thou}
touch it, my dear uncle \i{Toby}?

\patch[14pt]{\lsss{CHAP}.}

\chap{XXXVI}{28pt}{14pt}

\dropcap{N}{ow}, because I have once or twice\break
said, in my inconsiderate way of\break
\stick{talking, That I was confident the follow-}
\stck{ing memoirs of my uncle \toby's courtship}
\stick{of widow \i{Wadman}, whenever I got time}
\stick{to write them, would turn out one of}
\stick{the most compleat systems, both of the}
\stick{elementary and practical part of love and}
\stick{love-making, that ever was addressed to}
\stick{the world\tsk are you to imagine from}
\stick{thence, that I shall set out with a descrip-}
\stick{tion of \i{what love is?} whether part God}
\stick{and part Devil, as \i{Plotinus} will have}
it\tsh

\medskip

\tshh Or by a more critical equation,\break
\stick{and supposing the whole of love to be}
\stick{as ten\tshh to determine, with \i{Ficinus},}
\stick{\lqq\i{How many parts of it\tskk the one,\tskk and}}
\catch{\lqq \i{how}}
\stick{\lqq \i{how many the other};\,”\tskk or whether it}
\stick{is \i{all of it one great Devil}, from head to}
\stick{tail, as \i{Plato} has taken upon him to}
\stick{pronounce; concerning which conceit}
\stick{of his, I shall not offer my opinion:}
\stick{\tsk but my opinion of \i{Plato} is this; that}
\stick{he appears, from this instance, to have}
\stick{been a man of much the same temper}
\stick{and way of reasoning with doctor \i{Bayn}-}
\stick{\i{yard}, who being a great enemy to blis-}
\stick{ters, as imagining that half a dozen of}
\stick{’em on at once, would draw a man as}
\stick{surely to his grave,\, as a herse and six\tsk}
\stick{rashly concluded, that the Devil himself}
\stick{was nothing in the world, but one great}
bouncing \i{Cantharidis.}\tush

I have nothing to say to people who\break
\stick{allow themselves this monstrous liberty}
\stick{in arguing, but what \i{Nazianzen} cried out}
\i{(that is polemically)} to \i{Philagrius}\tsh

\indent\lqq \g{῏Ευγε!}” \i{O rare! ’tis fine reasoning},\break
\stick{\i{Sir}, \i{indeed!}\tskk \lqq \bgroup\porson ὅτι φιλοσοφεῖς ἐν Πάθεσι\egroup”}
\catch{\i{and}}
\stick{\tsk \i{and most nobly do you aim at truth}, \i{when}}
\stick{\i{you philosophize about it in your moods and}}
\i{passions}.

Nor is it to be imagined, for the same\break
\stick{reason, I should stop to inquire, whe-}
\stick{ther love is a disease,\tsh or embroil my-}
\stick{self with \i{Rhasis} and \i{Dioscorides}, whether}
\stick{the seat of it is in the brain or liver;\tsk}
\stick{because this would lead me on, to an}
\stick{examination of the two very opposite}
\stick{manners, in which patients have been}
\stick{treated\tsk the one, of \i{Aætius}, who}
\stick{always begun with a cooling glyster of}
\stick{hempseed and bruised cucumbers;\tsk and}
\stick{followed on with thin potations of water}
\stick{lillies and purslane\tsk to which he added}
\stick{a pinch of snuff, of the herb \i{Hanea};\tsk}
\stick{and where \i{Aætius} durst venture it,\tsk his}
topaz-ring.

\tsk The other, that of \i{Gordonius},\break
\stick{who (in his cap.\@ 15.\@ \i{de Amore)}\sic directs}
\catch{they}
\stick{they should be thrashed, \lqq\i{ad putorem}}
\i{usque},”\tsh till they stink again.

These are disquisitions, which my fa-\break
\stick{ther, who had laid in a great stock of}
\stick{knowledge of this kind, will be very}
\stick{busy with, in the progress of my uncle}
\stick{\toby's affairs: I must anticipate thus}
\stick{much, That from his theories of love,}
\stick{(with which, by the way, he contrived}
\stick{to crucify my uncle \toby's mind, almost}
\stick{as much as his amours themselves)\tsk he}
\stick{took a single step into practice;\tsk and}
\stick{by means of a camphorated cerecloth,}
\stick{which he found means to impose upon}
\stick{the taylor for buckram, whilst he was}
\stick{making my uncle \toby a new pair of}
\stick{breeches, he produced \i{Gordonius}’s effect}
\stick{upon my uncle \toby without the dis-}
grace.

What changes this produced, will be\break
\stick{read in its proper place: all that is need-}
\catch{ful}
\setlength{\baselineskip}{14pt}% 294/23 \simeq 12.7826  294/22 \simeq 13.3636
\stick{ful to be added to the anecdote, is this,}
\stick{\tsk That whatever effect it had upon}
\stick{my uncle \toby,\tsk it had a vile effect}
\stick{upon the house;\tsk and if my uncle}
\stick{\toby had not smoaked it down as he did,}
\stick{it might have had a vile effect upon my}
father too.

\chap{XXXVII}{28pt}{14pt}

\dropcap[ante=\lower5pt\hbox{\tsh\huge’}]{T}{will} come out of itself\break
by and bye.\tsk All I con-\break
\stick{tend for is, that I am not \i{obliged} to set}
\stick{out with a definition of what love is;}
\stick{and so long as I can go on with my story}
\stick{intelligibly, with the help of the word}
\stick{itself, without any other idea to it, than}
\stick{what I have in common with the rest of}
\stick{the world, why should I differ from it a}
\stick{moment before the time?\tsh When I}
\catch{can}
\baselineskip 12.8333pt
\stick{can get on no further,\tsk and find myself}
\stick{entangled on all sides of this mystick la-}
\stick{byrinth,\tsk my Opinion will then come}
in, in course,\tsk and lead me out.

At present, I hope I shall be suffici-\break
\stick{ently understood, in telling the reader,}
my uncle \i{Toby fell in love:}

\tsk Not that the phrase is at all to my\break
\stick{liking: for to say a man is \i{fallen} in love,}
\stick{\tsk or that he is \i{deeply} in love, \tsk or up to}
\stick{the ears in love,\tsk and sometimes even}
\stick{\i{over head and ears in it},\tsk carries an idio-}
\stick{matical kind of implication, that love is}
\stick{a thing \i{below} a man:\tsk this is recurring}
\stick{again to \i{Plato}’s opinion, which, with}
\stick{all his divinityship,\tsk I hold to be dam-}
\stick{nable and heretical:\tsk and so much for}
that.

Let love therefore be what it will,\tsk\break
my uncle \toby fell into it.
\patch{\tsk And}
\baselineskip 13.5pt

\topstrut\tsh And possibly, gentle reader, with\break
\stick{such a temptation\tsk so wouldst thou:}
\stick{For never did thy eyes behold, or thy}
\stick{concupiscence covet any thing in this}
\stick{world, more concupiscible than widow}
\i{Wadman.}

\chap{XXXVIII}{12pt}{8pt}
\baselineskip 12.5pt

\dropcap[findent=.5pt,lraise=.3,loversize=-.17]{T}{o} conceive this right,\tsk call for pen\break
\tight{and ink\tsk here’s paper ready to your}\break
\stick{hand.\tsk Sit down, Sir, paint her to}
\stick{your own mind\tsk as like your mistress}
\stick{as you can\tsk as unlike your wife as}
\stick{your conscience will let you\tsk ’tis all}
\stick{one to me\tsk please but your own fancy}
in it.

\vfill
\eject
\hbox{}
\vfill
\eject
\noindent\topstrut\enspace\tsh Was ever any thing in Nature\break
so sweet!\tsk so exquisite!

\tshh Then, dear Sir, how could my\break
uncle \toby resist it?

Thrice happy book! thou wilt have\break
\stick{one page, at least, within thy covers,}
\stick{which \s{Malice} will not blacken, and}
which \s{Ignorance} cannot misrepresent.

\chap{XXXIX}{14pt}{8pt}

\dropcap[findent=2pt]{A}{s} \susannah was informed by an
express from Mrs.\@ \i{Bridget}, of my\break
\stick{uncle \toby's falling in love with her}
\stick{mistress, fifteen days before it happened,}
\stick{\tsk the contents of which express, \i{Susan}-}
\stick{\i{nah} communicated to my mother the}
\stick{next day,\tsk it has just given me an op-}
\stick{portunity of entering upon my uncle}
\stick{\toby's amours a fortnight before their}
existence.

\patch{I have}
\baselineskip 13pt
\parskip 12pt

I have an article of news to tell you,\break
\stick{Mr.\@ \i{Shandy}, quoth my mother, which}
will surprise you greatly.\tush

Now my father was then holding one\break
\stick{of his second beds of justice, and was}
\stick{musing within himself about the hard-}
\stick{ships of matrimony, as my mother}
broke silence.\tush

“\tshh My brother \toby, quoth she,\break
\stick{\lqq is going to be married to Mrs.\@ \i{Wad-}}
\lqq \i{man}.\,”

\tshh Then he will never, quoth my\break
\stick{father, be able to lie \i{diagonally} in his bed}
again as long as he lives.

It was a consuming vexation to my\break
\stick{father, that my mother never asked the}
\stick{meaning of a thing she did not under-}
stand.
\patch{\tsh That}

\tstrut 24pt\tsh That she is not a woman of sci-\break
\stick{ence, my father would say\tsk is her mis-}
\tight{fortune\tsk but she might ask a question}.\tsk

My mother never did.\tshh In short,\break
\stick{she went out of the world at last without}
\stick{knowing whether it turned \i{round}, or}
\stick{stood \i{still}.\tsh My father had officiously}
\stick{told her above a thousand times which}
way it was,\tsk but she always forgot.

For these reasons a discourse seldom\break
\stick{went on much further betwixt them,}
\stick{than a proposition,\tsk a reply, and a re-}
\stick{joinder; at the end of which, it gene-}
\stick{rally took breath for a few minutes, (as}
\stick{in the affair of the breeches) and then}
went on again.

If he marries, ’twill be the worse for\break
us,\tsk quoth my mother.

\patch{Not}
\parskip 9pt

Not a cherry-stone, said my father,\tsk\break
\stick{he may as well batter away his means}
upon that, as any thing else,

\noindent\enspace\tshh To be sure, said my mother: so\break
\stick{here ended the proposition,\tsk the reply,}
\tsk and the rejoinder, I told you of.

It will be some amusement to him,\break
too,\tsh said my father.

A very great one, answered my mo-\break
ther, if he should have children.\tsh

\tsh Lord have mercy upon me,\tsk\break
\stick{said my father to himself\tsh\;\wastfill}
\stick{\wastfill}
\stick{\wastfill}
\stick{\wastfill}
\hbox to 88pt{\wastfill.}

\patch[20pt]{\lsss{CHAP}.}
\setlength{\baselineskip}{12.7826pt}% 294/23 \simeq 12.7826  294/22 \simeq 13.3636
\setlength{\parskip}{7pt plus 4pt minus 4pt}%
\chap{XL}{24pt}{9pt}

\initial{I}{Am} now beginning to get fairly into\break
my work; and by the help of a\break
\stick{vegitable diet, with a few of the cold}
\stick{seeds, I make no doubt but I shall be}
\stick{able to go on with my uncle \toby's story,}
\stick{and my own, in a tolerable straight line.}

\vskip -3pt

\stick{Now,\hfill}

\vskip 2pt

\centerline{\begin{mplibcode}
    beginfig(0);
path plot[];
plot1 = ((-147, 114) -- (-133.3, 114.5) {dir 70} .. (-125, 120) .. {dir -66} (-116, 115) -- (-94, 115.5)
{dir -70} .. (-86, 111) .. {dir 70} (-79, 115.5) -- (-45, 116)
.. (-60, 111) .. (-80, 105.6) .. (-105, 95.6) -- (-24, 116) -- (10, 116)
{dir 70} .. (17.6, 128) {up} .. (14, 131) .. (10.5, 128) {down} .. {dir -70} (22, 116)
-- (52, 116.7) -- (55, 107) -- (62, 116.6) -- (69, 116.6)
-- (74.5, 127) -- (77.3, 116.3) -- (98.7, 116) {up} .. (100, 121) .. (107, 115.5) .. (110, 130) {up}
.. (113, 133) .. (119, 127) {down} .. (114, 110) {down} .. (119, 108) .. {dir 100} (121, 115.5)
-- (140, 116.6) .. (145, 110) {down} .. {right} (146, 100) -- (151, 100) -- (152.5, 89)
{dir 200} .. {dir -40} (150, 80) .. (155, 79) -- (162, 68)
) scaled 1/2;

plot2 = ((-149.5,59.6) -- (-111,58) {dir -80} .. (-106,52.5) .. {dir 80}
(-100.5,57) -- (-88,57) {dir 110} .. {dir -40} (-83,61.4) {dir 110} .. {dir
-50} (-79,64.3) {dir 60} .. {dir -100} (-74,62.5) {dir 5} .. (-72,62) ..
{dir 225} (-77,53) {dir 5} -- (-68,53.4) {dir -80} .. (-61,48) {right} ..
(-61,41) {left} .. (-61,48) {right} .. {dir 78} (-55,54.5) -- (-45,54.5) --
(-30,54) .. (-7,61) {right} ..  (11,53) -- (32.5,53) {dir 90} .. (10,67)
{dir 184} .. {up} (-8,76) .. (0,90.5) .. (10,92) .. (20,90) ..  (30.5,80)
.. (35,70) {dir -80} .. (35,62) {down} .. (38.5,57) .. {down} (39,53) --
(56,53) {dir 85} .. (60,59) .. {dir -80} (66,53) -- (85,52) {dir 85} ..
(90,59) .. {dir -80} (96,52) -- (113.6,51.7) -- (138.5,78.5) -- (123,52)
-- (147.5,52.5)) scaled 1/2;

plot3 = ((-157, -2) -- (-140, -2) {dir 50} .. {dir -60} (-127, 4)
{dir 100} .. {dir -85} (-118, 4.5) {dir 90} .. {dir -110} (-110, 2.4)
{dir 72} .. {dir -120} (-104, 0) {dir 37} .. {dir -130} (-101, -6)
-- (-66, -6) {up} .. (-70, 5) .. (-84, 15) {up} .. (-76, 21) {right}
.. {down} (-62, 0) .. (-63, -10) .. (-59, -20) .. (-50, -24) .. (-44, -20) .. (-50, -12)
.. {up} (-59, -8) .. {dir -2} (-50, -4) .. (0, -5) .. (80, -6.8)
-- (168, 18) -- (150, 10) -- (93, -7.2) -- (106, -7.8) {dir -80} ..
{up} (113, -8) -- (122.5, -8) {dir 80} .. {down} (130, -8) -- (137, -8)
-- (140.5, -16) -- (145, -8) -- (159, -8)
) scaled 1/2;
plot4 = ((-160, -40) {dir 245} .. (-165, -44) .. (-166, -40) .. (-160, -31) ..
(-153, -40) .. (-158, -52) .. (-150, -50) .. (-145, -49) .. (-149, -60) .. (-140, -60) ..
(-132, -57) ..  (-133, -60) .. (-130, -69) .. {dir 45} (-126, -65) {dir -45} ..
(-120, -68.8) .. {dir 27} (-114, -65) {dir -8} .. {right} (-100, -66.6) -- (-70, -67)
--(-66, -120) -- (-60, -69) -- (-45, -69) {dir -60} .. {dir 70} (-36, -69) -- (-20, -69)
{dir -60} .. {dir 80} (-11, -70) -- (3, -70.7) {dir 70} .. {dir -60} (13, -72) -- (32, -72)
-- (32, -64) -- (80.5, -65) {dir 230} .. (70, -73) .. (70, -90) .. (80, -88)
.. {up} (80, -71) {right} .. {dir -5} (100, -72) {dir 50} ..(125, -61)
.. {dir -30} (148, -72) -- (170, -72)
) scaled 1/2;

    label.rt(TEX("\small\itshape Inv.T.S") scaled .8, (-86.5, -61.2));
    label.rt(TEX("\small\itshape Scul.T.S") scaled .8, (59.6, -62.1));

    draw subpath (26.7, 27.5) of plot1 shifted (1/4, 2) withpen pencircle
    scaled 3/4;
    undraw subpath(38, 39) of plot1
    rotatedabout(point 39 of plot1, -3.6)
    shifted (1,-1);

    pen goose; goose = pencircle xscaled 1 yscaled 1/2 rotated 60;
    draw plot1 withpen goose;
    draw plot2 withpen goose;
    draw plot3 withpen goose;
    draw plot4 withpen goose;
    currentpicture := currentpicture scaled 1.2;
endfig;
\end{mplibcode}}

\patch[6pt]{These}

\topstrut These were the four lines I moved in\break
\stick{through my first, second, third, and}
\stick{fourth volumes\tsh In the fifth volume}
\stick{I have been very good,\tsh the precise}
line I have described in it being this:

\bigskip
\centerline{\begin{mplibcode}
beginfig(1);
pen goose; goose = pencircle xscaled 1 yscaled 1/2 rotated 60;
draw (
    (-1982, -55) -- (-1694, -40) {dir 45} .. (-1500, 72) {right} .. {dir -36} (-1210, -30)
    --(-1149, -30) -- (-1150, -8) {dir 230} .. {dir 50} (-1184, 20)
    --(-1204, 60) -- (-1156, 59) -- (-1162, 102) -- (-1118, 74) -- (-1100, 123) {dir 50}
    .. {dir -70} (-1070, 92) {dir 48} .. (-1030, 104) .. {dir -95} (-1034, 85)
    {dir 42} .. (-995, 96) .. (-1005, 70) {dir 42} .. (-970, 64) .. {down} (-1000, 0)
    .. {dir 120} (-1040, 0) .. (-1065, -16) .. {right} (-1000, -22) -- (-244, 0)
    {down} .. (-204, -60) .. {up} (-170, 0) -- (8, 0) {dir 100} .. (70, 90) .. {dir -100} (110, 0)
    -- (500, 4) {dir -80} .. (550, -50) .. {dir 80} (590, 5)
    -- (695, 6) {dir 100} .. (730, 92) .. {dir -110} (770, 0)
    -- (880, 8) -- (910, -90) -- (980, 10) -- (1170, 15) {dir 120} .. (1350, 610) {right} ..
    (1590, 380) {down} ..
    {dir -130} (1320, 6) -- (1570, 6) -- (1650, 130) -- (1695, 0)
    .. {dir 42} (1822, 50) {dir 90} .. (1790,300) .. (1890, 395) {right}
    .. (1900, 0) {dir 240} .. (1700, -400) {down} .. (1800, -500) ..(1960, -300) .. (1957, -200)
    .. (1900, -144) .. (1800, -270) {down} .. (1850,-320) .. (1890, -280) .. (1860, -250)
) scaled 1/20 withpen goose;

label.urt("\large A", (-80, 8));
label.urt("\large B", (-60, 8));
label.urt("\itshape c", (-14, 4));
label.urt("\itshape c", (-1, 4));
label.urt("\itshape c", (24, 4));
label.urt("\itshape c", (35, 5));
label.urt("\itshape c", (42, 4));
label.urt("\large D", (62, 14));
endfig;
\end{mplibcode}}

\vskip-6pt

\noindent
\stick{By which it appears, that except at the}
\stick{curve, marked A.\@ where I took a trip}
\stick{to \i{Navarre},\tsk and the indented curve B.}
\stick{which is the short airing when I was}
\stick{there with the Lady \i{Baussiere} and her}
\stick{page,\tsk I have not taken the least frisk}
\stick{of a digression, till \i{John de la Casse}’s}
\stick{devils led me the round you see marked}
\stick{D.\tsk for as for \i{c c c c c} they are nothing}
\stick{but parentheses, and the common \i{ins}}
\stick{and \i{outs} incident to the lives of the great-}
\stick{est ministers of state; and when com-}
\catch{pared}
\setlength{\baselineskip}{14pt}% 308/22 = 14
\setlength{\parskip}{10pt plus 4pt minus 4pt}%
\stick{pared with what men have done,\tsk or}
\stick{with my own transgressions at the letters}
A B D\tsk they vanish into nothing.

In this last volume I have done better\break
\stick{still\tsk for from the end of \i{Le Fever}’s}
\stick{episode, to the beginning of my uncle}
\stick{\toby's campaigns,\tsk I have scarce step-}
ped a yard out of my way.

If I mend at this rate, it is not im-\break
\stick{possible\tshh by the good leave of his}
\stick{grace of \i{Benevento}’s devils\tsk but I}
\stick{may arrive hereafter at the excellency of}
\stick{going on even thus;\hfill}
\stickrule\break
\stick{which is a line drawn as straight as I}
\stick{could draw it, by a writing-master’s}
\stick{ruler, (borrowed for that purpose) turn-}
\stick{ing neither to the right hand or to the}
left.
\patch{This}
\setlength{\baselineskip}{12.3pt}% 308/22 = 14

This \i{right line},\tsk the path-way for\break
Christians to walk in! say divines\tsh

\tsh The emblem of moral rectitude!\break
says \i{Cicero}\tsh

\tsh \i{The best line!} say cabbage-plan-\break
\stick{ters\tsk is the shortest line, says \i{Archi}-}
\stick{\i{medes}, which can be drawn from one}
given point to another.\tsh

I wish your ladyships would lay\break
\stick{this matter to heart in your next birth-}
day suits!

\tsh What a journey!

Pray can you tell me,\tsk that is, with-\break
\stick{out anger, before I write my chapter}
\stick{upon straight lines\tsk by what mis-}
\stick{take\tsk who told them so\tsk or how it}
\stick{has come to pass, that your men of wit}
\stck{and genius have all along confounded this}
\stick{line, with the line of \s{gravitation}.\quad}

\smallskip

\centerline{\s{End} of the \s{Sixth Volume}.}
\vfill
\end{document}
