\documentclass[twoside]{article}
\usepackage[paperheight=395pt, paperwidth=240pt, width=166pt, height=308pt, top=42pt, headsep=2pt]{geometry}
\usepackage[noinfo, frame, a6, center, lualatex]{crop}
\usepackage{graphicx}
\usepackage{shandean}
%
%
\title{The Life and Opinions of Tristram Shandy, Gentleman}
\author{Laurence Sterne}
\date{1765}
\begin{document}
\def\vol{VIII}
\pagestyle{empty}
\null
\vfill
\eject
\null
\vfill
\eject
% full title
\noindent
\centerline{\fontsize{10}{14}\selectfont\lsv{80}{THE}}
\centerline{\tstrut 26pt\fontsize{18}{28}\selectfont\lsv{140}{LIFE}}
\centerline{\tstrut 23pt\fontsize{10}{14}\selectfont\lsv{80}{AND}}
\centerline{\tstrut 26pt\fontsize{18}{28}\selectfont\lsv{55}{OPINIONS}}
\centerline{\tstrut 23pt\fontsize{10}{14}\selectfont\lsv{80}{OF}}
\centerline{\tstrut 26pt\fontsize{14}{20}\selectfont\lsv{20}{TRISTRAM}\enspace\lsv{20}{SHANDY},}
\centerline{\tstrut 12pt\fontsize{12}{14}\selectfont\lsv{16}{\textsc{Gentleman}}.}

\vskip 4pt

\bgroup\footnotesize
\noindent
\stick{\enspace \i{Non enim excursus hic ejus, sed opus ipsum est.}\hfill}
\stick{\hfill \s{Plin.} Lib.\@ quintus Epistola sexta.}
\par\egroup

\vskip 13pt

\centerline{\lsss{VOL}.\quad\vol.}

\vskip 13pt

\noindent
\centerline{\lsv{100}{LONDON}:}
\centerline{\tstrut 14pt Printed for T.\@ \s{Becket} and P.\@ A.\@  \s{Dehont},\sic}
\centerline{in the Strand.\quad M DCC LXV.}

\vfill
\eject
\null
\vfill
\eject
\pagestyle{fancy}
\thispagestyle{empty}
\setcounter{page}{1}
\baselineskip 13pt

\noindent
\centerline{\lsv{60}{THE}}
\centerline{\tstrut 30pt\large \lsss{LIFE}\enspace and\enspace \lsss{OPINIONS}}
\centerline{\tstrut 26pt\lsv{70}{OF}}
\centerline{\tstrut 28pt\large \lss{TRISTRAM SHANDY}, Gent.}

\vskip 20pt
\hrule

\chap{I}{18pt}{0pt}

\dropcap[ante=\lower5pt\hbox{\tsh}]{B}{ut} softly\tsk for in these\break
sportive plains, and under\break
\stick{this genial sun, where at this instant all}
\stick{flesh is running out piping, fiddling, and}
\stick{dancing to the vintage, and every step}
\stick{that’s taken, the judgment is surprised by}
\stick{the imagination, I defy, notwithstanding}
\catch{all}
\stick{all that has been said upon \i{straight lines} \fnast}
\stick{in sundry pages of my book\tsk I defy the}
\stick{best cabbage planter that ever existed,}
\stick{whether he plants backwards or for-}
\stick{wards, it makes little difference in the ac-}
\stck{count (except that he will have more to an-}
\stk{swer for in the one case than in the other)\tsk}
\stick{I defy him to go on coolly, critically, and}
\stick{canonically, planting his cabbages one by}
\stick{one, in straight lines, and stoical distan-}
\stick{ces, especially if slits in petticoats are un-}
\stick{sew’d up\tsk without ever and anon strad-}
\stick{dling out, or sidling into some bastardly}
\stick{digression\tsh In \i{Freeze-land}, \i{Fog-land}, }
\stick{and some other lands I wot of\tsk it may}
be done\tsh

But in this clear climate of fantasy\break 
\stick{and perspiration, where every idea,}
\stick{\hfill sensible}
\stick{\tstrut 14pt\footnotesize \hfill \fnast\  Vid.\@ Vol.\@ VI.\, p.\@ 152.\hfill}
\vfill
\eject
\noindent
\stick{sensible and insensible, gets vent\tskk in} 
\stick{this land, my dear Eugenius\tskk in this}
\stick{fertile land of chivalry and romance,} 
\stck{where I now sit, unskrewing my ink-horn} 
\stick{to write my uncle Toby’s amours, and} 
\stick{with all the meanders of \s{Julia}’s track} 
\stick{in quest of her \s{Diego}, in full view of} 
\stick{my study window\tsk if thou comest not}
and takest me by the hand\tsh

What a work it is likely to turn\break
out!

Let us begin it.

\patch[70pt]{\lsss{CHAP}.}
\parskip 10pt

\chap{II}{28pt}{14pt}

\dropcap{I}{t} is with \s{love} as with \s{cuck}-\break
\s{oldom}\tsh

\tshh But now I am talking of begin-\break
\stick{ning a book, and have long had a thing}
\stick{upon my mind to be imparted to the}
\stick{reader, which if not imparted now, can}
\stick{never be imparted to him as long as I}
\stick{live (whereas the \s{comparison} may be}
\stick{imparted to him any hour in the day)\tsh}
\stick{I’ll just mention it, and begin in good}
earnest.

The thing is this.

That of all the several ways of begin-\break
\stick{ning a book which are now in practice}
\catch{throughout} 
\baselineskip 13.5pt
\parskip14pt
\stick{throughout the known world, I am con-}
\stick{fident my own way of doing it is the}
\stick{best\tsk I’m sure it is the most religious} 
\stick{\tsk for I begin with writing the first}
\stick{sentence\tsk and trusting to Almighty} 
God for the second.

’Twould cure an author for ever of\break
\stick{the fuss and folly of opening his street-}
\stick{door, and calling in his neighbours and} 
\stick{friends, and kinsfolk, with the devil}
\stick{and all his imps, with their hammers and} 
\stick{engines, \etc only to observe how one} 
\stick{sentence of mine follows another, and}
how the plan follows the whole.

I wish you saw me half starting out of\break
\stick{my chair, with what confidence, as I}
\stick{grasp the elbow of it, I look up\tshh}
\catch{catching}
\parskip 7pt
\stick{catching the idea, even sometimes before} 
it half way reaches me\tsh

I believe in my conscience I intercept\break
\stick{many a thought which heaven intended} 
for another man.

Pope and his Portrait \fnast\  are fools to\break
\stick{me\tshh no martyr is ever so full of faith}
\stick{or fire\tshh I wish I could say of good}
\stick{works too\tshh but I have no\hfill}
\stick{\kern 42pt Zeal\ \ or\ \ Anger\tshh or\hfill}
\stick{\kern 40pt Anger or Zeal\tshh\hfill}
\stick{And till gods and men agree together to}
\stick{call it by the same name\tshh the errant-}
\stick{est \s{Tartuffe}, in science\tskk in politics}
\stick{\tskk or in religion, shall never kindle a}
\stick{spark within me, or have a worse} 
\stick{word, or a more unkind greeting,}
\stick{\hfill than}
\stick{\tstrut 14pt \footnotesize \hfill \fnast\enspace Vid.\@ Pope’s Portrait.\hfill}
\vfill
\eject
\noindent
than what he will read in the next\break
chapter.

\chap{III}{21pt}{10pt}

\noindent
\stick{\indent\tsk Bon jour!\tsk good morrow?}
\stick{\tsk so you have got your cloak on}
\stick{betimes!\tsk but ’tis a cold morning,}
\stick{and you judge the matter rightly\tsk ’tis}
\stick{better to be well mounted, than go o’\thinspace foot}
\stick{\tsk and obstructions in the glands are}
\stick{dangerous\tsh And how goes it with thy }
\stick{concubine\tsk thy wife\tsk and thy little ones}
\stick{o’both sides? and when did you hear}
\stick{from the old gentleman and lady\tsk your}
\stick{sister, aunt, uncle and cousins\tsk I~hope}
\stick{they have got better of their colds,}
\stick{coughs, claps, tooth-aches, fevers, stran-}
\stck{guries, sciaticas, swellings, and sore-eyes.}
\catch{\tsh What}
\parskip 15pt
\stick{\tshh What a devil of an apothecary!} 
\stick{to take so much blood\tsk give such a vile}
\stick{purge\tsk puke\tsk poultice\tsk plaister\tsk night-}
\stick{draught\tsk glister\tsk blister?\tsh And why}
\stick{so many grains of calomel? santa Ma-}
\stick{ria!\@ and such a dose of opium!\@ peri-}
\stick{clitating, pardi!\@ the whole family of}
\stick{ye, from head to tail\tshh By my great}
\stick{aunt Dinah’s old black velvet mask!}
\stick{I think there was no occasion for it.}

Now this being a little bald about the\break
\stick{chin, by frequently putting off and on,}
\stick{\i{before} she was got with child by the}
\stick{coachman\tsk not one of our family would} 
\stick{wear it after. To cover the \s{mask} afresh,} 
\stick{was more than the mask was worth\tsh} 
\stick{and to wear a mask which was bald,}
\catch{or}
\stick{or which could be half seen through,}
\stick{was as bad as having no mask at all\tsh}

This is the reason, may it please your\break
\stick{reverences, that in all our numerous fa-}
\stick{mily, for these four generations, we} 
\stick{count no more than one archbishop, a}
\stick{Welch judge, some three or four alder-}
men, and a single mountebank\tush

In the sixteenth century, we boast of\break
no less than a dozen alchymists.
\patch[78pt]{\lsss{CHAP}.}
%---------
\parskip 12.5pt
\chap{IV}{28pt}{7pt}

\dropcap{\raise10pt\hbox{\large \lqq}I}{t} is with Love as with Cuckol-\break 
dom”\tsh the suffering party is at\break 
\stick{least the \i{third}, but generally the last in}
\stck{the house who knows any thing about the}
\stick{matter: this comes, as all the world} 
\stick{knows, from having half a dozen words}
\stick{for one thing; and so long, as what in}
\stick{this vessel of the human frame, is \i{Love}\tsk} 
\stick{may be \i{Hatred}, in that\tsk \i{Sentiment}}
\stick{half a yard higher\tsh and \i{Nonsense}\tsh}
\stick{\tsh\!\tsk no, Madam,\tsk not there\tsh I}
\stick{mean at the part I am now pointing to}
\stick{with my forefinger\tsh how can we help}
ourselves?

Of all mortal, and immortal men too,\break
\stick{if you please, who ever soliloquized upon}
\catch{this} 
\stick{this mystic subject, my uncle Toby was}
\stick{the worst fitted, to have push’d his re-}
\stick{searches, thro’ such a contention of feel-}
\stick{ings; and he had infallibly let them all} 
\stick{run on, as we do worse matters, so\sic see}
\stick{what they would turn out\tsh had not} 
\stick{Bridget’s pre-notification of them to}
\stick{Susannah, and Susannah’s repeated}  
\stick{manifesto’s thereupon to all the world,} 
\stick{made it necessary for my uncle Toby to} 
look into the affair.
\patch[78pt]{\lsss{CHAP}.}

\chap{V}{24pt}{0pt}

\dropcap{W}{hy} weavers, gardeners, and gla-\break
diators\tsk or a man with a pined\break
\stick{leg (proceeding from some ailment in}
\stick{the \i{foot})\tsk should ever have had some}
\stck{tender nymph breaking her heart in secret} 
\stck{for them, are points well and duely settled}
\stck{and accounted for, by ancient and modern}
physiologists.

A water-drinker, provided he is a pro-\break
\stick{fess’d one, and does it without fraud or} 
\stick{covin, is precisely in the same predica-}
\stick{ment: not that, at first sight, there is} 
\stick{any consequence, or shew of logic in it,}
\stick{\lqq That a rill of cold water dribbling}
\stick{\lqq through my inward parts, should light}
\lqq up a torch in my Jenny’s\tsk\sic

\patch{\tsk The}

\tsk\!\tsk The proposition does not strike\break
\stick{one; on the contrary, it seems to run op-}
\stick{posite to the natural workings of causes}
and effects\tsh

But it shews the weakness and imbeci-\break
lity of human reason.

\tsk \lqq And in perfect good health\break
\lqq with it?\,”

\tsk The most perfect\tsk Madam, that\break
friendship herself could wish me\tsh

\tsk \lqq And drink nothing!\tsk nothing\break
\lqq but water?\,”

\tsk Impetuous fluid! the moment thou\break 
\stick{presses against the flood-gates of the} 
brain\tsh see how they give way!\tsh
\patch{In}
\parskip8pt

\tighter{In swims \s{Curiosity}, beckoning to her}\break
\stick{damsels to follow\tsk they dive into the} 
centre of the current\tsh

\s{Fancy} sits musing upon the bank,\break
\stick{and with her eyes following the stream,} 
\stck{turns straws and bulrushes into masts and}
\stick{bowsprits\tsk And \s{Desire}, with vest}
\stick{held up to the knee in one hand, snatches} 
\stick{at them, as they swim by her, with the} 
other\tsh

O ye water-drinkers! is it then by this\break
\stick{delusive fountain, that ye have so often}
\stick{governed and turn’d this world about} 
\stick{like a mill-wheel\tsk grinding the faces of}
\stick{the impotent\tsk be-powdering their ribs\tsk} 
\stick{be-peppering their noses, and changing} 
\stick{sometimes even the very frame and face} 
of nature\tsh
\patch{\tsk If}
\parskip15pt

\tsk If I was you, quoth Yorick, I would\break
\stick{drink more water, Eugenius.\tsk And, if I} 
\stick{was you, Yorick, replied Eugenius, so}
would I.

Which shews they had both read\break
Longinus\tsh

For my own part, I am resolved never\break
to read any book but my own, as long\break
as I live.
\patch[78pt]{\lsss{CHAP}.}
\parskip10pt

\chap{VI}{26pt}{12pt}

\initial[nindent=0pt]{I}{Wish} my uncle Toby had been a\break
water-drinker; for then the thing had\break
\stick{been accounted for, That the first mo-}
\stick{ment Widow Wadman saw him, she felt} 
\stick{something stirring within her in his fa-}
vour\tsk Something!\tsk something.

\noindent
\stck{\hfill\tsk Something perhaps more than friend-}
\stick{ship\tsk less than love\tsk something\tsk no}
\stick{matter what\tsk no matter where\tsk I would}
\stick{not give a single hair off my mule’s tail,}
\stick{and be obliged to pluck it off myself}
\stick{(indeed the villain has not many to spare,}
\stick{and is not a little vicious into the bar-}
\stick{gain) to be let by your worships into the} 
secret\tsh

\patch[8pt]{But}

But the truth is, my uncle Toby was\break
\stick{not a water-drinker; he drank it neither}
\stick{pure nor mix’d, or any how, or any} 
\stck{where, except fortuitously upon some ad-}
\stick{vanced posts, where better liquor was not} 
\stick{to be had\tsh or during the time he was} 
\stick{under cure; when the surgeon telling} 
\stick{him it would extend the fibres, and}
\stick{bring them sooner into contact\tsh my}
uncle Toby drank it for quietness sake.

Now as all the world knows, that no\break 
\stick{effect in nature can be produced without} 
\stick{a cause and as it is as well known, that}
\stick{my uncle Toby, was neither a weaver\tsk}
\stick{a gardener, or a gladiator\tsk unless as a}
\stick{captain, you will needs have him one\tsk}
\stick{but then he was only a captain of foot\tsk}
\stick{and besides the whole is an equivocation}
\stick{\tsh There is nothing left for us to sup-}
\catch{pose,} 
\stick{pose, but that my uncle Toby’s leg\tsh}
\stick{but that will avail us little in the present}
\stick{hypothesis, unless it had proceeded from}
\stick{some ailment \i{in the foot}\tsk whereas his leg}
\stck{was not emaciated from any disorder in his} 
\stick{foot\tsk for my uncle Toby’s leg was not}
\stick{emaciated at all. It was a little stiff and} 
\stick{awkward, from a total disuse of it, for} 
\stick{the three years he lay confined at my}
\stick{father’s house in town; but it was plump}
\stick{and muscular, and in all other respects as}
good and promising a leg as the other.

I declare, I do not recollect any one\break
\stick{opinion or passage of my life, where my} 
\stck{understanding was more at a loss to make}
\stick{ends meet, and torture the chapter I had}
\stick{been writing, to the service of the chapter}
\stick{following it, than in the present case:}
\stick{one would think I took a pleasure in run-\sic}
\catch{ing} 
\stick{ing into difficulties of this kind, merely}
\stick{to make fresh experiments of getting out}
\stick{of ’em\tsh Inconsiderate soul that thou}
\stick{art! What! are not the unavoidable}
\stick{distresses with which, as an author and a}
\stick{man, thou art hemm’d in on every side}
\stick{of thee\tsk\!\tsk are they, Tristram, not suffi-}
\stick{cient, but thou must entangle thyself}
still more?

Is it not enough that thou art in debt,\break
\stick{and that thou hast ten cart-loads of thy} 
\stick{fifth and sixth volumes still\tsk still unsold,}
\stick{and art almost at thy wit’s ends, how to} 
get them off thy hands.

To this hour art thou not tormented\break
\stick{with the vile asthma thou gattest in} 
\stick{skating against the wind in Flanders?}
\stick{and is it but two months ago, that in a}
\catch{fit} 
\parskip 8pt
\stick{fit of laughter, on seeing a cardinal make}
\stick{water like a quirister (with both hands)} 
\stick{thou brakest a vessel in thy lungs, where-}
\stick{by, in two hours, thou lost as many}
\stick{quarts of blood; and hadst thou lost as}
\stick{much more, did not the faculty tell thee}
\stick{\tsh it would have amounted to a}
gallon?\tsh

\chap{VII}{2pt}{0pt}

\tshh But for heaven’s sake, let us not\break
\stick{talk of quarts or gallons\tsk let us take} 
\stick{the story straight before us; it is so nice}
\stick{and intricate a one, it will scarce bear}
\stick{the transposition of a single tittle; and}
\stick{some how or other, you have got me}
thrust almost into the middle of it\tsk

\tsk I beg we may take more care.

\patch[16pt]{\lsss{CHAP}.}

\chap{VIII}{20pt}{12pt}

\dropcap{M}{y} uncle Toby and the corporal\break
had posted down with so much\break 
\stick{heat and precipitation, to take possession} 
\stick{of the spot of ground we have so often} 
\stick{spoke of, in order to open their cam-}
\stck{paign as early as the rest of the allies; that}
\stick{they had forgot one of the most neces-}
\stick{sary articles of the whole affair; it was}
\stick{neither a pioneer’s spade, a pick-ax, or a}
shovel\tsk

\medskip

\tskk It was a bed to lie on: so that as\break 
\stick{Shandy Hall was at that time unfurnish-}
\stick{ed; and the little inn where poor Le}
\stick{Fever died, not yet built; my uncle} 
\stick{Toby was constrained to accept of a bed}
\stick{at Mrs.\@ Wadman’s, for a night or two,}
\catch{till}
\stick{till corporal Trim (who to the cha-}
\stick{racter of an excellent valet, groom, cook,}
\stick{sempster, surgeon and engineer, super-}
\stick{added that of an excellent upholsterer}
\stick{too) with the help of a carpenter and a}
\stick{couple of taylors, constructed one in my}
uncle Toby’s house.

A daughter of Eve, for such was wi-\break
\stick{dow Wadman, and ’tis all the character}
I intend to give of her\tsk 

\tsk\lqq\i{That she was a perfect woman};”

\noindent
\stick{had better be fifty leagues off\tsk or in her}
\stick{warm bed\tsk or playing with a case-knife}
\stick{\tsk or any thing you please\tsk than make a} 
\stick{man the object of her attention, when}
\stick{the house and all the furniture is her} 
own.
\patch{There}
\parskip 8pt

There is nothing in it out of doors\break
\stick{and in broad day-light, where a woman} 
\stick{has a power, physically speaking, of}
\stick{viewing a man in more lights than one\tsk}
\stick{but here, for her soul, she can see him in}
\stick{no light without mixing something of} 
\stick{her own goods and chattels along with} 
\stick{him\tshh till by reiterated acts of such}
\stick{combinations, he gets foisted into her in-}
ventory\tsh

\tsk And then good night.

But this is not matter of \s{System}; for\break
\stick{I have delivered that above\tshh nor is it} 
\stick{matter of \s{Breviary}\tshh for I make no}
\stick{man’s creed but my own\tshh nor matter} 
\stick{of \s{Fact}\tshh at least that I know of;}
\stick{but ’tis matter copulative and introduc-}
tory to what follows.
\patch[8pt]{\lsss{CHAP}.}

\chap{IX}{24pt}{8pt}

\initial{I}{Do} not speak it with regard to the\break
\tight{coarseness or cleanness of them\tsh or}\break
\stick{the strength of their gussets\tsh but pray}
\stick{do not night-shifts differ from day-shifts}
\stick{as much in this particular, as in any}
\stick{thing else in the world; That they so far}
\stick{exceed the others in length, that when}
\stick{you are laid down in them, they fall al-}
\stick{most as much below the feet, as the day-}
shifts fall short of them?

Widow Wadman’s night-shifts (as\break
\stick{was the mode I suppose in King Wil-}
\stick{liam’s and Queen Anne’s reigns) were}
\stick{cut however after this fashion; and if the}
\stick{fashion is changed, (for in Italy they are}
\stick{come to nothing)\tsh so much the worse}
\catch{for}
\stick{for the public; they were two Flemish} 
\stick{ells and a half in length; so that allow-}
\stick{ing a moderate woman two ells, she had} 
\stick{half an ell to spare, to do what she would} 
with.

Now from one little indulgence gain’d\break
\stick{after another, in the many bleak and de-}
\stick{cemberly nights of a seven years widow-}
\stick{hood, things had insensibly come to this} 
\stick{pass, and for the two last years had got} 
\stick{establish’d into one of the ordinances of}
\stick{the bed-chamber\tsk That as soon as Mrs.}
\stick{Wadman was put to bed, and had got her}
\stick{legs stretched down to the bottom of it, of}
\stick{which she always gave Bridget notice\tsk} 
\stick{Bridget with all suitable decorum, having}
\stick{first open’d the bed-cloaths at the feet,} 
\stick{took hold of the half ell of cloath we}
\stick{are speaking of, and having gently, and}
\catch{with}
\parskip 12pt
\stick{with both her hands, drawn it down-}
\stick{wards to its furthest extension, and then}
\stick{contracted it again side long by four or} 
\stick{five even plaits, she took a large corking} 
\stick{pin out of her sleeve, and with the}
\stick{point directed towards her, pin’d the} 
\stick{plaits all fast together a little above the} 
\stick{hem; which done, she tuck’d all in} 
\stick{tight at the feet, and wish’d her mistress}
a good night.

This was constant, and without any\break
\stick{other variation than this; that on shiv-}
\stck{ering and tempestuous nights, when Brid-}
\stick{get untuck’d the feet of the bed, \etc}
\stick{to do this\tsk she consulted no thermo-}
\stick{meter but that of her own passions;}
\stick{and so performed it standing\tsk kneeling}
\stick{\tsk or squatting, according to the different}
\stick{degrees of faith, hope, and charity, she}
\catch{was}
\stick{was in, and bore towards her mistress} 
\stick{that night. In every other respect, the} 
\stick{\i{etiquette} was sacred, and might have}
\stick{vied with the most mechanical one of} 
\stick{the most inflexible bed-chamber in}
Christendom.

The first night, as soon as the cor-\break
\stick{poral had conducted my uncle Toby up}
\stick{stairs, which was about ten\tshh Mrs.}
\stick{Wadman threw herself into her arm} 
\stick{chair, and crossing her left knee with} 
\stick{her right, which formed a resting-place}
\stick{for her elbow, she reclin’d her cheek} 
\stick{upon the palm of her hand, and leaning} 
\stick{forwards, ruminated till midnight upon}
both sides of the question.

The second night she went to her bu-\break
\stick{reau, and having ordered Bridget to}
\catch{bring}
\parskip 10pt
\stick{bring her up a couple of fresh candles}
\stick{and leave them upon the table, she took}
\stick{out her marriage-settlement, and read it} 
\stick{over with great devotion: and the third} 
\stick{night (which was the last of my uncle} 
\stick{Toby’s stay) when Bridget had pull’d} 
\stick{down the night-shift, and was assaying} 
to stick in the corking pin\tsh

\tsk With a kick of both heels at\break
\stick{once, but at the same time the most}
\stick{natural kick that could be kick’d in her}
\stick{situation\tsk for supposing\enspace\astWW6}
\stick{\astWW3\enspace to be the sun in its meridian, it}
\stick{was a north-east kick\tsh she kick’d the}
\stick{pin out of her fingers\tsh\!\tsk the \i{etiquette}}
\stick{which hung upon it, down\tsh down it} 
\stick{fell to the ground, and was shivered into}
a thousand atoms.
\patch[12pt]{From}

From all which it was plain that\break
\stick{widow Wadman was in love with my}
uncle Toby.

\chap{X}{24pt}{10pt}

\dropcap{M}{y} uncle Toby’s head at that\break
time was full of other matters,\break
\stick{so that it was not till the demolition of}
\stick{Dunkirk, when all the other civilities of}
\stick{Europe were settled, that he found lei-}
sure to return this.

This made an armistice (that is speak-\break
\stick{ing with regard to my uncle Toby\tskk} 
\stick{but with respect to Mrs.\@ Wadman, a}
\stick{vacancy)\tskk of almost eleven years. But} 
\stick{in all cases of this nature, as it is the}
\stick{second blow, happen at what distance of}
\catch{time} 
\stick{time it will, which makes the fray\tsh} 
\stick{I chuse for that reason to call these} 
\stick{the amours of my uncle Toby with}
\stick{Mrs.\@ Wadman, rather than the amours}
of Mrs.\@ Wadman with my uncle Toby.

This is not a distinction without a\break 
difference.

It is not like the affair of \i{an old hat}\break
\stick{\i{cock}’\i{d}\tshh and \i{a cock}’\i{d old hat}, about}
\stick{which your reverences have so often}
\stick{been at odds with one another\tshh but}
\stick{there is a difference here in the nature of}
things\tsh

And let me tell you, gentry, a wide\break 
one too.
\patch[26pt]{\lsss{CHAP}.}

\chap{XI}{28pt}{12pt}

\dropcap{N}{ow} as widow Wadman did love\break
my uncle Toby\tsh and my uncle\break
\stick{Toby did not love widow Wadman,}
\stick{there was nothing for widow Wadman}
\stick{to do, but to go on and love my uncle}
Toby\tsh or let it alone.

Widow Wadman would do neither\break 
the one or the other\tsh

\tsk Gracious heaven!\tsk but I for-\break
\stick{get I am a little of her temper myself;}
\stick{for whenever it so falls out, which it}
\stck{sometimes does about the equinoxes, that}
\stick{an earthly goddess is so much this, and}
\stick{that, and t’other, that I cannot eat my}
\catch{breakfast}
\stick{breakfast for her\tsh and that she careth}
\stick{not three halfpence whether I eat my}
breakfast or no\tsh

\tshh Curse on her! and so I send her\break
\stick{to Tartary, and from Tartary to \i{Terra}}
\stick{\i{del Fuogo}, and so on to the devil: in}
\stick{short there is not an infernal nitch where}
\stick{I do not take her divinityship and stick} 
it.

But as the heart is tender, and the\break
\stick{passions in these tides ebb and flow ten}
\stick{times in a minute, I instantly bring her}
\stick{back again; and as I do all things in} 
\stick{extremes, I place her in the very centre}
of the milky-way\tsh

Brightest of stars! thou wilt shed thy\break
influence upon some one\tush
\patch[8pt]{\tsh The}

\tshh The duce take her and her influ-\break
\stick{ence too\tshh for at that word I lose all}
\stick{patience\tshh much good may it do him!} 
\stick{\tshh By all that is hirsute and gashly!}
\stick{I cry, taking off my furr’d cap, and}
\stick{twisting it round my finger\tshh I would}
not give sixpence for a dozen such!

\tshh But ’tis an excellent cap too (put-\break
\stick{ting it upon my head, and pressing it} 
\stick{close to my ears)\tsk and warm\tsk and soft;}
\stick{especially if you stroke it the right way\tsk}
\stick{but alas! that will never be my luck\tshh}
\stick{(so here my philosophy is shipwreck’d} 
again)

\tshh No; I shall never have a finger\break
\stick{in the pye (so here I break my meta-}
phor)\tsh
\patch[12pt]{Crust}

Crust and Crumb\\
\null\indent Inside and out\\
\null\indent Top and bottom\tsh I detest it, I hate\break
\stick{it, I repudiate it\tsh I’m sick at the sight}
of it\tsh 

\noindent\hbox to 42pt{\hss ’Tis all }pepper,\\
\null\hbox to 42pt{\hss}garlick,\\
\null\hbox to 42pt{\hss}staragen,\\
\null\hbox to 42pt{\hss}salt, and\\
\null\hbox to 42pt{\hss}devil’s dung\tsk by the great\break
\stick{arch cook of cooks, who does nothing,}
\stick{I think, from morning to night, but sit} 
\stick{down by the fire-side and invent inflam-}
\stick{gatory\sic dishes for us, I would not touch}
it for the world\tsh

\tsk O Tristram! Tristram! cried\break
Jenny.
\patch[14pt]{O Jenny!}

O Jenny! Jenny! replied I, and so\break
went on with the twelfth chapter.

\chap{XII}{30pt}{0pt}

\tshh \lqq Not touch it for the world”\break
\indent did I say\tsh\sic

Lord, how I have heated my imagi-\break
nation with this metaphor!


\chap{XIII}{30pt}{0pt}

\dropcap[nindent=1pt]{W}{hich} shews, let your rever-\break
ences and worships say what you\break
\stick{will of it (for as for \i{thinking}\tsk all}
\stick{who \i{do} think\tsk think pretty much alike}
\stick{both upon it and other matters)\tshh}
\catch{\s{Love}} 
\stick{\s{Love} is certainly, at least alphabetically} 
speaking, one of the most

\noindent
\stick{\indent A\tspace gitating\hfill}
\stick{\indent B\tspace ewitching\hfill}
\stick{\indent C\tspace onfounded\hfill}
\stick{\indent D\tspace evilish affairs of life\tsh the most}
\stick{\indent E\tspace xtravagant\hfill}
\stick{\indent F\tspace utilitous\hfill}
\stick{\indent G\tspace alligaskinish\hfill}
\stick{\indent H\tspace andy-dandyish\hfill}
\stick{\indent I\tspace racundulous (there is no K to it) and}
\stick{\indent L\tspace yrical of all human passions: at}
\stick{\kern 4em the same time, the most\hfill}
\stick{\indent M\tspace isgiving\hfill}
\stick{\indent N\tspace innyhammering\hfill}
\stick{\indent O\tspace bstipating\hfill}
\stick{\indent P\tspace ragmatical\hfill}
\stick{\indent S\tspace tridulous\hfill}
\stick{\indent R\tspace idiculous\tsk though by the bye the}
\catch{R}
\stick{R should have gone first\tsk But in short}
\stick{’tis of such a nature, as my father once}
\stick{told my uncle Toby upon the close of a}
\stick{long dissertation upon the subject\tshh}
\stick{\lqq You can scarce,” said he, \lqq combine}
\stick{\lqq two ideas together upon it, brother}
\stick{\lqq Toby, without an hypallage”\tshh}
What’s that? cried my uncle Toby.

The cart before the horse, replied my\break
father\tsh

\tshh And what has he to do there?\break
cried my uncle Toby\tsh

Nothing, quoth my father, but to get\break
in\tsh or let it alone.

Now widow Wadman, as I told you\break
\stick{before, would do neither the one or the} 
other.
\patch{She}

\tight{She stood however ready harnessed and}\break
\stick{caparisoned at all points, to watch ac-}
cidents.

\chap{XIV}{26pt}{0pt}

\dropcap{T}{he} Fates, who certainly all fore-\break
knew of these amours of widow\break
\stick{Wadman and my uncle Toby, had,} 
\stick{from the first creation of matter and}
\stick{motion (and with more courtesy than}
\stick{they usually do things of this kind),}
\stick{established such a chain of causes and}
\stick{effects hanging so fast to one another,} 
\stick{that it was scarce possible for my uncle}
\stick{Toby to have dwelt in any other house}
\stick{in the world, or to have occupied any} 
\stick{other garden in Christendom, but the}
\stick{very house and garden which join’d and}
\catch{laid}
\stick{laid parallel to Mrs.\@ Wadman’s; this,}
\stick{with the advantage of a thickset arbour}
\stick{in Mrs.\@ Wadman’s garden, but planted}
\stick{in the hedge-row of my uncle Toby’s,}
\stick{put all the occasions into her hands}
\stick{which Love-militancy wanted; she could}
\stick{observe my uncle Toby’s motions, and}
\stick{was mistress likewise of his councils of}
\stick{war; and as his unsuspecting heart had}
\stick{given leave to the corporal, through the}
\stick{mediation of Bridget, to make her a}
\stick{wicker gate of communication to enlarge}
\stick{her walks, it enabled her to carry on}
\stick{her approaches to the very door of the}
\stick{sentry-box; and sometimes out of grati-}
\stick{tude, to make the attack, and endeavour}
\stick{to blow my uncle Toby up in the very}
sentry-box itself.

\patch[12pt]{\lsss{CHAP}.}

\chap{XV}{32pt}{0pt}

\dropcap{I}{t} is a great pity\tsh but ’tis certain\break
from every day’s observation of man,\break
\stick{that he may be set on fire like a candle,}
\stick{at either end\tsk provided there is a suffi-}
\stick{cient wick standing out; if there is not}
\stick{\tsk there’s an end of the affair; and if}
\stick{there is\tsk by lighting it at the bottom,}
\stick{as the flame in that case has the misfor-}
\stick{tune generally to put out itself\tsk there’s}
an end of the affair again.

For my part, could I always have the\break
\stick{ordering of it which way I would be} 
\stick{burnt myself\tskk for I cannot bear the}
\stick{thoughts of being burnt like a beast\tsk}
\stick{I would oblige a housewife constantly to}
\catch{light}
\stick{light me at the top; for then I should}
\stick{burn down decently to the socket; that} 
\stick{is, from my head to my heart, from my} 
\stick{heart to my liver, from my liver to my}
\stick{bowels, and so on by the meseraick} 
\stick{veins and arteries, through all the turns}
\stck{and lateral insertions of the intestines and}
their tunicles, to the blind gut\tsh

\vskip 4pt

\tsh I beseech you, doctor Slop, quoth\break
\stick{my uncle Toby, interrupting him as he}
\stick{mentioned the \i{blind gut}, in a discourse} 
\stick{with my father the night my mother} 
\stick{was brought to bed of me\tsk I be-}
\stick{seech you, quoth my uncle Toby, to} 
\stick{tell me which is the blind gut; for, old}
\stick{as I am, I vow I do not know to this}
day where it lies.
\patch[12pt]{The}

The \i{blind gut}, answered doctor Slop, lies betwixt the
\i{Illion} and \i{Colon}\tsh

\vskip 4pt

\tsh In a man? said my father.

\vskip 4pt

\tsk ’Tis precisely the same, cried\break
doctor Slop, in a woman\tsh

\vskip 4pt

That’s more than I know; quoth my father.
\patch[108pt]{\lsss{CHAP}.}
\parskip 12pt

\chap{XVI}{32pt}{0pt}

\tsh And so to make sure of both sys-\break
\stick{tems, Mrs. Wadman predetermined to} 
\stick{light my uncle Toby neither at this end or}
\stick{that; but, like a prodigal’s candle, to} 
\stick{light him, if possible, at both ends at}
once.

Now, through all the lumber rooms\break
\stick{of military furniture, including both of}
\stick{horse and foot, from the great arsenal} 
\stick{of Venice to the Tower of London}
\stick{(exclusive) if Mrs.\@ Wadman had been} 
\stick{rummaging for seven years together,}
\stick{and with Bridget to help her, she could}
\stick{not have found any one \i{blind} or \i{man}-}
\stick{\i{telet} so fit for her purpose, as that which}
\catch{the}
\stick{the expediency of my uncle Toby’s af-}
fairs had fix’d up ready to her hands.

I believe I have not told you\tsh but\break
\stick{I don’t know\tsh possibly I have\tsh be} 
\stick{it as it will, ’tis one of the number of} 
\stick{those many things, which a man had} 
\stick{better do over again, than dispute about}
\stick{it\tsk That whatever town or fortress the}
\stick{corporal was at work upon, during the}
\stick{course of their campaign, my uncle Toby}
\stick{always took care on the inside of his} 
\stick{sentry-box, which was towards his left}
\stick{hand, to have a plan of the place, fasten’d} 
\stick{up with two or three pins at the top, but} 
\stick{loose at the bottom, for the conveniency}
\stick{of holding it up to the eye, \etc .\@ .\@ .\@ as}
\stick{occasions required; so that when an} 
\stick{attack was resolved upon, Mrs.\@ Wadman}
\stick{had nothing more to do, when she had}
\catch{got} 
\parskip 10pt
\stick{got advanced to the door of the sentry-}
\stick{box, but to extend her right hand; and} 
\stick{edging in her left foot at the same move-}
\stick{ment, to take hold of the map or plan,}
\stick{or upright, or whatever it was, and with}
\stick{outstretched neck meeting it half way,} 
\stick{\tsk  to advance it towards her; on which}
\stick{my uncle Toby’s passions were sure to} 
\stick{catch fire\tsh for he would instantly take}
\stick{hold of the other corner of the map in} 
\stick{his left hand, and with the end of his}
\stick{pipe, in the other, begin an expla-}
nation.

When the attack was advanced to this\break
\stick{point;\tsh the world will naturally enter} 
\stick{into the reasons of Mrs.\@ Wadman’s next} 
\stick{stroke of generalship\tsh which was, to} 
\stick{take my uncle Toby’s tobacco-pipe out} 
\stick{of his hand as soon as she possibly could;}
\catch{which,}
\parskip 9pt
\stick{which, under one pretence or other, but}
\stick{generally that of pointing more distinctly}
\stick{at some redoubt or breast-work in the}
\stick{map, she would effect before my uncle}
\stick{Toby (poor soul!) had well march’d a-}
bove half a dozen toises with it.

\tsk  It obliged my uncle Toby to make\break
use of his forefinger.

The difference it made in the attack\break
\stick{was this; That in going upon it, as in}
\stick{the first case, with the end of her fore-}
\stick{finger against the end of my uncle Toby’s}
\stick{tobacco-pipe, she might have travelled}
\stick{with it, along the lines, from Dan to Beer-}
\stick{sheba, had my uncle Toby’s lines reach’d}
\stick{so far, without any effect: For as there}
\stick{was no arterial or vital heat in the end of}
\stick{the tobacco-pipe, it could excite no sen-}
\catch{timent} 
\stick{timent\tsh it could neither give fire by}
\stick{pulsation\tsh or receive it by sympathy}
\tsh ’twas nothing but smoak.

Whereas, in following my uncle\break
\stick{Toby’s forefinger with hers, close thro’}
\stick{all the little turns and indentings of his}
\stick{works\tsk pressing sometimes against} 
\stick{the side of it\tsk then treading upon it's\sic}
\stick{nail\tsk then tripping it up\tsk then} 
\stick{touching it here\tsk then there, and so}
\stick{on\tsk it set something at least in mo-}
tion.

This, tho’ slight skirmishing, and at\break
\stick{a distance from the main body, yet drew} 
\stick{on the rest; for here, the map usually} 
\stick{falling with the back of it, close to the} 
\stick{side of the sentry-box, my uncle Toby,} 
\stick{in the simplicity of his soul, would lay his}
\catch{hand}
\stick{hand flat upon it, in order to go on with}
\stick{his explanation; and Mrs.\@ Wadman, by a} 
\stick{manœuvre as quick as thought, would} 
\stck{as certainly place her’s close besides it; this}
\stick{at once opened a communication, large} 
\stick{enough for any sentiment to pass or re-}
\stick{pass, which a person skill’d in the ele-}
\stick{mentary and practical part of love-}
making, has occasion for\tsh

By bringing up her forefinger parallel\break 
\stick{(as before) to my uncle Toby’s\tsk it}
\stick{unavoidably brought the thumb into}
\stick{action\tsk and the forefinger and thumb}
\stick{being once engaged, as naturally brought}
\stick{in the whole hand. Thine, dear uncle}
\stick{Toby!\@ was never now in it’s\sic\ right place}
\stick{\tsh Mrs.\@ Wadman had it ever to take} 
\stick{up, or, with the gentlest pushings, pro-}
\stick{trusions, and equivocal compressions,}
\catch{that}
\stick{that a hand to be removed is capable of}
\stick{receiving\tsk to get it press’d a hair}
breadth of one side out of her way.

Whilst this was doing, how could she\break
\stck{forget to make him sensible, that it was her}
\stck{leg (and no one’s else) at the bottom of the} 
\stck{sentry-box, which slightly press’d against}
\stick{the calf of his\tsk So that my uncle} 
\stick{Toby being thus attacked and sore}
\stick{push’d on both his wings\tsh was it a}
\stick{wonder, if now and then, it put his}
centre into disorder?\tsh

\tsh The duce take it! said my uncle\break
Toby.
\patch[48pt]{\lsss{CHAP}.}

\chap{XVII}{28pt}{14pt}

\dropcap{T}{hese} attacks of Mrs.\@ Wadman,\break
you will readily conceive to be of\break
\stick{different kinds; varying from each other,} 
\stick{like the attacks which history is full of,}
\stick{and from the same reasons. A general}
\stick{looker on, would scarce allow them to be}
\stick{attacks at all\tshh or if he did, would}
\stick{confound them all together\tshh but I} 
\stick{write not to them: it will be time enough} 
\stick{to be a little more exact in my descrip-}
\stick{tions of them, as I come up to them,} 
\stick{which will not be for some chapters;}
\stick{having nothing more to add in this, but}
\stick{that in a bundle of original papers and} 
\stick{drawings which my father took care to} 
\stick{roll up by themselves, there is a plan of}
\catch{Bouchain}
\stick{Bouchain in perfect preservation (and}
\stick{shall be kept so, whilst I have power to}
\stck{preserve any thing) upon the lower corner}
\stick{of which, on the right hand side, there}
\stick{is still remaining the marks of a snuffy}
\stick{finger and thumb, which there is all the} 
\stick{reason in the world to imagine, were}
\stick{Mrs.\@ Wadman’s; for the opposite side}
\stick{of the margin, which I suppose to have}
\stck{been my uncle Toby’s, is absolutely clean:}
\stick{This seems an authenticated record of} 
\stck{one of these attacks; for there are vestigia}
\stick{of the two punctures partly grown up,}
\stick{but still visible on the opposite corner of} 
\stick{the map, which are unquestionably the}
\stick{very holes, through which it has been} 
pricked up in the sentry-box\tsh

By all that is priestly! I value this\break
\stick{precious relick, with it’s\sic \i{stigmata} and}
\catch{\i{pricks},}
\stick{\i{pricks}, more than all the relicks of the}
\stick{Romish church\tsh always excepting,}
\stick{when I am writing upon these matters,} 
\stick{the pricks which enter’d the flesh of St.}
\stick{\i{Radagunda} in the desert, which in your}
\stick{road from \s{Fesse} to \s{Cluny}, the nuns}
of that name will shew you for love.

\chap{XVIII}{26pt}{12pt}

\initial{I}{Think}, an’ please your honour, quoth\break
Trim, the fortifications are quite de-\break
\stck{stroyed\tshh and the bason is upon a level} 
\stick{with the mole\tshh I think so too; replied}
\stick{my uncle Toby with a sigh half sup-}
\stick{press’d\tshh but step into the parlour,} 
\stick{Trim, for the stipulation\tshh it lies upon}
the table.
\patch[12pt]{It}

\tight{It has lain there these six weeks, replied}\break
\stick{the corporal, till this very morning that}
\stick{the old woman kindled the fire with it\tsk}

\tsk Then, said my uncle Toby, there\break
\stick{is no further occasion for our services.} 
\stick{The more, an’ please your honour, the} 
\stick{pity, said the corporal; in uttering which} 
\stick{he cast his spade into the wheel-barrow,}
\stick{which was beside him, with an air the}
\stck{most expressive of disconsolation that can}
\stick{be imagined, and was heavily turning} 
\stick{about to look for his pick-ax, his pio-}
\stick{neer’s shovel, his picquets and other little} 
\stick{military stores, in order to carry them off}
\stick{the field\tsh when a heigh ho!\@ from the}
\stick{sentry-box, which, being made of thin}
\stick{slit deal, reverberated the sound more}
sorrowfully to his ear, forbad him.
\patch[12pt]{\tsk No;}

\tshh No; said the corporal to himself,\break 
\stick{I’ll do it before his honour rises to-mor-}
\stick{row morning; so taking his spade out of} 
\stick{the wheel-barrow again, with a little} 
\stick{earth in it, as if to level something at the}
\stick{foot of the glacis\tsh but with a real in-}
\stick{tent to approach nearer to his master, in}
\stick{order to divert him\tsh he loosen’d a sod} 
\stick{or two\tsh pared their edges with his}
\stick{spade, and having given them a gentle} 
\stick{blow or two with the back of it, he sat} 
\stick{himself down close by my uncle Toby’s} 
feet, and began as follows.

\patch[65pt]{\lsss{CHAP}.}
\parskip 14pt

\chap{XIX}{24pt}{8pt}

\dropcap{I}{t} was a thousand pities\tsh though I\break
believe, an’ please your honour, I am\break
\stick{going to say but a foolish kind of a thing} 
for a soldier\tsh

A soldier, cried my uncle Toby, inter-\break
\stick{rupting the corporal, is no more exempt} 
\stick{from saying a foolish thing, Trim, than}
\stick{a man of letters\tshh But not so often;} 
\stick{and please your honour, replied the cor-}
poral\tsh My uncle Toby gave a nod.

\vskip 4pt

It was a thousand pities then, said the\break
\stick{corporal, casting his eye upon Dunkirk,} 
\stick{and the mole, as \i{Servius Sulpicius}, in re-}
\stick{turning out of Asia (when he sailed from}
\catch{Ægina}
\parskip 10pt
\stick{Ægina towards Megara) did upon Co-}
rinth and Pyreus\tsh

\tsk \lqq It was a thousand pities, an’ please\break
\stick{your honour, to destroy these works\tsh}
\stick{and a thousand pities to have let them} 
stood.”\tsh

\tshh Thou art right, Trim, in both\break
\stick{cases: said my uncle Toby\tshh This,} 
\stck{continued the corporal, is the reason, that} 
\stick{from the beginning of their demolition} 
\stick{to the end\tshh I have never once whist-}
\stick{led, or sung, or laugh’d, or cry’d, or}
\stick{talk’d of pass'd done deeds, or told your} 
honour one story good or bad\tsh

\tshh Thou hast many excellencies,\break
\stick{Trim, said my uncle Toby, and I hold} 
\stick{it not the least of them, as thou happenest}
\catch{to}
\stick{to be a story-teller, that of the number}
\stick{thou hast told me, either to amuse me in} 
\stick{my painful hours, or divert me in my} 
\stick{grave ones\tsk thou hast seldom told me a} 
bad one\tsh

\tsh Because, an’ please your honour,\break
\stick{except one of a \i{King of Bohemia and his}}
\stick{\i{seven castles},\tsk they are all true; for they}
are about myself\tsh

I do not like the subject the worse,\break
\stick{Trim, said my uncle Toby, on that}
\stick{score: But prithee what is this story?}
thou hast excited my curiosity.

I’ll tell it your honour, quoth the cor-\break
\stick{poral directly\tsh Provided, said my uncle} 
\stick{Toby, looking earnestly towards Dun-}
\stick{kirk and the mole again\tsh provided it}
\catch{is} 
\stick{is not a merry one; to such, Trim, a man} 
\stick{should ever bring one half of the enter-}
\stick{tainment along with him; and the dis-}
\stick{position I am in at present would wrong}
\stick{both thee, Trim, and thy story\tsk It is}
\stick{not a merry one by any means, replied}
\stick{the corporal\tsk Nor would I have it alto-}
\stick{gether a grave one, added my uncle} 
\stick{Toby\tsh It is neither the one nor the}
\stick{other, replied the corporal, but will suit}
\stick{your honour exactly\tsh Then I’ll thank} 
\stick{thee for it with all my heart, cried my} 
uncle Toby, so prithee begin it, Trim.

The corporal made his reverence;\break
\stick{and though it is not so easy a matter as} 
\stick{the world imagines, to pull off a lank}
\stick{montero cap with grace\tsh or a whit less}
\stick{difficult, in my conceptions, when a man}
\stick{is sitting squat upon the ground, to make}
\catch{a}
\stick{a bow so teeming with respect as the} 
\stick{corporal was wont; yet by suffering the}
\stick{palm of his right hand, which was to-}
\stick{wards his master, to slip backwards upon}
\stick{\tight{the grass, a little beyond his body, in order}}
\stick{to allow it the greater sweep\tsh and by}
\stick{\tight{an unforced compression, at the same time,}}
\stick{of his cap with the thumb and the two}
\stick{forefingers of his left, by which the dia-}
\stick{meter of the cap became reduced, so that}
\stick{it might be said, rather to be insensibly}
\stick{squeez’d\tsk than pull’d off with a flatus}
\stick{\tsh the corporal acquitted himself of}
\stick{both, in a better manner than the posture} 
\stick{of his affairs promised; and having} 
\stick{hemmed twice, to find in what key his}
\stick{story would best go, and best suit his} 
\stick{master’s humour\tsk he exchanged a single}
\stick{look of kindness with him, and set off} 
thus.
\patch{The}
\parskip 8pt
\bgroup
\fontsize{11}{16}\selectfont
\stick{\tstrut 36pt The Story of the king of Bohemia}
\stick{\qquad\quad and his seven castles.\quad\qquad}
\par
\egroup
%---------
\vskip 12pt

\dropcap{T}{here} was a certain king of\break
Bo - - he\tsh

\tstrut16pt As the corporal was entering the con-\break
\stick{fines of Bohemia, my uncle Toby obliged} 
\stick{him to halt for a single moment; he} 
\stick{had set out bare-headed, having since he}
\stick{pull’d off his Montero-cap in the latter}
\stick{end of the last chapter, left it lying} 
beside him on the ground.

\tsh The eye of Goodness espieth all\break
\stick{things\tsk so that before the corporal}
\stick{had well got through the first five words}
\catch{of} 
\stick{of his story, had my uncle Toby twice}
\stick{touch’d his Montero-cap with the end of} 
\stick{his cane, interrogatively\tsh as much as} 
\stick{to say, Why don’t you put it on, Trim?}
\stick{Trim took it up with the most respect-}
\stick{ful slowness, and casting a glance of hu-}
\stick{miliation as he did it, upon the em-}
\stick{broidery of the fore-part, which being}
\stick{dismally tarnish’d and fray’d moreover}
\stick{in some of the principal leaves and bold-}
\stick{est parts of the pattern, he lay’d it} 
\stick{down again betwixt his two feet, in or-}
der to moralize upon the subject.

\vskip 8pt

\tsh ’Tis every word of it but too true,\break
\stick{cried my uncle Toby, that thou art about}
to observe\tsh


\patch[18pt]{\lqq\i{Nothing}}
\parskip 14pt

\lqq\i{Nothing in this world, Trim, is}\break
\lqq\i{made to last for ever.}”

\tshh But when tokens, dear Tom, of\break
\stick{thy love and remembrance wear out,}
said Trim, what shall we say?

There is no occasion, Trim, quoth my\break
\stick{uncle Toby, to say any thing else; and}
\stick{was a man to puzzle his brains till}
\stick{Doom’s day, I believe, Trim, it would be}
impossible.

The corporal perceiving my uncle\break
\stick{Toby was in the right, and that it would} 
\stick{be in vain for the wit of man to think} 
\stick{of extracting a purer moral from his cap,} 
\stick{without further attempting it, he put it} 
\stick{on; and passing his hand across his fore-}
\catch{head} 
\stick{head to rub out a pensive wrinkle,} 
\stick{which the text and the doctrine between} 
\stick{them had engender’d, he return’d, with} 
\stick{the same look and tone of voice, to his}
\stick{story of the king of Bohemia and his}
seven castles.
\patch[158pt]{The}

\null
\vskip 12pt

\centerline{The story of the king of Bohemia and}
\centerline{his seven castles, continued.}

\dropcap{T}{here} was a certain king of\break
Bohemia, but in whose reign, ex-\break
\stick{cept his own, I am not able to inform} 
your honour\tsh

\medskip

I do not desire it of thee, Trim, by\break
any means, cried my uncle Toby.

\medskip

\tsh It was a little before the time,\break
\stick{an’ please your honour, when giants} 
\stick{were beginning to leave off breeding;\tsk}
\stick{but in what year of our Lord that}
was\tsh

\patch[8pt]{\tsh I would}
\parskip 13pt

\tshh I would not give a half-penny\break
to know, said my uncle Toby.

\tshh Only, an’ please your honour, it\break
\stick{makes a story look the better in the}
face\tsh

\tshh ’Tis thy own, Trim, so orna-\break
\stick{ment it after thy own fashion; and take}
\stick{any date, continued my uncle Toby,}
\stick{looking pleasantly upon him\tsk take any}
\stick{date in the whole world thou choosest,} 
\stick{and put it to\tsk thou art heartily wel-}
come\tsh

The corporal bowed; for of every\break
\stick{century, and of every year of that cen-}
\stick{tury, from the first creation of the world}
\stick{down to Noah’s flood; and from Noah’s}
\catch{flood} 
\stick{flood to the birth of Abraham; through}
\stick{all the pilgrimages of the patriarchs, to}
\stick{the departure of the Israelites out of}
\stick{Egypt\tsk and throughout all the Dy-}
\stick{nasties, Olympiads, Urbecondita’s, and}
\stick{other memorable epochas of the differ-}
\stick{ent nations of the world, down to the}
\stick{coming of Christ, and from thence to} 
\stick{the very moment in which the cor-}
\stick{poral was telling his story\tsh had my}
\stick{uncle Toby subjected this vast empire of}
\stick{time and all its abysses at his feet; but}
\stick{as \s{modesty} scarce touches with a finger} 
\stick{what \s{liberality} offers her with both} 
\stck{hands open\tsk the corporal contented him-}
\stick{self with the very \i{worst year} of the} 
\stick{whole bunch; which, to prevent your} 
\stick{honours of the Majority and Minority} 
\stick{from tearing the very flesh off your bones} 
\catch{in} 
\stick{in contestation, ‘\kern 1pt Whether that year is not}
\stick{always the last cast-year of the last cast-}
\stick{almanack’\tsk I tell you plainly it was;} 
\stick{but from a different reason than you}
wot of\tsh

\tsk It was the year next him\tsk\break
\stick{which being the year of our Lord}
\stick{seventeen hundred and twelve, when}
\stick{the duke of Ormond was playing} 
\stick{the devil in Flanders\tsk the corpo-}
\stick{ral took it, and set out with it afresh}
on his expedition to Bohemia.
\patch[68pt]{The}
\parskip 10pt

\noindent
\centerline{\tstrut 24pt The story of the king of Bohemia and}
\centerline{his seven castles, continued.}

\smallskip

\dropcap{I}{n} the year of our Lord one thou-\break
\tight{sand seven hundred and twelve, there}\break
was, an’ please your honour\tsh

\vskip 2pt

\tshh To tell thee truly, Trim, quoth\break
\stick{my uncle Toby, any other date would}
\stick{have pleased me much better, not only}
\stick{on account of the sad stain upon our}
\stick{history that year, in marching off our}
\stick{troops, and refusing to cover the siege of}
\stick{Quesnoi, though Fagel was carrying on}
\stick{the works with such incredible vigour\tsk}
\stick{but likewise on the score, Trim, of thy}
\stick{own story; because if there are\tsk and}
\stick{which, from what thou hast dropt, I}
\catch{partly} 
\stick{partly suspect to be the fact\tsk if there}
are giants in it\tsh

There is but one, an’ please your ho-\break
nour\tsh

\tsh ’Tis as bad as twenty, replied my\break
\stick{uncle Toby\tsk thou should’st have} 
\stick{carried him back some seven or eight}
\stick{hundred years out of harm’s way, both} 
\stick{of criticks and other people; and there-}
\stick{fore I would advise thee, if ever thou}
tellest it again\tsh

\tsh If I live, an’ please your honour,\break
\stick{but once to get through it, I will ne-}
\stick{ver tell it again, quoth Trim, either to}
\stick{man, woman, or child\tsh Poo\tsk poo!} 
\stick{said my uncle Toby\tsk but with accents}
\stick{of such sweet encouragement did he}
\catch{utter}  
\stick{utter it, that the corporal went on with}
his story with more alacrity than ever.

\centerline{\tstrut 32pt The story of the king of Bohemia and}
\centerline{his seven castles, continued.}

\dropcap{T}{here} was, an’ please your\break
honour, said the corporal, raising\break
\stick{his voice and rubbing the palms of his}
\stick{two hands cheerily together as he begun,}
a certain king of Bohemia\tsh

\tshh Leave out the date entirely,\break
\stick{Trim, quoth my uncle Toby, leaning} 
\stick{forwards, and laying his hand gently}
\stick{upon the corporal’s shoulder to temper}
\stick{the interruption\tsk leave it out entirely,} 
\stick{Trim; a story passes very well without} 
\stick{these niceties, unless one is pretty sure}
\catch{of} 
\stick{of ’em\tshh Sure of ’em! said the cor-}
poral, shaking his head\tsh

Right; answered my uncle Toby, it\break
\stick{is not easy, Trim, for one, bred up as} 
\stick{thou and I have been to arms, who sel-}
\stck{dom looks further forward than to the end} 
\stick{of his musket, or backwards beyond his} 
\stick{knapsack, to know much about this}
\stick{matter\tsh God bless your honour! said} 
\stick{the corporal, won by the \i{manner} of my}
\stick{uncle Toby’s reasoning, as much as by} 
\stick{the reasoning itself, he has something}
\stick{else to do; if not on action, or a} 
\stick{march, or upon duty in his garrison\tsk he} 
\stick{has his firelock, an’ please your honour,}
\stck{to furbish\tsk his accoutrements to take care}
\stick{of\tsk his regimentals to mend\tsk himself to}
\stick{shave and keep clean, so as to appear al-}
\catch{ways} 
\stick{ways like what he is upon the parade;} 
\stick{what business, added the corporal tri-}
\stick{umphantly, has a soldier, an’ please your}
\stick{honour, to know any thing at all of \i{geo}-}
\i{graphy ?}

\tsh Thou would’st have said \i{chro}-\break
\stick{\i{nology}, Trim, said my uncle Toby;}
\stick{for as for geography, ’tis of absolute} 
\stick{use to him; he must be acquainted in-}
\stck{timately with every country and its boun-}
\stick{daries where his profession carries him;}
\stick{he should know every town and city,} 
\stick{and village and hamlet, with the canals,}
\stck{the roads, and hollow ways which lead up} 
\stick{to them; there is not a river or a rivu-}
\stick{let he passes, Trim, but he should be}
\stick{able at first sight to tell thee what is its}
\stick{name\tsk in what mountains it takes its}
\catch{rise} 
\stick{rise\tsk what is its course\tsk how far it is na-}
\stick{vigable\tsk where fordable\tsk where not; he}
\stick{should know the fertility of every val-}
\stick{ley, as well as the hind who ploughs it;}
\stick{and be able to describe, or, if it is re-}
\stick{quired, to give thee an exact map of all}
\stick{the plains and defiles, the forts, the ac-}
\stick{clivities, the woods and morasses, thro’}
\stick{and by which his army is to march;}
\stck{he should know their produce, their plants,}
\stick{their minerals, their waters, their ani-}
\stick{mals, their seasons, their climates, their}
\stick{heats and cold, their inhabitants, their}
\stick{customs, their language, their policy,} 
and even their religion.

\smallskip

Is it else to be conceived, corporal,\break
\stick{continued my uncle Toby, rising up in his}
\stick{sentry-box, as he began to warm in this}
\catch{part} 
\stick{part of his discourse\tsk how Marlborough} 
\stick{could have marched his army from the}
\stick{banks of the Maes to Belburg; from Bel-}
\stick{burg to Kerpenord\tsk (here the corporal}
\stick{could sit no longer) from Kerpenord,}
\stick{Trim, to Kalsaken; from Kalsaken to}
\stick{Newdorf; from Newdorf to Landen-}
\stick{bourg; from Landenbourg to Milden-}
\stick{heim; from Mildenheim to Elchingen;} 
\stick{from Elchingen to Gingen; from Gingen}
\stck{to Balmerchoffen; from Balmerchoffen to}
\stick{Skellenburg, where he broke in upon the}
\stick{enemy’s works; forced his passage over}
\stick{the \i{Danube}; cross’d the \i{Lech}\tsk pushed on}
\stick{his troops into the heart of the empire,} 
\stick{marching at the head of them through} 
\stick{Fribourg, Hokenwert, and Schonevelt,} 
\stick{to the plains of Blenheim and Hochstet?}
\stick{\tsh Great as he was, corporal, he could}
\catch{not}
\stick{not have advanced a step, or made one} 
\stick{single day’s march without the aids of}
\stick{\i{Geography}\tsh As for \i{Chronology}, I own,} 
\stick{Trim, continued my uncle Toby, sitting} 
\stick{down again coolly in his sentry-box,} 
\stck{that of all others, it seems a science which} 
\stick{the soldier might best spare, was it not} 
\stick{for the lights which that science must one} 
\stick{day give him, in determining the inven-}
\stick{tion of powder; the furious execution of}
\stick{which, renversing every thing like thun-}
\stick{der before it, has become a new æra to} 
\stck{us of military improvements, changing so}
\stick{totally the nature of attacks and defences}
\stick{both by sea and land, and awakening so} 
\stick{much art and skill in doing it, that the} 
\stick{world cannot be too exact in ascertaining}
\stick{the precise time of its discovery, or too} 
\stick{inquisitive in knowing what great man}
\catch{was}
\stick{was the discoverer, and what occasions} 
gave birth to it.

I am far from controverting, conti-\break
\stick{nued my uncle Toby, what historians} 
\stick{agree in, that in the year of our Lord} 
\stick{1380, under the reign of Wencelaus,} 
\stick{son of Charles the fourth\tsh a certain}
\stck{priest, whose name was Schwartz, shew’d}
\stick{the use of powder to the Venetians, in}
\stick{their wars against the Genoese; but ’tis}
\stick{certain he was not the first; because if}
\stick{we are to believe Don Pedro the bishop}
\stick{of Leon\tsk How came priests and bishops,}
\stick{an’ please your honour, to trouble their}
\stick{heads so much about gun-powder? God}
\stick{knows, said my uncle Toby\tsh his pro-}
\stick{vidence brings good out of every thing}
\stick{\tsk and he avers, in his chronicle of King}
\stick{Alphonsus, who reduced Toledo, That}
\catch{in} 
\stick{in the year 1343, which was full thirty} 
\stick{seven years before that time, the secret}
\stick{of powder was well known, and employ-}
\stick{ed with success, both by Moors and} 
\stick{Christians, not only in their sea-combats,} 
\stick{at that period, but in many of their most} 
\stick{memorable sieges in Spain and Barbary\tsk}
\stick{And all the world knows, that Friar} 
\stick{Bacon had wrote expressly about it, and}
\stick{had generously given the world a receipt}
\stick{to make it by, above a hundred and fifty}
\stick{years before even Schwartz was born\tsk} 
\stick{And that the Chinese, added my uncle} 
\stick{Toby, embarrass us, and all accounts of}
\stick{it still more, by boasting of the inven-}
\stick{tion some hundreds of years even before}
\stick{him\tsh\hfill}\\[4pt]
\stick{\indent\tsk They are a pack of liars, I believe,}
\stick{cried Trim\tsh\hfill}
\catch{\tsh They}
\stick{\indent\tshh They are some how or other de-}
\stick{ceived, said my uncle Toby, in this mat-}
\stick{ter, as is plain to me from the present}
\stick{miserable state of military architecture}
\stck{amongst them; which consists of nothing}
\stick{more than a fossè with a brick wall with-}
\stick{out flanks\tsk and for what they give us as}
\stick{a bastion at each angle of it, ’tis so bar-}
\stick{barously constructed, that it looks for all}
\stick{the world\tsk\!\tsh Like one of my} 
\stick{seven castles, an’ please your honour,}
quoth Trim.

\medskip

My uncle Toby, tho’ in the utmost\break
\stick{distress for a comparison, most courte-}
\stick{ously refused Trim’s offer\tsk till Trim tell-}
\stick{ing him, he had half a dozen more in} 
\stick{Bohemia, which he knew not how to get} 
\stick{off his hands\tsh my uncle Toby was so}
\catch{touch’d} 
\stick{touch’d with the pleasantry of heart}
\stick{of the corporal\tsh that he discontinued}
\stick{his dissertation upon gunpowder\tsh and} 
\stick{begged the corporal forthwith to go on}
\stick{with his story of the King of Bohemia}
and his seven castles.

\noindent
\stick{\tstrut21pt The story of the King of Bohemia and}
\centerline{his seven castles, continued.}

\dropcap{T}{his} \i{unfortunate} King of Bohemia,\break
said Trim\tshh Was he unfortu-\break
\stick{nate then? cried my uncle Toby, for he}
\stick{had been so wrapt up in his dissertation}
\stick{upon gun-powder and other military}
\stick{affairs, that tho’ he had desired the cor-}
\stick{poral to go on, yet the many interrup-}
\stick{tions he had given, dwelt not so strong}
\stick{upon his fancy, as to account for the}
\catch{epithet}
\stick{epithet\tsk Was he \i{unfortunate}, then,}
\stick{Trim? said my uncle Toby, pathetically}
\stick{\tsk The corporal, wishing first the \i{word}}
\stick{and all its synonimas at the devil, forth-}
\stick{with began to run back in his mind, the} 
\stick{principal events in the King of Bohemia’s}
\stick{story; from every one of which, it ap-}
\stick{pearing that he was the most fortunate}
\stick{man that ever existed in the world\tsk}
\stick{it put the corporal to a stand: for not} 
\stick{caring to retract his epithet\tsk and less,} 
\stick{to explain it\tsk and least of all, to twist}
\stick{his tale (like men of lore) to serve a sys-}
\stick{tem\tsk he looked up in my uncle} 
\stick{Toby’s face for assistance\tsk but seeing}
\stick{it was the very thing, my uncle Toby}
\stick{sat in expectation of himself\tsk after a} 
hum and a haw, he went on\tsh
\patch[12pt]{The}
\parskip 14pt

The King of Bohemia, an’ please your\break
\stick{honour, replied the corporal, was \i{unfor}-}
\stick{\i{tunate}, as thus\tsk That taking great} 
\stick{pleasure and delight in navigation and all} 
\stick{sort of sea-affairs\tsk and there \i{happen}-}
\stick{\i{ing} throughout the whole kingdom of} 
\stick{Bohemia, to be no sea-port town what-}
ever\tsh

How the duce should there\tsk Trim?\break
\stick{cried my uncle Toby; for Bohemia be-}
\stick{ing totally inland, it could have happen’d}
\stick{no otherwise\tsk It might, said Trim,}
if it had pleased God\tsh

My uncle Toby never spoke of the\break
\stick{being and natural attributes of God, but} 
with diffidence and hesitation\tsh

\patch[14pt]{\tsh I}


\tsk I believe not, replied my uncle\break
\stick{Toby, after some pause\tsk  for being}
\stick{inland, as I said, and having Silesia and}
\stick{Moravia to the east; Lusatia and Upper}
\stick{Saxony to the north; Franconia to the} 
\stick{west; and Bavaria to the south: Bohemia}
\stick{could not have been propell’d to the sea,} 
\stick{without ceasing to be Bohemia\tsh nor}
\stick{could the sea, on the other hand, have}
\stick{come up to Bohemia, without overflow-}
\stick{ing a great part of Germany, and de-}
\stick{stroying millions of unfortunate inhabi-}
\stick{tants who could make no defence against}
\stick{it\tsh Scandalous! cried Trim\tsk Which}
\stick{would bespeak, added my uncle Toby,}
\stick{mildly, such a want of compassion in}
\stick{him who is the father of it\tsk that, I}
\stick{think, Trim\tsk the thing could have}
happen’d no way.
\patch[12pt]{The}
\parskip 12pt

The corporal made the bow of un-\break
feigned conviction; and went on.

Now the King of Bohemia with his\break
\stick{queen and courtiers \i{happening} one fine}
\stick{summer’s evening to walk out\tsh Aye!} 
\stick{there the word \i{happening} is right, Trim,}
\stick{cried my uncle Toby; for the King of}
\stck{Bohemia and his queen might have walk’d}
\stick{out, or let it alone;\tsh ’twas a matter of}
\stick{contingency, which might happen, or} 
not, just as chance ordered it.

King William was of an opinion, an’\break
\stick{please your honour, quoth Trim, that} 
\stick{every thing was predestined for us in this} 
\stick{world; insomuch, that he would often} 
\stick{say to his soldiers, that \lqq every ball had}
\stick{it’s\sic billet.\,” He was a great man, said} 
\stick{my uncle Toby\tsh And I believe, con-}
\catch{tinued} 
\stick{tinued Trim, to this day, that the shot}
\stick{which disabled me at the battle of Lan-}
\stick{den, was pointed at my knee for no}
\stick{other purpose, but to take me out of}
\stick{his service, and place me in your ho-}
\stick{nour’s, where I should be taken so much}
\stick{better care of in my old age\tsh It shall}
\stick{never, Trim, be construed otherwise,} 
said my uncle Toby.

The heart, both of the master and the\break
\stick{man, were alike subject to sudden over-}
flowings;\tsh a short silence ensued.

\tight{Besides, said the corporal, resuming the}\break
\stick{discourse\tsk but in a gayer accent\tsh if it}
\stick{had not been for that single shot, I had}
\stick{never, ’an please your honour, been in}
love\tush
\patch[12pt]{So,}
\parskip 10pt

So, thou wast once in love, Trim! said\break
my uncle Toby, smiling\tsh

\tight{Souse!\@ replied the corporal\tsk over head}\break
\stick{and ears! an’ please your honour. Pri-}
\stick{thee when?\@ where?\tsk and how came it to}
\stick{pass?\tshh I never heard one word of it}
\stick{before; quoth my uncle Toby:\tsk I} 
\stick{dare say, answered Trim, that every}
\stick{drummer and serjeant’s son in the regi-}
\stick{ment knew of it\tsk Its\sic high time I} 
should\tsh\!\!\tsk said my uncle Toby.

\tight{Your honour remembers with concern,}\break
\stick{said the corporal, the total rout and} 
\stick{confusion of our camp and army at the}
\stick{affair of Landen; every one was left to} 
\stick{shift for himself; and if it had not been}
\stick{for the regiments of Wyndham, Lum-}
\stick{ley, and Galway, which covered the re-}
\catch{treat} 
\stick{treat over the bridge of Neerspeeken,}
\stick{the king himself could scarce have gain’d}
\stick{it\tshh he was press’d hard, as your ho-}
nour knows, on every side of him\tsh

\vskip 12pt

Gallant mortal! cried my uncle Toby, 
\stck{caught up with enthusiasm\tsk this moment,}
\stick{now that all is lost, I see him galloping}
\stick{across me, corporal, to the left, to bring}
\stick{up the remains of the English horse}
\stick{along with him to support the right, and}
\stick{tear the laurel from Luxembourg’s}
\stick{brows, if yet ’tis possible\tsh I see him}
\stick{with the knot of his scarfe just shot off,}
\stick{infusing fresh spirits into poor Galway’s}
\stick{regiment\tsk riding along the line\tsk then}
\stick{wheeling about, and charging Conti at}
\stick{the head of it\tsh Brave! brave by hea-}
\stick{ven! cried my uncle Toby\tsk he deserves}
\catch{a} 
\stick{a crown\tsh As richly, as a thief a halter;} 
shouted Trim.

\vskip 3pt

My uncle Toby knew the corporal’s\break
\stick{loyalty;\tsk otherwise the comparison was}
\stick{not at all to his mind\tshh it did not al-}
\stick{together strike the corporal’s fancy when}
\stick{he had made it\tshh but it could not be}
\stick{recall’d\tsh so he had nothing to do, but}
proceed.

As the number of wounded was pro-\break
\stick{digious, and no one had time to think of}
\stick{any thing, but his own safety\tsk Though} 
\stick{Talmash, said my uncle Toby, brought}
\stick{off the foot with great prudence\tshh}
\stick{But I was left upon the field, said the}
\stick{corporal. Thou wast so; poor fellow!}
\stick{replied my uncle Toby\tshh So that it was} 
\stick{noon the next day, continued the cor-}
\catch{poral,}
\stick{poral, before I was exchanged, and}
\stick{put into a cart with thirteen or fourteen} 
\stick{more, in order to be convey’d to our}
hospital.

There is no part of the body, an’\break
\stick{please your honour, where a wound oc-}
\stick{casions more intolerable anguish than} 
upon the knee\tsh

Except the groin; said my uncle Toby.\break
\stick{An’ please your honour, replied the cor-}
\stick{poral, the knee, in my opinion, must} 
\stick{certainly be the most acute, there being}
\stck{so many tendons and what-d’ye-call-’ems} 
all about it.

It is for that reason, quoth my uncle\break
\stick{Toby, that the groin is infinitely more}
\stick{sensible\tsh there being not only as ma-}
\catch{ny} 
\stick{ny tendons and what-d’ye-call-’ems (for}
\stick{I know their names as little as thou}
\stick{do’st)\tsk about it\tsk but moreover}
\astW3\tsh

Mrs.\@ Wadman, who had been all the\break
\stick{time in her arbour\tsk instantly stopp’d her}
\stick{breath\tsk unpinn’d her mob at the chin,}
and stood upon one leg\tsh

The dispute was maintained with ami-\break
\stick{cable and equal force betwixt my uncle} 
\stick{Toby and Trim for some time; till}
\stick{Trim at length recollecting that he had} 
\stick{often cried at his master’s sufferings,} 
\stick{but never shed a tear at his own\tsk was} 
\stick{for giving up the point, which my uncle}
\stick{Toby would not allow\tsh ’Tis a proof} 
\stick{of nothing, Trim, said he, but the ge-}
nerosity of thy temper\tsh
\patch[8pt]{So}

So that whether the pain of a wound\break
\stick{in the groin (cæteris paribus) is great-}
\stick{er than the pain of a wound in the} 
knee\tsh or

Whether the pain of a wound in the\break
\stick{knee is not greater than the pain of a}
\stick{wound in the groin\tsh are points which}
to this day remain unsettled.

\chap{XX}{24pt}{0pt}

\dropcap{T}{he} anguish of my knee, con-\break
tinued the corporal, was excessive\break
\stick{in itself; and the uneasiness of the cart,}
\stick{with the roughness of the roads which}
\stick{were terribly cut up\tsk making bad still}
\stick{worse\tsk every step was death to me: so}
\stick{that with the loss of blood, and the}
\catch{want} 
\stick{want of care-taking of me, and a fever} 
\stick{I~felt coming on besides\tsh (Poor soul!}
\stck{said my uncle Toby) all together, an’ please} 
\stick{your honour, was more than I could} 
sustain.

I was telling my sufferings to a young\break
\stick{woman at a peasant’s house, where our} 
\stick{cart, which was the last of the line, had} 
\stick{halted; they had help’d me in, and the}
\stick{young woman had taken a cordial out of}
\stick{her pocket and dropp’d it upon some su-}
\stick{gar, and seeing it had cheer’d me, she had}
\stick{given it me a second and a third time\tsh}
\stick{So I was telling her, an’ please your ho-}
\stick{nour, the anguish I was in, and was}
\stick{saying it was so intolerable to me, that I}
\stick{had much rather lie down upon the bed,}
\stick{turning my face towards one which was}
\stick{in the corner of the room\tsk and die,}
\catch{than}
\stick{than go on\tsh when, upon her attempt-}
\stick{ing to lead me to it, I fainted away in}
\stick{her arms. She was a good soul! as your}
\stick{honour, said the corporal, wiping his} 
eyes, will hear.

I thought \i{love} had been a joyous thing,\break
quoth my uncle Toby.

’Tis the most serious thing, an’ please\break
\stick{your honour (sometimes) that is in the}
world.

By the persuasion of the young wo-\break
\stick{man, continued the corporal, the cart} 
\stick{with the wounded men set off without} 
\stick{me: she had assured them I should ex-}
\stick{pire immediately if I was put into the} 
\stick{cart. So when I came to myself\tsk I}
\stick{found myself in a still quiet cottage, with}
\catch{no} 
\stick{no one but the young woman, and the} 
\stick{peasant and his wife. I was laid across} 
\stick{the bed in the corner of the room, with} 
\stick{my wounded leg upon a chair, and the} 
\stick{young woman beside me, holding the}
\stick{corner of her handkerchief dipp’d in vi-}
\stick{negar to my nose with one hand, and rub-}
bing my temples with the other.

I took her at first for the daughter of\break
\stick{the peasant (for it was no inn)\tsk so had}
\stick{offer’d her a little purse with eighteen}
\stick{florins, which my poor brother Tom} 
\stick{(here Trim wip’d his eyes) had sent me}
\stick{as a token, by a recruit, just before he set}
out for Lisbon\tsh

\tsh I never told your honour that pi-\break
\stick{teous story yet\tsh here Trim wiped his}
eyes a third time.
\patch{The}

The young woman call’d the old man\break
\stick{and his wife into the room, to shew}
\stick{them the money, in order to gain me}
\stick{credit for a bed and what little ne-}
\stick{cessaries I should want, till I should be}
\stick{in a condition to be got to the hospital}
\stick{\tsh Come then! said she, tying up the}
\stick{little purse\tsk I’ll be your banker\tsk but as}
\stick{that office alone will not keep me em-}
ploy’d, I’ll be your nurse too.

I thought by her manner of speaking\break
\stick{this, as well as by her dress, which I}
\stick{then began to consider more attentively}
\stick{\tsk that the young woman could not be the}
daughter of the peasant.

She was in black down to her toes,\break
\stick{with her hair conceal’d under a cambrick} 
\stick{border, laid close to her forehead: she}
\catch{was}
\stick{was one of those kind of nuns, an’ please}
\stick{your honour, of which, your honour}
\stick{knows, there are a good many in Flan-}
\stick{ders which they let go loose\tsh By thy} 
\stick{description, Trim, said my uncle Toby,} 
\stick{I dare say she was a young Beguine, of}
\stick{which there are none to be found any} 
\stick{where but in the Spanish Netherlands\tsk}
\stick{except at Amsterdam\tshh they differ} 
\stick{from nuns in this, that they can quit}
\stick{their cloister if they choose to marry;}
\stick{they visit and take care of the sick by}
\stick{profession\tsh I had rather, for my own}
part, they did it out of good-nature.

\tsh She often told me, quoth Trim,\break
\stick{she did it for the love of Christ\tsk I did}
\stick{not like it.\tsh I believe, Trim, we are}
\stick{both wrong, said my uncle Toby\tsk we’ll}
\catch{ask} 
\stick{ask Mr.\@ Yorick about it to-night at} 
\stick{my brother Shandy’s\tshh so put me in}
mind; added my uncle Toby.

\vskip 8pt

The young Beguine, continued the\break
\stick{corporal, had scarce given herself time} 
\stick{to tell me \lqq she would be my nurse,”}
\stick{when she hastily turned about to begin}
\stick{the office of one, and prepare something}
\stick{for me\tsh and in a short time\tsk though I}
\stick{thought it a long one\tsk she came back}
\stick{with flannels, \etc \etc and having fo-}
\stick{mented my knee soundly for a couple of}
\stick{hours, \etc and made me a thin basin of}
\stick{gruel for my supper\tsk she wish’d me rest,} 
\stick{and promised to be with me early in the}
\stick{morning.\tsh She wish’d me, an’ please} 
\stick{your honour, what was not to be had.}
\catch{My}
\stick{My fever ran very high that night\tsk}
\stick{her figure made sad disturbance within}
\stick{me\tsk I was every moment cutting the} 
\stick{world in two\tsk to give her half of it\tsk}
\stick{and every moment was I crying, That} 
\stick{I had nothing but a knapsack and eight-}
\stick{een florins to share with her\tsk The}
\stick{whole night long was the fair Beguine,}
\stick{like an angel, close by my bedside, hold-}
\stick{ing back my curtain and offering me}
\stick{cordials\tsk and I was only awakened from} 
\stick{my dream by her coming there at the} 
\stick{hour promised, and giving them in re-}
\stick{ality. In truth, she was scarce ever from}
\stick{me, and so accustomed was I to receive}
\stick{life from her hands, that my heart sick-}
\stick{ened, and I lost colour when she left the}
\stick{room: and yet, continued the corporal,} 
\catch{(making}
\stick{(making one of the strangest reflections}
upon it in the world)\tsh

\tsh \lqq\i{It was not love}”\tsh for du-\break
\stick{ring the three weeks she was almost con-}
\stick{stantly with me, fomenting my knee}
\stick{with her hand, night and day\tsk I can}
\stick{honestly say, an’ please your honour\tsk}
\stick{that\enspace \wastfill}
\hbox to 0.7\hsize{\wastfill\enspace once.}

That was very odd, Trim, quoth my\break
uncle Toby\tsh

I think so too\tsk said Mrs.\@ Wadman.

It never did, said the corporal.
\patch[48pt]{\lsss{CHAP}.}


\chap{XXI}{35pt}{10pt}

\tshh But ’tis no marvel, continued\break 
\stick{the corporal\tsk seeing my uncle Toby}
\stick{musing upon it\tsk for Love, an’ please} 
\stick{your honour, is exactly like war, in this;}
\stick{that a soldier, though he has escaped}
\stick{three weeks compleat o’\,Saturday-night,}
\stick{\tsk may nevertheless be shot through his}
\stick{heart on Sunday morning\tsh \i{It happen}-}
\stick{\i{ed so here}, an’ please your honour, with}
\stick{this difference only\tsk that it was on Sun-}
\stick{day in the afternoon, when I fell in love}
\stick{all at once with a sisserara\tsh it burst up-}
\stick{on me, an’ please your honour, like a}
\stick{bomb\tsh scarce giving me time to say,}
\lqq God bless me.”
\patch[8pt]{I thought,}

I thought, Trim, said my uncle\break
Toby, a man never fell in love so very\break
suddenly.

Yes, an’ please your honour, if\break
he is in the way of it\tsk replied\break
Trim.

I prithee, quoth my uncle Toby,\break
inform me how this matter happened.

\tsh With all pleasure, said the cor-\break
poral, making a bow.
\patch[78pt]{\lsss{CHAP}.}
\baselineskip 13pt
\parskip 9pt

\chap{XXII}{30pt}{9pt}

\initial{I}{Had} escaped, continued the corporal,\break
all that time from falling in love, and\break
\stick{had gone on to the end of the chapter,}
\stick{had it not been predestined otherwise\tsh}
there is no resisting our fate.

It was on a Sunday, in the afternoon,\break
as I told your honour\tsh

The old man and his wife had walked\break
out\tsh

Every thing was still and hush as\break
midnight about the house\tsh

There was not so much as a duck or\break
a duckling about the yard\tsh
\patch[10pt]{\tsh When}
\parskip 13.5pt

\tstrut18pt\tshh When the fair Beguine came in\break
to see me.

My wound was then in a fair way of\break
\stick{doing well\tsh the inflammation had been}
\stick{gone off for some time, but it was suc-}
\stick{ceeded with an itching both above and} 
\stick{below my knee, so insufferable, that I}
\stick{had not shut my eyes the whole night}
for it.

Let me see it, said she, kneeling down\break
\stick{upon the ground parallel to my knee,} 
\stick{and laying her hand upon the part below}
\stick{it\tsh It only wants rubbing a little, said} 
\stick{the Beguine; so covering it with the bed} 
\stick{cloaths, she began with the forefinger} 
\stick{of her right-hand to rub under my knee,} 
\stick{guiding her fore-finger backwards and}
\catch{forwards}
\baselineskip 13.5pt
\parskip 8pt
\stick{forwards by the edge of the flannel which}
kept on the dressing.

In five or six minutes I felt slightly the\break
\stick{end of her second finger\tsk and presently}
\stick{it was laid flat with the other, and she} 
\stick{continued rubbing in that way round and} 
\stick{round for a good while; it then came in-}
\stick{to my head, that I should fall in love\tsk}
\stick{I blush’d when I saw how white a hand} 
\stick{she had\tsk I shall never, an’ please your}
\stick{honour, behold another hand so white}
whilst I live\tsh

\tsh Not in that place: said my uncle\break
Toby\tsh

Though it was the most serious des-\break
\stick{pair in nature to the corporal\tsk he could} 
not forbear smiling.
\patch{The}

The young Beguine, continued the\break
\stick{corporal, perceiving it was of great ser-}
\stick{vice to me\tsk from rubbing, for some} 
\stick{time, with two fingers\tsk proceeded to rub}
\stick{at length, with three\tsk till by little and}
\stick{little she brought down the fourth, and} 
\stick{then rubb’d with her whole hand: I} 
\stick{will never say another word, an’ please}
\stick{your honour, upon hands again\tsk but it}
was softer than satin\tsh

\tshh Prithee, Trim, commend it as\break
\stick{much as thou wilt, said my uncle Toby; I}
\stick{shall hear thy story with the more delight}
\stick{\tshh The corporal thank’d his master}
\stick{most unfeignedly; but having nothing to} 
\stick{say upon the Beguine’s hand, but the} 
\stick{same over again\tsh he proceeded to the} 
effects of\; it.
\patch[14pt]{The}

The fair Beguine, said the corporal,\break
\stick{continued rubbing with her whole hand} 
\stick{under my knee\tsk till I fear’d her zeal} 
\stick{would weary her\tshh \lqq I would do a}
\stick{\lqq thousand times more,” said she, \lqq for}
\stick{\lqq the love of Christ”\tsh In saying which}
\stick{she pass’d her hand across the flannel, to}
\stick{the part above my knee, which I had}
\stick{equally complained of, and rubb’d it}
also.

I perceived, then, I was beginning to\break
be in love\tsh

As she continued rub-rub-rubbing\tsk I\break
\stick{felt it spread from under her hand, an’} 
\stick{please your honour, to every part of my}
frame\tsh

The more she rubb’d, and the longer\break
\stick{strokes she took\tshh the more the fire}
\catch{kindled} 
\parskip 10pt
\stick{kindled in my veins\tshh till at length,}
\stick{by two or three strokes longer than the}
\stick{rest\tshh my passion rose to the highest}
pitch\tshh I seiz’d her hand\tush

\tshh And then, thou clapped’st it to\break 
\stick{thy lips, Trim, said my uncle Toby\tshh}
and madest a speech.

Whether the corporal’s amour termi-\break
\stick{nated precisely in the way my uncle Toby}
\stick{described it, is not material; it is enough} 
\stick{that it contain'd in it the essence of all the} 
\stick{love-romances which ever have been}
wrote since the beginning of the world.

\patch[68pt]{\lsss{CHAP}.}

\chap{XXIII}{32pt}{10pt}

\dropcap[findent=0pt]{A}{s} 
soon as the corporal had finished\break
the story of his amour\tsk or ra-\break
\stick{ther my uncle Toby for him\tsk Mrs.}
\stick{Wadman silently sallied forth from her}
\stick{arbour, replaced the pin in her mob,} 
\stick{pass’d the wicker gate, and advanced} 
\stick{slowly towards my uncle Toby’s sentry-}
\stick{box: the disposition which Trim had}
\stick{made in my uncle Toby’s mind, was too} 
favourable a crisis to be let slipp’d\tsh

\tsh The attack was determin’d upon:\break
\stick{it was facilitated still more by my uncle}
\stick{Toby’s having ordered the corporal to} 
\stick{wheel off the pioneer’s shovel, the spade,} 
\stick{the pick-axe, the picquets, and other}
\catch{military} 
\parskip 10pt
\stick{military stores which lay scatter’d upon}
\stick{the ground where Dunkirk stood\tsk The} 
\stck{corporal had march’d\tsk the field was clear.}

Now consider, sir, what nonsense it is,\break
\stick{either in fighting, or writing, or any}
\stick{thing else (whether in rhyme to it, or} 
\stick{not) which a man has occasion to do\tsk}
\stick{to act by plan: for if ever Plan, inde-}
\stck{pendent of all circumstances, deserved re-}
\stick{gistering in letters of gold (I mean in the}
\stick{archives of Gotham)\tsk it was certainly} 
\stick{the \s{Plan} of Mrs.\@ Wadman’s attack of} 
\stick{my uncle Toby in his sentry-box, \s{by}}
\stick{\s{Plan}\tsh Now the Plan hanging up in} 
\stick{it at this juncture, being the Plan of Dun-}
\stick{kirk\tsk and the tale of Dunkirk a tale of} 
\stick{relaxation, it opposed every impression}
\stick{she could make: and besides, could she}
\stick{have gone upon it\tsk the manœuvre of}
\catch{fingers} 
\parskip 12pt
\stick{fingers and hands in the attack of the}
\stick{sentry-box, was so outdone by that of the} 
\stick{fair Beguine’s, in Trim’s story\tsk that just}
\stick{then, that particular attack, however}
\stck{successful before\tsk became the most heart-}
less attack that could be made\tsk\!\tsk

O! let woman alone for this. Mrs.\break
\stick{Wadman had scarce open’d the wicker-}
\stick{gate, when her genius sported with the}
change of circumstances.

\tshh She formed a new attack in a\break
moment.


\patch[65pt]{\lsss{CHAP}.}
\parskip 8pt

\chap{XXIV}{30pt}{16pt}

\tsk I am half distracted, captain\break
\stick{Shandy, said Mrs.\@ Wadman, holding} 
\stick{up her cambrick handkerchief to her left} 
\stick{eye, as she approach’d the door of my} 
\stick{uncle Toby’s sentry-box\tsh a mote\tsh}
\stick{or sand\tsh or something\tsh I know not}
\stick{what, has got into this eye of mine\tsk\!\tsk}
\stick{do look into it\tsk it is not in the white\tsk\ } 

In saying which, Mrs.\@ Wadman edged\break
\stick{herself close in beside my uncle Toby,} 
\stck{and squeezing herself down upon the cor-}
\stick{ner of his bench, she gave him an op-}
\stick{portunity of doing it without rising up} 
\tush Do look into it\tsk said she.
\patch[12pt]{Honest}
\parskip 12pt

Honest soul! thou didst look into it\break
\stick{with as much innocency of heart, as ever}
\stick{child look’d into a raree-shew-box; and}
’twere as much a sin to have hurt thee.

\tsh If a man will be peeping of his\break
\stick{own accord into things of that nature}
\tsh I’ve nothing to say to it\tsh

My uncle Toby never did: and I will\break
\stick{answer for him, that he would have sat} 
\stick{quietly upon a sopha from June to Ja-}
\stick{nuary, (which, you know, takes in both}
\stick{the hot and cold months) with an eye as}
\stick{fine as the Thracian \fnast\  Rodope’s besides}

\bgroup\fontsize{8.5}{11}\selectfont
\indent\fnast\enspace 
Rodope Thracia tam inevitabili fascino\break
\stick{instructa, tam exacte oculis intuens attraxit,}
\stick{ut si in illam quis incidesset,\sic fieri non posset,}
quin caperetur.\tsk\!\tsk I know not who.\par\egroup
\smallskip

\patch{him,}
\parskip 8pt
\noindent
\stick{him, without being able to tell, whether}
it was a black or a blue one.

The difficulty was to get my uncle\break
Toby, to look at one, at all.

’Tis surmounted. And

I see him yonder with his pipe pendu-\break
\stick{lous in his hand, and the ashes falling} 
\stick{out of it\tsk looking\tsk and looking\tsk}
\stick{then rubbing his eyes\tsh and looking} 
\stick{again, with twice the good nature that}
\stick{ever Gallileo look’d for a spot in the sun.}

\tshh In vain! for by all the powers\break
\stick{which animate the organ\tsk Widow}
\stick{Wadman’s left eye shines this moment} 
\stick{as lucid as her right\tsk there is neither}
\stick{mote, or sand, or dust, or chaff, or speck,}
\stick{or particle of opake matter floating in it}
\catch{\tsk There\sic} 
\stick{\tshh there is nothing, my dear paternal}
\stick{uncle! but one lambent delicious fire,} 
\stick{furtively shooting out from every part}
of it, in all directions, into thine\tsh

\tshh If thou lookest, uncle Toby, in\break
\stick{search of this mote one moment longer}
\tshh thou art undone.

\chap{XXV}{24pt}{0pt}

\dropcap[findent=-1pt,lraise=0]{A}{n}
eye is for all the world exactly\break
like a cannon, in this respect;\break
\stick{That it is not so much the eye or the can-}
\stick{non, in themselves, as it is the carriage of} 
\stick{the eye\tsh and the carriage of the can-}
\stick{non, by which both the one and the other}
\stick{are enabled to do so much execution.} 
\stick{I don’t think the comparison a bad one:}
\stick{However, as ’tis made and placed at the}
\catch{head} 
\parskip 9pt
\stick{head of the chapter, as much for use as}
\stick{ornament, all I desire in return, is, that}
\stick{whenever I speak of Mrs.\@ Wadman’s} 
\stick{eyes (except once in the next period) that} 
you keep it in your fancy.

I protest, Madam, said my uncle\break 
\stick{Toby, I can see nothing whatever in} 
your eye.

It is not in the white; said Mrs Wad-\break
\stick{man: my uncle Toby look’d with might}
and main into the pupil\tsh

Now of all the eyes which ever were\break
\stick{created\tsh from your own, Madam, up} 
\stick{to those of Venus herself, which certainly} 
\stick{were as venereal a pair of eyes as ever} 
\stick{stood in a head\tsh there never was an} 
\stick{eye of them all, so fitted to rob my uncle}
\catch{Toby} 
\stick{Toby of his repose, as the very eye, at}
\stick{which he was looking\tshh it was not,} 
\stick{Madam, a rolling eye\tshh a romping or}
\stick{a wanton one\tsk nor was it an eye spark-}
\stick{ling\tsk petulant or imperious\tsk of high}
\stick{claims and terrifying exactions, which}
\stick{would have curdled at once that milk of}
\stick{human nature, of which my uncle Toby}
\stick{was made up\tsh but ’twas an eye full of}
\stick{gentle salutations\tsh and soft responses}
\stick{\tsh speaking\tsh not like the trumpet}
\stick{stop of some ill-made organ, in which} 
\stick{many an eye I talk to, holds coarse con-}
\stick{verse\tsh but whispering soft\tsh like the}
\stick{last low accent of an expiring saint\tsh}
\stick{\lqq How can you live comfortless, cap-}
\stick{\lqq tain Shandy, and alone, without a}
\stick{\lqq bosom to lean your head on\tshh or}
\lqq trust your cares to?\,”
\patch[8pt]{It}

It was an eye\tsh

But I shall be in love with it myself,\break
if I say another word about it.

\tsh It did my uncle Toby’s business.

\chap{XXVI}{24pt}{10pt}

\dropcap{T}{here} is nothing shews the cha-\break
racter of my father and my uncle\break
\stick{Toby, in a more entertaining light, than} 
\stick{their different manner of deportment,}
\stick{under the same accident\tsk for I call}
\stick{not love a misfortune, from a persuasion,}
\stick{that a man’s heart is ever the better for it}
\stick{\tshh Great God! what must my uncle}
\stick{Toby’s have been, when ’twas all be-}
nignity without it.
\patch[14pt]{My}

My father, as appears from many of\break
\stick{his papers, was very subject to this pas-}
\stick{sion, before he married\tsh but from a} 
\stick{little subacid kind of drollish impatience} 
\stick{in his nature, whenever it fefell\sic him, he} 
\stick{would never submit to it like a christian;} 
\stick{but would pish, and huff, and bounce,} 
\stick{and kick, and play the Devil, and write}
\stick{the bitterest Philippicks against the eye}
\stick{that ever man wrote\tsh there is one in} 
\stick{verse upon some body’s eye or other, that}
\stick{for two or three nights together, had put} 
\stick{him by his rest; which in his first trans-}
\stick{port of resentment against it, he begins} 
thus:

\bigskip
\bgroup\fontsize{8}{10}\selectfont
\stick{\indent\lqq A Devil ’tis\tsk and mischief such doth work}
\stick{\indent\lqq As never yet did Pagan, Jew, or Turk.”\enspace\fnast\quad}
\egroup

\bgroup\fontsize{9}{11}\selectfont
\indent\fnast\enspace This will be printed with my father’s\break
life of Socrates, \etc \etc\par
\egroup

\patch{In}

In short, during the whole paroxism,\break
\stick{my father was all abuse and foul lan-}
\stick{guage, approaching rather towards male-}
\stick{diction\tsh only he did not do it with as}
\stick{much method as Ernulphus\tsh he was}
\stick{too impetuous; nor with Ernulphus’s} 
\stick{policy\tsk for tho’ my father, with the} 
\stick{most intolerant spirit, would curse both}
\stick{this and that, and every thing under}
\stick{heaven, which was either aiding or abet-}
\stick{ting to his love\tsh yet never concluded} 
\stick{his chapter of curses upon it, without} 
\stick{cursing himself in at the bargain, as one}
\stick{of the most egregious fools and cox-}
\stick{combs, he would say, that ever was let}
loose in the world.

\vskip 4pt

My uncle Toby, on the contrary, took\break
\stick{it like a lamb\tsh sat still and let the}
\catch{poison} 
\stick{poison work in his veins without resis-}
\stick{tance\tsh in the sharpest exacerbations of}
\stick{his wound (like that on his groin) he} 
\stick{never dropt one fretful or discontented} 
\stick{word\tsh he blamed neither heaven nor}
\stick{earth\tsh or thought or spoke an inju-}
\stick{rious thing of any body, or any part of} 
\stick{it; he sat solitary and pensive with his}
\stick{pipe\tsh looking at his lame leg\tsh}
\stck{then whiffing out a sentimental heigh ho!}
\stick{which mixing with the smoak, incom-}
moded no one mortal.

He took it like a lamb\tsh I say.

\vskip 12pt

In truth he had mistook it at first; for\break
\stick{having taken a ride with my father, that}
\stick{very morning, to save if possible a beau-}
\stick{tiful wood, which the dean and chapter}
\catch{were} 
\stick{were hewing down to give to the poor \fnast;}
\stick{which said wood being in full view of my} 
\stick{uncle Toby’s house, and of singular} 
\stick{service to him in his description of the}
\stick{battle of Wynnendale\tsk\kern0.5pt by trotting on}
\stick{too hastily to save it\tsh upon an uneasy} 
\stick{saddle\tsh worse horse, \etc \etc .\@ .\@  it had}
\stick{so happened, that the serous part of the} 
\stick{blood had got betwixt the two skins, in}
\stick{the nethermost part of my uncle Toby}
\stick{\tsk the first shootings of which (as my} 
\stick{uncle Toby had no experience of love)} 
\stick{he had taken for a part of the passion\tsk}
\stick{till the blister breaking in the one case\tsk}
\stick{and the other remaining\tsk my uncle} 
\stick{Toby was presently convinced, that his}

\vskip 18pt

\bgroup\footnotesize
\stick{\indent\fnast\enspace Mr.\@ Shandy must mean the poor \i{in spirit};}
\stick{inasmuch as they divided the money amongst}
\stick{themselves.\hfill}
\par\egroup
\patch{wound}
\noindent
\stick{wound was not a skin-deep wound\tshh}
but that it had gone to his heart.

\chap{XXVII}{14pt}{0pt}

\dropcap[lraise=0]{T}{he} world is ashamed of being vir-\break
tuous\tshh My uncle Toby knew\break
\stick{little of the world; and therefore when} 
\stick{he felt he was in love with widow Wad-}
\stick{man, he had no conception that the}
\stick{thing was any more to be made a mystery} 
\stick{of, than if Mrs.\@ Wadman, had given} 
\stick{him a cut with a gap’d knife across his} 
\stick{finger: Had it been otherwise\tsh yet as}
\stick{he ever look’d upon Trim as a humble} 
\stick{friend; and saw fresh reasons every day}
\stick{of his life, to treat him as such\tshh it}
\stick{would have made no variation in the} 
\stick{manner in which he informed him of the}
\stick{affair.\hfill}
\catch{\lqq I}
\stick{\indent\lqq I am in love, corporal!” quoth my}
uncle Toby.

\chap{XXVIII}{28pt}{0pt}

\dropcap{I}{n} love!\tsh said the corporal\tsk your\break
honour was very well the day before\break
\stick{yesterday, when I was telling your ho-}
\stick{nour of the story of the King of Bohemia}
\stick{\tsk Bohemia! said my uncle Toby - - - -}
\stick{musing a long time - - -\enspace What became} 
of that story, Trim?

\tskk We lost it, an’ please your honour,\break
\stick{somehow betwixt us\tsk but your honour} 
\stick{was as free from love then, as I am\tsh}
\stck{’twas, just whilst thou went’st off with the}
\stick{wheel-barrow\tskk with Mrs.\@ Wadman,}
\stick{quoth my uncle Toby\tsh She has left a}
\catch{ball}
\parskip 8pt
\stick{ball here\tsk added my uncle Toby\tsk point-} 
ing to his breast\tsh

\tsh She can no more, an’ please your\break
\stick{honour, stand a siege, than she can fly\tsk}
cried the corporal\tsh

\tsh But as we are neighbours, Trim,\break
\stick{\tsk the best way I think is to let her know}
it civilly first\tsk quoth my uncle Toby.

Now if I might presume, said the cor-\break
poral, to differ from your honour\tsh

\tsh Why else, do I talk to thee Trim:\break
said my uncle Toby, mildly\tsh

\tight{\tsh Then I would begin, an’ please your}\break
\stick{honour, with making a good thundering} 
\stick{attack upon her, in return\tsk and telling} 
\stick{her civilly afterwards\tsk for if she knows}
\catch{any}
\parskip 10pt
\stick{any thing of your honour’s being in love,}
\stick{before hand\tsh L\bzz{or}d help her!\tsk she}
\stick{knows no more at present of it, Trim,} 
\stick{said my uncle Toby\tsk than the child}
unborn\tsh

Precious souls!\tush

Mrs.\@ Wadman had told it with all its\break
\stick{circumstances, to Mrs.\@ Bridget twenty-}
\stick{four hours before; and was at that very}
\stick{moment sitting in council with her,} 
\stick{touching some slight misgivings with re-}
\stick{gard to the issue of the affair, which the} 
\stick{Devil, who never lies dead in a ditch, had} 
\stick{put into her head\tsk before he would allow} 
\stick{half time, to get quietly through her,\sic} 
\i{te Deum}\tsh

I am terribly afraid, said widow Wad-\break
\stick{man, in case I should marry him, Bridget}
\catch{\tsk that}    
\baselineskip 12.725pt
\parskip 8pt
\stick{\tsk that the poor captain will not enjoy his}
\stick{health, with the monstrous wound upon}
his groin\tsh

It may not, Madam, be so very large,\break
\stick{replied Bridget, as you think\tshh and I}
\stick{believe besides, added she\tsk that ’tis dried}
up\tsh

\tsh I could like to know\tsk merely for\break
his sake, said Mrs.\@ Wadman\tsh

\tsk We’ll know and long and the broad
\stick{of it, in ten days\tsk answered Mrs.\@ Bridget,}
\stick{for whilst the captain is paying his ad-}
\stick{dresses to you\tsk I’m confident Mr.  Trim}
\stick{will be for making love to me\tsk and I’ll}
\stick{let him as much as he will\tsk added}
Bridget\tsk to get it all out of him\tsh

The measures were taken at once\tsh\break
\stick{and my uncle Toby and the corporal} 
went on with theirs.
\patch{Now,}
\baselineskip 13.5pt
\parskip 10pt

Now, quoth the corporal, setting his\break
\stick{left hand a kimbo, and giving such a} 
\stick{flourish with his right, as just promised}
\stick{success\tsk and no more\tsh if your honour}
\stick{will give me leave to lay down the plan} 
of this attack\tsh

\tsh Thou wilt please me by it, Trim,\break
\stick{said my uncle Toby, exceedingly\tsk and as}
\stick{I foresee thou must act in it as my \i{aid de}}
\stick{\i{camp}, here’s a crown, corporal, to begin} 
with, to steep thy commission.

Then, an’ please your honour, said the\break
\stick{corporal (making a bow first for his com-}
\stck{mission)\tsk we will begin with getting your}
\stick{honour’s laced cloaths out of the great}
\stick{campaign-trunk, to be well-air’d, and}
\stick{have the blue and gold taken up at the}
\stick{sleeves\tsk and I’ll put your white ramallie-}
\catch{wig}
\parskip 8pt
\stick{wig fresh into pipes\tsk and send for a} 
\stick{taylor, to have your honour’s thin}
scarlet breeches turn’d\tsh

\tsh I had better take the red plush ones,\break
\stick{quoth my uncle Toby\tsk They will be}
too clumsy\tsk said the corporal.

\chap{XXIX}{18pt}{0pt}

\tsh Thou wilt get a brush and a little\break
\stick{chalk to my sword\tsk\!\tsk ’Twill be only in}
your honour’s way, replied Trim.

\chap{XXX}{18pt}{0pt}

\tsk But your honour’s two razors\break
\stick{shall be new set\tsk and I will get my Mon-}
\stick{tero cap furbish’d up, and put on poor} 
\stick{lieutenant \hbox{Le Fever’s} regimental coat,}
\stick{which your honour gave me to}
\catch{wear}
\stick{wear for his sake\tsk and as soon as your} 
\stick{honour is clean shaved\tsk and has got your}
\stick{clean shirt on, with your blue and gold,} 
\stick{or your fine scarlet\tsh sometimes one} 
\stick{and sometimes t’other\tsk and every thing} 
\stick{is ready for the attack\tsk we’ll march up}
\stick{boldly, as if ’twas to the face of a bas-}
\stick{tion; and whilst your honour engages} 
\stick{Mrs.\@  Wadman in the parlour, to the}
\stick{right\tsh I’ll attack Mrs.\@ Bridget in the}
\stick{kitchen, to the left; and having seiz’d}
\stick{that pass, I’ll answer for it, said the cor-}
\stick{poral, snapping his fingers over his head}
\tsk that the day is our own.

\vskip 2pt

I wish I may but manage it right; said\break
\stick{my uncle Toby\tsk but I declare, corporal,}
\stick{I had rather march up to the very edge of}
a trench\tsk\!\tsk
\patch[8pt]{\tsk A}

\tsk A woman is quite a different thing\break
\tsk said the corporal.

\tsk I suppose so, quoth my uncle Toby.

\chap{XXXI}{18pt}{0pt}

\dropcap{I}{f} any thing in this world, which my\break
father said, could have provoked my\break
\stick{uncle Toby, during the time he was in} 
\stick{love, it was the perverse use my father} 
\stick{was always making of an expression of} 
\stick{Hilarion the hermit; who, in speaking} 
\stick{of his abstinence, his watchings, flagel-}
\stick{lations, and other instrumental parts of}
\stick{his religion\tsk would say\tsk tho’ with more}
\stick{facetiousness than became an hermit\tsk} 
\stick{\lqq That they were the means he used, to} 
\stick{make his \i{ass} (meaning his body) leave}
off kicking.”
\patch[8pt]{It}

It pleased my father well; it was not\break
\stick{only a laconick way of expressing\tsh}
\stick{but of libelling, at the same time, the}
\stick{desires and appetites of the lower part of}
\stick{us; so that for many years of my fa-}
\stick{ther’s life, ’twas his constant mode of} 
\stick{expression\tsk he never used the word \i{pas}-}
\stick{\i{sions} once\tsk but \i{ass} always instead of}
\stick{them\tsh So that he might be said truly,} 
\stick{to have been upon the bones, or the back} 
\stick{of his own ass, or else of some other}
man’s, during all that time.

\vskip 2pt

I must here observe to you, the differ-\break
\stick{ence betwixt\hfill}
\stick{\kern 2.8em My father’s ass\hfill}
\stick{\kern 2.8em and my hobby-horse\tsk in order to}
\stick{keep characters as separate as may be, in} 
our fancies as we go along.
\patch[12pt]{For}

For my hobby-horse, if you recollect\break
\stick{a little, is no way a vicious beast; he has}
\stick{scarce one hair or lineament of the ass} 
\stick{about him\tshh ’Tis the sporting little} 
\stick{filly-folly which carries you out for the\sic} 
\stick{the present hour\tsk a maggot, a butterfly,}
\stick{a picture, a fiddle-stick\tsk an uncle Toby’s}
\stck{siege\tsk or an \i{any thing}, which a man makes}
\stick{a shift to get a stride on, to canter it away} 
\stick{from the cares and solicitudes of life\tsk} 
\stick{’Tis as useful a beast as is in the whole}
\stick{creation\tskk nor do I really see how the}
world could do without it\tush

\tsh But for my father’s ass\tush oh!\break
\stick{mount him\tsk mount him\tsk mount him\tsk}
\stick{(that’s three times, is it not?)\tsk mount}
\stick{him not:\tsk ’tis a beast concupiscent\tsk and} 
\stick{foul befall the man, who does not hinder}
him from kicking.
\patch{\lsss{CHAP}.}
\parskip 7pt

\chap{XXXII}{28pt}{10pt}

\dropcap{W}{ell}! dear brother Toby, said\break
my father, upon his first seeing\break
\stick{him after he fell in love\tsk and how goes} 
it with your \s{Asse} ?

Now my uncle Toby thinking more\break
\stick{of the \i{part} where he had had the blister,}
\stick{than of Hilarion’s metaphor\tsk and our}
\stick{preconceptions having (you know) as} 
\stick{great a power over the sounds of words}
\stick{as the shapes of things, he had imagined,}
\stick{that my father, who was not very cere-}
\stick{monious in his choice of words, had en-}
\stick{quired after the part by its proper name;} 
\stick{so notwithstanding my mother, doctor} 
\stick{Slop, and Mr.\@ Yorick, were sitting in}
\stick{the parlour, he thought it rather civil to}
\catch{conform} 
\parskip 10pt
\stick{conform to the term my father had} 
\stick{made use of than not. When a man is}
\stick{hemm’d in by two indecorums, and must}
\stick{commit one of ’em\tsk I always observe\tsk}
\stick{let him choose which he will, the world} 
\stick{will blame him\tsk so I should not be asto-}
nished if it blames my uncle Toby.

My A\bzz{rs}e, quoth my uncle Toby, is\break
\stick{much better\tsk brother Shandy\tsk My fa-}
\stick{ther had formed great expectations from}
\stick{his Asse in this onset; and would have} 
\stick{brought him on again; but doctor Slop} 
\stick{setting up an intemperate laugh\tsk and my} 
\stick{mother crying out L\bzz{ord} bless us!\tskk it}
\stick{drove my father’s Asse off the field\tsk and}
\stick{the laugh then becoming general\tsk there} 
\stick{was no bringing him back to the charge,} 
for some time\tsh
\patch[8pt]{And}
\parskip 9pt

And so the discourse went on without\break
him.

Every body, said my mother, says\break
\stick{you are in love, brother Toby\tsk and we}
hope it is true.

I am as much in love, sister, I believe,\break
\stick{replied my uncle Toby, as any man usu-}
\stick{ally is\tshh Humph! said my father\tshh}
\stick{and when did you know it? quoth my}
mother\tsh

\tsh When the blister broke; replied\break
my uncle Toby.

My uncle Toby’s reply put my father\break
\stick{into good temper\tsk so he charged o’foot.}

\patch[38pt]{\lsss{CHAP}.}

\chap{XXXIII}{24pt}{9pt}

\dropcap[findent=0pt]{A}{s} the antients agree, brother Toby,\break
said my father, that there are two\break
\stick{different and distinct kinds of \i{love}, ac-}
\stick{cording to the different parts which are} 
\stick{affected by it\tsk the Brain or Liver\tsh I}
\stick{think when a man is in love, it behoves}
\stick{him a little to consider which of the two} 
he is fallen into.

What signifies it, brother Shandy, re-\break
\stick{plied my uncle Toby, which of the two}
\stick{it is, provided it will but make a man} 
\stick{marry, and love his wife, and get a few}
children.

\tsh A few children! cried my father,\break
\stick{rising out of his chair, and looking full}
\catch{in} 
\baselineskip 13.3333pt
\parskip 9pt
\stick{in my mother’s face, as he forced his}
\stick{way betwixt her’s and doctor Slop’s\tsk a} 
\stick{few children! cried my father, repeating} 
\stick{my uncle Toby’s words as he walk’d to} 
and fro’\tsh

\tshh Not, my dear brother Toby, cried\break
\stick{my father, recovering himself all at once,}
\stick{and coming close up to the back of my} 
\stick{uncle Toby’s chair\tsk not that I should be}
\stick{sorry had’st thou a score\tsk on the con-}
\stick{trary, I should rejoice\tsk and be as kind,}
\stick{Toby, to every one of them as a father\tsk} 

My uncle Toby stole his hand unper-\break
\stick{ceived behind his chair, to give my fa-}
ther’s a squeeze\tsh

\tshh Nay, moreover, continued he,\break
\stick{keeping hold of my uncle Toby’s hand}
\stick{\tsk so much do’st thou possess, my dear}
\catch{Toby,}
\baselineskip 13.5pt
\parskip 12pt  
\stick{Toby, of the milk of human nature,}
\stick{and so little of its asperities\tsk ’tis piteous}
\stick{the world is not peopled by creatures}
\stick{which resemble thee; and was I an Asia-}
\stick{tick monarch, added my father, heating}
\stick{himself with his new project\tsk I would}
\stick{oblige thee, provided it would not impair}
\stick{thy strength\tsk or dry up thy radical mois-}
\stick{ture too fast\tsk or weaken thy memory or}
\stick{fancy, brother Toby, which these gym-}
\stick{nicks inordinately taken, are apt to do\tsk}
\stick{else, dear Toby, I would procure thee}
\stick{the most beautiful woman in my empire,}
\stick{and I would oblige thee, \i{nolens, volens}, to}
beget for me one subject every \i{month}\tsh

As my father pronounced the last\break
\stick{word of the sentence\tsk my mother took} 
a pinch of snuff.
\patch[12pt]{Now}
\parskip 6pt

Now I would not, quoth my uncle\break
\stick{Toby, get a child, \i{nolens, volens}, that}
\stick{is, whether I would or no, to please the}
greatest prince upon earth\tsh

\tsk And ’twould be cruel in me,\break
\stick{brother Toby, to compell thee; said}
\stick{my father\tsk but ’tis a case put to shew} 
\stick{thee, that it is not thy begetting a child}
\stick{\tsk in case thou should’st be able\tsk but the}
\stick{system of Love and marriage thou goest} 
upon, which I would set thee right in\tsh

There is at least, said Yorick, a great\break
\stick{deal of reason and plain sense in captain}
\stick{Shandy’s opinion of love; and ’tis} 
\stick{amongst the ill spent hours of my life}
\stick{which I have to answer for, that I have}
\stick{read so many flourishing poets and rhe-}
\stick{toricians in my time, from whom I never}
could extract so much\tsh
\patch{I wish,}

I wish, Yorick, said my father, you\break 
\stick{had read Plato; for there you would} 
\stick{have learnt that there are two \s{Loves}\tsk}
\stick{I know there were two \s{Religions}, re-}
\stick{plied Yorick, amongst the ancients\tsh}
\stick{one\tsk for the vulgar, and another for the}
\stick{learned;\tsk but I think \s{one Love} might}
have served both of them very well\tsk 

\vskip 4pt

It could not; replied my father\tsk and\break
\stick{for the same reasons: for of these Loves,}
\stick{according to Ficinus’s comment upon Ve-}
\stick{lasius, the one is \i{rational}\tsh\hfill}
\stick{\quad\tsh the other is \i{natural}\tsh\hfill}
\stick{the first ancient\tsh without mother\tsh}
\stick{where Venus had nothing to do: the} 
second, begotten of Jupiter and Dione\tsk

\tsh Pray brother, quoth my uncle\break
\stick{Toby, what has a man who believes in}
\stick{God to do with this? My father could}
\catch{not}
\parskip 2pt
\stick{not stop to answer, for fear of breaking} 
the thread of his discourse\tsh

This latter, continued he, partakes\break
wholly of the nature of Venus.

The first, which is the golden chain\break
\stick{let down from heaven, excites to love he-}
\stck{roic, which comprehends in it, and excites}
\stick{to the desire of philosophy and truth\tsh}
the second, excites to \i{desire}, simply\tsh
\parskip 6pt

\tshh \tight{I think the procreation of children}\break
\stick{as beneficial to the world, said Yorick, as} 
the finding out the longitude\tsh

\tshh To be sure, said my mother, \i{love}\break
keeps peace in the world\tsh

\tshh In the \i{house}\tsk my dear, I own\tsk\break
\stick{\tsh It replenishes the earth; said my}
mother\tsh

But it keeps heaven empty\tsk my dear;\break
replied my father.
\patch{\tsh ’Tis}
\parskip8pt

\tsk ’Tis Virginity, cried Slop, tri-\break
umphantly, which fills paradise.

Well push’d nun! quoth my father.

\chap{XXXIV}{16pt}{0pt}

\dropcap{M}{y} father had such a skirmishing,\break
cutting kind of a slashing way\break
\stick{with him in his disputations, thrusting} 
\stick{and ripping, and giving every one a}
\stick{stroke to remember him by in his turn\tsk}
\stick{that if there were twenty people in com-}
\stick{pany\tsk in less than half an hour he was}
\stick{\tight{sure to have every one of ’em against him.}\hss}

\noindent
\stick{\indent\tight{What did not a little contribute to leave}}
\stick{him thus without an ally, was, that if}
\stick{there was any one post more untenable}
\stick{than the rest, he would be sure to throw}
\stick{himself into it; and to do him justice,}
\stck{when he was once there, he would defend}
\catch{it} 
\stick{it so gallantly, that ’twould have been a} 
\stick{concern, either to a brave man, or a}
\stick{good-natured one, to have seen him}
driven out.

Yorick, for this reason, though he\break
\stick{would often attack him\tsk yet could never} 
bear to do it with all his force.

Doctor Slop’s \s{Virginity}, in the close\break
\stick{of the last chapter, had got him for} 
\stick{once on the right side of the rampart;} 
\stick{and he was beginning to blow up all the} 
\stick{convents in Christendom about Slop’s} 
\stick{ears, when corporal Trim came into the} 
\stick{parlour to inform my uncle Toby, that}
\stick{his thin scarlet breeches, in which the} 
\stick{attack was to be made upon Mrs.\@ Wad-}
\stick{man, would not do; for, that the tay-}
\stick{lor, in ripping them up, in order to turn} 
\stick{them, had found they had been turn’d}
\catch{before} 
\stick{before\tshh Then turn them again, bro-}
\stick{ther, said my father rapidly, for there}
\stick{will be many a turning of ’em yet be-}
\stick{fore all’s done in the affair\tshh They are}
\stick{as rotten as dirt, said the corporal\tsh}
\stick{Then by all means, said my father, be-}
\stick{speak a new pair, brother\tshh for though} 
\stick{I know, continued my father, turning}
\stick{himself to the company, that widow}
\stick{Wadman has been deeply in love with} 
\stick{my brother Toby for many years, and}
\stick{has used every art and circumvention of} 
\stick{woman to outwit him into the same} 
\stick{passion, yet now that she has caught}
\stick{him\tshh her fever will be pass’d it’s\sic}
height\tsh

\tsh She has gain’d her point.

In this case, continued my father,\break
\stick{which Plato, I am persuaded, never}
\catch{thought}
\parskip 12pt
\stick{thought of\tshh Love, you see, is not}
\stick{so much a \s{Sentiment} as a \s{Situation},} 
\stick{into which a man enters, as my brother} 
\stick{Toby would do, into a \i{corps}\tshh no} 
\stick{matter whether he loves the service or}
\stick{no\tshh being once in it\tskk he acts as if he} 
\stick{did; and takes every step to shew him-}
self a man of prowesse.

The hypothesis, like the rest of my\break 
\stick{father’s, was plausible enough, and my} 
\stick{uncle Toby had but a single word to ob-}
\stick{ject to it\tsk in which Trim stood ready to} 
\stick{second him\tshh but my father had not} 
drawn his conclusion\tsh

For this reason, continued my father\break
\stick{(stating the case over again) notwith-}
\stick{standing all the world knows, that Mrs.}
\catch{Wadman}
\stick{Wadman \i{affects} my brother Toby\tsk and}
\stick{my brother Toby contrariwise \i{affects}} 
\stick{Mrs.\@ Wadman, and no obstacle in nature} 
\stick{to forbid the music striking up this very} 
\stick{night, yet will I answer for it, that this}
\stick{self-same tune will not be play’d this} 
twelvemonth.

We have taken our measures badly,\break
\stick{quoth my uncle Toby, looking up in-}
terrogatively in Trim’s face.

I would lay my Montero cap, said\break
\stick{Trim\tshh Now Trim’s Montero-cap, as}
\stick{I once told you, was his constant wager;}
\stick{and having furbish’d it up that very} 
\stick{night, in order to go upon the attack\tsk}
\stick{it made the odds look more considerable}
\stick{\tshh I would lay, an’ please your ho-}
\stick{nour, my Montero-cap to a shilling\tsk}
\catch{was} 
\parskip 6pt
\stick{was it proper, continued Trim (making}
\stick{a bow) to offer a wager before your ho-}
nours\tsh

\tshh There is nothing improper in it,\break
\stick{said my father\tsk ’tis a mode of expres-}
\stick{sion; for in saying thou would’st lay thy}
\stck{Montero-cap to a shilling\tsk all thou mean-}
est is this\tsk that thou believest\tsh

\tshh Now, What do’st thou believe?

That Widow Wadman, an’ please your\break
worship, cannot hold it out ten days\tsh

\tight{And whence, cried Slop, jeeringly, hast}\break
\tight{thou all this knowledge of woman, friend?}

By falling in love with a popish cler-\break
gy-woman; said Trim.

’Twas a Beguine, said my uncle Toby.
\patch{Doctor}
\baselineskip 13pt

Doctor Slop was too much in wrath\break
\stick{to listen to the distinction; and my fa-}
\stick{ther taking that very crisis to fall in hel-}
\stick{ter-skelter upon the whole order of Nuns} 
\stick{and Beguines, a set of silly, fusty bag-}
\stick{gages\tshh Slop could not stand it\tshh}
\stick{and my uncle Toby having some mea-}
\stick{sures to take about his breeches\tsk and} 
\stick{Yorick about his fourth general division}
\stick{\tsk in order for their several attacks next}
\stick{day\tsk the company broke up: and my}
\stick{father being left alone, and having half} 
\stick{an hour upon his hands betwixt that and} 
\stick{bed-time; he called for pen, ink, and}
\stick{paper, and wrote my uncle Toby the fol-}
lowing letter of instructions.


\patch[40pt]{My}
\baselineskip 13.5pt
\tstrut 27pt My dear brother Toby,

\vskip -4pt

\dropcap[loversize=-0.15,lraise=0.3,nindent=1pt]{W}{hat} I am going to say to\break 
thee is upon the nature of wo-\break
\stick{men, and of love-making to them;}
\stick{and perhaps it is as well for thee\tsk tho’} 
\stick{not so well for me\tsk that thou hast occa-}
\stick{sion for a letter of instructions upon that} 
\stick{head, and that I am able to write it to}
thee.

Had it been the good pleasure of him\break
\stick{who disposes of our lots\tsk and thou no}
\stick{sufferer by the knowledge, I had been}
\stick{well content that thou should’st have}
\stick{dipp’d the pen this moment into the ink,}
\stick{instead of myself; but that not being the}
\stick{case\tsh\!\tsh Mrs Shandy being now} 
\stick{close besides me, preparing for bed\tsh}
\catch{I have}
\baselineskip 13pt
\stick{I have thrown together without order,}
\stick{and just as they have come into my mind,}
\stick{such hints and documents as I deem may}
\stick{be of use to thee; intending, in this, to} 
\stick{give thee a token of my love; not}
\stick{doubting, my dear Toby, of the manner} 
in which it will be accepted.

\vskip 6.5pt

In the first place, with regard to all\break
\stick{which concerns religion in the affair\tsh}
\stick{though I perceive from a glow in my} 
\stick{cheek, that I blush as I begin to speak}
\stick{to thee upon the subject, as well know-}
\stick{ing, notwithstanding thy unaffected se-}
\stick{crecy, how few of its offices thou neglect-}
\stick{est\tskk yet I would remind thee of one}
\stick{(during the continuance of thy courtship)} 
\stick{in a particular manner, which I would} 
\stick{not have omitted; and that is, never to}
\stick{go forth upon the enterprize, whether it}
\stick{be in the morning or the afternoon, with-}
\catch{out}
\parskip 8pt
\stick{out first recommending thyself to the}
\stick{protection of Almighty God, that he}
may defend thee from the evil one.

\tight{Shave the whole top of thy crown clean,}\break
\stick{once at least every four or five days, but}
\stick{oftner if convenient; lest in taking off} 
\stick{thy wig before her, thro’ absence of mind,}
\stick{she should be able to discover how much} 
\stick{has been cut away by Time\tsk how} 
much by Trim.

\tsh ’Twere better to keep ideas of bald-\break
ness out of her fancy.

Always carry it in thy mind, and act\break
upon it, as a sure maxim, Toby\tsh

\lqq\i{That women are timid:}” And ’tis\break
\stick{well they are\tsk else there would be no}
dealing with them.
\patch[13pt]{Let}
\parskip 10pt

Let not thy breeches be too tight, or\break
\stick{hang too loose about thy thighs, like the} 
trunk-hose of our ancestors.

\tshh A just medium prevents all con-\break
clusions.

Whatever thou hast to say, be it more\break
\stick{or less, forget not to utter it in a low soft}
\stick{tone of voice. Silence, and whatever}
\stick{approaches it, weaves dreams of mid-}
\stick{night secrecy into the brain: For this} 
\stick{cause, if thou canst help it, never throw} 
down the tongs and poker.

Avoid all kinds of pleasantry and face-
\stick{tiousness in thy discourse with her, and}
\stick{do whatever lies in thy power at the same}
\stick{time, to keep from her all books and}
\stick{writings which tend thereto: there are}
\stick{some devotional tracts, which if thou}
\catch{canst}
\stick{canst entice her to read over\tsk it will be}
\stick{well: but suffer her not to look into Ra-}
\stick{belais, or Scarron, or Don Quixote\tsh}

\tsh They are all books which excite\break
\stick{laughter; and thou knowest, dear Toby,} 
\stck{that there is no passion so serious, as lust.}

Stick a pin in the bosom of thy shirt,\break
before thou enterest her parlour.

\tight{And if thou art permitted to sit upon the}\break
\stick{same sopha with her, and she gives thee} 
\stick{occasion to lay thy hand upon hers\tsk be-}
\stick{ware of taking it\tsh thou can’st not lay}
\stick{thy hand on hers, but she will feel the} 
\stick{temper of thine. Leave that and as} 
\stick{many other things as thou canst, quite}
\stick{undetermined; by so doing, thou will}
\stick{have her curiosity on thy side; and if she}
\stick{is not conquer’d by that, and thy \s{Asse}}
\catch{continues}
\stick{continues still kicking, which there is} 
\stick{great reason to suppose\tshh Thou must}
\stick{begin, with first losing a few ounces of}
\stick{blood below the ears, according to the}
\stick{practice of the ancient Scythians, who} 
\stick{cured the most intemperate fits of the}
appetite by that means.

\vskip -4pt

Avicenna, after this, is for having the\break
\stick{part anointed with the syrrup of helle-}
\stick{bore, using proper evacuations and}
\stick{purges\tshh and I believe rightly. But} 
\stick{thou must eat little or no goat’s flesh,}
\stick{nor red deer\tshh nor even foal’s flesh by}
\stick{any means; and carefully abstain\tshh} 
\stick{that is, as much as thou canst, from pea-}
\stick{cocks, cranes, coots, didappers, and} 
water-hens\tsh

\vskip -6pt

As for thy drink\tsk I need not tell\break
\stick{thee, it must be the infusion of \s{Vervain}}
\stick{and the herb \s{Hanea}, of which Ælian}
\catch{relates}
\parskip 4pt
\stick{relates such effects\tsk but if thy stomach}
\stick{palls with it\tsk discontinue it from time to}
\stick{time, taking cucumbers, melons, purs-}
\stick{lane, water-lillies, woodbine, and lettice,}
in the stead of them.

There is nothing further for thee,\break
which occurs to me at present\tsh

\tskk Unless the breaking out of a fresh\break
\stick{war\tshh So wishing every thing, dear}
\stick{Toby, for best,\hfill}
\stick{\tstrut 14pt\indent I rest thy affectionate brother,\hfill}
\stick{\tstrut 14pt\hfill\s{Walter Shandy}.}

\chap{XXXV}{18pt}{0pt}

\dropcap{W}{hilst} my father was writing\break
his letter of instructions, my\break
\stick{uncle Toby and the corporal were busy}
\stick{in preparing every thing for the attack.}
\stick{As the turning of the thin scarlet}
\catch{breeches} 
\baselineskip 13.5pt
\parskip 8pt
\stick{breeches was laid aside (at least for the}
\stick{present) there was nothing which should}
\stick{put it off beyond the next morning; so}
\stick{accordingly it was resolv’d upon, for}
eleven o’clock.

Come, my dear, said my father to my\break
\stick{mother\tsk ’twill be but like a brother and}
\stick{sister, if you and I take a walk down to}
\stick{my brother Toby’s\tshh to countenance}
him in this attack of his.

My uncle Toby and the corporal had\break
\stick{been accoutred both some time, when}
\stick{my father and mother enter’d, and the}
\stick{clock striking eleven, were that moment}
\stick{in motion to sally forth\tsk but the account}
\stick{of this is worth more, than to be wove}
\stick{into the fag end of the eighth volume}
\stick{of such a work as this.\tshh My father}
\stick{had no time but to put the letter of in-}
\catch{structions} 
\stick{structions into my uncle Toby’s coat-}
\stick{pocket\tshh and join with my mother in}
wishing his attack prosperous.

I could like, said my mother, to look\break
\stick{through the key-hole out of \i{curiosity}\tsh}
\stick{Call it by it’s\sic\ right name, my dear,}
quoth my father\tsk 

\i{And look through the key-hole} as long as\break
you will.

\vskip 54pt

\noindent
\centerline{\hbox to 132pt{\s{End} of the \s{Eighth} \s{Volume}.}}
\vfill
\eject
\pagestyle{empty}
\null
\end{document}
