\documentclass[twoside]{article}
\usepackage[paperheight=395pt, paperwidth=240pt, width=166pt, height=294pt, headsep=2pt]{geometry}
\usepackage[noinfo, frame, a6, center, lualatex]{crop}
\usepackage{graphicx}
\usepackage{shandean}
\title{The Life and Opinions of Tristram Shandy, Gentleman}
\author{Laurence Sterne}
\begin{document}
\pagestyle{empty}
% half title
\hbox{}
\vfill
\vbox{\openup 16pt\halign to \textwidth{\hss #\hss\cr
\hbox to 3em{T H E}\cr
\hbox to 0.6\textwidth{\Huge L I F E}\cr
\hbox to 3em{A N D}\cr
\hbox to \textwidth{\Huge O P I N I O N S}\cr
\textsc{o f}\cr
\hbox{\Large \ls{TRISTRAM SHANDY},}\cr
\textsc{\lss{Gentleman}.}\cr
}}
\vfill
\newpage
\null
\newpage % full title
\pagestyle{empty}
\vbox{\openup 10pt\halign{\hss #\hss\cr
\hbox to 3em{T H E}\cr
\lower 2pt\hbox to 10em{\Large L I F E}\cr
\hbox to 3em{A N D}\cr
\lower 3pt\hbox to 15.3em{\Large O P I N I O N S}\cr
\textsc{o f}\cr
\hbox{\large\lsss{TRISTRAM SHANDY},}\cr
\lower-2pt\hbox{\textsc{\lss{Gentleman}.}}\cr
}}
\vfill
\vbox{\openup -2pt\halign to 180pt{\footnotesize #\cr
Multitudinis imperitæ non formido judicia; meis\cr
\quad tamen, rogo, parcant opusculis\tsh in quibus\cr
\quad fuit propositi semper, a jocis ad seria, a seriis\cr
\quad vicissim ad jocos transire.\hfill\cr
\hfill \textscl{Joan. Saresberiensis},\cr
\hfill \textit{Episcopus Lugdun.}\cr}}
\vfill
\centerline{\lsss{VOL}.\quad IV.}
\vfill
\centerline{\itshape\lsss{LONDON}:}
\centerline{\smaller Printed for R.\@ and J.\@ \lss{\textsc{Dodsley}} in \textit{Pall-Mall}.}
\centerline{M.\,DCC.\,LXI.}
\newpage
\null
\newpage
\null
\vfill
\noindent
\textsc{Vol. IV.}
\setcounter{page}{1}
\setlength{\baselineskip}{14pt}  % 21*14 = 294
\newpage
\centerline{\Large\lsss{SLAWKENBERGII}}
\smallskip
\centerline{\large\lss{\textsc{Fabella}}.}

\bgroup\itshape
\initial[depth=1,loversize=0.4, lraise=-0.25,findent=8pt]{V}{\lsss{ESPERA}} 
quâdam frigidulâ, po-\break
steriori in parte mensis {\normalshape Augusti},\break
peregrinus, mulo fusco colore insi-\break
\stick{dens, manticâ a tergo, paucis indusijs,}
\stick{binis calceis, braccisque sericis coccinejs re-}
pletâ {\normalshape Argentoratum} ingressus est.

Militi eum percontanti, quum portus in-\break
\stick{traret, dixit, se apud Nasorum promonto-}
\stick{rium fuisse, Francofurtum proficisci, et Ar-}
\stick{gentoratum, transitu ad fines Sarmatiæ}
\stick{mensis intervallo, reversurum.\hfill}
\stick{\hfill Miles}
\egroup

\vfill
\bgroup\fontsize{8}{10}\selectfont
\indent\fnast As \textit{Hafen Slawkenbergius de Nasis} is extremely\break
\stick{scarce, it may not be unacceptable to the learned}
\stick{reader to see the specimen of a few pages of his}
\stick{original; I will make no reflection upon it, but}
\stick{that his story-telling Latin is much more concise}
\stick{than his philosophic\tsk and, I think, has more of}
Latinity in it.\par
\egroup
\newpage
\centerline{\chapstrut\Large\lss{SLAWKENBERGIUS}'s}
\smallskip
\centerline{\large\lss{\textsc{Tale}}.}

\kern \baselineskip

\dropcap[lines=3]{I}{t} was one cool refreshing evening,\break
at the close of a very sultry day, in\break
\tight{the latter end of the month of \i{August},}\break
\stick{when a stranger, mounted upon a dark}
\stick{mule, with a small cloak-bag behind}
\stick{him, containing a few shirts, a pair of}
\stick{shoes, and a crimson-sattin pair of}
breeches, entred\sic the town of \i{Strasburg}.

He told the centinel, who questioned
\stick{him as he entered the gates, that he had}
\stick{been at the promontory of \s{Noses}\tsk was}
\stick{going on to \i{Frankfort}\tsh and should be}
\stick{back again at \i{Strasburg} that day month,}
\stick{in his way to the borders of \i{Crim}-}
\i{Tartary}.\patch{The}

\pagestyle{fancy}
\noindent
\stick{\indent\i{Miles peregrini in faciem suspexit\tsk Di}}
\stick{\i{boni, nova forma nasi!}\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{At multum mihi profuit, inquit pere-}}
\stick{\i{grinus, carpum amento extrahens, e quo}}
\stick{\i{pependit acinaces: Loculo manum inse-}}
\stick{\i{ruit}; \et\ \i{magnâ cum urbanitate, pilei parte}}
\stick{\i{anteriore tactâ manu sinistrâ, ut extendit}}
\stick{\i{dextram, militi florinum dedit et processit.}\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Dolet mihi, ait miles, tympanistam na-}}
\stick{\i{num et valgum alloquens, virum adeo ur-}}
\stick{\i{banum vaginam perdidisse}; \i{itinerari haud}}
\stick{\i{poterit nudâ acinaci, neque vaginam toto}}
\stick{Argentorato, \i{habilem inveniet.\tsh Nullam}}
\stick{\i{unquam habui, respondit peregrinus respi-}}
\natch{\i{ciens},}

The centinel looked up into the stran-\break
\stick{ger’s face\tsh never saw such a nose in his}
life!

\tsk I have made a very good venture of\break
\stick{it, quoth the stranger\tsk so slipping his}
\stick{wrist out of the loop of a black ribban,}
\stick{to which a short scymetar was hung: He}
\stick{put his hand into his pocket, and with}
\stick{great courtesy touching the forepart of}
\stick{his cap with his left-hand, as he ex-}
\stick{tended his right\tsk he put a florin into}
the centinel’s hand, and passed on.

It grieves me, said the centinel, speak-\break
\stick{ing to a little dwarfish bandy-leg’d drum-}
\stick{mer, that so courteous a soul should have}
\stick{lost his scabbard\tsk he cannot travel with-}
\stick{out one to his scymetar, and will not be}
\stick{able to get a scabbard to fit it in all}
\natch{\i{Strasburg}.\tsk}

%------
\stick{\i{ciens,\tsk seque comiter inclinans \tsk hoc more}}
\stick{\i{gesto, nudam acinacem elevans, mulo lentò}}
\stick{\i{progrediente, ut nasum tueri possim}.\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Non immerito, benigne peregrine, re-}}
\stick{\i{spondit miles}.\hfill}
\stick{\hfill}
\stick{\indent\i{Nihili æstimo, ait ille tympanista, e per-}}
\stick{\i{gamenâ factitius est}.\hfill}
\stick{\hfill}
\stick{\hfill}
\stck{\indent\i{Prout christianus sum, inquit miles, nasus}}
\stick{\i{ille, ni sexties major sit, meo esset con-}}
\stick{\i{formis}.\hfill}
\stick{\hfill}
\stick{\indent\i{Crepitare audivi ait tympanista}.\hfill}
\patch{‘ \i{Me}-}


\noindent
\stick{\i{Strasburg.}\tsh I never had one, replied}
\stick{the stranger, looking back to the centi-}
\stick{nel, and putting his hand up to his cap}
\stick{as he spoke\tsh I carry it, continued he}
\stck{thus\tsh holding up his naked scymetar, his}
\stick{mule moving on slowly all the time, on}
purpose to defend my nose.

It is well worth it, gentle stranger,\break 
replied the centinel.

\tsk ’Tis not worth a single stiver, said\break
\stick{the bandy-leg’d drummer\tsk ’tis a nose of}
parchment.

As I am a true catholic\tsk except that\break
\stick{it is six times as big\tsk ’tis a nose, said}
the centinel, like my own.

\tsk I heard it crackle, said the drum-\break 
mer.\patch{By}

%----------------
\stick{\indent\i{Mehercule! sanguinem emisit, respondit}}
\stick{\i{miles}.\hfill}
\stick{\hfill}
\stick{\indent\i{Miseret me, inquit tympanista, qui non}}
\stick{\i{ambo titigimus!\sic}\hfill}
\stick{\hfill}
\stick{\indent\i{Eodem temporis puncto, quo hæc res ar}-}
\stick{\i{gumentata fuit inter militem et tympani}-}
\stick{\i{stam, disceptabatur ibidem tubicine \et\ ux}-}
\stick{\i{ore suâ, qui tunc accesserunt, et peregrino}}
\stick{\i{prætereunte, restiterunt.}\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Quantus nasus! æque longus est, ait}}
\stick{\i{tubicina, ac tuba}.\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Et ex eodem metallo, ait tubicen, velut}}
\stick{\i{sternutamento audias}.\hfill}

\patch{\i{Tantum}}

By dunder, said the centinel, I saw it bleed.

What a pity, cried the bandy-legg’d\break
drummer, we did not both touch it!

At the very time that this dispute was\break
\stick{maintaining by the centinel and the}
\stick{drummer\tsk was the same point debating}
\stick{betwixt a trumpeter and a trumpeter’s}
\stick{wife, who were just then coming up,}
\stick{and had stopped to see the stranger pass}
by.

\i{Benedicity!}\tsh What a nose! ’tis as\break
\stick{long, said the trumpeter’s wife, as a}
trumpet.

And of the same mettle, said the\break
\stick{trumpeter, as you hear by its sneez-}
ing.

\patch{\tsk ’Tis}
%-----------------------

\stick{\indent\i{Tantum abest, respondit illa, quod fistu-}}
\stick{\i{lam dulcedine vincit}.\hfill}
\stick{\indent\i{Æneus est, ait tubicen}.\hfill}
\stick{\hfill}
\stick{\indent\i{Nequaquam, respondit uxor}.\hfill}
\stick{\hfill}
\stick{\indent\i{Rursum affirmo, ait tubicen, quod æneus}}
\stick{\i{est}.\hfill}
\stick{\hfill}
\stick{\indent\i{Rem penitus explorabo; prius, enim}}
\stick{\i{digito tangam, ait uxor, quam dormi-}}
\stick{\i{vero}.\hfill}
\stick{\hfill}
\stick{\indent\i{Mulus peregrini, gradu lento progressus}}
\stick{\i{est, ut unumquodque verbum controversiæ,}}
\stick{\i{non tantum inter militem et tympanistam,}}
\stick{\i{verum etiam inter tubicinem et uxorum ejus,}}
\stick{\i{audiret}.\hfill}
\stick{\hfill}\etp
\stick{\indent\i{Nequaquam, ait ille, in muli collum}}
\stick{\itshape fræna demittens, \et\ manibus ambabus in}
\natch{\i{pectus}}

\tsk ’Tis as soft as a flute, said she.

\tsk ’Tis brass, said the trumpeter.

\tsk ’Tis a pudding’s end\tsk said his wife.

I tell thee again, said the trumpeter,\break
’tis a brazen nose,

I’ll know the bottom of it, said the\break
\stick{trumpeter’s wife, for I will touch it with}
my finger before I sleep.

The stranger’s mule moved on at so\break
\stick{slow a rate, that he heard every word of}
\stick{the dispute, not only betwixt the centinel}
\stick{and the drummer; but betwixt the trum-}
peter and trumpeter’s wife.

No! said he, dropping his reins upon\break
\stick{his mule’s neck, and laying both his}
\natch{hands}
%------------------------------
\stick{\i{pectus positis, (mulo lentè progrediente)}}
\stick{\i{nequaquam ait ille, respiciens, non necesse}}
\stick{\i{est ut res isthæc dilucidata foret. Minime}}
\stick{\i{gentium! meus nasus nunquam tangetur,}}
\stick{\i{dum spiritus hos reget artus\tsk ad quid agen-}}
\stick{\i{dum? ait uxor burgomagistri}.\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Peregrinus illi non respondit. Votum}}
\stick{\i{faciebat tunc temporis sancto Nicolao, quo}}
\stick{\i{facto, sinum dextrum inserens, e quâ negli-}}
\stick{\i{genter pependit acinaces, lento gradu pro-}}
\stick{\i{cessit per plateam Argentorati latam quæ}}
\stick{\i{ad diversorium templo ex adversum ducit}.\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\natch{\i{Peregrinus}}
%------------------------------\break
\noindent
\stick{hands upon his breast, the one over the}
\stick{other in a saint-like position (his mule}
\stick{going on easily all the time) No!  said}
\stick{he, looking up,\tsk I am not such a deb-}
\stick{tor to the world\tsk slandered and disap-}
\stick{pointed as I have been\tsk\!\tsk as to give it}
\stick{that conviction\tsk no! said he, my nose}
\stck{shall never be touched whilst heaven gives}
\stick{me strength\tsk To do what? said a bur-}
gomaster’s wife.

The stranger took no notice of the\break
\stick{burgomaster’s wife\tsh he was making a}
\stick{vow to saint \i{Nicolas}; which done, hav-}
\stick{ing uncrossed his arms with the same so-}
\stick{lemnity with which he crossed them, he}
\stick{took up the reins of his bridle with his}
\stick{left-hand, and putting his right-hand in-}
\stick{to his bosom, with the scymetar hanging}
\stick{loosely to the wrist of it, he rode on as}
\stick{slowly as one foot of the mule could fol-}
\natch{low}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Peregrinus mulo descendens stabulo in-}}
\stick{\i{cludi, \et\ manticam inferri jussit: quâ aper-}}
\stick{\i{tâ et coccineis sericis femoralibus extractis}}
\stick{\i{cum argenteo laciniato} \g{Περιζοματὶ}, \i{his sese}}
\stick{\i{induit, statimque, acinaci in manu, ad}}
\stick{\i{forum deambulavit}.\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Quod ubi peregrinus esset ingressus, ux-}}
\stick{\i{orem tubicinis obviam euntem aspicit; illico}}
\stick{\i{cursum flectit, metuens ne nasus suus explo-}}
\stick{\i{raretur, atque ad diversorium regressus est}}
\stick{\i{\tsk exuit se vestibus; braccas coccineas se}-}
\natch{\i{ricas}}
%------------------------
\noindent
\stick{low another, thro’ the principal streets of}
\stick{\i{Strasburg}, till chance brought him to the}
\stck{great inn in the market-place over-against}
the church.

The moment the stranger alighted, he\break
\stick{ordered his mule to be led into the stable,}
\stick{and his cloak-bag to be brought in;}
\stick{then opening, and taking out of it, his}
\stick{crimson-sattin breeches, with a silver-}
\stick{fringed\tsk (appendage to them, which I}
\stick{dare not translate)\tsk he put his breeches,}
\stick{with his fringed cod-piece on, and forth-}
\stick{with with his short scymetar in his hand,}
walked out to the grand parade.

\tight{The stranger had just taken three turns}\break
\stick{upon the parade, when he perceived the}
\stick{trumpeter’s wife at the opposite side of}
\stick{it\tsk so turning short, in pain lest his nose}
\stick{should be attempted, he instantly went}
\natch{back}
%----------
\stick{\i{ricas manticæ imposuit mulumque educi}}
\stick{\i{jussit}.\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Francofurtum proficiscor, ait ille, et}}
\stick{\i{Argentoratum quatuor abhinc hebdomadis }}
\stick{\i{revertar}.\hfill}
\stick{\hfill}
\stick{\indent\i{Bene curasti hoc jumentam (ait) muli}}
\stick{\i{faciem manu demulcens\tsk me, manticam-}}
\stick{\i{que meam, plus sexcentis mille passibus por-}}
\stick{\i{tavit}.\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\hfill} \etp
\stick{\i{\indent Longa via est! respondet hospes, nisi}}
\stick{\i{plurimum esset negoti.\tsk Enimvero, ait}}
\stick{\i{peregrinus a nasorum promontorio redij,}}
\stk{\i{et nasum speciosissimum, egregiosissimumque}}
\natch{\i{quem}}
%---------------
\stick{back to his inn\tsh undressed himself,}
\stick{packed up his crimson-sattin breeches,}
\stick{\etc in his cloak-bag, and called for his}
mule.

I am going forwards, said the stranger,\break
\stick{for \i{Franckfort}\tsk\!\tsk and shall be back at}
\i{Strasburg} this day month.

I hope, continued the stranger, stro-\break
\stick{king down the face of his mule with his}
\stick{left-hand as he was going to mount it,}
\stick{that you have been kind to this faithful}
\stick{slave of mine\tsk it has carried me and}
\stick{my cloak-bag, continued he, tapping the}
mule’s back, above six hundred leagues.

\tsk ’Tis a long journey, Sir, replied the\break
\stick{master of the inn\tsk\!\tsk unless a man has}
\stick{great business.\tsk Tut!\@ tut!\@ said the stran-}
\stick{ger, I have been at the promontory of}
\natch{Noses;}
%--------
\stick{\i{quem unquam quisquam sortitus est, acqui}-}
\stick{\i{sivi?}\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Dum peregrinus hanc miram rationem,}}
\stick{\i{de seipso reddit, hospes et uxor ejus, oculis}}
\stick{\i{intentis, peregrini nasum contemplantur\tsk}}
\stick{\i{Per sanctos, sanctasque omnes, ait hospitis}}
\stick{\i{uxor, nasis duodecim maximis, in toto Ar-}}
\stick{\i{gentorato major est!\tsk estne ait illa mariti}}
\stick{\i{in aurem insusurrans, nonne est nasus præ-}}
\stick{\i{grandis?}\hfill}
\stick{\hfill}
\stick{\hfill}
\stick{\indent\i{Dolus inest, anime mi, ait hospes\tsk nasus}}
\stick{\i{est falsus}.\tsk\hfill}

\medskip

\stick{\indent\i{Verus est, respondit uxor.\tsk}\hfill}

\medskip

\stick{\indent\i{Ex abiete factus est, ait ille, terebinth-}}
\stick{\i{inum olet}\tsh\hfill}
\natch{\i{Carbunculus}}
\stick{Noses; and have got me one of the}
\stick{goodliest and jolliest, thank heaven, that}
ever fell to a single man’s lot.

Whilst the stranger was giving this\break
\stick{odd account of himself, the master of}
\stick{the inn and his wife kept both their}
\stick{eyes fixed full upon the stranger’s nose\tsk}
\stick{By saint \i{Radagunda}, said the inn-keeper’s}
\stick{wife to herself, there is more of it than}
\stick{in any dozen of the largest noses put to-}
\stick{gether in all \i{Strasburg!}\@ is it not, said}
\stick{she, whispering her husband in his ear,}
is it not a noble nose?

’Tis an imposture, my dear, said the\break
master of the inn\tsk ’tis a false nose.\tsk

’Tis a true nose, said his wife.\tsk

’Tis made of fir-tree, said he,\tsk I smell\break
the turpentine.\tsk\patch{’Tis\sic}
%-----------
\bgroup\itshape
Carbunculus inest, ait uxor.

Mortuus est nasus, respondit hospes.

\vskip .5\baselineskip

Vivus est, ait illa,\tsk\!\tsk \et\ si ipsa vivam\break
tangam.

\vskip .5\baselineskip

Votum feci sancto Nicolao, ait peregrinus,\break
\stck{nasum meum intactum fore usque ad\tsk Quod-}
nam tempus? illico respondit illa.

\vskip 1.5\baselineskip

Minime tangetur, inquit ille (manibus in\break
\stick{pectus compositis) usque ad illam horam\tsk}
\stick{Quam horam? ait illa.\tsk Nullam, respondit}
\stick{peregrinus, donec perveneo,\sic ad\tsk Quem lo-}
\stick{cum,\tsk obsecro? ait illa\tsh Peregrinus nil}
respondens mulo conscenso discessit.
\egroup\vfill\eject

There’s a pimple on it, said she.

’Tis a dead nose, replied the inn-\break
keeper.

’Tis a live nose, and if I am alive my-\break
\stick{self, said the inn-keeper’s wife, I will}
touch it.

I have made a vow to saint \i{Nicolas}\break
\stick{this day, said the stranger, that my nose}
\stick{shall not be touched till\tsk Here the stran-}
\stick{ger, suspending his voice, looked up\tsh}
Till when?\@ said she hastily.

It never shall be touched, said he,\break
\stick{clasping his hands and bringing them}
\stick{close to his breast, till that hour.\tsh}
\stick{What hour? cried the inn keeper’s wife.}
\stick{\tsh Never!\tsk never! said the stranger,}
\stick{never tell\sic\ I am got\tsk For heaven sake}
\stick{into what place? said she.\tsk The stranger}
rode away without saying a word.\patch{The}


The stranger had not got half a league\break
\stick{on his way towards \i{Frankfort}, before all}
\stick{the city of \i{Strasburg} was in an uproar}
\stick{about his nose. The \i{Compline}-bells were}
\stick{just ringing to call the \i{Strasburgers} to}
\stick{their devotions, and shut up the duties}
\stick{of the day in prayer:\tsk\!\tsk no soul in all}
\stick{\i{Strasburg} heard ’em\tsk the city was like}
\stick{a swarm of bees\tsk\!\tsk men, women, and}
\stick{children (the \i{Compline}-bells tinkling all}
\stick{the time) flying here and there\tsk in at}
\stick{one door, out at another\tsk this way and}
\stick{that way\tsk long ways and cross ways\tsk}
\stick{up one street, down another street\tsk in at}
\stick{this ally, out at that\tsk did you see}
\stick{it? did you see it? did you see it? O!}
\stick{did you see it?\tsk who saw it? who did}
see it? for mercy’s sake, who saw it?

Alack o’day! I was at vespers!\tsk I\break
\stick{was washing, I was starching, I was}
\natch{scouring,}
\stick{scouring, I was quilting\tsk God help me!}
\stick{I never saw it\tsk I never touch’d it!\tsh}
\stick{would I had been a centinel, a bandy-}
\stick{leg’d drummer, a trumpeter, a trumpe-}
\stick{ter’s wife, was the general cry and la-}
\stick{mentation in every street and corner of}
\i{Strasburg}.

Whilst all this confusion and disorder\break
\stick{triumphed throughout the great city of}
\stick{\i{Strasburg}, was the courteous stranger go-}
\stick{ing on as gently upon his mule in his way}
\stick{to \i{Frankfort}, as if he had had no concern}
\stick{at all in the affair\tsk talking all the way}
\stick{he rode in broken sentences, sometimes}
\stick{to his mule\tsk sometimes to himself\tsh}
sometimes to his Julia.

O Julia, my lovely Julia!\tsk nay I cannot\break
\stick{stop to let thee bite that thistle\tsk that ever}
\stck{the suspected tongue of a rival should have}
\natch{robbed}
\stick{robbed me of enjoyment when I was}
upon the point of tasting it.\tsk

\tsk Pugh!\tsk ’tis nothing but a thistle\tsk\break
\stick{never mind it\tsk thou shalt have a better}
supper at night.\tsk

\tsh Banish’d from my country\tsk my\break
friends\tsk from thee.\tsk

Poor devil, thou’rt sadly tired with\break
\stick{thy journey!\tsk come\tsk get on a little}
\stick{faster\tsk there’s nothing in my cloak-bag}
\stick{but two shirts\tsk a crimson-sattin pair of}
breeches, and a fringed\tsk Dear Julia!

\tsh But why to \i{Frankfort?}\tsk is it that\break
\stick{there is a hand unfelt, which secretly is}
\stick{conducting me through these meanders}
and unsuspected tracts!\tsk
\patch{\tsk Stumbling!}

\tsk Stumbling! by saint \i{Nicolas!} every\break
\stick{step\tsk\!\tsk why at this rate we shall be all}
night in getting in\tsh

\tsk To happiness\tsk or am I to be the\break
\stick{sport of fortune and slander\tsk destined to}
\stick{be driven forth unconvicted\tsk unheard\tsk}
\stick{untouched\tsk if so, why did I not stay}
\stick{at \i{Strasburg}, where justice\tsk\!\tsk but I had}
\stick{sworn!\tsk Come, thou shalt drink\tsk to \i{St}.}
\stick{\i{Nicolas}\tsk O Julia!\tsk\!\tsk What dost thou}
\stick{prick up thy ears at?\tsh ’tis nothing but a}
man, \etc\tsh

The stranger rode on communing in\break
\stick{this manner with his mule and Julia\tsk}
\stick{till he arrived at his inn, where, as soon}
\stick{as he arrived, he alighted\tsk saw his mule,}
\stick{as he had promised it, taken good care}
\stick{of\tsk\!\tsk took off his cloak bag, with his}
\stick{crimson-sattin breeches, \etc in it\tsh}
\natch{called}
\stick{called for an omelet to his supper, went}
\stick{to his bed about twelve o’clock, and in}
five minutes fell fast asleep.

It was about the same hour when\break
\stick{the tumult in \i{Strasburg} being abated}
\stick{for that night,\tsh the \i{Strasburgers} had}
\stick{all got quietly into their beds\tsk but not}
\stick{like the stranger, for the rest either of}
\stick{their minds or bodies; queen \i{Mab}, like}
\stick{an elf as she was, had taken the stranger’s}
\stick{nose, and without reduction of its bulk,}
\stick{had that night been at the pains of slitting}
\stick{and dividing it into as many noses of}
\stick{different cuts and fashions, as there were}
\stick{heads in \i{Strasburg} to hold them. The}
\stick{abbess of \i{Quedlingberg}, who, with the}
\stick{four great dignitaries of her chapter,}
\stick{the prioress, the deaness, the sub-chan-}
\stick{tress, and senior canoness, had that week}
\tight{come to \i{Strasburg} to consult the university}
\natch{upon}
\tight{upon a case of conscience relating to their}
placket holes\tsk was ill all the night.

\vskip -0.5\parskip

The courteous stranger’s nose had got\break
\stick{perched upon the top of the pineal gland}
\stick{of her brain, and made such rousing}
\stick{work in the fancies of the four great}
\stick{dignitaries of her chapter, they could not}
\stick{get a wink of sleep the whole night thro’}
\stick{for it\tsh there was no keeping a limb}
\stick{still amongst them\tsh in short, they got}
up like so many ghosts.

\vskip -0.5\parskip

The penitentiaries of the third order\break
\stick{of saint \i{Francis}\tsh the nuns of mount}
\stick{\i{Calvary}\tsk the \i{Præmonstratenses}\tsk\!\tsk the}
\stick{\i{Clunienses} \fnast  \tsk the \i{Carthusians}, and all the}
\stick{severer orders of nuns, who lay that}
\stick{night in blankets or hair-cloth, were still}
\stick{\hfill in}
\stick{\footnotesize \quad \fnast\  \i{Hafen Slawkenbergius} means the Benedictine}
\vskip -4pt
\stick{\footnotesize nuns of \i{Cluny}, founded in the year 940, by \i{Odo},}
\vskip -4pt
\stick{\footnotesize abbé de \i{Cluny.}\hfill}\eject
\noindent
\stick{in a worse condition than the abbess of}
\stick{\i{Quedlingberg}\tsk by tumbling and tossing,}
\stick{and tossing and tumbling from one side}
\stick{of their beds to the other the whole}
\stick{night long\tsk the several sisterhoods had}
\stick{scratch’d and mawl’d themselves all to}
\stick{death\tsk they got out of their beds almost}
\stick{flead alive\tsk every body thought saint}
\stick{\i{Antony} had visited them for probation}
\stick{with his fire\tsk\!\tsk they had never once, in}
\stick{short, shut their eyes the whole night}
long from vespers to matins.

The nuns of saint \i{Ursula} acted the\break
\stick{wisest\tsk they never attempted to go to}
bed at all.

The dean of \i{Strasburg}, the prebenda-\break
\stick{ries, the capitulars and domiciliars (ca-}
\stck{pitularly assembled in the morning to con-}
\stick{sider the case of butter’d buns) all wished}
\natch{they}
\stick{they had followed the nuns of saint}
\stick{\i{Ursula}’s example.\tsk\!\tsk In the hurry and}
\stick{confusion every thing had been in the}
\stick{night before, the bakers had all forgot}
\stick{to lay their leaven\tsk there were no but-}
\stick{ter’d buns to be had for breakfast in all}
\stick{\i{Strasburg}\tsk the whole close of the cathe-}
\stick{dral was in one eternal commotion\tsk such}
\stick{a cause of restlessness and disquietude,}
\stick{and such a zealous inquiry into the}
\stick{cause of the restlessness, had never hap-}
\stick{pened in \i{Strasburg}, since \i{Martin Luther},}
\stick{with his doctrines, had turned the city}
up-side down.

If the stranger’s nose took this liberty\break
\stick{of thrusting itself thus into the dishes \fnast\ of}
\stick{\hfill religious}\etp

\kern -0.5\baselineskip

\bgroup\fontsize{8}{10}\selectfont\noindent
\stick{\indent \fnast\ Mr.\@ \i{Shandy}’s compliments to orators\tsk is very}
\stick{sensible that \i{Slawkenbergius} has here changed his}
\stick{metaphor\tsk which he is very guilty of;\tsk that as a}
\stick{translator, Mr.\@ \i{Shandy} has all along done what he}
\stick{could to make him stick to it\tsk but that here ’twas}
impossible.\par\egroup
\newpage\noindent\etp
\stick{religious orders, \etc what a carnival did}
\stick{his nose make of it, in those of the laity!}
\stick{\tsk ’tis more than my pen, worn to the}
\stick{stump as it is, has power to describe; tho’}
\stick{I acknowledge, (\i{cries} Slawkenbergius,}
\stick{\i{with more gaiety of thought than I could}}
\stick{\i{have expected from him)}\sic that there is many}
\stick{a good simile now subsisting in the world}
\stick{which might give my countrymen some}
\stick{idea of it; but at the close of such a folio}
\stick{as this, wrote for their sakes, and in}
\stick{which I have spent the greatest part of}
\stick{my life\tsh tho’ I own to them the simile is}
\stick{in being, yet would it not be unreason-}
\stick{able in them to expect I should have ei-}
\stick{ther time or inclination to search for it?}
\stick{Let it suffice to say, that the riot and dis-}
\stck{order it occasioned in the \i{Strasburgers} fan-}
\stick{tacies was so general\tsk such an overpow-}
\stick{ering mastership had it got of all the}
\stick{faculties of the \i{Strasburgers} minds\tsk so}
\natch{many}
\stick{many strange things, with equal confi-}
\stick{\tighter{dence on all sides, and with equal eloquence}}
\stick{in all places, were spoken and sworn to}
\stick{concerning it, that turned the whole}
\stick{stream of all discourse and wonder to-}
\stick{wards it\tsk every soul, good and bad\tsk}
\stick{rich and poor\tsk learned and unlearned\tsk}
\stick{doctor and student\tsk mistress and maid\tsk}
\stick{gentle and simple\tsk nun’s flesh and wo-}
\stick{man’s flesh in \i{Strasburg} spent their}
\stick{time in hearing tidings about it\tsk every}
\stick{eye in \i{Strasburg} languished to see it\tsk}
\stick{every finger\tsk every thumb in \i{Strasburg}}
burned to touch it.

Now what might add, if any thing\break
\stick{may be thought necessary to add to so}
\stick{vehement a desire\tsk was this, that the}
\stick{centinel, the bandy-legg’d drummer, the}
\stick{trumpeter, the trumpeter’s wife, the}
\stick{burgo-master’s widow, the master of the}
\natch{inn,} \etp
\stick{inn, and the master of the inn’s wife,}
\stick{how widely soever they all differed every}
\stick{one from another in their testimonies}
\stick{and description of the stranger’s nose\tsk}
\stick{they all agreed together in two points\tsk}
\stick{namely, that he was gone to \i{Frankfort},}
\stick{and would not return to \i{Strasburg} till}
\stick{that day month; and secondly, whether}
\stick{his nose was true or false, that the stran-}
\stick{ger himself was one of the most perfect}
\stck{paragons of beauty\raise0.4pt\hbox{-}-the finest made man!}
\stick{\tsk the most genteel!\tsk the most generous}
\stick{of his purse\tsk the most courteous in his}
\stick{carriage that had ever entered the gates}
\stick{of \i{Strasburg}\tsk that as he rode, with his}
\stick{scymetar slung loosely to his wrist, thro’}
\stck{the streets\tsk and walked with his crimson-}
\stick{sattin breeches across the parade\tsk ’twas}
\stick{with so sweet an air of careless modesty,}
\stick{and so manly withal\tsk as would have put}
\stick{the heart in jeopardy (had his nose not}
\natch{stood}
\stick{stood in his way) of every virgin who}
had cast her eyes upon him.

I call not upon that heart which is a\break
\stick{stranger to the throbs and yearnings of}
\stick{curiosity, so excited to justify the abbess}
\stick{of \i{Quedlingberg}, the prioress, the deaness,}
\stck{and subchantress for sending at noon-day}
\stick{for the trumpeter’s wife: she went}
\stick{through the streets of \i{Strasburg} with}
\stick{her husband’s trumpet in her hand;\tsk}
\stick{the best apparatus the straitness of the}
\stick{time would allow her, for the illustration}
\stick{of her theory\tsk she staid no longer than}
three days.

The centinel and the bandy-legg’d\break
\stick{drummer!\tsk nothing on this side of old}
\stick{\i{Athens} could equal them! they read}
\stick{their lectures under the city gates to}
\stick{comers and goers, with all the pomp}
\natch{of}
\stick{of a \i{Chrysippus} and a \i{Crantor} in their}
porticos.

The master of the inn, with his ostler\break
\stick{on his left-hand, read his also in the same}
\stck{stile,\tsk under the portico or gateway of his}
\stck{stable-yard\tsk his wife, hers more privately}
\stk{in a back room: all flocked to their lectures;}
\stick{not promiscuously\tsk but to this or that,}
\stick{as is ever the way, as faith and credulity}
\stick{marshal’d them\tsk in a word, each \i{Stras}-}
\stick{\i{burger} came crouding for intelligence\tsk}
\stck{and every \i{Strasburger} had the intelligence}
he wanted.

Tis worth remarking, for the benefit\break
\stick{of all demonstrators in natural philoso-}
\stick{phy, \etc that as soon as the trumpeter’s}
\stick{wife had finished the abbess of \i{Quedlin}-}
\stick{\i{berg}’s private lecture, and had begun to}
\stick{read in public, which she did upon a}
\natch{stool}
\stick{stool in the middle of the great parade\tsh}
\stck{she incommoded the other demonstrators}
\stick{mainly, by gaining incontinently the}
\stick{most fashionable part of the city of \i{Stras}-}
\stick{\i{burg} for her auditory\tsk But when a de-}
\stick{monstrator in philosophy (cries \i{Slawken}-}
\stick{\i{bergius)}\sic has a \i{trumpet} for an apparatus,}
\stick{pray what rival in science can pretend to}
be heard besides him?

Whilst the unlearned, thro’ these con-\break
\stick{duits of intelligence, were all busied in}
\stick{getting down to the bottom of the well,}
\stick{where \s{Truth} keeps her little court\tsk}
\stick{were the learned in their way as busy in}
\stick{pumping her up thro’ the conduits of}
\stick{dialect induction\tsk they concerned them-}
\tight{selves not with facts\tsk they reasoned\tsk}

Not one profession had thrown more\break
\stick{light upon this subject than the faculty\tsk}
\natch{had}
\stick{had not all their disputes about it run into}
\stick{the affair of \i{Wens} and œdematous swel-}
\stick{lings, they could not keep clear of them}
\stck{for their bloods and souls\tsh the stranger’s}
\stick{nose had nothing to do either with wens}
or œdematous swellings.

It was demonstrated however very sa-\break
\stck{tisfactorily, that such a ponderous mass of}
\stick{heterogenous matter could not be con-}
\stick{gested and conglomerated to the nose,}
\stick{whilst the infant was \i{in Utero}, without}
\stick{destroying the statical balance of the}
\stick{fœtus, and throwing it plump upon its}
head nine months before the time.\tsh

\tsk The opponents granted the theory\tsk\break
they denied the cousequences.\sic

And if a suitable provision of veins,\break
\stick{arteries, \etc said they, was not laid in,}
\natch{for}
\stick{for the due nourishment of such a nose,}
\stick{in the very first stamina and rudiments}
\stick{of its formation  before it came into the}
\stick{world (bating the case of Wens) it could}
\stick{not regularly grow and be sustained af-}
terwards.

\tight{This was all answered by a dissertation}\break
\stick{upon nutriment, and the effect which}
\stick{nutriment had in extending the vessels,}
\stick{and in the increase and prolongation of}
\stick{the muscular parts to the greatest growth}
\stick{and expansion imaginable\tsk In the tri-}
\stick{umph of which theory, they went so far}
\stick{as to affirm, that there was no cause in}
\stick{nature, why a nose might not grow to}
the size of the man himself.

The respondents satisfied the world\break
\stick{this event could never happen to them so}
\stick{long as a man had but one stomach and}
\natch{one}
\stick{one pair of lungs\tsk For the stomach, said}
\stick{they, being the only organ destined for}
\stick{the reception of food, and turning it into}
\stick{chyle,\tsk and the lungs the only engine of}
\stck{sanguification\tsk it could possibly work off}
\stick{no more, than what the appetite brought}
\stick{it: or admitting the possibility of a man’s}
\stick{overloading his stomach, nature had set}
\stick{bounds however to his lungs\tsk the engine}
\stick{was of a determined size and strength,}
\stck{and could elaborate but a certain quantity}
\stick{in a given time\tsk that is, it could produce}
\stick{just as much blood as was sufficient for}
\stick{one single man, and no more; so that,}
\stick{if there was as much nose as man\tsk they}
\stick{proved a mortification must necessarily}
\stick{ensue; and forasmuch as there could not}
\stick{be a support for both, that the nose must}
\stick{either fall off from the man, or the man}
inevitably fall off from his nose.\patch{Nature}

Nature accommodates herself to these\break
\stick{emergencies, cried the  opponents\tsk else}
\stick{what do you say to the case of a whole}
\stick{stomach\tsk a whole pair of lungs, and but}
\stick{\i{half} a man, when both his legs have}
been unfortunately shot off?\tsk

He dies of a plethora, said they\tsk or\break
\stick{must spit blood, and in a fortnight or}
three weeks go off in a consumption\tsk

\tsk It happens otherways\tsk replied the\break
opponents.\tsh

It ought not, said they.

The more curious and intimate inqui-\break
\stick{rers after nature and her doings, though}
\stick{they went hand in hand a good way to-}
\stick{gether, yet they all divided about the}
\stick{nose at last, almost as much as the fa-}
culty itself.\patch{They}

They amicably laid it down, that there\break
\stick{was a just and geometrical arrangement}
\stick{and proportion of the several parts of the}
\stick{human frame to its several destinations,}
\stick{offices, and functions, which could not}
\stick{be transgressed but within certain limits}
\stick{\tsk that nature, though she sported\tsk she}
\stick{sported within a certain circle;\tsk and they}
could not agree about the diameter of it.

The logicians stuck much closer to the\break
\stick{point before them than any of the classes}
\stick{of the literati;\tsh they began and ended}
\stick{with the word nose; and had it not been}
\stick{for a \i{petitio principii}, which one of the}
\stick{ablest of them ran his head against in the}
\stick{beginning of the combat, the whole con-}
troversy had been settled at once.

A nose, argued the logician, cannot\break
\stick{bleed without blood\tsk and not only blood}
\natch{\tsk but}
\stick{\tsk but blood circulating in it to supply}
\stick{the phænomenon with a succession of}
\stick{drops\tsk (a stream being but a quicker}
\stick{succession of drops, that is included, said}
\stick{he)\tsk Now death, continued the logician,}
\stick{being nothing but the stagnation of the}
blood\tsk

I deny the definition\tsh Death is the\break
\stick{separation of the soul from the body,}
\stick{said his antagonist\tsk Then we don’t agree}
\stick{about our weapon, said the logician\tsk}
\stick{Then there is an end of the dispute, re-}
plied the antagonist.

The civilians were still more concise;\break
\stick{what they offered being more in the na-}
ture of a decree\tsk than a dispute.

\tsk Such a monstrous nose, said they,\break
\stck{had it been a true nose, could not possibly}
\stick{have been suffered in civil society\tsk and if}
\natch{false}
\stick{false\tsk to impose upon society with such}
\stick{false signs and tokens, was a still greater}
\stick{violation of its rights, and must have}
had still less mercy shewn it.

The only objection to this was, that\break
\stick{if it proved any thing, it proved the}
\tight{stranger’s nose was neither true nor false.}

This left room for the controversy to\break
\stick{go on. It was maintained by the advo-}
\stick{cates of the ecclesiastic court, that there}
\stick{was nothing to inhibit a decree, since}
\stick{the stranger \i{ex mero motu} had confessed}
\stick{he had been at the Promontory of Noses,}
\stick{and had got one of the goodliest, \etc\etc}
\stick{\tsk To this it was answered, it was impossi-}
\stick{ble there should be such a place as the}
\stick{Promontory of Noses, and the learned be}
\stick{ignorant where it lay. The commissary}
\stick{of the bishop of \i{Strasburg} undertook the}
\natch{advocates,}
\stick{advocates, explained this matter in a}
\stck{treatise upon proverbial phrases, shewing}
\stick{them, that the Promontory of Noses was}
\stck{a mere allegoric expression, importing no}
\stick{more than that nature had given him a}
\stick{long nose: in proof of which, with great}
\stick{learning, he cited the underwritten au-}
\stick{thorities \fnast, which had decided the point}
\rightline{incon-}

\vfill\noindent\etp
\begingroup\fontsize{8}{10}\selectfont\frenchspacing
\stick{\indent\fnast\enspace Nonnulli ex nostratibus eadem loquendi for-}
\stick{mulâ utun.\enspace Quinimo et Logistæ \& Canonistæ\tsk}
\stick{Vid. Parce Bar e Jas in d. L. Provincial.Constitut.}
\stick{de conjec. vid. Vol. Lib.~4. Titul.~1. N.~7. quà}
\stick{etiam in re conspir. Om. de Promontorio Nas.}
\stick{Tichmak. ff. d. tit.~3. fol.~189. passim. Vid. Glos.}
\stick{de contrahend. empt.\etc nec non J. Scrudr. in cap.}
\stick{§. refut. ff. per totum. cum his cons. Rever. J.}
\stick{Tubal, Sentent. \& Prov. cap.~9. ff.~11, 12. obiter.}
\stick{V. et Librum, cui Tit. de Terris \& Phras. Belg.}
\stick{ad finem cum, Comment. N. Bardy Belg. Vid.}
\stick{Scrip. Argentotarens. de Antiq. Ecc. in Episc. Ar-}
\stick{chiv. fid. coll. per Von Jacobum Koinshoven Fo-}
\stick{lio Argent. 1583, præcip. ad finem.  Quibus add.}
\stick{Rebuff in L. obvenire de Signif. Nom. ff. fol. \&}
\stick{de Jure, Gent. \& Civil. de protib. aliena feud. per}
\stick{federa, test. Joha. Luxius in prolegom. quem velim}
videas, de Analy. Cap. 1, 2, 3. Vid. Idea.\par\endgroup

\newpage\noindent
\stick{incontestably, had it not appeared that a}
\stck{dispute about some franchises of dean and}
\stick{chapter-lands had been determined by it}
nineteen years before.

It happened\tsk I must not say unluckily\break
\stick{for Truth, because they were giving her}
\stick{a lift another way in so doing; that the}
\stick{two universities of \i{Strasburg}\tsk the \i{Luthe}-}
\stick{\i{ran}, founded in the year 1538 by \i{Jacobus}}
\stick{\i{Sturmius}, counsellor of the senate,\tsk and}
\stick{the \i{Popish}, founded by \i{Leopold}, arch-duke}
\stick{of \i{Austria}, were, during all this time,}
\stick{employing the whole depth of their}
\stick{knowledge (except just what the affair}
\stick{of the abbess of \i{Quedlinberg}’s placket-}
\stick{\stck{holes required)\tsk in determining the point}}
of \i{Martin Luther}’s damnation.

The \i{Popish} doctors had undertaken to\break
\stick{demonstrate \i{a priori}; that from the ne-}
\natch{cessary}
\stick{cessary influence of the planets on the}
\stick{twenty-second day of \i{October} 1483\tsh}
\stick{when the moon was in the twelfth house}
\stick{\tsk \i{Jupiter}, \i{Mars}, and \i{Venus} in the third,}
\stick{the \i{Sun}, \i{Saturn}, and \i{Mercury}, all got}
\stick{together in the fourth\tsk that he must in}
\stick{course, and unavoidably be a damn’d}
\stick{man\tsk and that his doctrines, by a direct}
corollary, must be damn’d doctrines too.

\tight{By inspection into his horoscope, where}
\stick{five planets were in coition all at once}
\stick{with scorpio \raise1pt\hbox{\ast} (in reading this my father}
\rightline{would}
\vfill\etp
\noindent
\begingroup\fontsize{8}{10}\selectfont
\stick{\indent\fnast\enspace Haec mira, satisque horrenda. Planetarum}
\stick{coitio sub Scarpio\sic Asterismo in nonâ cœli statione,}
\stick{quam Arabes religioni deputabant efficit \i{Martinum}}
\stick{\i{Lutherum} sacrilegum hereticum, christianæ reli-}
\stick{gionis hostem acerrimum atque prophanum, ex}
\stick{horoscopi directione ad Martis coitum, religiosis-}
\stick{simus obiit, ejus Anima seclestissima\sic\ ad infernos}
\stick{navigavit\tsk ab Alecto, Tisiphone et Magera fla-}
\stick{gellis igneis cruciata pereniter.\hfill}
\stick{\indent\tsk Lucas Gauricus in Tractatu astrologico de}
\stick{præteritis multorum hominum accidentibus per}
genituras examinatis.\par\endgroup

\newpage\noindent
\stick{would always shake his head) in the}
\stick{ninth house which the \i{Arabians} allotted}
\stick{to religion\tsk it appeared that \i{Martin Lu}-}
\stick{\i{ther} did not care one stiver about the}
\stick{matter\tsk and that from the horoscope}
\stick{directed to the conjunction of \i{Mars}\tsk}
\stick{they made it plain likewise he must die}
\stick{\lower.3ex\hbox{c}ursing\sic and blaspheming\tsk with the blast}
\stick{of which his soul (being steep’d in guilt)}
\stick{sailed before the wind, into the lake of}
hell fire.

The little objection of the \i{Lutheran}\break
\stick{doctors to this, was, that it must certainly}
\stick{be the soul of another man, born \i{Oct}.}
\stick{22, \quad\sic 83, which was forced to sail down}
\stick{\tight{before the wind in that manner\tsk inasmuch}}
\stick{as it appeared from the register of \i{Islaben}}
\stick{in the county of \i{Mansfelt}, that \i{Luther}}
\stick{was not born in the year 1483, but in}
\stick{84; and not on the 22d day of \i{October},}
\natch{but}
\stick{but on the 10th of \i{November}, the eve of}
\stick{\i{Martinmas}-day, from whence he had the}
name of \i{Martin}.

[\tsh I must break off my translation for\break
\stick{a moment; for if I did not, I know I}
\stick{should no more be able to shut my eyes}
\stick{in bed, than the abbess of \i{Quedlinberg}\tsk}
\stick{It is to tell the reader, that my father}
\stick{never read this passage of \i{Slawkenbergius}}
\stick{to my uncle \toby but with triumph\tsk not}
\stick{over my uncle \toby, for he never opposed}
him in it\tsk but over the whole world.

\tight{\tsk Now you see, brother \toby, he would}\break
\stick{say, looking up, \lqq that christian names}
\stick{\lqq are not such indifferent things;\,”\tsk}
\stick{had \i{Luther} here been called by any other}
\stick{name but \i{Martin}, he would have been}
\stick{damned to all eternity\tsk Not that I look}
\stick{upon \i{Martin}, he would add, as a good}
\natch{name}
\stick{name\tsk far from it\tsk ’tis something better}
\stick{than a neutral, and but a little\tsk yet little}
\stick{as it is, you see it was of some service to}
him.

My father knew the weakness of this\break
\stick{prop to his hypothesis, as well as the best}
\stick{logician could shew him\tsk yet so strange}
\stick{is the weakness of man at the same time,}
\stick{as it fell in his way, he could not for his}
\stick{life but make use of it; and it was cer-}
\stick{tainly for this reason, that though there}
\stck{are many stories in \i{Hafen Slawkenbergius}’s}
\stick{Decads full as entertaining as this I am}
\stick{translating, yet there is not one amongst}
\stick{them which my father read over with}
\stick{half the delight\tsk it flattered two of his}
\stk{strangest hypotheses together\tsk his \s{Names}}
\stick{and his \s{Noses}\tsk I will be bold to say,}
\stick{he might have read all the books in the}
\stick{\i{Alexandrian} library, had not fate taken}
\natch{other}
\stick{other care of them, and not have met}
\stick{with a book or a passage in one, which}
\stick{hit two such nails as these upon the head}
at one stroke.]

The two universities of \i{Strasburg} were\break
\stick{hard tugging at this affair of \i{Luther}’s}
\stick{navigation. The Protestant doctors had}
\stck{demonstrated, that he had not sailed right}
\stick{before the wind, as the Popish doctors}
\stick{had pretended; and as every one knew}
\stick{there was no sailing full in the teeth of}
\stick{it,\tsk they were going to settle, in case he}
\stick{had sailed, how many points he was off;}
\stick{whether \i{Martin} had doubled the cape,}
\stick{or had fallen upon a lee-shore; and no}
\stick{doubt, as it was an enquiry of much edi-}
\stick{fication, at least to those who understood}
\stk{this sort of \s{navigation}, they had gone on}
\stick{with it in spite of the size of the stranger’s}
\stick{nose, had not the size of the stranger’s}
\natch{nose}
\stick{nose drawn off the attention of the world}
\stick{from what they were about\tsk it was their}
business to follow.\tsh

\vskip -2pt

\tight{The abbess of \i{Quedlinberg} and her four}
\stick{dignitaries was no stop; for the enormity}
\stick{\tight{of the stranger’s nose running full as much}}
\stick{in their fancies as their case of conscience}
\stick{\tsk The affair of their placket holes kept}
\stick{cold\tsk In a word, the printers were or-}
\stick{dered to distribute their types\tsk all con-}
troversies dropp’d.

\vskip -2pt

’Twas a square cap with a silk tassel\break
\stick{upon the crown of it\tsk to a nut shell\tsk}
\stick{to have guessed on which side of the nose}
the two universities would split.

\vskip -2pt

’Tis above reason, cried the doctors on\break one side.

\vskip -2pt

’Tis below reason, cried the others.

\vskip -2pt

’Tis faith, cried the one.\patch{’Tis}

’Tis a fiddle-stick, said the other.

’Tis possible, cried the one.

’Tis impossible, said the other.

God’s power is infinite, cried the No-\break
sarians, he can do any thing.

He can do nothing, replied the Anti-\break
nosarians, which implies contradictions.

He can make matter think, said the\break
Nosarians.

\tighter{As certainly as you can make a velvet cap}\break
\stk{out of a sow’s ear, replied the Antinosarians.}

He cannot\sic make two and two five,\break
\stick{replied the Popish doctors.\tsh ’Tis false,}
said their opponents.\tsk

Infinite power is infinite power, said\break
\stick{the doctors who maintained the \i{reality}}
\natch{of}
\stick{of the nose.\tsh It extends only to all}
possible things, replied the \i{Lutherans}.

By God in heaven, cried the Popish\break
\stick{doctors, he can make a nose, if he thinks}
fit, as big as the steeple of \i{Strasburg}.

Now the steeple of \i{Strasburg} being the\break
\stick{biggest and the tallest church-steeple to}
\stick{be seen in the whole world, the Antino-}
\stick{sarians denied that a nose of 575 geome-}
\stick{trical feet in length could be worn, at}
\stick{least by a middle-siz’d man\tsk The Popish}
\stick{doctors swore it could\tsk The \i{Lutheran}}
doctors said No;\tsk it could not.

This at once started a new dispute,\break
\stick{which they pursued a great way upon the}
\stick{extent and limitation of the moral and}
\stick{natural attributes of God\tsk That contro-}
\stick{versy led them naturally into \i{Thomas}}
\natch{\i{Aquinas},}
\stick{\i{Aquinas}, and \i{Thomas Aquinas} to the}
devil.

\tight{The stranger’s nose was no more heard}\break
\stck{of in the dispute\tsk it just served as a frigate}
\stick{to launch them into the gulph of school-}
\stick{divinity,\tsk and then they all sailed before}
the wind.

Heat is in proportion to the want of\break
true knowledge.

The controversy about the attributes,\break
\stick{\etc instead of cooling, on the contrary}
\stick{had inflamed the \i{Strasburgers} imagina-}
\stick{tions to a most inordinate degree\tsk The}
\stick{less they understood of the matter, the}
\stick{greater was their wonder about it\tsk they}
\stick{were left in all the distresses of desire}
\stick{unsatisfied\tsk saw their doctors, the \i{Parch}-}
\stick{\i{mentarians}, the \i{Brassarians}, the \i{Turpen}-}
\stick{\i{tarians}, on one side\tsk the Popish doctors}
\natch{on}
\stick{on the other, like \i{Pantagruel} and his}
\stick{companions in quest of the oracle of the}
bottle, all embarked and out of sight.

\tsh The poor \i{Strasburgers} left upon\break
the beach!

\tsk What was to be done?\tsk No delay\break
\stick{\tsk the uproar increased\tsk every one in dis-}
order\tsk the city gates set open.\tsk

Unfortunate \i{Strasburgers!} was there\break
\stick{in the store-house of nature\tsk was there}
\stick{in the lumber-rooms of learning\tsk was}
\stick{there in the great arsenal of chance, one}
\stck{single engine left undrawn forth to torture}
\stick{your curiosities, and stretch your desires,}
\stick{which was not pointed by the hand of}
\stick{fate to play upon your hearts?\tsk I dip}
\stick{not my pen into my ink to excuse the}
\stck{surrender of yourselves\tsk ’tis to write your}
\natch{panegyrick.}
\stick{panegyrick. Shew me a city so macera-}
\stick{ted with expectation\tsh who neither eat,}
\stick{or drank, or slept, or prayed, or heark-}
\stick{ned to the calls either of religion or na-}
\stick{ture, for seven and twenty days together,}
who could have held out one day longer.

On the twenty-eighth the courteous\break
\stick{stranger had promised to return to \i{Stras-}}
\i{burg.}


Seven thousand coaches (\i{Slawkenber}-\break
\stick{\i{gius} must certainly have made some mis-}
\stick{take in his numeral characters) 7000}
\stick{coaches\tsk 15000 single horse chairs\tsh}
\stick{20000 waggons, crouded as full as they}
\stick{could all hold with senators, counsellors,}
\stick{syndicks\tsk beguines, widows, wives, vir-}
\stick{gins, canons, concubines, all in their}
\stick{coaches\tsk The abbess of \i{Quedlinberg}, with}
\stck{the prioress, the deaness and sub-chantress}
\natch{leading}
\stick{leading the procession in one coach, and}
\stick{the dean of \i{Strasburg}, with the four great}
\stick{dignitaries of his chapter on her left-}
\stck{hand\tsk the rest following higglety-pigglety}
\stick{as they could; some on horseback\tsh}
\stick{some on foot\tsk some led\tsk some driven\tsk}
\stick{some down the \i{Rhine}\tsk some this way\tsk}
\stick{some that\tsk all set out at sun-rise to meet}
the courteous stranger on the road.

\tight{Haste we now towards the catastrophe}\break
\stick{of my tale\tsk I say \i{Catastrophe} (cries \i{Slaw}-}
\stick{\i{kenbergius)}\sic inasmuch as a tale, with parts}
\stick{rightly disposed, not only rejoiceth \i{(gau}-}
\stick{\i{det)} in the \i{Catastrophe} and \i{Peripeitia} of}
\stick{a \s{Drama}, but rejoiceth moreover in all}
\stick{the essential and integrant parts of it\tsk}
\stick{it has its \i{Protasis}, \i{Epistasis},\sic \i{Catastasis},}
\stick{\lower.2ex\hbox{i}ts\sic \i{Catastrophe} or \i{Peripeitia} growing one}
\stick{out of the other in it, in the order \i{Aristotle}}
\stick{first planted them\tsh without which a tale}
\natch{had}
\stick{had better never be told at all, says}
\stick{\i{Slawkenbergius}, but be kept to a man’s}
self.

In all my ten tales, in all my ten de-\break
\stick{cads, have I, \i{Slawkenbergius}, tied down}
\stick{every tale of them as tightly to this rule,}
\stick{as I have done this of the stranger and}
his nose.

\tsk From his first parley with the centi-\break
\stick{nel, to his leaving the city of \i{Strasburg},}
\stick{after pulling off his crimson-sattin pair of}
\stick{breeches, is the \i{Protasis} or first entrance}
\stick{\tsk where the characters of the \i{Personæ}}
\stick{\i{Dramatis} are just touched in, and the}
subject slightly begun.

The \i{Epitasis}, wherein the action is\break
\stick{more fully entered upon and heightened,}
\stick{till it arrives at its state or height called}
\natch{the}
\stick{the \i{Catastasis}, and which usually takes}
\stick{up the 2d and 3d act, is included within}
\stick{that busy period of my tale, betwixt the}
\stick{first night’s uproar about the nose, to}
\stick{the conclusion of the trumpeter’s wife’s}
\stick{lectures upon it in the middle of the}
\stick{grand parade; and from the first em-}
\stick{barking of the learned in the dispute\tsk}
\stick{to the doctors finally sailing away, and}
\stick{leaving the \i{Strasburgers} upon the beach}
\stick{in distress, is the \i{Catastasis} or the ripen-}
\stick{ing of the incidents and passions for their}
bursting forth in the fifth act.

This commences with the setting out\break
\stick{of the \i{Strasburgers} in the \i{Frankfort} road,}
\stick{and terminates in unwinding the laby-}
\stick{rinth and bringing the hero out of a}
\stick{state of agitation (as \i{Aristotle} calls it)}
to a state of rest and quietness.\patch{This,}

This, says \i{Hafen Slawkenbergius}, con-\break
\stick{stitutes the catastrophe or peripeitia of}
\stick{my tale\tsk and that is the part of it I am}
going to relate.

We left the stranger behind the curtain\break
asleep\tsk he enters now upon the stage.

\tsk What dost thou prick up thy ears\break
\stick{at?\tsk ’tis nothing but a man upon a horse}
\stick{\tsk was the last word the stranger uttered}
\stick{to his mule. It was not proper then to}
\stick{tell the reader, that the mule took his}
\stick{master’s word for it; and without any}
\stick{more \i{ifs} or \i{ands}, let the traveller and his}
horse pass by.

The traveller was hastening with all\break
\stick{diligence to get to \i{Strasburg} that night}
\stick{\tsk What a fool am I, said the traveller}
\stick{to himself, when he had rode about a}
\natch{league}
\stick{league farther, to think of getting into}
\stick{\i{Strasburg} this night.\tsk \i{Strasburg!}\tsk the}
\stick{great \i{Strasburg!}\tsk \i{Strasburg}, the capital}
\stick{of all \i{Alsatia!} \i{Strasburg}, an imperial}
\stick{city! \i{Strasburg}, a sovereign state! \i{Stras}-}
\stick{\i{burg}, garrisoned with five thousand of}
\stick{the best troops in all the world!\tsk Alas!}
\stick{if I was at the gates of \i{Strasburg} this}
\stick{moment, I could not gain admittance into}
\stick{it for a ducat,\tsk nay a ducat and half\tsk}
\stick{’tis too much\tsk better go back to the last}
\stick{inn I have passed\tsk than lie I know not}
\stick{where\tsk or give I know not what. The}
\stick{traveller, as he made these reflections in}
\stick{his mind, turned his horse’s head about,}
\stick{and three minutes after the stranger had}
\stick{been conducted into his chamber, he ar-}
rived at the same inn.\etp

\tsk We have bacon in the house, said\break
\stick{the host, and bread\tsh and till eleven}
\stick{o’clock this night had three eggs in it\tsk}
\natch{but}
\stick{but a stranger, who arrived an hour ago,}
\stick{has had them dressed into an omlet, and}
we have nothing.\tsh


\tsk Alas! said the traveller, harassed \break
\stick{as I am, I want nothing but a bed\tsk I}
\stick{have one as soft as is in \i{Alsatia}, said the}
host.

\tsk The stranger, continued he, should\break
\stick{have slept in it, for ’tis my best bed, but}
\stick{upon the score of his nose\tsk He has got}
\stick{a defluxion, said the traveller\tsk Not that}
\stick{I know, cried the host\tsk But ’tis a camp-}
\stick{bed, and \i{Jacinta}, said he, looking to-}
\stick{wards the maid, imagined there was not}
\stick{room in it to turn his nose in\tsk Why so?}
\stick{cried the traveller starting back\tsk It is so}
\stick{long a nose, replied the host\tsk The tra-}
\stick{veller fixed his eyes upon \i{Jacinta}, then}
\stick{upon the ground\tsk kneeled upon his right}
\natch{knee}
\stick{knee\tsk had just got his hand laid upon}
\stick{his breast\tsk Trifle not with my anxiety,}
\stick{said he, rising up again\tsk ’Tis no trifle,}
\stick{said \i{Jacinta}, ’tis the most glorious nose!}
\stick{\tsk The traveller fell upon his knee again\tsk}
\stck{laid his hand upon his breast\tsk then said he,}
\stick{looking up to heaven! thou hast con-}
\stick{ducted me to the end of my pilgrimage}
\tsh ’Tis \i{Diego!}

The traveller was the brother of the \break
\stick{Julia, so often invoked that night by the}
\stick{stranger as he rode from \i{Strasburg} upon}
\stick{his mule; and was come, on her part,}
\stick{in quest of him. He had accompanied}
\stick{his sister from \i{Valadolid} across the \i{Pyre}-}
\stick{\i{nean} mountains thro’ \i{France}, and had}
\stick{many an entangled skein to wind off in}
\stick{pursuit of him thro’ the many meanders}
\stick{and abrupt turnings of a lover’s thorny}
tracks.\patch{\tsk Julia}

\tsk Julia had sunk under it\tsk and had\break
\stick{not been able to go a step farther than to}
\stck{\i{Lyons}, where, with the many disquietudes}
\stick{of a tender heart, which all talk of\tsk but}
\stick{few feel\tsk she sicken’d, but had just}
\stick{strength to write a letter to \i{Diego}; and}
\stick{having conjured her brother never to see}
\stick{her face till he had found him out, and}
\stick{put the letter into his hands, Julia took}
to her bed.

\i{Fernandez} (for that was her brother’s\break
\stick{name)\tsh tho’ the camp-bed was as soft as}
\stick{any one in \i{Alsace}, yet he could not shut}
\stick{his eyes in it.\tsk As soon as it was day he}
\stick{rose, and hearing \i{Diego} was risen too,}
\stick{he enter’d his chamber, and discharged}
his sister’s commission.

The letter was as follows:\patch{Seig.}

\indent\quad Seig. \textsc{Diego},\\
\indent\lqq Whether my suspicions of your nose\break
\stick{\lqq were justly excited or not\tsk ’tis not now}
\stick{\lqq to inquire\tsk it is enough I have not}
\stick{\lqq had firmness to put them to farther}
\lqq tryal.

\indent\lqq How could I know so little of my-\break
\stick{\lqq self, when I sent my \i{Duenna} to forbid}
\stick{\lqq your coming more under my lattice?}
\stick{\lqq or how could I know so little of you,}
\stick{\lqq \i{Diego}, as to imagine you would not}
\stick{\lqq have staid one day in \i{Valadolid} to have}
\stick{\lqq given ease to my doubts?\tsk Was I to}
\stick{\lqq be abandoned, \i{Diego}, because I was}
\stick{\lqq deceived? or was it kind to take me}
\stick{\lqq at my word, whether my suspicions}
\stick{\lqq were just or no, and leave me, as you}
\stick{\lqq did, a prey to much uncertainty and}
\lqq sorrow.\patch{\lqq In}

\indent\lqq In what manner Julia has resented\break
\stick{\lqq this\tsk my brother, when he puts this}
\stick{\lqq letter into your hands, will tell you:}
\stick{\lqq He will tell you in how few moments}
\stick{\lqq she repented of the rash message she}
\stick{\lqq had sent you\tsk in what frantic haste}
\stick{\lqq she flew to her lattice, and how many}
\stick{\lqq days and nights together she leaned}
\stick{\lqq immoveably upon her elbow, looking}
\stick{\lqq thro’ it towards the way which \i{Diego}}
\lqq was wont to come.\hfill

\indent\lqq He will tell you, when she heard\break
\stick{\lqq of your departure\tsk how her spirits de-}
\stick{\lqq serted her\tsk how her heart sicken’d\tsk}
\stick{\lqq how piteously she mourn’d\tsh how low}
\stick{\lqq she hung her head. O \i{Diego!} how}
\stick{\lqq many weary steps has my brother’s}
\stick{\lqq pity led me by the hand languishing}
\stick{\lqq to trace out yours! how far has desire}
\stick{\lqq carried me beyond strength\tsh and how}
\natch{\lqq oft}
\stick{\lqq oft have I fainted by the way, and}
\stick{\lqq sunk into his arms, with only power}
\lqq to cry out\tsk O my \i{Diego!}

\indent\lqq If the gentleness of your carriage\break
\stick{\lqq has not belied your heart, you will fly}
\stick{\lqq to me, almost as fast as you fled from}
\stick{\lqq me\tsk haste as you will, you will arrive}
\stick{\lqq but to see me expire.\tsk ’Tis a bitter}
\stick{\lqq draught, \i{Diego}, but oh!\@ ’tis embitter’d}
\lqq still more by dying \i{un}\tsh.”

She could proceed no farther.

\i{Slawkenbergius} supposes the word in-\break
\stick{tended was \i{unconvinced}, but her strength}
would not enable her to finish her letter.

The heart of the courteous \i{Diego}\break
\stick{overflowed as he read the letter\tsk he or-}
\stick{dered his mule forthwith and \i{Fernandez}’s}
\natch{horse}
\stick{horse to be saddled; and as no vent in}
\stick{prose is equal to that of poetry in such}
\stick{conflicts\tsk chance, which as often directs}
\stick{us to remedies as to \i{diseases}, having}
\stick{thrown a piece of charcoal into the wind-}
\stick{ow\tsk \i{Diego} availed himself of it, and}
\stick{whilst the ostler was getting ready his}
\stick{mule, he eased his mind against the wall}
as follows.

\bigskip
\centerline{\lsss{ODE}.}
\vskip-0.5\parskip
\bgroup\itshape\noindent
\null\quad Harsh and untuneful are the notes of love,\\
\null\qquad Unless my Julia strikes the key,\\
\null\quad Her hand alone can touch the part,\\
\null\qquad Whose dulcet move-\\
\null\quad -ment\sic charms the heart,\\
\null\qquad And governs all the man with sympa-\\
\null\qquad\quad thetic sway.
\par\egroup

\medskip
\centerline{2d.}
\vskip-0.5\parskip
\bgroup\itshape\small\noindent
\null\qquad O Julia!
\par\egroup
\rightline{The}
\newpage
The lines were very natural\tsk for they\break
\stick{were nothing at all to the purpose, says}
\stick{\i{Slawkenbergius}, and ’tis a pity there were}
\stick{no more of them; but whether it was}
\stick{that Seig.\@ \i{Diego} was slow in composing}
\stick{verses\tsk or the ostler quick in saddling}
\stick{mules\tsk is not averred; certain it was,}
\stick{that \i{Diego}’s mule and \i{Fernandez}’s horse}
\stick{were ready at the door of the inn, before}
\stick{\i{Diego} was ready for his second stanza;}
\stick{so without staying to finish his ode, they}
\stick{both mounted, sallied forth, passed the}
\stick{\i{Rhine}, traversed \i{Alsace}, shaped their}
\stick{course towards \i{Lyons}, and before the}
\stck{\i{Strasburgers} and the abbess of \i{Quedlinberg}}
\stick{had set out on their cavalcade, had \i{Fer}-}
\stick{\i{nandez}, \i{Diego}, and his \i{Julia}, crossed}
\stick{the \i{Pyrenean} mountains, and got safe to}
\i{Valadolid.}\patch{’Tis}

’Tis needless to inform the geographi-\break
\stick{cal reader, that when \i{Diego} was in \i{Spain},}
\stick{it was not possible to meet the courteous}
\stick{stranger in the \i{Frankfort} road; it is}
\stick{enough to say, that of all restless desires,}
\stick{curiosity being the strongest\tsk the \i{Stras}-}
\stick{\i{burgers} felt the full force of it; and that}
\stck{for three days and nights they were tossed}
\stick{to and fro in the \i{Frankfort} road, with}
\stick{the tempestuous fury of this passion, be-}
\stick{fore they could submit to return home\tsk}
\stick{When alas! an event was prepared for}
\stick{them, of all others the most grievous}
that could befal a free people.

As this revolution of the \i{Strasburgers}\break
\stick{affairs is often spoken of, and little un-}
\stick{derstood, I will, in ten words, says \i{Slawk}-}
\stick{\i{enbergius}, give the world an explanation}
\stick{of it, and with it put an end to my}
tale.\patch{Every}

Every body knows of the grand sy-\break
\stick{stem of Universal Monarchy, wrote by}
\stick{order of Mons.\@ \i{Colbert}, and put in ma-}
\stick{nuscript into the hands of \i{Lewis} the}
fourteenth, in the year 1664.

’Tis as well known, that one branch\break
\stick{out of many of that system, was the}
\stick{getting possession of \i{Strasburg}, to favour}
\stick{an entrance at all times into \i{Suabia}, in}
\stick{order to disturb the quiet of \i{Germany}\tsk}
\stick{and that in consequence of this plan,}
\stick{\i{Strasburg} unhappily fell at length into}
their hands.

It is the lot of few to trace out the\break
\stick{true springs of this and such like revolu-}
\stick{tions\tsk The vulgar look too high for}
\stick{them\tsk Statesmen look too low\tsk Truth}
(for once) lies in the middle.\patch{What}

What a fatal thing is the popular pride\break
\stick{of a free city!\@ cries one historian\tsk The}
\stick{\i{Strasburgers} deemed it a diminution of}
\stick{their freedom to receive an imperial gar-}
\tight{rison\tsk and so fell a prey to a \i{French} one}.

The fate, says another, of the \i{Stras}-\break
\stick{\i{burgers}, may be a warning to all free}
\stick{people to save their money\tsk They anti-}
\stick{cipated their revenues\tsk brought them-}
\stick{selves under taxes, exhausted their}
\stick{strength, and in the end became so weak}
\stick{a people, they had not strength to keep}
\stick{their gates shut, and so the \i{French} pushed}
them open.

Alas!\@ alas!\@ cries \i{Slawkenbergius}, ’twas\break
\stick{not the \i{French}\tsk ’twas \s{curiosity} pushed}
\stick{them open\tsk The \i{French} indeed, who are}
\stick{ever upon the catch, when they saw the}
\stick{\i{Strasburgers}, men, women, and children,}
\natch{all}
\stick{all marched out to follow the stranger’s}
\stick{nose\tsk each man followed his own, and}
marched in.

Trade and manufactures have decayed\break
\stick{and gradually grown down ever since\tsk}
\stick{but not from any cause which commer-}
\stick{cial heads have assigned; for it is owing}
\stick{to this only, that Noses have ever so run}
\stick{in their heads, that the \i{Strasburgers} could}
not follow their business.

Alas! alas! cries \i{Slawkenbergius}, ma-\break
\stick{king an exclamation\tsk it is not the first\tsk}
\stick{and I fear will not be the last fortress}
\stick{that has been either won\tsh or lost by}
\s{Noses}.

\bigskip
\centerline{The\enspace \lsss{END}\enspace of}
\centerline{\i{Slawkenbergius}’s \lss{\textsc{Tale}}.}

\bigskip

\rightline{\ls{CHAP}.}
\newpage
\section{\chapstrut\lsss{CHAP}.\enspace I.}

\dropcap{W}{ith} all this learning upon Noses\break
running perpetually in my fa-\break
\stick{ther’s facy\sic\tsk with so many family preju-}
\stick{dices\tsk and ten decads of such tales run-}
\stick{ning on for ever along with them\tsk how}
\stick{was it possible with such exquisite\tsk was}
\stick{it a true nose?\tsk That a man with such}
\stick{exquisite feelings as my father had, could}
\stick{bear the shock at all below stairs\tsk or in-}
\stick{deed above stairs, in any other posture,}
but the very posture I have described.

\tsk Throw yourself down upon the bed,\break
\stick{a dozen times\tsk taking care only to place}
\stick{a looking-glass first in a chair on one side}
\stick{of it, before you do it\tsh But was the}
\natch{stran-}
\stick{stranger’s nose a true nose\tsk or was it a}
false one?

To tell that before-hand, madam,\break
\stick{would be to do injury to one of the best}
\stick{tales in the christian world; and that is}
\stick{the tenth of the tenth decad which imme-}
diately follows this.

This tale, crieth \i{Slawkenbergius} some-\break
\stick{what exultingly, has been reserved by}
\stick{me for the concluding tale of my whole}
\stick{work; knowing right well, that when I}
\stick{shall have told it, and my reader shall}
\stick{have read it thro’\tsk ’twould be even high}
\stick{time for both of us to shut up the book;}
\stick{inasmuch, continues \i{Slawkenbergius}, as I}
\stick{know of no tale which could possibly}
ever go down after it.

\tsh ’Tis a tale indeed!\patch{This}
% -- tighter leading...  294/22 \simeq 13.3636 
\fontsize{10}{13.3636}\selectfont
This sets out with the first interview\break
\stick{in the inn at \i{Lyons}, when \i{Fernandez} left}
\stick{the courteous stranger and his sister \i{Julia}}
alone in her chamber, and is overwritten,

\bigskip
\centerline{The\enspace\lss{\textsc{Intricacies}}}
\centerline{of}
\centerline{\i{Diego} and \i{Julia.}}

Heavens! thou art a strange creature
\stick{\i{Slawkenbergius!} what a whimsical view}
\stick{of the involutions of the heart of woman}
\stick{hast thou opened! how this can ever be}
\stick{translated, and yet if this specimen of}
\stick{\i{Slawkenbergius}’s tales, and the exquisi-}
\stick{tiveness of his moral should please the}
\stick{world\tsk translated shall a couple of vo-}
\stick{lumes be.\tsk Else, how this can ever be}
\stick{translated into good \i{English}, I have no}
\stick{sort of conception.\tsk There seems in some}
\stick{passages to want a sixth sense to do it}
\natch{rightly.}
\stick{rightly.\tsh What can he mean by the}
\stick{lambent pupilability of slow, low, dry}
\stick{chat, five notes below the natural tone,}
\stick{\tsk which you know, madam, is little}
\stick{more than a whisper? The moment I}
\stick{pronounced the words, I could perceive}
\stick{an attempt towards a vibration in the}
\stick{strings, about the region of the heart.\tsk}
\stick{The brain made no acknowledgment.\tsk}
\stick{There’s often no good understanding}
\stick{betwixt ’em\tsk I felt as if I understood}
\stick{it.\tsk I had no ideas.\tsk The movement}
\stick{could not be without cause.\tsk I’m lost.}
\stick{I can make nothing of it,\tsk unless, may}
\stick{it please your worships, the voice, in}
\stick{that case being little more than a whis-}
\stick{per, unavoidably forces the eyes to}
\stick{approach not only within six inches}
\stick{of each other\tsk but to look into the}
\stick{pupils\tsk is not that dangerous?\tsh But it}
\stick{can’t be avoided\tsk for to look up to the} 
\natch{cieling,}
\stick{cieling, in that case the two chins un-}
\stick{avoidably meet\tsk and to look down into}
\stick{each others laps, the foreheads come into}
\stick{immediate contact, which at once puts}
\stick{an end to the conference\tsk I mean to the}
\stick{sentimental part of it.\tsh What is left,}
madam, is not worth stooping for.

\section{\lsss{CHAP}.\enspace II.}

\dropcap{M}{y} father lay stretched across the\break
bed as still as if the hand of death\break
\stick{had pushed him down, for a full hour}
\stick{and a half, before he began to play upon}
\stick{the floor with the toe of that foot which}
\stick{hung over the bed-side; my uncle \toby's}
\stick{heart was a pound lighter for it.\tsk In a}
\stick{few moments, his left-hand, the knuckles}
\stick{of which had all the time reclined upon}
\stick{the handle of the chamber-pot, came}
\stick{to its feeling\tsk he thrust it a little more}
\natch{within}
\stick{within the valance\tsk drew up his hand,}
\stick{when he had done, into his bosom\tsk gave}
\stick{a hem!\tsk My good uncle \toby, with infi-}
\stick{nite pleasure, answered it; and full gladly}
\stck{would have ingrafted a sentence of conso-}
\stick{lation upon the opening it afforded: but}
\stick{having no talents, as I said, that way,}
\stick{and fearing moreover that he might set}
\stick{out with something which might make a}
\stick{bad matter worse, he contented himself}
\stick{with resting his chin placidly upon the}
cross of his crutch.

\tighter{Now whether the compression shortened}\break
\stick{my uncle \toby's face into a more plea-}
\stick{sureable oval,\tsk or that the philanthropy}
\stick{of his heart, in seeing his brother begin-}
\stick{ning to emerge out of the sea of his af-}
\stick{flictions, had braced up his muscles,\tsk}
\stick{so that the compression upon his chin}
\stick{only doubled the benignity which was}
\natch{there}
\stick{there before, is not hard to decide.\tsk My}
\stick{father, in turning his eyes, was struck}
\stick{with such a gleam of sun-shine in his}
\stick{face, as melted down the sullenness of}
his grief in a moment.

He broke silence as follows.

\section{\chapstrut\lsss{CHAP}.\enspace III.}

\dropcap[findent=-.5pt]{D}{id} ever man, brother \toby, cried\break
my father, raising himself up upon\break
\stick{his elbow, and turning himself round to}
\stick{the opposite side of the bed where my}
\stick{uncle \toby was sitting in his old fringed}
\stick{chair, with his chin resting upon his}
\stick{crutch\tsk did ever a poor unfortunate man,}
\stick{brother \toby, cried my father, receive so}
\stick{many lashes?\tsk The most I ever saw}
\stick{given, quoth my uncle \toby, (ringing}
\natch{the}
\stick{the bell at the bed’s head for \i{Trim)}\sic was}
\stick{to a grenadier, I think in \i{Makay}’s regi-}
ment.

\tsk Had my uncle \toby shot a bullet\break
\stick{thro’ my father’s heart, he could not}
\stick{have fallen down with his nose upon the}
quilt more suddenly.

Bless me! said my uncle \toby.

\bigskip
\section{\lsss{CHAP}.\enspace IV.}

\dropcap{W}{as} it \i{Makay}’s regiment, quoth\break
my uncle \toby, where the poor\break
\stck{grenadier was so unmercifully whipp’d at}
\stick{\i{Bruges} about the ducats.\tsk O Christ! he}
\stick{was innocent! cried \i{Trim} with a deep}
\stick{sigh.\tsk And he was whipp’d, may it}
\stick{please your honour, almost to death’s}
\natch{door.}
\stick{door.\tsk They had better have shot him}
\stick{outright as he begg’d, and he had gone}
\stick{directly to heaven, for he was as innocent}
\stick{as your honour.\tsh I thank thee, \i{Trim},}
\stick{quoth my uncle \toby.\quad I never think of}
\stick{his, continued \i{Trim}, and my poor bro-}
\stick{ther \i{Tom}’s misfortunes, for we were all}
\stick{three school-fellows, but I cry like a}
\stick{coward.\tsk Tears are no proof of cowar-}
\stick{dice, \i{Trim.}\tsk I drop them oft-times my-}
\stick{self, cried my uncle \toby.\tsk I know your}
\stick{honour does, replied \i{Trim}, and so am}
\stick{not ashamed of it myself.\tsk But to think,}
\stick{may it please your honour, continued}
\stick{\i{Trim}, a tear stealing into the corner of}
\stick{his eye as he spoke\tsk to think of two vir-}
\stick{tuous lads with hearts as warm in their}
\stick{bodies, and as honest as God could make}
\stick{them\tsk the children of honest people,}
\stick{going forth with gallant spirits to seek}
\stick{their fortunes in the world\tsk and fall into}
\natch{such}
\stick{such evils!\tsk poor \i{Tom!} to be tortured}
\stick{upon a rack for nothing\tsk but marrying}
\stck{a \i{Jew}’s widow who sold sausages\tsk honest}
\stick{\i{Dick Johnson}’s soul to be scourged out of}
\stick{his body, for the ducats another man put}
\stick{into his knapsack!\tsk O!\tsk these are mis-}
\stick{fortunes, cried \i{Trim}, pulling out his}
\stck{handkerchief\tsk these are misfortunes, may}
\stick{it please your honour, worth lying down}
and crying over.

\tsk My father could not help blushing.

\tsk ’Twould be a pity, \i{Trim}, quoth my\break
\stck{uncle \toby, thou shouldst ever feel sorrow}
\stick{of thy own\tsk thou feelest it so tenderly}
\stick{for others.\tsk Alack-o-day, replied the}
\stick{corporal, brightening up his face\tsk your}
\stck{honour knows I have neither wife or child}
\stick{\tsh I can have no sorrows in this world.}
\stick{\tsk My father could not help smiling.\tsk}
\natch{As}
\stick{As few as any man, \i{Trim}, replied my}
\stick{uncle \toby; nor can I see how a fellow}
\stick{of thy light heart can suffer, but from}
\stick{the distress of poverty in thy old age\tsk}
\stick{when thou art passed all services, \i{Trim},\tsk}
\stck{and hast out-lived thy friends\tsk An’ please}
\stick{your honour, never fear, replied \i{Trim}}
\stick{chearily\tsk But I would have thee never}
\stick{fear, \i{Trim}, replied my uncle; and there-}
\stick{fore, continued my uncle \toby, throwing}
\stick{down his crutch, and getting up upon}
\stick{his legs as he uttered the word \i{there}-}
\stick{\i{fore}\tsk in recompence, \i{Trim}, of thy long}
\stick{fidelity to me, and that goodness of thy}
\stick{heart I have had such proofs of\tsk whilst}
\stick{thy master is worth a shilling\tsk thou shalt}
\stick{never ask elsewhere, \i{Trim}, for a penny.}
\stick{\i{Trim} attempted to thank my uncle \toby,}
\stick{\tsk but had not power\tsk tears trickled}
\stick{down his cheeks faster than he could}
\stick{wipe them off\tsk He laid his hands upon}
\natch{his}
\stick{his breast\tsh made a bow to the ground,}
and shut the door.

\tsk I have left \i{Trim} my bowling-green,\break
\stick{cried my uncle \i{Toby}\tsk My father smiled}
\stick{\tsk I have left him moreover a pension,}
\stick{continued my uncle \i{Toby}\tsk My father}
looked grave.

\section{\lsss{CHAP}.\enspace V.}

\dropcap{I}{s} this a fit time, said my father to\break
himself, to talk of \s{pensions} and\break
\ls{\textsc{grenadiers}?}

\section{\lsss{CHAP}.\enspace VI.}

\dropcap{W}{hen} my uncle \toby first men-\break
tioned the grenadier, my father,\break
\stick{I said, fell down with his nose flat to the}
\stick{quilt, and as suddenly as if my uncle}
\natch{\toby}
\stick{\toby had shot him; but it was not added,}
\stick{that every other limb and member of my}
\stck{father instantly relapsed with his nose into}
\stick{the same precise attitude in which he lay}
\stick{first described; so that when corporal}
\stick{\i{Trim} left the room, and my father found}
\stick{himself disposed to rise off the bed,\tsk he}
\stick{had all the little preparatory movements}
\stick{to run over again, before he could do it.}
\stick{\tsk  Attitudes are nothing, madam,\tsk ’tis the}
\stick{transition from one attitude to another\tsk}
\stick{like the preparation and resolution of the}
discord into harmony, which is all in all.

For which reason my father played the\break
\stick{same jig over again with his toe upon the}
\stick{floor\tsk pushed the chamber-pot still a}
\stick{little farther within the valance\tsk gave a}
\stick{hem\tsk raised himself up upon his elbow}
\stick{\tsk and was just beginning to address}
\stick{himself to my uncle \i{Toby}\tsk when recol-}
\natch{lecting}
\stick{lecting the unsuccessfulness of his first}
\stick{effort in that attitude,\tsk he got upon his}
\stick{legs, and in making the third turn across}
\stick{the room, he stopped short before my}
\stick{uncle \toby; and laying the three first}
\stick{fingers of his right-hand in the palm of}
\stick{his left, and stooping a little, he addressed}
himself to my uncle \toby as follows.

\section{\chapstrut\lsss{CHAP}.\enspace VII.}

\dropcap{W}{hen} I reflect, brother \toby,\break
upon \s{man}; and take a view of\break
\stick{that dark side of him which represents}
\stick{his life as open to so many causes of}
\stick{trouble\tsk when I consider, brother \toby,}
\stick{how oft we eat the bread of affliction,}
\stick{and that we are born to it, as to the}
\stick{portion of our inheritance\tsk I was born}
\stick{to nothing, quoth my uncle \toby, inter-}
\natch{rupting}
\stick{rupting my father\tsk but my commission.}
\stick{Zooks! said my father, did not my uncle}
\stick{leave you a hundred and twenty pounds}
\stick{a year?\tsk What could I have done with-}
\stick{out it? replied my uncle \toby.\tsk That’s}
\stick{another concern, said my father testily\tsk}
\stick{But I say, \toby, when one runs over the}
\stick{catalogue of all the cross reckonings and}
\stick{sorrowful \i{items} with which the heart of}
\stick{man is overcharged, ’tis wonderful by}
\stick{\tight{what hidden resources the mind is enabled}}
\stick{to stand it out, and bear itself up, as it}
\stick{does against the impositions laid upon}
\stick{our nature.\tsh ’Tis by the assistance of}
\stick{Almighty God, cried my uncle \toby,}
\stick{looking up, and pressing the palms of}
\stick{his hands close together\tsk ’tis not from}
\stick{our own strength, brother \i{Shandy}\tsk a sen-}
\stick{tinel in a wooden centry-box, might as}
\stick{well pretend to stand it out against a}
\stick{detachment of fifty men,\tsk we are upheld}
\natch{by}
\stick{by the grace and the assistance of the best}
of Beings.

\tsk That is cutting the knot, said my\break
\stick{father, instead of untying it.\tsk But give}
\stick{me leave to lead you, brother \toby, a}
little deeper into the mystery.

With all my heart, replied my uncle\break\toby.

My father instantly exchanged the at-\break
\stick{titude he was in, for that in which \i{So}-}
\stick{\i{crates} is so finely painted by \i{Raffael} in}
\stick{his school of \i{Athens}; which your con-}
\stick{noisseurship knows is so exquisitely ima-}
\stick{gined, that even the particular manner}
\stick{of the reasoning of \i{Socrates} is expressed}
\stick{by it\tsk for he holds the fore-finger of his}
\stick{left-hand between the fore-finger and the}
\stick{thumb of his right, and seems as if he}
\natch{was}
\stick{was saying to the libertine he is re-}
\stick{claiming\tsk \lqq\i{You grant me} this\tsk and this:}
\stick{\lqq and this, and this, I don’t ask of}
\stick{\lqq you\tsk they follow of themselves in}
\lqq course.\,”

So stood my father, holding fast his\break
\stick{fore-finger betwixt his finger and his}
\stick{thumb, and reasoning with my uncle}
\stick{\toby as he sat in his old fringed chair,}
\stick{valanced around with party-coloured}
\stick{worsted bobs\tsk O \i{Garrick!} what a rich}
\stick{scene of this would thy exquisite powers}
\stick{make!\@ and how gladly would I write such}
\stick{another to avail myself of thy immor-}
tality, and secure my own behind it.\patch{\lss{CHAP}.}

\section{\chapstrut\lsss{CHAP}.\enspace VIII.}

\dropcap{T}{hough} man is of all others\break
the most curious vehicle, said my\break
\stick{father, yet at the same time ’tis of so}
\stick{slight a frame and so totteringly put to-}
\stck{gether, that the sudden jerks and hard jost-}
\stick{lings it unavoidably meets with in this}
\stick{rugged journey, would overset and tear}
\stick{it to pieces a dozen times a day\tsk was it}
\stick{not, brother \toby, that there is a secret}
\stick{spring within us\tsk Which spring, said my}
\stick{uncle \toby, I take to be Religion.\tsk Will}
\stick{that set my child’s nose on?  cried my}
\stick{father, letting go his finger, and striking}
\stick{one hand against the other\tsk It makes}
\stick{every thing straight for us, answered my}
\stick{uncle \i{Toby}\tsk Figuratively speaking, dear}
\stick{\toby, it may, for aught I know, said}
\natch{my}
\stick{my father; but the spring I am speaking}
\stick{of, is that great and elastic power within}
\stick{us of counterbalancing evil, which like}
\stck{a secret spring in a well-ordered machine,}
\stick{though it can’t prevent the shock\tsh at}
least it imposes upon our sense of it.

Now, my dear brother, said my fa-\break
\stick{ther, replacing his fore-finger, as he was}
\stick{coming closer to the point,\tsk had my}
\stick{child arrived safe into the world, un-}
\stick{martyr’d in that precious part of him\tsk}
\stick{fanciful and extravagant as I may appear}
\stick{to the world in my opinion of christian}
\stick{names, and of that magic bias which}
\stick{good or bad names irresistibly impress}
\stick{upon our characters and conducts\tsk hea-}
\stck{ven is witness! that in the warmest trans-}
\stick{ports of my wishes for the prosperity of}
\stick{my child, I never once wished to crown}
\stick{his head with more glory and honour,}
\natch{than}
\stick{than what \s{George} or \textscl{Edward} would}
have spread around it.

But alas!\@ continued my father, as the\break
\stick{greatest evil has befallen him\tsk I must}
\stick{counteract and undo it with the greatest}
good.

He shall be christened \i{Trismegistus},\break
brother.

I wish it may answer\tsk replied my\break
uncle \toby, rising up.

\section{\chapstrut\lsss{CHAP}.\enspace IX.}

\dropcap{W}{hat} a chapter of chances, said\break
my father, turning himself about\break
\stick{upon the first landing, as he and my}
\stick{uncle \toby were going down stairs\tsh}
\stick{what a long chapter of chances do the}
\natch{events}
\stick{events of this world lay open to us!}
\stick{Take pen and ink in hand, brother \toby,}
\stick{and calculate it fairly\tsk I know no more}
\stick{of calculations than this balluster, said}
\stick{my uncle \toby, (striking short of it with}
\stick{his crutch, and hitting my father a de-}
\stick{sperate blow souse upon his shin-bone)\tsk}
\stick{’Twas a hundred to one\tsk cried my uncle}
\stick{\toby.\tsk\!\tsk I thought, quoth my father,}
\stick{(\,rubbing his shin\,) you had known}
\stick{nothing of calculations, brother \toby.}
\stick{\tsk ’Twas a meer chance, said my uncle}
\stick{\i{Toby}\tsk Then it adds one to the chapter}
\tsk replied my father.

The double success of my father’s re-\break
\stick{partees tickled off the pain of his shin at}
\stick{once\tsk it was well it so fell out\tsk (chance!}
\stick{again)\tsk or the world to this day had}
\stick{never known the subject of my father’s}
\stick{calculation\tsk to guess it\tsk there was no}
\natch{chance}
\stick{chance\tsk What a lucky chapter of chances}
\stick{has this turned out! for it has saved me}
\stick{the trouble of writing one express, and}
\stick{in truth I have anew\sic\ already upon my}
\stick{hands without it\tsh Have not I pro-}
\stick{mised the world a chapter of knots?}
\stick{two chapters upon the right and the}
\stick{wrong end of a woman? a chapter upon}
\stick{whiskers? a chapter upon wishes?\tsk a}
\stick{chapter of noses?\tsk No, I have done}
\stick{that\tsk a chapter upon my uncle \toby's}
\stick{modesty: to say nothing of a chapter}
\stick{upon chapters, which I will finish before}
\stick{I sleep\tsk by my great grandfather’s whis-}
\stick{kers, I shall never get half of ’em through}
this year.

Take pen and ink in hand, and calcu-\break
\stick{late it fairly, brother \toby, said my fa-}
\stick{ther, and it will turn out a million to one,}
\stick{that of all the parts of the body, the edge}
\natch{of}
\stick{of the forceps should have the ill luck}
\stick{just to fall upon and break down that}
\stick{one part, which should break down the}
fortunes of our house with it.

It might have been worse, replied my\break
\stick{uncle \i{Toby}\tsk I don’t comprehend, said}
\stick{my father\tsk Suppose the hip had present-}
\stick{ed, replied my uncle \toby, as \drslop}
foreboded.

My father reflected half a minute\tsk\break
\stick{looked down\tsk touched the middle of his}
forehead slightly with his finger\tsk

\tsk True, said he.\patch{\lss{CHAP}.}

\null\smallskip
\section{\lsss{CHAP}.\enspace X.}

\dropcap[findent=0.75pt]{I}{s}
it not a shame to make two chapters\break
of what passed in going down one\break
\stick{pair of stairs? for we are got no farther}
\stick{yet than to the first landing, and there}
\stick{are fifteen more steps down to the bot-}
\stick{tom; and for aught I know, as my fa-}
\stick{ther and my uncle \toby are in a talking}
\stick{humour, there may be as many chapters}
\stick{as steps;\tsk let that be as it will, Sir, I}
\stick{can no more help it than my destiny:\tsk}
\stick{A sudden impulse comes across me\tsh}
\stick{drop the curtain, \i{Shandy}\tsk I drop it\tsh}
\stck{Strike a line here across the paper, \i{Tristram}}
\stick{\tsk I strike it\tsk and hey for a new chapter? \sic}

The duce of any other rule have I to\break
\stick{govern myself by in this affair\tsk and if I}
\natch{had}
\stick{had one\tsk as I do all things out of all}
\stick{rule\tsk I would twist it and tear it to}
\stick{pieces, and throw it into the fire when I}
\stick{had done\tsk Am I warm? I am, and the}
\stick{cause demands it\tsk a pretty story! is a}
\stick{man to follow rules\tsh or rules to follow}
him?

Now this, you must know, being my\break
\stick{chapter upon chapters, which I promised}
\stick{to write before I went to sleep, I thought}
\stick{it meet to ease my conscience entirely}
\stick{before I lay’d down, by telling the world}
\stick{all I knew about the matter at once: Is}
\stick{not this ten times better than to set out}
\stck{dogmatically with a sententious parade of}
\stick{wisdom, and telling the world a story of}
\stick{a roasted horse\tsk that chapters relieve the}
\stick{mind\tsk that they assist\tsk or impose upon}
\stick{the imagination\tsk and that in a work of}
\stck{this dramatic cast they are as necessary as}
\natch{the}
\stick{the shifting of scenes\tsk with fifty other}
\stick{cold conceits, enough to extinguish the}
\stick{fire which roasted him.\tsk O! but to un-}
\stick{derstand this, which is a puff at the fire}
\stick{of \i{Diana}’s temple\tsk you must read \i{Lon}-}
\stick{\i{ginus}\tsk read away\tsk if you are not a jot}
\stck{the wiser by reading him the first time over}
\stick{\tsk never fear\tsk read him again\tsk \i{Avicenna}}
\stick{and \i{Licetus}, read \i{Aristotle}’s metaphysicks}
\stick{forty times through a piece, and never}
\stick{understood a single word.\tsk But mark the}
\stick{consequence\tsk \i{Avicenna} turned out a de-}
\stick{sperate writer at all kinds of writing\tsk}
\stick{for he wrote books \i{de omni scribili}; and}
\stick{for \i{Licetus (Fortunio)} though all the}
\stick{world knows he was born a fœtus \fnast, of}
\stick{\hfill no}

\bgroup\fontsize{8}{10}\selectfont\noindent
\stick{\enspace\fnast\ \i{Ce Fœtus} n’étoit pas plus grand que la paúme}
\stick{de la main; mais son pere l’ayant éxaminè en qua-}
\stick{litè de Médecin, \& ayant trouvé que c’etoit quel-}
\stick{que chose de plus qu’un Embryon, le fit transporter}
\stick{tout vivant à Rapallo, ou il le fit voir à Jerôme}
\stick{\hfill Bardi}\par\egroup
\newpage\noindent
\stick{no more than five inches and a half in}
\stick{length, yet he grew to that astonishing}
\stick{height in literature, as to write a book}
\stick{with a title as long as himself\tsh the}
\stick{learned know I mean his \i{Gonopsychan}-}
\stick{\i{thropologia}, upon the origin of the human}
\stick{soul.\hfill}
\stick{\hfill So}

\vfill
\bgroup\fontsize{8}{10.8}\selectfont\noindent
\stick{Bardi \& à d’autres Medecins du lieu.  On trouva}
\stick{qu’il ne lui manquoit rien d’essentiel a la vie; \&}
\stick{son pere pour faire voir un essai de son expérience,}
\stick{entreprit d’achever l’ouvrage de la Nature, \& de}
\stick{travailler a la formation de l’Enfant avec le même}
\stick{artifice que celui dont on se sert pour faire éclorre}
\stick{les Poulets en Egypte. Il instruisit une Nourisse}
\stick{de tout ce qu’elle avoit à faire, \& ayant fait mettre}
\stick{son fil dans un four proprement accommodè, il}
\stick{reussit à l’élever et a lui faire prendre ses accroisse-}
\stick{mens  necessaires, par l’uniformité d’une chaleur}
\stick{étrangére mesurée éxactement sur les dégrés d’un}
\stick{Thermométre, ou d’un autre instrument équivalent.\hfill}
\stick{(Vide Mich.\@  Giustinian, ne gli Scritt.\@ Liguri á}
\stick{Cart~223.\@ 488.)\hfill}
\stick{\hfill On}
\par\egroup
\newpage
So much for my chapter upon chap-\break
\stick{ters, which I hold to be the best chapter}
\stick{in my whole work; and take my word,}
\stick{whoever reads it, is full as well employed,}
as in picking straws.

\bigskip

\rightline{\ls{CHAP}.}

\bigskip

\bgroup\fontsize{8}{10.8}\selectfont
\stick{\quad On auroit toujours été très-satisfait de l’industrie}
\stick{d’un Pere si experimenté dans l’Art de la Genera-}
\stick{tion, quand il n’auroit pû prolonger la vie a son}
\stick{fils que pour quelques mois, ou pour peu d’années.\hfill}
\stick{\quad Mais quand on se represente que l’Enfant a vecu}
\stick{pres de quatre-vingts ans, \& que il a composé}
\stick{quatre-vingts Ouvrages differents tous fruits d’une}
\stick{longue lecture,\tsk il faut convenir que tout ce qui}
\stick{est incroyable n’est pas toujours faux, \& que la}
\stick{\i{Vraisemblance n’est pas toujours du coté de la Veritè}.\hfill}
\stick{\quad Il n’avoit que dix-neuf ans lors qu’il composa}
\stick{Gonopsychanthropologia de Origine Animæ hu-}
\stick{manæ.\hfill}
\stick{\quad (Les Enfans celebres, revûs \& corriges par M.}
\stick{De la Monnoye de l’Academie Françoise.)\hfill}
\par\egroup

\vfill

\newpage
\null\smallskip
\section{\lsss{CHAP}.\enspace XI.}

\dropcap{W}{e} shall bring all things to rights,\break
said my father, setting his foot\break
\stick{upon the first step from the landing\tsh}
\stick{This \i{Trismegistus}, continued my father,}
\stick{drawing his leg back, and turning to}
\stick{my uncle \i{Toby}\tsk was the greatest \thinspace\i{(Toby\,)}}
\stick{of all earthly beings\tsk he was the greatest}
\stick{king\tsk the greatest lawgiver\tsk the greatest}
\stick{philosopher\tsk and the greatest priest\tsh}
and engineer\tsk said my uncle \toby.\tsk

\tsk In course, said my father.\patch{\lss{CHAP}.}

\section{\chapstrut\lsss{CHAP}.\enspace XII.}

\dropcap[findent=-1pt]{\Usk{16pt}A}{nd} how does your mistress?\break
cried my father, taking the\break
\stick{same step over again from the landing,}
\stick{and calling to \i{Susannah}, whom he saw}
\stick{passing by the foot of the stairs with a}
\stick{\tight{huge pin-cushion in her hand\tsk how does}}
\stick{your mistress? As well, said \i{Susannah},}
\stick{tripping by, but without looking up, as}
\stick{can be expected\tsk What a fool am I,}
\stick{said my father! drawing his leg back}
\stick{again\tsk let things be as they will, brother}
\stick{\toby, ’tis ever the precise answer\tsk And}
\stick{how is the child, pray?\tsk No answer.}
\stick{And where is doctor \i{Slop}?  added my}
\stick{father, raising his voice aloud, and look-}
\stck{ing over the ballusters\tsk \i{Susannah} was out}
of hearing.\patch{Of}

Of all the riddles of a married life,\break
\stick{said my father, crossing the landing, in}
\stick{order to set his back against the wall,}
\stick{whilst he propounded it to my uncle}
\stick{\i{Toby}\tsk of all the puzzling riddles, said}
\stick{he, in a marriage state,\tsk of which you}
\stick{may trust me, brother \toby, there are}
\stick{more asses loads than all \i{Job}’s stock of}
\stick{asses could have carried\tsk there is not one}
\stick{that has more intricacies in it than this\tsk}
\stick{that from the very moment the mistress}
\stick{of the house is brought to bed, every}
\stck{female in it, from my lady’s gentlewoman}
\stick{down to the cinder-wench, becomes an}
\stick{inch taller for it; and give themselves}
\stick{more airs upon that single inch, than all}
their other inches put together.

I think rather, replied my uncle \toby,\break
\stick{that ’tis we who sink an inch lower.\tsh}
\stick{If I meet but a woman with child\tsk I do}
\natch{it}
\stick{it\tsk ’Tis a heavy tax upon that half of}
\stick{our fellow-creatures, brother \i{Shandy},}
\stick{said my uncle \i{Toby}\tsk ’Tis a piteous bur-}
\stick{den upon ’em, continued he, shaking his}
\stick{head.\tsk Yes, yes, ’tis a painful thing\tsk}
\stick{said my father, shaking his head too\tsk}
\stick{but certainly since shaking of heads came}
\stick{into fashion, never did two heads shake}
\stick{together, in concert, from two such dif-}
ferent springs.

\indent$\vcenter{\halign{#\rlap{\phantom{y}}\hss\cr
God bless\cr
Duce take\cr}}\,\Big\}
\vcenter{\halign{#\cr
’em all\tsk said my uncle\cr
\;\toby and my father, each\cr}}$
\\[4pt] to himself.

\smallskip
\section{\lsss{CHAP}.\enspace XIII.}

\dropcap{H}{olla!}\tsk you chairman!\tsk here’s\break
sixpence\tsk do step into that book-\break
\stick{seller’s shop, and call me a \i{day-tall}}
\stick{critick. I am very willing to give any}
\natch{one}
\stick{one of ’em a crown to help me with his}
\stick{tackling, to get my father and my uncle}
\stick{\toby off the stairs, and to put them to}
bed.\tsk

\tsk ’Tis even high time; for except a\break
\stick{short nap, which they both got whilst}
\stick{\i{Trim} was boring the jack-boots\tsk and}
\stick{which, by the bye, did my father no sort}
\stick{of good upon the score of the bad hinge}
\stick{\tsk they have not else shut their eyes, since}
\stick{nine hours before the time that doctor}
\stick{\i{Slop} was led into the back parlour in that}
dirty pickle by \i{Obadiah}.

Was every day of my life to be as\break
\stick{busy a day as this,\tsk and to take up,\tsk}
truce\tsk

\tight{I will not finish that sentence till I have}\break
\stick{made an observation upon the strange}
\natch{state}
\stick{state of affairs between the reader and}
\stick{myself, just as things stand at present\tsk}
\stick{an observation never applicable before to}
\stick{any one biographical writer since the}
\stick{creation of the world, but to myself\tsk}
\stick{and I believe will never hold good to}
\stick{any other, until its final destruction\tsh}
\stick{and therefore, for the very novelty of it}
\stick{alone, it must be worth your worships}
attending to.

I am this month one whole year older\break
\stick{than I was this time twelve-month; and}
\stick{having got, as you perceive, almost into}
\stick{the middle of my fourth volume\tsk and}
\stick{no farther than to my first day’s life\tsk}
\stick{’tis demonstrative that I have three hun-}
\stick{dred and sixty-four days more life to}
\stick{write just now, than when I first set out;}
\stick{so that instead of advancing, as a com-}
\stick{mon writer, in my work with what I}
\natch{have}
\stick{have been doing at it\tsk on the contrary,}
\stick{I am just thrown so many volumes back}
\stick{\tsk was every day of my life to be as busy}
\stick{a day as this\tsk And why not?\tsk and the}
\stick{transactions and opinions of it to take}
\stick{up as much description\tsk And for what}
\stick{reason should they be cut short? as at}
\stick{this rate I should just live 364 times}
\stick{faster than I should write\tsk It must fol-}
\stick{low, an’ please your worships, that the}
\stick{more I write, the more I shall have to}
\stick{write\tsk and consequently, the more your}
\stick{worships read, the more your worships}
will have to read.

Will this be good for your worships\break eyes?

It will do well for mine; and, was it\break
\stick{not that my \s{Opinions} will be the death}
\stick{of me, I perceive shall\sic lead a fine life of}
\natch{it}
\stick{it out of this self-same life of mine; or,}
\stick{in other words, shall lead a couple of fine}
lives together.

As for the proposal of twelve volumes\break
\stick{a year, or a volume a month, it no way}
\stick{alters my prospect\tsk write as I will, and}
\stick{rush as I may into the middle of things,}
\stick{as \i{Horace} advises,\tsk I shall never overtake}
\stick{myself\tsk whipp’d and driven to the last}
\stick{pinch, at the worst I shall have one day}
\stick{the start of my pen\tsk and one day is}
\stick{enough for two volumes\tsk and two vo-}
lumes will be enough for one year.\tsk

Heaven prosper the manufactures of\break
\stick{paper under this propitious reign, which}
\stick{is now open’d to us,\tsk as I trust its pro-}
\stick{vidence will prosper every thing else in}
it that is taken in hand.\tsk\patch{As}

As for the propagation of Geese\tsk I\break
\stick{give myself no concern\tsk Nature is all}
\stick{bountiful\tsk I shall never want tools to}
work with.

\tsk So then, friend! you have got my\break
\stick{father and my uncle \toby off the stairs,}
\stck{and seen them to bed?\tsk And how did you}
\stick{manage it?\tsk You dropp’d a curtain at}
\stick{the stairs foot\tsk I thought you had no}
\stick{other way for it\tsk Here’s a crown for}
your trouble.

\section{\chapstrut\lsss{CHAP}.\enspace XIV.}

\dropcap{\Usk{16pt}T}{hen} reach me my breeches\break
off the chair, said my father to\break
\stick{\i{Susannah}\tsh There is not a moment’s}
\stick{time to dress you, Sir, cried \i{Susannah}\tsk}
\stick{the child is as black in the face as my\tsh}
\natch{As}
\stick{As your, what? said my father, for like}
\stick{all orators, he was a dear searcher into}
\stick{comparisons\tsk Bless me, Sir, said \i{Susan-}}
\stick{\i{nah}, the child’s in a fit\tsk And where’s}
\stick{Mr.\@ \i{Yorick}\tsk Never where he should be,}
\stick{said \i{Susannah}, but his curate’s in the}
\stick{dressing-room, with the child upon his}
\stick{arm, waiting for the name\tsk and my}
\stick{mistress bid me run as fast as I could to}
\stick{know, as captain \i{Shandy} is the godfather,}
\stick{whether it should not be called after}\break
him.

Were one sure, said my father to him-\break
\stick{self, scratching his eye-brow, that the}
\stick{child was expiring, one might as well}
\stick{compliment my brother \toby as not\tsk}
\stick{and ’twould be a pity, in such a case, to}
\stck{throw away so great a name as \i{Trismegistus}}
upon him\tsk But he may recover.\patch{No,}

No, no,\tsk said my father to \i{Susannah},\break
\stick{I’ll get up\tsk There is no time, cried}
\stick{\i{Susannah}, the child’s as black as my shoe.}
\stick{\i{Trismegistus}, said my father\tsk But stay}
\stick{\tsh thou art a leaky vessel, \i{Susannah}, ad-}
\stick{ded my father; canst thou carry \i{Trisme}-}
\stick{\i{gistus} in thy head, the length of the gal-}
\stick{lery without scattering\tsk Can I? cried}
\stick{\i{Susannah}, shutting the door in a huff\tsk}
\stick{If she can, I’ll be shot, said my father,}
\stick{bouncing out of bed in the dark, and}
groping for his breeches.

\i{Susannah} ran with all speed along the\break
gallery.

My father made all possible speed to\break
find his breeches.

\i{Susannah} got the start, and kept it\tsk\break
\stick{’Tis \i{Tris}\tsk something, cried \i{Susannah}\tsk}
\natch{There}
\stick{There is no christian name in the world,}
\stick{said the curate, beginning with \i{Tris}\tsk}
\stick{but \i{Tristram.} Then ’tis  \i{Tristram-gistus},}
quoth \i{Susannah}.

\tsk There is no \i{gistus} to it, noodle!\tsk\break
\stick{’tis my own name, replied the curate,}
\stick{dipping his hand as he spoke into the}
\stick{bason\tsk \i{Tristram!} said he, \etc \etc \etc}
\stick{\etc so \i{Tristram} was I called, and \i{Tristram}}
shall I be to the day of my death.

My father followed \i{Susannah} with his\break
\stick{night-gown across his arm, with nothing}
\stick{more than his breeches on, fastened}
\stick{through haste with but a single button,}
\stick{and that button through haste thrust}
only half into the button-hole.

\tsh She has not forgot the name, cried\break
\stick{my father, half opening the door\tsk No,}
\natch{no,}
\stick{no, said the curate, with a tone of intel-}
\stick{ligence\tsk And the child is better, cried}
\stick{\i{Susannah}\tsk And how does your mistress?}
\stick{As well, said \i{Susannah}, as can be ex-}
\stick{pected\tsk Pish!\@ said my father, the button}
\stick{of his breeches slipping out of the button-}
\stick{hole\tsk So that whether the interjection}
\stick{was levelled at \i{Susannah}, or the button-}
\stick{hole,\tsk whether pish was an interjection}
\stck{of contempt or an interjection of modesty,}
\stick{is a doubt, and must be a doubt till I}
\stick{shall have time to write the three follow-}
\stick{ing favorite chapters, that is, my chap-}
\stick{ter of \i{chamber-maids}\tsk my chapter of}
\i{pishes}, and my chapter of \i{button-holes}.

All the light I am able to give the\break
\stick{reader at present is this, that the moment}
\stick{my father cried Pish! he whisk’d him-}
\stick{self about\tsk and with his breeches held}
\stick{up by one hand, and his night-gown}
\natch{thrown}
\stick{thrown across the arm of the other, he}
\stick{returned along the gallery to bed, some-}
thing slower than he came.

\section{\chapstrut\lsss{CHAP}.\enspace XV.}

\initial{I}{Wish} I could write a chapter upon\break
sleep.

A fitter occasion could never have pre-\break
\stick{sented itself, than what this moment of-}
\stick{fers, when all the curtains of the family}
\stick{are drawn\tsk the candles put out\tsk and no}
\stick{creature’s eyes are open but a single one,}
\stick{for the other has been shut these twenty}
years, of my mother’s nurse.

It is a fine subject!\patch{And}

And yet, as fine as it is, I would un-\break
\stick{dertake to write a dozen chapters upon}
\stick{button-holes, both quicker and with}
\stick{more fame than a single chapter upon}
this.

Button-holes!\tsh there is something\break
\stick{lively in the very idea of ’em\tsk and trust}
\stick{me, when I get amongst ’em\tsk You}
\stick{gentry with great beards\tsk look as}
\stick{grave as you will\tsk I’ll make merry}
\stick{work with my button-holes\tsk I shall have}
\stick{’em all to myself\tsk ’tis a maiden subject}
\stick{\tsk I shall run foul of no man’s wisdom}
or fine sayings in it.

But for sleep\tsk I know I shall make\break
\stick{nothing of it before I begin\tsk I am no}
\stick{dab at your fine sayings in the first place}
\stick{\tsk and in the next, I cannot for my soul}
\natch{set}
\stick{set a grave face upon a bad matter, and}
\stick{tell the world\tsk ’tis the refuge of the}
\stick{unfortunate\tsk the enfranchisement of the}
\stick{prisoner\tsk the downy lap of the hopeless,}
\stick{the weary and the broken-hearted; nor}
\stick{could I set out with a lye in my mouth,}
\stick{by affirming, that of all the soft and}
\stick{delicious functions of our nature, by}
\stick{which the great Author of it, in his}
\stick{bounty, has been pleased to recompence}
\stick{the sufferings wherewith his justice and}
\stick{his good pleasure has wearied us,\tsh that}
\stick{this is the chiefest (I know pleasures}
\stick{worth ten of it) or what a happiness it is}
\stick{to man, when the anxieties and passions}
\stick{of the day are over, and he lays down}
\stick{upon his back, that his soul shall be so}
\stick{seated within him, that which ever way}
\stick{she turns her eyes, the heavens shall look}
\stick{calm and sweet above her\tsk no desire\tsk}
\stick{or fear\tsk or doubt that troubles the air,}
\natch{nor}
\stick{nor any difficulty pass’d, present, or to}
\stick{come, that the imagination may not}
\stick{pass over without offence, in that sweet}
secession.

\tsk\lqq God’s blessing, said \i{Sancho Panca},\break
\stick{\lqq be upon the man who first invented}
\stick{\lqq this self-same thing called sleep\tsk\!\tsk it}
\stick{\lqq covers a man all over like a cloak.\,”}
\stick{Now there is more to me in this, and it}
\stck{speaks warmer to my heart and affections,}
\stick{than all the dissertations squeez’d out of}
\stick{the heads of the learned together upon}
the subject.

\tsk Not that I altogether disapprove of\break
\stick{what \i{Montaigne} advances upon it\tsk ’tis}
\stick{admirable in its way.\tsk (I quote by}
memory.)\patch{The}

\tight{The world enjoys other pleasures, says}\break
\stick{he, as they do that of sleep, without}
\stick{tasting or feeling it as it slips and passes}
\stick{by\tsk We should study and ruminate up-}
\stick{on it, in order to render proper thanks}
\stick{to him who grants it to us\tsk for this end}
\stick{I cause myself to be disturbed in my}
\stick{sleep, that I may the better and more}
\stick{sensibly relish it\tsk And yet I see few,}
\stick{says he again, who live with less sleep}
\stck{when need requires; my body is capable}
\stick{of a firm, but not of a violent and sud-}
\stick{den agitation\tsk I evade of late all violent}
\stick{exercises\tsk I am never weary with walk-}
\stick{ing\tsk but from my youth, I never liked}
\stick{to ride upon pavements. I love to lie}
\stick{hard and alone, and even without my}
\stick{wife\tsk This last word may stagger the}
\stick{faith of the world\tsk but remember, \lqq La}
\stick{\lqq Vraisemblance (\,as \i{Baylet} says in the}
\stick{\lqq affair of \i{Liceti}\,\i{)}\sic n’est pas toujours}
\natch{\lqq du}
\stick{\lqq du Cotè de la Verité.\,”\quad And so much}
for sleep.

\section{\chapstrut\lsss{CHAP}.\enspace XVI.}

\dropcap{I}{f} my wife will but venture him\tsk\break
brother \toby, \i{Trismegistus} shall be\break
\stick{dress’d and brought down to us, whilst}
\stick{you and I are getting our breakfasts}
together.\tsk

\tsk Go, tell \i{Susannah}, \i{Obadiah}, to step\break
here.

\tight{She is run up stairs, answered \i{Obadiah},}\break
\stick{this very instant, sobbing and crying,}
\stick{and wringing her hands as if her heart}
would break.\tsh\patch{We}

We shall have a rare month of it,\break
\stick{said my father, turning his head from}
\stick{\i{Obadiah}, and looking wistfully in my}
\stck{uncle \toby's face for some time\tsk we shall}
\stick{have a devilish month of it, brother \toby,}
\stick{said my father, setting his arms a-kimbo,}
\stick{and shaking his head; fire, water, wo-}
\stick{men, wind\tsk brother \i{Toby!}\tsk ’Tis some}
\stick{misfortune, quoth my uncle \i{Toby}\tsk That}
\stick{it is, cried my father,\tsk to have so many}
\stick{jarring elements breaking loose, and}
\stick{riding triumph in every corner of a gen-}
\stick{tleman’s house\tsk Little boots it to the}
\stick{peace of a family, brother \toby, that you}
\stck{and I possess ourselves, and sit here silent}
\stick{and unmoved,\tsk whilst such a storm is}
whistling over our heads.\tsh

\tsk And what’s the matter, \i{Susannah?}\break
\stick{They have called the child \i{Tristram}\tsh}
\stick{and my mistress is just got out of an}
\natch{hysterick}
\stick{hysterick fit about it\tsk No!\tsk ’tis not my}
\stick{fault, said \i{Susannah}\tsk I told him it was}
\i{Tristram-gistus}.

\tsh Make tea for yourself, brother\break
\stick{\toby, said my father, taking down his}
\stick{hat\tsk but how different from the sallies}
\stick{and agitations of voice and members}
which a common reader would imagine!

\tsk For he spake in the sweetest modu-\break
\stick{lation\tsk and took down his hat with the}
\stick{gentlest movement of limbs, that ever}
\stick{affliction harmonized and attuned toge-}
ther.

\tsk Go to the bowling-green for corpo-\break
\stick{ral \i{Trim}, said my uncle \toby, speaking}
\stick{to \i{Obadiah}, as soon as my father left the}
room.\patch{\ls{CHAP}.}

\section{\chapstrutt\lsss{CHAP}.\enspace XVII.}

\dropcap{W}{hen} the misfortune of my\break
\s{nose} fell so heavily upon my\break
\stick{father’s head,\tsk the reader remembers}
\stick{that he walked instantly up stairs, and}
\stick{cast himself down upon his bed; and}
\stick{from hence, unless he has a great insight}
\stick{into human nature, he will be apt to}
\stick{expect a rotation of the same ascending}
\stick{and descending movements from him,}
\stick{upon this misfortune of my \s{name};\tsh}\break
no.

The different weight, dear Sir\tsh nay\break
\stick{even the different package of two vexa-}
\stick{tions of the same weight,\tsh makes a very}
\stick{wide difference in our manner of bearing}
\stick{and getting through with them.\tsk It is}
\natch{not}
\stick{not half an hour ago, when (in the great}
\stick{hurry and precipitation of a poor devil’s}
\stick{writing for daily bread) I threw a fair}
\stick{sheet, which I had just finished, and care-}
\stick{fully wrote out, slap into the fire, instead}
of the foul one.

Instantly I snatch’d off my wig, and\break
\stick{threw it perpendicularly, with all imagi-}
\stick{nable violence, up to the top of the room}
\stick{\tsk indeed I caught it as it fell\tsk but there}
\stick{was an end of the matter; nor do I}
\stick{think any think else in \i{Nature}, would}
\stick{have given such immediate ease: She,}
\stick{dear Goddess, by an instantaneous im-}
\stick{pulse, in all \i{provoking cases}, determines}
\stick{us to a sally of this or that member\tsk or}
\stick{else she thrusts us into this or that place,}
\stick{or posture of body, we know not why\tsk}
\stick{But mark, madam, we live amongst}
\stick{riddles and mysteries\tsk the most obvious}
\natch{things,}
\stick{things, which come in our way, have}
\stick{dark sides, which the quickest sight can-}
\stick{not penetrate into; and even the clearest}
\stck{and most exalted understandings amongst}
\stick{us find ourselves puzzled and at a loss in}
\stick{almost every cranny of nature’s works;}
\stick{so that this, like a thousand other things,}
\stick{falls out for us in a way, which tho’ we}
\stick{cannot reason upon it,\tsk yet we find the}
\stick{good of it, may it please your reverences}
\stick{and your worships\tsh and that’s enough}
for us.

Now, my father could not lie down\break
\stick{with this affliction for his life\tsk nor could}
\stick{he carry it up stairs like the other\tsk He}
\stick{walked composedly out with it to the}
fish-pond.

Had my father leaned his head upon\break
\stick{his hand, and reasoned an hour which}
\natch{way}
\stick{way to have gone\tsk reason, with all her}
\stick{force, could not have directed him to}
\stick{any think like it: there is something,}
\stick{Sir, in fish-ponds\tsk but what it is, I leave}
\stck{to system builders and fish pond diggers}
\stick{betwixt ’em to find out\tsk but there is}
\stick{something, under the first disorderly}
\stick{transport of the humours, so unaccount-}
\stick{ably becalming in an orderly and a sober}
\stick{walk towards one of them, that I have}
\stick{often wondered that neither \i{Pythagoras},}
\stick{nor \i{Plato}, nor \i{Solon}, nor \i{Licurgus}, nor}
\stick{\i{Mahomet}, nor any one of your noted law-}
givers, ever gave order about them.

\section{\chapstrut\lsss{CHAP}.\enspace XVIII.}

\dropcap{Y}{our} honour, said \i{Trim}, shutting\break
the parlour door before he began\break
\stick{to speak, has heard, I imagine, of this}
\natch{unlucky}
\stick{unlucky accident\tsk\!\tsk O yes, \i{Trim!} said}
\stick{my uncle \toby, and it gives me great}
\stick{concern\tsk I am heartily concerned too,}
\stick{but I hope your honour, replied \i{Trim},}
\stick{will do me the justice to believe, that it}
\stick{was not in the least owing to me\tsk To}
\stick{thee\tsk \i{Trim!}\tsk cried my uncle \toby, look-}
\stick{ing kindly in his face\tsk ’twas \i{Susannah}’s}
\stick{and the curate’s folly betwixt them\tsk}
\stick{What business could they have together,}
\stick{an’ please your honour, in the garden?\tsk}
\stick{In the gallery, thou meanest, replied}
my uncle \toby.

Trim found he was upon a wrong\break
\stick{scent, and stopped short with a low bow}
\stick{\tsk Two misfortunes, quoth the corporal}
\stick{to himself, are twice as many at least as}
\stick{are needful to be talked over at one time,}
\stick{\tsk the mischief the cow has done in}
\stick{breaking into the fortifications, may be}
\natch{told}
\stck{told his honour hereafter\tsk \i{Trim}’s casuistry}
\stick{and address, under the cover of his low}
\stick{bow, prevented all suspicion in my uncle}
\stick{\toby, so he went on with what he had}
to say to \i{Trim} as follows.

\tsk For my own part, \i{Trim}, though I\break
\stick{can see little or no difference betwixt my}
\stick{nephew’s being called \i{Tristram} or \i{Trisme}-}
\stick{\i{gistus}\tsk yet as the thing sits so near my}
\stick{brother’s heart, \i{Trim},\tsk I would freely}
\stick{have given a hundred pounds rather}
\stick{than it should have happened\tsk A hun-}
\stick{dred pounds, an’ please your honour,  re-}
\stick{plied \i{Trim},\tsk I would not give a cherry-}
\stick{stone to boot\tsk Nor would I, \i{Trim}, upon}
\stick{my own account, quoth my uncle \i{Toby}\tsk}
\stck{but my brother, whom there is no arguing}
\stick{with in this case\tsk maintains that a great}
\stick{deal more depends, \i{Trim}, upon christian}
\stick{names, than what ignorant people ima-}
\natch{gine;}
\stick{gine;\tsh for he says there never was a}
\stick{great or heroic action performed since}
\stick{the world began by one called \i{Tristram}}
\stick{\tsk nay he will have it, \i{Trim}, that a man}
\stick{can neither be learned, or wise, or brave}
\stick{\tsk ’Tis all a fancy, an’ please your honour}
\stick{\tsk I fought just as well, replied the cor-}
\stick{poral, when the regiment called me \i{Trim},}
\stick{as when they called me \i{James Butler}\tsh}
\stick{And for my own part, said my uncle}
\stick{\toby, though I should blush to boast of}
\stick{myself, \i{Trim},\tsk yet had my name been}
\stick{\i{Alexander}, I could have done no more}
\stick{at \i{Namur} than my duty\tsk Bless your ho-}
\stick{nour! cried \i{Trim}, advancing three steps}
\stick{as he spoke, does a man think of his}
\stick{christian name when he goes upon the}
\stck{attack?\tsk Or when he stands in the trench,}
\stick{\i{Trim?} cried my uncle \toby, looking}
\stick{firm\tsk Or when he enters a breach? said}
\stick{\i{Trim}, pushing in between two chairs\tsk}
\natch{Or\sic}
\stick{\tsk Or forces the lines? cried my uncle,}
\stick{rising up, and pushing his crutch like a}
\stick{pike\tsk Or facing a platoon, cried \i{Trim},}
\stick{presenting his stick like a firelock\tsk Or}
\stick{when he marches up the glacis, cried my}
\stick{uncle \toby, looking warm and setting}
his foot upon his stool.\tsh


\section{\chapstrut\lsss{CHAP}.\enspace XIX.}

\dropcap{M}{y} father was returned from his\break
walk to the fish-pond\tsk and open-\break
\stick{ed the parlour-door in the very height of}
\stick{the attack, just as my uncle \toby was}
\stick{marching up the glacis\tsk \i{Trim} recovered}
\stick{his arms\tsk never was my uncle \toby}
\stick{caught riding at such a desperate rate in}
\stick{his life! Alas! my uncle \i{Toby!} had not}
\stick{a weightier matter called forth all the}
\stick{ready eloquence of my father\tsk how hadst}
\natch{thou}
\stick{thou then and thy poor \s{Hobby-Horse}}
too have been insulted!

My father hung up his hat with the\break
\stck{same air he took it down; and after giving}
\stick{a slight look at the disorder of the room,}
\stick{he took hold of one of the chairs which}
\stick{had formed the corporal’s breach, and}
\stick{placing it over-against my uncle \toby,}
\stick{he sat down in it, and as soon as the tea-}
\stick{things were taken away and the door}
\stick{shut, he broke out in a lamentation as}
follows.

\section{\textsc{My\enspace Father's\enspace\lss{Lamentation}}.}

\dropcap{I}{t} is in vain longer, said my father, ad-\break
\tighter{dressing himself as much to \i{Ernulphus}’s}\break
\stick{curse, which was laid upon the corner of}
\stick{the chimney-piece,\tsk as to my uncle \toby}
\stick{who sat under it\tsk it is in vain longer,}
\natch{said}
\stick{said my father, in the most querulous}
\stick{monotony imaginable, to struggle as I}
\stick{have done against this most uncomfort-}
\stick{able of human persuasions\tsk I see it}
\stick{plainly, that either for my own sins,}
\stick{brother \toby, or the sins and follies of}
\stick{the \i{Shandy}-family, heaven has thought}
\stick{fit to draw forth the heaviest of its artil-}
\stick{lery against me; and that the prosperity}
\stick{of my child is the point upon which the}
\stick{whole force of it is directed to play\tsh}
\stick{Such a thing would batter the whole}
\stick{universe about our ears, brother \i{Shandy},}
\stick{said my uncle \toby,\tsk if it was so\tsk Un-}
\stick{happy \i{Tristram!} child of wrath! child}
\stick{of decrepitude!  interruption! mistake!}
\stick{and discontent! What one misfortune or}
\stick{disaster in the book of embryotic evils,}
\stick{that could unmechanize thy frame, or}
\stick{entangle thy filaments!  which has not}
\stick{fallen upon thy head, or ever thou camest}
\natch{into}
\stick{into the world\tsk what evils in thy passage}
\stick{into it!\tsk What evils since!\tsk produced}
\stick{into being, in the decline of thy father’s}
\stck{days\tsk when the powers of his imagination}
\stick{and of his body were waxing feeble\tsk}
\stick{when radical heat and radical moisture,}
\stck{the elements which should have temper’d}
\stick{thine, were drying up; and nothing left}
\stick{to found thy stamina in, but negations\tsk}
\stick{\tsk ’tis pitiful\tsk brother \toby, at the best,}
\stick{and called out for all the little helps}
\stick{that care and attention on both sides}
\stick{could give it. But how were we de-}
\stick{feated! You know the event, brother}
\stick{\toby,\tsk ’tis too melancholy a one to be}
\stick{repeated now,\tsk when the few animal}
\stick{spirits I was worth in the world, and}
\stick{with which memory, fancy, and quick}
\stick{parts should have been convey’d,\tsk were}
\stck{all dispersed, confused, confounded, scat-}
\stick{tered, and sent to the devil.\tsk\hfill}
\natch{Here}

Here then was the time to have put a\break
\stick{stop to this persecution against him;\tsh}
\stck{and tried an experiment at least\tsk whether}
\stick{calmness and serenity of mind in your}
\stick{sister, with a due attention, brother \toby,}
\stick{to her evacuations and repletions\tsk and}
\stick{the rest of her non-naturals, might not,}
\stick{in a course of nine months gestation,}
\stick{have set all things to rights.\tsk My child}
\stick{was bereft of these!\tsk What a teazing}
\stck{life did she lead herself, and consequently}
\stick{her fœtus too, with that nonsensical}
\stick{anxiety of hers about lying in in town?}
\stick{I thought my sister submitted with the}
\stick{greatest patience, replied my uncle \toby}
\stick{\tsk I never heard her utter one fretful}
\stick{word about it\tsk She fumed inwardly, cried}
\stick{my father; and that, let me tell you,}
\stick{brother, was ten times worse for the}
\stick{child\tsk and then!\@ what battles did she}
\stick{fight with me, and what perpetual storms}
\natch{about}
\stick{about the midwife\tsk There she gave vent,}
\stick{said my uncle \i{Toby}\tsk Vent!\@ cried my fa-}
ther, looking up\tsk

But what was all this, my dear \toby,\break
\stick{to the injuries done us by my child’s}
\stick{coming head foremost into the world,}
\stick{when all I wished in this general wreck}
\stick{of his frame, was to have saved this}
little casket unbroke, unrifled\tsk

With all my precautions, how was\break
\stick{my system turned topside turvy in the}
\stick{womb with my child! his head exposed}
\stick{to the hand of violence, and a pressure}
\stick{of 470 pounds averdupois weight acting}
\stick{so perpendicularly upon its apex\tsk that}
\stck{at this hour ’tis ninety \i{per Cent}. insurance,}
\stick{that the fine network of the intellectual}
\stick{web be not rent and torn to a thousand}
tatters.\patch{\tsk Still}

\tsk Still we could have done.\tsh Fool,\break
\stick{coxcomb, puppy\tsk give him but a \s{nose}}
\stick{\tsk Cripple, Dwarf, Driveller, Goosecap}
\stick{\tsk (shape him as you will) the door of}
\stick{Fortune stands open\tsk O \i{Licetus! Licetus!}}
\stick{had I been blest with a fœtus five inches}
\stick{long and a half, like thee\tsk fate might}
have done her worst.

Still, brother \toby, there was one cast\break
\stick{of the dye left for our child after all\tsk}
O \i{Tristram! Tristram!  Tristram!}

We will send for Mr.\@ \i{Yorick}, said my\break
uncle \toby.

\tsk You may send for whom you will,\break
replied my father.\patch{\lss{CHAP}.}

\section{\chapstrutt\lsss{CHAP}.\enspace XX.}

\dropcap{W}{hat} a rate have I gone on at,\break
curvetting and striking it away,\break
\stick{two up and two down for four volumes}
\stick{together, without looking once behind,}
\stick{or even on one side of me, to see whom}
\stick{I trod upon!\tsk I’ll tread upon no one,\tsk}
\stick{quoth I to myself when I mounted\tsk I’ll}
\stick{take a good rattling gallop; but I’ll not}
\stck{hurt the poorest jack-ass upon the road\tsk}
\stick{So off I set\tsk up one lane\tsk down another,}
\stick{through this turn-pike\tsk over that, as if}
\stick{the arch-jockey of jockeys had got be-}
hind me.

Now ride at this rate with what good\break
\stick{intention and resolution you may,\tsk ’tis}
\stick{a million to one you’ll do some one a}
\natch{mischief,}
\stick{mischief, if not yourself\tsk He’s flung\tsk}
\stick{he’s off\tsk he’s lost his seat\tsk he’s down\tsk}
\stick{he’ll break his neck\tsk see!\tsk if he has not}
\stick{galloped full amongst the scaffolding of}
\stick{the undertaking criticks!\tsk he’ll knock}
\stick{his brains out against some of their posts}
\stick{\tsk he’s bounced out!\tsk look\tsk he’s now}
\stick{riding like a madcap full tilt through a}
\stick{whole crowd of painters, fiddlers, poets,}
\stick{biographers, physicians, lawyers, logi-}
\stick{cians, players, schoolmen, churchmen,}
\stck{statesmen, soldiers, casuists, connoisseurs,}
\stick{prelates, popes, and engineers\tsk Don’t}
\stick{fear, said I\tsk I’ll not hurt the poorest}
\stick{jack-ass upon the king’s high-way\tsk But}
\stk{your horse throws dirt; see you’ve splash’d}
\stick{a bishop\tsk I hope in God, ’twas only}
\stick{\i{Ernulphus}, said I\tsk But you have squirted}
\stick{full in the faces of Mess.\@ \i{Le Moyne}, \i{De}}
\stick{\i{Romigny}, and \i{De Marcilly}, doctors of}
\stick{the \i{Sorbonne}\tsk That was last year, re-}
\natch{plied}
\stick{plied I\tsk But you have trod this moment}
\stick{upon a king.\tsh Kings have bad times}
\stick{on’t, said I, to be trod upon by such}
people as me.

\tsk You have done it, replied my ac-\break
cuser.

I deny it, quoth I, and so have got off,\break
\stick{and here am I standing with my bridle}
\stick{in one hand, and with my cap in the}
\stick{other, to tell my story\tsh And what is it?}
You shall hear in the next chapter.


\section{\chapstrut\lsss{CHAP}.\enspace XXI.}

\dropcap[findent=0pt]{A}{s} \i{Francis} the first of \i{France} was\break
one winterly night warming him-\break
\stick{self over the embers of a wood fire, and}
\stick{talking with his firft\sic minister of sundry}
\natch{things}
\stick{things for the good of the state\tsk it\sic}
\stick{would not be amiss, said the king, stir-}
\stick{ring up the embers with his cane, if this}
\stick{good understanding betwixt ourselves}
\stick{and \i{Switzerland} was a little strengthened}
\stick{\tsk There is no end, Sire, replied the mi-}
\stick{nister, in giving money to these peo-}
\stick{ple\tsk they would swallow up the trea-}
\stick{sury of \i{France}\tsk Poo!\@ poo!\@  answered the}
\stick{king\tsk\!\tsk there are more ways, Mons.\@ \i{le}}
\stick{\i{Premier}, of bribing states, besides that}
\stick{of giving money\tsk\!\tsk I’ll pay \i{Switzer}-}
\stick{\i{land} the honour of standing godfather}
\stick{for my next child\tsk Your majesty, said}
\stick{the minister, in so doing, would have all}
\stick{the grammarians in \i{Europe} upon your}
\stick{back;\tsk \i{Switzerland}, as a republick, be-}
\stick{ing a female, can in no construction be}
\stick{godfather\tsk She may be godmother, re-}
\stick{plied \i{Francis}, hastily\tsk so announce my}
\stick{\hfill intentions}\etp\break
\stick{\footnotesize\qquad\fnast\ Vide Menagiana, vol. 1.\hfill}

\newpage
\noindent
intentions by a courier to morrow morn-\break
ing.

I am astonished, said \i{Francis} the First,
\stick{(that day fortnight) speaking to his mi-}
\stick{nister as he entered the closet, that we}
\stick{have had no answer from \i{Switzerland}\tsk}
\stick{Sire, I wait upon you this moment, said}
\stick{Mons.\@ \i{le Premier}, to lay before you my}
\stick{dispatches upon that business.\tsk They}
\stick{take it kindly? said the king\tsk They do,}
\stick{Sire, replied the minister, and have the}
\stick{highest sense of the honour your majesty}
\stick{has done them\tsk but the republick, as}
\stick{godmother, claims her right in this case,}
of naming the child.

In all reason, quoth the king\tsk she will\break
\stick{christen him \i{Francis}, or \i{Henry}, or \i{Lewis},}
\stick{or some name that she knows will be}
\stick{agreeable to us. Your majesty is de-}
\natch{ceived,}
\stick{ceived, replied the minister\tsk I have this}
\stick{hour received a dispatch from our resi-}
\stick{dent, with the determination of the re-}
\stick{publick on that point also\tsk And what}
\stick{name has the republick fixed upon for}
\stick{the Dauphin?\tsk \i{Shadrach}, \i{Mesech}, and}
\stick{\i{Abed-nego}, replied the minister\tsk By saint}
\stick{\i{Peter}’s girdle, I will have nothing to do}
\stick{with the \i{Swiss}, cried \i{Francis} the First,}
\stick{pulling up his breeches and walking}
hastily across the floor.

Your majesty, replied the minister\break
calmly, cannot bring yourself off.

We’ll pay them in money\tsk said the\break
king.

Sire, there are not sixty thousand\break
\stick{crowns in the treasury, answered the}
\stick{minister\tsh I’ll pawn the best jewel}
\natch{in}
\stick{in my crown, quoth \i{Francis} the}
First.

Your honour stands pawn’d already\break
\stick{in this matter, answered Monsieur \i{le}}
\i{Premier}.

Then, Mons.\@ \i{le Premier}, said the king,\break
by\tsh we’ll go to war with ’em.


\section{\chapstrut\lsss{CHAP}.\enspace XXII.}

\dropcap[findent=2pt]{A}{lbeit}, gentle reader, I have\break
lusted earnestly, and endeavoured\break
\stick{carefully (according to the measure of}
\stick{such slender skill as God has vouchsafed}
\stick{me, and as convenient leisure from other}
\stick{occasions of needful profit and healthful}
\stick{pastime have permitted) that these little}
\stick{books, which I here put into thy hands,}
\natch{might}
\stck{might stand instead of many bigger books}
\stick{\tsk yet have I carried myself towards thee}
\stick{in such fanciful guise of careless disport,}
\stick{that right sore am I ashamed now to en-}
\stick{\lower1.618pt\hbox{t}reat\sic thy lenity seriously\tsk in beseeching}
\stick{thee to believe it of me, that in the}
\stick{story of my father and his christen-}
\stick{names,\tsk I had no thoughts of treading}
\stck{upon \i{Francis} the First\tsk nor in the affair of}
\stck{the nose\tsk upon \i{Francis} the Ninth\tsk nor in}
\stick{the character of my uncle \i{Toby}\tsk of cha-}
\stick{racterizing the militiating spirits of my}
\stick{country\tsk the wound upon his groin, is}
\stick{a wound to every comparison of that}
\stick{kind,\tsk nor by \i{Trim},\tsk that I meant the}
\stick{duke of \i{Ormond}\tsk or that my book is}
\stick{wrote against predestination, or free will,}
\stick{or taxes\tsk If ’tis wrote against any thing,}
\stick{\tsh ’tis wrote, an’ please your worships,}
\stick{against the spleen; in order, by a more}
\natch{frequent}
\stick{frequent and a more convulsive elevation}
\stick{and depression of the diaphragm, and}
\stick{the succussations of the intercostal and}
\stick{abdominal muscles in laughter, to drive}
\stick{the \i{gall} and other \i{bitter juices} from the}
\stick{gall bladder, liver, and sweet-bread of}
\stick{his majesty’s subjects, with all the inimi-}
\stick{citious passions which belong to them,}
down into their duodenums.



\section{\chapstrut\lsss{CHAP}.\enspace XXIII.}

\dropcap{\Usk{16pt}B}{ut} can the thing be undone,\break
\i{Yorick?} said my father\tsk for in\break
\stick{my opinion, continued he, it cannot. I}
\stick{am a vile canonist, replied \i{Yorick}\tsk but}
\stick{of all evils, holding suspense to be the}
\stick{most tormenting, we shall at least know}
\stick{the worst of this matter. I hate these}
\stick{great dinners\tsk said my father\tsk The size}
\natch{of}
\stick{of the dinner is not the point, answered}
\stick{\i{Yorick}\tsk we want, Mr.\@ \i{Shandy}, to dive}
\stick{into the bottom of this doubt, whether}
\stick{the name can be changed or not\tsk and as}
\stick{the beards of so many commissaries,}
\stick{officials, advocates, proctors, registers,}
\stick{and of the most able of our school-}
\stick{divines, and others, are all to meet in the}
\stick{middle of one table, and \i{Didius} has so}
\stick{pressingly invited you,\tsk who in your}
\stck{distress would miss such an occasion? All}
\stick{that is requisite, continued \i{Yorick}, is to}
\stick{apprize \i{Didius}, and let him manage a con-}
\stick{versation after dinner so as to introduce}
\stick{the subject\tsk Then my brother \toby, cried}
\stick{my father, clapping his two hands toge-}
ther, shall go with us.\patch{\tsk Let}

\tsk Let my old tye wig, quoth my uncle\break
\stick{\toby, and my laced regimentals, be hung}
to the fire all night, \i{Trim.}

\vfill
\rightline{\ls{CHAP}.}
\vfill
\eject
\pagestyle{plusnine}  % so that we effectively skip to p.156
\section{\chapstrutt\lsss{CHAP}.\enspace XXV.}

\dropcap{\Usk{16pt}N}{o} doubt, Sir\tsk there is a whole\break
chapter wanting here\tsk and a\break
\stick{chasm of ten pages made in the book}
\stick{by it\tsk but the book-binder is neither a}
\stick{fool, or a knave, or a puppy\tsk nor is the}
\stick{book a jot more imperfect, (at least upon}
\stick{that score)\tsk but, on the contrary, the}
\stick{book is more perfect and complete by}
\stick{wanting the chapter, than having it, as}
\stick{I shall demonstrate to your reverences in}
\stick{this manner\tsk I question first by the bye,}
\stick{whether the same experiment might not}
\stick{be made as successfully upon sundry}
\stick{other chapters\tsh but there is no end,}
\stick{an’ please your reverences, in trying ex-}
\stick{periments upon chapters\tsh we have had}
\natch{enough}
\stick{enough of it\tsh So there’s an end of that}
matter.

But before I begin my demonstration,\break
\stick{let me only tell you, that the chapter}
\stick{which I have torn out, and which other-}
\stick{wise you would all have been reading}
\stick{just now, instead of this,\tsk was the de-}
\stick{scription of my father’s, my uncle \toby's,}
\stick{\i{Trim}’s, and \i{Obadiah}’s setting out and}
journeying to the visitation at \raise3pt\hbox{\astiv}.

We’ll go in the coach, said my father\break
\stick{\tsk Prithee, have the arms been altered,}
\stick{\i{Obadiah}?\tsk It would have made my story}
\stick{much better, to have begun with telling}
\stick{you, that at the time my mother’s arms}
\stick{were added to the \i{Shandy}’s, when the}
\stick{coach was repainted upon my father’s}
\stick{marriage, it had so fallen out, that the}
\stick{coach painter, whether by performing}
\natch{all}
\stick{all his works with the left hand, like}
\stick{\i{Turpilius} the \i{Roman}, or \i{Hans Holbein} of}
\stick{\i{Basil}\tsk or whether ’twas more from the}
\stick{blunder of his head than hand\tsk or whe-}
\stick{ther, lastly, it was from the sinister turn,}
\stick{which every thing relating to our family}
\stick{was apt to take\tsk It so fell out, however,}
\stick{to our reproach, that instead of the \i{bend}}
\stick{\i{dexter}, which since \i{Harry} the Eighth’s}
\stick{reign was honestly our due\tsk a \i{bend}}
\stick{\i{sinister}, by some of these fatalities, had}
\stick{been drawn quite across the field of the}
\stick{\i{Shandy}-arms.  ’Tis scarce credible that}
\stick{the mind of so wise a man as my father}
\stick{was, could be so much incommoded with}
\stick{so small a matter. The word coach\tsk let}
\stick{it be whose it would\tsk or coach-man, or}
\stick{coach-horse, or coach-hire, could never}
\stick{be named in the family, but he constantly}
\stick{complained of carrying this vile mark of}
\stick{illegitimacy upon the door of his own;}
\natch{he}
\stick{he never once was able to step into the}
\stick{coach, or out of it, without turning}
\stick{round to take a view of the arms, and}
\stick{making a vow at the same time, that it}
\stick{was the last time he would ever set his}
\stick{foot in it again, till the \i{bend-sinister} was}
\stick{taken out\tsk but like the affair of the}
\stick{hinge, it was one of the many things}
\stick{which the \i{Destinies} had set down in their}
\stick{books\tsk ever to be grumbled at (and in}
\stick{wiser families than ours)\tsk but never to}
be mended.

\tsk Has the \i{bend-sinister} been brush’d\break
\stick{out, I say? said my father\tsk There has}
\stick{been nothing brush’d out, Sir, answered}
\stick{\i{Obadiah}, but the lining. We’ll go}
\stick{o’horse-back, said my father, turning to}
\stick{\i{Yorick}\tsk Of all things in the world, ex-}
\stick{cept politicks, the clergy know the least}
\stick{of heraldry, said \i{Yorick.}\tsk No matter for}
\natch{that,}
\stick{that, cried my father\tsk I should be sorry}
\stick{to appear with a blot in my escutcheon}
\stick{before them\tsh Never mind the \i{bend}-}
\stick{\i{sinister}, said my uncle \toby, putting on}
\stick{his tye-wig\tsh No, indeed, said my father,}
\stick{\tsk you may go with my aunt \i{Dinah} to a}
\stick{visitation with a \i{bend-sinister}, if you think}
\stick{fit\tsk My poor uncle \toby blush’d. My}
\stick{father was vexed at himself\tsk No\tsk my}
\stick{dear brother \toby, said my father,}
\stick{changing his tone\tsk but the damp of the}
\stick{coach-lining about my loins, may give}
\stick{me the Sciatica again, as it did \i{December},}
\stick{\i{January}, and \i{February} last winter\tsk so if}
\stick{you please you shall ride my wife’s pad}
\stick{\tsk and as you are to preach, \i{Yorick}, you}
\stick{had better make the best of your way}
\stick{before,\tsk and leave me to take care of}
\stick{my brother \toby, and to follow at our}
own rates.\patch{Now}

Now the chapter I was obliged to tear\break
\stick{out, was the description of this cavalcade,}
\stick{in which corporal \i{Trim} and \i{Obadiah},}
\stick{upon two coach-horses a-breast, led the}
\stick{way as slow as a patrole\tsk whilst my}
\stick{uncle \toby, in his laced regimentals and}
\stick{tye-wig, kept his rank with my father,}
\stick{\stck{in deep roads and dissertations alternately}}
\stick{upon the advantage of learning and arms,}
as each could get the start.


\tsk But the painting of this journey,\break
\stick{upon reviewing it, appears to be so much}
\stick{above the stile and manner of any thing}
\stick{else I have been able to paint in this}
\stick{book, that it could not have remained}
\stick{in it, without depreciating every other}
\stick{scene; and destroying at the same time}
\stick{that necessary equipoise and balance,}
\stick{(whether of good or bad) betwixt chap-}
\stick{ter and chapter, from whence the just}
\natch{proportions}
\stick{proportions and harmony of the whole}
\stick{work results. For my own part, I am}
\stick{but just set up in the business, so know}
\stick{little about it\tsk but, in my opinion, to}
\stick{write a book is for all the world like}
\stick{humming a song\tsk be but in tune with}
\stick{yourself, madam, ’tis no matter how}
high or how low you take it.\tsk


\tsk This is the reason, may it please\break
\stick{your reverences, that some of the lowest}
\stick{and flattest compositions pass off very}
\stick{well\tsk (as \i{Yorick} told my uncle \toby one}
\stick{night) by siege\tsk My uncle \toby looked}
\stick{brisk at the sound of the word \i{siege}, but}
could make neither head or tail of it.

I’m to preach at court next Sunday,\break
\stick{said \i{Homenas}\tsk run over my notes\tsk so I}
\stick{humm’d over doctor \i{Homenas}’s notes\tsk}
\stick{the modulation’s very well\tsk ’twill do,}
\natch{\i{Homenas},}
\stick{\i{Homenas}, if it holds on at this rate\tsk so}
\stick{on I humm’d\tsk and a tolerable tune I}
\stick{thought it was; and to this hour, may}
\stick{it please your reverences, had never}
\stick{found out how low, how flat, how spi-}
\stick{ritless and jejune it was, but that all of}
\stick{a sudden, up started an air in the middle}
\stick{of it, so fine, so rich, so heavenly\tsk it car-}
\stick{ried my soul up with it into the other}
\stick{world; now had I, (as \i{Montaigne} com-}
\stick{plained in a parallel accident)\tsk had I}
\stick{found the declivity easy, or the ascent}
\stick{accessible\tsk certes I had been outwitted}
\stick{\tsk Your notes, \i{Homenas}, I should have}
\stick{said, are good notes,\tsk but it was so per-}
\stick{pendicular a precipice\tsk so wholly cut off}
\stick{from the rest of the work, that by the}
\stick{first note I humm’d, I found myself}
\stick{flying into the other world, and from}
\stick{thence discovered the vale from whence}
\stick{I came, so deep, so low, and dismal,}
\natch{that}
\stick{that I shall never have the heart to de-}
scend into it again.

\indent
\fist A dwarf who brings a standard\break
\stick{along with him to measure his own size}
\stick{\tsk take my word, is a dwarf in more ar-}
\stick{ticles than one\tsk And so much for tear-}
ing out of chapters.

\bigskip

\section{\lsss{CHAP}.\enspace XXVI.}

\dropcap[findent=0pt]{\Usk{12pt}S}{ee}  if he is not cutting it all into\break
slips, and giving them about him\break
\stick{to light their pipes!\tsk ’Tis abominable,}
\stick{answered \i{Didius}; it should not go unno-}
\stick{ticed, said doctor \i{Kysarcius}\tsk \fist he was}
of  the \i{Kysarcij} of the low countries.

Methinks, said \i{Didius}, half rising\break
\stick{from his chair, in order to remove a}
\natch{bottle}
\stick{bottle and a tall decanter, which stood}
\stick{in a direct line betwixt him and \i{Yorick}}
\stck{\tsk you might have spared this sarcastick}
\stick{stroke, and have hit upon a more proper}
\stick{place, Mr.\@ \i{Yorick}\tsk or at least upon a}
\stck{more proper occasion to have shewn your}
\stick{contempt of what we have been about:}
\stick{If the Sermon is of no better worth than}
\stick{to light pipes with\tsk ’twas certainly, Sir,}
\stick{not good enough to be preached before}
\stick{so learned a body; and if ’twas good}
\stick{enough to be preached before so learned}
\stick{a body\tsk ’twas certainly, Sir, too good}
to light their pipes with afterwards.


\tsk I have got him fast hung up, quoth\break
\stick{\i{Didius} to himself, upon one of the two}
\stick{horns of my dilemma\tsk let him get off}
as he can.\patch{I have}

I have undergone such unspeakable\break
\stick{torments, in bringing forth this sermon,}
\stick{quoth \i{Yorick}, upon this occasion,\tsk that}
\stick{I declare, \i{Didius}, I would suffer martyr-}
\stick{dom\tsk and if it was possible my horse}
\stick{with me, a thousand times over, before I}
\stick{would sit down and make such another:}
\stick{I was delivered of it at the wrong end of}
\stick{me\tsk it came from my head instead of}
\stick{my heart\tsk and it is for the pain it gave}
\stick{me, both in the writing and preaching}
\stick{of it, that I revenge myself of it, in}
\stick{this manner.\tsk To preach, to shew the}
\stick{extent of our reading, or the subtleties}
\stick{of our wit\tsk to parade it in the eyes of}
\stick{the vulgar with the beggarly accounts}
\stick{of a little learning, tinseled over with a}
\stick{few words which glitter, but convey little}
\stick{light and less warmth\tsk is a dishonest use}
\stk{of the poor single half hour in a week which}
\stick{is put into our hands\tsk ’Tis not preaching}
\natch{the}
\stick{the gospel\tsk but ourselves\tsk For my own}
\stick{part, continued \i{Yorick}, I had rather}
\stick{direct five words point blank to the}
heart\tsk

As \i{Yorick} pronounced the word \i{point}\break
\stick{\i{blank}, my uncle \toby rose up to say}
\stick{something upon projectiles\tsh when a}
\stick{single word, and no more, uttered from}
\stick{the opposite side of the table drew}
\stick{every one’s ears towards it\tsk a word of}
\stick{all others in the dictionary the last in}
\stick{that place to be expected\tsk a word I am}
\stick{ashamed to write\tsk yet must be written}
\stick{\tsk must be read;\tsk illegal\tsk uncanonical\tsk}
\stick{guess ten thousand guesses, multiplied}
\stick{into themselves\tsk rack\tsk torture your in-}
\stick{vention for ever, you’re where you was\tsk}
In short, I’ll tell it in the next chapter.\patch{\ls{CHAP}.}

\null\smallskip
\section{\lsss{CHAP}.\enspace XXVII.}

\dropcap{Z}{ounds!}\enspace\tsh\!\tsh\!\tsh\!\tsh\break
\tsh\!\tsh\!\tsh\!\tsh\!\tsk\break
\tsh\!\tsh\!\tsk\enspace Z\tsh ds!  cried \i{Phu}-\break
\stick{\i{tatorius}, partly to himself\tsk and yet high}
\stick{enough to be heard\tsk and what seemed}
\stick{odd, ’twas uttered in a construction of}
\stick{look, and in a tone of voice, somewhat}
\stick{between that of a man in amazement,}
and of one in bodily pain.

One or two who had very nice ears,\break
\stick{and could distinguish the expression and}
\stick{mixture of the two tones as plainly as a}
\stick{\i{third} or a \i{fifth}, or any other chord in}
\stick{musick\tsk were the most puzzled and per-}
\stick{plexed with it\tsk the \i{concord} was good in}
\stick{itself\tsk but then ’twas quite out of the}
\natch{key,}
\stick{key, and no way applicable to the subject}
\stck{started;\tsk so that with all their knowledge,}
\stick{they could not tell what in the world to}
make of it.

Others who knew nothing of musical\break
\stick{expression, and merely lent their ears to}
\stick{the plain import of the \i{word}, imagined}
\stick{that \i{Phutatorius}, who was somewhat of a}
\stick{cholerick spirit, was just going to snatch}
\stick{the cudgels out of \i{Didius}’s hands, in order}
\stick{to bemawl \i{Yorick} to some purpose\tsk and}
\stick{that the desperate monosyllable Z\tsk\!\tsk\!ds}
\stick{was the exordium to an oration, which,}
\stick{as they judged from the sample, presaged}
\stick{but a rough kind of handling of him;}
\stick{so that my uncle \toby's good nature felt}
\stick{a pang for what \i{Yorick} was about to}
\stick{undergo. But seeing \i{Phutatorius} stop}
\stick{short, without any attempt or desire to}
\stick{go on\tsk a third party began to suppose,}
\natch{that}
\stick{that it was no more than an involuntary}
\stick{respiration, casually forming itself into}
\stick{the shape of a twelve-penny oath\tsk with-}
out the sin or substance of one.

Others, and especially one or two who\break
\stick{sat next him, looked upon it on the con-}
\stick{trary,  as a real and substantial oath pro-}
\stick{pensly  formed against \i{Yorick}, to whom}
\stick{he was known to bear no good liking\tsk}
\stick{which said oath, as my father philoso-}
\stick{phized upon it, actually lay fretting and}
\stick{fuming at that very time in the upper}
\stick{regions of \i{Phutatorius}’s purtenance; and}
\stick{so was naturally, and according to the}
\stick{due course of things, first squeezed out}
\stick{by the sudden influx of blood, which}
\stick{was driven into the right ventricle of}
\stick{\i{Phutatorius}’s heart, by the stroke of sur-}
\stick{prize  which so strange a theory of}
preaching had excited.\patch{How}

How finely we argue upon mistaken\break facts!

There was not a soul busied in all\break
\stick{these various reasonings upon the mono-}
\stick{syllable which \i{Phutatorius} uttered,\tsk who}
\stick{did not take this for granted, proceeding}
\stick{upon it as from an axiom, namely, that}
\stick{\i{Phutatorius}’s mind was intent upon the}
\stick{subject of debate which was arising be-}
\stick{tween \i{Didius} and \i{Yorick}; and indeed as}
\stick{he looked first towards the one, and then}
\stick{towards the other, with the air of a man}
\stick{listening to what was going forwards,\tsk}
\stick{who would not have thought the same?}
\stick{But the truth was, that \i{Phutatorius} knew}
\stick{not one word or one syllable of what was}
\stick{passing\tsk but his whole thoughts and at-}
\stick{tention were taken up with a transaction}
\stick{which was going forwards at that very}
\stick{instant within the precincts of his own}
\natch{\i{Galli}-}
\stick{\i{Galligaskins}, and in a part of them, where}
\stick{of all others he stood most interested to}
\stick{watch accidents: So that notwithstand-}
\stick{ing he looked with all the attention in}
\stick{the world, and had gradually skrewed}
\stick{up every nerve and muscle in his face,}
\stick{to the utmost pitch the instrument would}
\stick{bear, in order, as it was thought, to give}
\stick{a sharp reply to \i{Yorick}, who sat over-}
\stick{against him\tsk Yet I say, was \i{Yorick} never}
\stick{once in any one domicile of \i{Phutatorius}’s}
\stick{brain\tsk but the true cause of his excla-}
mation lay at least a yard below.

This I will endeavour to explain to\break
you with all imaginable decency.

You must be informed then, that\break
\stick{\i{Gastripheres}, who had taken a turn into}
\stick{the kitchen a little before dinner, to see}
\stck{how things went on\tsk observing a wicker-}
\natch{basket}
\stick{basket of fine chesnuts standing upon the}
\stick{dresser, had ordered that a hundred or}
\stick{two of them might be roasted and sent}
\stick{in, as soon as dinner was over\tsk \i{Gastri}-}
\stick{\i{pheres} inforcing his orders about them,}
\stick{that \i{Didius}, but \i{Phutatorius} especially,}
were particularly fond of ’em.

About two minutes before the time\break
\stick{that my uncle \toby interrupted \i{Yorick}’s}
\stick{harangue\tsk \i{Gastripheres}’s chesnuts were}
\stck{brought in\tsk and as \i{Phutatorius}’s fondness}
\stick{for ’em, was uppermost in the waiter’s}
\stick{head, he laid them directly before \i{Phuta}-}
\stick{\i{torius}, wrapt up hot in a clean damask}
napkin.

Now whether it was physically impos-\break
\stick{sible, with half a dozen hands all thrust}
\stick{into the napkin at a time\tsk but that some}
\stick{one chesnut, of more life and rotundity}
\natch{than}
\stick{than the rest, must be put in motion\tsk}
\stick{it so fell out, however, that one was}
\stick{actually sent rolling off the table; and}
\stick{as \i{Phutatorius} sat straddling under\tsk it}
\stick{fell perpendicularly into that particular}
\stick{aperture of \i{Phutatorius}’s breeches, for}
\stick{which, to the shame and indelicacy of}
\stick{our language be it spoke, there is no}
\stick{chaste word throughout all \i{Johnson}’s}
\stick{dictionary\tsk let it suffice to say\tsk it was}
\stck{that particular aperture, which in all good}
\stick{societies, the laws of decorum do strictly}
\stick{require, like the temple of \i{Janus} (in}
peace at least) to be universally shut up.

The neglect of this punctilio in \i{Phuta}-
\stick{\i{torius} (which by the bye should be a}
\stick{warning to all mankind) had opened a}
door to this accident.\tsk\patch{\tsk Accident,}

\tsk Accident, I call it, in compliance to\break
\stick{a received mode of speaking,\tsk but in}
\stick{no opposition to the opinion either of}
\stick{\i{Acrites} or \i{Mythogeras} in this matter; I}
\stick{know they were both prepossessed and}
\stick{fully persuaded of it\tsk and are so to this}
\stick{hour, That there was nothing of acci-}
\stick{dent in the whole event\tsk but that the}
\stick{chesnut’s taking that particular course,}
\stick{and in a manner of its\sic accord\tsk and then}
\stick{falling with all its heat directly into that}
\stick{one particular place, and no other\tsh}
\stick{was a real judgment upon \i{Phutatorius},}
\stick{for that filthy and obscene treatise \i{de}}
\stick{\i{Concubinis retinendis}, which \i{Phutatorius}}
\stick{had published about twenty years ago\tsk}
\stick{and was that identical week going to give}
the world a second edition of.

It is not my business to dip my pen in\break
\stick{this controversy\tsh much undoubtedly}
\natch{may}
\stick{may be wrote on both sides of the}
\stick{question\tsk all that concerns me as an hi-}
\stick{storian, is to represent the mtter of}
\stick{fact, and render it credible to the reader,}
\stick{that the hiatus in \i{Phutatorius}’s breeches}
\stick{was sufficiently wide to receive the ches-}
\stick{nut;\tsk and that the chesnut, some how}
\stick{or other, did fall perpendicularly and}
\stick{piping hot into it, without \i{Phutatorius}’s}
perceiving it, or any one else at that time.

The genial warmth which the chesnut\break
\stick{imparted, was not undelectable for the}
\stick{first twenty or five and twenty seconds,\tsk}
\stick{and did no more than gently solicit \i{Phu}-}
\stick{\i{tatorius}’s attention towards the part:\tsk}
\stick{But the heat gradually increasing, and}
\stick{in a few seconds more getting beyond}
\stick{the point of all sober pleasure, and then}
\stick{advancing with all speed into the regions}
\stick{of pain,\tsk the soul of \i{Phutatorius}, toge-}
\natch{ther}
\stick{ther with all his ideas, his thoughts, his}
\stick{attention, his imagination, judgment,}
\stick{resolution, deliberation, ratiocination,}
\stick{memory, fancy, with ten battalions of}
\stick{animal spirits, all tumultuously crouded}
\stick{down, through different defiles and cir-}
\stick{cuits, to the place of danger, leaving all}
\stick{his upper regions, as you may imagine,}
as empty as my purse.

With the best intelligence which all\break
\stick{these messengers could bring him back,}
\stick{\i{Phutatorius} was not able to dive into the}
\stick{secret of what was going forwards below,}
\stck{nor could he make any kind of conjecture,}
\stick{what the devil was the matter with it:}
\stick{However, as he knew not what the true}
\stick{cause might turn out, he deemed it most}
\stick{prudent, in the situation he was in at}
\stick{present, to bear it, if possible, like a}
\stick{stoick; which, with the help of some}
\natch{wry}
\stick{wry faces and compursions of the mouth,}
\stick{he had certainly accomplished, had his}
\stick{imagination continued neuter\tsh but the}
\stick{sallies of the imagination are ungovern-}
\stick{able in things of this kind\tsk a thought}
\stick{instantly darted into his mind, that tho’}
\stick{the anguish had the sensation of glowing}
\stick{heat\tsk it might, notwithstanding that,}
\stick{be a bite as well as a burn; and if so,}
\stick{that possibly a \i{Newt} or an \i{Asker}, or some}
\stick{such detested reptile, had crept up, and}
\stick{was fastening his teeth\tsh the horrid idea}
\stick{of which, with a fresh glow of pain}
\stick{arising that instant from the chesnut,}
\stick{seized \i{Phutatorius} with a sudden panick,}
\stick{and in the first terrifying disorder of the}
\stick{passion it threw him, as it has done the}
\stick{best generals upon earth, quite off his}
\stick{guard:\tsk the effect of which was this,}
\stick{that he leapt incontinently up, uttering}
\stick{as he rose that interjection of surprise so}
\natch{much}
\stick{much discanted upon, with the aposio-}
\stick{pestick break after it, marked thus,}
\stick{Z\tsh ds\tsk which, though not strictly}
\stick{canonical, was still as little as any man}
\stick{could have said upon the occasion;\tsh}
\stick{and which, by the bye, whether cano-}
\stick{nical or not, \i{Phutatorius} could no more}
help than he could the cause of it.

Though this has taken up some time\break
\stick{in the narrative, it took up little more}
\stick{time in the transaction, than just to al-}
\stick{low time for \i{Phutatorius} to draw forth}
\stick{the chesnut, and throw it down with}
\stick{violence upon the floor\tsk and for \i{Yorick},}
\stick{to rise from his chair, and pick the}
chesnut up.

It is curious to observe the triumph\break
\stck{of slight incidents over the mind:\tsk What}
\stick{incredible weight they have in forming}
\natch{and}
\stick{and governing our opinions, both of men}
\stick{and things,\tsk that trifles light as air,}
\stick{shall waft a belief into the soul, and}
\stick{plant it so immoveably within it,\tsk that}
\stick{\i{Euclid}’s demonstrations, could they be}
\stick{brought to batter it in breach, should not}
all have power to overthrow it.

\i{Yorick}, I said, picked up the chesnut\break
\stick{which \i{Phutatorius}’s wrath had flung}
\stick{down\tsk the action was trifling\tsk I am}
\stick{ashamed to account for it\tsk he did it,}
\stick{for no reason, but that he thought the}
\stick{chesnut not a jot worse for the adven-}
\stick{ture\tsk and that he held a good chesnut}
\stick{worth stooping for.\tsk But this incident,}
\stick{trifling as it was, wrought differently in}
\stick{\i{Phutatorius}’s head: He considered this}
\stick{act of \i{Yorick}’s, in getting off his chair,}
\stick{and picking up the chesnut, as a plain}
\stick{acknowledgment in him, that the ches-}
\natch{nut}
\stick{nut was originally his,\tsk and in course,}
\stick{that it must have been the owner of the}
\stick{chesnut, and no one else, who could}
\stick{have played him such a prank with it:}
\stick{What greatly confirmed him in this opi-}
\stick{nion, was this, that the table being pa-}
\stick{rallelogramical and very narrow, it af-}
\stick{forded a fair opportunity for \i{Yorick},}
\stick{who sat directly over-against \i{Phutatorius},}
\stick{of slipping the chesnut in\tsk and conse-}
\stick{quently that he did it. The look of}
\stick{something more than suspicion, which}
\stick{\i{Phutatorius} cast full upon \i{Yorick} as these}
\stick{thoughts arose, too evidently spoke his}
\stck{opinion\tsk and as \i{Phutatorius} was naturally}
\stick{supposed to know more of the matter}
\stick{than any person besides, his opinion at}
\stick{once became the general one;\tsk and for}
\stick{a reason very different from any which}
\stick{have been yet given\tsk in a little time it}
was put out of all manner of dispute.\etp\patch{When}

When great or unexpected events fall\break
\stick{out upon the stage of this sublunary}
\stick{world\tsk the mind of man, which is an}
\stick{inquisitive kind of a substance, naturally}
\stick{takes a flight, behind the scenes, to see}
\stick{what is the cause and first spring of}
\stick{them\tsk The search was not long in this}
instance.

It was well known that \i{Yorick} had\break
\stck{never a good opinion of the treatise which}
\stick{\i{Phutatorius} had wrote \i{de Concubinis reti}-}
\stick{\i{nendis}, as a thing which he feared had}
\stick{done hurt in the world\tsk and ’twas easily} 
\stick{found out, that there was a mystical} 
\stick{meaning in \i{Yorick}’s prank\tsk and that his}
\stick{chucking the chesnut hot into \i{Phutato}-}
\stick{\i{rius}’s \hbox{\astw3\tsk\astw5}, was a sarcastical fling} 
\stick{at his book\tsk the doctrines of which,}
\stick{they said, had enflamed many an honest} 
man in the same place.\patch{This}

This conceit awaken’d \i{Somnolentus}\tsk\break
\stick{made \i{Agelastes} smile\tsk and if you can}
\stick{recollect the precise look and air of a}
\stick{man’s face intent in finding out a riddle\tsk}
\stick{it threw \i{Gastripheres}’s into that form\tsk} 
\stick{and in short was thought by many to be} 
a master-stroke of arch-wit.

This, as the reader has seen from one\break
\stick{end to the other, was as groundless as}
\stick{the dreams of philosophy: \i{Yorick}, no}
\stick{doubt, as \i{Shakespear} said of his ancestor}
\stick{\tsk\lqq\i{was a man of jest},\,” but it was tem-}
\stck{per’d with something which withheld him}
\stick{from that, and many other ungracious}
\stick{pranks, of which he as undeservedly}
\stick{bore the blame;\tsk but it was his misfor-}
\stick{tune all his life long to bear the impu-}
\stick{tation of saying and doing a thousand}
\stck{things of which (unless my esteem blinds} 
\stick{me) his nature was incapable. All I}
\natch{blame}
\stick{blame him for\tsk or rather, all I blame} 
\stick{and alternately like him for, was that}
\stick{singularity of his temper, which would} 
\stick{never suffer him to take pains to set a} 
\stick{story right with the world, however in} 
\stick{his power. In every ill usage of that} 
\stick{sort, he acted precisely as in the affair of} 
\stick{his lean horse\tsk he could have explained}
\stick{it to his honour, but his spirit was} 
\stick{above it; and besides he ever looked} 
\stick{upon the inventor, the propagator and} 
\stick{believer of an illiberal report alike so in-}
\stick{jurious to him,\tsk he could not stoop to} 
\stick{tell his story to them\tsk and so trusted to} 
time and truth to do it for him.

This heroic cast produced him incon-\break
\stick{veniences in many respects\tsk in the pre-}
\stick{sent, it was followed by the fixed resent-}
\stick{ment of \i{Phutatorius}, who, as \i{Yorick} had} 
\stick{just made an end of his chesnut, rose up}
\natch{from} 
\stick{from his chair a second time, to let him} 
\stick{know it\tsk which indeed he did with a} 
\stick{smile; saying only\tsk that he would en-}
deavour not to forget the obligation.

But you must mark and carefully se-\break
\stick{parate and distinguish these two things} 
in your mind.

\tsk The smile was for the company.

\tsk The threat was for \i{Yorick}.


\section{\chapstrut\lsss{CHAP}.\enspace XXVIII.}

\dropcap[findent=2pt,nindent=1pt]{\Usk{16pt}C}{an} you tell me, quoth \i{Phuta}-
\i{torius}, speaking to \i{Gastripheres}\break
\stick{who sat next to him,\tsk for one would}
\stick{not apply to a surgeon in so foolish an} 
\stick{affair,\tsk can you tell me, \i{Gastripheres},}
\natch{what} 
\stick{what is best to take out the fire?\tsk Ask}
\stick{\i{Eugenius}, said \i{Gastripheres}\tsk That greatly} 
\stick{depends, said \i{Eugenius}, pretending igno-}
\stick{rance of the adventure, upon the nature} 
\stick{of the part\tsk If it is a tender part, and}
\stick{a part which can conveniently be wrapt}
\stick{up\tsk It is both the one and the other,} 
\stick{replied \i{Phutatorius}, laying his hand as}
\stick{he spoke, with an emphatical nod of his} 
\stick{head upon the part in question, and lift-}
\stick{ing up his right leg at the same time to}
\stick{ease and ventilate it\tsh If that is the case,} 
\stick{said \i{Eugenius}, I would advise you, \i{Phu}-}
\stick{\i{tatorius}, not to tamper with it by any}
\stick{means; but if you will send to the next}
\stick{printer, and trust your cure to such a} 
\stick{simple thing as a soft sheet of paper just}
\stick{come off the press\tsk you need do nothing} 
\stick{more than twist it round\tsk The damp}
\stick{paper, quoth \i{Yorick} (who sat next to} 
\stick{his friend \i{Eugenius}) though I know it}
\natch{has} 
\stick{has a refreshing coolness in it\tsk yet I pre-}
\stick{sume is no more than the vehicle\tsk and} 
\stick{that the oil and lamp-black with which} 
\stick{the paper is so strongly impregnated,}
\stick{does the business\tsk Right, said \i{Eugenius},}
\stick{and is of any outward application I would} 
\stick{venture to recommend the most anodyne}
and safe.

Was it my case, said \i{Gastripheres}, as\break
\stick{the main thing is the oil and lamp-black,} 
\stick{I should spread them thick upon a rag,} 
\stick{and clap it on directly. That would}
\stick{make a very devil of it, replied \i{Yorick}\tsk}
\stick{And besides, added \i{Eugenius}, it would} 
\stick{not answer the intention, which is the} 
\stick{extreame neatness and elegance of the}
\stick{prescription, which the faculty hold to}
\stick{be half in half\tsk for consider, if the type}
\stick{is a very small one, (which it should be)}
\stick{the sanative particles, which come into}
\natch{contact} 
\stick{contact in this form, have the advantage}
\stck{of being spread so infinitely thin and with}
\stick{such a mathematical equality (fresh pa-}
\stick{ragraphs and large capitals excepted) as} 
\stick{no art or management of the spatula can} 
\stick{come up to. It falls out very luckily,}
\stick{replied \i{Phutatorius}, that the second edi-}
\stck{tion of my treatise \i{de Concubinis retinendis},}
\stick{is at this instant in the press\tsk You may} 
\stick{take any leaf of it, said \i{Eugenius}\tsk No}
\stick{matter which\tsk provided, quoth \i{Yorick},}
there is no bawdry in it\tsk

They are just now, replied \i{Phutatorius},\break
\stick{printing off the ninth chapter\tsk which is}
\stick{the last chapter but one in the book\tsk}
\stick{Pray what is the title of that chapter,} 
\stick{said \i{Yorick}, making a respectful bow to}
\stick{\i{Phutatorius} as he spoke\tsk I think, an-}
\stick{swered \i{Phutatorius}, ’tis that, \i{de re con}-}
\i{cubinariâ}.\patch{For}

For heaven’s sake keep out of that\break
chapter, quoth \i{Yorick}.

\tsk By all means\tsk added \i{Eugenius}.


\section{\chapstrut\lsss{CHAP}.\enspace XXIX.}

\dropcap[nindent=0pt]{\Usk{12pt}N}{ow},
quoth \i{Didius}, rising up,\break
and laying his right-hand with\break
\stick{his fingers spread upon his breast\tsk had}
\stick{such a blunder about a christian-name}
\stick{happened before the reformation\tsk (It}
\stck{happened the day before yesterday, quoth}
\stick{my uncle \toby to himself \!) and when}
\stick{baptism was administer’d in \i{Latin}\tsh}
\stick{(’Twas all in \i{English}, said my uncle \!)}
\stick{\tsk Many things might have coincided}
\stick{with it, and upon the authority of sundry} 
\stick{decreed cases, to have pronounced the}
\stick{baptism null, with a power of giving}
\natch{the}
\stick{the child a new name\tsk Had a priest, for}
\stck{instance, which was no uncommon thing,}
\stick{through ignorance of the \i{Latin} tongue,}
\stick{baptized a child of Tom-o’Stiles, \i{in}}
\stick{\i{nomino patriæ \et\ filia \et\ spiritum sanctos},}
\stick{\tsk the baptism was held null\tsk I beg your} 
\stick{pardon, replied \i{Kysarcius},\tsk in that case,}
\stick{as the mistake was only the \i{termina}-}
\stick{\i{tions}, the baptism was valid\tsk and to}
\stick{have rendered it null, the blunder of the} 
\stick{priest should have fallen upon the first}
\stick{syllable of each noun\tsk and not, as in}
your case, upon the last.\tsk

My father delighted in subtleties of\break
\stick{this kind, and listen’d with infinite at-}
tention.

\i{Gastripheres}, for example, continued\break
\stick{\i{Kysarcius}, baptizes a child of \i{John}}
\stick{\i{Stradling}’s, \i{in Gomine} gatris, \etc \et{\itshape c}.}
\natch{instead}
\stick{instead of \i{in Nomine} patris, \etc\!\tsk Is this}
\stick{a baptism? No,\tsk say the ablest cano-}
\stick{nists; inasmuch as the radix of each}
\stick{word is hereby torn up, and the sense}
\stick{and meaning of them removed and}
\stick{changed quite to another object; for}
\stick{\i{Gomine} does not signify a name, nor}
\stick{\i{gatris} a father\tsk What do they signify?}
\stick{said my uncle \i{Toby}\tsk Nothing at all\tsk}
\stick{quoth \i{Yorick}\tsk Ergo, such a baptism is}
\stick{null, said \i{Kysarcius}\tsk In course, answered}
\stick{\i{Yorick}, in a tone two parts jest and one}
part earnest\tsk

But in the case cited, continued \i{Kysar}-\break
\stick{\i{cius}, where \i{patrim} is put for \i{patris}, \i{filia}}
\stick{for \i{filij}, and so on\tsk as it is a fault only} 
\stick{in the declension, and the roots of the} 
\stick{words continue untouch’d, the inflexions}
\stick{of their branches, either this way or}
\stick{that, does not in any sort hinder the}
\natch{baptism,} 
\stick{baptism, inasmuch as the same sense}
\stick{continues in the words as before\tsk But} 
\stick{then, said \i{Didius}, the intention of the} 
\stck{priest’s pronouncing them grammatically,}
\stick{must have been proved to have gone} 
\stick{along with it\tsk Right, answered \i{Kysar}-}
\stick{\i{cius}; and of this, brother \i{Didius}, we}
\stick{have an instance in a decree of the de-}
\stick{cretals of Pope \i{Leo} the IIId.\tsk But my}
\stick{brother’s child, cried my uncle \toby,}
\stick{has nothing to do with the Pope\tsk ’tis}
\stick{the plain child of a Protestant gentle-}
\stick{man, christen’d \i{Tristram} against the wills}
\stick{and wishes both of its father and mother,}
and all who are a-kin to it\tsk

If the wills and wishes, said \i{Kysarcius}, 
\stick{interrupting my uncle \toby, of those}
\stick{only who stand related to Mr.\@ \i{Shandy}’s} 
\stick{child, were to have weight in this mat-}
\stick{ter, Mrs.\@ \i{Shandy}, of all people, has the}
\natch{least}
\stick{least to do in it.\tsk My uncle \toby lay’d}
\stick{down his pipe, and my father drew his} 
\stick{chair still closer to the table to hear the} 
\tight{conclusion of so strange an introduction.}

It has not only been a question,\break 
\stick{captain \i{Shandy}, amongst the \fnast\ best law-}
\stick{yers and civilians in this land, continued}
\stick{\i{Kysarcius}, \lqq\i{Whether the mother be of kin}}
\stick{\lqq \i{to her child},\,”\tsk but after much dispas-}
\stick{sionate enquiry and jactitation of the} 
\stick{arguments on all sides,\tsk it has been ad-}
\stick{judged for the negative,\tsk namely, \lqq\i{That}}
\stick{\lqq \i{the mother is not of kin to her child}\,\dagger.”}
\stick{My father instantly clapp’d his hand}
\stick{upon my uncle \toby's mouth, under co-}
\stick{lour of whispering in his ear\tsk the truth}
\stick{was, he was alarmed for \i{Lillabullero}\tsk}
\stick{and having a great desire to hear more}
\stick{\hfill of}
\vskip 2pt
\stick{\fontsize{8}{8}\selectfont \quad\fnast\ Vid. Swinburn on Testaments, Part 7. § 8.\hfill}
\vskip -3pt
\stick{\fontsize{8}{8}\selectfont \quad\dagger\ Vid. Brook Abridg. Tit. Administr. N.\@ 47.\hfill}

\newpage\noindent
\stick{of so curious an argument\tsk he begg’d my}
\stick{uncle \toby, for heaven’s sake, not to} 
\stick{disappoint him in it\tsk My uncle \toby} 
\stick{gave a nod\tsk resumed his pipe, and con-}
\stck{tenting himself with whistling \i{Lillabullero}}
\stick{inwardly\tsk \i{Kysarcius}, \i{Didius}, and \i{Trip}-}
\stick{\i{tolemus} went on with the discourse as} 
follows.

This determination, continued \i{Kysar}-
\stick{\i{cius}, how contrary soever it may seem}
\stick{to run to the stream of vulgar ideas,}
\stick{yet had reason strongly on its side; and} 
\stick{has been put out of all manner of dis-}
\stick{pute from the famous case, known com-}
\stick{monly by the name of the Duke of \i{Suf}-}
\stick{\i{folk}’s case:\tsk It is cited in \i{Brook}, said} 
\stick{\i{Triptolemus}\tsk And taken notice of by} 
\stick{Lord \i{Coke}, added \i{Didius}\tsk And you may} 
\stick{find it in \i{Swinburn} on Testaments, said}
\i{Kysarcius.}\patch{The}

The case, Mr.\@ \i{Shandy}, was this.

In the reign of \i{Edward} the Sixth,\break
\stick{\i{Charles} Duke of \i{Suffolk} having issue a} 
\stick{son by one venter, and a daughter by}
\stick{another venter, made his last will,}
\stick{wherein he devised goods to his son,}
\stick{and died; after whose death the son died} 
\stick{also\tsk but without will, without wife,} 
\stick{and without child\tsk his mother and his}
\stick{sister by the father’s side (for she was}
\stick{born of the former venter) then living.} 
\stick{The mother took the administration of}
\stick{her son’s goods, according to the statute} 
\stick{of the 21st of \i{Harry} the Eighth, whereby} 
\stick{it is enacted, That in case any person}
\stick{die intestate, the administration of his}
\stick{goods shall be committed to the next}
of kin.\patch{The}

The administration being thus (sur-\break
\stick{reptitiously) granted to the mother, the} 
\stick{sister by the father’s side commenced a}
\stick{suit before the Ecclesiastical Judge, al-}
\stick{ledging, 1st, That she herself was next} 
\stick{of kin; and 2dly, That the mother was}
\stick{not of kin at all to the party deceased;}
\stick{and therefore pray’d the court, that the}
\stick{administration granted to the mother}
\stick{might be revoked, and be committed}
\stick{unto her, as next of kin to the deceased,} 
by force of the said statute.

Hereupon, as it was a great cause,\break
\stick{and much depending upon its issue\tsk} 
\stick{and many causes of great property likely}
\stick{to be decided in times to come, by the} 
\stick{precedent to be then made\tsk the most}
\stick{learned, as well in the laws of this realm,}
\stick{as in the civil law, were consulted toge-}
\stick{ther, whether the mother was of kin to}
\natch{her} 
\stick{her son, or no. \tsk Whereunto not only} 
\stick{the temporal lawyers\tsk but the church-}
\stick{lawyers\tsk the juris-consulti\tsk the juris-}
\stick{prudentes\tsk the civilians\tsk the advocates}
\stick{\tsk the commissaries\tsk the judges of the}
\stick{consistory and prerogative courts of \i{Can}-}
\stick{\i{terbury} and \i{York}, with the master of} 
\stick{the faculties, were all unanimously of} 
\stick{opinion, That the mother was not of \fnast}
kin to her child\tsk

And what said the Duchess of \i{Suffolk}\break
to it? said my uncle \toby.

The unexpectedness of my uncle\break
\stick{\toby's question, confounded \i{Kysarcius}}
\stick{more than the ablest advocate\tsh He}
\stick{stopp’d a full minute, looking in my}
\stick{uncle \toby's face without replying\tsk}
\stick{\hfill and}

\begingroup\footnotesize\indent\fnast\
Mater non numeratur \vrule width0pt depth 4pt inter consanguineos.\break
Bald.\@ in ult.\@ C.\@ de Verb.\@ signific.\par\endgroup

\newpage\noindent
\stick{and in that single minute \i{Triptolemus} put}
by him, and took the lead as follows.

’Tis a ground and principle in the\break
\stick{law, said \i{Triptolemus}, that things do not} 
\stick{ascend, but descend in it; and I make no}
\stick{doubt ’tis for this cause, that however}
\stick{true it is, that the child may be of the} 
\stick{blood or seed of its parents\tsk that the}
\stick{parents, nevertheless, are not of the}
\stick{blood and seed of it; inasmuch as the}
\stick{parents are not begot by the child, but}
\stick{the child by the parents\tsk For so they} 
\stick{write, \i{Liberi sunt de sanguine patris \et}}
\stick{\i{matris, sed pater et mater non sunt de}}
\i{sanguine liberorum}.

\tsk But this, \i{Triptolemus}, cried \i{Didius},\break
\stick{proves too much\tsk for from this autho-}
\stick{rity cited it would follow, not only}
\stick{what indeed is granted on all sides,}
\natch{that}
\stick{that the mother is not of kin to her}
\stick{child\tsk but the father likewise\tsh It is}
\stick{held, said \i{Triptolemus}, the better opi-}
\stick{nion; because the father, the mother,}
\stick{and the child, though they be three}
\stick{persons, yet are they but (\i{una caro} \fnast)}
\stick{one flesh; and consequently no degree}
\stick{of kindred\tsk or any method of acquiring}
\stick{one \i{in nature}\tsk There you push the ar-}
\stick{gument again too far, cried \i{Didius}\tsk}
\stick{for there is no prohibition \i{in nature},}
\stick{though there is in the levitical law,\tsk}
\stick{but that a man may beget a child upon}
\stick{his grandmother\tsh in which case, sup-}
\stick{posing the issue a daughter, she would} 
\stick{stand in relation both of\tsh But who}
\stick{ever thought, cried \i{Kysarcius}, of laying} 
\stick{with his grandmother?\tsh The young}
\stick{gentleman, replied \i{Yorick}, whom \i{Selden}}
\stick{speaks of\tsk who not only thought of it,}
\stick{\hfill but}\etp\break
\stick{\footnotesize\enspace\fnast\ Vide Brook Abridg. tit. Administr. N.\@ 47.\hfill}

\newpage\noindent
\stick{but justified his intention to his father}
\stick{by the argument drawn from the law}
\stick{of retaliation\tsh \lqq You lay’d, Sir,}
\stick{\lqq with my mother, said the lad\tsk why}
\stick{\lqq may not I lay with yours?”\tsk ’Tis}
\stick{the \i{Argumentum commune}, added \i{Yorick.}}
\stick{\tsk ’Tis as good, replied \i{Eugenius}, taking} 
down his hat, as they deserve.

The company broke up\tsh


\section{\chapstrut\lsss{CHAP}.\enspace XXX.}

\dropcap[findent=0pt]{\Usk{16pt}A}{nd} pray, said my uncle \toby,\break
leaning upon \i{Yorick}, as he\break
\stick{and my father were helping him leisurely} 
\stick{down the stairs\tsk don’t be terrified, ma-}
\stick{dam, this stair-case conversation is not} 
\stick{so long as the last\tsk And pray, \i{Yorick},} 
\stick{said my uncle \toby, which way is this}
\natch{said}
\stick{said\sic affair of \i{Tristram} at length settled}
\stck{by these learned men?  Very satisfactorily,}
\stick{replied \i{Yorick}; no mortal, Sir, has any}
\stick{concern with it\tsk for Mrs.\@ \i{Shandy} the}
\stick{mother is nothing at all akin to him\tsk}
\stick{and as the mother’s is the surest side\tsk}
\stick{Mr.\@ \i{Shandy}, in course, is still less than}
\stick{nothing\tsk In short, he is not as much}
akin to him, Sir, as I am\tsk

\tsk That may well be, said my father,\break
shaking his head.

\tsk Let the learned say what they will,\break
\stick{there must certainly, quoth my uncle}
\stick{\toby, have been some sort of consan-}
\stick{guinity betwixt the duchess of \i{Suffolk}}
and her son\tsk

The vulgar are of the same opinion,\break
quoth \i{Yorick}, to this hour.\patch{\ls{CHAP}.}


\section{\chapstrutt\lsss{CHAP}.\enspace XXXI.}

\dropcap{T}{hough} my father was hugely\break
tickled with the subtleties of these\break
\stick{learned discourses\tsk ’twas still but like}
\stick{the anointing of a broken bone\tsk The}
\stick{moment he got home, the weight of}
\stick{his afflictions returned upon him but so} 
\stick{much the heavier, as is ever the case}
\stick{when the staff we lean on slips from}
\stick{under us\tsk He became pensive\tsk walked} 
\stick{frequently forth to the fish-pond\tsk let}
\stick{down one loop of his hat\tsk sigh’d often}
\stick{\tsk forbore to snap\tsk and, as the hasty}
\stick{sparks of temper, which occasion snap-}
\stick{ping, so much assist perspiration and}
\stick{digestion, as \i{Hippocrates} tells us\tsk he had} 
\stick{certainly fallen ill with the extinction of} 
\stick{them, had not his thoughts been criti-}
\natch{cally} 
\stick{cally drawn off, and his health rescued} 
\stick{by a fresh train of disquietudes left him,} 
\stick{with a legacy of a thousand pounds by}
my aunt \i{Dinah}\tsk 

My father had scarce read the letter,\break
\stick{when taking the thing by the right end,}
\stick{he instantly began to plague and puzzle}
\stick{his head how to lay it out mostly to}
\stick{the honour of his family.\tsk A hundred}
\stick{and fifty odd projects took possession}
\stick{of his brains by turns\tsk he would do}
\stick{this, and that, and\sic to’ther\tsk He would}
\stick{go to \i{Rome}\tsk he would go to law\tsk he}
\stick{would buy stock\tsk he would buy \i{John}}
\stick{\i{Hobson}’s farm\tsk he would new fore-front}
\stick{his house, and add a new wing to make}
\stick{it even\tsk There was a fine water-mill on}
this side, and he would build a wind-mill on the
other side of the river in full view to answer it\tsk But above
all
\natch{things} things in the world, he would inclose the great \i{Ox-moor},
and send out my brother \i{Bobby} immediately upon his
travels.

But as the sum was \i{finite}, and conse-\break
\stick{quently could not do every thing\tsk and} 
\stick{in truth very few of these to any pur-}
\stick{pose,\tsk of all the projects which offered}
\stick{themselves upon this occasion, the two}
\stick{last seemed to make the deepest impres-}
\stick{sion; and he would infallibly have de-}
\stick{termined upon both at once, but for}
\stick{the small inconvenience hinted at above,}
\stick{which absolutely put him under a ne-}
\stick{cessity of deciding in favour either of}
the one or the other.


This was not altogether so easy to be\break
\stick{done; for though ’tis certain my father} 
\stick{had long before set his heart upon this} 
\stick{necessary part of my brother’s education,}
\natch{and} 
\stick{and like a prudent man had actually}
\stick{determined to carry it into execution,}
\stick{with the first money that returned from}
\stick{the second creation of actions in the} 
\stick{\i{Missisippi}-scheme, in which he was an}
\stick{adventurer\tsk yet the \i{Ox-moor}, which was} 
\stick{a fine, large, whinny, undrained, unim-}
\stck{proved common, belonging to the \i{Shandy}-}
\stick{estate, had almost as old a claim upon}
\stick{him: He had long and affectionately set}
\stck{his heart upon turning it likewise to some}
account.

\tight{But having never hitherto been pressed}\break
\stick{with such a conjuncture of things, as}
\stick{made it necessary to settle either the}
\stick{priority or justice of their claims,\tsk like} 
\stick{a wise man he had refrained entering} 
\stick{into any nice or critical examination} 
\stick{about them: So that upon the dismission} 
\stick{of every other project at this crisis,\tsh}
\natch{the} 
\stick{the two old projects, the \s{Ox-moor} and} 
\stick{my \s{brother}, divided him again; and so}
\stick{equal a match were they for each other,}
\stick{as to become the occasion of no small} 
\stick{contest in the old gentleman’s mind,\tsk} 
\stick{which of the two should be set o’going} 
first.

\tsk People may laugh as they will\tsh\break
but the case was this.

It had ever been the custom of the\break
\stick{family, and by length of time was al-}
\stick{most become a matter of common right,}
\stick{that the eldest son of it should have free} 
\stick{ingress, egress, and regress into foreign}
\stick{parts before marriage,\tsk not only for the}
\stick{sake of bettering his own private parts,} 
\stick{by the benefit of exercise and change of}
\stick{so much air\tsk but simply for the mere}
\stick{delectation of his fancy, by the feather}
\natch{put}
\stick{put into his cap, of having been abroad}
\stick{\tsk \i{tantum valet}, my father would say,}
\i{quantum sonat}.

Now as this was a reasonable, and in\break
\stick{course a most christian indulgence\tsk to}
\stick{deprive him of it, without why or}
\stick{wherefore,\tsk and thereby make an ex-}
\stick{ample of him, as the first \i{Shandy} un-}
\stick{whirl’d about \i{Europe} in a post-chaise,}
\stick{and only because he was a heavy lad\tsk}
\stick{would be using him ten times worse than} 
a \i{Turk}.

On the other hand, the case of the\break
\i{Ox-moor} was full as hard.

Exclusive of the original purchase-\break
\stick{money, which was eight hundred pounds}
\stick{\tsk it had cost the family eight hundred}
\stick{pounds more in a law-suit about fifteen}
\natch{years}
\stick{years before\tsk besides the Lord knows}
what trouble and vexation.

It had been moreover in possession of\break
\stick{the \i{Shandy}-family ever since the middle} 
\stick{of the last century; and though it lay} 
\stick{full in view before the house, bounded} 
\stick{on one extremity by the water-mill,} 
\stick{and on the other by the projected wind-}
\stick{mill spoken of above,\tsk and for all these} 
\stick{reasons seemed to have the fairest title} 
\stick{of any part of the estate to the care and}
\stick{protection of the family\tsk yet by an un-}
\stick{accountable fatality, common to men,} 
\stick{as well as the ground they tread on,\tsk it}
\stick{had all along most shamefully been over-}
\stick{look’d; and to speak the truth of it,}
\stick{had suffered so much by it, that it} 
\stick{would have made any man’s heart have}
\stick{bled \i{(\sic Obadiah} said ) who understood} 
\stick{the value of the land, to have rode over}
\natch{it,}
\stick{it, and only seen the condition it was} 
in.

However, as neither the purchasing\break
\stick{this tract of ground\tsk nor indeed the} 
\stick{placing of it where it lay, were either of}
\stick{them, properly speaking, of my father’s}
\stick{doing\tsk he had never thought himself} 
\stick{any way concerned in the affair\tsk till the}
\stick{fifteen years before, when the breaking}
\stick{out of that cursed law-suit mentioned}
\stick{above (and which had arose about its}
\stick{boundaries)\tsk which being altogether my}
\stick{father’s own act and deed, it naturally}
\stick{awakened every other argument in its}
\stick{favour; and upon summing them all up}
\stick{together, he saw, not merely in interest,}
\stick{but in honour, he was bound to do}
\stick{something for it\tsk and that now or never}
was the time.\patch{I think}

I think there must certainly have been\break
\stick{a mixture of ill-luck in it, that the rea-}
\stick{sons on both sides should happen to be}
\stick{so equally balanced by each other; for}
\stick{though my father weigh’d them in all}
\stick{humours and conditions\tsk spent many an}
\stick{anxious hour in the most profound and}
\stick{abstracted meditation upon what was}
\stick{best to be done\tsh reading books of}
\stick{farming one day\tsk books of travels an-}
\stick{other\tsk laying aside all passion whatever}
\stick{\tsk viewing the arguments on both sides}
\stick{in all their lights and circumstances\tsk}
\stick{communing every day with my uncle}
\stick{\i{Toby}\tsk arguing with \i{Yorick}, and talking}
\stick{over the whole affair of the \i{Ox-moor}}
\stick{with \i{Obadiah}\tsk yet nothing in all that} 
\stick{time appeared so strongly in behalf of}
\stick{the one, which was not either strictly}
\stick{applicable to the other, or at least so}
\stick{far counterbalanced by some considera-}
\natch{tion} 
\stick{tion of equal weight, as to keep the}
scales even.

For to be sure, with proper helps,\break
\stick{and in the hands of some people, tho’}
\stick{the \i{Ox-moor} would undoubtedly have}
\stick{made a different appearance in the world}
\stick{from what it did, or ever would do in} 
\stick{the condition it lay\tsk yet every tittle of} 
\stick{this was true, with regard to my bro-}
\stick{ther \i{Bobby}\tsk let \i{Obadiah} say what he}
would.\tsh

In point of interest\tsk the contest, I\break
\stick{own, at first sight, did not appear so}
\stick{undecisive betwixt them; for whenever}
\stick{my father took pen and ink in hand,}
\stick{and set about calculating the simple ex-}
\stick{pence of paring and burning, and fence-}
\stick{ing in the \i{Ox-moor}, \&c.\@ \&c.\tsk with the}
\stick{certain profit it would bring him in}
\natch{return} 
\stick{return\tsk the latter turned out so prodi-}
\stick{giously in his way of working the ac-}
\stick{count, that you would have sworn the}
\stick{\i{Ox-moor} would have carried all before} 
\stick{it. For it was plain he should reap} 
\stick{a hundred lasts of rape, at twenty}
\stick{pounds a last, the very first year\tsk be-}
\stick{sides an excellent crop of wheat the} 
\stick{year following\tsk and the year after that,} 
\stick{to speak within bounds, a hundred\tsh} 
\stick{but, in all likelihood, a hundred and} 
\stick{fifty\tsk if not two hundred quarters of}
\stick{pease and beans\tsk besides potatoes with-}
\stick{out end\tsk But then, to think he was all} 
\stick{this while breeding up my brother like}
\stick{a hog to eat them\tsk knocked all on the}
\stick{head again, and generally left the old}
\stick{gentleman in such a state of suspence\tsk} 
\stick{that, as he often declared to my uncle} 
\stick{\i{Toby}\tsk he knew no more than his heels} 
what to do.\etp\patch{No}

No body, but he who has felt it, can\break
\stick{conceive what a plaguing thing it is to}
\stick{have a man’s mind torn asunder by two}
\stick{projects of equal strength, both obsti-}
\stick{nately pulling in a contrary direction at}
\stick{the same time: For to say nothing of}
\stick{the havock, which by a certain conse-}
\stick{quence is unavoidably made by it all}
\stick{over the finer system of the nerves,}
\stick{which you know convey the animal}
\stick{spirits and more subtle juices from the}
\stick{heart to the head, and so on\tsk\!\tsk It is}
\stick{not to be told in what a degree such a}
\stick{wayward kind of friction works upon} 
\stick{the more gross and solid parts, wasting} 
\stick{the fat and impairing the strength of a}
\stick{man every time as it goes backwards}
and forwards.

My father had certainly sunk under\break
\stick{this evil, as certainly as he had done}
\natch{under}
\stick{under that of my \s{christian name}\tsk}
\stick{had he not been rescued out of it as}
\stick{he was out of that, by a fresh evil\tsk}
\stick{the misfortune of my brother \i{Bobby}’s}
death.

What is the life of man! Is it not to\break
\stck{shift from side to side?\tsk from sorrow to} 
\stick{sorrow?\tsk\!\!\tsk to button up one cause of}
vexation!\tsk and unbutton another!


\section{\chapstrut\lsss{CHAP}.\enspace XXXII.}

\dropcap{F}{rom} this moment I am to be\break
considered as heir-apparent to the\break
\stick{\i{Shandy} family\tsk and it is from this point} 
\stick{properly, that the story of my \s{Life}}
\stick{and my \s{Opinions} sets out; with all}
\stick{my hurry and precipitation, I have but}
\natch{been}
\stick{been clearing the ground to raise the}
\stick{building\tsh and such a building do I}
\stick{foresee it will turn out, as never was}
\stick{planned, and as never was executed since}
\stick{\i{Adam}. In less than five minutes I shall}
\stick{have thrown my pen into the fire, and} 
\stick{the little drop of thick ink which is left}
\stick{remaining at the bottom of my ink-}
\stick{horn, after it\tsk I have but half a score}
\stick{things to do in the time\tsh I have a}
\stick{thing to name\tsk a thing to lament\tsk a}
\stick{thing to hope\tsk a thing to promise, and}
\stick{a thing to threaten\tsk I have a thing to}
\stick{suppose\tsk a thing to declare\tsk a thing}
\stick{to conceal\tsk a thing to chuse, and a}
\stick{thing to pray for.\tsk This chapter, there-}
\stick{fore, I \i{name} the chapter of \s{Things}\tsk}
\stick{and my next chapter to it, that is, the}
\stick{first chapter of my next volume, if I live,}
\stick{shall be my chapter upon \s{whiskers},}
\natch{in} 
\stick{in order to keep up some sort of con-}
nection in my works.

The thing I lament is, that things\break
\stick{have crowded in so thick upon me,}
\stick{that I have not been able to get into}
\stick{that part of my work, towards which,}
\stick{I have all the way, looked forwards,}
\stick{with so much earnest desire; and that is}
\stick{the campaigns, but especially the amours}
\stick{of my uncle \toby, the events of which}
\stick{are of so singular a nature, and so Cer-}
\stick{vantick a cast, that if I can so manage}
\stick{it, as to convey but the same impressions}
\stick{to every other brain, which the occur-}
\stick{rences themselves excite in my own\tsh}
\stick{I will answer for it the book shall make}
\stick{its way in the world, much better}
\stick{than its master has done before it\tsh}
\stick{Oh \i{Tristram!} \i{Tristram!} can this but}
\stick{be once brought about\tsh the credit,}
\natch{which}
\stick{which will attend thee as an author,}
\stick{\tight{shall counterbalance the many evils which}}
\stick{have befallen thee as a man\tsk thou wilt}
\stick{feast upon the one\tsk when thou hast}
\stick{lost all sense and remembrance of the}
other!\tsh

No wonder I itch so much as I do,\break 
\stick{to get at these amours\tsk They are the}
\stick{choicest morsel of my whole story! and}
\stick{when I do get at ’em\tsk assure yourselves,}
\stick{good folks,\tsk (nor do I value whose}
\stick{squeamish stomach takes offence at it)}
\stick{I shall not be at all nice in the choice of}
\stick{my words;\tsk\!\tsk and that’s the thing I}
\stick{have to \i{declare}.\tsk I shall never get all} 
\stick{through in five minutes, that I fear\tsk} 
\stick{and the thing I \i{hope} is, that your wor-}
\stick{ships and reverences are not offended\tsk}
\stick{if you are, depend upon’t I’ll give you}
\stick{something, my good gentry, next year,}
\natch{to} 
\stick{to be offended at\tsk\!\tsk that’s my dear}
\stick{\i{Jenny}’s way\tsk but who my \i{Jenny} is\tsk}
\stick{and which is the right and which the} 
\stick{wrong end of a woman, is the thing} 
\stick{to be \i{concealed}\tsk it shall be told you} 
\stick{the next chapter but one, to my chapter} 
\stick{of button-holes,\tsk and not one chapter}
before.


And now that you have just got to\break
\stick{the end of these four volumes\tsk\!\tsk the}
\stick{thing I have to \i{ask} is, how you feel}
\stick{your heads? my own akes dismally\tsk}
\stick{as for your healths, I know, they are}
\stick{much better\tsk\!\tsk True \i{Shandeism}, think}
\stick{what you will against it, opens the}
\stick{heart and lungs, and like all those af-}
\stick{fections which partake of its nature, it} 
\stick{forces the blood and other vital fluids of}
\stick{the body to run freely thro’ its channels,}
\natch{and} 
\stick{and makes the wheel of life run long} 
and chearfully round.

Was I left like \i{Sancho Pança}, to\break
\stick{chuse my kingdom, it should not be}
\stick{maritime\tsk or a kingdom of blacks to}
\stick{make a penny of\tsh no, it should}
\stick{be a kingdom of hearty laughing sub-}
\stick{jects: And as the bilious and more sa-}
\stick{turnine passions, by creating disorders}
\stick{in the blood and humours, have as bad}
\stick{an influence, I see, upon the body politick}
\stick{as body natural\tsk and as nothing but a}
\stick{habit of virtue can fully govern those}
\stick{passions, and subject them to reason\tsk I}
\stick{should add to my prayer\tsk that God}
\stick{would give my subjects grace to be as} 
\stick{\s{wise} as they were \s{merry}; and then}
\stick{should I be the happiest monarch, and} 
they the happiest people under heaven\tsk\patch{And}

And so, with this moral for the pre-\break
\stick{sent, may it please your worships and}
\stick{your reverences, I take my leave of you}
\stick{till this time twelve-month, when (unless}
\stick{this vile cough kills me in the mean}
\stick{time) I’ll have another pluck at your}
\stick{beards, and lay open a story to the}
world you little dream of.

\vskip 72pt
\centerline{\large\itshape\lsv{52}{F I N I S}.}
\end{document}
